### 2023-05-23 article/Fig_tree.tex
Formulas + BD optimiser

\documentclass{standalone}
\usepackage{amsmath,amsthm,amssymb,graphicx,mathtools,tikz,hyperref,amsthm}
\usepackage{tikz}
\usepackage{standalone}
\usepackage{caption}
\usetikzlibrary{positioning}
\tikzset{main node/.style={fill=none,draw=none,minimum size=.5cm,inner sep=0pt}}
\tikzset{label node/.style={color=black,fill=none,draw=none,minimum size=.2cm,inner sep=0pt}}
\usepackage[final]{pdfpages}

\definecolor{tip}{rgb}{0.84,0.1,0.11}
\definecolor{branch}{rgb}{.48, .2, .58}
\definecolor{node}{rgb}{.17, .48, .71}
\definecolor{T}{rgb}{.65,.38,.1}

\begin{document}

  \begin{tikzpicture}[node distance = 1cm and 1cm, line width=2pt, edge/.style = {-}]  
    \node (Tree) {
        \begin{tikzpicture}[line width=2pt, edge/.style = {-}]

            \node[draw, circle, color=node, scale=.5] (root) {$0$}; 
            \coordinate[above=.2cm of root] (t_root);

            \coordinate[right=1cm of root] (root_n3);    
            \node[draw, circle, color=node, scale=.5] (n3) [below = 1cm of root_n3] {$01$};          
            \draw[edge, color=node] (root) to (root_n3);
            \draw[edge, color=branch] (root_n3) to node[right, color=branch, scale=.5] {} (n3);

            \coordinate[left=.5cm of n3] (n3_t4);    
            \node[draw, circle, color=tip, scale=.5] (t4) [below = 1cm of n3_t4] {$010$};          
            \draw[edge, color=node] (n3) to (n3_t4);
            \draw[edge, color=branch] (n3_t4) to node[right, color=branch, scale=.5] {} (t4);
            \coordinate[right=.5cm of n3] (n3_t5);    
            \node[draw, circle, color=tip, scale=.5] (t5) [below = .8cm of n3_t5] {$011$};          
            \draw[edge, color=node] (n3) to (n3_t5);
            \draw[edge, color=branch] (n3_t5) to node[right, color=branch, scale=.5] {} (t5);

            \coordinate[left=1cm of root] (root_n2);    
            \node[draw, circle, color=node, scale=.5] (n2) [below = .4cm of root_n2] {$00$};          
            \draw[edge, color=node] (root) to (root_n2);
            \draw[edge, color=branch] (root_n2) to node[right, color=branch, scale=.5] {} (n2);

            \coordinate[left=1cm of n2] (n2_n4);    
            \node[draw, circle, color=node, scale=.45] (n4) [below = 1.2cm of n2_n4] {$000$};          
            \draw[edge, color=node] (n2) to (n2_n4);
            \draw[edge, color=branch] (n2_n4) to node[right, color=branch, scale=.5] {} (n4);
            \coordinate[right=.5cm of n2] (n2_t3);    
            \node[draw, circle, color=tip, scale=.5] (t3) [below = .1cm of n2_t3] {$001$};          
            \draw[edge, color=node] (n2) to (n2_t3);
            \draw[edge, color=branch] (n2_t3) to node[right, color=branch, scale=.5] {} (t3);

            \coordinate[left=.5cm of n4] (n4_t1);    
            \node[draw, circle, color=tip, scale=.45] (t1) [below = 1.6cm of n4_t1] {$0000$};          
            \draw[edge, color=node] (n4) to (n4_t1);
            \draw[edge, color=branch] (n4_t1) to node[right, color=branch, scale=.5] {} (t1);
            \coordinate[right=.5cm of n4] (n4_t2);    
            \node[draw, circle, color=tip, scale=.45] (t2) [below = .8cm of n4_t2] {$0001$};          
            \draw[edge, color=node] (n4) to (n4_t2);
            \draw[edge, color=branch] (n4_t2) to node[right, color=branch, scale=.5] {} (t2);

                \coordinate[left=3.5cm of root] (then);
    			\coordinate[below=3.85cm of then] (now);

            \draw[edge, color=gray, <-, line width=.5] (now) to node[left, color=gray, scale=.75] {} (then);

            \coordinate[left=.12cm of then] (c_then); 
            \coordinate[left=.12cm of now] (c_now); 
            \coordinate[right=6cm of c_now] (c_now_c);

            \node[color=gray, scale=.75] () [left = 0cm of c_then] {$t=t_0=0$};
            \draw[edge, color=gray, dotted, line width=.7] (c_then) to (root_n2);

            \coordinate[left=1.425cm of n2_n4] (c_n2); 
            \node[color=gray, scale=.75] () [left = 0cm of c_n2] {$t=t_{00}$};
            \draw[edge, color=gray, dotted, line width=.7] (c_n2) to (n2_n4);

            \coordinate[left=3.09cm of t3] (c_01); 
            \node[color=gray, scale=.75] () [left = 0cm of c_01] {$t=t_{001}$};
            \draw[edge, color=gray, dotted, line width=.7] (c_01) to (t3);

            \coordinate[left=.7cm of n4_t1] (c_n4); 
            \node[color=gray, scale=.75] () [left = 0cm of c_n4] {$t=t_{000}$};
            \draw[edge, color=gray, dotted, line width=.7] (c_n4) to (n4_t1);

            \coordinate[left=4.26cm of n3_t4] (c_n3); 
            \node[color=gray, scale=.75] () [left = 0cm of c_n3] {$t=t_{01}$};
            \draw[edge, color=gray, dotted, line width=.7] (c_n3) to (n3_t4);

            \coordinate[left=5.43cm of t5] (c_e); 
            \node[color=gray, scale=.75] () [left = 0cm of c_e] {$t=t_{011}$};
            \draw[edge, color=gray, dotted, line width=.7] (c_e) to (t5);

            \coordinate[left=4.02cm of t4] (c_000); 
            \node[color=gray, scale=.75] () [left = 0cm of c_000] {$t=t_{010}$};
            \draw[edge, color=gray, dotted, line width=.7] (c_000) to (t4);

            \coordinate[left=1.88cm of t2] (c_00); 
            \node[color=gray, scale=.75] () [left = 0cm of c_00] {$t=t_{0001}$};
            \draw[edge, color=gray, dotted, line width=.7] (c_00) to (t2);

            \node[color=gray, scale=.75] () [left = 0cm of c_now] {$t=t_{0000}=T$};

            \draw[edge, color=gray, dotted, line width=.7] (c_now) to node[midway, color=gray, fill=white, scale=.75, opacity=1, text opacity=1] {end of sampling period} (c_now_c);

        \end{tikzpicture}
    };
  \end{tikzpicture}

 \end{document}

### 2023-05-23 article/article.tex
Formulas + BD optimiser

\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,mathrsfs}

\usepackage{standalone}
\usepackage{caption}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\usetikzlibrary{backgrounds}
\usepackage{tikzsymbols}
\usepackage{hieroglf}
\usepackage{nameref}
\usepackage{algpseudocode}
\usepackage{hyperref}
\usepackage{natbib}

\definecolor{tip}{rgb}{0.84,0.1,0.11}
\definecolor{branch}{rgb}{.48, .2, .58}
\definecolor{node}{rgb}{.17, .48, .71}
\definecolor{T}{rgb}{.65,.38,.1}

%opening
\title{MTBD models with partner notification}
\author{Anna Zhukova}

\begin{document}

\maketitle

%\begin{abstract}
%
%\end{abstract}

\section{Introduction}
The interaction of epidemiological and evolutionary processes leaves a footprint in pathogen genomes. Phylodynamics leverages this footprint to estimate epidemiological parameters~\cite{Grenfell2004a,Volz2013}. It relies on models that bridge the gap between traditional epidemiology and sequence data by estimating such parameters as the basic reproduction number, $R_0$, from topology and branch lengths of pathogen phylogenies (i.e. genealogies of the pathogen population, approximating the transmission trees) combined with metadata on the samples. This is particularly useful for emerging epidemics, when not enough data (e.g. incidence curves) might be gathered for accurate estimations with classic epidemiological methods, while using genetic data for phylodynamic estimations can provide valuable insights and help prevent epidemic spreads (e.g. accurate estimation of the infectious period is crucial for adjusting health policies for self-isolation).

Phylodynamic models can be classified into two main families:  coalescent~\cite{Volz2009a,Drummond2005,Pybus2000a} and birth-death (BD)~\cite{Kendall1948,Maddison2007,Stadler2009,Stadler2010}. Coalescent models are often preferred for estimating deterministic population dynamics, however for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\cite{Macpherson2021}. In the classic BD model with incomplete sampling (BDS~\cite{Stadler2009}), births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g. due to healing, self-isolation, starting a treatment, or death, modelled with a constant removal rate).

The epidemic spread and detection can however be non-homogenous and depend on different factors, including governmental health policies (e.g. quarantine measures or development of pathogen detection tests), pathogen evolution over time (e.g. some SARS-CoV-2 variants are more transmissible than other),  differences between host individuals (e.g. due to their immune system particularities or to their behaviour).

To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extention of the classical BDS model was developed by Stadler \textit{et al.}~\cite{Stadler2013a}. MTBD framework allows for different types of individual states (and changes between them, modelled with state change rates). These models are phylodynamic analogies of compartmental models in classical epidemiology (e.g. SIR, Susceptible-Infectious-Removed).  Examples of the MTBD family models include the birth-death exposed-infectious (BDEI) model, which was designed for pathogens featuring an incubation period between the moments of infection and of becoming infectious (e.g. Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen more than the others~\cite{Stadler2014}.

To allow for heterogeneity over time, Stadler~\textit{et~al.}~\cite{Stadler2013} developed one-state Bayesian birth-death skyline plot (BDSKY) that divides the time into intervals and allows for different piecewise constant rates on them. K\"{u}hnert \textit{et al.}~\cite{Kuhnert2016} combined the MTBD model with the BDSKY to allow for both piecewise-constant rate changes over time and multiple individual types. In particular the skyline approach can be useful to account for changes in the sampling policies (e.g. before/after HIV discovery or before/after PCR test spread for SARS-CoV-2).

Another important source of heterogeneity that none of these models account for is inflicted by non-random sampling. We can propose different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability over time intervals with the BDSKY, however, none of them can account for non-randomness of sampling, in particular due to contact tracing and partner notification.

In this study we propose an extension of the MTBD framework that allows to model non-random sampling due to partner notification, MTBD-PN. In the rest of the article, we describe the MTBD-PN model family and its assumptions, implement a transmission tree simulator under these models, derive the equations for their likelihood calculation, implement a maximum likelihood parameter estimator and test it on simulated data. We also propose a test allowing to distinguish between MTBD and MTBD-PN models. Finally, we apply it to the study of HIV epidemic in Cuba.

\section{MTBD-PN Framework}
In a pathogen transmission tree $\mathscr{T}$ (approximated by a time-scaled pathogen phylogeny) the tips represent sampled pathogens, patient state transitions occur along the branches, and bifurcations (i.e. internal nodes) correspond to pathogen transmissions (Fig.~\ref{fig:tt}). The tree branches are measured in units of time, where $T$ is the time that passed between the tree root (the beginning of the (sub-)epidemic) and the last sampled tip.

\begin{figure*}[tbhp]
\centering 
%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
\includestandalone[width=1\textwidth]{Fig_tree}
\caption{A transmission tree $\mathscr{T}$ with $n=5$ external nodes (i.e. tips, which correspond to sampling events: $0000, 0001, 001, 010, 011$), $n-1=4$ internal nodes (which correspond to transmissions: $0$ (the root) and $00, 01, 000$) and $2n - 2 = 8$ branches (plus the root branch of zero length). %The lengths of the branches are shown on the right of each branch: $t_i$ is the lengths of the branch connecting the node $i$ to its parent. 
Time $t$ starts at the root of the tree ($t=0$) and goes till the last sampled tip. The times of the nodes are shown on the left, e.g. $t_{0001}$ is the time of tip $0001$ (when $0001$'s pathogen was sampled). $T$ corresponds to the end of the sampling period (when the most recent tip, $0000$, was sampled).}
\label{fig:tt} 
\end{figure*}

\subsection{MTBD Framework}
A general MTBD model describes $m$ possible individual states. An individual in state $k \in 1:m$ can be removed (i.e., become non-infectious, e.g., due to healing, starting a treatment, self-isolation, or death) at a rate $\psi_k$, change their state to a state $l \in 1:m$ at a rate $\mu_{kl}$ (where $\mu_{kk} = 0$), and transmit their pathogen to an individual in a state $l \in 1:m$ at rate $\lambda_{kl}$. Upon $i$'s removal, their pathogen can be sampled (and hence observed as a tip in the transmission tree) with a probability $\rho$.

The MTBD model can be described with master equations representing the likelihood densities of evolving as on the observed transmission tree, with time $t$ going forward from the root ($t=0$) till the time of the last sampled tip ($t=T$). These likelihood densities depend on the probability $U_s(t)$ of an unobserved transmission tree that started from one individual at time $t$ in state $s \in 1:m$ and evolved till $T$, without them or any of their induced cases being sampled:

\begin{equation}
\scriptsize
\begin{cases}
\dot{U}_{s}(t) = &\Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) U_{s}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\ 
    & - \sum\limits_{k=1}^{m} \mu_{sk} U_{k}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, folowed by unsampled evolution from the new state }\\
    &- \sum\limits_{k=1}^{m} \lambda_{sk} U_{s}(t)U_{k}(t) \;  \textit{\color{gray} $\leftarrow$ transmission, followed by unsampled evolutions of both subtrees}\\
    &- \psi_s (1 - \rho)\;  \textit{\color{gray} $\leftarrow$ removal without sampling}\\
U_{s}(T) = & 1\;  \textit{\color{gray} $\leftarrow$ the probability to stay unsampled over time 0 is 1} \label{eq:u}
\end{cases}
\end{equation}

With the equations~(\ref{eq:u}), we can write down the equations for the probability densities $p_{s,si}^{(i)}(t)$ of evolving as on an observed tree branch that finishes at time $t_i$ in state $si$, starting  on this branch at a time $t \leq t_i$ in state $s$. The master equations for the MTBD models were initially developed by ~\citet{Stadler2013a}, however here we present their parallelizable version that we recently proposed~\cite{zhukovaFastAccurateMaximumLikelihood2022}:

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}_{s,si}^{(i)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, folowed by evolution from the new state }\\
    & - \sum\limits_{k=1}^{m} \lambda_{sk} p_{s,si}^{(i)}(t)U_{k}(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
    & - \sum\limits_{k=1}^{m} \lambda_{sk} p_{k,si}^{(i)}(t)U_{s}(t)\;  \textbf{\textit{\color{gray} $\leftarrow$ transmission, where the donor subtree stayed unsampled}}\\
p_{s,si}^{(i)}(t_i) = & 
    \begin{cases}
    0 & \textit{\color{gray} if $i$ is in state $k \neq si$}\\
    1 & \textit{\color{gray} if $i$ is in state $si$}
    \end{cases}\label{eq:p}
\end{cases}
\end{equation}

If the states of all the observed tree $\mathscr{T}$ nodes are known, then its likelihood density under parameter values $\Theta$ can be calculated as $L(\mathscr{T}|\Theta)$:

{\scriptsize
\begin{alignat*}{3}
log L(\mathscr{T}|\Theta) =  \sum\limits_{i \in tips}  log(&\psi_{si}  \rho) & \textit{\color{gray} $\leftarrow$ sampling of $n$ tips} \\
 +\sum\limits_{i \in \textit{internal nodes}} log \Big(&\sum\limits_{k=1}^{m} \lambda_{si,k}  & \textit{\color{gray} $\leftarrow$ $n - 1$ transmission events}\\
 & \cdot \big(p_{si,si0}^{(i0)}(t_i)p_{k,si1}^{(i1)}(t_i) + p_{k,si0}^{(i0)}(t_i)p_{si,si1}^{(i1)}(t_i)\big)\Big) & \textit{\color{gray}$\leftarrow$ child branch evolutions}  \label{eq:mtbd_tree_likelihood}
\end{alignat*}}

\subsection{PN extension}

We now propose a non-Markovian extension of the MTBD framework that adds partner notification (PN), for a tree with known node states.  In MTBD-PN, at the moment of sampling the sampled individual might notify their most recent partner with a given probability $\rho_n$. Upon notification, partner is removed almost instantaneously (modelled via a notified removal rate $\psi_n >> \psi_k \;\forall k \in 1:m$). MTBD-PN hence has $2k^2 + 1$ classical MTBD and 2 additional parameters:
\begin{itemize}
 \item $\mu_{kl} \;\forall k,l \in 1:m, k \neq l$ -- state change rate from $k$ to $l$;
 \item $\lambda_{kl} \;\forall k,l \in 1:m$ -- transmission rate: an infected individual in state $k$ transmits their pathogen (which corresponds to an internal node in the transmission tree), the newly infected individual is in state $l$;
 \item $\psi_k \;\forall k \in 1:m$ -- removal rate: an infected individual in state $k$ stops being infectious at this rate (exponential distribution), which corresponds to a tip in the transmission tree (observed or hidden);
 \item $\rho$ -- sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
 \item $\rho_n$ -- notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
 \item $\psi_n >> \psi_k  \;\forall k \in 1:m$ -- notified removal rate: a notified individual stops being infectious and gets observed with this rate (exponential distribution), which corresponds to an observed tip in the transmission tree.
\end{itemize}

This model is a simplification of the notification process, in particular it makes 5 assumptions, which might need to be relaxed in the future work:
\begin{enumerate}
\item only observed individuals can notify (instead of any removed individual);
\item notified individuals are always observed upon removal;
\item notified individuals do not notify further;
\item after the notification notified individuals do not transmit further;
\item only the most recent partner can get notified.
\end{enumerate}

These assumptions make the mathematics behind the model manageable, however they can be easily relaxed in a simulator (e.g. to be used with a deep learning estimator~\cite{Voznica2021} after).

\subsection{Mathematical formulation}
$p_{s,si}^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along a tree branch, provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$. They could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.

We will use an \textbf{oriented probability} $P^{(i,o)}_{s,si}(t)$ for the cases when the individual at the end of the branch is the same as at time $t$:

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}_{s,si}^{(i,o)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i,o)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, folowed by evolution from the new state }\\
    & - \sum\limits_{k=1}^{m} \lambda_{sk} p_{s,si}^{(i,o)}(t)U_{k}(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
p_{s,si}^{(i,o)}(t_i) = & 
    \begin{cases}
    0 & \textit{\color{gray} if $i$ is in state $k \neq si$}\\
    1 & \textit{\color{gray} if $i$ is in state $si$}
    \end{cases}\label{eq:p-o}
\end{cases}
\end{equation}

Additionally we define a \textbf{non-hidden probability} $p^{(i,nh)}_{s,si}(t)$ for the cases when no hidden transmission occurred along the branch:

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}_{s,si}^{(i,nh)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i,nh)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i,nh)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, folowed by evolution from the new state }\\
p_{s,si}^{(i,nh)}(t_i) = & 
    \begin{cases}
    0 & \textit{\color{gray} if $i$ is in state $k \neq si$}\\
    1 & \textit{\color{gray} if $i$ is in state $si$}
    \end{cases}\label{eq:p-nh}
\end{cases}
\end{equation}

Under the MTBD-PN model, the probability density $q_{si,sj}(i)$ of a branch connecting a node $i$ in state $si$ to its parent node $j$ in state $sj$ can be written as:

\begin{equation}
\scriptsize
q_{sj,si}(i) = 
\begin{cases}
q^{(u)}_{sj,si}(i) = p^{(i)}_{sj,si}(t_j) \textit{\color{gray} ~~if $i$ is an \textbf{unnotified non-notifier}}\\
q_{sj,si}^{(n)}(i) = 
p_{sj,si}^{(i,nh)}(t_j) \textit{\color{gray} ~~if $i$ is a tip corresponding to a \textbf{notifier}}\\
~~~~~~~~~~~~ \textit{\color{gray}(parent is the last transmission, no hidden transmission along the branch)}
\\
q^{(p)}_{sj,si}(i)  = p^{(i, o)}_{sj,si}(t_j)  \textit{\color{gray} ~~if $i$ is an internal node who will be notified after $t_i$}
\\
q^{(p)}_{sj,si}(i, r)  = p^{(r, o)}_{sj,si}(t_j) e^{- \psi_n (t_i - t_r)} \textit{\color{gray} if $i$ is a \textbf{partner} tip notified by $r$ at time $t_r < t_i$}\\
\end{cases}
\label{eq:branch}
\end{equation}

Using the formulas~(\ref{eq:branch}), we can calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}, where for each visited node $i$ we calculate two values for each possible state $sj \in 1:m$ at the beginning of the branch connecting it to its parent $j$: the likelihood density of its subtree if its branch corresponds to (1) an unnotified individual $l^{(u)}_{sj}(i)$; (2) a notifier $l^{(u)}_{sj}(i)$ (for tips only); or (3) an (eventually) notified partner $l^{(p)}_{sj}(i, r)$ (notified by $r$).

For a tip $i$, these values can be calculated by considering all its possible states $si \in 1:m$, as:
\begin{equation}
\scriptsize
\begin{array}{ll}
l^{(u)}_{sj}(i) &= \sum\limits_{si=1}^{m} q^{(u)}_{sj,si}(i)\psi_{si}\rho (1-\rho_{n})\\
l^{(n)}_{sj}(i) &= \sum\limits_{si=1}^{m} q^{(n)}_{sj,si}(i)\psi_{si}\rho \rho_{n}\\
l^{(p)}_{sj}(i, r) &= \sum\limits_{si=1}^{m} q^{(p)}_{sj,si}(i,r)\psi_{n}\\
\end{array}
\label{eq:tip}
\end{equation}

Let us now consider an internal node $i$ in state $si$ with two child nodes, $i0$ and $i1$. We will start with the configuration where the $i$'s branch corresponds to an unnotified individual:

\begin{equation}
\scriptsize
\begin{split}
l^{(u)}_{sj}(i) = &\sum\limits_{si=1}^{m}q^{(u)}_{sj,si}(i) \sum\limits_{k=1}^{m} \lambda_{si,k} \\
\cdot \Big(~&
l^{(u)}_{si}(i0)l^{(u)}_{k}(i1) + l^{(u)}_{k}(i0)l^{(u)}_{si}(i1)  \textit{\color{gray} $\leftarrow$ no notification on child branches} \\
& + is\_tip(i0) \big(l^{(n)}_{si}(i0)l^{(p)}_{k}(i1,i0) + l^{(n)}_{k}(i0)l^{(p)}_{si}(i1,i0) \big)\textit{\color{gray} $\leftarrow$ notification of someone in $i1$'s subtree by $i0$} \\
&+ is\_tip(i1) \big(l^{(p)}_{si}(i0,i1)l^{(n)}_{k}(i1) + l^{(p)}_{k}(i0,i1)l^{(n)}_{si}(i1) \big)~\Big) \textit{\color{gray} $\leftarrow$ notification of someone in $i0$'s subtree by $i1$},\\
\textit{where } & is\_tip(r) = 1 \textit{ if r is a tip and $= 0$ if r is an internal node.}
\end{split}\label{eq:lu}
\end{equation}

Note that a notification will only be possible if (1) the notifying branch is external (as the notifier only notifies the most recent partner); (2) there is at least one tip in the notified subtree that can be a partner, i.e. was sampled after the notifier and whose parent's time was before notification (i.e. there was no transmission between notification and sampling).

In the other case (when an internal nodes's branch corresponds to a partner who will eventually get notified by a node $r$), the individual on the path connecting this branch to the notified tip stays the same (the partner). Hence, all the transmissions on this path are oriented and correspond to the partner transmitting to another recipient individual. Moreover, this recipient's branch cannot be notified by our partner (as in our model notified partners do not notify further):

\begin{equation}
\scriptsize
\begin{split}
l^{(p)}_{sj}(i, r) = \sum\limits_{si=1}^{m}q^{(p)}_{sj,si}(i,r) \sum\limits_{k=1}^{m}  \lambda_{si,k}\Big(&l^{(p)}_{si}(i0,r)l^{(u)}_{k}(i1) \textit{\color{gray} $\leftarrow$ notification in $i0$'s subtree} \\
 &+ l^{(u)}_{k}(i0)l^{(p)}_{si}(i1,r)\Big)  \textit{\color{gray} $\leftarrow$ notification in $i1$'s subtree} 
\end{split}\label{eq:lu}
\end{equation}

Note that $l^{(p)}_{sj}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time is known, i.e. once we have climbed to an internal node corresponding to the transmission between the notifier (corresponding to one of its children, must be a tip) and the partner (the other child). As only the contributions coming from the partner tip branches depend on the notification times, while climbing the tree for each internal node $i$ and each of the possible states $sj \in 1:m$ at the beginning of its branch, we can store an array of pre-calculated values for notified internal paths for each of its tips (e.g. $a$) and each of the states $v \in 1:m$ at the beginning of $a$'s branch, describing the tip's sampling time $t_a$ and a configuration $C_{sj}(i,a,v)$:

\begin{equation}
\scriptsize
\begin{split}
C_{sj}(i,a,v) = &\begin{cases}
q^{(p)}_{sj,v}(i)
\sum\limits_{k=1}^{m}\lambda{v,k} \cdot l_{k}^{(u)}\Big(child(i, \not a)\Big) \textit{\color{gray}~~~if $i$ is a direct parent of $a$}
\\
\sum\limits_{si=1}^{m}q^{(p)}_{sj,si}(i)
\sum\limits_{k=1}^{m}\lambda{si,k} \cdot l_{k}^{(u)}\Big(child(i, \not a)\Big)C_{si}\Big(child(i, a),a,v\Big) \textit{\color{gray}~~otherwise}\\ 
\end{cases},
\\ &\textit{where child(i,a) is the child of i whose subtree contains a},
\\
 &\textit{~~~~~~~~child(i,$\not a$) is the child of i whose subtree does not contain a.}
\end{split}
\label{eq:c}
\end{equation}

%Note that the path connecting $j$ to the parent of $i$ in this configuration corresponds to the same partner individual, hence this individual is the donor at all transmissions (internal nodes) on this path, hence at the moments of these transmissions the individual's state is known (as we only consider trees with known node states). At each internal node in this path the other branch (leading to a non-$j$'s subtree) corresponds to a recipient of the transmission occurring at that node, where the donor is the partner individual corresponding to $j$ (and $i$) and hence will not notify the recipient (as in our model's assumptions notified partners do not notify further).

Once a potential notifier tip $r$ is identified (i.e., $r$ is a sister tip of the currently visited node $i$), we can calculate $l^{(p)}_{sj}(i, r)$ as:
\begin{equation}
\scriptsize 
l^{(p)}_{sj}(i, r) = \sum\limits_{a \in tips(i)}\sum\limits_{k=1}^{m} C_{sj}(i, a, k) l^{(p)}_{k}(a, r)\label{eq:lp}
\end{equation}

Finally, the tree $\mathscr{T}$ likelihood density $\mathscr{L}(\mathscr{T}|\Theta)$ under MTBD-PN model with parameters $\Theta$ can be calculated as $l^{(u)}_{state(root)}(root)$, as the root can only correspond to an unnotified individual (since there was no observed transmission before it).

%The computational of $\mathscr{L}(\mathscr{T}|\Theta)$ includes $O(N^2)$ resolutions of system~(\ref{eq:p}) (for different times and initial conditions) in the worst case (caterpillar tree) and $O(NlogN)$ in the best case (for a balanced tree), where $N$ is the number of tips in $\mathscr{T}$.

\subsection{BD-PN}
A particular example of the MTBD model generating trees with known node states is the basic birth-death model~\citep{Stadler2009} where all the individuals are in the same state. They can transmit with a constant rate $\lambda$, get removed with a constant rate $\psi$, be sampled upon removal with a probability $\rho$. Below we describe the special case of equations~(\ref{eq:u}-\ref{eq:lp}) for its PN extension BD-PN.

The master equation for this model are a simplified version of the general case:

\begin{equation}
\scriptsize
\begin{cases}
\dot{U}(t) &= \Big(\lambda + \psi\Big) U(t) - \lambda U^2(t) - \psi (1 - \rho)\\\\
U(T) &=  1\\
\dot{p}^{(i)}(t) &=  \Big(\lambda + \psi\Big) p^{(i)}(t) - 2 \lambda p^{(i)}(t)U(t)\\
\dot{p}^{(i,o)}(t) &=  \Big(\lambda + \psi\Big) p^{(i)}(t) - \lambda p^{(i)}(t)U(t)\\
\dot{p}^{(i,nh)}(t) &=  \Big(\lambda + \psi\Big) p^{(i)}(t)\\
p^{(i)}(t_i) &= p^{(i,o)}(t_i) = p^{(i,nh)}(t_i) = 1
\end{cases}
\end{equation}

These equations have the following closed form solutions:
\begin{equation}
\scriptsize
\begin{split}
&\begin{cases}
U(t) = \frac{\lambda + \psi + c_1 \frac{c_2 e^{t c_1} - 1}{c_2 e^{tc_1} + 1}}{2\lambda}\\
\\
p^{(i)}(t) = \frac{c_3^2 e^{c_1 (t - t_i)}}{(c_2e^{c_1 t} + 1)^2} \\
\\
p^{(i,o)}(t) =  \frac{c_3 e^{1/2 (\lambda + \psi + c_1)(t - t_i)}}{c_2e^{c_1 t} + 1} \\
\\
p^{(i,nh)}(t) =  e^{(\lambda + \psi)(t - t_i)}
\end{cases},\\
& \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi}e^{-Tc_1},\; c_3 = c_2 e^{c_1 t_i} + 1
\end{split}
\end{equation}

The likelihood of a tree $\mathscr{T}$ under BD-PN model with parameter values $\Theta$ can be calculated as $\mathscr{L}(\mathscr{T}|\Theta) = l^{(u)}(root)$, where $l^{(u)}(i)$ is a likelihood of node $i$'s subtree if the branch connecting $i$ to its parent node (of length zero for the root) correspond to an unnotified individual. It can be calculated, along with $l^{(p)}(i, r)$ (likelihood when $i$'s branch corresponds to an eventually notified partner, notified by $r$) for each tree node $i$ with a pruning algorithm.

For a tip $i$ whose branch starts at time $t_j$ and finishes at time $t_i$, these values can be calculated as:
\begin{equation}
\scriptsize
\begin{array}{ll}
l^{(u)}(i) &= p^{(i)}(t_j)\psi\rho (1-\rho_{n})\\
l^{(n)}(i) &= p^{(i,nh)}(t_j)\psi\rho \rho_{n}\\
l^{(p)}(i, r) &= p^{(r,o)}(t_j)e^{-\psi_n (t_i - t_j)}\psi_{n}\\
\end{array}
\end{equation}

For an internal node $i$ whose branch starts at time $t_j$ and finishes at time $t_i$, with two child nodes, $i0$ and $i1$, these values can be calculated as:

\begin{equation}
\scriptsize
\begin{array}{ll}
l^{(u)}(i) &= p^{(i)}(t_j) 2\lambda 
\Big(l^{(u)}(i0)l^{(u)}(i1) 
+ is\_tip(i0) l^{(n)}(i0)l^{(p)}(i1,i0)
+ is\_tip(i1) l^{(p)}(i0,i1)l^{(n)}(i1)\Big),\\
& \textit{where } is\_tip(x) = 1 \textit{ if x is a tip and $= 0$ if x is an internal node.}\\
l^{(p)}(i, r) &= p^{(i,o)}(t_j) \lambda\Big(l^{(p)}(i0,r)l^{(u)}(i1) + l^{(u)}(i0)l^{(p)}(i1,r)\Big),\\
& \textit{where r is the notifier.}\\
\end{array}
\end{equation}

As in the general MTBD-PN case, $l^{(p)}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time is known, i.e. once we have climbed to an internal node corresponding to the transmission between the notifier (corresponding to one of its children, must be a tip) and the partner (the other child). As only the contributions coming from the partner tip branches depend on the notification times, while climbing the tree for each internal node $i$ and the time of its branch start $t_j$, we can store an array of pre-calculated values for notified internal paths for each of its tips (e.g. $a$), describing the tip's sampling time $t_a$ and a configuration $C(i,a)$:

\begin{equation}
\scriptsize
\begin{split}
C(i,a) = &\begin{cases}
1 \textit{\color{gray}~~~if $i == a$}
\\
p^{(i,o)}(t_j) \lambda \cdot l^{(u)}\Big(child(i, \not a)\Big)C\Big(child(i, a),a\Big) \textit{\color{gray}~~otherwise}\\ 
\end{cases},
\\ &\textit{where child(i,a) is the child of i whose subtree contains a},
\\
 &\textit{~~~~~~~~child(i,$\not a$) is the child of i whose subtree does not contain a.}
\end{split}
\end{equation}

Once a potential notifier tip $r$ is identified (i.e., $r$ is a sister tip of the currently visited node $i$), we can calculate $l^{(p)}(i, r)$ as:
\begin{equation}
\scriptsize 
l^{(p)}(i, r) = \sum\limits_{a \in tips(i)}C(i, a) l^{(p)}(a, r)
\end{equation}

\subsection{BDEI-PN}
Another particular example of the MTBD model generating trees with known node states is the basic birth-death exposed-infectious model (BDEI)~\citep{Stadler2014} where   infectious ($I$) individuals can transmit (at a rate $\lambda$) or be removed (at a rate $\psi$), and hence create an internal or external node in the transmission tree. Upon removal they can be sampled with a probability $\rho$. The recipient of the transmission is at first in state $E$ (exposed, already infected but not yet infectious), they can eventually become infectious ($I$) at a state-change rate $\mu$. No other type of state transitions, transmissions or removals are allowed. Below we describe the special case of equations~(\ref{eq:u}-\ref{eq:lp}) for its PN extension BDEI-PN, assuming only infectious individuals can be removed and sampled even with notification (e.g., because the viral load of the exposed individuals is yet undetectable).

The master equations for the BDEI-PN model are a special case of the MTBD-PN equations:

\begin{equation}
\scriptsize
\begin{cases}
\dot{U_E}(t) = &\mu U_E(t)- \mu U_I(t)\\    
\dot{U_I}(t) = &\Big(\lambda + \psi\Big) U_I(t)- \lambda U_I(t)U_E(t)- \psi (1 - \rho)\\
U_E(T) = & U_I(T) = 1\\
\dot{p}_E^{(i)}(t) = & \mu p_E^{(i)}(t) - \mu p_I^{(i)}(t)\\
p_E^{(i)}(t_i) = & 0\\
p_E^{(i,o)}(t_i) = & p_E^{(i,nh)}(t_i) = p_E^{(i)}(t_i)\\
\dot{p}_I^{(i)}(t) = & \Big(\lambda + \psi\Big) p^{(i)}(t) - \lambda \Big(p_I^{(i)}(t)U_E(t) + p_E^{(i)}(t)U_I(t)\Big)\\
\dot{p}_I^{(i,o)}(t) = & \Big(\lambda + \psi\Big) p^{(i,o)}(t) - \lambda p_I^{(i,o)}(t)U_E(t)\\
\dot{p}_I^{(i,nh)}(t) = & \Big(\lambda + \psi\Big) p^{(i,nh)}(t)\\
p_I^{(i)}(t_i) = & p_I^{(i,o)}(t_i) = p^{(i,nh)}(t_i) = 1\label{eq:up-bd}
\end{cases}
\end{equation}

The likelihood of a tree $\mathscr{T}$ under BDEI-PN model with parameter values $\Theta$ can be calculated as $\mathscr{L}(\mathscr{T}|\Theta) = l_I^{(u)}(root)$, where $l_s^{(u)}(i)$ is a likelihood of node $i$'s subtree if the branch connecting $i$ to its parent node (of length zero for the root) starts in state $s \in \{E, I\}$ and correspond to an unnotified individual. It can be calculated, along with $l_s^{(p)}(i, notifier)$ (likelihood when $i$'s branch corresponds to an eventually notified partner) for each tree node $i$ with a pruning algorithm.

For a tip $i$ whose branch starts at time $t_j$ in state $s \in \{E, I\}$ and finishes at time $t_i$ in state $I$, these values can be calculated as:
\begin{equation}
\scriptsize
\begin{array}{ll}
l^{(u)}_{s}(i) &= p_s^{(i)}(t_j)\psi\rho (1-\rho_{n})\\
l^{(n)}_{s}(i) &= p_s^{(i,nh)}(t_j)\psi\rho \rho_{n}\\
l^{(p)}_{s}(i, r) &= p_s^{(r,o)}(t_j)e^{-\psi_n (t_i - t_j)}\psi_{n}\\
\end{array}
\end{equation}

For an internal node $i$ whose branch starts at time $t_j$ is state $s$ and finishes at time $t_i$ in state $I$, with two child nodes, $i0$ and $i1$, these values can be calculated as:

\begin{equation}
\scriptsize
\begin{array}{ll}
l^{(u)}(i) =& p_s^{(i)}(t_j) \lambda \\
& \cdot \Big(\big(l_I^{(u)}(i0)l_E^{(u)}(i1) + l_E^{(u)}(i0)l_I^{(u)}(i1)\big) \\
&~~+ is\_tip(i0) \big(l_I^{(n)}(i0)l_E^{(p)}(i1,i0) + l_E^{(n)}(i0)l_I^{(p)}(i1,i0)\big)\\
&~~+ is\_tip(i1) \big(l_I^{(p)}(i0,i1)l_E^{(n)}(i1) + l_E^{(p)}(i0,i1)l_I^{(n)}(i1)\big)\Big),\\
& \textit{where } is\_tip(x) = 1 \textit{ if x is a tip and $= 0$ if x is an internal node.}\\
l^{(p)}(i, r) &= p_s^{(i,o)}(t_j) \lambda\Big(l_I^{(p)}(i0,r)l_E^{(u)}(i1) + l_E^{(u)}(i0)l_I^{(p)}(i1,r)\Big),\\
& \textit{where r is the notifier.}\\
\end{array}
\end{equation}

As in the general MTBD-PN case, $l_s^{(p)}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time is known, i.e. once we have climbed to an internal node corresponding to the transmission between the notifier (corresponding to one of its children, must be a tip) and the partner (the other child). As only the contributions coming from the partner tip branches depend on the notification times, while climbing the tree for each internal node $i$ and each state $s \in \{E, I\}$ at the time of its branch start $t_j$, we can store an array of pre-calculated values for notified internal paths for each of its tips (e.g. $a$), describing the tip's sampling time $t_a$ and a configuration $C_s(i,a)$. Note, that the partner branch for any transmission after the transmission between the notifier and the partner, is always the donor one (hence, starting in state $I$).

\begin{equation}
\scriptsize
\begin{split}
C_s(i,a) = &\begin{cases}
1 \textit{\color{gray}~~~if $i==a$}
\\
p_s^{(i,o)}(t_j) \lambda \cdot l_E^{(u)}\Big(child(i, \not a)\Big)C_I\Big(child(i, a),a\Big) \textit{\color{gray}~~otherwise}\\ 
\end{cases},
\\ &\textit{where child(i,a) is the child of i whose subtree contains a},
\\
 &\textit{~~~~~~~~child(i,$\not a$) is the child of i whose subtree does not contain a.}
\end{split}
\end{equation}

Once a potential notifier tip $r$ is identified (i.e., $r$ is a sister tip of the currently visited node $i$), we can calculate $l_s^{(p)}(i, r)$ as:
\begin{equation}
\scriptsize 
l_s^{(p)}(i, r) = 
\begin{cases}
\sum\limits_{a \in tips(i)}C_s(i, a) l_I^{(p)}(a, r)\textit{\color{gray}~~if $i$ is an internal node or $s=I$}\\
l_E^{(p)}(a, r)\textit{\color{gray}~~if $i=a$ and $s=E$}
\end{cases}
\end{equation}

\section{Tree simulator}
We implemented a quick tree simulator that generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states). The simulator is Gillespie-based, generates state change, transmission and removal times, and only reconstructs the sampled parts of the tree to save memory and increase speed.

The simulator iterates through occurring events till either time or sampled tip number limit is reached. 
It keeps updating an array $C = [c_1, \ldots, c_m]$ of counts of currently infected individuals (i.e. non-removed) in different states; an array $S = [s_1, \ldots, s_m]$ of counts of sampled individuals in different states; as well as a mapping $M$ between states and ids of currently infected individuals in those states, and a mapping $N$ between states and ids of sampled individuals in those states.  In the beginning ($t=0$) the only infected individual corresponds to the root: $c_k = 0, \;M_k = \emptyset \; \forall k \in 1:m, \; k \neq state(root)$, $c_{state(root)} = 1, \; M_{state(root)} = \{root\}$, and no individual is sampled: $s_k=0,\;N_k = \emptyset \; \forall k \in 1:m$.

At each iteration the algorithm (1) calculates the time of the next event; (2) chooses the type of this event and the the individual involved in it; (3) updates the counts and the mappings according to the event and records its time.

At step 1, to calculate the time of the next event, we (i) calculate the total rate $r$ as a sum of total state change, transmission and transition rates: $r = r^{(\mu)} + r^{(\lambda)} + r^{(\psi)}$, where $r^{(\mu)} = \sum\limits_{k=1}^{m} c_k \sum\limits_{r=1}^{m} \mu_{kr}$, $r^{(\lambda)} = \sum\limits_{k=1}^{m} c_k \sum\limits_{r=1}^{m} \lambda_{kr}$, and $r^{(\psi)} = \sum\limits_{k=1}^{m} c_k \psi_{k}$; (ii) draw $\Delta t$ from the exponential distribution with rate $r$; (iii) update the current time to $t + \Delta t$.

At step 2, to chose the type of the event, we draw a value $v$ (uniformly) from an interval $[0, r[$. This interval can be seen as composed of subintervals corresponding to each possible event, where the width of each subinterval is defined by the corresponding rate and infected individual count, e.g. the subinterval corresponding to a transmission from $k$ to $r$ has a width of $\lambda_{kr}c_k$. The subinterval in which $v$ is located defines the event type.

At step 3, we randomly draw an individual $i$ of type $k$ (selected at step 2) from $M[k]$, and proceed depending on the event type. If it is a state-change from $k$ to $r$, then we decrease $c_k$ by one, increase $c_r$ by one, remove $i$ from $M[k]$ and put $i$ into $M[r]$. If it is a transmission from $k$ to $r$, then we increase $c_r$ by one, create a new id $j$ for the recipient and put $j$ into $M[r]$. If it is a removal, then we decrease $c_k$ by one, remove $i$ from $M[k]$, and draw a value $p$ (uniformly) from an interval $[0, 1[$: if $p < \rho$, the individual gets sampled and we put $i$ into $N[k]$.

For transmissions and sampling events we also record their time and ids of the involved individuals. These values are used to reconstruct the sampled parts of the tree once the tree simulation is finished.

For the PN version of this simulator, we add an additional $m+1^{st}$ state for notified partners. $m + 1$'s state-change and transmission rates are set to zero: $\mu_{k,m+1} = \mu_{m+1,k} = \lambda_{k,m+1} = \lambda_{m+1,k} = 0\; \forall k \in 1:m$, while its removal rate is set to the notification removal rate: $\psi_{m+1} = \psi_n >> \psi_k\; \forall k \in 1:m$. We keep a mapping of most recent partners $P$ for each infectious individual, and update it for both the donor and the recipient at each transmission event. We also keep a mapping between each infectious individual and their state, and update it at each state-change event. At each removal event, we check whether the individual $i$ is a notified partner (i.e. in state $m+1$). If yes, we put them into $N[m+1]$ without performing the sampling probability draw. If not, and if $i$ gets sampled upon removal, we draw a value $p$ (uniformly) from an interval $[0, 1[$. If $p < \rho_n$, and if $i$'s most recent partner ($j = P[i]$) is not yet sampled (not in $N[state(j)]$), we change $j$'s state to $m + 1$ as in a state-change event.

\subsection*{Code availability}
The simulator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\cite{Huerta-Cepas2016} and NumPy package for array operations~\cite{harris_array_2020}.

It is available as a command-line program and a Python 3 package via PyPi (\href{https://pypi.org/project/treesimulator}{treesimulator}), and via Docker/Singularity (\href{https://hub.docker.com/r/evolbioinfo/treesimulator/tags}{evolbioinfo/treesimulator}). Its source code and the installation and usage documentation are available on GitHub at \href{https://github.com/evolbioinfo/treesimulator}{github.com/evolbioinfo/treesimulator}.

\end{document}

### 2023-05-23 src/bdpn.py

@@ -1,0 +1,172 @@
+import numpy as np
+import pandas as pd
+from ete3 import Tree
+from scipy.optimize import NonlinearConstraint, LinearConstraint, minimize
+
+
+def annotate_tree(tree):
+    for n in tree.traverse('preorder'):
+        if n.is_root():
+            p_time = 0
+        else:
+            p_time = getattr(n.up, 'time')
+        n.add_feature('time', p_time + n.dist)
+
+
+def get_c1(la, psi, rho):
+    return np.power(np.power((la - psi), 2) + 4 * la * psi * rho, 0.5)
+
+
+def get_c2(la, psi, rho, T):
+    c1 = get_c1(la, psi, rho)
+    return (c1 + la - psi) / (c1 - la + psi) * np.exp(-T * c1)
+
+
+def get_c3(la, psi, rho, T, ti):
+    c1 = get_c1(la, psi, rho)
+    c2 = get_c2(la, psi, rho, T)
+    return c2 * np.exp(c1 * ti) + 1
+
+
+def get_u(t, la, psi, rho, T):
+    c1 = get_c1(la, psi, rho)
+    c2 = get_c2(la, psi, rho, T)
+    exp_t_c1 = np.exp(t * c1)
+    return (la + psi + c1 * (c2 * exp_t_c1 - 1) / (c2 * exp_t_c1 + 1)) / (2 * la)
+
+
+def get_p(t, la, psi, rho, T, ti):
+    c1 = get_c1(la, psi, rho)
+    c2 = get_c2(la, psi, rho, T)
+    c3 = get_c3(la, psi, rho, T, ti)
+    return (np.power(c3, 2) * np.exp(c1 * (t - ti))) / np.power(c2 * np.exp(c1 * t) + 1, 2)
+
+
+def loglikelihood(tree, la, psi, rho, T):
+    n = len(tree)
+    res = n * np.log(psi * rho) + (n - 1) * np.log(2 * la)
+    for n in tree.traverse('preorder'):
+        if not n.is_leaf():
+            ti = getattr(n, 'time')
+            c1, c2 = n.children
+            res += np.log(get_p(ti, la, psi, rho, T, getattr(c1, 'time'))) \
+                   + np.log(get_p(ti, la, psi, rho, T, getattr(c2, 'time')))
+    return res
+
+
+def optimize_likelihood_params(tree, T, la=None, psi=None, rho=None):
+    """
+    Optimizes the likelihood parameters for a given forest and a given MTBD model.
+
+
[…]

### 2023-05-23 src/bdpn.py
minor

@@ -157,7 +157,7 @@ def optimize_likelihood_params(tree, T, la=None, psi=None, rho=None):
     return get_real_params_from_optimised(x0)
 
 
-nwk = '/home/azhukova/projects/bdei/bd/large/trees/tree.1.nwk'
+nwk = '/home/azhukova/projects/bdpn/trees/bd/tree.1.nwk'
 tree = Tree(nwk)
 annotate_tree(tree)
 df = pd.read_csv(nwk.replace('.nwk', '.log'))

### 2023-06-02 src/bd.py
split bd and bdpn

@@ -42,86 +42,19 @@ def get_p(t, la, psi, rho, T, ti):
     return (np.power(c3, 2) * np.exp(c1 * (t - ti))) / np.power(c2 * np.exp(c1 * t) + 1, 2)
 
 
-def get_p_o(t, la, psi, rho, T, ti):
-    c1 = get_c1(la, psi, rho)
-    c2 = get_c2(la, psi, rho, T)
-    c3 = get_c3(la, psi, rho, T, ti)
-    return (c3 * np.exp(1/2 * (la + psi + c1) * (t - ti))) / (c2 * np.exp(c1 * t) + 1)
+def loglikelihood(tree, la, psi, rho, T):
+    n = len(tree)
+    res = n * np.log(psi * rho) + (n - 1) * np.log(2 * la)
+    for n in tree.traverse('preorder'):
+        if not n.is_leaf():
+            ti = getattr(n, 'time')
+            c1, c2 = n.children
+            res += np.log(get_p(ti, la, psi, rho, T, getattr(c1, 'time'))) \
+                   + np.log(get_p(ti, la, psi, rho, T, getattr(c2, 'time')))
+    return res
 
 
-def get_p_nh(t, la, psi, ti):
-    return np.exp((la + psi) * (t - ti))
-
-
-def get_lu(i, la, psi, psi_n, rho, rho_n, T, attribute=True):
-    ti = getattr(i, 'time')
-    tj = ti - i.dist
-    pi_tj = get_p(tj, la, psi, rho, T, ti)
-    if i.is_leaf():
-        return pi_tj * psi * rho * (1 - rho_n)
-    i0, i1 = i.children
-    lu_0 = getattr(i0, 'LU') if attribute else get_lu(i0, la, psi, psi_n, rho, rho_n, T, attribute)
-    lu_1 = getattr(i1, 'LU') if attribute else get_lu(i1, la, psi, psi_n, rho, rho_n, T, attribute)
-    branch_probs = lu_0 * lu_1
-    if i0.is_leaf():
-        branch_probs += get_ln(i0, la, psi, rho, rho_n) * get_lp(i1, i0, la, psi, psi_n, rho, rho_n, T, attribute)
-    if i1.is_leaf():
-        branch_probs += get_lp(i0, i1, la, psi, psi_n, rho, rho_n, T, attribute) * get_ln(i1, la, psi, rho, rho_n)
-    return pi_tj * 2 * la * branch_probs
-
-
[…]

### 2023-06-05 article/article.tex
BDPN estimator and simulation pipelines

@@
 \begin{cases}
-\dot{p}_{s,si}^{(i,o)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+\dot{p}_{s,si}^{(i,o)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i,o)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
     & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i,o)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, folowed by evolution from the new state }\\
@@
 l^{(n)}(i) &= p^{(i,nh)}(t_j)\psi\rho \rho_{n}\\
-l^{(p)}(i, r) &= p^{(r,o)}(t_j)e^{-\psi_n (t_i - t_j)}\psi_{n}\\
+l^{(p)}(i, r) &= p^{(r,o)}(t_j)e^{-\psi_n (t_i - t_r)}\psi_{n}\\
 \end{array}
@@
 l^{(n)}_{s}(i) &= p_s^{(i,nh)}(t_j)\psi\rho \rho_{n}\\
-l^{(p)}_{s}(i, r) &= p_s^{(r,o)}(t_j)e^{-\psi_n (t_i - t_j)}\psi_{n}\\
+l^{(p)}_{s}(i, r) &= p_s^{(r,o)}(t_j)e^{-\psi_n (t_i - t_r)}\psi_{n}\\
 \end{array}

### 2023-06-05 bdpn/bd.py

@@ -1,0 +1,62 @@
+import numpy as np
+
+from bdpn.parameter_estimator import get_rough_rate_etimates
+from bdpn.tree_manager import TIME
+
+
+def get_c1(la, psi, rho):
+    return np.power(np.power((la - psi), 2) + 4 * la * psi * rho, 0.5)
+
+
+def get_c2(la, psi, rho, T):
+    c1 = get_c1(la, psi, rho)
+    return (c1 + la - psi) / (c1 - la + psi) * np.exp(-T * c1)
+
+
+def get_c3(la, psi, rho, T, ti):
+    c1 = get_c1(la, psi, rho)
+    c2 = get_c2(la, psi, rho, T)
+    return c2 * np.exp(c1 * ti) + 1
+
+#
+# def get_u_wa(t, la, psi, rho, T):
+#     x = t - T
+#     c = np.power(-np.power(la, 2) - 4 * la * psi * rho + 2 * la * psi - np.power(psi, 2), 0.5)
+#     return 1 / (2 * la) * (c * np.tan(0.5 * (-x * c + 2 * np.arctan((la - psi) / c))) + la + psi)
+
+
+def get_u(t, la, psi, rho, T):
+    c1 = get_c1(la, psi, rho)
+    c2 = get_c2(la, psi, rho, T)
+    exp_t_c1 = np.exp(t * c1)
+    return (la + psi + c1 * (c2 * exp_t_c1 - 1) / (c2 * exp_t_c1 + 1)) / (2 * la)
+
+
+def get_log_p(t, la, psi, rho, T, ti):
+    c1 = get_c1(la, psi, rho)
+    c2 = get_c2(la, psi, rho, T)
+    c3 = get_c3(la, psi, rho, T, ti)
+    # res1 = (np.power(c3, 2) * np.exp(c1 * (t - ti))) / np.power(c2 * np.exp(c1 * t) + 1, 2)
+    res2 = (2 * np.log(c3) + (c1 * (t - ti))) - (2 * np.log(c2 * np.exp(c1 * t) + 1))
+    # assert(np.log(res1) == res2)
+    return res2
+
+
+def loglikelihood(tree, la, psi, rho, T):
+    n = len(tree)
+    res = n * np.log(psi * rho) + (n - 1) * np.log(2 * la)
+    for n in tree.traverse('preorder'):
+        if not n.is_leaf():
+            ti = getattr(n, TIME)
+            c1, c2 = n.children
[…]

### 2023-06-05 bdpn/bdpn.py

@@ -1,0 +1,126 @@
+import os
+
+import numpy as np
+from ete3 import Tree
+
+from bdpn.bd import get_c1, get_c2, get_c3, get_log_p
+from bdpn.parameter_estimator import get_rough_rate_etimates, optimize_likelihood_params, rescale_log
+from bdpn.tree_manager import TIME
+
+
+def get_log_p_o(t, la, psi, rho, T, ti):
+    c1 = get_c1(la, psi, rho)
+    c2 = get_c2(la, psi, rho, T)
+    c3 = get_c3(la, psi, rho, T, ti)
+    # return (c3 * np.exp(1/2 * (la + psi + c1) * (t - ti))) / (c2 * np.exp(c1 * t) + 1)
+    return np.log(c3) + (1/2 * (la + psi + c1) * (t - ti)) - np.log(c2 * np.exp(c1 * t) + 1)
+
+
+def get_log_p_nh(t, la, psi, ti):
+    # return np.exp((la + psi) * (t - ti))
+    return (la + psi) * (t - ti)
+
+
+def loglikelihood(tree, la, psi, psi_n, rho, rho_n, T):
+
+    for n in tree.traverse():
+        ti = getattr(n, TIME)
+        t = ti - n.dist
+        n.add_feature('p', get_log_p(t, la, psi, rho, T, ti))
+        n.add_feature('po', get_log_p_o(t, la, psi, rho, T, ti))
+        if n.is_leaf():
+            n.add_feature('pnh', get_log_p_nh(t, la, psi, ti))
+
+    for n in tree.traverse('postorder'):
+        ti = getattr(n, TIME)
+        t = ti - n.dist
+        if n.is_leaf():
+            n.add_feature('LU', getattr(n, 'p') + np.log(psi * rho * (1 - rho_n)))
+            n.add_feature('LN', getattr(n, 'pnh') + np.log(psi * rho * rho_n))
+            n.add_feature('CP', [(t, ti, 1)])
+            continue
+            
+        c1, c2 = n.children
+        result = []
+        po_log_la = getattr(n, 'po') + np.log(la)
+        po_log_la_lu2 = po_log_la + getattr(c2, 'LU')
+        po_log_la_lu1 = po_log_la + getattr(c1, 'LU')
+        for (t, ti, c) in getattr(c1, 'CP'):
+            result.append((t, ti, po_log_la_lu2 + c))
+        for (t, ti, c) in getattr(c2, 'CP'):
[…]

### 2023-06-05 bdpn/main.py

@@ -1,0 +1,26 @@
+import pandas as pd
+from ete3 import Tree
+
+import bdpn.bd as bd
+import bdpn.bdpn as bdpn
+from bdpn.parameter_estimator import optimize_likelihood_params
+
+nwk = '/home/azhukova/projects/bdpn/trees/bdpn/tree.1.nwk'
+tree = Tree(nwk)
+df = pd.read_csv(nwk.replace('.nwk', '.log'))
+rho = df.loc[0, 'sampling probability']
+rho_n = df.loc[0, 'notification probability']
+R0 = df.loc[0, 'R0']
+it = df.loc[0, 'infectious time']
+rt = df.loc[0, 'removal time after notification']
+psi = 1 / it
+psi_n = 1 / rt
+la = R0 * psi
+vs = optimize_likelihood_params(tree, input_parameters=[la, None, None, None, None],
+                                loglikelihood=bdpn.loglikelihood, get_bounds_start=bdpn.get_bounds_start)
+
+vs_bd = optimize_likelihood_params(tree, input_parameters=[la, None, None],
+                                   loglikelihood=bd.loglikelihood, get_bounds_start=bd.get_bounds_start)
+print('Real params: {}'.format([la, psi, psi_n, rho, rho_n]))
+print('Found BDPN params: {}'.format(vs))
+print('Found BD params: {}'.format(vs_bd))

### 2023-06-05 bdpn/mtbd.py

@@ -1,0 +1,232 @@
+import numpy as np
+
+
+RTOL = 100 * np.finfo(np.float64).eps
+from scipy.integrate import odeint
+
+N_U_STEPS = int(1e7)
+
+
+def find_index_within_bounds(sol, start, stop, upper_bound, lower_bound=0):
+    """
+    Find an index i in a given array sol (of shape n x m),
+    such that all the values of sol[i, :] are withing the given bounds
+    and at least one of them is above the lower bound.
+
+    Note that such index might not be unique. The search starts with the middle of the array
+    and if needed proceeds with the binary search in the lower half of the array.
+
+    :param sol: an array where to search
+    :param start: start position in the array (inclusive)
+    :param stop: stop position in the array (exclusive)
+    :param upper_bound: upper bound
+    :param lower_bound: lower bound
+    :return: the index i that satisfies the above conditions
+    """
+    i = start + ((stop - start) // 2)
+    if i == start or stop <= start:
+        return start
+    value = sol[i, :]
+    if np.all(value >= lower_bound) and np.all(value <= upper_bound) and np.any(value > lower_bound):
+        return i
+    return find_index_within_bounds(sol, start, i, upper_bound)
+
+
+def find_time_index(v, tt, start, stop):
+    """
+    Searches for an index i in time array tt, such that tt[i] >= v > tt[i + 1], using a binary search.
+
+    :param v: a time value for which the index is searched for
+    :param tt: a time array [t_n, ..., t_0] such that t_{i + 1} > t_i.
+    :param start: start index for the search (inclusive)
+    :param stop: stop index for the search (exclusive)
+    :return: an index i, such that tt[i] >= v > tt[i + 1].
+    """
+    i = start + ((stop - start) // 2)
+    if i == start or i == stop - 1:
+        return i
+    if tt[i] >= v:
+        if tt[i + 1] < v:
+            return i
+        return find_time_index(v, tt, i + 1, stop)
+    if tt[i - 1] >= v:
+        return i - 1
+    return find_time_index(v, tt, start, i)
+
+
[…]

### 2023-06-05 bdpn/parameter_estimator.py

@@ -1,0 +1,110 @@
+import numpy as np
+from scipy.optimize import minimize
+
+from bdpn.tree_manager import TIME, annotate_tree
+
+MIN_VALUE = np.log(np.finfo(np.float64).eps)
+MAX_VALUE = np.log(np.finfo(np.float64).max)
+
+
+
+def rescale_log(loglikelihood_array):
+    """
+    Rescales the likelihood array if it gets too small/large, by multiplying it by a factor of e.
+    :param loglikelihood_array: numpy array containing the loglikelihood to be rescaled
+    :return: float, factor of e by which the likelihood array has been multiplies.
+    """
+
+    max_limit = MAX_VALUE
+    min_limit = MIN_VALUE
+
+    non_zero_loglh_array = loglikelihood_array[loglikelihood_array > -np.inf]
+    min_lh_value = np.min(non_zero_loglh_array)
+    max_lh_value = np.max(non_zero_loglh_array)
+
+    factors = 0
+    if max_lh_value > max_limit:
+        factors = max_limit - max_lh_value - 1
+    elif min_lh_value < min_limit:
+        factors = min(min_limit - min_lh_value + 1, max_limit - max_lh_value - 1)
+    loglikelihood_array += factors
+    return factors
+
+
+def optimize_likelihood_params(tree, input_parameters, loglikelihood, get_bounds_start):
+    """
+    Optimizes the likelihood parameters for a given forest and a given MTBD model.
+
+
+    :param forest: a list of ete3.Tree trees, annotated with node states and times via features STATE_K and TI.
+    :param T: time at end of the sampling period
+    :param model: MTBD model containing starting parameter values
+    :param optimise: MTBD model whose rates indicate which parameters need to optimized:
+        positive rates correspond to optimized parameters
+    :param u: number of hidden trees, where no tip got sampled
+    :return: the values of optimized parameters and the corresponding loglikelihood: (MU, LA, PSI, RHO, best_log_lh)
+    """
+    annotate_tree(tree)
+    T = max(getattr(_, TIME) for _ in tree)
+    bounds, start_parameters = get_bounds_start(tree, *input_parameters)
+
+    def get_real_params_from_optimised(ps):
+        ps = np.maximum(np.minimum(ps, bounds[:, 1]), bounds[:, 0])
+        result = np.zeros(len(input_parameters))
+
+        i = 0
+        for par_index, par in enumerate(input_parameters):
+            if par is None:
+                result[par_index] = ps[i]
[…]

### 2023-06-05 bdpn/tree_manager.py

@@ -1,0 +1,33 @@
+TIME = 'time'
+
+
+def annotate_tree(tree):
+    for n in tree.traverse('preorder'):
+        if n.is_root():
+            p_time = 0
+        else:
+            p_time = getattr(n.up, TIME)
+        n.add_feature(TIME, p_time + n.dist)
+
+
+def sort_tree(tree):
+    """
+    Reorganise the tree in such a way that the oldest tip (with the minimal time) is always on the left.
+    The tree must be time-annotated.
+
+    :param tree:
+    :return:
+    """
+    for n in tree.traverse('postorder'):
+        ot_feature = 'oldest_tip'
+        if n.is_leaf():
+            n.add_feature(ot_feature, getattr(n, TIME))
+            continue
+        c1, c2 = n.children
+        t1, t2 = getattr(c1, ot_feature), getattr(c2, ot_feature)
+        if t1 > t2:
+            n.children = [c2, c1]
+        delattr(c1, ot_feature)
+        delattr(c2, ot_feature)
+        if not n.is_root():
+            n.add_feature(ot_feature, min(t1, t2))

### 2023-06-05 setup.py

@@ -1,0 +1,28 @@
+from setuptools import setup, find_packages
+
+setup(
+    name='bdpn',
+    packages=find_packages(),
+    long_description=open('README.md').read(),
+    long_description_content_type='text/markdown',
+    classifiers=[
+        'Development Status :: 4 - Beta',
+        'Environment :: Console',
+        'Intended Audience :: Developers',
+        'Topic :: Scientific/Engineering :: Bio-Informatics',
+        'Topic :: Software Development',
+        'Topic :: Software Development :: Libraries :: Python Modules',
+    ],
+    version='0.1',
+    description='Estimation of BDPN parameters from phylogenetic trees.',
+    author='Anna Zhukova',
+    author_email='anna.zhukova@pasteur.fr',
+    url='https://github.com/evolbioinfo/bdpn',
+    keywords=['phylogenetics', 'multitype birth-death model', 'partner notification'],
+    install_requires=['six', 'ete3', 'numpy', 'scipy'],
+    entry_points={
+            'console_scripts': [
+                'bdpn_infer = bdpn.bdpn:main',
+            ]
+    },
+)

### 2023-06-05 simulations/py/plot_error.py

@@ -1,0 +1,167 @@
+import logging
+from collections import defaultdict
+
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sns
+from matplotlib.offsetbox import TextArea, HPacker, AnchoredOffsetbox, VPacker
+from statsmodels.stats.weightstats import CompareMeans
+
+RATE_PARAMETERS = ['lambda', 'psi', 'psi_p', 'p', 'pn']
+EPIDEMIOLOGIC_PARAMETERS = ['R_naught', 'infectious_time', 'partner_removal_time']
+par2greek = {'lambda': u'\u03bb', 'psi': u'\u03c8', 'psi_p': u'\u03c8p', 'p': '\u03c1', 'pn': '\u03c1n',
+             'R_naught': u'\u0052\u2080' + '=' + u'\u03bb\u002F\u03c8',
+             'infectious_time': 'infectious time 1' + u'\u002F\u03c8', 'partner_removal_time': 'partner removal time 1' + u'\u002F\u03c8p'}
+PARAMETERS = RATE_PARAMETERS + EPIDEMIOLOGIC_PARAMETERS
+
+if __name__ == "__main__":
+    import argparse
+
+    parser = argparse.ArgumentParser(description="Plots errors.")
+    parser.add_argument('--estimates', type=str, help="estimated parameters")
+    parser.add_argument('--pdf', type=str, help="plot")
+    parser.add_argument('--tab', type=str, help="error table")
+    params = parser.parse_args()
+
+    df = pd.read_csv(params.estimates, sep='\t', index_col=0)
+
+    real_df = df.loc[df['type'] == 'real', :]
+    df = df.loc[df['type'] != 'real', :]
+    types = sorted(df['type'].unique(), key=lambda _: next(i for (i, par) in enumerate(RATE_PARAMETERS) if par in _))
+
+    for type in types:
+        mask = df['type'] == type
+        for par in PARAMETERS:
+            # df.loc[mask, '{}_error'.format(par)] = (df.loc[mask, par] - real_df[par]) / real_df[par]
+            if par != 'p' and par != 'pn':
+                df.loc[mask, '{}_error'.format(par)] = (df.loc[mask, par] - real_df[par]) / real_df[par]
+            else:
+                df.loc[mask, '{}_error'.format(par)] = (df.loc[mask, par] - real_df[par])
+
[…]

### 2023-06-05 simulations/py/summary_table.py

@@ -1,0 +1,63 @@
+import logging
+import re
+
+import pandas as pd
+
+
+if __name__ == "__main__":
+    import argparse
+
+    parser = argparse.ArgumentParser(description="Summarize errors.")
+    parser.add_argument('--estimated_p', nargs='+', type=str, help="estimated parameters")
+    parser.add_argument('--estimated_psi', nargs='+', type=str, help="estimated parameters")
+    parser.add_argument('--real', nargs='+', type=str, help="real parameters")
+    parser.add_argument('--tab', type=str, help="estimate table")
+    params = parser.parse_args()
+
+    logging.getLogger().handlers = []
+    logging.basicConfig(level=logging.INFO, format='%(asctime)s: %(message)s', datefmt="%Y-%m-%d %H:%M:%S")
+
+    df = pd.DataFrame(columns=['type', 'sampled_tips',
+                               'lambda',
+                               'psi',
+                               'psi_p',
+                               'R_naught',
+                               'infectious_time',
+                               'p',
+                               'pn',
+                               'partner_removal_time'])
+
+    for real in params.real:
+        i = int(re.findall(r'[0-9]+', real)[0])
+        ddf = pd.read_csv(real)
+        # R0,infectious time,sampling probability,notification probability,removal time after notification,tips,time,hidden_trees
+        R0, it, p, pn, rt, tips, T, h_trees \
+            = ddf.loc[next(iter(ddf.index)), :]
+        df.loc['{}.real'.format(i),
+               ['R_naught', 'infectious_time', 'partner_removal_time',
+                'lambda', 'psi', 'psi_p', 'p', 'pn', 'sampled_tips', 'type']] \
+            = [R0, it, rt, R0 / it, 1 / it, 1 / rt, p, pn, tips, 'real']
+
+    for (est_list, fixed) in ((params.estimated_psi, 'psi'), (params.estimated_p, 'p')):
+        for est in est_list:
+            i = int(re.findall(r'[0-9]+', est)[0])
+            ddf = pd.read_csv(est)
+            est_label = 'BDPN({})'.format(fixed)
+            estimates = ddf.loc[next(iter(ddf.index)), :]
+            R0, it, p, pn, rt = ddf.loc[next(iter(ddf.index)), :]
+            df.loc['{}.{}'.format(i, est_label),
+            ['R_naught', 'infectious_time', 'partner_removal_time',
[…]

### 2023-06-05 simulations/py/tree_generator.py

@@ -1,0 +1,44 @@
+import numpy as np
+from treesimulator import save_log, save_forest
+from treesimulator.generator import generate
+from treesimulator.mtbd_models import BirthDeathModel, PNModel
+
+
+def random_float(min_value=0, max_value=1):
+    """
+    Generate a random float in ]min_value, max_value]
+    :param max_value: max value
+    :param min_value: min value
+    :return: the generated float
+    """
+    return min_value + (1 - np.random.random(size=1)[0]) * (max_value - min_value)
+
+
+if __name__ == "__main__":
+    import argparse
+
+    parser = argparse.ArgumentParser(description="Generates random BDPN tree.")
+    parser.add_argument('--log', default='/home/azhukova/projects/bdpn/simulations/trees/BDPN/tree.0.log', type=str, help="Nwk")
+    parser.add_argument('--nwk', default='/home/azhukova/projects/bdpn/simulations/trees/BDPN/tree.0.nwk', type=str, help="Log")
+    parser.add_argument('--min_tips', default=5000, type=int, help="Min number of tips")
+    parser.add_argument('--max_tips', default=10000, type=int, help="Max number of tips")
+    params = parser.parse_args()
+
+    R0 = random_float(1, 5)
+    psi = random_float(1 / 20, 1 / 5)
+    la = psi * R0
+    psi_ratio = random_float(10, 100)
+    psi_n = psi * psi_ratio
+    rho = random_float(0.01, 1)
+    rho_n = random_float(0.1, 1)
+
+    print(la, psi, psi_n, rho, rho_n)
+
+    model = PNModel(model=BirthDeathModel(p=rho, la=la, psi=psi), pn=rho_n, removal_rate=rho_n)
+
+    forest, (total_tips, u, T), _ = generate(model, params.min_tips, params.max_tips)
+
+    save_forest(forest, params.nwk)
+    save_log(model, total_tips, T, u, params.log)
+
+

### 2023-06-05 src/bd.py

@@ -1,172 +1,0 @@
-import numpy as np
-import pandas as pd
-from ete3 import Tree
-from scipy.optimize import NonlinearConstraint, LinearConstraint, minimize
-
-
-def annotate_tree(tree):
-    for n in tree.traverse('preorder'):
-        if n.is_root():
-            p_time = 0
-        else:
-            p_time = getattr(n.up, 'time')
-        n.add_feature('time', p_time + n.dist)
-
-
-def get_c1(la, psi, rho):
-    return np.power(np.power((la - psi), 2) + 4 * la * psi * rho, 0.5)
-
-
-def get_c2(la, psi, rho, T):
-    c1 = get_c1(la, psi, rho)
-    return (c1 + la - psi) / (c1 - la + psi) * np.exp(-T * c1)
-
-
-def get_c3(la, psi, rho, T, ti):
-    c1 = get_c1(la, psi, rho)
-    c2 = get_c2(la, psi, rho, T)
-    return c2 * np.exp(c1 * ti) + 1
-
-
-def get_u(t, la, psi, rho, T):
-    c1 = get_c1(la, psi, rho)
-    c2 = get_c2(la, psi, rho, T)
-    exp_t_c1 = np.exp(t * c1)
-    return (la + psi + c1 * (c2 * exp_t_c1 - 1) / (c2 * exp_t_c1 + 1)) / (2 * la)
-
-
-def get_p(t, la, psi, rho, T, ti):
-    c1 = get_c1(la, psi, rho)
-    c2 = get_c2(la, psi, rho, T)
-    c3 = get_c3(la, psi, rho, T, ti)
-    return (np.power(c3, 2) * np.exp(c1 * (t - ti))) / np.power(c2 * np.exp(c1 * t) + 1, 2)
-
-
-def loglikelihood(tree, la, psi, rho, T):
-    n = len(tree)
-    res = n * np.log(psi * rho) + (n - 1) * np.log(2 * la)
-    for n in tree.traverse('preorder'):
-        if not n.is_leaf():
-            ti = getattr(n, 'time')
-            c1, c2 = n.children
-            res += np.log(get_p(ti, la, psi, rho, T, getattr(c1, 'time'))) \
-                   + np.log(get_p(ti, la, psi, rho, T, getattr(c2, 'time')))
-    return res
-
-
-def optimize_likelihood_params(tree, T, la=None, psi=None, rho=None):
-    """
-    Optimizes the likelihood parameters for a given forest and a given MTBD model.
-
-
[…]

### 2023-06-05 src/bdpn.py

@@ -1,257 +1,0 @@
-import numpy as np
-import pandas as pd
-from ete3 import Tree
-from scipy.optimize import NonlinearConstraint, LinearConstraint, minimize
-
-
-def annotate_tree(tree):
-    for n in tree.traverse('preorder'):
-        if n.is_root():
-            p_time = 0
-        else:
-            p_time = getattr(n.up, 'time')
-        n.add_feature('time', p_time + n.dist)
-
-
-def get_c1(la, psi, rho):
-    return np.power(np.power((la - psi), 2) + 4 * la * psi * rho, 0.5)
-
-
-def get_c2(la, psi, rho, T):
-    c1 = get_c1(la, psi, rho)
-    return (c1 + la - psi) / (c1 - la + psi) * np.exp(-T * c1)
-
-
-def get_c3(la, psi, rho, T, ti):
-    c1 = get_c1(la, psi, rho)
-    c2 = get_c2(la, psi, rho, T)
-    return c2 * np.exp(c1 * ti) + 1
-
-
-def get_u(t, la, psi, rho, T):
-    c1 = get_c1(la, psi, rho)
-    c2 = get_c2(la, psi, rho, T)
-    exp_t_c1 = np.exp(t * c1)
-    return (la + psi + c1 * (c2 * exp_t_c1 - 1) / (c2 * exp_t_c1 + 1)) / (2 * la)
-
-
-def get_p(t, la, psi, rho, T, ti):
-    c1 = get_c1(la, psi, rho)
-    c2 = get_c2(la, psi, rho, T)
-    c3 = get_c3(la, psi, rho, T, ti)
-    return (np.power(c3, 2) * np.exp(c1 * (t - ti))) / np.power(c2 * np.exp(c1 * t) + 1, 2)
-
-
-def get_p_o(t, la, psi, rho, T, ti):
-    c1 = get_c1(la, psi, rho)
-    c2 = get_c2(la, psi, rho, T)
-    c3 = get_c3(la, psi, rho, T, ti)
-    return (c3 * np.exp(1/2 * (la + psi + c1) * (t - ti))) / (c2 * np.exp(c1 * t) + 1)
-
-
-def get_p_nh(t, la, psi, ti):
-    return np.exp((la + psi) * (t - ti))
-
-
-def get_lu(i, la, psi, psi_n, rho, rho_n, T, attribute=True):
-    ti = getattr(i, 'time')
-    tj = ti - i.dist
[…]

### 2023-06-05 tests/BDPNTest.py

@@ -1,0 +1,94 @@
+import numpy as np
+from matplotlib import pyplot as plt
+from scipy.integrate import odeint
+from treesimulator.mtbd_models import BirthDeathModel
+
+from bdpn.bdpn import get_log_p_o, get_log_p_nh
+from bdpn.bd import get_log_p, get_u
+from bdpn.mtbd import compute_U
+
+
+def plot_U(get_U, T, LA, PSI, RHO):
+    tt = np.linspace(0, T, 1001)
+    esol = [get_U(_)[0] for _ in tt]
+
+    y = [get_u(t, LA[0], PSI[0], RHO[0], T) for t in tt]
+
+    plt.plot(tt, esol, 'b', label='U_automatic(t)', alpha=0.5)
+    plt.plot(tt, y, 'r', label='U_formulas(t)', alpha=0.5)
+    plt.legend(loc='best')
+    plt.xlabel('t')
+    plt.grid()
+    plt.show()
+
+
+def plot_P(get_U, ti, t0, MU, LA, PSI, RHO, T):
+    tt = np.linspace(ti, t0, 1001)
+
+    def pdf_P(P, t):
+        U = get_U(t)
+        return (MU.sum(axis=1) + LA.dot(1 - U) + PSI) * P - (MU + U * LA).dot(P)
+
+    y0 = np.zeros(len(PSI), np.float64)
+    y0[0] = 1
+    sol = odeint(pdf_P, y0, tt)
+
+    y = [np.exp(get_log_p(t, LA[0], PSI[0], RHO[0], T, ti)) for t in tt]
+
+    plt.plot(tt, sol[:, 0], 'b', label='P_automatic(t)', alpha=0.5)
+    plt.plot(tt, y, 'r:', label='P_formulas(t)', alpha=0.5)
+
+    def pdf_Po(P, t):
+        U = get_U(t)
+        return (MU.sum(axis=1) + LA.dot(1 - U) + PSI) * P - MU.dot(P)
+
+    y0 = np.zeros(len(PSI), np.float64)
+    y0[0] = 1
+    sol = odeint(pdf_Po, y0, tt)
+    y = [np.exp(get_log_p_o(t, LA[0], PSI[0], RHO[0], T, ti)) for t in tt]
+
+    plt.plot(tt, sol[:, 0], 'g', label='Po_automatic(t)', alpha=0.5)
[…]

### 2023-06-05 tests/BDTest.py

@@ -1,0 +1,66 @@
+import numpy as np
+from matplotlib import pyplot as plt
+from scipy.integrate import odeint
+from treesimulator.mtbd_models import BirthDeathModel
+
+from bdpn.bd import get_log_p, get_u
+from bdpn.mtbd import compute_U
+
+
+def plot_U(get_U, T, LA, PSI, RHO):
+    tt = np.linspace(0, T, 1001)
+    esol = [get_U(_)[0] for _ in tt]
+
+    y = [get_u(t, LA[0], PSI[0], RHO[0], T) for t in tt]
+
+    plt.plot(tt, esol, 'b', label='U_automatic(t)', alpha=0.5)
+    plt.plot(tt, y, 'r', label='U_formulas(t)', alpha=0.5)
+    plt.legend(loc='best')
+    plt.xlabel('t')
+    plt.grid()
+    plt.show()
+
+
+def plot_P(get_U, ti, t0, MU, LA, PSI, RHO, T):
+    tt = np.linspace(ti, t0, 1001)
+
+    def pdf_Pany_l(P, t):
+        U = get_U(t)
+        return (MU.sum(axis=1) + LA.dot(1 - U) + PSI) * P - (MU + U * LA).dot(P)
+
+    y0 = np.zeros(len(PSI), np.float64)
+    y0[0] = 1
+    sol = odeint(pdf_Pany_l, y0, tt)
+
+    y = [np.exp(get_log_p(t, LA[0], PSI[0], RHO[0], T, ti)) for t in tt]
+
+    plt.plot(tt, sol[:, 0], 'b', label='P_automatic(t)', alpha=0.5)
+    plt.plot(tt, y, 'r', label='P_formulas(t)', alpha=0.5)
+    plt.legend(loc='best')
+    plt.xlabel('t')
+    plt.grid()
+    plt.show()
+
+
+def random_bt_0_and_1():
+    return 1 - np.random.random(size=1)[0]
+
+
+if __name__ == '__main__':
+
+    p = random_bt_0_and_1()
+    R0 = random_bt_0_and_1() * 5
+    real_psi = random_bt_0_and_1()
+    real_la = real_psi * R0
+    T = 20
+    model = BirthDeathModel(la=real_la, psi=real_psi, p=p)
+
[…]

### 2023-06-09 article/article.tex
Added cherry tests

@@
 \usepackage[utf8]{inputenc}
-\usepackage{amsmath,mathrsfs}
+\usepackage{amsmath,mathrsfs}\usepackage{amsmath,amsfonts,amssymb,float,graphicx,mathtools,mathrsfs,tikz,amsthm}
 
@@
 \usepackage{natbib}
+\usepackage{booktabs}
 
@@
 We also propose a test allowing to distinguish between MTBD and MTBD-PN models.
-Finally, we apply it to the study of HIV epidemic in Cuba.
+Finally, we apply it to the study of HIV epidemic in the UK.
 
@@
 
+\subsection{Extension to forests}
+
+As we previously described in~\citep{zhukovaFastAccurateMaximumLikelihood2022}, the likelihood calculation with MTBD models can be easily extended to forests.
+This extension applies also to MTBD-PN models.
+Forests are useful for cases when a (sub-)epidemic started with several infected individuals (e.g.
+dur to multiple pathogen introductions to a country of interest or due to a change of health policies leading to a change in parameter values).
+
+In this case the (sub-)epidemic leads to a forest $\mathscr{F}$ of $f$ observed trees: $\mathscr{T}_1, \ldots, \mathscr{T}_f$.
+If all the trees started at the same time (e.g.
+due to health policy change), the forest might also include a certain number $u$ of unobserved trees, i.e., individuals who were infected at the beginning of the (sub-)epidemic, but whose trees stayed unobserved  as none of their tips got sampled.
+Forest likelihood formula hence combines the likelihoods of $f$ observed and $u$ hidden trees, and can be represented in logarithmic form~(\ref{eq:forest_likelihood}).
+Tree likelihood formula %[\ref{eq:tree_likelihood}]
+is its special case, where $f=1$ and $u=0$.
+
+%As our application to Ebola data (see~Application) shows, $u$ is an important parameter: while the results are not very sensible to slight $u$ variations, ignoring it completely (e.g.
+using $u=0$ instead of $u \approx 500$ in our example) can change the inferences.
+
+\begin{equation}
+logL(\mathscr{F}|\Theta,u)=u\,logU_{hidden}(\Theta) + \sum\limits_{j=1}^f logL(\mathscr{T}_j|\Theta), \textit{ where } U_{hidden}(\Theta)=\sum\limits_{s=1}^{d}\pi_s U_s(0) \label{eq:forest_likelihood}
+\end{equation}
+
+If all the sub-epidemics started at the same time, for given model parameter values $\Theta$ we can estimate the number of hidden trees $u$ from the number of observed trees $f$ as:
+\begin{equation}
+u = f \frac{U_{hidden}(\Theta)}{1 - U_{hidden}(\Theta)}\label{eq:u} 
+\end{equation}
+
 \section{Tree simulator}
@@
 
+\section{PN test}
+To access whether a give tree is generated under a classical MTBD model or MTBD-PN, we developed a non-parametric test based on cherries (two tips having a common parent). 
+The intuition behind the test is that in the presence of partner notification the tree will contain more cherries whose tips are close in time.
+
+The test sorts the cherries by the times of their roots, and splits them into blocks of 100 cherries.
+For each cherry in a block, the test calculates the difference between its tip times.
+It then generates a collection of random cherry tip differences for the same block: For each original cherry root it picks 5 cherries with the roots that are the closest in time, randomly selects two tips among their tips, and calculates their time difference.
+The random cherry tip difference generation is repeated 100 times.
+Finally, the test reports the proportion of cases where the first quantile of the random cherry differences was smaller than that of the real cherries.
+
+The test therefore reports a probability of partner notification being present at time interval corresponding to each block.
+To estimate the probability of partner notification being present in the whole tree, it averages the block probabilities.
+
+\subsection{Performance on simulated data}
+To illustrate the PN test performance, we applied it to the trees simulated under BD and under BDPn models.
+
+... details coming soon ...
+
+\section{BD-PN parameter and CI estimator}
+We implemented a parameter estimator for the BD-PN model (which we called bdpn).
+It estimates the BD-PN model parameters  $\Theta = (\lambda,\psi,\psi_n,\rho,\rho_n) \in \mathbb{R}^5$ for a forest $\mathscr{F}$ comprising $f \geq 1$ observed trees in the maximum-likelihood framework, where one of the BD parameters ($\lambda,\psi$ or $\rho$) is fixed (for identifiability reasons).
+
+Once the optimal parameter values are found, we calculate their confidence intervals (CIs) using Wilks' method~\citep{Wilks1938}.
+For each non-fixed parameter $p \in \Theta$, we calculate its $95\%$-CI as including the values $\tilde{p}$ such that $log L(\mathscr{F}|\Theta_{opt|p=\tilde{p}}) > log L(\mathscr{F}| \Theta_{opt}) - \chi^2_1(0.95) / 2$, where $\chi^2_1(0.95)$ is the value of chi-squared distribution with 1 degree of freedom corresponding to the significance level of $0.95$ (i.e., $\sim3.84$).
+$\Theta_{opt|p=\tilde{p}}$ corresponds to the maximum-likelihood value for the other non-fixed parameters when $p = \tilde{p}$.
+
+\subsection*{Code availability}
+The BD-PN parameter estimator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\cite{Huerta-Cepas2016} and NumPy package for array operations~\cite{harris_array_2020}.
+
+It is available as a command-line program and a Python 3 package via PyPi (\href{https://pypi.org/project/bdpn}{bdpn}), and via Docker/Singularity (\href{https://hub.docker.com/r/evolbioinfo/bdpn/tags}{evolbioinfo/bdpn}).
+Its source code and the installation and usage documentation are available on GitHub at \href{https://github.com/evolbioinfo/bdpn}{github.com/evolbioinfo/bdpn}.
+
+\subsection{Performance on simulated data}
+
+To assess the performance of our maximum-likelihood estimator for the BD-PN model, we simulated 100 trees with 5\,000--10\,000 tips under the BD-PN model with our tree simulator.
+For each tree the parameter values were drawn randomly within the following bounds:
+$R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
+removal rate $\psi \in ]1 / 20, 1 / 5]$,
+ratio between the partner removal rate after notification and the removal rate $\frac{\psi_n}{\psi} \in ]10, 100]$,
+sampling probability $\rho \in ]0.01, 0.9]$,
+partner notification probability $\rho_n \in ]0.1, 0.9]$.
+We applied bdpn to each of the trees three times: fixing each of the BD parameters ($\lambda,\psi$ or $\rho$) to its real value (for identifiability).
+
+As expected with a maximum likelihood estimator, the tree likelihoods for the estimated parameter values were higher than the tree likelihoods for the real parameter values in all the settings.
+
+We calculated the relative error (normalized distance between the estimated and the target values: $\frac{|estimated - target|}{target}$) and the relative bias ($\frac{estimated - target}{target}$) for each parameter on each tree (Fig.~\ref{fig:sim}).
+Average relative errors were the lowest when $\psi$ was fixed: $\leq 5\%$  for all the parameters but $\psi_n$, which turned to be the hardest parameter to estimate in all the settings (relative errors $\approx 15\%$, with a tendency to underestimation).
+The worst performance was obtained with $\lambda$ being fixed: in this setting the estimator tends to underestimate $\psi$ and overestimate $\rho$.
+As with the point estimates, CI estimates were the best when $\psi$ was fixed: the target values of $\lambda, \psi_{n} \rho$ and $\rho_n$ were within the estimated CIs in correspondingly 97\%, 95\%, 87\%, and 76\% of cases.
+The values for the other settings are shown in Table~\ref{tbl:ci}.
+
+\begin{figure}[!pht]
+\centering 
+\includegraphics[width=1\textwidth]{Fig_errors.eps}
+\caption{Comparison of inference accuracy of bdpn on a data set of 100 trees of 5\,000--10\,000 tips each, with either $\lambda$ (blue), $\psi$ (orange) or $\rho$ (green) parameter being fixed to its real value.
+We show the swarmplots (coloured by fixed parameter) of relative errors for each test tree and parameter, which are measured as the normalized distance between the estimate and the real value.
+Average relative error (and in parentheses average bias, calculated on normalized values) are displayed for each parameter and method below their swarmplot.
+For $\rho$ absolute values are shown instead of normalized ones.
+}
+\label{fig:sim} 
+\end{figure}
+
+ \begin{table}[!h]\centering
+\small
+\caption{Percentage of simulated trees for which the real parameter values were withing the estimated 95\% CIs. \smallskip}
+\begin{tabular}{c|ccc}
+\textbf{parameter} & \textbf{$\lambda$ fixed} & \textbf{$\psi$ fixed} & \textbf{$\rho$ fixed} \\
+\toprule 
+ $\lambda$ & -- &  97\% & 96\% \\
+ $\psi$ & 78\% & -- & 83\% \\
+ $\rho$ & 28\% & 87\%  & -- \\
+ $\psi_n$ & 94\% & 95\% & 95\% \\
+ $\rho_n$ & 75\% & 76\% & 77\% \\
+\bottomrule
+\end{tabular}
+\label{tbl:ci}
+\end{table}
+
+\subsection{Application: HIV-B epidemic in the UK}
+We applied our estimator to asses the partner notification in HIV-infected patients in the UK.
+We used the dated phylogenetic tree from our recent study of HIV drug resistance in the UK~\citep{zhukovaModelingDrugResistance2023}.
+The tree represents the UK HIV-B epidemic between 1960s and 2017, its tips represent viruses sampled from 39\,159 individuals between 1996 and 2016.
+The samples come from the UK HIV Drug resistance database~\citep{Dunn2007}, which stores sequences from about 50\% of infected individuals in the UK~\citep{zhukovaModelingDrugResistance2023}.
+We therefore estimated the sampling probability $\rho$ as 0.5.
+
+We cut this tree at year 2005, and estimated the BD-PN parameters on the 2005-on forests, in order to have a more homogeneous setting in terms of access to antiretroviral treatment, awareness of the HIV infection, etc.
+The 2005-on forest contained 18\,483 trees with a total of 31\,648 tips.
+We estimated the following parameter values: $\lambda=0.262$, $\psi=0.186$, $\psi_n=186.921$, $\rho_n=0.033$, which corresponds to $R_e = 1.4$, infectious time $\frac{1}{\psi} = 5.4$ years, and about 2 days between partner notification and their sampling.
+
+The average time of HIV progression to AIDs in the absence of treatment is 10 years, which could serve as an estimate of the infectious time.
+However, in the UK, 95\% of infected population is on antiretroviral treatment, which, when taken properly, prevents transmission.
+Hence, the average infectious time represents the average time before starting treatment.
+Our estimate is approximately 5 years.
+
+To assess the robustness of these predictions with respect to $\rho$, we additionally estimated the parameter values for $\rho=0.4$ and for $\rho=0.6$.
+The results were very similar: $R_e$ of respectively $1.5$ and
+$1.3$, infectious time of $5.1$ and $5.6$ years, partner notification probability of $0.033$ in both cases and about 2 days between partner notification and their sampling.
+
 \end{document}

### 2023-06-09 bdpn/bd.py

@@ -1,6 +1,5 @@
 import numpy as np
 
-from bdpn.parameter_estimator import get_rough_rate_etimates
 from bdpn.tree_manager import TIME, annotate_tree
 
 
@@ -47,41 +46,25 @@ def get_log_p(t, la, psi, rho, T, ti):
     return 2 * np.log(c3) + (c1 * (t - ti)) - 2 * np.log((c1 + la - psi) / (c1 - la + psi) * np.exp(c1 * (t - T)) + 1)
 
 
-def loglikelihood(tree, la, psi, rho, T=None):
-    if not hasattr(tree, TIME):
-        annotate_tree(tree)
+def loglikelihood(forest, la, psi, rho, T=None):
+    for tree in forest:
+        if not hasattr(tree, TIME):
+            annotate_tree(tree)
     if T is None:
-        T = max(getattr(_, TIME) for _ in tree)
+        T = 0
+        for tree in forest:
+            T = max(T, max(getattr(_, TIME) for _ in tree))
 
-    n = len(tree)
-    res = n * np.log(psi * rho) + (n - 1) * np.log(2 * la)
-    for n in tree.traverse('preorder'):
-        if not n.is_leaf():
-            ti = getattr(n, TIME)
-            c1, c2 = n.children
-            res += get_log_p(ti, la, psi, rho, T, getattr(c1, TIME)) \
-                   + get_log_p(ti, la, psi, rho, T, getattr(c2, TIME))
+    res = 0
+    for tree in forest:
+        n = len(tree)
+        res += n * np.log(psi * rho) + (n - 1) * np.log(2 * la)
+        for n in tree.traverse('preorder'):
+            if not n.is_leaf():
+                ti = getattr(n, TIME)
+                c1, c2 = n.children
+                res += get_log_p(ti, la, psi, rho, T, getattr(c1, TIME)) \
+                       + get_log_p(ti, la, psi, rho, T, getattr(c2, TIME))
     u = get_u(0, la, psi, rho, T)
-    return res + u / (1 - u) * np.log(u)
-    # return res
+    return res + len(forest) * u / (1 - u) * np.log(u)
 
-
-def get_bounds_start(tree, la, psi, rho):
-    bounds = []
-    l_ext = []
-    l_int = []
-    for n in tree.traverse('preorder'):
-        if n.dist:
[…]

### 2023-06-09 bdpn/bdpn.py

@@ -1,10 +1,10 @@
 import os
 
 import numpy as np
-from ete3 import Tree
+from pastml.tree import read_forest
 
 from bdpn.bd import get_c1, get_c3, get_log_p, get_u
-from bdpn.parameter_estimator import get_rough_rate_etimates, optimize_likelihood_params, rescale_log
+from bdpn.parameter_estimator import optimize_likelihood_params, rescale_log
 from bdpn.tree_manager import TIME, annotate_tree
 
 
@@ -17,111 +17,84 @@ def get_log_p_o(t, la, psi, rho, T, ti):
     return np.log(c3) + (1 / 2 * (la + psi + c1) * (t - ti)) \
         - np.log((c1 + la - psi) / (c1 - la + psi) * np.exp(c1 * (t - T)) + 1)
 
+
 def get_log_p_nh(t, la, psi, ti):
     # return np.exp((la + psi) * (t - ti))
     return (la + psi) * (t - ti)
 
 
-def loglikelihood(tree, la, psi, psi_n, rho, rho_n, T=None):
-    if not hasattr(tree, TIME):
-        annotate_tree(tree)
+def loglikelihood(forest, la, psi, psi_n, rho, rho_n, T=None):
+    for tree in forest:
+        if not hasattr(tree, TIME):
+            annotate_tree(tree)
     if T is None:
-        T = max(getattr(_, TIME) for _ in tree)
-
-    for n in tree.traverse():
-        ti = getattr(n, TIME)
-        t = ti - n.dist
-        n.add_feature('p', get_log_p(t, la, psi, rho, T, ti))
-        n.add_feature('po', get_log_p_o(t, la, psi, rho, T, ti))
-        if n.is_leaf():
-            n.add_feature('pnh', get_log_p_nh(t, la, psi, ti))
+        T = 0
+        for tree in forest:
+            T = max(T, max(getattr(_, TIME) for _ in tree))
 
     log_sampled = np.log(psi) + np.log(rho)
     log_sampled_and_notified = log_sampled + np.log(rho_n)
     log_sampled_not_notified = log_sampled + np.log(1 - rho_n)
-    for n in tree.traverse('postorder'):
-        ti = getattr(n, TIME)
-        t = ti - n.dist
-        if n.is_leaf():
-            n.add_feature('LU', getattr(n, 'p') + log_sampled_not_notified)
-            n.add_feature('LN', getattr(n, 'pnh') + log_sampled_and_notified)
[…]

### 2023-06-09 bdpn/model_distinguisher.py

@@ -9,10 +9,11 @@ import numpy as np
 import scipy
 from matplotlib import rcParams
 from matplotlib.pyplot import hist, legend, savefig, clf, title, show
+from pastml.tree import read_forest
 from scipy.stats import stats
 from wquantiles import quantile
 
-from bdpn.tree_manager import TIME
+from bdpn.tree_manager import TIME, annotate_tree
 
 
 class SortedSet():
@@ -329,89 +330,57 @@ def pick_communities(tree, pn_delay):
         yield Motif(clustered_tips=clustered_tips)
 
 
-
-def pick_motifs(tree, pn_lookback, pn_delay, include_trivial=True, list_all_potential_notifiers=True):
+def main():
     """
-    Picks motifs that satisfy the given values of pn_lookback, pn_delay in the given tree.
-    :param tree: ete3.Tree, the tree of interest
-    :param pn_lookback: float, max value of PN lookback
-    (how old can a transmission be for a detected case to notify this partner)
-    :param pn_delay: float, max delay between the notification and the sampling
-    :param include_trivial: bool, whether to include trivial one-node motifs in the output
-    :param list_all_potential_notifiers: bool, whether to (True) list all potential notifiers
-    for each detected notified node, or (False) just one (the closest in time)
-    :return: iterator of Motif motifs
+    Entry point for PN test with command-line arguments.
+    :return: void
     """
-    # sorted by TIME
-    sorted_tips = SortedSet(tree.get_leaves(), key=lambda _: getattr(_, TIME))
+    import argparse
 
-    while sorted_tips:
-        # the oldest unprocessed tip
-        index_tip = sorted_tips.pop()
-        index_TIME = getattr(index_tip, TIME)
+    parser = \
+        argparse.ArgumentParser(description="Assess if the tree was generated under a -PN model.")
+    parser.add_argument('--log', required=True, type=str, help="output log file")
+    parser.add_argument('--nwk', required=True, type=str, help="input tree file")
+    parser.add_argument('--block_size', default=100, type=int, help="number of cherries per block")
+    parser.add_argument('--report_blocks', action="store_true", help="report p-values for each block as well")
+    params = parser.parse_args()
 
-        clustered_tips = [index_tip]
+    forest = read_forest(params.nwk)
+    for tree in forest:
+        if not hasattr(tree, TIME):
+            annotate_tree(tree)
 
-        cur_root = climb_up_tree(index_tip, pn_lookback)
[…]

### 2023-06-09 bdpn/parameter_estimator.py

@@ -1,5 +1,6 @@
 import numpy as np
 from scipy.optimize import minimize
+from scipy.stats import stats
 
 from bdpn.tree_manager import TIME, annotate_tree
 
@@ -32,7 +33,7 @@ def rescale_log(loglikelihood_array):
     return factors
 
 
-def optimize_likelihood_params(tree, input_parameters, loglikelihood, bounds, start_parameters):
+def optimize_likelihood_params(forest, input_parameters, loglikelihood, bounds, start_parameters, cis=False):
     """
     Optimizes the likelihood parameters for a given forest and a given MTBD model.
 
@@ -45,10 +46,12 @@ def optimize_likelihood_params(tree, input_parameters, loglikelihood, bounds, st
     :param u: number of hidden trees, where no tip got sampled
     :return: the values of optimized parameters and the corresponding loglikelihood: (MU, LA, PSI, RHO, best_log_lh)
     """
-    if not hasattr(tree, TIME):
-        annotate_tree(tree)
-    T = max(getattr(_, TIME) for _ in tree)
-    # bounds, start_parameters = get_bounds_start(tree, *input_parameters)
+    for tree in forest:
+        if not hasattr(tree, TIME):
+            annotate_tree(tree)
+    T = 0
+    for tree in forest:
+        T = max(T, max(getattr(_, TIME) for _ in tree))
     # print('Bounds are {}'.format(bounds))
     # print('Starting parameter values are {}'.format(start_parameters))
 
@@ -76,7 +79,7 @@ def optimize_likelihood_params(tree, input_parameters, loglikelihood, bounds, st
         if np.any(np.isnan(ps)):
             return np.nan
         ps = get_real_params_from_optimised(ps)
-        res = loglikelihood(tree, *ps, T)
+        res = loglikelihood(forest, *ps, T)
         # print("{}\t-->\t{:g}".format(ps, res))
         return -res
 
@@ -99,19 +102,45 @@ def optimize_likelihood_params(tree, input_parameters, loglikelihood, bounds, st
                 best_log_lh = -fres.fun
                 break
         # print('Attempt {} of trying to optimise the parameters: {}.'.format(i, -fres.fun))
-    return get_real_params_from_optimised(x0)
+    optimised_parameters = get_real_params_from_optimised(x0)
 
+    n = len(optimised_parameters)
+    optimised_cis = np.zeros((n, 2))
+    optimised_cis[:, 0] = np.array(optimised_parameters)
[…]

### 2023-06-09 examples/cut_tree.py

@@ -1,0 +1,41 @@
+import pandas as pd
+from ete3 import Tree
+
+if __name__ == "__main__":
+    import argparse
+
+    parser = argparse.ArgumentParser(description="Splits a tree into the root part and the bottom forest.")
+    parser.add_argument('--in_nwk', required=True, type=str, help="Nwk tree to cut")
+    parser.add_argument('--out_forest', type=str, help="Nwk containing a forest of cut subtrees")
+    parser.add_argument('--out_root', type=str, help="Nwk containing a the root subtree")
+    parser.add_argument('--threshold', type=float, help="Date at which to cut")
+    params = parser.parse_args()
+
+    tree = Tree(params.in_nwk)
+    for n in tree.traverse('preorder'):
+        parent_time = 0 if n.is_root() else getattr(n.up, 'time')
+        n.add_feature('time', parent_time + n.dist)
+    height = max(getattr(t, 'time') for t in tree)
+    df.loc[0, 'T'] = height * (1 - params.threshold)
+    height = height * params.threshold
+    # tips	f	u	T	mu	la	psi	p
+    f = 0
+    tips = 0
+    with open(params.out_nwk, 'w+') as file:
+        todo = [tree]
+        while todo:
+            n = todo.pop()
+            time = getattr(n, 'time')
+            if time >= height:
+                n.detach()
+                n.dist = time - height
+                file.write(n.write(format_root_node=True, format=5) + '\n')
+                f += 1
+                tips += len(n)
+            else:
+                todo.extend(n.children)
+    df.loc[0, 'f'] = f
+    df.loc[0, 'tips'] = tips
+    rho = df.loc[0, 'p']
+    df.loc[0, 'u'] = round(f / rho * (1 - rho))
+    df.to_csv(params.out_log, sep='\t', index=False)

### 2023-06-09 examples/main.py

@@ -25,27 +25,27 @@ psi_n = 1 / rt
 la = R0 * psi
 
 
-lk_real_bdpn = bdpn.loglikelihood(tree, la, psi, psi_n, rho, rho_n)
-lk_real_bd = bd.loglikelihood(tree, la, psi, rho)
+lk_real_bdpn = bdpn.loglikelihood([tree], la, psi, psi_n, rho, rho_n)
+lk_real_bd = bd.loglikelihood([tree], la, psi, rho)
 print('Real params: {}\t loglikelihood-bdpn: {}\t loglikelihood-bd: {}'.format([la, psi, psi_n, rho, rho_n], lk_real_bdpn, lk_real_bd))
 
 input_parameters = np.array([None, None, None, rho, None])
 # bounds = np.array([[la / 10, la * 10], [psi / 10, psi * 10], [psi_n / 10, psi_n * 10], [1e-3, 1], [0, 1]])
 bounds = np.array([[0.05, 1], [0.05, 0.2], [0.5, 20], [0.01, 0.99999], [0.1, 0.99999]])
 start_parameters = (bounds[:, 0] + bounds[:, 1]) / 2
-vs_bdpn = optimize_likelihood_params(tree, input_parameters=input_parameters,
+vs_bdpn, cis_bdpn = optimize_likelihood_params([tree], input_parameters=input_parameters,
                                      loglikelihood=bdpn.loglikelihood,
                                      bounds=bounds[input_parameters == None],
-                                     start_parameters=start_parameters)
-lk_bdpn = bdpn.loglikelihood(tree, *vs_bdpn)
-print('Found BDPN params: {}\t loglikelihood: {}, AIC: {}'.format(vs_bdpn, lk_bdpn, AIC(4, lk_bdpn)))
+                                     start_parameters=start_parameters, cis=True)
+lk_bdpn = bdpn.loglikelihood([tree], *vs_bdpn)
+print('Found BDPN params: {}\n\t{}\t loglikelihood: {}, AIC: {}'.format(vs_bdpn, cis_bdpn, lk_bdpn, AIC(4, lk_bdpn)))
 
 bd_indices = [0, 1, 3]
 input_parameters_bd = input_parameters[bd_indices]
 bounds_bd = bounds[bd_indices]
 start_parameters_bd = (bounds_bd[:, 0] + bounds_bd[:, 1]) / 2
-vs_bd = optimize_likelihood_params(tree, input_parameters=input_parameters_bd,
[…]

### 2023-06-09 examples/main_cherry_test.py

@@ -1,6 +1,5 @@
 import logging
 import os
-from collections import defaultdict
 
 from ete3 import Tree
 
@@ -10,56 +9,28 @@ from bdpn.tree_manager import annotate_tree, TIME
 DATA_DIR = os.path.abspath(os.path.join(os.path.dirname(__file__), '..', 'trees'))
 NWK_BD = os.path.join(DATA_DIR, 'bd', 'tree.1.nwk')
 NWK_BDPN = os.path.join(DATA_DIR, 'bdpn', 'tree.1.nwk')
-min_cherry_num = 50
+min_cherry_num = 100
 
 if __name__ == '__main__':
     logging.basicConfig(level=logging.INFO, format='%(asctime)s: %(message)s', datefmt="%Y-%m-%d %H:%M:%S")
 
     for nwk in (NWK_BD, NWK_BDPN):
         logging.info('==============={}============'.format(nwk.replace(DATA_DIR, '')))
+
+        results = []
         tree = Tree(nwk)
         annotate_tree(tree)
-        T = max(getattr(_, TIME) for _ in tree)
-        dT = T / 100
 
-        all_cherries = list(pick_cherries(tree, include_polytomies=True))
+        all_cherries = sorted(list(pick_cherries(tree, include_polytomies=True)), key=lambda _: getattr(_.root, TIME))
         logging.info('Picked {} cherries with {} roots.'.format(len(all_cherries), len({_.root for _ in all_cherries})))
-        year2cherries = defaultdict(list)
-        for _ in all_cherries:
-            year2cherries[getattr(_.root, TIME) // dT].append(_)
-
-        total_cherries = len(all_cherries)
-
-        year = min(year2cherries.keys())
-        cherries = []
-        max_year = max(year2cherries.keys())
-        while year <= max_year:
-            start_year = year
-            while len(cherries) < min_cherry_num and year <= max_year:
-                cherries.extend(year2cherries[year])
-                year += 1
-            # drop the last few cherries if they are not the only ones we have
-            if len(cherries) < min_cherry_num / 2 and min_cherry_num <= total_cherries:
-                break
-            result = nonparametric_cherry_diff(cherries, repetitions=1e4)
+        for i in range(0, len(all_cherries) // min_cherry_num):
+            cherries = all_cherries[i * min_cherry_num: (i + 1) * min_cherry_num]
+            result = nonparametric_cherry_diff(cherries, repetitions=1e3)
+            results.append(result)
 
             logging.info(
[…]

### 2023-06-09 examples/main_cuba.py

@@ -13,7 +13,7 @@ DATA_DIR = os.path.abspath(os.path.join(os.path.dirname(__file__), '..', 'trees'
 
 # gotree resolve -i D_CRF_19_timetree.nwk | gotree brlen setmin -l 0.0027397260273972603 -o D_CRF_19_timetree.resolved.nwk
 NWK_CUBA = os.path.join(DATA_DIR, 'D_CRF_19_timetree.resolved.nwk')
-min_cherry_num = 50
+min_cherry_num = 25
 
 if __name__ == '__main__':
     logging.basicConfig(level=logging.INFO, format='%(asctime)s: %(message)s', datefmt="%Y-%m-%d %H:%M:%S")

### 2023-06-09 examples/main_uk_psi.py

@@ -17,26 +17,6 @@ NWK_UK = os.path.join(DATA_DIR, 'UK_B_raxml.lsd2.resolved.nwk')
 min_cherry_num = 50
 
 
-def save_estimates(log, vs, cis):
-    with open(log, 'w+') as f:
-        f.write('\t{}\n'.format(','.join(['R0', 'infectious time', 'sampling probability', 'notification probability',
-                                          'removal time after notification',
-                                          'transmission rate', 'removal rate', 'partner removal rate'])))
-        la, psi, psi_p, rho, rho_p = vs
-        R0 = la / psi
-        rt = 1 / psi
-        prt = 1 / psi_p
-        (la_min, la_max), (psi_min, psi_max), (psi_p_min, psi_p_max), (rho_min, rho_max), (rho_p_min, rho_p_max) = cis
-        R0_min, R0_max = la_min / psi, la_max / psi
-        rt_min, rt_max = 1 / psi_max, 1 / psi_min
-        prt_min, prt_max = 1 / psi_p_max, 1 / psi_p_min
-        f.write('value,{}\n'.format(','.join(str(_) for _ in [R0, rt, rho, rho_p, prt, la, psi, psi_p])))
-        f.write('CI_min,{}\n'.format(
-            ','.join(str(_) for _ in [R0_min, rt_min, rho_min, rho_p_min, prt_min, la_min, psi_min, psi_p_min])))
-        f.write('CI_max,{}\n'.format(
-            ','.join(str(_) for _ in [R0_max, rt_max, rho_max, rho_p_max, prt_max, la_max, psi_max, psi_p_max])))
-
-
 def cut_tree(tree, threshold_date):
     forest = []
     todo = [tree]
@@ -105,11 +85,11 @@ if __name__ == '__main__':
     cut_year = 2005
     root, forest = cut_tree(tree, cut_year)
 
-    for rho in (0.4, 0.5, 0.6, 0.7):
-        print("=======rho={}============".format(rho))
+    for psi in (1/3, 1/5, 1/7, 1/9):
+        print("=======psi={}============".format(psi))
         bounds = np.array([[0.1, 12], [0.05, 1], [0.5, 365], [0.0001, 0.99999], [0.0001, 0.99999]])
         start_parameters = (bounds[:, 0] + bounds[:, 1]) / 2
-        input_parameters = np.array([None, None, None, rho, None])
+        input_parameters = np.array([None, psi, None, None, None])
[…]

### 2023-06-09 setup.py

@@ -23,6 +23,7 @@ setup(
     entry_points={
             'console_scripts': [
                 'bdpn_infer = bdpn.bdpn:main',
+                'pn_test = bdpn.model_distinguisher:main',
             ]
     },
 )

### 2023-06-09 simulations/py/assess_CIs.py

@@ -1,0 +1,45 @@
+import os
+
+import numpy as np
+import pandas as pd
+
+RATE_PARAMETERS = ['lambda', 'psi', 'psi_p']
+PARAMETERS = RATE_PARAMETERS + ['p', 'pn']
+
+
+if __name__ == "__main__":
+    import argparse
+
+    parser = argparse.ArgumentParser(description="Access CIs.")
+    parser.add_argument('--estimates', type=str, help="estimated parameters",
+                        default=os.path.join(os.path.dirname(__file__), '..', 'trees', 'BDPN', 'estimates.tab'))
+    parser.add_argument('--log', type=str, help="output log",
+                        default=os.path.join(os.path.dirname(__file__), '..', 'trees', 'BDPN', 'CIs.log'))
+    params = parser.parse_args()
+
+    df = pd.read_csv(params.estimates, sep='\t', index_col=0)
+
+    real_df = df.loc[df['type'] == 'real', :]
+    df = df.loc[df['type'] != 'real', :]
+    with open(params.log, 'w+') as f:
+        for type in df['type'].unique():
+            mask = df['type'] == type
+            print('\n================{}==============='.format(type))
+            f.write('\n================{}===============\n'.format(type))
+            n_observations = sum(mask)
+            for par in PARAMETERS:
+                df.loc[mask, '{}_within_CI'.format(par)] \
+                    = (np.less_equal(df.loc[mask, '{}_min'.format(par)], real_df[par])
+                       | np.less_equal(np.abs(real_df[par] - df.loc[mask, '{}_min'.format(par)]), 1e-6)) \
+                      & (np.less_equal(real_df[par], df.loc[mask, '{}_max'.format(par)])
+                         | np.less_equal(np.abs(df.loc[mask, '{}_max'.format(par)] - real_df[par]), 1e-6))
+                print('{}:\t{:.1f}% within CIs'
+                      .format(par, 100 * sum(df.loc[mask, '{}_within_CI'.format(par)]) / n_observations))
+                f.write('{}:\t{:.1f}% within CIs\n'
+                      .format(par, 100 * sum(df.loc[mask, '{}_within_CI'.format(par)]) / n_observations))
+                df.loc[mask, '{}_CI_relative_width'.format(par)] \
+                    = 100 * (df.loc[mask, '{}_max'.format(par)] - df.loc[mask, '{}_min'.format(par)]) / real_df[par]
+                print('{}:\t{:.1f}% median CI width'
[…]

### 2023-06-09 simulations/py/likelihood_comparison.py

@@ -1,0 +1,64 @@
+import pandas as pd
+from ete3 import Tree
+
+from bdpn.bdpn import loglikelihood
+import numpy as np
+from scipy.stats import binomtest
+
+PYBDEI = 'PyBDEI'
+
+REAL_TYPE = 'real'
+
+
+if __name__ == "__main__":
+    import argparse
+
+    parser = argparse.ArgumentParser(description="Calculates likelihoods.")
+    parser.add_argument('--estimates', type=str, help="estimated parameter", required=True)
+    parser.add_argument('--tree_pattern', type=str, help="trees", required=True)
+    parser.add_argument('--likelihoods', type=str, help="likelihood table")
+    parser.add_argument('--log', type=str, help="likelihood stats")
+    params = parser.parse_args()
+
+    df = pd.read_csv(params.estimates, sep='\t', index_col=0)
+    NON_REAL_TYPES = [_ for _ in df['type'].unique() if _ != REAL_TYPE]
+    ALL_TYPES = NON_REAL_TYPES + [REAL_TYPE]
+
+    df.index = df['type'] + '_' + df.index.map(str)
+    lk_df, index = [], []
+    for i in range(len(df) // len(ALL_TYPES)):
+        tree = Tree(params.tree_pattern.format(i))
+        p = df.loc['real_{}'.format(i), 'p']
+        type2lk = {}
+        for type in ALL_TYPES:
+            la, psi, psi_n, rho, rho_n = df.loc['{}_{}'.format(type, i), ['lambda', 'psi', 'psi_p', 'p', 'pn']]
+            type2lk[type] = loglikelihood([tree], la, psi, psi_n, rho, rho_n)
+            index.append(i)
+        best_lk = max(type2lk.values())
+        for _ in ALL_TYPES:
+            lk_df.append([_, type2lk[_], best_lk - type2lk[_], type2lk[_] - type2lk['real']])
+    lk_df = pd.DataFrame(data=lk_df, columns=['type', 'loglk', 'diff with max', 'diff with real'], index=index)
+    lk_df.to_csv(params.likelihoods, sep='\t')
+
+    with open(params.log, 'w+') as f:
+
+        for i in range(len(ALL_TYPES)):
+            type1 = ALL_TYPES[i]
+            log_lk_t1 = lk_df[lk_df['type'] == type1]
+            for j in range(i + 1, len(ALL_TYPES)):
+                type2 = ALL_TYPES[j]
+                log_lk_t2 = lk_df[lk_df['type'] == type2]
[…]

### 2023-06-09 simulations/py/summary_table.py

@@ -19,14 +19,14 @@ if __name__ == "__main__":
     logging.basicConfig(level=logging.INFO, format='%(asctime)s: %(message)s', datefmt="%Y-%m-%d %H:%M:%S")
 
     df = pd.DataFrame(columns=['type', 'sampled_tips',
-                               'lambda',
-                               'psi',
-                               'psi_p',
-                               'R_naught',
-                               'infectious_time',
-                               'p',
-                               'pn',
-                               'partner_removal_time'])
+                               'lambda', 'lambda_min', 'lambda_max',
+                               'psi', 'psi_min', 'psi_max',
+                               'psi_p', 'psi_p_min', 'psi_p_max',
+                               'R_naught', 'R_naught_min', 'R_naught_max',
+                               'infectious_time', 'infectious_time_min', 'infectious_time_max',
+                               'p', 'p_min', 'p_max',
+                               'pn', 'pn_min', 'pn_max',
+                               'partner_removal_time', 'partner_removal_time_min', 'partner_removal_time_max'])
 
     for real in params.real:
         i = int(re.findall(r'[0-9]+', real)[0])
@@ -42,23 +42,24 @@ if __name__ == "__main__":
     for (est_list, fixed) in ((params.estimated_la, 'lambda'), (params.estimated_psi, 'psi'), (params.estimated_p, 'p')):
         for est in est_list:
             i = int(re.findall(r'[0-9]+', est)[0])
-            ddf = pd.read_csv(est)
+            ddf = pd.read_csv(est, index_col=0)
             est_label = 'BDPN({})'.format(fixed)
-            estimates = ddf.loc[next(iter(ddf.index)), :]
-            R0, it, p, pn, rt = ddf.loc[next(iter(ddf.index)), :]
+            R0, rt, rho, rho_p, prt, la, psi, psi_p = ddf.loc['value', :]
             df.loc['{}.{}'.format(i, est_label),
             ['R_naught', 'infectious_time', 'partner_removal_time',
              'lambda', 'psi', 'psi_p', 'p', 'pn', 'type']] \
-                = [R0, it, rt, R0 / it, 1 / it, 1 / rt, p, pn, est_label]
+                = [R0, rt, prt, la, psi, psi_p, rho, rho_p, est_label]
[…]

### 2023-06-12 article/article.tex
cherry test on simulated data

@@
 \subsection{Performance on simulated data}
-To illustrate the PN test performance, we applied it to the trees simulated under BD and under BDPn models.
+To illustrate the PN test performance, we applied it to the trees simulated under BD and under BD-PN models.
+We simulated 100 trees with 5\,000--10\,000 tips under the BD-PN model with our tree simulator.
+For each tree the parameter values were drawn randomly within the following bounds:
+$R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
+removal rate $\psi \in ]1 / 20, 1 / 5]$,
+ratio between the partner removal rate after notification and the removal rate $\frac{\psi_n}{\psi} \in ]10, 100]$,
+sampling probability $\rho \in ]0.01, 0.9]$,
+partner notification probability $\rho_n \in ]0.1, 0.9]$.
+We also simulated 100 trees of 5\,000--10\,000 tips under the BD model.
+For each tree in the BD data set the parameter values ($\lambda, \psi,$ and $\rho$) were the same as the ones selected for the corresponding tree of the BD-PN dataset.
 
-... details coming soon ...
+We then run the PN tests on each tree of the both datasets.
+For all the trees in the BD dataset the PN test value was above $0.05$ and varied between $0.33$ and $0.63$ (i.e.
+100\% true negative results).
+For 97 (out of 100) trees in the BD-PN dataset the PN test value was below $0.05$ (i.e.
+97\% true positive and 3\% false negative results).
+Our test therefore showed absolute specificity and high sensitivity.
 
@@
 
-To assess the performance of our maximum-likelihood estimator for the BD-PN model, we simulated 100 trees with 5\,000--10\,000 tips under the BD-PN model with our tree simulator.
-For each tree the parameter values were drawn randomly within the following bounds:
-$R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
-removal rate $\psi \in ]1 / 20, 1 / 5]$,
-ratio between the partner removal rate after notification and the removal rate $\frac{\psi_n}{\psi} \in ]10, 100]$,
-sampling probability $\rho \in ]0.01, 0.9]$,
-partner notification probability $\rho_n \in ]0.1, 0.9]$.
+To assess the performance of our maximum-likelihood estimator for the BD-PN model, we used the 100 trees with 5\,000--10\,000 tips simulated under the BD-PN model as described in the "PN test" section.
+
 We applied bdpn to each of the trees three times: fixing each of the BD parameters ($\lambda,\psi$ or $\rho$) to its real value (for identifiability).

### 2023-06-12 simulations/py/summary_table_cherries.py

@@ -1,0 +1,37 @@
+import re
+
+import pandas as pd
+
+if __name__ == "__main__":
+    import argparse
+
+    parser = argparse.ArgumentParser(description="Summarize cherry tests.")
+    parser.add_argument('--logs', nargs='+', type=str, help="cherry test results")
+    parser.add_argument('--tab', type=str, help="summary table")
+    params = parser.parse_args()
+
+    df = pd.DataFrame(columns=['model', 'tree', 'PN_test'])
+
+    for log in params.logs:
+        i = int(re.findall(r'[0-9]+', log)[0])
+        with open(log, 'r') as f:
+            pval = float(f.readline().strip('\n').split('\t')[1])
+        model = re.findall(r'BDPN|BD', log)[0]
+        df.loc['{}.{}'.format(i, model), :] = [model, i, pval]
+
+    for model in df['model'].unique():
+        ddf = df[df['model'] == model]
+        mean_pval = ddf['PN_test'].mean()
+        min_pval = ddf['PN_test'].min()
+        max_pval = ddf['PN_test'].max()
+        percentage_significant = 100 * len(ddf[ddf['PN_test'] < 0.05]) / len(ddf)
+        print('{}: avg pval {} [{}-{}], {}% have pval < 0.05'
+              .format(model, mean_pval, min_pval, max_pval, percentage_significant))
+        df.loc['{}.mean'.format(model), :] = [model, 'mean', mean_pval]
+        df.loc['{}.min'.format(model), :] = [model, 'min', min_pval]
+        df.loc['{}.max'.format(model), :] = [model, 'max', max_pval]
+        df.loc['{}.perc'.format(model), :] = [model, 'percent < 0.05', percentage_significant]
+
+    df.sort_values(by=['model', 'tree'], inplace=True)
+    df.to_csv(params.tab, sep='\t', index=False)
+

### 2023-07-25 article/article.tex
Simplified BD-Pn equation solutions

@@
 Coalescent models are often preferred for estimating deterministic population dynamics, however for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\cite{Macpherson2021}.
-In the classic BD model with incomplete sampling (BDS~\cite{Stadler2009}), births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g.
+In the classic BD model with incomplete sampling~\cite{Stadler2009}, births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g.
 due to healing, self-isolation, starting a treatment, or death, modelled with a constant removal rate).
@@
 
-In this study we propose an extension of the MTBD framework that allows to model non-random sampling due to partner notification, MTBD-PN.
-In the rest of the article, we describe the MTBD-PN model family and its assumptions, implement a transmission tree simulator under these models, derive the equations for their likelihood calculation, implement a maximum likelihood parameter estimator and test it on simulated data.
-We also propose a test allowing to distinguish between MTBD and MTBD-PN models.
-Finally, we apply it to the study of HIV epidemic in the UK.
+In this study we propose an extension of the MTBD framework that allows for modeling non-random sampling due to partner notification (PN), MTBD-PN.
+In the rest of the article, we describe the MTBD-PN model family and its assumptions, present a transmission tree simulator under these models, derive the equations for MTBD-PN likelihood calculation, and propose a test for detecting partner notification in pathogen phylogenetic trees.
+We also present a maximum likelihood parameter estimator for the simplest model of the MTBD-PN family, BD-PN, and test it on simulated data.
+Finally, we apply it to the study of HIV-1 B epidemic in the UK.
 
@@
     &- \psi_s (1 - \rho)\;  \textit{\color{gray} $\leftarrow$ removal without sampling}\\
-U_{s}(T) = & 1\;  \textit{\color{gray} $\leftarrow$ the probability to stay unsampled over time 0 is 1} \label{eq:u}
+U_{s}(T) = & 1\;  \textit{\color{gray} $\leftarrow$ the probability to stay unsampled over time 0 is 1} \label{eq:Us}
 \end{cases}
@@
 
-With the equations~(\ref{eq:u}), we can write down the equations for the probability densities $p_{s,si}^{(i)}(t)$ of evolving as on an observed tree branch that finishes at time $t_i$ in state $si$, starting  on this branch at a time $t \leq t_i$ in state $s$.
-The master equations for the MTBD models were initially developed by ~\citet{Stadler2013a}, however here we present their parallelizable version that we recently proposed~\cite{zhukovaFastAccurateMaximumLikelihood2022}:
+With the equations~(\ref{eq:Us}), we can write down the equations for the probability densities $p_{s,si}^{(i)}(t)$ of evolving as on an observed tree branch that finishes at time $t_i$ in state $si$, starting  on this branch at a time $t \leq t_i$ in state $s$.
+The master equations for the MTBD models were initially developed by ~\citet{Stadler2013a}, however what we present here, is their parallelizable version that we proposed in~\cite{zhukovaFastAccurateMaximumLikelihood2022}:
 
@@
 
-If the states of all the observed tree $\mathscr{T}$ nodes are known, then its likelihood density under parameter values $\Theta$ can be calculated as $L(\mathscr{T}|\Theta)$:
+If the states of all the observed nodes of the tree $\mathscr{T}$ are known, then its likelihood density under parameter values $\Theta$ can be calculated as $L(\mathscr{T}|\Theta)$:
 
@@
 In MTBD-PN, at the moment of sampling the sampled individual might notify their most recent partner with a given probability $\rho_n$.
-Upon notification, partner is removed almost instantaneously (modelled via a notified removal rate $\psi_n >> \psi_k \;\forall k \in 1:m$).
+Upon notification, partner is removed almost instantaneously (modeled via a notified removal rate $\psi_n >> \psi_k \;\forall k \in 1:m$).
 MTBD-PN hence has $2k^2 + 1$ classical MTBD and 2 additional parameters:
 \begin{itemize}
- \item $\mu_{kl} \;\forall k,l \in 1:m, k \neq l$ -- state change rate from $k$ to $l$;
- \item $\lambda_{kl} \;\forall k,l \in 1:m$ -- transmission rate: an infected individual in state $k$ transmits their pathogen (which corresponds to an internal node in the transmission tree), the newly infected individual is in state $l$;
- \item $\psi_k \;\forall k \in 1:m$ -- removal rate: an infected individual in state $k$ stops being infectious at this rate (exponential distribution), which corresponds to a tip in the transmission tree (observed or hidden);
- \item $\rho$ -- sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
- \item $\rho_n$ -- notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
- \item $\psi_n >> \psi_k  \;\forall k \in 1:m$ -- notified removal rate: a notified individual stops being infectious and gets observed with this rate (exponential distribution), which corresponds to an observed tip in the transmission tree.
+ \item $\mu_{kl} \;\forall 1 \leq k,l  \leq m, k \neq l$ -- constant state change rate from $k$ to $l$;
+ \item $\lambda_{kl} \;\forall 1 \leq k,l \leq m$ -- constant transmission rate: an infected individual in state $k$ transmits their pathogen (which corresponds to an internal node in the transmission tree), the newly infected individual is in state $l$;
+ \item $\psi_k \;\forall 1 \leq k \leq m$ -- constant removal rate: an infected individual in state $k$ stops being infectious at this rate.
+This corresponds to a tip in the transmission tree (observed or hidden);
+ \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
+ \item $\rho_n$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
+ \item $\psi_n >> \psi_k  \;\forall 1 \leq k \leq m$ -- constant notified removal rate: a notified individual (if not yet removed via standard procedure) stops being infectious and gets observed with this rate.
+This creates an observed tip in the transmission tree.
 \end{itemize}
 
-This model is a simplification of the notification process, in particular it makes 5 assumptions, which might need to be relaxed in the future work:
+This model is a simplification of the notification process, in particular it makes 5 assumptions:
 \begin{enumerate}
@@
 \item notified individuals do not notify further;
-\item after the notification notified individuals do not transmit further;
+\item after the notification, notified individuals do not transmit further;
 \item only the most recent partner can get notified.
@@
 
-These assumptions make the mathematics behind the model manageable, however they can be easily relaxed in a simulator (e.g. to be used with a deep learning estimator~\cite{Voznica2021} after).
+These assumptions make the mathematics behind the model manageable, however they can be easily relaxed in a tree simulator (e.g., to be used with a deep learning estimator~\cite{Voznica2021}, which does not require likelihood calculation).
 
-\subsection{Mathematical formulation}
+\subsubsection{Mathematical formulation}
 $p_{s,si}^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along a tree branch, provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$.
-They could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.
+These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.
 
-We will use an \textbf{oriented probability} $P^{(i,o)}_{s,si}(t)$ for the cases when the individual at the end of the branch is the same as at time $t$:
+We will use an \textbf{oriented probability} $p^{(i,o)}_{s,si}(t)$ for the cases when the individual at the end of the branch is the same as at time $t$:
 
@@
 \dot{p}_{s,si}^{(i,o)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i,o)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i,o)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, folowed by evolution from the new state }\\
+    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i,o)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, followed by evolution from the new state }\\
     & - \sum\limits_{k=1}^{m} \lambda_{sk} p_{s,si}^{(i,o)}(t)U_{k}(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
@@
 \dot{p}_{s,si}^{(i,nh)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i,nh)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i,nh)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, folowed by evolution from the new state }\\
+    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i,nh)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, followed by evolution from the new state }\\
 p_{s,si}^{(i,nh)}(t_i) = & 
@@
 
-Under the MTBD-PN model, the probability density $q_{si,sj}(i)$ of a branch connecting a node $i$ in state $si$ to its parent node $j$ in state $sj$ can be written as:
+Under the MTBD-PN model, the probability density $q_{si,sj}(i)$ of a branch connecting a node $i$ in state $si$ to its parent node $j$ in state $sj$ (without taking into account the event at the end of the branch) can be written as:
 
@@
 \begin{cases}
-q^{(u)}_{sj,si}(i) = p^{(i)}_{sj,si}(t_j) \textit{\color{gray} ~~if $i$ is an \textbf{unnotified non-notifier}}\\
+q^{(u)}_{sj,si}(i) = p^{(i)}_{sj,si}(t_j) \textit{\color{gray} ~if $i$ is an \textbf{unnotified non-notifier}}\\
 q_{sj,si}^{(n)}(i) = 
@@
 \\
-q^{(p)}_{sj,si}(i)  = p^{(i, o)}_{sj,si}(t_j)  \textit{\color{gray} ~~if $i$ is an internal node who will be notified after $t_i$}
+q^{(p)}_{sj,si}(i)  = p^{(i, o)}_{sj,si}(t_j)  \textit{\color{gray} ~if $i$ is an internal node whose individual will be notified after $t_i$}
 \\
-q^{(p)}_{sj,si}(i, r)  = p^{(r, o)}_{sj,si}(t_j) e^{- \psi_n (t_i - t_r)} \textit{\color{gray} if $i$ is a \textbf{partner} tip notified by $r$ at time $t_r < t_i$}\\
+q^{(p)}_{sj,si}(i, r)  = 
+\begin{cases}
+p^{(r, o)}_{sj,si}(t_j) e^{- \psi_n (t_i - t_r)} \textit{\color{gray} if $i$ is a \textbf{partner} tip notified by $r$ at time $t_r \leq t_i$}\\
+p^{(i, o)}_{sj,si}(t_j) \textit{\color{gray}~~ if $i$ is a \textbf{partner} tip sampled via a standard procedure} \\
+~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\textit{\color{gray} before the notification by $r$: $t_i < t_r$}\\
+\end{cases}
 \end{cases}
@@
 
-Using the formulas~(\ref{eq:branch}), we can calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}, where for each visited node $i$ we calculate two values for each possible state $sj \in 1:m$ at the beginning of the branch connecting it to its parent $j$: the likelihood density of its subtree if its branch corresponds to (1) an unnotified individual $l^{(u)}_{sj}(i)$; (2) a notifier $l^{(u)}_{sj}(i)$ (for tips only); or (3) an (eventually) notified partner $l^{(p)}_{sj}(i, r)$ (notified by $r$).
+Using the formulas~(\ref{eq:branch}), we can calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}, where for each visited node $i$, and for each possible state $1 \leq sj \leq m$ at the beginning of the branch connecting it to its parent $j$  we calculate the likelihood density of its subtree if its branch corresponds to (1) an unnotified individual $l^{(u)}_{sj}(i)$; (2) a notifier $l^{(n)}_{sj}(i)$ (for tips only); or (3) an (eventually) notified partner $l^{(p)}_{sj}(i, r)$ (notified by the individual corresponding to the tip $r$ at time $t_r$).
 
-For a tip $i$, these values can be calculated by considering all its possible states $si \in 1:m$, as:
+For a tip $i$ in state $si$ these values can be calculated as:
 \begin{equation}
@@
 \begin{array}{ll}
-l^{(u)}_{sj}(i) &= \sum\limits_{si=1}^{m} q^{(u)}_{sj,si}(i)\psi_{si}\rho (1-\rho_{n})\\
-l^{(n)}_{sj}(i) &= \sum\limits_{si=1}^{m} q^{(n)}_{sj,si}(i)\psi_{si}\rho \rho_{n}\\
-l^{(p)}_{sj}(i, r) &= \sum\limits_{si=1}^{m} q^{(p)}_{sj,si}(i,r)\psi_{n}\\
+l^{(u)}_{sj}(i) &= q^{(u)}_{sj}(i)\psi_{si}\rho (1-\rho_{n})\\
+l^{(n)}_{sj}(i) &= q^{(n)}_{sj}(i)\psi_{si}\rho \rho_{n}\\
+l^{(p)}_{sj}(i, r) &= q^{(p)}_{sj}(i,r)\psi_{n}\\
 \end{array}
@@
 \begin{split}
-l^{(u)}_{sj}(i) = &\sum\limits_{si=1}^{m}q^{(u)}_{sj,si}(i) \sum\limits_{k=1}^{m} \lambda_{si,k} \\
+l^{(u)}_{sj}(i) = &q^{(u)}_{sj}(i) \sum\limits_{k=1}^{m} \lambda_{si,k} \\
 \cdot \Big(~&
@@
 
-Note that a notification will only be possible if (1) the notifying branch is external (as the notifier only notifies the most recent partner); (2) there is at least one tip in the notified subtree that can be a partner, i.e.
+Note that a notification will only be possible if the notifying branch is external as the notifier only notifies the most recent partner, i.e.
+the transmission corresponds to the notifier's parent node.
+%; (2) there is at least one tip in the notified subtree that can be a partner, i.e.
 was sampled after the notifier and whose parent's time was before notification (i.e.
@@
 \begin{split}
-l^{(p)}_{sj}(i, r) = \sum\limits_{si=1}^{m}q^{(p)}_{sj,si}(i,r) \sum\limits_{k=1}^{m}  \lambda_{si,k}\Big(&l^{(p)}_{si}(i0,r)l^{(u)}_{k}(i1) \textit{\color{gray} $\leftarrow$ notification in $i0$'s subtree} \\
+l^{(p)}_{sj}(i, r) = q^{(p)}_{sj}(i,r) \sum\limits_{k=1}^{m}  \lambda_{si,k}\Big(&l^{(p)}_{si}(i0,r)l^{(u)}_{k}(i1) \textit{\color{gray} $\leftarrow$ notification in $i0$'s subtree} \\
  &+ l^{(u)}_{k}(i0)l^{(p)}_{si}(i1,r)\Big)  \textit{\color{gray} $\leftarrow$ notification in $i1$'s subtree} 
@@
 once we have climbed to an internal node corresponding to the transmission between the notifier (corresponding to one of its children, must be a tip) and the partner (the other child).
-As only the contributions coming from the partner tip branches depend on the notification times, while climbing the tree for each internal node $i$ and each of the possible states $sj \in 1:m$ at the beginning of its branch, we can store an array of pre-calculated values for notified internal paths for each of its tips (e.g.
-$a$) and each of the states $v \in 1:m$ at the beginning of $a$'s branch, describing the tip's sampling time $t_a$ and a configuration $C_{sj}(i,a,v)$:
+In the partner subtree only the contribution coming from the partner tip branch depends on the notification time.
+Hence while climbing the tree from tips to the root, for each internal node $i$ and each of the possible states $sj \in 1:m$ at the beginning of its branch, we can store an array of pre-calculated values for notified internal paths leading to each of its tips.
+The values of this array will store for each tip (e.g.
+$a$) its sampling time $t_a$ and a configuration $C_{sj}(i,a)$, representing the subtree likelihood provided $a$ corresponds to the same notified partner individual as $i$ (but without the $a$'s tip branch contribution):
 
@@
 \begin{split}
-C_{sj}(i,a,v) = &\begin{cases}
-q^{(p)}_{sj,v}(i)
-\sum\limits_{k=1}^{m}\lambda{v,k} \cdot l_{k}^{(u)}\Big(child(i, \not a)\Big) \textit{\color{gray}~~~if $i$ is a direct parent of $a$}
+C_{sj}(i,a) = &\begin{cases}
+1 \textit{\color{gray}~~~if $i = a$}
 \\
-\sum\limits_{si=1}^{m}q^{(p)}_{sj,si}(i)
-\sum\limits_{k=1}^{m}\lambda{si,k} \cdot l_{k}^{(u)}\Big(child(i, \not a)\Big)C_{si}\Big(child(i, a),a,v\Big) \textit{\color{gray}~~otherwise}\\ 
+q^{(p)}_{sj}(i)
+\sum\limits_{k=1}^{m}\lambda{si,k} \cdot l_{k}^{(u)}\Big(child(i, \not a)\Big)C_{si}\Big(child(i, a),a\Big) \textit{\color{gray}~~otherwise}\\ 
 \end{cases},
@@
 \scriptsize 
-l^{(p)}_{sj}(i, r) = \sum\limits_{a \in tips(i)}\sum\limits_{k=1}^{m} C_{sj}(i, a, k) l^{(p)}_{k}(a, r)\label{eq:lp}
+l^{(p)}_{sj}(i, r) = \sum\limits_{a \in tips(i)}C_{sj}(i, a) l^{(p)}_{sa}(a, r)\label{eq:lp}
 \end{equation}
@@
 They can transmit with a constant rate $\lambda$, get removed with a constant rate $\psi$, be sampled upon removal with a probability $\rho$.
-Below we describe the special case of equations~(\ref{eq:u}-\ref{eq:lp}) for its PN extension BD-PN.
+Below we describe the special case of equations~(\ref{eq:Us}-\ref{eq:lp}) for its PN extension BD-PN.
+
+%\begin{equation}
+%\scriptsize
+%\begin{cases}
+%\dot{U}(t) &= \Big(\lambda + \psi\Big) U(t) - \lambda U^2(t) - \psi (1 - \rho)\\
+%U(T) &=  1\\
+%\dot{p}^{(i)}(t) &=  \Big(\lambda + \psi\Big) p^{(i)}(t) - 2 \lambda p^{(i)}(t)U(t)\\
+%p^{(i)}(t_i) &= 1
+%\end{cases}
+%\end{equation}
+%
+%\begin{equation}
+%L(\mathscr{T}|\{\lambda, \psi, \rho\}) = n log(\psi\rho) + \sum\limits_{i \in \substack{internal\\nodes}} 2\lambda p^{(i0)}(t_i) p^{(i1)}(t_i)
+%\end{equation}
+%
 
@@
 \begin{cases}
-\dot{U}(t) &= \Big(\lambda + \psi\Big) U(t) - \lambda U^2(t) - \psi (1 - \rho)\\\\
+\dot{U}(t) &= \Big(\lambda + \psi\Big) U(t) - \lambda U^2(t) - \psi (1 - \rho)\\
 U(T) &=  1\\
 \dot{p}^{(i)}(t) &=  \Big(\lambda + \psi\Big) p^{(i)}(t) - 2 \lambda p^{(i)}(t)U(t)\\
-\dot{p}^{(i,o)}(t) &=  \Big(\lambda + \psi\Big) p^{(i)}(t) - \lambda p^{(i)}(t)U(t)\\
-\dot{p}^{(i,nh)}(t) &=  \Big(\lambda + \psi\Big) p^{(i)}(t)\\
+\dot{p}^{(i,o)}(t) &=  \Big(\lambda + \psi\Big) p^{(i,o)}(t) - \lambda p^{(i,o)}(t)U(t)\\
+\dot{p}^{(i,nh)}(t) &=  \Big(\lambda + \psi\Big) p^{(i,nh)}(t)\\
 p^{(i)}(t_i) &= p^{(i,o)}(t_i) = p^{(i,nh)}(t_i) = 1
@@
 \end{equation}
+
+%
+%These equations have the following closed form solutions:
+%\begin{equation}
+%\scriptsize
+%\begin{split}
+%&\begin{cases}
+%U(t) = \frac{\lambda + \psi + c_1 \frac{c_2 e^{t c_1} - 1}{c_2 e^{tc_1} + 1}}{2\lambda}\\
+%\\
+%p^{(i)}(t) = \frac{c_3^2 e^{c_1 (t - t_i)}}{(c_2e^{c_1 t} + 1)^2} \\
+%\\
+%p^{(i,o)}(t) =  \frac{c_3 e^{1/2 (\lambda + \psi + c_1)(t - t_i)}}{c_2e^{c_1 t} + 1} \\
+%\\
+%p^{(i,nh)}(t) =  e^{(\lambda + \psi)(t - t_i)}
+%\end{cases},\\
+%& \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi}e^{-Tc_1},\; c_3 = c_2 e^{c_1 t_i} + 1
+%\end{split}
+%\end{equation}
+%
+%
+%These equations have the following closed form solutions:
+%\begin{equation}
+%\scriptsize
+%\begin{split}
+%&\begin{cases}
+%U(t) = \frac{\lambda + \psi + c_1 \frac{c_2 e^{ c_1 (t-T)} - 1}{c_2 e^{c_1 (t-T)} + 1}}{2\lambda}\\
+%\\
+%p^{(i)}(t) = \Big(\frac{c_2 e^{c_1 (t_i - T)} + 1}{c_2e^{c_1 (t - T)} + 1}\Big)^2e^{c_1 (t - t_i)} \\
+%\\
+%p^{(i,o)}(t) =  \frac{(c_2 e^{c_1 (t_i -T)} + 1) \Big(e^{c_1 (t -t_i)} e^{(\lambda + \psi)(t - t_i)}\Big)^{1/2}}{c_2e^{c_1 (t - T)} + 1} \\
+%\\
+%p^{(i,nh)}(t) =  e^{(\lambda + \psi)(t - t_i)}
+%\end{cases},\\
+%& \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi},
+%\end{split}
+%\end{equation}
 
@@
 &\begin{cases}
-U(t) = \frac{\lambda + \psi + c_1 \frac{c_2 e^{t c_1} - 1}{c_2 e^{tc_1} + 1}}{2\lambda}\\
+U(t) = \frac{\lambda + \psi}{2\lambda} +  \frac{c_1}{2\lambda}\Big(\frac{E_1 - 1}{E_1 + 1}\Big)\\
 \\
-p^{(i)}(t) = \frac{c_3^2 e^{c_1 (t - t_i)}}{(c_2e^{c_1 t} + 1)^2} \\
+p^{(i)}(t) = \Big(\frac{E_2 + 1}{E_1 + 1}\Big)^2E_3 \\
 \\
-p^{(i,o)}(t) =  \frac{c_3 e^{1/2 (\lambda + \psi + c_1)(t - t_i)}}{c_2e^{c_1 t} + 1} \\
+p^{(i,o)}(t) =  \frac{E_2 + 1}{E_1 + 1}\sqrt{E_3E_4} \\
 \\
-p^{(i,nh)}(t) =  e^{(\lambda + \psi)(t - t_i)}
+p^{(i,nh)}(t) =  E_4
 \end{cases},\\
-& \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi}e^{-Tc_1},\; c_3 = c_2 e^{c_1 t_i} + 1
+& \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi},\\
+& ~~~~~~~~E_1 = c_2 e^{c_1 (t -T)},\; E_2 = c_2 e^{c_1 (t_i -T)},\; E_3 = e^{c_1 (t -t_i)}, \; E_4 =  e^{(\lambda + \psi) (t -t_i)}\\
 \end{split}
@@
 The likelihood of a tree $\mathscr{T}$ under BD-PN model with parameter values $\Theta$ can be calculated as $\mathscr{L}(\mathscr{T}|\Theta) = l^{(u)}(root)$, where $l^{(u)}(i)$ is a likelihood of node $i$'s subtree if the branch connecting $i$ to its parent node (of length zero for the root) correspond to an unnotified individual.
-It can be calculated, along with $l^{(p)}(i, r)$ (likelihood when $i$'s branch corresponds to an eventually notified partner, notified by $r$) for each tree node $i$ with a pruning algorithm.
+It can be calculated, along with $l^{(n)}(i)$ (likelihood when $i$ is a notifier  tip) and $l^{(p)}(i, r)$ (likelihood when $i$'s branch corresponds to an eventually notified partner, notified by $r$) for each tree node $i$ with a pruning algorithm.
 
@@
 l^{(n)}(i) &= p^{(i,nh)}(t_j)\psi\rho \rho_{n}\\
-l^{(p)}(i, r) &= p^{(r,o)}(t_j)e^{-\psi_n (t_i - t_r)}\psi_{n}\\
+l^{(p)}(i, r) &=\begin{cases}
+p^{(r, o)}(t_j) e^{- \psi_n (t_i - t_r)} \textit{\color{gray} if $i$ is a \textbf{partner} tip notified by $r$ at time $t_r \leq t_i$}\\
+p^{(i, o)}(t_j) \textit{\color{gray}~~ if $i$ is a \textbf{partner} tip sampled via a standard procedure} \\
+~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\textit{\color{gray} before the notification by $r$: $t_i < t_r$}\\
+\end{cases}
 \end{array}
@@
 once we have climbed to an internal node corresponding to the transmission between the notifier (corresponding to one of its children, must be a tip) and the partner (the other child).
-As only the contributions coming from the partner tip branches depend on the notification times, while climbing the tree for each internal node $i$ and the time of its branch start $t_j$, we can store an array of pre-calculated values for notified internal paths for each of its tips (e.g.
-$a$), describing the tip's sampling time $t_a$ and a configuration $C(i,a)$:
+As in the general MTBD-PN case, while climbing the tree from tips to the root, for each internal node $i$ we can pre-calculate and store an array of values for notified internal paths leading to each of its tips.
+The values of this array will store for each tip (e.g.
+$a$) its sampling time $t_a$ and a configuration $C(i,a)$, representing the subtree likelihood provided $a$ corresponds to the same notified partner individual as $i$ (but without the $a$'s tip branch contribution):
 
@@
 
-\subsection{BDEI-PN}
-Another particular example of the MTBD model generating trees with known node states is the basic birth-death exposed-infectious model (BDEI)~\citep{Stadler2014} where   infectious ($I$) individuals can transmit (at a rate $\lambda$) or be removed (at a rate $\psi$), and hence create an internal or external node in the transmission tree.
+%\subsection{BDEI-PN}
+%Another particular example of the MTBD model generating trees with known node states is the basic birth-death exposed-infectious model (BDEI)~\citep{Stadler2014} where   infectious ($I$) individuals can transmit (at a rate $\lambda$) or be removed (at a rate $\psi$), and hence create an internal or external node in the transmission tree.
 Upon removal they can be sampled with a probability $\rho$.
@@
 Below we describe the special case of equations~(\ref{eq:u}-\ref{eq:lp}) for its PN extension BDEI-PN, assuming only infectious individuals can be removed and sampled even with notification (e.g., because the viral load of the exposed individuals is yet undetectable).
-
-The master equations for the BDEI-PN model are a special case of the MTBD-PN equations:
-
-\begin{equation}
-\scriptsize
-\begin{cases}
-\dot{U_E}(t) = &\mu U_E(t)- \mu U_I(t)\\    
-\dot{U_I}(t) = &\Big(\lambda + \psi\Big) U_I(t)- \lambda U_I(t)U_E(t)- \psi (1 - \rho)\\
-U_E(T) = & U_I(T) = 1\\
-\dot{p}_E^{(i)}(t) = & \mu p_E^{(i)}(t) - \mu p_I^{(i)}(t)\\
-p_E^{(i)}(t_i) = & 0\\
-p_E^{(i,o)}(t_i) = & p_E^{(i,nh)}(t_i) = p_E^{(i)}(t_i)\\
-\dot{p}_I^{(i)}(t) = & \Big(\lambda + \psi\Big) p^{(i)}(t) - \lambda \Big(p_I^{(i)}(t)U_E(t) + p_E^{(i)}(t)U_I(t)\Big)\\
-\dot{p}_I^{(i,o)}(t) = & \Big(\lambda + \psi\Big) p^{(i,o)}(t) - \lambda p_I^{(i,o)}(t)U_E(t)\\
-\dot{p}_I^{(i,nh)}(t) = & \Big(\lambda + \psi\Big) p^{(i,nh)}(t)\\
-p_I^{(i)}(t_i) = & p_I^{(i,o)}(t_i) = p^{(i,nh)}(t_i) = 1\label{eq:up-bd}
-\end{cases}
-\end{equation}
-
-The likelihood of a tree $\mathscr{T}$ under BDEI-PN model with parameter values $\Theta$ can be calculated as $\mathscr{L}(\mathscr{T}|\Theta) = l_I^{(u)}(root)$, where $l_s^{(u)}(i)$ is a likelihood of node $i$'s subtree if the branch connecting $i$ to its parent node (of length zero for the root) starts in state $s \in \{E, I\}$ and correspond to an unnotified individual.
+%
+%
+%The master equations for the BDEI-PN model are a special case of the MTBD-PN equations:
+%
+%\begin{equation}
+%\scriptsize
+%\begin{cases}
+%\dot{U_E}(t) = &\mu U_E(t)- \mu U_I(t)\\    
+%\dot{U_I}(t) = &\Big(\lambda + \psi\Big) U_I(t)- \lambda U_I(t)U_E(t)- \psi (1 - \rho)\\
+%U_E(T) = & U_I(T) = 1\\
+%\dot{p}_E^{(i)}(t) = & \mu p_E^{(i)}(t) - \mu p_I^{(i)}(t)\\
+%p_E^{(i)}(t_i) = & 0\\
+%p_E^{(i,o)}(t_i) = & p_E^{(i,nh)}(t_i) = p_E^{(i)}(t_i)\\
+%\dot{p}_I^{(i)}(t) = & \Big(\lambda + \psi\Big) p^{(i)}(t) - \lambda \Big(p_I^{(i)}(t)U_E(t) + p_E^{(i)}(t)U_I(t)\Big)\\
+%\dot{p}_I^{(i,o)}(t) = & \Big(\lambda + \psi\Big) p^{(i,o)}(t) - \lambda p_I^{(i,o)}(t)U_E(t)\\
+%\dot{p}_I^{(i,nh)}(t) = & \Big(\lambda + \psi\Big) p^{(i,nh)}(t)\\
+%p_I^{(i)}(t_i) = & p_I^{(i,o)}(t_i) = p^{(i,nh)}(t_i) = 1\label{eq:up-bd}
+%\end{cases}
+%\end{equation}
+%
+%The likelihood of a tree $\mathscr{T}$ under BDEI-PN model with parameter values $\Theta$ can be calculated as $\mathscr{L}(\mathscr{T}|\Theta) = l_I^{(u)}(root)$, where $l_s^{(u)}(i)$ is a likelihood of node $i$'s subtree if the branch connecting $i$ to its parent node (of length zero for the root) starts in state $s \in \{E, I\}$ and correspond to an unnotified individual.
 It can be calculated, along with $l_s^{(p)}(i, notifier)$ (likelihood when $i$'s branch corresponds to an eventually notified partner) for each tree node $i$ with a pruning algorithm.
-
-For a tip $i$ whose branch starts at time $t_j$ in state $s \in \{E, I\}$ and finishes at time $t_i$ in state $I$, these values can be calculated as:
-\begin{equation}
-\scriptsize
-\begin{array}{ll}
-l^{(u)}_{s}(i) &= p_s^{(i)}(t_j)\psi\rho (1-\rho_{n})\\
-l^{(n)}_{s}(i) &= p_s^{(i,nh)}(t_j)\psi\rho \rho_{n}\\
-l^{(p)}_{s}(i, r) &= p_s^{(r,o)}(t_j)e^{-\psi_n (t_i - t_r)}\psi_{n}\\
-\end{array}
-\end{equation}
-
-For an internal node $i$ whose branch starts at time $t_j$ is state $s$ and finishes at time $t_i$ in state $I$, with two child nodes, $i0$ and $i1$, these values can be calculated as:
-
-\begin{equation}
-\scriptsize
-\begin{array}{ll}
-l^{(u)}(i) =& p_s^{(i)}(t_j) \lambda \\
-& \cdot \Big(\big(l_I^{(u)}(i0)l_E^{(u)}(i1) + l_E^{(u)}(i0)l_I^{(u)}(i1)\big) \\
-&~~+ is\_tip(i0) \big(l_I^{(n)}(i0)l_E^{(p)}(i1,i0) + l_E^{(n)}(i0)l_I^{(p)}(i1,i0)\big)\\
-&~~+ is\_tip(i1) \big(l_I^{(p)}(i0,i1)l_E^{(n)}(i1) + l_E^{(p)}(i0,i1)l_I^{(n)}(i1)\big)\Big),\\
-& \textit{where } is\_tip(x) = 1 \textit{ if x is a tip and $= 0$ if x is an internal node.}\\
-l^{(p)}(i, r) &= p_s^{(i,o)}(t_j) \lambda\Big(l_I^{(p)}(i0,r)l_E^{(u)}(i1) + l_E^{(u)}(i0)l_I^{(p)}(i1,r)\Big),\\
-& \textit{where r is the notifier.}\\
-\end{array}
-\end{equation}
-
-As in the general MTBD-PN case, $l_s^{(p)}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time is known, i.e.
+%
+%
+%For a tip $i$ whose branch starts at time $t_j$ in state $s \in \{E, I\}$ and finishes at time $t_i$ in state $I$, these values can be calculated as:
+%\begin{equation}
+%\scriptsize
+%\begin{array}{ll}
+%l^{(u)}_{s}(i) &= p_s^{(i)}(t_j)\psi\rho (1-\rho_{n})\\
+%l^{(n)}_{s}(i) &= p_s^{(i,nh)}(t_j)\psi\rho \rho_{n}\\
+%l^{(p)}_{s}(i, r) &= p_s^{(r,o)}(t_j)e^{-\psi_n (t_i - t_r)}\psi_{n}\\
+%\end{array}
+%\end{equation}
+%
+%For an internal node $i$ whose branch starts at time $t_j$ is state $s$ and finishes at time $t_i$ in state $I$, with two child nodes, $i0$ and $i1$, these values can be calculated as:
+%
+%\begin{equation}
+%\scriptsize
+%\begin{array}{ll}
+%l^{(u)}(i) =& p_s^{(i)}(t_j) \lambda \\
+%& \cdot \Big(\big(l_I^{(u)}(i0)l_E^{(u)}(i1) + l_E^{(u)}(i0)l_I^{(u)}(i1)\big) \\
+%&~~+ is\_tip(i0) \big(l_I^{(n)}(i0)l_E^{(p)}(i1,i0) + l_E^{(n)}(i0)l_I^{(p)}(i1,i0)\big)\\
+%&~~+ is\_tip(i1) \big(l_I^{(p)}(i0,i1)l_E^{(n)}(i1) + l_E^{(p)}(i0,i1)l_I^{(n)}(i1)\big)\Big),\\
+%& \textit{where } is\_tip(x) = 1 \textit{ if x is a tip and $= 0$ if x is an internal node.}\\
+%l^{(p)}(i, r) &= p_s^{(i,o)}(t_j) \lambda\Big(l_I^{(p)}(i0,r)l_E^{(u)}(i1) + l_E^{(u)}(i0)l_I^{(p)}(i1,r)\Big),\\
+%& \textit{where r is the notifier.}\\
+%\end{array}
+%\end{equation}
+%
+%As in the general MTBD-PN case, $l_s^{(p)}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time is known, i.e.
 once we have climbed to an internal node corresponding to the transmission between the notifier (corresponding to one of its children, must be a tip) and the partner (the other child).
@@
 Note, that the partner branch for any transmission after the transmission between the notifier and the partner, is always the donor one (hence, starting in state $I$).
-
-\begin{equation}
-\scriptsize
-\begin{split}
-C_s(i,a) = &\begin{cases}
-1 \textit{\color{gray}~~~if $i==a$}
-\\
-p_s^{(i,o)}(t_j) \lambda \cdot l_E^{(u)}\Big(child(i, \not a)\Big)C_I\Big(child(i, a),a\Big) \textit{\color{gray}~~otherwise}\\ 
-\end{cases},
-\\ &\textit{where child(i,a) is the child of i whose subtree contains a},
-\\
- &\textit{~~~~~~~~child(i,$\not a$) is the child of i whose subtree does not contain a.}
-\end{split}
-\end{equation}
-
-Once a potential notifier tip $r$ is identified (i.e., $r$ is a sister tip of the currently visited node $i$), we can calculate $l_s^{(p)}(i, r)$ as:
-\begin{equation}
-\scriptsize 
-l_s^{(p)}(i, r) = 
-\begin{cases}
-\sum\limits_{a \in tips(i)}C_s(i, a) l_I^{(p)}(a, r)\textit{\color{gray}~~if $i$ is an internal node or $s=I$}\\
-l_E^{(p)}(a, r)\textit{\color{gray}~~if $i=a$ and $s=E$}
-\end{cases}
-\end{equation}
+%
+%\begin{equation}
+%\scriptsize
+%\begin{split}
+%C_s(i,a) = &\begin{cases}
+%1 \textit{\color{gray}~~~if $i==a$}
+%\\
+%p_s^{(i,o)}(t_j) \lambda \cdot l_E^{(u)}\Big(child(i, \not a)\Big)C_I\Big(child(i, a),a\Big) \textit{\color{gray}~~otherwise}\\ 
+%\end{cases},
+%\\ &\textit{where child(i,a) is the child of i whose subtree contains a},
+%\\
+% &\textit{~~~~~~~~child(i,$\not a$) is the child of i whose subtree does not contain a.}
+%\end{split}
+%\end{equation}
+%
+%Once a potential notifier tip $r$ is identified (i.e., $r$ is a sister tip of the currently visited node $i$), we can calculate $l_s^{(p)}(i, r)$ as:
+%\begin{equation}
+%\scriptsize 
+%l_s^{(p)}(i, r) = 
+%\begin{cases}
+%\sum\limits_{a \in tips(i)}C_s(i, a) l_I^{(p)}(a, r)\textit{\color{gray}~~if $i$ is an internal node or $s=I$}\\
+%l_E^{(p)}(a, r)\textit{\color{gray}~~if $i=a$ and $s=E$}
+%\end{cases}
+%\end{equation}
 
@@
 Forests are useful for cases when a (sub-)epidemic started with several infected individuals (e.g.
-dur to multiple pathogen introductions to a country of interest or due to a change of health policies leading to a change in parameter values).
+due to multiple pathogen introductions to a country of interest or due to a change of health policies leading to a change in parameter values).
 
 In this case the (sub-)epidemic leads to a forest $\mathscr{F}$ of $f$ observed trees: $\mathscr{T}_1, \ldots, \mathscr{T}_f$.
-If all the trees started at the same time (e.g.
-due to health policy change), the forest might also include a certain number $u$ of unobserved trees, i.e., individuals who were infected at the beginning of the (sub-)epidemic, but whose trees stayed unobserved  as none of their tips got sampled.
+The forest $\mathscr{F}$ might also include a certain number $u$ of unobserved trees, for which none of their tips got sampled.
 Forest likelihood formula hence combines the likelihoods of $f$ observed and $u$ hidden trees, and can be represented in logarithmic form~(\ref{eq:forest_likelihood}).
@@
 \begin{equation}
-logL(\mathscr{F}|\Theta,u)=u\,logU_{hidden}(\Theta) + \sum\limits_{j=1}^f logL(\mathscr{T}_j|\Theta), \textit{ where } U_{hidden}(\Theta)=\sum\limits_{s=1}^{d}\pi_s U_s(0) \label{eq:forest_likelihood}
+\begin{split}
+logL(\mathscr{F}|\Theta,u)&=u\,logU_{hidden}(\Theta) + \sum\limits_{j=1}^f logL(\mathscr{T}_j|\Theta), \\
+\textit{ where }& U_{hidden}(\Theta)=\sum\limits_{s=1}^{d}\pi_s U_s(t_{start}),\\
+\textit{ and }& t_{start} \textit{ is the (potentially averaged) start time of the hidden trees.}
+\end{split} \label{eq:forest_likelihood} 
 \end{equation}
 
-If all the sub-epidemics started at the same time, for given model parameter values $\Theta$ we can estimate the number of hidden trees $u$ from the number of observed trees $f$ as:
+For given model parameter values $\Theta$ we can estimate the number of hidden trees $u$ from the number of observed trees $f$ as:
 \begin{equation}
@@
 non-removed) in different states; an array $S = [s_1, \ldots, s_m]$ of counts of sampled individuals in different states; as well as a mapping $M$ between states and ids of currently infected individuals in those states, and a mapping $N$ between states and ids of sampled individuals in those states.
-In the beginning ($t=0$) the only infected individual corresponds to the root: $c_k = 0, \;M_k = \emptyset \; \forall k \in 1:m, \; k \neq state(root)$, $c_{state(root)} = 1, \; M_{state(root)} = \{root\}$, and no individual is sampled: $s_k=0,\;N_k = \emptyset \; \forall k \in 1:m$.
+In the beginning ($t=0$) the only infected individual corresponds to the root: $c_k = 0 \;\forall k \neq state(root), \;c_{state(root)} = 1; \;M_k = \emptyset \; \forall k \neq state(root)$, $ \; M_{state(root)} = \{root\}$, and no individual is sampled: $s_k=0,\;N_k = \emptyset \; \forall 1 \leq k \leq m$.
 
@@
 
-For transmissions and sampling events we also record their time and ids of the involved individuals.
-These values are used to reconstruct the sampled parts of the tree once the tree simulation is finished.
+For transmission and sampling events we also record their time and ids of the involved individuals. These values are used to reconstruct the sampled parts of the tree once the simulation is finished.
 
-For the PN version of this simulator, we add an additional $m+1^{st}$ state for notified partners.
-$m + 1$'s state-change and transmission rates are set to zero: $\mu_{k,m+1} = \mu_{m+1,k} = \lambda_{k,m+1} = \lambda_{m+1,k} = 0\; \forall k \in 1:m$, while its removal rate is set to the notification removal rate: $\psi_{m+1} = \psi_n >> \psi_k\; \forall k \in 1:m$.
+For the PN version of this simulator, we add an additional $m+1^{th}$ state for notified partners.
+$m + 1$'s state-change and transmission rates are set to zero: $\mu_{k,m+1} = \mu_{m+1,k} = \lambda_{k,m+1} = \lambda_{m+1,k} = 0\; \forall 1 \leq k \leq m$, while its removal rate is set to the notification removal rate: $\psi_{m+1} = \psi_n >> \psi_k\; \forall 1 \leq k \leq m$.
 We keep a mapping of most recent partners $P$ for each infectious individual, and update it for both the donor and the recipient at each transmission event.
@@
 
-The test sorts the cherries by the times of their roots, and splits them into blocks of 100 cherries.
-For each cherry in a block, the test calculates the difference between its tip times.
-It then generates a collection of random cherry tip differences for the same block: For each original cherry root it picks 5 cherries with the roots that are the closest in time, randomly selects two tips among their tips, and calculates their time difference.
-The random cherry tip difference generation is repeated 100 times.
-Finally, the test reports the proportion of cases where the first quantile of the random cherry differences was smaller than that of the real cherries.
+The test sorts the cherries by the times of their roots, and splits them into blocks of 100 cherries (adjustable parameter).
+For each cherry in a block, the test calculates the difference between its tip times, hence obtaining an array of 100 cherry tip differences.
+It then generates a collection of random cherry tip differences for the same block: For each original cherry root it picks 5 more cherries with the roots that are the closest in time, randomly selects two tips among their tips, and calculates their time difference.
+An array of 100 reshuffled cherry tip differences is thus obtained fro the same block.
+The reshuffled cherry tip difference array generation is repeated 100 times (adjustable parameter).
+Finally, the test reports the proportion of reshuffled arrays whose first quantile is smaller than that of the real cherry array.
 
@@
 removal rate $\psi \in ]1 / 20, 1 / 5]$,
-ratio between the partner removal rate after notification and the removal rate $\frac{\psi_n}{\psi} \in ]10, 100]$,
-sampling probability $\rho \in ]0.01, 0.9]$,
-partner notification probability $\rho_n \in ]0.1, 0.9]$.
+ratio between the partner removal rate after notification and the removal rate $\frac{\psi_n}{\psi} \in ]50, 500]$,
+sampling probability $\rho \in ]0.1, 0.9]$,
+partner notification probability $\rho_n \in ]0.01/\rho, 0.9]$.
 We also simulated 100 trees of 5\,000--10\,000 tips under the BD model.
@@
 
-To assess the performance of our maximum-likelihood estimator for the BD-PN model, we used the 100 trees with 5\,000--10\,000 tips simulated under the BD-PN model as described in the "PN test" section.
+To assess the performance of our maximum-likelihood estimator for the BD-PN model, we used the 100 trees with 5\,000--10\,000 tips simulated under the BD-PN model as described in the ``PN test'' section.
 
@@
 We used the dated phylogenetic tree from our recent study of HIV drug resistance in the UK~\citep{zhukovaModelingDrugResistance2023}.
-The tree represents the UK HIV-B epidemic between 1960s and 2017, its tips represent viruses sampled from 39\,159 individuals between 1996 and 2016.
-The samples come from the UK HIV Drug resistance database~\citep{Dunn2007}, which stores sequences from about 50\% of infected individuals in the UK~\citep{zhukovaModelingDrugResistance2023}.
-We therefore estimated the sampling probability $\rho$ as 0.5.
+The tree represents the UK HIV-B epidemic between 1960s and 2016, its tips represent viruses sampled from 39\,130 individuals between 1996 and 2016.
+The samples used for the tree reconstruction were obtained from the UK HIV Drug resistance database~\citep{Dunn2007}~\citep{zhukovaModelingDrugResistance2023}.
 
-We cut this tree at year 2005, and estimated the BD-PN parameters on the 2005-on forests, in order to have a more homogeneous setting in terms of access to antiretroviral treatment, awareness of the HIV infection, etc.
-The 2005-on forest contained 18\,483 trees with a total of 31\,648 tips.
+Between 2012 and 2015, in the UK antiretroviral treatment (ART) of HIV-infected individuals  was initiated when their CD4 count dropped below 350 cells/mL~\citep{williamsBritishHIVAssociation2012}.
+However ``if a patient with a CD4 cell count > 350 cells/mL wishes to start ART to reduce the risk of transmission to partners, this decision is respected and ART is started''~\citep{williamsBritishHIVAssociation2012}.
+Before 2012, treatment started with an even lower CD4 (i.e.
+later).
+The British HIV Association 2015 guidelines recommend all individuals with suspected or diagnosed primary HIV infection ``are offered immediate ART''~\citep{churchillBritishHIVAssociation2016}.
+To have a more homogeneous setting in terms of access to ART, awareness of the HIV infection, etc., we cut the dated tree between 2012 and 2015, and estimated the BD-PN parameters on the 2012-2015 forests.
+
+According to the \href{https://webarchive.nationalarchives.gov.uk/ukgwa/20181112132123mp_/https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/602942/HIV_in_the_UK_report.pdf}{``Towards elimination of HIV transmission, AIDS and HIV-related deaths in the UK''}
+ report, the estimated total number of people living with HIV in the UK in 2015 was 101\,200 (97\,500 to 105\,700).
+% in 2012 this number was 98\,400 (93\,500-104\,300) [\href{HIV in the United Kingdom: 2013 Report}{https://webarchive.nationalarchives.gov.uk/ukgwa/20181112133715mp_/https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/326601/HIV_annual_report_2013.pdf}].
+Moreover these document report about 400 HIV-related deaths per year.
+According to our estimate~\citep{zhukovaModelingDrugResistance2023}, 66.5\% of the HIV-infected individuals in the UK were infected with HIV-B.
+We hence estimated that in 2015 67\,298 (64\,837, 70\,291) people were living with HIV-B.
+Our tree contains 39\,047 tips sampled by the end of 2015.
+Hence, it accounts for about 58\% (56\%, 60\%) of the total number of individuals living with HIV-B in the UK.
+However this estimate does not take into account deaths of the HIV-infected individuals over the years, which would increase the total number of HIV-infected individuals by 2015, and decrease the proportion of individuals represented by our tree.
+We therefore tested several sampling probabilities $\rho=$0.6; 0.5; 0.4.
+
  We estimated the following parameter values: $\lambda=0.262$, $\psi=0.186$, $\psi_n=186.921$, $\rho_n=0.033$, which corresponds to $R_e = 1.4$, infectious time $\frac{1}{\psi} = 5.4$ years, and about 2 days between partner notification and their sampling.

### 2023-07-25 hiv_b_uk/py/cut_tree.py

@@ -1,0 +1,52 @@
+import numpy as np
+import pandas as pd
+
+
+def date2years(d, default_min_date=1900, default_max_date=2017):
+    """
+    Converts date to integers provided the dates are month-specific
+    :param d: input date to be converted
+    :param default_min_date: default min date
+    :param default_max_date: default max date
+    :return: date (None if min_date < max_date), min_date, max_date
+    """
+    if pd.notnull(d):
+        first_jan_this_year = pd.datetime(year=d.year, month=1, day=1)
+        day_of_this_year = d - first_jan_this_year
+        first_jan_next_year = pd.datetime(year=d.year + 1, month=1, day=1)
+        days_in_this_year = first_jan_next_year - first_jan_this_year
+        min_date = d.year + day_of_this_year / days_in_this_year
+        next_month_d = pd.datetime(year=d.year + (0 if d.month < 12 else 1),
+                                   month=(d.month if d.month < 12 else 0) + 1, day=1)
+        nm_day_of_this_year = (next_month_d - first_jan_this_year).days - 1
+        max_date = d.year + nm_day_of_this_year / days_in_this_year.days
+        date = min_date + np.random.random(1)[0] * (max_date - min_date)
+        return date, date, date
+    else:
+        return None, default_min_date, default_max_date
+
+
+if '__main__' == __name__:
+    import argparse
+
+    parser = argparse.ArgumentParser()
+
+    parser.add_argument('--data', default='/home/azhukova/projects/HIV1-UK/data/metadata/metadata.uk.tab', type=str)
+    parser.add_argument('--dates', default='/home/azhukova/projects/HIV1-UK/data/metadata/lsd2.interval.dates', type=str)
+    parser.add_argument('--date_col', default='sampledate_my', type=str)
+    params = parser.parse_args()
+
+    df = pd.read_csv(params.data, index_col=0, sep='\t')[[params.date_col]]
+    df[params.date_col] = pd.to_datetime(df[params.date_col].astype(str).str.replace('.0', '', regex=False),
+                                         infer_datetime_format=True)
+    m_date = date2years(df[~pd.isna(df[params.date_col])][params.date_col].min())[1]
+    M_date = 2017
+    df[['date', 'lower', 'upper']] = \
+        df[params.date_col].apply(lambda _: date2years(_, m_date, M_date)).apply(pd.Series)
[…]

### 2023-07-25 hiv_b_uk/py/extract_dates_lsd.py

@@ -1,0 +1,66 @@
+import numpy as np
+import pandas as pd
+
+
+INTERVAL = 'interval'
+
+INPUT = 'input'
+
+
+def date2years(d, default_min_date=1900, default_max_date=2017, type=INPUT):
+    """
+    Converts date to integers provided the dates are month-specific
+    :param d: input date to be converted
+    :param default_min_date: default min date
+    :param default_max_date: default max date
+    :param type: interval or input: if interval is specified the min and max dates returns would correspond
+        to the 1st day of the date's month and the 1st day of the month following the date's one.
+        Otherwise, the middle of the month will be returned.
+    :return: date (None if min_date < max_date), min_date, max_date
+    """
+    if pd.notnull(d):
+        first_jan_this_year = pd.datetime(year=d.year, month=1, day=1)
+        day_of_this_year = d - first_jan_this_year
+        first_jan_next_year = pd.datetime(year=d.year + 1, month=1, day=1)
+        days_in_this_year = first_jan_next_year - first_jan_this_year
+        min_date = d.year + day_of_this_year / days_in_this_year
+        next_month_d = pd.datetime(year=d.year + (0 if d.month < 12 else 1),
+                                   month=(d.month if d.month < 12 else 0) + 1, day=1)
+        nm_day_of_this_year = (next_month_d - first_jan_this_year).days - 1
+        max_date = d.year + nm_day_of_this_year / days_in_this_year.days
+        if INTERVAL == type:
+            # return None, min_date, max_date
+            date = min_date + np.random.random(1)[0] * (max_date - min_date)
+            return date, date, date
+        else:
+            date = (min_date + max_date) / 2
+            return date, date, date
+    else:
+        return None, default_min_date, default_max_date
+
+
+if '__main__' == __name__:
+    import argparse
+
+    parser = argparse.ArgumentParser()
+
+    parser.add_argument('--data', default='/home/azhukova/projects/HIV1-UK/data/metadata/metadata.uk.tab', type=str)
+    parser.add_argument('--dates', default='/home/azhukova/projects/HIV1-UK/data/metadata/lsd2.interval.dates', type=str)
+    parser.add_argument('--date_col', default='sampledate_my', type=str)
+    parser.add_argument('--type', required=False, choices=(INTERVAL, INPUT), type=str, default=INTERVAL)
+    params = parser.parse_args()
+
[…]

### 2023-07-26 article/article.tex
Partners can also notify

@@
 \usepackage{booktabs}
+\usepackage{stackengine}
 
@@
 
-Another important source of heterogeneity that none of these models account for is inflicted by non-random sampling.
+Another important source of heterogeneity that none of these models accounts for is inflicted by non-random sampling.
 We can propose different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability over time intervals with the BDSKY, however, none of them can account for non-randomness of sampling, in particular due to contact tracing and partner notification.
 
-In this study we propose an extension of the MTBD framework that allows for modeling non-random sampling due to partner notification (PN), MTBD-PN.
-In the rest of the article, we describe the MTBD-PN model family and its assumptions, present a transmission tree simulator under these models, derive the equations for MTBD-PN likelihood calculation, and propose a test for detecting partner notification in pathogen phylogenetic trees.
-We also present a maximum likelihood parameter estimator for the simplest model of the MTBD-PN family, BD-PN, and test it on simulated data.
+In this study we propose an extension of the BD model that allows for modeling non-random sampling due to partner notification (PN), BD-PN.
+In the rest of the article, we describe the BD-PN model and its assumptions, present a transmission tree simulator under this model, derive the equations for BD-PN likelihood calculation, and propose a test for detecting partner notification in pathogen phylogenetic trees.
+We also present a maximum likelihood parameter estimator for BD-PN, and test it on simulated data.
 Finally, we apply it to the study of HIV-1 B epidemic in the UK.
 
-\section{MTBD-PN Framework}
-In a pathogen transmission tree $\mathscr{T}$ (approximated by a time-scaled pathogen phylogeny) the tips represent sampled pathogens, patient state transitions occur along the branches, and bifurcations (i.e.
+\section{BD-PN Framework}
+In a pathogen transmission tree $\mathscr{T}$ (approximated by a time-scaled pathogen phylogeny) the tips represent sampled pathogens %, patient state transitions occur along the branches, 
+while bifurcations (i.e.
 internal nodes) correspond to pathogen transmissions (Fig.~\ref{fig:tt}).
@@
 %\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
-\includestandalone[width=1\textwidth]{Fig_tree}
+\includestandalone[width=0.6\textwidth]{Fig_tree}
 \caption{A transmission tree $\mathscr{T}$ with $n=5$ external nodes (i.e.
@@
 
-\subsection{MTBD Framework}
-A general MTBD model describes $m$ possible individual states.
+\subsection{BD Model with incomplete sampling}
+Under the basic birth-death model~\citep{Stadler2009} all the individuals are in the same state.
+They can transmit with a constant rate $\lambda$, get removed with a constant rate $\psi$, and their pathogen can be sampled upon removal with a constant probability $\rho$.
+
+%A general MTBD model describes $m$ possible individual states.
 An individual in state $k \in 1:m$ can be removed (i.e., become non-infectious, e.g., due to healing, starting a treatment, self-isolation, or death) at a rate $\psi_k$, change their state to a state $l \in 1:m$ at a rate $\mu_{kl}$ (where $\mu_{kk} = 0$), and transmit their pathogen to an individual in a state $l \in 1:m$ at rate $\lambda_{kl}$.
@@
 
-The MTBD model can be described with master equations representing the likelihood densities of evolving as on the observed transmission tree, with time $t$ going forward from the root ($t=0$) till the time of the last sampled tip ($t=T$).
-These likelihood densities depend on the probability $U_s(t)$ of an unobserved transmission tree that started from one individual at time $t$ in state $s \in 1:m$ and evolved till $T$, without them or any of their induced cases being sampled:
+This model can be described with master equations representing the likelihood densities of evolving as on the observed transmission tree, with time $t$ going forward from the root ($t=0$) till the time of the last sampled tip ($t=T$).
+These likelihood densities depend on the probability $U(t)$ of an unobserved transmission tree that started from one individual at time $t$ and evolved till $T$, without them or any of their induced cases being sampled:
 
@@
 \begin{cases}
-\dot{U}_{s}(t) = &\Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) U_{s}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-    & - \sum\limits_{k=1}^{m} \mu_{sk} U_{k}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, folowed by unsampled evolution from the new state }\\
-    &- \sum\limits_{k=1}^{m} \lambda_{sk} U_{s}(t)U_{k}(t) \;  \textit{\color{gray} $\leftarrow$ transmission, followed by unsampled evolutions of both subtrees}\\
-    &- \psi_s (1 - \rho)\;  \textit{\color{gray} $\leftarrow$ removal without sampling}\\
-U_{s}(T) = & 1\;  \textit{\color{gray} $\leftarrow$ the probability to stay unsampled over time 0 is 1} \label{eq:Us}
+\dot{U}(t) = &\Big(\lambda + \psi\Big) U(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+    &- \lambda U^2(t) \;  \textit{\color{gray} $\leftarrow$ transmission, followed by unsampled evolutions of both subtrees}\\
+    &- \psi (1 - \rho)\;  \textit{\color{gray} $\leftarrow$ removal without sampling}\\
+U(T) = & 1\;  \textit{\color{gray} $\leftarrow$ the probability to stay unsampled over time 0 is 1} \label{eq:Us}
 \end{cases}
@@
 
-With the equations~(\ref{eq:Us}), we can write down the equations for the probability densities $p_{s,si}^{(i)}(t)$ of evolving as on an observed tree branch that finishes at time $t_i$ in state $si$, starting  on this branch at a time $t \leq t_i$ in state $s$.
-The master equations for the MTBD models were initially developed by ~\citet{Stadler2013a}, however what we present here, is their parallelizable version that we proposed in~\cite{zhukovaFastAccurateMaximumLikelihood2022}:
+With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch that finishes at time $t_i$, starting on this branch at a time $t \leq t_i$.
+The master equations for the BD models were initially developed by ~\citet{Stadler2009}, however what we present here, is their branch-specific formulation that we proposed in~\cite{zhukovaFastAccurateMaximumLikelihood2022}:
 
@@
 \begin{cases}
-\dot{p}_{s,si}^{(i)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, folowed by evolution from the new state }\\
-    & - \sum\limits_{k=1}^{m} \lambda_{sk} p_{s,si}^{(i)}(t)U_{k}(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
-    & - \sum\limits_{k=1}^{m} \lambda_{sk} p_{k,si}^{(i)}(t)U_{s}(t)\;  \textbf{\textit{\color{gray} $\leftarrow$ transmission, where the donor subtree stayed unsampled}}\\
-p_{s,si}^{(i)}(t_i) = & 
-    \begin{cases}
-    0 & \textit{\color{gray} if $i$ is in state $k \neq si$}\\
-    1 & \textit{\color{gray} if $i$ is in state $si$}
+\dot{p}^{(i)}(t) = & \Big(\lambda + \psi\Big) p^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+    & - 2 \lambda p^{(i)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where one of the subtrees stayed unsampled}\\
+p^{(i)}(t_i) = 1
 \end{cases}\label{eq:p}
-\end{cases}
 \end{equation}
 
-If the states of all the observed nodes of the tree $\mathscr{T}$ are known, then its likelihood density under parameter values $\Theta$ can be calculated as $L(\mathscr{T}|\Theta)$:
+Using this equation, we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ as:
 
-{\scriptsize
-\begin{alignat*}{3}
-log L(\mathscr{T}|\Theta) =  \sum\limits_{i \in tips}  log(&\psi_{si}  \rho) & \textit{\color{gray} $\leftarrow$ sampling of $n$ tips} \\
- +\sum\limits_{i \in \textit{internal nodes}} log \Big(&\sum\limits_{k=1}^{m} \lambda_{si,k}  & \textit{\color{gray} $\leftarrow$ $n - 1$ transmission events}\\
- & \cdot \big(p_{si,si0}^{(i0)}(t_i)p_{k,si1}^{(i1)}(t_i) + p_{k,si0}^{(i0)}(t_i)p_{si,si1}^{(i1)}(t_i)\big)\Big) & \textit{\color{gray}$\leftarrow$ child branch evolutions}  \label{eq:mtbd_tree_likelihood}
-\end{alignat*}}
+\begin{equation}
+\scriptsize
+\begin{split}
+log L(\mathscr{T}|\Theta) =  &\sum\limits_{i \in tips}  log(\psi \rho)  \textit{\color{gray} $\leftarrow$ sampling of $n$ tips} \\
+ &+\sum\limits_{i \in \stackanchor{\tiny\textit{internal}}{\tiny\textit{nodes}}} log \big(2 \lambda p^{(i0)}(t_i)p^{(i1)}(t_i)\big)   \textit{\color{gray} $\leftarrow$ $n - 1$ transmission events,}\\
+ & ~~~~~~~~~~~~~~~\textit{\color{gray} followed by child branch evolutions (each can be a donor)}  \label{eq:mtbd_tree_likelihood}
+\end{split}
+\end{equation}
+
+The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g.
+by finding parameter values that maximize it, in maximum likelihood framework).
+In particular, it permits inference of the following epidemiological parameters:
+
+\begin{itemize}
+\item \textit{effective reproduction number} $R_e = \frac{\lambda}{\psi}$, expected number of individuals directly infected by an infectious case;
+\item \textit{infectious time} $\frac{1}{\psi}$, time during which an infectious individual can further spread the epidemic.
+\end{itemize}
+
+Importantly, the BD model is asymptotically unidentifiable (see Remark 3.4 in~\citep{Stadler2009}), but to become identifiable it requires one of the parameters to be fixed.
+In practice, it is often the sampling probability, as it may be approximated from epidemiological data $\rho$ (e.g., the proportion of sampled cases among the declared ones) or the infectious time $\frac{1}{\psi}$ (estimated from observations of infected cases).
 
@@
 
-We now propose a non-Markovian extension of the MTBD framework that adds partner notification (PN), for a tree with known node states.
-In MTBD-PN, at the moment of sampling the sampled individual might notify their most recent partner with a given probability $\rho_n$.
-Upon notification, partner is removed almost instantaneously (modeled via a notified removal rate $\psi_n >> \psi_k \;\forall k \in 1:m$).
-MTBD-PN hence has $2k^2 + 1$ classical MTBD and 2 additional parameters:
+We now propose a non-Markovian extension of the BD model that adds partner notification (PN).
+In BD-PN, at the moment of sampling the sampled individual might notify their most recent partner with a given probability $\rho_n$.
+Upon notification, partner is removed almost instantaneously (modeled via a notified removal rate $\psi_n >> \psi$).
+BD-PN hence has 3 classical BD and 2 additional parameters:
 \begin{itemize}
- \item $\mu_{kl} \;\forall 1 \leq k,l  \leq m, k \neq l$ -- constant state change rate from $k$ to $l$;
- \item $\lambda_{kl} \;\forall 1 \leq k,l \leq m$ -- constant transmission rate: an infected individual in state $k$ transmits their pathogen (which corresponds to an internal node in the transmission tree), the newly infected individual is in state $l$;
- \item $\psi_k \;\forall 1 \leq k \leq m$ -- constant removal rate: an infected individual in state $k$ stops being infectious at this rate.
-This corresponds to a tip in the transmission tree (observed or hidden);
+ \item $\lambda$ -- constant transmission rate: an infected individual transmits their pathogen to a newly infected individual (which corresponds to an internal node in the transmission tree);
+ \item $\psi$ -- constant removal rate: an infected individual stops being infectious at this rate. This corresponds to a tip in the transmission tree (observed or hidden);
  \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
  \item $\rho_n$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
- \item $\psi_n >> \psi_k  \;\forall 1 \leq k \leq m$ -- constant notified removal rate: a notified individual (if not yet removed via standard procedure) stops being infectious and gets observed with this rate.
+ \item $\psi_n >> \psi$ -- constant notified removal rate: a notified individual (if not yet removed via standard procedure) stops being infectious and gets observed with this rate.
 This creates an observed tip in the transmission tree.
@@
 
-This model is a simplification of the notification process, in particular it makes 5 assumptions:
+This model is a simplification of the notification process, in particular it makes 4 assumptions:
 \begin{enumerate}
@@
 \item notified individuals are always observed upon removal;
-\item notified individuals do not notify further;
+%\item notified individuals do not notify further;
 \item after the notification, notified individuals do not transmit further;
@@
 
+Note that the most recent partner relationship is not necessarily symmetrical, as the most recent partner of $i$ might have transmitted their virus further after the contact with $i$.
+
 \subsubsection{Mathematical formulation}
-$p_{s,si}^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along a tree branch, provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$.
+$p^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along a tree branch, provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$.
 These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.
 
-We will use an \textbf{oriented probability} $p^{(i,o)}_{s,si}(t)$ for the cases when the individual at the end of the branch is the same as at time $t$:
+We will use an \textbf{oriented probability} $p^{(i,o)}(t)$ for the cases when the individual at the end of the branch is the same as at time $t$:
 
@@
 \begin{cases}
-\dot{p}_{s,si}^{(i,o)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i,o)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i,o)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, followed by evolution from the new state }\\
-    & - \sum\limits_{k=1}^{m} \lambda_{sk} p_{s,si}^{(i,o)}(t)U_{k}(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
-p_{s,si}^{(i,o)}(t_i) = & 
-    \begin{cases}
-    0 & \textit{\color{gray} if $i$ is in state $k \neq si$}\\
-    1 & \textit{\color{gray} if $i$ is in state $si$}
-    \end{cases}\label{eq:p-o}
+\dot{p}^{(i,o)}(t) = & \Big(\lambda + \psi\Big) p^{(i,o)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+    & - \lambda p^{(i,o)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
+p^{(i,o)}(t_i) = 1\label{eq:p-o}
 \end{cases}
@@
 
-Additionally we define a \textbf{non-hidden probability} $p^{(i,nh)}_{s,si}(t)$ for the cases when no hidden transmission occurred along the branch:
+Additionally we define a \textbf{non-hidden probability} $p^{(i,nh)}(t)$ for the cases when no hidden transmission occurred along the branch:
 
@@
 \begin{cases}
-\dot{p}_{s,si}^{(i,nh)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i,nh)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i,nh)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, followed by evolution from the new state }\\
-p_{s,si}^{(i,nh)}(t_i) = & 
-    \begin{cases}
-    0 & \textit{\color{gray} if $i$ is in state $k \neq si$}\\
-    1 & \textit{\color{gray} if $i$ is in state $si$}
-    \end{cases}\label{eq:p-nh}
+\dot{p}^{(i,nh)}(t) = & \Big(\lambda + \psi\Big) p^{(i,nh)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+p^{(i,nh)}(t_i) = 1\label{eq:p-nh}
 \end{cases}
@@
 
-Under the MTBD-PN model, the probability density $q_{si,sj}(i)$ of a branch connecting a node $i$ in state $si$ to its parent node $j$ in state $sj$ (without taking into account the event at the end of the branch) can be written as:
-
-\begin{equation}
-\scriptsize
-q_{sj,si}(i) = 
-\begin{cases}
-q^{(u)}_{sj,si}(i) = p^{(i)}_{sj,si}(t_j) \textit{\color{gray} ~if $i$ is an \textbf{unnotified non-notifier}}\\
-q_{sj,si}^{(n)}(i) = 
-p_{sj,si}^{(i,nh)}(t_j) \textit{\color{gray} ~~if $i$ is a tip corresponding to a \textbf{notifier}}\\
-~~~~~~~~~~~~ \textit{\color{gray}(parent is the last transmission, no hidden transmission along the branch)}
-\\
-q^{(p)}_{sj,si}(i)  = p^{(i, o)}_{sj,si}(t_j)  \textit{\color{gray} ~if $i$ is an internal node whose individual will be notified after $t_i$}
-\\
-q^{(p)}_{sj,si}(i, r)  = 
-\begin{cases}
-p^{(r, o)}_{sj,si}(t_j) e^{- \psi_n (t_i - t_r)} \textit{\color{gray} if $i$ is a \textbf{partner} tip notified by $r$ at time $t_r \leq t_i$}\\
-p^{(i, o)}_{sj,si}(t_j) \textit{\color{gray}~~ if $i$ is a \textbf{partner} tip sampled via a standard procedure} \\
-~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\textit{\color{gray} before the notification by $r$: $t_i < t_r$}\\
-\end{cases}
-\end{cases}
-\label{eq:branch}
-\end{equation}
-
-Using the formulas~(\ref{eq:branch}), we can calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}, where for each visited node $i$, and for each possible state $1 \leq sj \leq m$ at the beginning of the branch connecting it to its parent $j$  we calculate the likelihood density of its subtree if its branch corresponds to (1) an unnotified individual $l^{(u)}_{sj}(i)$; (2) a notifier $l^{(n)}_{sj}(i)$ (for tips only); or (3) an (eventually) notified partner $l^{(p)}_{sj}(i, r)$ (notified by the individual corresponding to the tip $r$ at time $t_r$).
-
-For a tip $i$ in state $si$ these values can be calculated as:
-\begin{equation}
-\scriptsize
-\begin{array}{ll}
-l^{(u)}_{sj}(i) &= q^{(u)}_{sj}(i)\psi_{si}\rho (1-\rho_{n})\\
-l^{(n)}_{sj}(i) &= q^{(n)}_{sj}(i)\psi_{si}\rho \rho_{n}\\
-l^{(p)}_{sj}(i, r) &= q^{(p)}_{sj}(i,r)\psi_{n}\\
-\end{array}
-\label{eq:tip}
-\end{equation}
-
-Let us now consider an internal node $i$ in state $si$ with two child nodes, $i0$ and $i1$. We will start with the configuration where the $i$'s branch corresponds to an unnotified individual:
-
-\begin{equation}
-\scriptsize
-\begin{split}
-l^{(u)}_{sj}(i) = &q^{(u)}_{sj}(i) \sum\limits_{k=1}^{m} \lambda_{si,k} \\
-\cdot \Big(~&
-l^{(u)}_{si}(i0)l^{(u)}_{k}(i1) + l^{(u)}_{k}(i0)l^{(u)}_{si}(i1)  \textit{\color{gray} $\leftarrow$ no notification on child branches} \\
-& + is\_tip(i0) \big(l^{(n)}_{si}(i0)l^{(p)}_{k}(i1,i0) + l^{(n)}_{k}(i0)l^{(p)}_{si}(i1,i0) \big)\textit{\color{gray} $\leftarrow$ notification of someone in $i1$'s subtree by $i0$} \\
-&+ is\_tip(i1) \big(l^{(p)}_{si}(i0,i1)l^{(n)}_{k}(i1) + l^{(p)}_{k}(i0,i1)l^{(n)}_{si}(i1) \big)~\Big) \textit{\color{gray} $\leftarrow$ notification of someone in $i0$'s subtree by $i1$},\\
-\textit{where } & is\_tip(r) = 1 \textit{ if r is a tip and $= 0$ if r is an internal node.}
-\end{split}\label{eq:lu}
-\end{equation}
-
-Note that a notification will only be possible if the notifying branch is external as the notifier only notifies the most recent partner, i.e.
-the transmission corresponds to the notifier's parent node.
-%; (2) there is at least one tip in the notified subtree that can be a partner, i.e.
-was sampled after the notifier and whose parent's time was before notification (i.e.
-there was no transmission between notification and sampling).
-
-In the other case (when an internal nodes's branch corresponds to a partner who will eventually get notified by a node $r$), the individual on the path connecting this branch to the notified tip stays the same (the partner).
-Hence, all the transmissions on this path are oriented and correspond to the partner transmitting to another recipient individual.
-Moreover, this recipient's branch cannot be notified by our partner (as in our model notified partners do not notify further):
-
-\begin{equation}
-\scriptsize
-\begin{split}
-l^{(p)}_{sj}(i, r) = q^{(p)}_{sj}(i,r) \sum\limits_{k=1}^{m}  \lambda_{si,k}\Big(&l^{(p)}_{si}(i0,r)l^{(u)}_{k}(i1) \textit{\color{gray} $\leftarrow$ notification in $i0$'s subtree} \\
- &+ l^{(u)}_{k}(i0)l^{(p)}_{si}(i1,r)\Big)  \textit{\color{gray} $\leftarrow$ notification in $i1$'s subtree} 
-\end{split}\label{eq:lu}
-\end{equation}
-
-Note that $l^{(p)}_{sj}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time is known, i.e.
-once we have climbed to an internal node corresponding to the transmission between the notifier (corresponding to one of its children, must be a tip) and the partner (the other child).
-In the partner subtree only the contribution coming from the partner tip branch depends on the notification time.
-Hence while climbing the tree from tips to the root, for each internal node $i$ and each of the possible states $sj \in 1:m$ at the beginning of its branch, we can store an array of pre-calculated values for notified internal paths leading to each of its tips.
-The values of this array will store for each tip (e.g.
-$a$) its sampling time $t_a$ and a configuration $C_{sj}(i,a)$, representing the subtree likelihood provided $a$ corresponds to the same notified partner individual as $i$ (but without the $a$'s tip branch contribution):
-
-\begin{equation}
-\scriptsize
-\begin{split}
-C_{sj}(i,a) = &\begin{cases}
-1 \textit{\color{gray}~~~if $i = a$}
-\\
-q^{(p)}_{sj}(i)
-\sum\limits_{k=1}^{m}\lambda{si,k} \cdot l_{k}^{(u)}\Big(child(i, \not a)\Big)C_{si}\Big(child(i, a),a\Big) \textit{\color{gray}~~otherwise}\\ 
-\end{cases},
-\\ &\textit{where child(i,a) is the child of i whose subtree contains a},
-\\
- &\textit{~~~~~~~~child(i,$\not a$) is the child of i whose subtree does not contain a.}
-\end{split}
-\label{eq:c}
-\end{equation}
-
-%Note that the path connecting $j$ to the parent of $i$ in this configuration corresponds to the same partner individual, hence this individual is the donor at all transmissions (internal nodes) on this path, hence at the moments of these transmissions the individual's state is known (as we only consider trees with known node states).
-At each internal node in this path the other branch (leading to a non-$j$'s subtree) corresponds to a recipient of the transmission occurring at that node, where the donor is the partner individual corresponding to $j$ (and $i$) and hence will not notify the recipient (as in our model's assumptions notified partners do not notify further).
-
-Once a potential notifier tip $r$ is identified (i.e., $r$ is a sister tip of the currently visited node $i$), we can calculate $l^{(p)}_{sj}(i, r)$ as:
-\begin{equation}
-\scriptsize 
-l^{(p)}_{sj}(i, r) = \sum\limits_{a \in tips(i)}C_{sj}(i, a) l^{(p)}_{sa}(a, r)\label{eq:lp}
-\end{equation}
-
-Finally, the tree $\mathscr{T}$ likelihood density $\mathscr{L}(\mathscr{T}|\Theta)$ under MTBD-PN model with parameters $\Theta$ can be calculated as $l^{(u)}_{state(root)}(root)$, as the root can only correspond to an unnotified individual (since there was no observed transmission before it).
-
-%The computational of $\mathscr{L}(\mathscr{T}|\Theta)$ includes $O(N^2)$ resolutions of system~(\ref{eq:p}) (for different times and initial conditions) in the worst case (caterpillar tree) and $O(NlogN)$ in the best case (for a balanced tree), where $N$ is the number of tips in $\mathscr{T}$.
-
-\subsection{BD-PN}
-A particular example of the MTBD model generating trees with known node states is the basic birth-death model~\citep{Stadler2009} where all the individuals are in the same state.
-They can transmit with a constant rate $\lambda$, get removed with a constant rate $\psi$, be sampled upon removal with a probability $\rho$.
-Below we describe the special case of equations~(\ref{eq:Us}-\ref{eq:lp}) for its PN extension BD-PN.
-
-%\begin{equation}
-%\scriptsize
-%\begin{cases}
-%\dot{U}(t) &= \Big(\lambda + \psi\Big) U(t) - \lambda U^2(t) - \psi (1 - \rho)\\
-%U(T) &=  1\\
-%\dot{p}^{(i)}(t) &=  \Big(\lambda + \psi\Big) p^{(i)}(t) - 2 \lambda p^{(i)}(t)U(t)\\
-%p^{(i)}(t_i) &= 1
-%\end{cases}
-%\end{equation}
-%
-%\begin{equation}
-%L(\mathscr{T}|\{\lambda, \psi, \rho\}) = n log(\psi\rho) + \sum\limits_{i \in \substack{internal\\nodes}} 2\lambda p^{(i0)}(t_i) p^{(i1)}(t_i)
-%\end{equation}
-%
-
-The master equation for this model are a simplified version of the general case:
-
-\begin{equation}
-\scriptsize
-\begin{cases}
-\dot{U}(t) &= \Big(\lambda + \psi\Big) U(t) - \lambda U^2(t) - \psi (1 - \rho)\\
-U(T) &=  1\\
-\dot{p}^{(i)}(t) &=  \Big(\lambda + \psi\Big) p^{(i)}(t) - 2 \lambda p^{(i)}(t)U(t)\\
-\dot{p}^{(i,o)}(t) &=  \Big(\lambda + \psi\Big) p^{(i,o)}(t) - \lambda p^{(i,o)}(t)U(t)\\
-\dot{p}^{(i,nh)}(t) &=  \Big(\lambda + \psi\Big) p^{(i,nh)}(t)\\
-p^{(i)}(t_i) &= p^{(i,o)}(t_i) = p^{(i,nh)}(t_i) = 1
-\end{cases}
-\end{equation}
-
-%
-%These equations have the following closed form solutions:
-%\begin{equation}
-%\scriptsize
-%\begin{split}
-%&\begin{cases}
-%U(t) = \frac{\lambda + \psi + c_1 \frac{c_2 e^{t c_1} - 1}{c_2 e^{tc_1} + 1}}{2\lambda}\\
-%\\
-%p^{(i)}(t) = \frac{c_3^2 e^{c_1 (t - t_i)}}{(c_2e^{c_1 t} + 1)^2} \\
-%\\
-%p^{(i,o)}(t) =  \frac{c_3 e^{1/2 (\lambda + \psi + c_1)(t - t_i)}}{c_2e^{c_1 t} + 1} \\
-%\\
-%p^{(i,nh)}(t) =  e^{(\lambda + \psi)(t - t_i)}
-%\end{cases},\\
-%& \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi}e^{-Tc_1},\; c_3 = c_2 e^{c_1 t_i} + 1
-%\end{split}
-%\end{equation}
-%
-%
-%These equations have the following closed form solutions:
-%\begin{equation}
-%\scriptsize
-%\begin{split}
-%&\begin{cases}
-%U(t) = \frac{\lambda + \psi + c_1 \frac{c_2 e^{ c_1 (t-T)} - 1}{c_2 e^{c_1 (t-T)} + 1}}{2\lambda}\\
-%\\
-%p^{(i)}(t) = \Big(\frac{c_2 e^{c_1 (t_i - T)} + 1}{c_2e^{c_1 (t - T)} + 1}\Big)^2e^{c_1 (t - t_i)} \\
-%\\
-%p^{(i,o)}(t) =  \frac{(c_2 e^{c_1 (t_i -T)} + 1) \Big(e^{c_1 (t -t_i)} e^{(\lambda + \psi)(t - t_i)}\Big)^{1/2}}{c_2e^{c_1 (t - T)} + 1} \\
-%\\
-%p^{(i,nh)}(t) =  e^{(\lambda + \psi)(t - t_i)}
-%\end{cases},\\
-%& \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi},
-%\end{split}
-%\end{equation}
-
-These equations have the following closed form solutions:
+The equations~\ref{eq:Us}, \ref{eq:p}, \ref{eq:p-o}, and \ref{eq:p-nh} have the following closed form solutions:
 \begin{equation}
@@
 & ~~~~~~~~E_1 = c_2 e^{c_1 (t -T)},\; E_2 = c_2 e^{c_1 (t_i -T)},\; E_3 = e^{c_1 (t -t_i)}, \; E_4 =  e^{(\lambda + \psi) (t -t_i)}\\
-\end{split}
+\end{split}\label{eq:ps}
 \end{equation}
 
-The likelihood of a tree $\mathscr{T}$ under BD-PN model with parameter values $\Theta$ can be calculated as $\mathscr{L}(\mathscr{T}|\Theta) = l^{(u)}(root)$, where $l^{(u)}(i)$ is a likelihood of node $i$'s subtree if the branch connecting $i$ to its parent node (of length zero for the root) correspond to an unnotified individual.
-It can be calculated, along with $l^{(n)}(i)$ (likelihood when $i$ is a notifier  tip) and $l^{(p)}(i, r)$ (likelihood when $i$'s branch corresponds to an eventually notified partner, notified by $r$) for each tree node $i$ with a pruning algorithm.
+%Under the BD-PN model, the probability density $q(i)$ of a branch connecting a node $i$ to its parent node $j$ (without taking into account the event at the end of the branch) can be written as:
+%
+%
+%\begin{equation}
+%\scriptsize
+%q(i) = 
+%\begin{cases}
+%q^{(u)}(i) = p^{(i)}(t_j) \textit{\color{gray} ~if $i$ is an \textbf{unnotified non-notifier}}\\
+%q^{(n)}(i) = 
+%p^{(i,nh)}(t_j) \textit{\color{gray} ~~if $i$ is a tip corresponding to a \textbf{notifier}}\\
+%~~~~~~~~~~~~ \textit{\color{gray}(parent is the last transmission, no hidden transmission along the branch)}
+%\\
+%q^{(p)}(i)  = p^{(i, o)}(t_j)  \textit{\color{gray} ~if $i$ is an internal or tip node whose individual will be notified after $t_i$}
+%\\
+%q^{(p)}(i, r)  = p^{(r, o)}(t_j) e^{- \psi_n (t_i - t_r)} \textit{\color{gray} if $i$ is a \textbf{partner} tip notified by $r$ at time $t_r \leq t_i$}
+%\end{cases}
+%\label{eq:branch}
+%\end{equation}
 
-For a tip $i$ whose branch starts at time $t_j$ and finishes at time $t_i$, these values can be calculated as:
+Using the formulas~(\ref{eq:ps}), we can calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}, where for each visited node $i$ we calculate the likelihood density of its subtree if the branch connecting it to its parent node $j$ corresponds to (1) an unnotified individual $l^{(u)}(i)$; or (2) an (eventually) notified partner $l^{(p)}(i, r)$ (notified by the individual corresponding to the tip $r$ at time $t_r$).
+For each visited tip we additionally consider whether the corresponding individual notified their most recent partner or not.
+
+%For a tip $i$ in state $si$ these values can be calculated as:
+%\begin{equation}
+%\scriptsize
+%\begin{array}{ll}
+%l^{(u)}(i) &= \begin{cases} l^{(u,n)}(i) = q^{(u)}(i)\psi\rho
+%\rho_{n} \textit{\color{gray} ~~~~~~~~if $i$ notified their most recent patner}\\
+%l^{(u,\not{n})}(i) = q^{(u)}(i)\psi\rho (1-\rho_{n}) \textit{\color{gray} ~if $i$ did not notify}
+%\end{cases}\\
+%l^{(p)}(i, r) &= q^{(p)}(i,r)\psi_{n}\cdot \begin{cases}
+%\rho_{n} \textit{\color{gray} ~~~~~~~~if $i$ notified their most recent patner}\\
+%(1-\rho_{n}) \textit{\color{gray} ~if $i$ did not notify}
+%\end{cases}\\
+%\end{array}
+%\label{eq:tip}
+%\end{equation}
+
+For a tip $i$ with a parent node $j$ these values can be calculated as:
 \begin{equation}
@@
 \begin{array}{ll}
-l^{(u)}(i) &= p^{(i)}(t_j)\psi\rho (1-\rho_{n})\\
-l^{(n)}(i) &= p^{(i,nh)}(t_j)\psi\rho \rho_{n}\\
-l^{(p)}(i, r) &=\begin{cases}
-p^{(r, o)}(t_j) e^{- \psi_n (t_i - t_r)} \textit{\color{gray} if $i$ is a \textbf{partner} tip notified by $r$ at time $t_r \leq t_i$}\\
-p^{(i, o)}(t_j) \textit{\color{gray}~~ if $i$ is a \textbf{partner} tip sampled via a standard procedure} \\
-~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\textit{\color{gray} before the notification by $r$: $t_i < t_r$}\\
-\end{cases}
+l^{(u,n)}(i) &= p^{(i,nh)}(t_j)\psi\rho
+\rho_{n} \textit{\color{gray} ~if $i$ was not notified but notified their most recent patner}\\
+l^{(u,\not{n})}(i) &= p^{i}(t_j)\psi\rho (1-\rho_{n}) \textit{\color{gray} ~if $i$ was not notified and did not notify}\\
+l^{(p,n)}(i,r) &= \begin{cases}
+p^{(i, nh)}(t_j)\psi\rho\rho_{n} \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and notified their most recent patner}\\
+p^{(r, nh)}(t_j) e^{- \psi_n (t_i - t_r)} \psi_n \rho_{n} \textit{\color{gray} ~if $i$ got notified at time $t_r \leq t_i$ and notified}
+\end{cases}\\
+l^{(p,\not{n})}(i,r) &= \begin{cases}
+p^{(i, o)}(t_j)\psi\rho(1 - \rho_{n}) \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and did not notify}\\
+p^{(r, o)}(t_j) e^{- \psi_n (t_i - t_r)} \psi_n (1-\rho_{n}) \textit{\color{gray} ~if $i$ got notified at time $t_r \leq t_i$ and did not notify}
+\end{cases}\\
 \end{array}
+\label{eq:tip}
 \end{equation}
 
-For an internal node $i$ whose branch starts at time $t_j$ and finishes at time $t_i$, with two child nodes, $i0$ and $i1$, these values can be calculated as:
+Let us now consider an internal node $i$ with two child nodes, $i_0$ and $i_1$.
+We will start with the configuration where the $i$'s branch corresponds to an unnotified individual.
+We will consider three possibilities: (i) when both $i_0$ and $i_1$ are internal nodes; (ii) when both $i_0$ and $i_1$ are tips; and (iii) when one of them is an internal node and the other one is a tip.
+
+%Note that a notification will only be possible if the notifying branch is external as the notifier only notifies the most recent partner, i.e.
+the transmission corresponds to the notifier's parent node.
+
+First, let's assume that (i) $i_0$ and $i_1$ are internal. Then none of them can be a notifier (since notifiers correspond to tips), and hence none of them is notified:
 
@@
 \scriptsize
-\begin{array}{ll}
-l^{(u)}(i) &= p^{(i)}(t_j) 2\lambda 
-\Big(l^{(u)}(i0)l^{(u)}(i1) 
-+ is\_tip(i0) l^{(n)}(i0)l^{(p)}(i1,i0)
-+ is\_tip(i1) l^{(p)}(i0,i1)l^{(n)}(i1)\Big),\\
-& \textit{where } is\_tip(x) = 1 \textit{ if x is a tip and $= 0$ if x is an internal node.}\\
-l^{(p)}(i, r) &= p^{(i,o)}(t_j) \lambda\Big(l^{(p)}(i0,r)l^{(u)}(i1) + l^{(u)}(i0)l^{(p)}(i1,r)\Big),\\
-& \textit{where r is the notifier.}\\
-\end{array}
+l^{(u)}(i) = p^{i}(t_j) 2\lambda l^{(u)}(i_0)l^{(u)}(i_1) \label{eq:lu-int-int}
 \end{equation}
 
-As in the general MTBD-PN case, $l^{(p)}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time is known, i.e.
-once we have climbed to an internal node corresponding to the transmission between the notifier (corresponding to one of its children, must be a tip) and the partner (the other child).
-As in the general MTBD-PN case, while climbing the tree from tips to the root, for each internal node $i$ we can pre-calculate and store an array of values for notified internal paths leading to each of its tips.
-The values of this array will store for each tip (e.g.
-$a$) its sampling time $t_a$ and a configuration $C(i,a)$, representing the subtree likelihood provided $a$ corresponds to the same notified partner individual as $i$ (but without the $a$'s tip branch contribution):
+Secondly, let's assume that (ii) $i_0$ and $i_1$ are tips. Then each of them could have notified the other one:
 
@@
 \begin{split}
-C(i,a) = &\begin{cases}
-1 \textit{\color{gray}~~~if $i == a$}
-\\
-p^{(i,o)}(t_j) \lambda \cdot l^{(u)}\Big(child(i, \not a)\Big)C\Big(child(i, a),a\Big) \textit{\color{gray}~~otherwise}\\ 
-\end{cases},
-\\ &\textit{where child(i,a) is the child of i whose subtree contains a},
-\\
- &\textit{~~~~~~~~child(i,$\not a$) is the child of i whose subtree does not contain a.}
-\end{split}
+l^{(u)}(i) = p^{i}(t_j) 2\lambda \cdot 
+\Big(&l^{(u,\not{n})}(i_0)l^{(u,\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ neigher $i_0$ nor $i_1$ notified} \\
+&+ l^{(u,n)}(i_0)l^{(p,\not{n})}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $i_0$ notified $i_1$ and $i_1$ did not notify} \\
+&+ l^{(u,n)}(i_1)l^{(p,\not{n})}(i_0,i_1) \textit{\color{gray}~$\leftarrow$ $i_1$ notified $i_0$ and $i_0$ did not notify} \\
+&+ l^{(p,n)}(i_0,i1)l^{(p,n)}(i_1,i_0)\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ and $i_0$ notified each other} \\
+\end{split}\label{eq:lu-tip-tip}
 \end{equation}
 
-Once a potential notifier tip $r$ is identified (i.e., $r$ is a sister tip of the currently visited node $i$), we can calculate $l^{(p)}(i, r)$ as:
+Lastly, let's assume that (iii) $i_0$ is a tip and $i_1$ is internal (the scenario where $i_1$ is a tip and $i_0$ is internal can be obtained by swapping the labels).
+Then $i_0$ could have notified $i_1$:
+
 \begin{equation}
 \scriptsize
-l^{(p)}(i, r) = \sum\limits_{a \in tips(i)}C(i, a) l^{(p)}(a, r)
+\begin{split}
+l^{(u)}(i) = p^{i}(t_j) 2\lambda \cdot 
+\Big(&l^{(u,\not{n})}(i_0)l^{(u)}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ did not notify $i_1$ } \\
+&+ l^{(u,n)}(i_0)l^{(p)}(i_1,i_0)\Big) \textit{\color{gray}~$\leftarrow$ $i_0$ notified $i_1$} \\
+\end{split}\label{eq:lu-int-tip}
 \end{equation}
 
-%\subsection{BDEI-PN}
-%Another particular example of the MTBD model generating trees with known node states is the basic birth-death exposed-infectious model (BDEI)~\citep{Stadler2014} where   infectious ($I$) individuals can transmit (at a rate $\lambda$) or be removed (at a rate $\psi$), and hence create an internal or external node in the transmission tree.
-Upon removal they can be sampled with a probability $\rho$.
-The recipient of the transmission is at first in state $E$ (exposed, already infected but not yet infectious), they can eventually become infectious ($I$) at a state-change rate $\mu$.
-No other type of state transitions, transmissions or removals are allowed.
-Below we describe the special case of equations~(\ref{eq:u}-\ref{eq:lp}) for its PN extension BDEI-PN, assuming only infectious individuals can be removed and sampled even with notification (e.g., because the viral load of the exposed individuals is yet undetectable).
-%
-%
-%The master equations for the BDEI-PN model are a special case of the MTBD-PN equations:
-%
 %\begin{equation}
 %\scriptsize
-%\begin{cases}
-%\dot{U_E}(t) = &\mu U_E(t)- \mu U_I(t)\\    
-%\dot{U_I}(t) = &\Big(\lambda + \psi\Big) U_I(t)- \lambda U_I(t)U_E(t)- \psi (1 - \rho)\\
-%U_E(T) = & U_I(T) = 1\\
-%\dot{p}_E^{(i)}(t) = & \mu p_E^{(i)}(t) - \mu p_I^{(i)}(t)\\
-%p_E^{(i)}(t_i) = & 0\\
-%p_E^{(i,o)}(t_i) = & p_E^{(i,nh)}(t_i) = p_E^{(i)}(t_i)\\
-%\dot{p}_I^{(i)}(t) = & \Big(\lambda + \psi\Big) p^{(i)}(t) - \lambda \Big(p_I^{(i)}(t)U_E(t) + p_E^{(i)}(t)U_I(t)\Big)\\
-%\dot{p}_I^{(i,o)}(t) = & \Big(\lambda + \psi\Big) p^{(i,o)}(t) - \lambda p_I^{(i,o)}(t)U_E(t)\\
-%\dot{p}_I^{(i,nh)}(t) = & \Big(\lambda + \psi\Big) p^{(i,nh)}(t)\\
-%p_I^{(i)}(t_i) = & p_I^{(i,o)}(t_i) = p^{(i,nh)}(t_i) = 1\label{eq:up-bd}
+%l^{(u)}(i) = p^{i}(t_j) 2\lambda \\
+%\cdot \begin{cases}
+%l^{(u)}(i_0)l^{(u)}(i_1) \textit{\color{gray} ~if both $i_0$ and $i_1$ are internal} \\
+%l^{(u,\not{n})}(i_0)l^{(u)}(i_1) + l^{(u,n)}(i_0)l^{(p)}(i_1,i_0) \textit{\color{gray}~if $i_0$ is a tip and $i_1$ is internal} \\
+%l^{(u,\not{n})}(i_1)l^{(u)}(i_0) + l^{(u,n)}(i_1)l^{(p)}(i_0,i_1) \textit{\color{gray}~if $i_1$ is a tip and $i_0$ is internal} \\
+%l^{(u,\not{n})}(i_0)l^{(u,\not{n})}(i_1) \\
+%~~~~+ l^{(u,n)}(i_0)l^{(p,\not{n})}(i_1,i_0) + l^{(u,n)}(i_1)l^{(p,\not{n})}(i_0,i_1)\\
+%~~~~+ l^{(p,n)}(i_0,i1)l^{(p,n)}(i_1,i_0) \textit{\color{gray}~if both $i_0$ and $i_1$ are tips} \\
 %\end{cases}
+%%& + is\_tip(i_0)l^{(u,n)}(i_0)l^{(p)}(i_1,i_0)\textit{\color{gray} $\leftarrow$ notification of someone in $i_1$'s subtree by $i_0$} \\
+%%&+ is\_tip(i_1)l^{(u,n)}(i_1)l^{(p)}(i_0,i_1)~\Big) \textit{\color{gray} $\leftarrow$ notification of someone in $i_0$'s subtree by $i_1$},\\
+%%\textit{where } & is\_tip(r) = 1 \textit{ if r is a tip and $= 0$ if r is an internal node.}\label{eq:lu}
 %\end{equation}
-%
-%The likelihood of a tree $\mathscr{T}$ under BDEI-PN model with parameter values $\Theta$ can be calculated as $\mathscr{L}(\mathscr{T}|\Theta) = l_I^{(u)}(root)$, where $l_s^{(u)}(i)$ is a likelihood of node $i$'s subtree if the branch connecting $i$ to its parent node (of length zero for the root) starts in state $s \in \{E, I\}$ and correspond to an unnotified individual.
-It can be calculated, along with $l_s^{(p)}(i, notifier)$ (likelihood when $i$'s branch corresponds to an eventually notified partner) for each tree node $i$ with a pruning algorithm.
-%
-%
-%For a tip $i$ whose branch starts at time $t_j$ in state $s \in \{E, I\}$ and finishes at time $t_i$ in state $I$, these values can be calculated as:
+
+In the other case an internal node $i$ corresponds to a partner who will eventually get notified by a node $r$.
+In this case the individual on the path connecting this branch to the notified tip stays the same (the partner).
+Hence, all the transmissions on this path are oriented and correspond to the partner transmitting to another recipient individual.
+%Moreover, this recipient's branch cannot be notified by our partner (as in our model notified partners do not notify further):
+Again, we will consider three possibilities for $i$'s child nodes: (i) when both $i_0$ and $i_1$ are internal nodes; (ii) when both $i_0$ and $i_1$ are tips; and (iii) when one of them is an internal node and the other one is a tip.
+
+In any of these scenarios, the notification must have happened after the partner tip branch start (as one of the model assumptions is that partners do not transmit after notification). Hence:
+\begin{equation}
+\scriptsize
+l^{(p)}(i, r) = 0 \textit{~if $i$ is internal and $t_i > t_r$ or $i$ is a tip and $t_{parent(i)} > t_r$} \\
+\label{eq:lp-zero}
+\end{equation}
+
+First, let's assume that (i) $i_0$ and $i_1$ are internal. Then none of them can be a notifier (since notifiers correspond to tips), and hence exactly one of them is notified (by $r$):
+
+\begin{equation}
+\scriptsize
+\begin{split}
+l^{(p)}(i, r) = p^{(i,o)}(t_j) \lambda \cdot
+\Big(&l^{(p)}(i_0,r)l^{(u)}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ is notified by $r$} \\
+& + l^{(p)}(i_1,r)l^{(u)}(i_0)\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ is notified by $r$} \\
+ \end{split}
+\label{eq:lp-int-int}
+\end{equation}
+
+Secondly, let's assume that (ii) $i_0$ and $i_1$ are tips. Then each of them could have notified the other one (in addition to one of them being notified by $r$):
+
+\begin{equation}
+\scriptsize
+\begin{split}
+l^{(p)}(i, r) = p^{(i,o)}(t_j) \lambda \cdot
+\Big(&l^{(p,\not{n})}(i_0,first(r,i_1))l^{(u,n)}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by both $r$ and $i_1$, but did not notify $i_1$} \\
+&+ l^{(p,n)}(i_0,first(r,i_1))l^{(p,n)}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by both $r$ and $i_1$, and notified $i_1$} \\
+&+l^{(p,\not{n})}(i_1,first(r,i_0))l^{(u,n)}(i_0) \textit{\color{gray}~$\leftarrow$ $i_1$ got notified by both $r$ and $i_0$, but did not notify $i_0$} \\
+&+ l^{(p,n)}(i_1,first(r,i_0))l^{(p,n)}(i_0, i_1) \textit{\color{gray}~$\leftarrow$ $i_1$ got notified by both $r$ and $i_0$, and notified $i_0$} \\
+&+l^{(p,n)}(i_0,r)l^{(p,\not{n})}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by $r$, and notified $i_1$} \\
+& + l^{(p,n)}(i_1,r)l^{(p,\not{n})}(i_0, i_1) \textit{\color{gray}~$\leftarrow$ $i_1$ got notified by $r$, and notified $i_0$} \\
+&+l^{(p,\not{n})}(i_0,r)l^{(u,\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by $r$, and did not notify $i_1$} \\
+& + l^{(p,\not{n})}(i_1,r)l^{(u,\not{n})}(i_0)\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ got notified by $r$, and did not notify $i_0$} \\
+\textit{\color{gray}~where } &\color{gray}first(a,b) = 
+\begin{cases}
+a \textit{~if $t_a \leq t_b$}\\
+b \textit{~if $t_a > t_b$.}
+\end{cases} 
+ \end{split}
+\label{eq:lu-tip-tip}
+\end{equation}
+
+Lastly, let's assume that (iii) $i_0$ is a tip and $i_1$ is internal (the scenario where $i_1$ is a tip and $i_0$ is internal can be obtained by swapping the labels).
+Then $i_0$ could have notified $i_1$ (in addition to one of them being notified by $r$):
+
+\begin{equation}
+\scriptsize
+\begin{split}
+l^{(p)}(i, r) = p^{(i,o)}(t_j) \lambda \cdot
+\Big(&l^{(p,n)}(i_0,r)l^{(p)}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by $r$, and notified $i_1$} \\
+& + l^{(p,\not{n})}(i_0,r)l^{(u)}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by $r$, and did not notify $i_1$} \\
+& +l^{(p)}(i_1,first(r,i_0))l^{(u, n)}(i_0) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by both $r$ and $i_0$} \\
+& + l^{(p)}(i_1,r)l^{(u, \not{n})}(i_0)\Big) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by $r$} \\
+\textit{\color{gray}~where } &\color{gray}first(a,b) = 
+\begin{cases}
+a \textit{~if $t_a \leq t_b$}\\
+b \textit{~if $t_a > t_b$.}
+\end{cases} 
+ \end{split}
+\label{eq:lp-tip-int}
+\end{equation}
+
 %\begin{equation}
 %\scriptsize
-%\begin{array}{ll}
-%l^{(u)}_{s}(i) &= p_s^{(i)}(t_j)\psi\rho (1-\rho_{n})\\
-%l^{(n)}_{s}(i) &= p_s^{(i,nh)}(t_j)\psi\rho \rho_{n}\\
-%l^{(p)}_{s}(i, r) &= p_s^{(r,o)}(t_j)e^{-\psi_n (t_i - t_r)}\psi_{n}\\
-%\end{array}
+%l^{(p)}(i, r) = p^{(i,o)}(t_j) \lambda
+%\begin{cases}
+%l^{(p)}(i_0,r)l^{(u)}(i_1) + l^{(p)}(i_1,r)l^{(u)}(i_0) \textit{\color{gray} ~if both $i_0$ and $i_1$ are internal} \\
+%l^{(p,n)}(i_0,r)l^{(p)}(i_1,i_0) + l^{(p,\not{n})}(i_0,r)l^{(u)}(i_1) \\
+%~~~~~~+ \begin{cases}
+%l^{(p)}(i_1,first(r,i_0))l^{(u, n)}(i_0) \textit{\color{gray} ~if $t_{i0} \geq t_{i1}$}  \\
+%0 \textit{\color{gray} ~if $t_{i0} < t_{i1}$}
+% \end{cases}\\
+%~~~~~~+ l^{(p)}(i_1,r)l^{(u, \not{n})}(i_0) \textit{\color{gray} ~if $i_0$ is a tip and $i_1$ is internal} \\
+%l^{(p,n)}(i_1,r)l^{(p)}(i_0,i_1) + l^{(p,\not{n})}(i_1,r)l^{(u)}(i_0) \\
+%~~~~~~+ \begin{cases}
+%l^{(p)}(i_0,first(r,i_1))l^{(u, n)}(i_1) \textit{\color{gray} ~if $t_{i1} \geq t_{i0}$}  \\
+%0 \textit{\color{gray} ~if $t_{i1} < t_{i0}$}
+% \end{cases}\\
+%~~~~~~+ l^{(p)}(i_0,r)l^{(u), \not{n}}(i_1) \textit{\color{gray} ~if $i_1$ is a tip and $i_0$ is internal} \\
+%l^{(p,\not{n})}(i_0,first(r,i_1))l^{(u,n)}(i_1) + l^{(p,n)}(i_0,first(r,i_1))l^{(p,n)}(i_1, i_0)\\ 
+%~~~~~~+l^{(p,\not{n})}(i_1,first(r,i_0))l^{(u,n)}(i_0) + l^{(p,n)}(i_1,first(r,i_0))l^{(p,n)}(i_0, i_1)\\
+%~~~~~~+l^{(p,n)}(i_0,r)l^{(p,\not{n})}(i_1, i_0) + l^{(p,n)}(i_1,r)l^{(p,\not{n})}(i_0, i_1)\\ 
+%~~~~~~+l^{(p,\not{n})}(i_0,r)l^{(u,\not{n})}(i_1) + l^{(p,\not{n})}(i_1,r)l^{(u,\not{n})}(i_0)
+%\textit{\color{gray} ~if both $i_0$ and $i_1$ are tips} \\
+% \end{cases}
+%\label{eq:lu}
 %\end{equation}
-%
-%For an internal node $i$ whose branch starts at time $t_j$ is state $s$ and finishes at time $t_i$ in state $I$, with two child nodes, $i0$ and $i1$, these values can be calculated as:
-%
-%\begin{equation}
-%\scriptsize
-%\begin{array}{ll}
-%l^{(u)}(i) =& p_s^{(i)}(t_j) \lambda \\
-%& \cdot \Big(\big(l_I^{(u)}(i0)l_E^{(u)}(i1) + l_E^{(u)}(i0)l_I^{(u)}(i1)\big) \\
-%&~~+ is\_tip(i0) \big(l_I^{(n)}(i0)l_E^{(p)}(i1,i0) + l_E^{(n)}(i0)l_I^{(p)}(i1,i0)\big)\\
-%&~~+ is\_tip(i1) \big(l_I^{(p)}(i0,i1)l_E^{(n)}(i1) + l_E^{(p)}(i0,i1)l_I^{(n)}(i1)\big)\Big),\\
-%& \textit{where } is\_tip(x) = 1 \textit{ if x is a tip and $= 0$ if x is an internal node.}\\
-%l^{(p)}(i, r) &= p_s^{(i,o)}(t_j) \lambda\Big(l_I^{(p)}(i0,r)l_E^{(u)}(i1) + l_E^{(u)}(i0)l_I^{(p)}(i1,r)\Big),\\
-%& \textit{where r is the notifier.}\\
-%\end{array}
-%\end{equation}
-%
-%As in the general MTBD-PN case, $l_s^{(p)}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time is known, i.e.
+
+Note that $l^{(p)}(a, b)$ depends on the notifier $b$ and can only be calculated once the notification time is known, i.e.
 once we have climbed to an internal node corresponding to the transmission between the notifier (corresponding to one of its children, must be a tip) and the partner (the other child).
-As only the contributions coming from the partner tip branches depend on the notification times, while climbing the tree for each internal node $i$ and each state $s \in \{E, I\}$ at the time of its branch start $t_j$, we can store an array of pre-calculated values for notified internal paths for each of its tips (e.g.
-$a$), describing the tip's sampling time $t_a$ and a configuration $C_s(i,a)$.
-Note, that the partner branch for any transmission after the transmission between the notifier and the partner, is always the donor one (hence, starting in state $I$).
+The calculation is done recursively by descending into the subtree.
+However, as all the unotified equation parts ($l^{(u)}(i)$) only depend on the corresponding node $i$, they can be pre-calculated and reused to speed up the recursive steps.
+
+%In the partner subtree only the contribution coming from the partner tip branch depends on the notification time.
+Hence while climbing the tree from tips to the root, for each internal node $i$ we can store an array of pre-calculated values for notified internal paths leading to each of its tips.
+The values of this array will store for each tip (e.g.
+$a$) its sampling time $t_a$ and a configuration $C(i,a)$, representing the subtree likelihood provided $a$ corresponds to the same notified partner individual as $i$ (but without the $a$'s tip branch contribution):
 %
@@
 %\begin{split}
-%C_s(i,a) = &\begin{cases}
-%1 \textit{\color{gray}~~~if $i==a$}
+%C_{sj}(i,a) = &\begin{cases}
+%1 \textit{\color{gray}~~~if $i = a$}
 %\\
-%p_s^{(i,o)}(t_j) \lambda \cdot l_E^{(u)}\Big(child(i, \not a)\Big)C_I\Big(child(i, a),a\Big) \textit{\color{gray}~~otherwise}\\ 
+%q^{(p)}_{sj}(i)
+%\sum\limits_{k=1}^{m}\lambda{si,k} \cdot l_{k}^{(u)}\Big(child(i, \not a)\Big)C_{si}\Big(child(i, a),a\Big) \textit{\color{gray}~~otherwise}\\ 
 %\end{cases},
@@
 %\end{split}
+%\label{eq:c}
 %\end{equation}
-%
-%Once a potential notifier tip $r$ is identified (i.e., $r$ is a sister tip of the currently visited node $i$), we can calculate $l_s^{(p)}(i, r)$ as:
+
+%Note that the path connecting $j$ to the parent of $i$ in this configuration corresponds to the same partner individual, hence this individual is the donor at all transmissions (internal nodes) on this path, hence at the moments of these transmissions the individual's state is known (as we only consider trees with known node states).
+At each internal node in this path the other branch (leading to a non-$j$'s subtree) corresponds to a recipient of the transmission occurring at that node, where the donor is the partner individual corresponding to $j$ (and $i$) and hence will not notify the recipient (as in our model's assumptions notified partners do not notify further).
+
+%Once a potential notifier tip $r$ is identified (i.e., $r$ is a sister tip of the currently visited node $i$), we can calculate $l^{(p)}_{sj}(i, r)$ as:
 %\begin{equation}
 %\scriptsize 
-%l_s^{(p)}(i, r) = 
-%\begin{cases}
-%\sum\limits_{a \in tips(i)}C_s(i, a) l_I^{(p)}(a, r)\textit{\color{gray}~~if $i$ is an internal node or $s=I$}\\
-%l_E^{(p)}(a, r)\textit{\color{gray}~~if $i=a$ and $s=E$}
-%\end{cases}
+%l^{(p)}_{sj}(i, r) = \sum\limits_{a \in tips(i)}C_{sj}(i, a) l^{(p)}_{sa}(a, r)\label{eq:lp}
 %\end{equation}
+
+\bigskip
+
+The tree $\mathscr{T}$ likelihood density $\mathscr{L}(\mathscr{T}|\Theta)$ under BD-PN model with parameters $\Theta=\{\lambda, \psi, \rho, \psi_n, \rho_n\}$ can be calculated as $l^{(u)}(root)$, as the root can only correspond to an unnotified individual (since there was no observed transmission before it).
+
+%The computational of $\mathscr{L}(\mathscr{T}|\Theta)$ includes $O(N^2)$ resolutions of system~(\ref{eq:p}) (for different times and initial conditions) in the worst case (caterpillar tree) and $O(NlogN)$ in the best case (for a balanced tree), where $N$ is the number of tips in $\mathscr{T}$.
 
@@
 As we previously described in~\citep{zhukovaFastAccurateMaximumLikelihood2022}, the likelihood calculation with MTBD models can be easily extended to forests.
-This extension applies also to MTBD-PN models.
+This extension applies also to the BD-PN model.
 Forests are useful for cases when a (sub-)epidemic started with several infected individuals (e.g.
@@
 To illustrate the PN test performance, we applied it to the trees simulated under BD and under BD-PN models.
-We simulated 100 trees with 5\,000--10\,000 tips under the BD-PN model with our tree simulator.
+We simulated 100 trees with 500--1\,000 tips under the BD-PN model with our tree simulator.
 For each tree the parameter values were drawn randomly within the following bounds:
@@
 partner notification probability $\rho_n \in ]0.01/\rho, 0.9]$.
-We also simulated 100 trees of 5\,000--10\,000 tips under the BD model.
+We also simulated 100 trees of 500--1\,000 tips under the BD model.
 For each tree in the BD data set the parameter values ($\lambda, \psi,$ and $\rho$) were the same as the ones selected for the corresponding tree of the BD-PN dataset.
@@
 We then run the PN tests on each tree of the both datasets.
-For all the trees in the BD dataset the PN test value was above $0.05$ and varied between $0.33$ and $0.63$ (i.e.
-100\% true negative results).
-For 97 (out of 100) trees in the BD-PN dataset the PN test value was below $0.05$ (i.e.
-97\% true positive and 3\% false negative results).
-Our test therefore showed absolute specificity and high sensitivity.
+For 98 out of 100 trees in the BD dataset the PN test value was above $0.05$ (i.e.
+98\% true negative and 2\% false positive results), the mean PN test value was $0.468$.
+For 96 (out of 100) trees in the BD-PN dataset the PN test value was below $0.05$ (i.e.
+96\% true positive and 4\% false negative results), the mean PN test value was $0.008$.
+Our test therefore showed both high specificity and sensitivity.
 
@@
 
-To assess the performance of our maximum-likelihood estimator for the BD-PN model, we used the 100 trees with 5\,000--10\,000 tips simulated under the BD-PN model as described in the ``PN test'' section.
+To assess the performance of our maximum-likelihood estimator for the BD-PN model, we used the 100 trees with 500--1\,000 tips simulated under the BD-PN model as described in the ``PN test'' section.
 
@@
 
-As expected with a maximum likelihood estimator, the tree likelihoods for the estimated parameter values were higher than the tree likelihoods for the real parameter values in all the settings.
+As expected with a maximum likelihood estimator, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the real parameter values in all the settings.
 

### 2023-07-26 bdpn/bd.py

@@ -1,52 +1,10 @@
 import numpy as np
 
 from bdpn.tree_manager import TIME, annotate_tree
+from bdpn.formulas import get_c1, get_c2, get_E2, get_E1, get_log_p, get_u
 
 
-def get_c1(la, psi, rho):
-    return np.power(np.power((la - psi), 2) + 4 * la * psi * rho, 0.5)
-
-
-def get_c2(la, psi, rho, T, c1=None):
-    if c1 is None:
-        c1 = get_c1(la, psi, rho)
-    return (c1 + la - psi) / (c1 - la + psi) * np.exp(-T * c1)
-
-
-def get_c3(la, psi, rho, T, ti, c1=None):
-    if c1 is None:
-        c1 = get_c1(la, psi, rho)
-    # c2 = get_c2(la, psi, rho, T)
-    # return c2 * np.exp(c1 * ti) + 1
-    # let's inline c2 here to reduce overflow
-    return (c1 + la - psi) / (c1 - la + psi) * np.exp(c1 * (ti - T)) + 1
-
-#
-# def get_u_wa(t, la, psi, rho, T):
-#     x = t - T
-#     c = np.power(-np.power(la, 2) - 4 * la * psi * rho + 2 * la * psi - np.power(psi, 2), 0.5)
-#     return 1 / (2 * la) * (c * np.tan(0.5 * (-x * c + 2 * np.arctan((la - psi) / c))) + la + psi)
-
-
-def get_u(t, la, psi, rho, T):
-    c1 = get_c1(la, psi, rho)
-    # c2 = get_c2(la, psi, rho, T, c1=c1)
-    # Inline c2 to avoid overflow
-    c2_exp_c1 = (c1 + la - psi) / (c1 - la + psi) * np.exp(c1 * (t - T))
-    # return (la + psi + c1 * ((c2_exp_c1 - 1) / (c2_exp_c1 + 1))) / (2 * la)
-    return (la + psi) / (2 * la) + c1 / (2 * la) * ((c2_exp_c1 - 1) / (c2_exp_c1 + 1))
-
-
-def get_log_p(t, la, psi, rho, T, ti):
-    c1 = get_c1(la, psi, rho)
-    c3 = get_c3(la, psi, rho, T, ti, c1=c1)
-    # c2 = get_c2(la, psi, rho, T, c1=c1)
-    # return (2 * np.log(c3) + (c1 * (t - ti))) - 2 * np.log(c2 * np.exp(c1 * t) + 1)
-    # Inline c2 to avoid overflow
[…]

### 2023-07-26 bdpn/bdpn.py

@@ -1,29 +1,43 @@
 import os
+from multiprocessing.pool import ThreadPool
 
 import numpy as np
 from pastml.tree import read_forest
+from scipy import integrate
 
-from bdpn.bd import get_c1, get_c3, get_log_p, get_u
 from bdpn.parameter_estimator import optimize_likelihood_params, rescale_log
 from bdpn.tree_manager import TIME, annotate_tree
+from bdpn.formulas import get_log_p, get_c1, get_c2, get_E2, get_E1, get_log_po, get_log_pnh, get_log_po_from_p_pnh, \
+    get_u
 
 
-def get_log_p_o(t, la, psi, rho, T, ti):
-    c1 = get_c1(la, psi, rho)
-    c3 = get_c3(la, psi, rho, T, ti, c1=c1)
-    # c2 = get_c2(la, psi, rho, T, c1=c1)
-    # return np.log(c3) + (1/2 * (la + psi + c1) * (t - ti)) - np.log(c2 * np.exp(c1 * t) + 1)
-    # Inline c2 to avoid underflow
-    return np.log(c3) + (1 / 2 * (la + psi + c1) * (t - ti)) \
-        - np.log((c1 + la - psi) / (c1 - la + psi) * np.exp(c1 * (t - T)) + 1)
+# def get_log_hidden_partner_notification(t, la, psi, rho, psi_n, T, ti, c1=None, c3=None, c4=None):
+#     # return np.log(
+#     #     integrate.quad(lambda tau: np.exp(get_log_p(t, la, psi, rho, T, tau, c1=c1) + np.log(2 * la)
+#     #                                       + get_log_p_nh(tau, la, psi, ti)
+#     #                                       + get_log_p_o(tau, la, psi, rho, T, ti, c1=c1, c3=c3)
+#     #                                       + (psi_n * (ti - T))), t, ti)[0])
+#     # we inline everything we can in the above expression
+#
+#     if c4 is None:
+#         c4 = (c1 + la - psi) / (c1 - la + psi)
+#     la_plus_psi = la + psi
+#     c5 = (3 * la_plus_psi + c1) / 2
+#     non_tau_part = c1 * t - 2 * np.log(c4 * np.exp(c1 * (t - T)) + 1) \
+#                    - c5 * ti \
+#                    + np.log(c3) \
+#                    + psi_n * (ti - T)
+#
+#     def get_v(tau, non_tau_part):
[…]

### 2023-07-26 bdpn/formulas.py

@@ -1,0 +1,52 @@
+import numpy as np
+
+
+def get_c1(la, psi, rho):
+    return np.power(np.power(la - psi, 2) + 4 * la * psi * rho, 0.5)
+
+
+def get_c2(la, psi, c1):
+    la_minus_psi = la - psi
+    return (c1 + la_minus_psi) / (c1 - la_minus_psi)
+
+
+def get_E1(c1, c2, t, T):
+    return c2 * np.exp(c1 * (t - T))
+
+
+def get_E2(c1, c2, ti, T):
+    return c2 * np.exp(c1 * (ti - T))
+
+
+def get_E3(c1, t, ti):
+    return np.exp(c1 * (t - ti))
+
+
+def get_E4(la, psi, t, ti):
+    return np.exp((la + psi) * (t - ti))
+
+
+def get_u(la, psi, c1, E1):
+    two_la = 2 * la
+    return (la + psi) / two_la + c1 / two_la * ((E1 - 1) / (E1 + 1))
+
+
+def get_log_p(c1, t, ti, E1, E2):
+    # return np.power((E2 + 1) / (E1 + 1), 2) * E3
+    # return 2 * (np.log(E2 + 1) - np.log(E1 + 1)) + np.log(E3)
+    return 2 * (np.log(E2 + 1) - np.log(E1 + 1)) + c1 * (t - ti)
+
+
+def get_log_po(la, psi, c1, t, ti, E1, E2):
+    # return (E2 + 1) / (E1 + 1) * np.power(E3 * E4, 0.5)
+    # return np.log(E2 + 1) - np.log(E1 + 1) + (np.log(E3) + np.log(E4)) / 2
+    return np.log(E2 + 1) - np.log(E1 + 1) + (c1 + la + psi) * (t - ti) / 2
+
+
+def get_log_po_from_p_pnh(log_p, log_pnh):
+    return (log_p + log_pnh) / 2
+
+
+def get_log_pnh(la, psi, t, ti):
+    # return E4
+    return (la + psi) * (t - ti)

### 2023-07-26 bdpn/model_distinguisher.py

@@ -355,11 +355,15 @@ def main():
         all_cherries.extend(pick_cherries(tree, include_polytomies=True))
     all_cherries = sorted(all_cherries, key=lambda _: getattr(_.root, TIME))
 
-    logging.info('Picked {} cherries with {} roots.'.format(len(all_cherries), len({_.root for _ in all_cherries})))
+    n_cherries = len(all_cherries)
+    logging.info('Picked {} cherries with {} roots.'.format(n_cherries, len({_.root for _ in all_cherries})))
 
     results = []
     root_times = []
-    for i in range(0, len(all_cherries) // params.block_size):
+    if n_cherries < params.block_size:
+        raise ValueError('The block size (argument --block_size) must be smaller '
+                         'than the number of cherries in the input tree ({}).'.format(n_cherries))
+    for i in range(0, n_cherries // params.block_size):
         cherries = all_cherries[i * params.block_size: (i + 1) * params.block_size]
         result = nonparametric_cherry_diff(cherries, repetitions=1e3)
         results.append(result)

### 2023-07-26 bdpn/parameter_estimator.py

@@ -1,6 +1,6 @@
 import numpy as np
 from scipy.optimize import minimize
-from scipy.stats import stats
+from scipy.stats import chi2
 
 from bdpn.tree_manager import TIME, annotate_tree
 
@@ -20,7 +20,7 @@ def rescale_log(loglikelihood_array):
 
     non_zero_loglh_array = loglikelihood_array[loglikelihood_array > -np.inf]
     if len(non_zero_loglh_array) == 0:
-        raise ValueError('Underflow')
+        return 0
     min_lh_value = np.min(non_zero_loglh_array)
     max_lh_value = np.max(non_zero_loglh_array)
 
@@ -33,18 +33,13 @@ def rescale_log(loglikelihood_array):
     return factors
 
 
-def optimize_likelihood_params(forest, input_parameters, loglikelihood, bounds, start_parameters, cis=False):
+def optimize_likelihood_params(forest, input_parameters, loglikelihood, bounds, start_parameters, cis=False, threads=1):
     """
     Optimizes the likelihood parameters for a given forest and a given MTBD model.
 
 
-    :param forest: a list of ete3.Tree trees, annotated with node states and times via features STATE_K and TI.
-    :param T: time at end of the sampling period
-    :param model: MTBD model containing starting parameter values
-    :param optimise: MTBD model whose rates indicate which parameters need to optimized:
-        positive rates correspond to optimized parameters
-    :param u: number of hidden trees, where no tip got sampled
-    :return: the values of optimized parameters and the corresponding loglikelihood: (MU, LA, PSI, RHO, best_log_lh)
+    :param forest: a list of ete3.Tree trees
+    :return: tuple: (the values of optimized parameters, CIs)
     """
     for tree in forest:
         if not hasattr(tree, TIME):
@@ -79,7 +74,7 @@ def optimize_likelihood_params(forest, input_parameters, loglikelihood, bounds, 
         if np.any(np.isnan(ps)):
             return np.nan
         ps = get_real_params_from_optimised(ps)
-        res = loglikelihood(forest, *ps, T)
+        res = loglikelihood(forest, *ps, T, threads=threads)
         # print("{}\t-->\t{:g}".format(ps, res))
         return -res
 
@@ -110,35 +105,39 @@ def optimize_likelihood_params(forest, input_parameters, loglikelihood, bounds, 
     optimised_cis[:, 1] = np.array(optimised_parameters)
 
     if cis:
[…]

### 2023-07-26 examples/main.py

@@ -1,8 +1,11 @@
 import os
+from datetime import datetime
 
 import numpy as np
 import pandas as pd
 from ete3 import Tree
+from matplotlib.pyplot import plot, show, title
+from scipy.stats import chi2
 
 import bdpn.bd as bd
 import bdpn.bdpn as bdpn
@@ -12,7 +15,7 @@ from bdpn.parameter_estimator import optimize_likelihood_params, AIC
 # nwk = os.path.join(DATA_DIR, 'tree.1.nwk')
 
 DATA_DIR = os.path.abspath(os.path.join(os.path.dirname(__file__), '..', 'simulations', 'trees', 'BDPN'))
-nwk = os.path.join(DATA_DIR, 'tree.6.nwk')
+nwk = os.path.join(DATA_DIR, 'tree.2.nwk')
 tree = Tree(nwk)
 df = pd.read_csv(nwk.replace('.nwk', '.log'))
 rho = df.loc[0, 'sampling probability']
@@ -24,19 +27,88 @@ psi = 1 / it
 psi_n = 1 / rt
 la = R0 * psi
 
+lk_real_bdpn = bdpn.loglikelihood([tree], la, psi, psi_n, rho, rho_n, threads=12)
+# lk_real_bd = bd.loglikelihood([tree], la, psi, rho)
+real_params = [la, psi, psi_n, rho, rho_n]
+# print('Real params: {}\t loglikelihood-bdpn: {}\t loglikelihood-bd: {}'.format(real_params, lk_real_bdpn, lk_real_bd))
+print('Real params: {}\t loglikelihood: {}'.format(real_params, lk_real_bdpn))
 
-lk_real_bdpn = bdpn.loglikelihood([tree], la, psi, psi_n, rho, rho_n)
-lk_real_bd = bd.loglikelihood([tree], la, psi, rho)
-print('Real params: {}\t loglikelihood-bdpn: {}\t loglikelihood-bd: {}'.format([la, psi, psi_n, rho, rho_n], lk_real_bdpn, lk_real_bd))
+input_parameters = np.array([None, None, None, rho, None])
+# bounds = np.array([[la / 10, la * 10], [psi / 10, psi * 10], [psi_n / 10, psi_n * 10], [1e-3, 1], [0, 1]])
+bounds = np.array([[0.05, 1], [0.05, 0.2], [2.5, 100], [0.01, 0.99999], [0.1, 0.99999]])
[…]

### 2023-07-26 examples/main_cuba.py

@@ -1,13 +1,13 @@
 import logging
 import os
-from collections import defaultdict
 
+import numpy as np
 from pastml.tree import read_forest, annotate_dates, DATE
 
 from bdpn import bdpn, bd
-from bdpn.model_distinguisher import pick_cherries, nonparametric_cherry_diff, pick_motifs
+from bdpn.model_distinguisher import pick_cherries, nonparametric_cherry_diff
 from bdpn.parameter_estimator import optimize_likelihood_params, AIC
-from bdpn.tree_manager import annotate_tree
+from bdpn.tree_manager import annotate_tree, TIME
 
 DATA_DIR = os.path.abspath(os.path.join(os.path.dirname(__file__), '..', 'trees'))
 
@@ -15,61 +15,99 @@ DATA_DIR = os.path.abspath(os.path.join(os.path.dirname(__file__), '..', 'trees'
 NWK_CUBA = os.path.join(DATA_DIR, 'D_CRF_19_timetree.resolved.nwk')
 min_cherry_num = 25
 
+
+def cut_tree(tree, threshold_date):
+    forest = []
+    todo = [tree]
+    while todo:
+        n = todo.pop()
+        date = getattr(n, DATE)
+        if date >= threshold_date:
+            parent = n.up
+            n.detach()
+            n.dist = date - threshold_date
+            forest.append(n)
+            if parent and len(parent.children) == 1:
+                child = parent.children[0]
+                if not parent.is_root():
+                    grandparent = parent.up
+                    grandparent.remove_child(parent)
+                    grandparent.add_child(child, dist=child.dist + parent.dist)
+                else:
+                    child.dist += parent.dist
+                    child.detach()
+                    tree = child
+        else:
+            todo.extend(n.children)
+    print("Cut the tree into a root tree of {} tips and {} {}-on trees of {} tips in total"
+          .format(len(tree), len(forest), threshold_date, sum(len(_) for _ in forest)))
+    return tree, forest
+
+
+def check_cherries(forest, block_size=20, log=None):
+    all_cherries = []
+    for tree in forest:
+        all_cherries.extend(pick_cherries(tree, include_polytomies=True))
+    all_cherries = sorted(all_cherries, key=lambda _: getattr(_.root, TIME))
+
+    logging.info('Picked {} cherries with {} roots.'.format(len(all_cherries), len({_.root for _ in all_cherries})))
+
[…]

### 2023-07-26 examples/main_uk.py

@@ -3,17 +3,20 @@ import os
 from collections import defaultdict
 
 import numpy as np
-from pastml.tree import read_forest, annotate_dates, DATE
+from matplotlib.pyplot import plot, show
+from pastml.tree import read_forest, annotate_dates, DATE, parse_nexus
 
 from bdpn import bdpn, bd
-from bdpn.model_distinguisher import pick_cherries, nonparametric_cherry_diff, pick_motifs
+from bdpn.model_distinguisher import pick_cherries, nonparametric_cherry_diff
 from bdpn.parameter_estimator import optimize_likelihood_params, AIC
-from bdpn.tree_manager import annotate_tree
+from bdpn.tree_manager import annotate_tree, TIME
 
 DATA_DIR = os.path.abspath(os.path.join(os.path.dirname(__file__), '..', 'trees'))
 
-# gotree resolve -i UK_B_raxml.lsd2.nwk | gotree brlen setmin -l 0.0027397260273972603 -o UK_B_raxml.lsd2.resolved.nwk
-NWK_UK = os.path.join(DATA_DIR, 'UK_B_raxml.lsd2.resolved.nwk')
+# min len of 1 h
+# lsd2 -i rooted_raxml.nwk -d lsd2.interval.dates -e 3 -s 3102 -u 0.0027397260273972603 -l 0 -m 20000
+NWK_UK = os.path.join(DATA_DIR, 'UK', 'timetree.nexus')
+PN_TEST_LOG = os.path.join(DATA_DIR, 'UK', 'cherry_test.txt')
 min_cherry_num = 50
 
 
@@ -48,7 +51,7 @@ def cut_tree(tree, threshold_date):
             n.detach()
             n.dist = date - threshold_date
             forest.append(n)
-            if len(parent.children) == 1:
+            if parent and len(parent.children) == 1:
                 child = parent.children[0]
                 if not parent.is_root():
                     grandparent = parent.up
@@ -65,59 +68,84 @@ def cut_tree(tree, threshold_date):
     return tree, forest
 
 
-def check_cherries(tree):
-    all_cherries = list(pick_cherries(tree, include_polytomies=True))
+def check_cherries(forest, block_size=100, log=None):
+    all_cherries = []
+    for tree in forest:
+        all_cherries.extend(pick_cherries(tree, include_polytomies=True))
+    all_cherries = sorted(all_cherries, key=lambda _: getattr(_.root, TIME))
+
[…]

### 2023-07-26 hiv_b_uk/py/cut_tree.py

@@ -1,29 +1,32 @@
-import numpy as np
-import pandas as pd
+from pastml.tree import parse_nexus, annotate_dates, DATE
 
 
-def date2years(d, default_min_date=1900, default_max_date=2017):
-    """
-    Converts date to integers provided the dates are month-specific
-    :param d: input date to be converted
-    :param default_min_date: default min date
-    :param default_max_date: default max date
-    :return: date (None if min_date < max_date), min_date, max_date
-    """
-    if pd.notnull(d):
-        first_jan_this_year = pd.datetime(year=d.year, month=1, day=1)
-        day_of_this_year = d - first_jan_this_year
-        first_jan_next_year = pd.datetime(year=d.year + 1, month=1, day=1)
-        days_in_this_year = first_jan_next_year - first_jan_this_year
-        min_date = d.year + day_of_this_year / days_in_this_year
-        next_month_d = pd.datetime(year=d.year + (0 if d.month < 12 else 1),
-                                   month=(d.month if d.month < 12 else 0) + 1, day=1)
-        nm_day_of_this_year = (next_month_d - first_jan_this_year).days - 1
-        max_date = d.year + nm_day_of_this_year / days_in_this_year.days
-        date = min_date + np.random.random(1)[0] * (max_date - min_date)
-        return date, date, date
-    else:
-        return None, default_min_date, default_max_date
+def cut_tree(tree, threshold_date):
+    forest = []
+    todo = [tree]
+    while todo:
+        n = todo.pop()
+        date = getattr(n, DATE)
+        if date >= threshold_date:
+            parent = n.up
+            n.detach()
+            n.dist = date - threshold_date
+            forest.append(n)
+            if parent and len(parent.children) == 1:
+                child = parent.children[0]
+                if not parent.is_root():
+                    grandparent = parent.up
+                    grandparent.remove_child(parent)
+                    grandparent.add_child(child, dist=child.dist + parent.dist)
+                else:
+                    child.dist += parent.dist
+                    child.detach()
+                    tree = child
+        else:
+            todo.extend(n.children)
+    print("Cut the tree into a root tree of {} tips and {} {}-on trees of {} tips in total"
+          .format(len(tree), len(forest), threshold_date, sum(len(_) for _ in forest)))
+    return tree, forest
[…]

### 2023-07-26 hiv_b_uk/py/extract_dates_lsd.py

@@ -2,20 +2,12 @@ import numpy as np
 import pandas as pd
 
 
-INTERVAL = 'interval'
-
-INPUT = 'input'
-
-
-def date2years(d, default_min_date=1900, default_max_date=2017, type=INPUT):
+def date2years(d, default_min_date=1900, default_max_date=2017):
     """
     Converts date to integers provided the dates are month-specific
     :param d: input date to be converted
     :param default_min_date: default min date
     :param default_max_date: default max date
-    :param type: interval or input: if interval is specified the min and max dates returns would correspond
-        to the 1st day of the date's month and the 1st day of the month following the date's one.
-        Otherwise, the middle of the month will be returned.
     :return: date (None if min_date < max_date), min_date, max_date
     """
     if pd.notnull(d):
@@ -28,13 +20,8 @@ def date2years(d, default_min_date=1900, default_max_date=2017, type=INPUT):
                                    month=(d.month if d.month < 12 else 0) + 1, day=1)
         nm_day_of_this_year = (next_month_d - first_jan_this_year).days - 1
         max_date = d.year + nm_day_of_this_year / days_in_this_year.days
-        if INTERVAL == type:
-            # return None, min_date, max_date
-            date = min_date + np.random.random(1)[0] * (max_date - min_date)
-            return date, date, date
-        else:
-            date = (min_date + max_date) / 2
-            return date, date, date
+        date = min_date + np.random.random(1)[0] * (max_date - min_date)
+        return date, date, date
     else:
         return None, default_min_date, default_max_date
 
@@ -47,7 +34,6 @@ if '__main__' == __name__:
     parser.add_argument('--data', default='/home/azhukova/projects/HIV1-UK/data/metadata/metadata.uk.tab', type=str)
     parser.add_argument('--dates', default='/home/azhukova/projects/HIV1-UK/data/metadata/lsd2.interval.dates', type=str)
     parser.add_argument('--date_col', default='sampledate_my', type=str)
-    parser.add_argument('--type', required=False, choices=(INTERVAL, INPUT), type=str, default=INTERVAL)
     params = parser.parse_args()
 
     df = pd.read_csv(params.data, index_col=0, sep='\t')[[params.date_col]]
@@ -56,7 +42,7 @@ if '__main__' == __name__:
     m_date = date2years(df[~pd.isna(df[params.date_col])][params.date_col].min())[1]
     M_date = 2017
[…]

### 2023-07-26 simulations/py/plot_error.py

@@ -22,12 +22,17 @@ if __name__ == "__main__":
     parser.add_argument('--estimates', type=str, help="estimated parameters")
     parser.add_argument('--pdf', type=str, help="plot")
     parser.add_argument('--tab', type=str, help="error table")
+    parser.add_argument('--pn_min', type=float, default=0., help="Only display cases with pn greater or equal to this value")
+    parser.add_argument('--pn_max', type=float, default=1, help="Only display cases with pn smaller than this value")
     params = parser.parse_args()
 
     df = pd.read_csv(params.estimates, sep='\t', index_col=0)
 
     real_df = df.loc[df['type'] == 'real', :]
+    real_df = real_df[(real_df['pn'] * real_df['p'] >= params.pn_min) & (real_df['pn'] * real_df['p'] < params.pn_max)]
+
     df = df.loc[df['type'] != 'real', :]
+    df = df.loc[real_df.index, :]
     types = sorted(df['type'].unique(), key=lambda _: next(i for (i, par) in enumerate(RATE_PARAMETERS) if par in _))
 
     for type in types:

### 2023-07-26 simulations/py/summary_table.py

@@ -8,9 +8,9 @@ if __name__ == "__main__":
     import argparse
 
     parser = argparse.ArgumentParser(description="Summarize errors.")
-    parser.add_argument('--estimated_p', nargs='+', type=str, help="estimated parameters")
-    parser.add_argument('--estimated_psi', nargs='+', type=str, help="estimated parameters")
-    parser.add_argument('--estimated_la', nargs='+', type=str, help="estimated parameters")
+    parser.add_argument('--estimated_p', nargs='*', default=[], type=str, help="estimated parameters")
+    parser.add_argument('--estimated_psi', nargs='*', default=[], type=str, help="estimated parameters")
+    parser.add_argument('--estimated_la', nargs='*', default=[], type=str, help="estimated parameters")
     parser.add_argument('--real', nargs='+', type=str, help="real parameters")
     parser.add_argument('--tab', type=str, help="estimate table")
     params = parser.parse_args()
@@ -40,6 +40,8 @@ if __name__ == "__main__":
             = [R0, it, rt, R0 / it, 1 / it, 1 / rt, p, pn, tips, 'real']
 
     for (est_list, fixed) in ((params.estimated_la, 'lambda'), (params.estimated_psi, 'psi'), (params.estimated_p, 'p')):
+        if not est_list:
+            continue
         for est in est_list:
             i = int(re.findall(r'[0-9]+', est)[0])
             ddf = pd.read_csv(est, index_col=0)

### 2023-07-26 simulations/py/tg.py

@@ -1,0 +1,8 @@
+from treesimulator.generator import generate
+from treesimulator.mtbd_models import BirthDeathModel, PNModel
+import logging
+
+logging.basicConfig(level=logging.INFO, format='%(asctime)s: %(message)s', datefmt="%Y-%m-%d %H:%M:%S")
+la, psi, p, psi_p, pn = (0.21666016107187575, 0.06376138425081969, 0.5614842560422901, 8.765449373319328, 0.2)
+model = PNModel(model=BirthDeathModel(p=p, la=la, psi=psi), pn=pn, removal_rate=psi_p)
+forest, (total_tips, u, T), _ = generate(model, 1000, 5000)

### 2023-07-26 simulations/py/tree_generator.py

@@ -1,3 +1,5 @@
+import logging
+
 import numpy as np
 from treesimulator import save_log, save_forest
 from treesimulator.generator import generate
@@ -24,13 +26,15 @@ if __name__ == "__main__":
     parser.add_argument('--max_tips', default=10000, type=int, help="Max number of tips")
     params = parser.parse_args()
 
+    logging.basicConfig(level=logging.INFO, format='%(asctime)s: %(message)s', datefmt="%Y-%m-%d %H:%M:%S",
+                        filename=params.nwk.replace('.nwk', '.txt'), filemode='w+')
+
     R0 = random_float(1, 5)
     psi = random_float(1 / 20, 1 / 5)
     la = psi * R0
-    psi_ratio = random_float(10, 100)
-    psi_n = psi * psi_ratio
-    rho = random_float(0.01, 0.9)
-    rho_n = random_float(0.1, 0.9)
+    psi_n = psi * random_float(50, 500)
+    rho = random_float(0.1, 0.9)
+    rho_n = random_float(0.01 / rho, 0.9)
 
     print(la, psi, psi_n, rho, rho_n)
 

### 2023-07-26 article/article.tex
Updated README.md

@@
 In BD-PN, at the moment of sampling the sampled individual might notify their most recent partner with a given probability $\rho_n$.
-Upon notification, partner is removed almost instantaneously (modeled via a notified removal rate $\psi_n >> \psi$).
+Upon notification, partner is removed almost instantaneously (modeled via a notified removal rate $\psi_p >> \psi$).
 BD-PN hence has 3 classical BD and 2 additional parameters:
@@
  \item $\rho_n$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
- \item $\psi_n >> \psi$ -- constant notified removal rate: a notified individual (if not yet removed via standard procedure) stops being infectious and gets observed with this rate.
+ \item $\psi_p >> \psi$ -- constant notified removal rate: a notified individual (if not yet removed via standard procedure) stops being infectious and gets observed with this rate.
 This creates an observed tip in the transmission tree.
@@
 %\\
-%q^{(p)}(i, r)  = p^{(r, o)}(t_j) e^{- \psi_n (t_i - t_r)} \textit{\color{gray} if $i$ is a \textbf{partner} tip notified by $r$ at time $t_r \leq t_i$}
+%q^{(p)}(i, r)  = p^{(r, o)}(t_j) e^{- \psi_p (t_i - t_r)} \textit{\color{gray} if $i$ is a \textbf{partner} tip notified by $r$ at time $t_r \leq t_i$}
 %\end{cases}
@@
 %\end{cases}\\
-%l^{(p)}(i, r) &= q^{(p)}(i,r)\psi_{n}\cdot \begin{cases}
+%l^{(p)}(i, r) &= q^{(p)}(i,r)\psi_p\cdot \begin{cases}
 %\rho_{n} \textit{\color{gray} ~~~~~~~~if $i$ notified their most recent patner}\\
@@
 p^{(i, nh)}(t_j)\psi\rho\rho_{n} \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and notified their most recent patner}\\
-p^{(r, nh)}(t_j) e^{- \psi_n (t_i - t_r)} \psi_n \rho_{n} \textit{\color{gray} ~if $i$ got notified at time $t_r \leq t_i$ and notified}
+p^{(r, nh)}(t_j) e^{- \psi_p (t_i - t_r)} \psi_p \rho_{n} \textit{\color{gray} ~if $i$ got notified at time $t_r \leq t_i$ and notified}
 \end{cases}\\
@@
 p^{(i, o)}(t_j)\psi\rho(1 - \rho_{n}) \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and did not notify}\\
-p^{(r, o)}(t_j) e^{- \psi_n (t_i - t_r)} \psi_n (1-\rho_{n}) \textit{\color{gray} ~if $i$ got notified at time $t_r \leq t_i$ and did not notify}
+p^{(r, o)}(t_j) e^{- \psi_p (t_i - t_r)} \psi_p (1-\rho_{n}) \textit{\color{gray} ~if $i$ got notified at time $t_r \leq t_i$ and did not notify}
 \end{cases}\\
@@
 
-The tree $\mathscr{T}$ likelihood density $\mathscr{L}(\mathscr{T}|\Theta)$ under BD-PN model with parameters $\Theta=\{\lambda, \psi, \rho, \psi_n, \rho_n\}$ can be calculated as $l^{(u)}(root)$, as the root can only correspond to an unnotified individual (since there was no observed transmission before it).
+The tree $\mathscr{T}$ likelihood density $\mathscr{L}(\mathscr{T}|\Theta)$ under BD-PN model with parameters $\Theta=\{\lambda, \psi, \rho, \psi_p, \rho_n\}$ can be calculated as $l^{(u)}(root)$, as the root can only correspond to an unnotified individual (since there was no observed transmission before it).
 
@@
 For the PN version of this simulator, we add an additional $m+1^{th}$ state for notified partners.
-$m + 1$'s state-change and transmission rates are set to zero: $\mu_{k,m+1} = \mu_{m+1,k} = \lambda_{k,m+1} = \lambda_{m+1,k} = 0\; \forall 1 \leq k \leq m$, while its removal rate is set to the notification removal rate: $\psi_{m+1} = \psi_n >> \psi_k\; \forall 1 \leq k \leq m$.
+$m + 1$'s state-change and transmission rates are set to zero: $\mu_{k,m+1} = \mu_{m+1,k} = \lambda_{k,m+1} = \lambda_{m+1,k} = 0\; \forall 1 \leq k \leq m$, while its removal rate is set to the notification removal rate: $\psi_{m+1} = \psi_p >> \psi_k\; \forall 1 \leq k \leq m$.
 We keep a mapping of most recent partners $P$ for each infectious individual, and update it for both the donor and the recipient at each transmission event.
@@
 removal rate $\psi \in ]1 / 20, 1 / 5]$,
-ratio between the partner removal rate after notification and the removal rate $\frac{\psi_n}{\psi} \in ]50, 500]$,
+ratio between the partner removal rate after notification and the removal rate $\frac{\psi_p}{\psi} \in ]50, 500]$,
 sampling probability $\rho \in ]0.1, 0.9]$,
@@
 We implemented a parameter estimator for the BD-PN model (which we called bdpn).
-It estimates the BD-PN model parameters  $\Theta = (\lambda,\psi,\psi_n,\rho,\rho_n) \in \mathbb{R}^5$ for a forest $\mathscr{F}$ comprising $f \geq 1$ observed trees in the maximum-likelihood framework, where one of the BD parameters ($\lambda,\psi$ or $\rho$) is fixed (for identifiability reasons).
+It estimates the BD-PN model parameters  $\Theta = (\lambda,\psi,\psi_p,\rho,\rho_n) \in \mathbb{R}^5$ for a forest $\mathscr{F}$ comprising $f \geq 1$ observed trees in the maximum-likelihood framework, where one of the BD parameters ($\lambda,\psi$ or $\rho$) is fixed (for identifiability reasons).
 
@@
 We calculated the relative error (normalized distance between the estimated and the target values: $\frac{|estimated - target|}{target}$) and the relative bias ($\frac{estimated - target}{target}$) for each parameter on each tree (Fig.~\ref{fig:sim}).
-Average relative errors were the lowest when $\psi$ was fixed: $\leq 5\%$  for all the parameters but $\psi_n$, which turned to be the hardest parameter to estimate in all the settings (relative errors $\approx 15\%$, with a tendency to underestimation).
+Average relative errors were the lowest when $\psi$ was fixed: $\leq 10\%$  for all the parameters.
 The worst performance was obtained with $\lambda$ being fixed: in this setting the estimator tends to underestimate $\psi$ and overestimate $\rho$.
-As with the point estimates, CI estimates were the best when $\psi$ was fixed: the target values of $\lambda, \psi_{n} \rho$ and $\rho_n$ were within the estimated CIs in correspondingly 97\%, 95\%, 87\%, and 76\% of cases.
-The values for the other settings are shown in Table~\ref{tbl:ci}.
+%The hardest parameter to estimate in all the settings (relative errors $\approx 15\%$, with a tendency to underestimation).
+
+As with the point estimates, CI estimates were the best when $\psi$ was fixed: the target values of $\lambda, \psi_p, \rho$ and $\rho_n$ were within the estimated CIs in correspondingly 82\%, 94\%, 51\%, and 67\% of cases.
+The CI estimates when $\lambda$ was fixed were the worst (see Table~\ref{tbl:ci}).
+The relative width of the CIs ($\frac{value_{97.5\%} - value_{2.5\%}}{value}$) varied between 11\% for $\lambda$ and 33\% for $\psi_p$ (in all the settings).
 
@@
 \includegraphics[width=1\textwidth]{Fig_errors.eps}
-\caption{Comparison of inference accuracy of bdpn on a data set of 100 trees of 5\,000--10\,000 tips each, with either $\lambda$ (blue), $\psi$ (orange) or $\rho$ (green) parameter being fixed to its real value.
+\caption{Comparison of inference accuracy of bdpn on a data set of 100 trees of 500--1\,000 tips each, with either $\lambda$ (blue), $\psi$ (orange) or $\rho$ (green) parameter being fixed to its real value.
 We show the swarmplots (coloured by fixed parameter) of relative errors for each test tree and parameter, which are measured as the normalized distance between the estimate and the real value.
@@
 \toprule 
- $\lambda$ & -- &  97\% & 96\% \\
- $\psi$ & 78\% & -- & 83\% \\
- $\rho$ & 28\% & 87\%  & -- \\
- $\psi_n$ & 94\% & 95\% & 95\% \\
- $\rho_n$ & 75\% & 76\% & 77\% \\
+ $\lambda$ & -- &  82\% & 92\% \\
+ $\psi$ & 34\% & -- & 78\% \\
+ $\rho$ & 20\% & 51\%  & -- \\
+ $\psi_p$ & 95\% & 94\% & 95\% \\
+ $\rho_n$ & 69\% & 67\% & 64\% \\
 \bottomrule
@@
 The tree represents the UK HIV-B epidemic between 1960s and 2016, its tips represent viruses sampled from 39\,130 individuals between 1996 and 2016.
-The samples used for the tree reconstruction were obtained from the UK HIV Drug resistance database~\citep{Dunn2007}~\citep{zhukovaModelingDrugResistance2023}.
+The samples used for the tree reconstruction were obtained from the UK HIV Drug resistance database~\citep{Dunn2007}.
 
@@
 The British HIV Association 2015 guidelines recommend all individuals with suspected or diagnosed primary HIV infection ``are offered immediate ART''~\citep{churchillBritishHIVAssociation2016}.
-To have a more homogeneous setting in terms of access to ART, awareness of the HIV infection, etc., we cut the dated tree between 2012 and 2015, and estimated the BD-PN parameters on the 2012-2015 forests.
+To have a more homogeneous setting in terms of access to ART, awareness of the HIV infection, etc., we cut the dated tree between 2012 and 2015, and estimated the BD-PN parameters on the 2012--2015 forests.
 
@@
 
- We estimated the following parameter values: $\lambda=0.262$, $\psi=0.186$, $\psi_n=186.921$, $\rho_n=0.033$, which corresponds to $R_e = 1.4$, infectious time $\frac{1}{\psi} = 5.4$ years, and about 2 days between partner notification and their sampling.
+Using $\rho=0.6$, we estimated the following parameter values: $\lambda=0.262$, $\psi=0.186$, $\psi_p=186.921$, $\rho_n=0.033$, which corresponds to $R_e = 1.4$, infectious time $\frac{1}{\psi} = 5.4$ years, and about 2 days between partner notification and their sampling.
+The results for $\rho=0.4$ and for $\rho=0.6$ were similar: $R_e$ of respectively $1.5$ and
+$1.3$, infectious time of $5.1$ and $5.6$ years, partner notification probability of $0.033$ in both cases and about 2 days between partner notification and their sampling.
 
@@
 
-To assess the robustness of these predictions with respect to $\rho$, we additionally estimated the parameter values for $\rho=0.4$ and for $\rho=0.6$.
-The results were very similar: $R_e$ of respectively $1.5$ and
-$1.3$, infectious time of $5.1$ and $5.6$ years, partner notification probability of $0.033$ in both cases and about 2 days between partner notification and their sampling.
-
 \end{document}

### 2023-07-26 simulations/py/assess_CIs.py

@@ -30,9 +30,9 @@ if __name__ == "__main__":
             for par in PARAMETERS:
                 df.loc[mask, '{}_within_CI'.format(par)] \
                     = (np.less_equal(df.loc[mask, '{}_min'.format(par)], real_df[par])
-                       | np.less_equal(np.abs(real_df[par] - df.loc[mask, '{}_min'.format(par)]), 1e-6)) \
+                       | np.less_equal(np.abs(real_df[par] - df.loc[mask, '{}_min'.format(par)]), 1e-3)) \
                       & (np.less_equal(real_df[par], df.loc[mask, '{}_max'.format(par)])
-                         | np.less_equal(np.abs(df.loc[mask, '{}_max'.format(par)] - real_df[par]), 1e-6))
+                         | np.less_equal(np.abs(df.loc[mask, '{}_max'.format(par)] - real_df[par]), 1e-3))
                 print('{}:\t{:.1f}% within CIs'
                       .format(par, 100 * sum(df.loc[mask, '{}_within_CI'.format(par)]) / n_observations))
                 f.write('{}:\t{:.1f}% within CIs\n'

### 2023-07-26 simulations/py/plot_error.py

@@ -92,7 +92,7 @@ if __name__ == "__main__":
                         pval_abs = 1
                     par2types2pval[par][(type_1, type_2)] = pval_abs
 
-        ERROR_COL = 'relative error' if 'p' not in pars else 'relative or absolute (for {}) error'.format(par2greek['p'])
+        ERROR_COL = 'relative error' if 'p' not in pars else 'relative or absolute (for {} and {}) error'.format(par2greek['p'], par2greek['pn'])
         plot_df = pd.DataFrame(data=data, columns=['parameter', ERROR_COL, 'config'])
 
         # if 'BEAST2' not in types:

### 2023-07-27 bdpn/formulas.py
Added comments to formulas

@@ -2,51 +2,102 @@ import numpy as np
 
 
 def get_c1(la, psi, rho):
+    """
+    Return c1 = ((la − psi)^2 + 4 * la * psi * rho)^1/2
+    """
     return np.power(np.power(la - psi, 2) + 4 * la * psi * rho, 0.5)
 
 
 def get_c2(la, psi, c1):
+    """
+    Return c2 = (c1 + la - psi) / (c1 - la + psi)
+    """
     la_minus_psi = la - psi
     return (c1 + la_minus_psi) / (c1 - la_minus_psi)
 
 
 def get_E1(c1, c2, t, T):
+    """
+    Returns E1 = c2 * exp(c1 * (t - T))
+    """
     return c2 * np.exp(c1 * (t - T))
 
 
 def get_E2(c1, c2, ti, T):
+    """
+    Returns E2 = c2 * exp(c1 * (ti - T))
+    """
     return c2 * np.exp(c1 * (ti - T))
 
 
 def get_E3(c1, t, ti):
+    """
+    Returns E3 = exp(c1 * (t - ti))
+    """
     return np.exp(c1 * (t - ti))
 
 
-def get_E4(la, psi, t, ti):
-    return np.exp((la + psi) * (t - ti))
-
-
 def get_u(la, psi, c1, E1):
+    """
+    Returns U(t), where
+    dU(t)/dt = la + psi * U(t) − la U^2(t) − psi * (1 − rho)
+    U(T) = 1
+
+    :param c1: c1 = ((la − psi)^2 + 4 * la * psi * rho)^1/2
+    :param E1: E1 = c2 * exp(c1 * (t - T)), where c2 = (c1 + la - psi) / (c1 - la + psi)
+    """
     two_la = 2 * la
     return (la + psi) / two_la + c1 / two_la * ((E1 - 1) / (E1 + 1))
 
 
 def get_log_p(c1, t, ti, E1, E2):
-    # return np.power((E2 + 1) / (E1 + 1), 2) * E3
-    # return 2 * (np.log(E2 + 1) - np.log(E1 + 1)) + np.log(E3)
+    """
+    Returns log(p(t)), where
+    dp(t)/dt = la + psi * p(t) − 2 * la * p(t) * U(t)
+    p(ti) = 1
+    dU(t)/dt = la + psi * U(t) − la U^2(t) − psi * (1 − rho)
+    U(T) = 1.
+
[…]

### 2023-07-27 bdpn/bdpn.py
Checked the formulas in BDPN, added comments and removed unused code

@@ -3,38 +3,11 @@ from multiprocessing.pool import ThreadPool
 
 import numpy as np
 from pastml.tree import read_forest
-from scipy import integrate
 
 from bdpn.parameter_estimator import optimize_likelihood_params, rescale_log
 from bdpn.tree_manager import TIME, annotate_tree
 from bdpn.formulas import get_log_p, get_c1, get_c2, get_E2, get_E1, get_log_po, get_log_pnh, get_log_po_from_p_pnh, \
     get_u
-
-
-# def get_log_hidden_partner_notification(t, la, psi, rho, psi_n, T, ti, c1=None, c3=None, c4=None):
-#     # return np.log(
-#     #     integrate.quad(lambda tau: np.exp(get_log_p(t, la, psi, rho, T, tau, c1=c1) + np.log(2 * la)
-#     #                                       + get_log_p_nh(tau, la, psi, ti)
-#     #                                       + get_log_p_o(tau, la, psi, rho, T, ti, c1=c1, c3=c3)
-#     #                                       + (psi_n * (ti - T))), t, ti)[0])
-#     # we inline everything we can in the above expression
-#
-#     if c4 is None:
-#         c4 = (c1 + la - psi) / (c1 - la + psi)
-#     la_plus_psi = la + psi
-#     c5 = (3 * la_plus_psi + c1) / 2
-#     non_tau_part = c1 * t - 2 * np.log(c4 * np.exp(c1 * (t - T)) + 1) \
-#                    - c5 * ti \
-#                    + np.log(c3) \
-#                    + psi_n * (ti - T)
-#
-#     def get_v(tau, non_tau_part):
-#         c3_p = get_c3(la, psi, rho, T, tau, c1)
-#         tau_part = 2 * np.log(c3_p) + (c5 - c1) * tau \
-#                    - np.log(c4 * np.exp(c1 * (tau - T)) + 1)
-#         return np.exp(non_tau_part + tau_part)
-#
-#     return np.log(integrate.quad(lambda tau: get_v(tau, non_tau_part), t, ti, limit=20)[0])
 
 
 def loglikelihood(forest, la, psi, psi_n, rho, rho_n, T=None, threads=1):
@@ -48,20 +21,6 @@ def loglikelihood(forest, la, psi, psi_n, rho, rho_n, T=None, threads=1):
 
     c1 = get_c1(la=la, psi=psi, rho=rho)
[…]

### 2023-07-27 article/article.tex
Added references to the article

@@
 The interaction of epidemiological and evolutionary processes leaves a footprint in pathogen genomes.
-Phylodynamics leverages this footprint to estimate epidemiological parameters~\cite{Grenfell2004a,Volz2013}.
+Phylodynamics leverages this footprint to estimate epidemiological parameters~\citep{Grenfell2004a,Volz2013}.
 It relies on models that bridge the gap between traditional epidemiology and sequence data by estimating such parameters as the basic reproduction number, $R_0$, from topology and branch lengths of pathogen phylogenies (i.e.
@@
 
-Phylodynamic models can be classified into two main families:  coalescent~\cite{Volz2009a,Drummond2005,Pybus2000a} and birth-death (BD)~\cite{Kendall1948,Maddison2007,Stadler2009,Stadler2010}.
-Coalescent models are often preferred for estimating deterministic population dynamics, however for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\cite{Macpherson2021}.
-In the classic BD model with incomplete sampling~\cite{Stadler2009}, births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g.
+Phylodynamic models can be classified into two main families:  coalescent~\citep{Volz2009a,Drummond2005,Pybus2000a} and birth-death (BD)~\citep{Kendall1948,Maddison2007,Stadler2009,Stadler2010}.
+Coalescent models are often preferred for estimating deterministic population dynamics, however for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\citep{Macpherson2021}.
+In the classic BD model with incomplete sampling~\citep{Stadler2009}, births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g.
 due to healing, self-isolation, starting a treatment, or death, modelled with a constant removal rate).
@@
 
-To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extention of the classical BDS model was developed by Stadler \textit{et al.}~\cite{Stadler2013a}.
+To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extention of the classical BDS model was developed by~\citet{Stadler2013a}.
 MTBD framework allows for different types of individual states (and changes between them, modelled with state change rates).
@@
 Examples of the MTBD family models include the birth-death exposed-infectious (BDEI) model, which was designed for pathogens featuring an incubation period between the moments of infection and of becoming infectious (e.g.
-Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen more than the others~\cite{Stadler2014}.
+Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen more than the others~\citep{Stadler2014}.
 
-To allow for heterogeneity over time, Stadler~\textit{et~al.}~\cite{Stadler2013} developed one-state Bayesian birth-death skyline plot (BDSKY) that divides the time into intervals and allows for different piecewise constant rates on them.
-K\"{u}hnert \textit{et al.}~\cite{Kuhnert2016} combined the MTBD model with the BDSKY to allow for both piecewise-constant rate changes over time and multiple individual types.
+To allow for heterogeneity over time, \citet{Stadler2013} developed one-state Bayesian birth-death skyline plot (BDSKY) that divides the time into intervals and allows for different piecewise constant rates on them.
+\citet{Kuhnert2016} combined the MTBD model with the BDSKY to allow for both piecewise-constant rate changes over time and multiple individual types.
 In particular the skyline approach can be useful to account for changes in the sampling policies (e.g.
@@
 With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch that finishes at time $t_i$, starting on this branch at a time $t \leq t_i$.
-The master equations for the BD models were initially developed by ~\citet{Stadler2009}, however what we present here, is their branch-specific formulation that we proposed in~\cite{zhukovaFastAccurateMaximumLikelihood2022}:
+The master equations for the BD models were initially developed by ~\citet{Stadler2009}, however what we present here, is their branch-specific formulation that we proposed in~\citep{zhukovaFastAccurateMaximumLikelihood2022}:
 
@@
 
-These assumptions make the mathematics behind the model manageable, however they can be easily relaxed in a tree simulator (e.g., to be used with a deep learning estimator~\cite{Voznica2021}, which does not require likelihood calculation).
+These assumptions make the mathematics behind the model manageable, however they can be easily relaxed in a tree simulator (e.g., to be used with a deep learning estimator~\citep{Voznica2021}, which does not require likelihood calculation).
 
@@
 
-Using the formulas~(\ref{eq:ps}), we can calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}, where for each visited node $i$ we calculate the likelihood density of its subtree if the branch connecting it to its parent node $j$ corresponds to (1) an unnotified individual $l^{(u)}(i)$; or (2) an (eventually) notified partner $l^{(p)}(i, r)$ (notified by the individual corresponding to the tip $r$ at time $t_r$).
+Using the formulas~(\ref{eq:ps}), we can calculate the likelihood density of a tree using a pruning algorithm~\citep{10.1093/sysbio/22.3.240}, where for each visited node $i$ we calculate the likelihood density of its subtree if the branch connecting it to its parent node $j$ corresponds to (1) an unnotified individual $l^{(u)}(i)$; or (2) an (eventually) notified partner $l^{(p)}(i, r)$ (notified by the individual corresponding to the tip $r$ at time $t_r$).
 For each visited tip we additionally consider whether the corresponding individual notified their most recent partner or not.
@@
 \subsection*{Code availability}
-The simulator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\cite{Huerta-Cepas2016} and NumPy package for array operations~\cite{harris_array_2020}.
+The simulator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\citep{Huerta-Cepas2016} and NumPy package for array operations~\citep{harris_array_2020}.
 
@@
 \subsection*{Code availability}
-The BD-PN parameter estimator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\cite{Huerta-Cepas2016} and NumPy package for array operations~\cite{harris_array_2020}.
+The BD-PN parameter estimator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\citep{Huerta-Cepas2016} and NumPy package for array operations~\citep{harris_array_2020}.
 
@@
 
+\bibliographystyle{plainnat}
+\bibliography{refs}
+
 \end{document}

### 2023-07-27 simulations/py/tg.py

@@ -1,8 +1,0 @@
-from treesimulator.generator import generate
-from treesimulator.mtbd_models import BirthDeathModel, PNModel
-import logging
-
-logging.basicConfig(level=logging.INFO, format='%(asctime)s: %(message)s', datefmt="%Y-%m-%d %H:%M:%S")
-la, psi, p, psi_p, pn = (0.21666016107187575, 0.06376138425081969, 0.5614842560422901, 8.765449373319328, 0.2)
-model = PNModel(model=BirthDeathModel(p=p, la=la, psi=psi), pn=pn, removal_rate=psi_p)
-forest, (total_tips, u, T), _ = generate(model, 1000, 5000)

### 2023-07-28 article/article.tex
Completed the article

@@
 We now propose a non-Markovian extension of the BD model that adds partner notification (PN).
-In BD-PN, at the moment of sampling the sampled individual might notify their most recent partner with a given probability $\rho_n$.
-Upon notification, partner is removed almost instantaneously (modeled via a notified removal rate $\psi_p >> \psi$).
+In BD-PN, at the moment of sampling the sampled individual might notify their most recent partner with a constant probability $\rho_n$.
+Upon notification, the partner is removed almost instantaneously (modeled via a constant notified removal rate $\psi_p >> \psi$).
 BD-PN hence has 3 classical BD and 2 additional parameters:
@@
 Note that the most recent partner relationship is not necessarily symmetrical, as the most recent partner of $i$ might have transmitted their virus further after the contact with $i$.
+
+In Appendix we discuss how the -PN extension can be generalized to multi-type birth-death models (MTBD-PN).
 
@@
 
-\section{Tree simulator}
-We implemented a quick tree simulator that generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states).
+\section{PN test}~\label{sec:test}
+To access whether a give tree is generated under a classical MTBD model or MTBD-PN, we developed a non-parametric test based on cherries (two tips having a common parent). 
+The intuition behind the test is that in the presence of partner notification the tree will contain more cherries whose tips are close in time.
+
+The test sorts the cherries by the times of their roots, and splits them into blocks of 100 cherries (adjustable parameter).
+For each cherry in a block, the test calculates the difference between its tip times, hence obtaining an array of 100 cherry tip differences.
+It then generates a collection of random cherry tip differences for the same block: For each original cherry root it picks 5 (adjustable parameter) cherries with the roots that are the closest in time, randomly selects two tips among their tips, and calculates their time difference.
+An array of 100 reshuffled cherry tip differences is thus obtained for the same block.
+The reshuffled cherry tip difference array generation is repeated 1\,000 times (adjustable parameter).
+Finally, the test reports the proportion of reshuffled arrays whose first quantile (adjustable parameter) is smaller than that of the real cherry array.
+
+The test therefore reports a probability of partner notification being present at time interval corresponding to each block.
+To estimate the probability of partner notification being present in the whole tree, it averages the block probabilities.
+
+\subsection*{Code availability}
+We implemented the PN-test in Python 3. It is available as a part of BDPN parameter estimator (see section~\ref{sec:sim}).
+
+\subsection{Performance on simulated data}~\label{sec:test:sim}
+To illustrate the PN test performance, we applied it to the trees simulated under BD and under BD-PN models.
+We implemented a MTBD/MTBD-PN tree simulator (see Appendix for details) and used it to generate 100 trees with 500--1\,000 tips under the BD-PN model.
+For each tree the parameter values were drawn randomly within the following bounds:
+$R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
+removal rate $\psi \in ]1 / 20, 1 / 5]$,
+ratio between the partner removal rate after notification and the removal rate $\frac{\psi_p}{\psi} \in ]50, 500]$,
+sampling probability $\rho \in ]0.1, 0.9]$,
+partner notification probability $\rho_n \in ]0.01/\rho, 0.9]$.
+We also generated 100 trees of 500--1\,000 tips under the BD model.
+For each tree in the BD data set the parameter values ($\lambda, \psi,$ and $\rho$) were the same as the ones selected for the corresponding tree of the BD-PN dataset.
+
+We then run the PN tests on each tree of the both datasets.
+For 98 out of 100 trees in the BD dataset the PN test value was above $0.05$ (i.e.
+98\% true negative and 2\% false positive results), the mean PN test value was $0.468$.
+For 96 (out of 100) trees in the BD-PN dataset the PN test value was below $0.05$ (i.e.
+96\% true positive and 4\% false negative results), the mean PN test value was $0.008$.
+Our test therefore showed both high specificity and sensitivity.
+
+\section{BD-PN parameter and CI estimator}\label{sec:sim}
+We implemented a parameter estimator for the BD-PN model (which we called bdpn).
+It estimates the BD-PN model parameters  $\Theta = (\lambda,\psi,\psi_p,\rho,\rho_n) \in \mathbb{R}^5$ for a forest $\mathscr{F}$ comprising $f \geq 1$ observed trees in the maximum-likelihood framework, where one of the BD parameters ($\lambda,\psi$ or $\rho$) is fixed (for identifiability reasons).
+
+Once the optimal parameter values are found, we calculate their confidence intervals (CIs) using Wilks' method~\citep{Wilks1938}.
+For each non-fixed parameter $p \in \Theta$, we calculate its $95\%$-CI as including the values $\tilde{p}$ such that $log L(\mathscr{F}|\Theta_{opt|p=\tilde{p}}) > log L(\mathscr{F}| \Theta_{opt}) - \chi^2_1(0.95) / 2$, where $\chi^2_1(0.95)$ is the value of chi-squared distribution with 1 degree of freedom corresponding to the significance level of $0.95$ (i.e., $\sim3.84$).
+$\Theta_{opt|p=\tilde{p}}$ corresponds to the maximum-likelihood value for the other non-fixed parameters when $p = \tilde{p}$.
+
+\subsection*{Code availability}
+The BD-PN parameter estimator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\citep{Huerta-Cepas2016} and NumPy package for array operations~\citep{harris_array_2020}.
+
+It is available as a command-line program and a Python 3 package via PyPi (\href{https://pypi.org/project/bdpn}{bdpn}), and via Docker/Singularity (\href{https://hub.docker.com/r/evolbioinfo/bdpn/tags}{evolbioinfo/bdpn}).
+Its source code and the installation and usage documentation are available on GitHub at \href{https://github.com/evolbioinfo/bdpn}{github.com/evolbioinfo/bdpn}.
+
+\subsection{Performance on simulated data}
+
+To assess the performance of our maximum-likelihood estimator for the BD-PN model, we used the 100 trees with 500--1\,000 tips simulated under the BD-PN model as described in subsection~\ref{sec:test:sim}.
+
+We applied bdpn to each of the trees three times: fixing each of the BD parameters ($\lambda,\psi$ or $\rho$) to its real value (for identifiability).
+
+As expected with a maximum likelihood estimator, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the real parameter values in all the settings.
+
+We calculated the relative error (normalized distance between the estimated and the target values: $\frac{|estimated - target|}{target}$) and the relative bias ($\frac{estimated - target}{target}$) for each parameter on each tree (Fig.~\ref{fig:sim}).
+Average relative errors were the lowest when $\psi$ or $\rho$ were fixed: $\leq 10\%$  for all the parameters.
+The worst performance was obtained with $\lambda$ being fixed: in this setting the estimator tends to underestimate $\psi$ and overestimate $\rho$.
+%The hardest parameter to estimate in all the settings (relative errors $\approx 15\%$, with a tendency to underestimation).
+
+As with the point estimates, CI estimates were the best when $\psi$ or $\rho$ were fixed, and the worst when $\lambda$ was fixed (see Table~\ref{tbl:ci}).
+The relative width of the CIs ($\frac{value_{97.5\%} - value_{2.5\%}}{value}$) varied between 11\% for $\lambda$ estimates and 32\% for $\psi_p$ estimates (in all the settings).
+
+\begin{figure}[!pht]
+\centering 
+\includegraphics[width=1\textwidth]{Fig_errors.eps}
+\caption{Comparison of inference accuracy of bdpn on a data set of 100 trees of 500--1\,000 tips each, with either $\lambda$ (blue), $\psi$ (orange) or $\rho$ (green) parameter being fixed to its real value.
+We show the swarmplots (coloured by fixed parameter) of relative errors for each test tree and parameter, which are measured as the normalized distance between the estimate and the real value.
+Average relative error (and in parentheses average relative bias) are displayed for each parameter and method below their swarmplot.
+For $\rho$ and $\rho_n$ absolute values are shown instead of normalized ones.
+}
+\label{fig:sim} 
+\end{figure}
+
+ \begin{table}[!h]\centering
+\small
+\caption{Percentage of simulated trees for which the real parameter values were withing the estimated 95\% CIs. \smallskip}
+\begin{tabular}{c|ccc}
+\textbf{parameter} & \textbf{$\lambda$ fixed} & \textbf{$\psi$ fixed} & \textbf{$\rho$ fixed} \\
+\toprule 
+ $\lambda$ & -- &  76\% & 96\% \\
+ $\psi$ & 22\% & -- & 78\% \\
+ $\rho$ & 16\% & 55\%  & -- \\
+ $\psi_p$ & 70\% & 93\% & 94\% \\
+ $\rho_n$ & 21\% & 68\% & 63\% \\
+\bottomrule
+\end{tabular}
+\label{tbl:ci}
+\end{table}
+
+\subsection{Application: HIV-B epidemic in the UK}
+We applied our estimator to asses the partner notification in HIV-infected patients in the UK.
+We used the phylogenetic tree from our recent study of HIV drug resistance in the UK~\citep{zhukovaModelingDrugResistance2023}.
+The tree tips represent viruses sampled from 40\,055 individuals between 1996 and 2016.
+The samples used for the tree reconstruction were obtained from the UK HIV Drug resistance database~\citep{Dunn2007}.
+The sampling date metadata for these samples were month-specific (e.g.
+Aug 2015), hence potentially creating simultaneous sampling dates for cases, which were sampled on different days of the month.
+To avoid this issue we randomly selected a date within the specified month for each sample, and time-scaled the tree with LSD2~\citep{To2016} (v2.4.1, strict molecular clock with outlier removal).
+We repeated this time-scaling procedure 10 times, obtaining 10 trees with slightly different branch lengths and root dates between Oct 5 and Nov 11, 1962.
+
+Between 2012 and 2015, in the UK antiretroviral treatment (ART) of HIV-infected individuals  was initiated when their CD4 count dropped below 350 cells/mL~\citep{williamsBritishHIVAssociation2012}.
+However ``if a patient with a CD4 cell count > 350 cells/mL wishes to start ART to reduce the risk of transmission to partners, this decision is respected and ART is started''~\citep{williamsBritishHIVAssociation2012}.
+Before 2012, treatment started with an even lower CD4 (i.e.
+later).
+The British HIV Association 2015 guidelines recommend all individuals with suspected or diagnosed primary HIV infection ``are offered immediate ART''~\citep{churchillBritishHIVAssociation2016}.
+To have a more homogeneous setting in terms of access to ART, awareness of the HIV infection, etc., we cut the dated trees between 2012 and 2015.
+We then applied the PN-test and estimated the BD-PN parameters on these 2012--2015 forests.
+
+The PN-test values were below 0.05 for all the forests (value between $0.01$ and $0.04$), suggesting the presence of partner notification.
+
+To estimate the BD-PN parameter values, we needed to fix one of the parameters for model identifiability.
+We therefore estimated the base sampling probability $\rho$ as follows.
+According to the \href{https://webarchive.nationalarchives.gov.uk/ukgwa/20181112132123mp_/https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/602942/HIV_in_the_UK_report.pdf}{``Towards elimination of HIV transmission, AIDS and HIV-related deaths in the UK''}
+ report, the estimated total number of people living with HIV in the UK in 2015 was 101\,200 (97\,500 to 105\,700).
+% in 2012 this number was 98\,400 (93\,500-104\,300) [\href{HIV in the United Kingdom: 2013 Report}{https://webarchive.nationalarchives.gov.uk/ukgwa/20181112133715mp_/https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/326601/HIV_annual_report_2013.pdf}].
+Moreover these document report about 400 HIV-related deaths per year.
+According to our estimate~\citep{zhukovaModelingDrugResistance2023}, 66.5\% of the HIV-infected individuals in the UK were infected with HIV-B.
+We hence estimated that in 2015 67\,298 (64\,837, 70\,291) people were living with HIV-B.
+Our trees contained 39\,047 tips sampled by the end of 2015, and therefore account for about 58\% (56\%, 60\%) of the total number of individuals living with HIV-B in the UK.
+We hence fixed the sampling probability $\rho$ to 0.58.
+%However this estimate does not take into account deaths of the HIV-infected individuals over the years, which would increase the total number of HIV-infected individuals by 2015, and decrease the proportion of individuals represented by our tree.
+We therefore tested several sampling probabilities $\rho=$0.6; 0.5; 0.4.
+
+Using $\rho=0.58$, we estimated the following parameter values: $\lambda=0.53$, $\psi=0.43$, $\psi_p=23--59$, $\rho_n=0.006--0.007$, which corresponds to $R_e = 1.2$, infectious time $\frac{1}{\psi} = 2.3$ years, and about 6--16 days between partner notification and their sampling.
+The results are consistent between 10 forests and shown in Table~\ref{tbl:uk}.
+%The results for $\rho=0.4$ and for $\rho=0.6$ were similar: $R_e$ of respectively $1.5$ and
+%$1.3$, infectious time of $5.1$ and $5.6$ years, partner notification probability of $0.033$ in both cases and about 2 days between partner notification and their sampling.
+
+The average time of HIV progression to AIDs in the absence of treatment is 10 years, which could serve as an estimate of the infectious time.
+However, in the UK, 95\% of infected population is on antiretroviral treatment, which, when taken properly, prevents transmission.
+Hence, the average infectious time represents the average time before starting treatment.
+Our estimate is approximately 5 years.
+
+\begin{table}
+\caption{BD-PN model parameters estimated for UK HIV-1 B epidemic between 2012 and 2015, assuming that our data represents 58\% of infected cases ($\rho=0.58$).}
+\label{tbl:uk}
+\begin{tabular}{c|cc|cccc}
+&&&&infectious&partner&\\
+fo-&num.&num.&&time&removal t.&\\
+rest&tips&trees&$R_e$&$\frac{1}{\psi}$ [years]& $\frac{1}{\psi_p}$ [days]&$\rho_n$\\
+\toprule
+ $1$ & $11\,117$ & $8\,838$& $1.24\;[1.22-1.27]$& $2.34\;[2.29-2.39]$& $8\;[5-12]$& $0.007\;[0.004-0.010]$ \\
+ $2$ & $11\,109$ & $8\,833$& $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.39]$& $6\;[4-10]$& $0.006\;[0.003-0.009]$ \\
+ $3$ & $11\,122$ & $8\,844$& $1.24\;[1.22-1.27]$& $2.34\;[2.28-2.39]$& $16\;[3-18]$& $0.006\;[0.003-0.010]$ \\
+ $4$ & $11\,119$ & $8\,838$& $1.24\;[1.22-1.27]$& $2.34\;[2.28-2.39]$& $7\;[5-11]$& $0.007\;[0.004-0.010]$ \\
+ $5$ & $11\,116$ & $8\,829$& $1.24\;[1.22-1.27]$& $2.33\;[2.28-2.39]$& $9\;[6-13]$& $0.006\;[0.003-0.009]$ \\
+ $6$ & $11\,111$ & $8\,827$& $1.24\;[1.22-1.27]$& $2.34\;[2.28-2.39]$& $8\;[5-12]$& $0.006\;[0.003-0.009]$ \\
+ $7$ & $11\,119$ & $8\,838$& $1.24\;[1.22-1.27]$& $2.33\;[2.28-2.39]$& $7\;[5-11]$& $0.006\;[0.003-0.009]$ \\
+ $8$ & $11111$ & $8\,830$& $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.39]$& $6\;[4-11]$& $0.006\;[0.003-0.008]$ \\
+ $9$ & $11\,120$ & $8\,833$& $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.39]$& $7\;[4-12]$& $0.006\;[0.003-0.009]$ \\
+ $10$ & $11\,112$ & $8\,827$& $1.25\;[1.22-1.27]$& $2.34\;[2.29-2.39]$& $7\;[5-11]$& $0.006\;[0.004-0.009]$ \\
+ \bottomrule
+ \end{tabular}
+ \end{table}
+
+\section{Discussion}
+\label{disc}
+
+We proposed an extension of the phylodynamic birth-death model that accounts for non-random sampling due to partner notification (BD-PN).
+In real-life epidemics, detected infected individuals might notify the people whom they think they might have infected or got infected by, and who in turn might get tested and detected in a much faster and non-random way.
+Our model accounts for this scenario.
+
+The -PN extension adds two constant parameters to the classical BD model: the probability of partner notification $p_n$ and the partner detection rate $\psi_p$.
+It can also be applied to a more general, multi-type birth-death (MTBD) model framework, which accounts for heterogeneity in the infected population (see Appendix).
+
+We developed a mathematical framework for calculation of the likelihood density of a transmission tree (approximated by a time-scaled phylogenetic tree) under the BD-PN model, and implemented a maximum-likelihood BD-PN parameter estimator based on this framework.
+The advantage of the BD-PN model is that the differential equations describing it have closed-form solutions, which are quick to calculate and do not require numerical approximation (as is the case for more complex MTBD models).
+Tests on simulated data show that our estimator robustly infers model parameter values and their CIs in the setting when the duration of the infectious period $\frac{1}{psi}$ or the base sampling probability $\rho$ if fixed.
+Fixing one of the model parameters is needed for identifiability.
+In practice, $\rho$ and $\frac{1}{psi}$ may often be approximated from epidemiological data, e.g., as the proportion of sampled cases among the declared ones (for $\rho$), or estimated from observations of infected cases ($\frac{1}{\psi}$).
+
+We also developed a PN-test allowing to distinguish between trees generated under models with and without partner notification.
+Its application to simulated data showed high sensitivity and specificity.
+
+We applied the PN-test our estimator to the HIV-1 B epidemic in the UK, between 2012 and 2015.
+The test detected the presence of partner notification.
+The obtained estimates of epidemiological parameters are in agreement with what we now know about this epidemic (e.g., $R_e \approx 1.2$ ).
+We estimated a modest partner notification probability $\rho_p \in [0.003-0.1]$, and rapid notified partner detection ($3-18$ days vs $\approx2.3$ years for detection under the standard random procedure).
+
+The BD-PN model is the first steps towards accounting for non-random sampling in phylodynamic models.
+In order to reduce its mathematical complexity (having closed-form solutions for its differential equations) and make parameter estimation fast, we made several assumptions: (i) that only detected individuals can notify their partner; (ii) only the most recent partner can get notified; (iii) after the notification, notified partners do not transmit further; and (iv) notified partners are always sampled upon removal (i.e., observed in the transmission tree).
+These assumptions may be too simple to fully grasp the real partner notification process and should be relaxed in the future work.
+Another needed extension should account for population heterogeneity (i.e., MTBD-PN).
+As it is much easier to implement a tree simulator with a complex model than to derive its likelihood, a promising path to take for future -PN model developments would be using deep learning (e.g., adding new models to PhyloDeep~\citep{Voznica2021}).
+
+\bibliographystyle{plainnat}
+\bibliography{refs}
+
+\section{Appendix}
+\subsection{MTBD-PN models}
+MTBD models~\citep{Stadler2013a} add population structure to the BD model by allowing different individual states, transmissions between them and state changes.
+A general MTBD model with $m$ individual states has $2m^2 + m$ parameters: An individual in state $k \in \{1, \ldots, m\}$ can be removed at a constant average rate $\psi_k$ (with pathogen sampling probability $\rho_k$), change their state to state $l \in \{1, \ldots, m\}$ at a constant average rate $\mu_{kl}$ (where $\mu_{kk} = 0$), and transmit their pathogen to an individual in state $l$ at a constant average rate $\lambda_{kl}$.
+The time between events of the same type is hence modeled with exponential distribution.
+
+The -PN extension can be easily generalized from BD to MTBD models: at the moment of sampling an individual of any type can notify their last partner with the probability $\rho_n$, who will then stop transmitting and get removed at a constant removal rate $\psi_n >> \psi_k \; \forall k \in \{1, \ldots, m\}$.
+The -PN extension therefore adds two parameters to an MTBD model: $\psi_p$ and $\rho_n$.
+
+\subsection{MTBD-PN tree simulator}
+
+We implemented a tree simulator that generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states).
 The simulator is Gillespie-based, generates state change, transmission and removal times, and only reconstructs the sampled parts of the tree to save memory and increase speed.
@@
 
-\subsection*{Code availability}
+\subsubsection*{Code availability}
 The simulator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\citep{Huerta-Cepas2016} and NumPy package for array operations~\citep{harris_array_2020}.
@@
 
-\section{PN test}
-To access whether a give tree is generated under a classical MTBD model or MTBD-PN, we developed a non-parametric test based on cherries (two tips having a common parent). 
-The intuition behind the test is that in the presence of partner notification the tree will contain more cherries whose tips are close in time.
-
-The test sorts the cherries by the times of their roots, and splits them into blocks of 100 cherries (adjustable parameter).
-For each cherry in a block, the test calculates the difference between its tip times, hence obtaining an array of 100 cherry tip differences.
-It then generates a collection of random cherry tip differences for the same block: For each original cherry root it picks 5 more cherries with the roots that are the closest in time, randomly selects two tips among their tips, and calculates their time difference.
-An array of 100 reshuffled cherry tip differences is thus obtained fro the same block.
-The reshuffled cherry tip difference array generation is repeated 100 times (adjustable parameter).
-Finally, the test reports the proportion of reshuffled arrays whose first quantile is smaller than that of the real cherry array.
-
-The test therefore reports a probability of partner notification being present at time interval corresponding to each block.
-To estimate the probability of partner notification being present in the whole tree, it averages the block probabilities.
-
-\subsection{Performance on simulated data}
-To illustrate the PN test performance, we applied it to the trees simulated under BD and under BD-PN models.
-We simulated 100 trees with 500--1\,000 tips under the BD-PN model with our tree simulator.
-For each tree the parameter values were drawn randomly within the following bounds:
-$R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
-removal rate $\psi \in ]1 / 20, 1 / 5]$,
-ratio between the partner removal rate after notification and the removal rate $\frac{\psi_p}{\psi} \in ]50, 500]$,
-sampling probability $\rho \in ]0.1, 0.9]$,
-partner notification probability $\rho_n \in ]0.01/\rho, 0.9]$.
-We also simulated 100 trees of 500--1\,000 tips under the BD model.
-For each tree in the BD data set the parameter values ($\lambda, \psi,$ and $\rho$) were the same as the ones selected for the corresponding tree of the BD-PN dataset.
-
-We then run the PN tests on each tree of the both datasets.
-For 98 out of 100 trees in the BD dataset the PN test value was above $0.05$ (i.e.
-98\% true negative and 2\% false positive results), the mean PN test value was $0.468$.
-For 96 (out of 100) trees in the BD-PN dataset the PN test value was below $0.05$ (i.e.
-96\% true positive and 4\% false negative results), the mean PN test value was $0.008$.
-Our test therefore showed both high specificity and sensitivity.
-
-\section{BD-PN parameter and CI estimator}
-We implemented a parameter estimator for the BD-PN model (which we called bdpn).
-It estimates the BD-PN model parameters  $\Theta = (\lambda,\psi,\psi_p,\rho,\rho_n) \in \mathbb{R}^5$ for a forest $\mathscr{F}$ comprising $f \geq 1$ observed trees in the maximum-likelihood framework, where one of the BD parameters ($\lambda,\psi$ or $\rho$) is fixed (for identifiability reasons).
-
-Once the optimal parameter values are found, we calculate their confidence intervals (CIs) using Wilks' method~\citep{Wilks1938}.
-For each non-fixed parameter $p \in \Theta$, we calculate its $95\%$-CI as including the values $\tilde{p}$ such that $log L(\mathscr{F}|\Theta_{opt|p=\tilde{p}}) > log L(\mathscr{F}| \Theta_{opt}) - \chi^2_1(0.95) / 2$, where $\chi^2_1(0.95)$ is the value of chi-squared distribution with 1 degree of freedom corresponding to the significance level of $0.95$ (i.e., $\sim3.84$).
-$\Theta_{opt|p=\tilde{p}}$ corresponds to the maximum-likelihood value for the other non-fixed parameters when $p = \tilde{p}$.
-
-\subsection*{Code availability}
-The BD-PN parameter estimator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\citep{Huerta-Cepas2016} and NumPy package for array operations~\citep{harris_array_2020}.
-
-It is available as a command-line program and a Python 3 package via PyPi (\href{https://pypi.org/project/bdpn}{bdpn}), and via Docker/Singularity (\href{https://hub.docker.com/r/evolbioinfo/bdpn/tags}{evolbioinfo/bdpn}).
-Its source code and the installation and usage documentation are available on GitHub at \href{https://github.com/evolbioinfo/bdpn}{github.com/evolbioinfo/bdpn}.
-
-\subsection{Performance on simulated data}
-
-To assess the performance of our maximum-likelihood estimator for the BD-PN model, we used the 100 trees with 500--1\,000 tips simulated under the BD-PN model as described in the ``PN test'' section.
-
-We applied bdpn to each of the trees three times: fixing each of the BD parameters ($\lambda,\psi$ or $\rho$) to its real value (for identifiability).
-
-As expected with a maximum likelihood estimator, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the real parameter values in all the settings.
-
-We calculated the relative error (normalized distance between the estimated and the target values: $\frac{|estimated - target|}{target}$) and the relative bias ($\frac{estimated - target}{target}$) for each parameter on each tree (Fig.~\ref{fig:sim}).
-Average relative errors were the lowest when $\psi$ was fixed: $\leq 10\%$  for all the parameters.
-The worst performance was obtained with $\lambda$ being fixed: in this setting the estimator tends to underestimate $\psi$ and overestimate $\rho$.
-%The hardest parameter to estimate in all the settings (relative errors $\approx 15\%$, with a tendency to underestimation).
-
-As with the point estimates, CI estimates were the best when $\psi$ was fixed: the target values of $\lambda, \psi_p, \rho$ and $\rho_n$ were within the estimated CIs in correspondingly 82\%, 94\%, 51\%, and 67\% of cases.
-The CI estimates when $\lambda$ was fixed were the worst (see Table~\ref{tbl:ci}).
-The relative width of the CIs ($\frac{value_{97.5\%} - value_{2.5\%}}{value}$) varied between 11\% for $\lambda$ and 33\% for $\psi_p$ (in all the settings).
-
-\begin{figure}[!pht]
-\centering 
-\includegraphics[width=1\textwidth]{Fig_errors.eps}
-\caption{Comparison of inference accuracy of bdpn on a data set of 100 trees of 500--1\,000 tips each, with either $\lambda$ (blue), $\psi$ (orange) or $\rho$ (green) parameter being fixed to its real value.
-We show the swarmplots (coloured by fixed parameter) of relative errors for each test tree and parameter, which are measured as the normalized distance between the estimate and the real value.
-Average relative error (and in parentheses average bias, calculated on normalized values) are displayed for each parameter and method below their swarmplot.
-For $\rho$ absolute values are shown instead of normalized ones.
-}
-\label{fig:sim} 
-\end{figure}
-
- \begin{table}[!h]\centering
-\small
-\caption{Percentage of simulated trees for which the real parameter values were withing the estimated 95\% CIs. \smallskip}
-\begin{tabular}{c|ccc}
-\textbf{parameter} & \textbf{$\lambda$ fixed} & \textbf{$\psi$ fixed} & \textbf{$\rho$ fixed} \\
-\toprule 
- $\lambda$ & -- &  82\% & 92\% \\
- $\psi$ & 34\% & -- & 78\% \\
- $\rho$ & 20\% & 51\%  & -- \\
- $\psi_p$ & 95\% & 94\% & 95\% \\
- $\rho_n$ & 69\% & 67\% & 64\% \\
-\bottomrule
-\end{tabular}
-\label{tbl:ci}
-\end{table}
-
-\subsection{Application: HIV-B epidemic in the UK}
-We applied our estimator to asses the partner notification in HIV-infected patients in the UK.
-We used the dated phylogenetic tree from our recent study of HIV drug resistance in the UK~\citep{zhukovaModelingDrugResistance2023}.
-The tree represents the UK HIV-B epidemic between 1960s and 2016, its tips represent viruses sampled from 39\,130 individuals between 1996 and 2016.
-The samples used for the tree reconstruction were obtained from the UK HIV Drug resistance database~\citep{Dunn2007}.
-
-Between 2012 and 2015, in the UK antiretroviral treatment (ART) of HIV-infected individuals  was initiated when their CD4 count dropped below 350 cells/mL~\citep{williamsBritishHIVAssociation2012}.
-However ``if a patient with a CD4 cell count > 350 cells/mL wishes to start ART to reduce the risk of transmission to partners, this decision is respected and ART is started''~\citep{williamsBritishHIVAssociation2012}.
-Before 2012, treatment started with an even lower CD4 (i.e.
-later).
-The British HIV Association 2015 guidelines recommend all individuals with suspected or diagnosed primary HIV infection ``are offered immediate ART''~\citep{churchillBritishHIVAssociation2016}.
-To have a more homogeneous setting in terms of access to ART, awareness of the HIV infection, etc., we cut the dated tree between 2012 and 2015, and estimated the BD-PN parameters on the 2012--2015 forests.
-
-According to the \href{https://webarchive.nationalarchives.gov.uk/ukgwa/20181112132123mp_/https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/602942/HIV_in_the_UK_report.pdf}{``Towards elimination of HIV transmission, AIDS and HIV-related deaths in the UK''}
- report, the estimated total number of people living with HIV in the UK in 2015 was 101\,200 (97\,500 to 105\,700).
-% in 2012 this number was 98\,400 (93\,500-104\,300) [\href{HIV in the United Kingdom: 2013 Report}{https://webarchive.nationalarchives.gov.uk/ukgwa/20181112133715mp_/https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/326601/HIV_annual_report_2013.pdf}].
-Moreover these document report about 400 HIV-related deaths per year.
-According to our estimate~\citep{zhukovaModelingDrugResistance2023}, 66.5\% of the HIV-infected individuals in the UK were infected with HIV-B.
-We hence estimated that in 2015 67\,298 (64\,837, 70\,291) people were living with HIV-B.
-Our tree contains 39\,047 tips sampled by the end of 2015.
-Hence, it accounts for about 58\% (56\%, 60\%) of the total number of individuals living with HIV-B in the UK.
-However this estimate does not take into account deaths of the HIV-infected individuals over the years, which would increase the total number of HIV-infected individuals by 2015, and decrease the proportion of individuals represented by our tree.
-We therefore tested several sampling probabilities $\rho=$0.6; 0.5; 0.4.
-
-Using $\rho=0.6$, we estimated the following parameter values: $\lambda=0.262$, $\psi=0.186$, $\psi_p=186.921$, $\rho_n=0.033$, which corresponds to $R_e = 1.4$, infectious time $\frac{1}{\psi} = 5.4$ years, and about 2 days between partner notification and their sampling.
-The results for $\rho=0.4$ and for $\rho=0.6$ were similar: $R_e$ of respectively $1.5$ and
-$1.3$, infectious time of $5.1$ and $5.6$ years, partner notification probability of $0.033$ in both cases and about 2 days between partner notification and their sampling.
-
-The average time of HIV progression to AIDs in the absence of treatment is 10 years, which could serve as an estimate of the infectious time.
-However, in the UK, 95\% of infected population is on antiretroviral treatment, which, when taken properly, prevents transmission.
-Hence, the average infectious time represents the average time before starting treatment.
-Our estimate is approximately 5 years.
-
-\bibliographystyle{plainnat}
-\bibliography{refs}
-
 \end{document}

### 2023-07-28 hiv_b_uk/py/summary_table.py
UK results

@@ -1,0 +1,122 @@
+import os
+import re
+
+import pandas as pd
+
+from bdpn.tree_manager import read_forest
+
+
+def latexify_values(ci=True):
+    df['p'] = ' $' + df['p'].apply(str) + '$'
+    if ci:
+        df['infectious_time'] = ' $' + df['infectious_time'].apply(str) + '\;[' + df['infectious_time_min'].apply(
+            str) + '-' + df['infectious_time_max'].apply(str) + ']$'
+        df['partner_removal_time'] = ' $' + df['partner_removal_time'].apply(str) + '\;[' + df[
+            'partner_removal_time_min'].apply(
+            str) + '-' + df['partner_removal_time_max'].apply(str) + ']$'
+        df['R'] = ' $' + df['R'].apply(str) + '\;[' + df['R_min'].apply(str) + '-' + df['R_max'].apply(str) + ']$'
+        df['lambda'] = ' $' + df['lambda'].apply(str) + '\;[' + df['lambda_min'].apply(str) + '-' + df['lambda_max'].apply(
+            str) + ']$'
+        df['psi'] = ' $' + df['psi'].apply(str) + '\;[' + df['psi_min'].apply(str) + '-' + df['psi_max'].apply(
+            str) + ']$ '
+        df['psi_p'] = ' $' + df['psi_p'].apply(str) + '\;[' + df['psi_p_min'].apply(str) + '-' + df['psi_p_max'].apply(
+            str) + ']$ '
+        df['pn'] = ' $' + df['pn'].apply(str) + '\;[' + df['pn_min'].apply(str) + '-' + df['pn_max'].apply(
+            str) + ']$ \\\\'
+    else:
+        df['infectious_time'] = ' $' + df['infectious_time'].apply(str) + '$'
+        df['partner_removal_time'] = ' $' + df['partner_removal_time'].apply(str) + '$  '
+        df['R'] = ' $' + df['R'].apply(str) + '$'
+        df['lambda'] = ' $' + df['lambda'].apply(str) + '$'
+        df['psi'] = ' $' + df['psi'].apply(str) + '$ '
+        df['psi_p'] = ' $' + df['psi_p'].apply(str) + '$ '
+        df['pn'] = ' $' + df['pn'].apply(str) + '$   \\\\'
+    df['repetition'] = ' $' + df['repetition'].apply(str) + '$ '
+    df['observed_trees'] = ' $' + df['observed_trees'].apply(str) + '$ '
[…]

### 2023-07-28 bdpn/bd.py
minor: delimiter fixed in BD/BDPN output

@@ -115,7 +115,7 @@ def infer(forest, la=None, psi=None, p=None,
 def save_results(vs, cis, log, ci=False):
     os.makedirs(os.path.dirname(os.path.abspath(log)), exist_ok=True)
     with open(log, 'w+') as f:
-        f.write('\t{}\n'.format(','.join(['R0', 'infectious time', 'sampling probability',
+        f.write(',{}\n'.format(','.join(['R0', 'infectious time', 'sampling probability',
                                           'transmission rate', 'removal rate'])))
         la, psi, rho = vs
         R0 = la / psi

### 2023-07-28 bdpn/bdpn.py

@@ -283,7 +283,7 @@ def loglikelihood(forest, la, psi, psi_n, rho, rho_n, T=None, threads=1):
 def save_results(vs, cis, log, ci=False):
     os.makedirs(os.path.dirname(os.path.abspath(log)), exist_ok=True)
     with open(log, 'w+') as f:
-        f.write('\t{}\n'.format(','.join(['R0', 'infectious time', 'sampling probability', 'notification probability',
+        f.write(',{}\n'.format(','.join(['R0', 'infectious time', 'sampling probability', 'notification probability',
                                           'removal time after notification',
                                           'transmission rate', 'removal rate', 'partner removal rate'])))
         la, psi, psi_p, rho, rho_p = vs

### 2024-02-23 article/article.tex
Rewrote PN test and added a figure for Pn trees

@@
 %opening
-\title{MTBD models with partner notification}
-\author{Anna Zhukova}
+\title{Epidemiological birth-death models with partner notification}
+\author{Anna Zhukova et al.}
 
@@
 genealogies of the pathogen population, approximating the transmission trees) combined with metadata on the samples.
-This is particularly useful for emerging epidemics, when not enough data (e.g.
-incidence curves) might be gathered for accurate estimations with classic epidemiological methods, while using genetic data for phylodynamic estimations can provide valuable insights and help prevent epidemic spreads (e.g.
-accurate estimation of the infectious period is crucial for adjusting health policies for self-isolation).
+Classic epidemiological methods make inferences from incidence curves, which in the presence of import cases might be biased and give appearance of elevated local transmission.
+In contrast, phylodynamic approach can overcome such biases as it uses the transmission structure as one of its information sources~\citep{vaughanEstimatesEarlyOutbreakspecific2024}.
+Therefore, using genomic data with phylodynamic estimations can provide valuable insights for preventing epidemic spreads.
 
@@
 Another important source of heterogeneity that none of these models accounts for is inflicted by non-random sampling.
-We can propose different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability over time intervals with the BDSKY, however, none of them can account for non-randomness of sampling, in particular due to contact tracing and partner notification.
+We can propose different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability over time intervals with the BDSKY, however, none of them can account for non-randomness of sampling, in particular due to partner notification (contact tracing).
+Partner notification or contact tracing is a process in which contacts of detected infected individuals are identified and offered testing for infection.
+It has been applied to various infectious diseases, including syphilis, HIV, tuberculosis, Ebola and COVID-19, and shown to successfully reduce transmission~\citep{el-sadrContactTracingBarriers2022}.
+For instance, for HIV,  partner notification showed increased early referral and initiation of treatment, and several countries include it in their national HIV testing services policies~\citep{worldhealthorganizationCountryPolicyReview2016}.
 
-In this study we propose an extension of the BD model that allows for modeling non-random sampling due to partner notification (PN), BD-PN.
-In the rest of the article, we describe the BD-PN model and its assumptions, present a transmission tree simulator under this model, derive the equations for BD-PN likelihood calculation, and propose a test for detecting partner notification in pathogen phylogenetic trees.
-We also present a maximum likelihood parameter estimator for BD-PN, and test it on simulated data.
-Finally, we apply it to the study of HIV-1 B epidemic in the UK.
+In this study we propose an extension of MTBD models that allows for modeling non-random sampling due to partner notification (PN), MTBD-PN.
+We describe the -PN extension and its assumptions, and present a transmission tree simulator under MTBD-PN models.
+We propose a test for detecting partner notification in pathogen phylogenetic trees.
+For the simplest representative of this model family, the BD-PN model, we also derive the equations for its likelihood calculation and present a maximum likelihood parameter estimator.
+We test our BD-PN parameter estimator on simulated data, and apply it to the study of HIV-1 B epidemic in the UK.
 
@@
     & - 2 \lambda p^{(i)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where one of the subtrees stayed unsampled}\\
-p^{(i)}(t_i) = 1
+p^{(i)}(t_i) =  &1\;  \textit{\color{gray} $\leftarrow$ the probability of no change over zero time is 1}
 \end{cases}\label{eq:p}
@@
 log L(\mathscr{T}|\Theta) =  &\sum\limits_{i \in tips}  log(\psi \rho)  \textit{\color{gray} $\leftarrow$ sampling of $n$ tips} \\
- &+\sum\limits_{i \in \stackanchor{\tiny\textit{internal}}{\tiny\textit{nodes}}} log \big(2 \lambda p^{(i0)}(t_i)p^{(i1)}(t_i)\big)   \textit{\color{gray} $\leftarrow$ $n - 1$ transmission events,}\\
- & ~~~~~~~~~~~~~~~\textit{\color{gray} followed by child branch evolutions (each can be a donor)}  \label{eq:mtbd_tree_likelihood}
+ &+\sum\limits_{i \in \stackanchor{\tiny\textit{internal}}{\tiny\textit{nodes}}} log \big(2 \lambda p^{(i0)}(t_i)p^{(i1)}(t_i)\big)   \textit{\color{gray} $\leftarrow$ $n - 1$ transmission events, followed by child }\\
+ & ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\textit{\color{gray} branch evolutions (each can be a donor)}  \label{eq:mtbd_tree_likelihood}
 \end{split}
@@
 \item \textit{effective reproduction number} $R_e = \frac{\lambda}{\psi}$, expected number of individuals directly infected by an infectious case;
-\item \textit{infectious time} $\frac{1}{\psi}$, time during which an infectious individual can further spread the epidemic.
+\item \textit{infectious time} $\frac{1}{\psi}$, expected time during which an infectious individual can further spread the epidemic.
 \end{itemize}
 
-Importantly, the BD model is asymptotically unidentifiable (see Remark 3.4 in~\citep{Stadler2009}), but to become identifiable it requires one of the parameters to be fixed.
+Importantly, the BD model is asymptotically unidentifiable (see Remark 3.4 in~\citep{Stadler2009}).
+To become identifiable it requires one of its parameters to be fixed.
 In practice, it is often the sampling probability, as it may be approximated from epidemiological data $\rho$ (e.g., the proportion of sampled cases among the declared ones) or the infectious time $\frac{1}{\psi}$ (estimated from observations of infected cases).
+
+\subsection{MTBD models}
+MTBD models~\citep{Stadler2013a} add population structure to the BD model by allowing different individual states, transmissions between them and state changes.
+A general MTBD model with $m$ individual states has $2m^2 + m$ parameters: An individual in state $k \in \{1, \ldots, m\}$ can be removed at a constant average rate $\psi_k$ (with pathogen sampling probability $\rho_k$), change their state to state $l \in \{1, \ldots, m\}$ at a constant average rate $\mu_{kl}$ (where $\mu_{kk} = 0$), and transmit their pathogen to an individual in state $l$ at a constant average rate $\lambda_{kl}$.
+The time between events of the same type is hence modeled with exponential distribution.
 
@@
 \begin{enumerate}
-\item only observed individuals can notify (instead of any removed individual);
-\item notified individuals are always observed upon removal;
+\item only observed individuals can notify (instead of any removed individual);\\
+
+This is a realistic hypothesis in which an individual is ``observed'' because he or she enters a diagnostic and follow-up process that includes partner notification.
+Without entering this process, he or she is not observed and does not notify.
+
+\item notified individuals are always observed upon removal;\\
+
+Realistically again, we assume here that the diagnosis and follow-up process works as expected, sampling notified patients.
+
 %\item notified individuals do not notify further;
-\item after the notification, notified individuals do not transmit further;
-\item only the most recent partner can get notified.
+\item after the notification, notified individuals do not transmit further (but might notify in their turn);\\
+
+We assume here that the diagnosis and follow-up process works as expected, so that the notified patients are either isolated or provided with treatment or other means of preventing further transmission.
+Note that the most recent partner relationship is not necessarily symmetrical.
+The most recent partner $j$ of individual $i$ might have transmitted their virus further (to individual $k$) after the contact with $i$ and before the notification by $i$.
+Hence, while $j$ is the most recent partner of $i$, $j$'s most recent partner is $k$, and, once notified, $j$ might notify $k$.
+
+\item only the most recent partner can get notified;\\
+
+This assumption might need to be removed or relaxed (e.g.
+to up to a certain fixed number of most recent partners, or all the partners within a certain time interval) in the future work.
+It is however useful to make the mathematics behind the model manageable.
+In the case of sexually transmitted diseases, notifying only the most recent partner could roughly correspond to notifying the ``main'' partner (e.g.
+spouse).
 \end{enumerate}
 
-These assumptions make the mathematics behind the model manageable, however they can be easily relaxed in a tree simulator (e.g., to be used with a deep learning estimator~\citep{Voznica2021}, which does not require likelihood calculation).
+These assumptions can be easily relaxed in a tree simulator (e.g., to be used with a deep learning estimator~\citep{Voznica2021}, which does not require likelihood calculation).
 
-Note that the most recent partner relationship is not necessarily symmetrical, as the most recent partner of $i$ might have transmitted their virus further after the contact with $i$.
+The -PN extension can be easily generalized from BD to MTBD models: at the moment of sampling an individual of any type can notify their last partner with the probability $\rho_n$, who will then stop transmitting and get removed at a constant removal rate $\psi_n >> \psi_k \; \forall k \in \{1, \ldots, m\}$.
+The -PN extension therefore adds two parameters to an MTBD model: $\psi_p$ and $\rho_n$.
 
-In Appendix we discuss how the -PN extension can be generalized to multi-type birth-death models (MTBD-PN).
+\subsection{MTBD-PN tree simulator}
 
-\subsubsection{Mathematical formulation}
-$p^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along a tree branch, provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$.
+We implemented a tree simulator that generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states). We give its algorithmic details in Appendix.
+
+\subsubsection*{Code availability}
+The simulator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\citep{Huerta-Cepas2016} and NumPy package for array operations~\citep{harris_array_2020}.
+
+It is available as a command-line program and a Python 3 package via PyPi (\href{https://pypi.org/project/treesimulator}{treesimulator}), and via Docker/Singularity (\href{https://hub.docker.com/r/evolbioinfo/treesimulator/tags}{evolbioinfo/treesimulator}).
+Its source code and the installation and usage documentation are available on GitHub at \href{https://github.com/evolbioinfo/treesimulator}{github.com/evolbioinfo/treesimulator}.
+
+\section{PN test}~\label{sec:test}
+To access whether a give tree is generated in a presence of partner notification, we developed a non-parametric test based on cherries (two tips having a common parent).
+
+The intuition behind the test is that in the presence of partner notification the tree will contain more cherries (two-tip clades) whose tips are close in time.
+Indeed, in a transmission tree internal nodes correspond to transmissions while tips correspond to sampling events.
+Hence if a transmission tree is complete, a cherry corresponds to a transmission from an infected individual to their most recent partner, followed by sampling of both.
+If there is no partner notification, these two sampling events are independent, while in the presence of partner notification the first sampling might trigger the second one, and they will be close in time.
+For an incomplete transmission tree (where some patients were not sampled), the only difference would be that some cherries would include hidden transmissions, and hence not correspond to the two most recent partners.
+
+More formally, in the absence of partner notification (null hypothesis, H0), we assume that branch lengths of two cherry tips are independently drawn from the same distribution (which is the case for MTBD models).
+In the presence of partner notification (alternative hypothesis, H1), we assume that for some of the cherries (including notified partners) their branch lengths are not independent and are similar.
+%We give control to the user on what should be considered ``some of the cherries'' and ``similar branch lengths'' for their data, via adjustable test parameter values.
+
+Let us assume there are N cherries in the tree.
+For each cherry $i\;(1 \leq i \leq N)$, our test calculates the difference between its tip times $d_i$.
+It then generates a reshuffled cherry $i'$: It picks 5 (adjustable parameter) cherries with the roots that are the closest in time to the root of $i$, randomly selects two tips among their tips, and calculates their time difference $d_{i'}$.
+The test then compares $d_{i'}$ to $d_{i'}$: $sign(d_{i'} - d_i) = 1$ if $d_{i'}$ is larger than $d_i$; $= -1$ if it is smaller; or $= 0$ if they are equal.
+Note that if cherry tip branch lengths are independently drawn from the same distribution, $d_{i'}$ will be smaller than $d_i$ for about half of the cherries: $S' =  \sum\limits_{i,i'=1}^N sign(d_{i'} - d_i) \approx 0$.
+However, if some cherries are ``notified'' their real tip differences should be smaller that reshuffled ones, biasing the total sum: $S' > 0$.
+
+The reason for which we are constructing a random cherry from the neighbouring instead of any cherries (in terms of root date), is that cherry branch lengths depend on the date of the cherry root.
+As tip sampling times are bounded from above by the end of the sampling period, the closer the cherry root is to the end of the sampling period, the shorter on average its branches will be (and less time difference they will be able to have in any case).
+
+The reshuffled cherry tip difference generation is repeated 1\,000 times (adjustable parameter), hence obtaining 1\,000 array comparisons $S'^{(1)}, \ldots, S'^{(1\,000)}$.
+
+Finally, the test reports the proportion of reshuffled arrays whose comparison was negative ($S'^{(k)} < 0$). Note that in the presence of partner notification, this proportion should be very low. 
+The test therefore reports the probability of partner notification being present if the tree.
+
+\subsection*{Code availability}
+We implemented the PN-test in Python 3. It is available as a part of BDPN parameter estimator (see section~\ref{sec:sim}).
+
+\subsection{Performance on simulated data}~\label{sec:test:sim}
+To illustrate the PN test performance, we applied it to the trees simulated under three classical MTBD-family models: BD, BDEI (Birth-Death Exposed-Infectious~~\citep{Stadler2014}) and BDSS (Birth-Death with SuperSpreading~\citep{Stadler2013a}, see Fig.~1 in \citet{Voznica2021} for a graphical representation of these models and their parameters), and under their -PN versions: BD-PN, BDEI-PN and BDSS-PN models.
+
+For each model, we generated 100 trees with 500--1\,000 tips with our MTBD/MTBD-PN tree simulator. For each tree the parameter values were drawn randomly within the following bounds:
+\begin{itemize}
+\item $R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
+\item removal rate $\psi \in ]1 / 20, 1 / 5]$,
+\item sampling probability $\rho \in ]0.1, 0.9]$,
+\item \textit{(for -PN models only)} ratio between the partner removal rate after notification and the removal rate $\frac{\psi_p}{\psi} \in ]50, 500]$,
+\item \textit{(for -PN models only)} partner notification probability $\rho_n \in ]0.01/\rho, 0.9]$ (the lower bound was picked this way to ensure that the probability to be a notified partner is at least 0.01 $=\rho \rho_n$),
+\item \textit{(for BDEI models only)} becoming-infectious rate (at which an exposed individual becomes infectious) $\mu \in ]0.2, 10]$,
+\item \textit{(for BDSS models only)} fraction of superspreaders $f_{SS} = \frac{\lambda_{SS}}{\lambda_{SS} + \lambda_{SN}} \in ]0.05, 0.2]$,
+\item \textit{(for BDSS models only)} superspreading transmission ratio $x_{SS} = \frac{\lambda_{SS}}{\lambda_{NS}} = \frac{\lambda_{SN}}{\lambda_{NN}} \in ]3, 10]$.
+\end{itemize}
+
+The common parameters were consistent between models, in a sense that the $i$-th tree under the BDSS model had the same $R_e$, $\psi$ and $\rho$ values as the $i$-th trees under BDSS-PN, BD(-PN) and BDEI(-PN) models; the $i$-th tree under the BDSS-PN model had the same $\psi_p$ and $\rho_n$ values as the $i$-th trees under BD-PN and BDEI-PN models.
+
+We then run the PN tests on each tree of these 6 datasets.
+For 100/99/100 out of 100 trees in the BD/BDEI/BDSS dataset the PN test p-value was above $0.05$ (i.e.
+100\%/99\%/100\% true negative and 0\%/1\%/0\% false positive results).
+For 97/99/96 out of 100 trees in the BD-PN/BDEI-PN/BDSS-PN dataset the PN test p-value was below $0.05$ (i.e.
+97\%/99\%/96\% true positive and 3\%/1\%/4\% false negative results).
+These results are summarised in Table~\ref{tbl:pntest}.
+Our test therefore showed both high specificity and sensitivity.
+
+\begin{table}[!h]\centering
+\small
+\caption{PN test performance on trees generated under BD, BDEI, BDSS, BD-PN, BDEI-PN and BDSS-PN models (100 trees per model).
+We report the number of trees for each the PN-test p-value was $<0.05$, $\geq 0.05$, and its mean p-value for each model dataset.}
+\begin{tabular}{l|r|r|c}
+ & \textbf{p-value $<0.05$} & \textbf{p-value $\geq0.05$} & \textbf{mean p-value} \\
+  \midrule
+\textbf{BD}& 0 & 100 & 0.424 \\
+\textbf{BD-PN}& 97 & 3 & 0.004 \\
+  \midrule
+\textbf{BDEI}& 1 & 99 & 0.493 \\
+\textbf{BDEI-PN}& 99 & 1 & 0.001 \\
+  \midrule
+\textbf{BDSS}& 0 & 100 & 0.455 \\
+\textbf{BD-PN}& 96 & 4 & 0.004 \\
+\end{tabular}
+\label{tbl:pntest}
+\end{table}
+
+\section{Mathematical formulation of BD-PN model}
+In the classical BD model with incomplete sampling, a branch of the tree corresponds to a virus evolution between the transmission corresponding to the node at the beginning of this branch and either a sampling event (if the branch is external) or a transmission corresponding to the node at the end of the branch (if it is internal).
+Note, that since sampling is incomplete, hidden transmissions (leading to fully unsampled subtrees) could occur along this branch, where the unsampled subtree could correspond to either donor or recipient.
+Hence, we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at any time $t < t_i$ along the branch.
 These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.
+$p^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along such a tree branch (with potentially different individuals at times $t$ and (the branch end) $t_i$).
 
-We will use an \textbf{oriented probability} $p^{(i,o)}(t)$ for the cases when the individual at the end of the branch is the same as at time $t$:
+In trees generated by the BD-PN model, however, two other types of branches are possible.
+Let us consider a tip corresponding to a person A who got sampled at time $t_A$ and successfully notified their partner B (see Fig.~\ref{fig:pn-branches}).
+According to our model assumption 2, notified individuals are always observed.
+Hence B must be sampled too (at time $t_B \geq t_A$) and correspond to a tree tip, while the common ancestor of A and B (AB) corresponds to the transmission between A and B.
+Therefore the individual on the path between AB and B cannot change (is always B) and only hidden transmissions (from B to someone else) where the recipient subtree stayed unobserved are possible along the branch(es) leading from AB to B (before $t_A$).
+Note, that non-hidden transmissions are also possible before the notification time $t_A$, and in such a case we know which subtree corresponds to the donor (the one that contains B).
+Between $t_A$ and $t_B$ no transmission is possible on the branch of B as, according to assumption 3, after notification notified individuals do not transmit further.
+
+We will use an \textbf{oriented probability} $p^{(i,o)}(t)$ for cases when the individual at the end of the branch (or considered time interval) is the same as at time $t$ but might have transmitted to someone else (unobserved) along the way:
 
@@
 
-Additionally we define a \textbf{non-hidden probability} $p^{(i,nh)}(t)$ for the cases when no hidden transmission occurred along the branch:
+We will use an \textbf{partner probability once notified} $p^{(i,np)}(t)$ for cases when the individual at the end of the branch is the partner who got notified at time $t_n < t$:
+
+\begin{equation}
+\scriptsize
+\begin{cases}
+\dot{p}^{(i,np)}(t) = & \psi_p p^{(i,nt)}(t)\; \textit{\color{gray} $\leftarrow$ no sampling in the next infinitesimal time $\Delta t$ }\\
+p^{(i,o)}(t_i) = 1\label{eq:p-np}
+\end{cases}
+\end{equation}
+
+Moreover, according to our model assumption 4, only the most recent partner can get notified.
+Hence, AB must be the parent node of A, and no hidden tranmission is possible along the branch AB-A (as otherwise there would have been a partner of A who is more recent than B).
+
+We will use a \textbf{non-hidden probability} $p^{(i,nh)}(t)$ for such cases (when no hidden transmission occurred along the branch):
 
@@
 
-The equations~\ref{eq:Us}, \ref{eq:p}, \ref{eq:p-o}, and \ref{eq:p-nh} have the following closed form solutions:
+Note that A could have also ``unsuccessfully'' notified B, if by the time $t_A$, when A decided to notify B, B already got sampled via the standard procedure or was notified by someone else.
+In this case nothing changes for the AB-A branch, and the AB-B branch(es) stay oriented.
+
+\begin{figure*}[tbhp]
+\centering 
+%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
+\includegraphics[width=0.6\textwidth]{Fig_PNbranches}
+\caption{An example of partner notification in a transmission tree: an individual A got sampled at time $t_A$ (violet tip) and notified their most recent partner B (orange tip).
+The internal node AB corresponds to the transmission between A and B.
+The fact that A is the notifier of B implies that there is no hidden event along its branch AB-A, and that in any transmission on the path AB-B the donor is B.
+Moreover, between the moment of notification ($t_A$) and the moment of sampling  of B ($t_B$) there are no hidden events on the branch B.}
+\label{fig:pn-branches} 
+\end{figure*}
+
+%For the BD-PN model we will also need to consider the case where the individual at the end of the branch is the same as at time $t$.
+This corresponds to a notified partner branch, as according to our model assumptions if a partner is notified, he or she is sampled (i.e.
+observed in our tree): Hence the individual at the beginning of the branch (internal node corresponding to transmission) should be the same as at its end (sampled partner if it finishes with a tip, or a transmission where the partner is the donor).
+%
+%
+%We will use an \textbf{oriented probability} $p^{(i,o)}(t)$ for the cases when the individual at the end of the branch is the same as at time $t$:
+%
+%\begin{equation}
+%\scriptsize
+%\begin{cases}
+%\dot{p}^{(i,o)}(t) = & \Big(\lambda + \psi\Big) p^{(i,o)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+%    & - \lambda p^{(i,o)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
+%p^{(i,o)}(t_i) = 1\label{eq:p-o}
+%\end{cases}
+%\end{equation}
+%
+%
+%Additionally we define a \textbf{non-hidden probability} $p^{(i,nh)}(t)$ for the cases when no hidden transmission occurred along the branch:
+%
+%\begin{equation}
+%\scriptsize
+%\begin{cases}
+%\dot{p}^{(i,nh)}(t) = & \Big(\lambda + \psi\Big) p^{(i,nh)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+%p^{(i,nh)}(t_i) = 1\label{eq:p-nh}
+%\end{cases}
+%\end{equation}
+
+The equations~\ref{eq:Us}, \ref{eq:p}, \ref{eq:p-o}-\ref{eq:p-nh} have the following closed form solutions:
 \begin{equation}
@@
 p^{(i,nh)}(t) =  E_4
+\\
+p^{(i,np)}(t) =  e^{\psi_p (t -t_i)}
 \end{cases},\\
@@
 
-\section{PN test}~\label{sec:test}
-To access whether a give tree is generated under a classical MTBD model or MTBD-PN, we developed a non-parametric test based on cherries (two tips having a common parent). 
-The intuition behind the test is that in the presence of partner notification the tree will contain more cherries whose tips are close in time.
-
-The test sorts the cherries by the times of their roots, and splits them into blocks of 100 cherries (adjustable parameter).
-For each cherry in a block, the test calculates the difference between its tip times, hence obtaining an array of 100 cherry tip differences.
-It then generates a collection of random cherry tip differences for the same block: For each original cherry root it picks 5 (adjustable parameter) cherries with the roots that are the closest in time, randomly selects two tips among their tips, and calculates their time difference.
-An array of 100 reshuffled cherry tip differences is thus obtained for the same block.
-The reshuffled cherry tip difference array generation is repeated 1\,000 times (adjustable parameter).
-Finally, the test reports the proportion of reshuffled arrays whose first quantile (adjustable parameter) is smaller than that of the real cherry array.
-
-The test therefore reports a probability of partner notification being present at time interval corresponding to each block.
-To estimate the probability of partner notification being present in the whole tree, it averages the block probabilities.
-
-\subsection*{Code availability}
-We implemented the PN-test in Python 3. It is available as a part of BDPN parameter estimator (see section~\ref{sec:sim}).
-
-\subsection{Performance on simulated data}~\label{sec:test:sim}
-To illustrate the PN test performance, we applied it to the trees simulated under BD and under BD-PN models.
-We implemented a MTBD/MTBD-PN tree simulator (see Appendix for details) and used it to generate 100 trees with 500--1\,000 tips under the BD-PN model.
-For each tree the parameter values were drawn randomly within the following bounds:
-$R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
-removal rate $\psi \in ]1 / 20, 1 / 5]$,
-ratio between the partner removal rate after notification and the removal rate $\frac{\psi_p}{\psi} \in ]50, 500]$,
-sampling probability $\rho \in ]0.1, 0.9]$,
-partner notification probability $\rho_n \in ]0.01/\rho, 0.9]$.
-We also generated 100 trees of 500--1\,000 tips under the BD model.
-For each tree in the BD data set the parameter values ($\lambda, \psi,$ and $\rho$) were the same as the ones selected for the corresponding tree of the BD-PN dataset.
-
-We then run the PN tests on each tree of the both datasets.
-For 98 out of 100 trees in the BD dataset the PN test value was above $0.05$ (i.e.
-98\% true negative and 2\% false positive results), the mean PN test value was $0.468$.
-For 96 (out of 100) trees in the BD-PN dataset the PN test value was below $0.05$ (i.e.
-96\% true positive and 4\% false negative results), the mean PN test value was $0.008$.
-Our test therefore showed both high specificity and sensitivity.
-
 \section{BD-PN parameter and CI estimator}\label{sec:sim}
@@
 
-Using $\rho=0.58$, we estimated the following parameter values: $\lambda=0.53$, $\psi=0.43$, $\psi_p=23--59$, $\rho_n=0.006--0.007$, which corresponds to $R_e = 1.2$, infectious time $\frac{1}{\psi} = 2.3$ years, and about 6--16 days between partner notification and their sampling.
+Using $\rho=0.58$, we estimated the following parameter values: $\lambda=0.53$, $\psi=0.43$, $\psi_p=23-59$, $\rho_n=0.006-0.007$, which corresponds to $R_e = 1.2$, infectious time $\frac{1}{\psi} = 2.3$ years, and about 6--16 days between partner notification and their sampling.
 The results are consistent between 10 forests and shown in Table~\ref{tbl:uk}.
@@
 Hence, the average infectious time represents the average time before starting treatment.
-Our estimate is approximately 5 years.
+Our estimate is approximately 2.3 years.
 
@@
 The advantage of the BD-PN model is that the differential equations describing it have closed-form solutions, which are quick to calculate and do not require numerical approximation (as is the case for more complex MTBD models).
-Tests on simulated data show that our estimator robustly infers model parameter values and their CIs in the setting when the duration of the infectious period $\frac{1}{psi}$ or the base sampling probability $\rho$ if fixed.
+Tests on simulated data show that our estimator robustly infers model parameter values and their CIs in the setting when the duration of the infectious period $\frac{1}{\psi}$ or the base sampling probability $\rho$ if fixed.
 Fixing one of the model parameters is needed for identifiability.
-In practice, $\rho$ and $\frac{1}{psi}$ may often be approximated from epidemiological data, e.g., as the proportion of sampled cases among the declared ones (for $\rho$), or estimated from observations of infected cases ($\frac{1}{\psi}$).
+In practice, $\rho$ and $\frac{1}{\psi}$ may often be approximated from epidemiological data, e.g., as the proportion of sampled cases among the declared ones (for $\rho$), or estimated from observations of infected cases ($\frac{1}{\psi}$).
 
@@
 
-The BD-PN model is the first steps towards accounting for non-random sampling in phylodynamic models.
+The BD-PN model is the first step towards accounting for non-random sampling in phylodynamic models.
 In order to reduce its mathematical complexity (having closed-form solutions for its differential equations) and make parameter estimation fast, we made several assumptions: (i) that only detected individuals can notify their partner; (ii) only the most recent partner can get notified; (iii) after the notification, notified partners do not transmit further; and (iv) notified partners are always sampled upon removal (i.e., observed in the transmission tree).
@@
 \section{Appendix}
-\subsection{MTBD-PN models}
-MTBD models~\citep{Stadler2013a} add population structure to the BD model by allowing different individual states, transmissions between them and state changes.
-A general MTBD model with $m$ individual states has $2m^2 + m$ parameters: An individual in state $k \in \{1, \ldots, m\}$ can be removed at a constant average rate $\psi_k$ (with pathogen sampling probability $\rho_k$), change their state to state $l \in \{1, \ldots, m\}$ at a constant average rate $\mu_{kl}$ (where $\mu_{kk} = 0$), and transmit their pathogen to an individual in state $l$ at a constant average rate $\lambda_{kl}$.
-The time between events of the same type is hence modeled with exponential distribution.
-
-The -PN extension can be easily generalized from BD to MTBD models: at the moment of sampling an individual of any type can notify their last partner with the probability $\rho_n$, who will then stop transmitting and get removed at a constant removal rate $\psi_n >> \psi_k \; \forall k \in \{1, \ldots, m\}$.
-The -PN extension therefore adds two parameters to an MTBD model: $\psi_p$ and $\rho_n$.
 
@@
 
-We implemented a tree simulator that generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states).
-The simulator is Gillespie-based, generates state change, transmission and removal times, and only reconstructs the sampled parts of the tree to save memory and increase speed.
+Our tree simulator generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states).
+It is Gillespie-based, generates state change, transmission and removal times, and only reconstructs the sampled parts of the tree to save memory and increase speed.
 
@@
 
-\subsubsection*{Code availability}
-The simulator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\citep{Huerta-Cepas2016} and NumPy package for array operations~\citep{harris_array_2020}.
-
-It is available as a command-line program and a Python 3 package via PyPi (\href{https://pypi.org/project/treesimulator}{treesimulator}), and via Docker/Singularity (\href{https://hub.docker.com/r/evolbioinfo/treesimulator/tags}{evolbioinfo/treesimulator}).
-Its source code and the installation and usage documentation are available on GitHub at \href{https://github.com/evolbioinfo/treesimulator}{github.com/evolbioinfo/treesimulator}.
-
 \end{document}

### 2024-02-23 bdpn/model_distinguisher.py
Changed PN test to use the whole tree

@@ -32,8 +32,7 @@ class CherryLikeMotif(object):
             .format(self.root.name, len(self.clustered_tips), ', '.join(_.name for _ in self.clustered_tips))
 
 
-def nonparametric_cherry_diff(cherries, k=DEFAULT_NEIGHBOURHOOD_SIZE, repetitions=RANDOM_REPETITIONS,
-                              percentile=DEFAULT_PERCENTILE):
+def nonparametric_cherry_diff(cherries, k=DEFAULT_NEIGHBOURHOOD_SIZE, repetitions=RANDOM_REPETITIONS):
     """
     Calculates the branch differences in real (pseudo)cherries and in random ones.
     Random cherry configuration is generated by replacing each real (pseudo)cherry
@@ -41,13 +40,19 @@ def nonparametric_cherry_diff(cherries, k=DEFAULT_NEIGHBOURHOOD_SIZE, repetition
     The k-neighbourhood contains the k cherries whose roots are the closest in time to the (pseudo)cherry in question,
     including itself.
 
-    We compare the first percentile of the real (pseudo)cherry branch differences
+    We compare the real (pseudo)cherry branch differences
     to the distribution of those for the random configurations.
+    Each real [d1, d2, ..., dn] vs random [d'1, d'2, ..., d'n] configuration comparison is performed by:
+        * comparing one by one each corresponding cherry and calculating the sign:
+        -1 if the random one if smaller, +1 if the real one is smaller or 0 if they are equal
+        * calculating the sum of signs (so that for similar distributions the sum should be close to 0)
+    Finally we report the proportion of sing sums that are smaller than 1.
 
     :param repetitions: how many times a random (pseudo)cherry configuration should be generated.
     :param k: the size of the k-neighbourhood.
     :return: the proportion, between 0 and 1,
-        of random 1st quantiles of (pseudo)cherry diffs less or equal to the real one.
+        of random (pseudo)cherry diffs that have more than half elements that are smaller
+        than the corresponding elements in the real one.
     """
     cherries = sorted(cherries, key=lambda _: getattr(_.root, TIME))
     n = len(cherries)
@@ -82,15 +87,16 @@ def nonparametric_cherry_diff(cherries, k=DEFAULT_NEIGHBOURHOOD_SIZE, repetition
             yield np.random.choice(reduce(lambda l1, l2: l1 + l2, (_.clustered_tips for _ in neighbours), []),
[…]

### 2024-02-23 examples/main_cherry_test.py

@@ -1,36 +1,18 @@
 import logging
 import os
 
-from ete3 import Tree
+from bdpn.model_distinguisher import pn_test
+from bdpn.tree_manager import read_forest
 
-from bdpn.model_distinguisher import pick_cherries, nonparametric_cherry_diff, pick_motifs
-from bdpn.tree_manager import annotate_tree, TIME
-
-DATA_DIR = os.path.abspath(os.path.join(os.path.dirname(__file__), '..', 'trees'))
-NWK_BD = os.path.join(DATA_DIR, 'bd', 'tree.1.nwk')
-NWK_BDPN = os.path.join(DATA_DIR, 'bdpn', 'tree.1.nwk')
+DATA_DIR = os.path.abspath(os.path.join(os.path.dirname(__file__), '..', 'simulations', 'trees'))
+NWK_BD = os.path.join(DATA_DIR, 'BD', 'tree.{}.nwk')
+NWK_BDPN = os.path.join(DATA_DIR, 'BDPN', 'tree.{}.nwk')
 min_cherry_num = 100
 
 if __name__ == '__main__':
     logging.basicConfig(level=logging.INFO, format='%(asctime)s: %(message)s', datefmt="%Y-%m-%d %H:%M:%S")
 
-    for nwk in (NWK_BD, NWK_BDPN):
-        logging.info('==============={}============'.format(nwk.replace(DATA_DIR, '')))
-
-        results = []
-        tree = Tree(nwk)
-        annotate_tree(tree)
-
-        all_cherries = sorted(list(pick_cherries(tree, include_polytomies=True)), key=lambda _: getattr(_.root, TIME))
-        logging.info('Picked {} cherries with {} roots.'.format(len(all_cherries), len({_.root for _ in all_cherries})))
-        for i in range(0, len(all_cherries) // min_cherry_num):
-            cherries = all_cherries[i * min_cherry_num: (i + 1) * min_cherry_num]
-            result = nonparametric_cherry_diff(cherries, repetitions=1e3)
-            results.append(result)
-
-            logging.info(
-                'For {n} cherries with roots in [{start}-{stop}), PN test {res}.'
-                .format(res=result, start=getattr(cherries[0].root, TIME), stop=getattr(cherries[-1].root, TIME), n=len(cherries)))
-            cherries = []
-
-        print(sum(results) / len(results))
+    for tree_id in range(100):
[…]

### 2024-02-23 simulations/py/summary_table_cherries.py

@@ -16,7 +16,7 @@ if __name__ == "__main__":
         i = int(re.findall(r'[0-9]+', log)[0])
         with open(log, 'r') as f:
             pval = float(f.readline().strip('\n').split('\t')[1])
-        model = re.findall(r'BDPN|BD', log)[0]
+        model = re.findall(r'BDPN|BDEIPN|BDSSPN|BDEI|BDSS|BD', log)[0]
         df.loc['{}.{}'.format(i, model), :] = [model, i, pval]
 
     for model in df['model'].unique():

### 2024-04-30 article/article.tex
Sign test

@@
 genealogies of the pathogen population, approximating the transmission trees) combined with metadata on the samples.
-Classic epidemiological methods make inferences from incidence curves, which in the presence of import cases might be biased and give appearance of elevated local transmission.
+Classical epidemiological methods make inferences from incidence curves, which in the presence of import cases might be biased and give appearance of elevated local transmission.
 In contrast, phylodynamic approach can overcome such biases as it uses the transmission structure as one of its information sources~\citep{vaughanEstimatesEarlyOutbreakspecific2024}.
@@
 Examples of the MTBD family models include the birth-death exposed-infectious (BDEI) model, which was designed for pathogens featuring an incubation period between the moments of infection and of becoming infectious (e.g.
-Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen more than the others~\citep{Stadler2014}.
+Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen (e.g.
+HIV) more than the others~\citep{Stadler2014}.
 
@@
 
-%A general MTBD model describes $m$ possible individual states.
-An individual in state $k \in 1:m$ can be removed (i.e., become non-infectious, e.g., due to healing, starting a treatment, self-isolation, or death) at a rate $\psi_k$, change their state to a state $l \in 1:m$ at a rate $\mu_{kl}$ (where $\mu_{kk} = 0$), and transmit their pathogen to an individual in a state $l \in 1:m$ at rate $\lambda_{kl}$.
-Upon $i$'s removal, their pathogen can be sampled (and hence observed as a tip in the transmission tree) with a probability $\rho$.
-
-This model can be described with master equations representing the likelihood densities of evolving as on the observed transmission tree, with time $t$ going forward from the root ($t=0$) till the time of the last sampled tip ($t=T$).
-These likelihood densities depend on the probability $U(t)$ of an unobserved transmission tree that started from one individual at time $t$ and evolved till $T$, without them or any of their induced cases being sampled:
-
-\begin{equation}
-\scriptsize
-\begin{cases}
-\dot{U}(t) = &\Big(\lambda + \psi\Big) U(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-    &- \lambda U^2(t) \;  \textit{\color{gray} $\leftarrow$ transmission, followed by unsampled evolutions of both subtrees}\\
-    &- \psi (1 - \rho)\;  \textit{\color{gray} $\leftarrow$ removal without sampling}\\
-U(T) = & 1\;  \textit{\color{gray} $\leftarrow$ the probability to stay unsampled over time 0 is 1} \label{eq:Us}
-\end{cases}
-\end{equation}
-
-With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch that finishes at time $t_i$, starting on this branch at a time $t \leq t_i$.
-The master equations for the BD models were initially developed by ~\citet{Stadler2009}, however what we present here, is their branch-specific formulation that we proposed in~\citep{zhukovaFastAccurateMaximumLikelihood2022}:
-
-\begin{equation}
-\scriptsize
-\begin{cases}
-\dot{p}^{(i)}(t) = & \Big(\lambda + \psi\Big) p^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-    & - 2 \lambda p^{(i)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where one of the subtrees stayed unsampled}\\
-p^{(i)}(t_i) =  &1\;  \textit{\color{gray} $\leftarrow$ the probability of no change over zero time is 1}
-\end{cases}\label{eq:p}
-\end{equation}
-
-Using this equation, we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ as:
-
-\begin{equation}
-\scriptsize
-\begin{split}
-log L(\mathscr{T}|\Theta) =  &\sum\limits_{i \in tips}  log(\psi \rho)  \textit{\color{gray} $\leftarrow$ sampling of $n$ tips} \\
- &+\sum\limits_{i \in \stackanchor{\tiny\textit{internal}}{\tiny\textit{nodes}}} log \big(2 \lambda p^{(i0)}(t_i)p^{(i1)}(t_i)\big)   \textit{\color{gray} $\leftarrow$ $n - 1$ transmission events, followed by child }\\
- & ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\textit{\color{gray} branch evolutions (each can be a donor)}  \label{eq:mtbd_tree_likelihood}
-\end{split}
-\end{equation}
-
+Under this model we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ (see \nameref{math}).
 The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g.
@@
 
-Importantly, the BD model is asymptotically unidentifiable (see Remark 3.4 in~\citep{Stadler2009}).
-To become identifiable it requires one of its parameters to be fixed.
-In practice, it is often the sampling probability, as it may be approximated from epidemiological data $\rho$ (e.g., the proportion of sampled cases among the declared ones) or the infectious time $\frac{1}{\psi}$ (estimated from observations of infected cases).
+%A general MTBD model describes $m$ possible individual states.
+An individual in state $k \in 1:m$ can be removed (i.e., become non-infectious, e.g., due to healing, starting a treatment, self-isolation, or death) at a rate $\psi_k$, change their state to a state $l \in 1:m$ at a rate $\mu_{kl}$ (where $\mu_{kk} = 0$), and transmit their pathogen to an individual in a state $l \in 1:m$ at rate $\lambda_{kl}$.
+Upon $i$'s removal, their pathogen can be sampled (and hence observed as a tip in the transmission tree) with a probability $\rho$.
 
@@
 A general MTBD model with $m$ individual states has $2m^2 + m$ parameters: An individual in state $k \in \{1, \ldots, m\}$ can be removed at a constant average rate $\psi_k$ (with pathogen sampling probability $\rho_k$), change their state to state $l \in \{1, \ldots, m\}$ at a constant average rate $\mu_{kl}$ (where $\mu_{kk} = 0$), and transmit their pathogen to an individual in state $l$ at a constant average rate $\lambda_{kl}$.
-The time between events of the same type is hence modeled with exponential distribution.
+The time between events % of the same type
+is hence modeled with exponential distribution.
 
@@
 
-We now propose a non-Markovian extension of the BD model that adds partner notification (PN).
-In BD-PN, at the moment of sampling the sampled individual might notify their most recent partner with a constant probability $\rho_n$.
-Upon notification, the partner is removed almost instantaneously (modeled via a constant notified removal rate $\psi_p >> \psi$).
+In this section we propose an %non-Markovian 
+extension of the MTBD models that adds partner notification (PN).
+In the simplest case, the BD-PN model, at the moment of sampling the sampled individual might notify their most recent partner with a constant probability $\upsilon$.
+Upon notification, the partner is removed almost instantaneously (modeled via a constant notified removal rate $\phi >> \psi$).
 BD-PN hence has 3 classical BD and 2 additional parameters:
@@
  \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
- \item $\rho_n$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
- \item $\psi_p >> \psi$ -- constant notified removal rate: a notified individual (if not yet removed via standard procedure) stops being infectious and gets observed with this rate.
+ \item $\upsilon$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
+ \item $\phi >> \psi$ -- constant notified removal rate: a notified individual (if not yet removed via standard procedure) stops being infectious and gets observed with this rate.
 This creates an observed tip in the transmission tree.
@@
 
-The -PN extension can be easily generalized from BD to MTBD models: at the moment of sampling an individual of any type can notify their last partner with the probability $\rho_n$, who will then stop transmitting and get removed at a constant removal rate $\psi_n >> \psi_k \; \forall k \in \{1, \ldots, m\}$.
-The -PN extension therefore adds two parameters to an MTBD model: $\psi_p$ and $\rho_n$.
+The -PN extension can be generalized from BD to MTBD models: at the moment of sampling, an individual of any type can notify their last partner with the probability $\upsilon$, who will then stop transmitting and get removed at a constant removal rate $\phi >> \psi_k \; \forall k \in \{1, \ldots, m\}$.
+The -PN extension therefore adds two parameters to an MTBD model: $\phi$ and $\upsilon$.
 
@@
 
-We implemented a tree simulator that generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states). We give its algorithmic details in Appendix.
-
-\subsubsection*{Code availability}
-The simulator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\citep{Huerta-Cepas2016} and NumPy package for array operations~\citep{harris_array_2020}.
-
-It is available as a command-line program and a Python 3 package via PyPi (\href{https://pypi.org/project/treesimulator}{treesimulator}), and via Docker/Singularity (\href{https://hub.docker.com/r/evolbioinfo/treesimulator/tags}{evolbioinfo/treesimulator}).
-Its source code and the installation and usage documentation are available on GitHub at \href{https://github.com/evolbioinfo/treesimulator}{github.com/evolbioinfo/treesimulator}.
+We implemented a tree simulator that generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states).
+It is Gillespie-based, generates state change, transmission and removal times, and only reconstructs the sampled parts of the tree to save memory and increase speed.
+We give its algorithmic details in Appendix.
 
 \section{PN test}~\label{sec:test}
-To access whether a give tree is generated in a presence of partner notification, we developed a non-parametric test based on cherries (two tips having a common parent).
+To access whether a given tree is generated in a presence of partner notification, we developed a non-parametric test based on cherries (two-tip clades).
 
-The intuition behind the test is that in the presence of partner notification the tree will contain more cherries (two-tip clades) whose tips are close in time.
+The intuition behind the test is that in the presence of partner notification the tree will contain more cherries whose tips are close in time.
 Indeed, in a transmission tree internal nodes correspond to transmissions while tips correspond to sampling events.
@@
 If there is no partner notification, these two sampling events are independent, while in the presence of partner notification the first sampling might trigger the second one, and they will be close in time.
-For an incomplete transmission tree (where some patients were not sampled), the only difference would be that some cherries would include hidden transmissions, and hence not correspond to the two most recent partners.
+For an incomplete transmission tree (where some infected individuals were not sampled), the only difference would be that some cherries might include hidden transmissions.
 
@@
 In the presence of partner notification (alternative hypothesis, H1), we assume that for some of the cherries (including notified partners) their branch lengths are not independent and are similar.
+Therefore, in the absence of partner notification two external branches taken randomly from the tree and two external branches belonging to the same cherry should be indistinguishable, if they all start at about the same time.
+Being close in starting time is important as the tree is censored by the end of the sampling period: $t \leq T$.
+Hence the external branches that start close to the end of the sampling period would be on average shorter (as if they are not sampled before $T$ they will stay hidden) than those that start closer to the root~\citep{mooersBranchLengthsBirth2012}.
 %We give control to the user on what should be considered ``some of the cherries'' and ``similar branch lengths'' for their data, via adjustable test parameter values.
 
-Let us assume there are N cherries in the tree.
-For each cherry $i\;(1 \leq i \leq N)$, our test calculates the difference between its tip times $d_i$.
-It then generates a reshuffled cherry $i'$: It picks 5 (adjustable parameter) cherries with the roots that are the closest in time to the root of $i$, randomly selects two tips among their tips, and calculates their time difference $d_{i'}$.
-The test then compares $d_{i'}$ to $d_{i'}$: $sign(d_{i'} - d_i) = 1$ if $d_{i'}$ is larger than $d_i$; $= -1$ if it is smaller; or $= 0$ if they are equal.
-Note that if cherry tip branch lengths are independently drawn from the same distribution, $d_{i'}$ will be smaller than $d_i$ for about half of the cherries: $S' =  \sum\limits_{i,i'=1}^N sign(d_{i'} - d_i) \approx 0$.
-However, if some cherries are ``notified'' their real tip differences should be smaller that reshuffled ones, biasing the total sum: $S' > 0$.
+Let us assume that there are N cherries in the tree.
+For each cherry $i\;(1 \leq i \leq N)$ with tips $i0$ and $i1$, we calculate the difference between its tip times $d_i = |t_{i0} - t_{i1}|$.
+We then generate a collection of reshuffled cherries by (1) ordering the cherries by their root dates; (2) for each cherry $i$ randomly selecting one of its tips $i* \in \{i0, i1\}$; (3) swapping the selected tips between the neighboring cherries: the tip $\{2k\}*$ gets swapped with the tip $\{2k-1\}*$ ($1 \leq k \leq N / 2$) as illustrated in Fig.~\ref{fig:tipswap}.
+If the total number of cherries N is odd, instead of swapping the selected tips of the last two cherries, we swap the selected tips of the last three cherries in a cycle (see Fig.~\ref{fig:tipswap}).
+We then calculate the time differences $d_{i'}$ between the tips of these reshuffles cherries.
+The test then compares $d_i$ to $d_{i'}$: $sign(d_{i'} - d_i) = 1$ if $d_{i'}$ is larger than $d_i$; $= -1$ if it is smaller; or $= 0$ if they are equal (this could happen as in real datasets dates are often truncated, e.g.
+to month or even a year).
+Note that if cherry tip branch lengths are independently drawn from the same distribution, $d_{i'}$ will be smaller than $d_i$ for about half of the cherries.
+However, if some cherries are ``notified'' their real tip differences should be systematically smaller than reshuffled ones.
+We therefore output the pvalue calculated by the sign test.
 
-The reason for which we are constructing a random cherry from the neighbouring instead of any cherries (in terms of root date), is that cherry branch lengths depend on the date of the cherry root.
+\begin{figure*}[tbhp]
+\centering 
+\includegraphics[width=0.8\textwidth]{Fig_cherryswap.png}
+\caption{Reshuffled cherry generation.
+\textbf{a.} Cherries of the original tree sorted by their root times (times of the roots of the two oldest cherries, $t_{r1}$ and $t_{r2}$, are shown on the left).
+The randomly selected tip is shown with a bold branch for each cherry (e.g.
+the tip on the right for the leftmost cherry).
+The tip sampling time differences are shown for the two leftmost cherries ($d_1$ and $d_2$).
+\textbf{b.} Reshuffled cherries, which were obtained from the original ones by swapping the selected tips between the neighbouring cherry couples (as shown with arrows in a.): the first and the second; the third and the forth.
+As the total number of cherries (7) is odd, the last three (instead of two) cherries exchanged their selected tips in a cycle.
+The reshuffled tip sampling time differences are shown for the two leftmost cherries ($d'_1$ and $d'_2$).}
+\label{fig:tipswap} 
+\end{figure*}
+
+%The reason for which we are constructing a random cherry from the neighbouring instead of any cherries (in terms of root date), is that cherry branch lengths depend on the date of the cherry root.
 As tip sampling times are bounded from above by the end of the sampling period, the closer the cherry root is to the end of the sampling period, the shorter on average its branches will be (and less time difference they will be able to have in any case).
-
-The reshuffled cherry tip difference generation is repeated 1\,000 times (adjustable parameter), hence obtaining 1\,000 array comparisons $S'^{(1)}, \ldots, S'^{(1\,000)}$.
-
-Finally, the test reports the proportion of reshuffled arrays whose comparison was negative ($S'^{(k)} < 0$). Note that in the presence of partner notification, this proportion should be very low. 
-The test therefore reports the probability of partner notification being present if the tree.
 
@@
 \subsection{Performance on simulated data}~\label{sec:test:sim}
-To illustrate the PN test performance, we applied it to the trees simulated under three classical MTBD-family models: BD, BDEI (Birth-Death Exposed-Infectious~~\citep{Stadler2014}) and BDSS (Birth-Death with SuperSpreading~\citep{Stadler2013a}, see Fig.~1 in \citet{Voznica2021} for a graphical representation of these models and their parameters), and under their -PN versions: BD-PN, BDEI-PN and BDSS-PN models.
+To illustrate the PN test performance, we applied it to the trees simulated under three classical MTBD-family models: BD, BDEI (Birth-Death Exposed-Infectious~~\citep{Stadler2014}) and BDSS (Birth-Death with SuperSpreading~\citep{Stadler2013a}), and under their -PN versions: BD-PN, BDEI-PN and BDSS-PN models.
+(See Fig.~1 in \citet{Voznica2021} for a graphical representation of BD, BDEI and DBSS models and their parameters).
 
@@
 \item sampling probability $\rho \in ]0.1, 0.9]$,
-\item \textit{(for -PN models only)} ratio between the partner removal rate after notification and the removal rate $\frac{\psi_p}{\psi} \in ]50, 500]$,
-\item \textit{(for -PN models only)} partner notification probability $\rho_n \in ]0.01/\rho, 0.9]$ (the lower bound was picked this way to ensure that the probability to be a notified partner is at least 0.01 $=\rho \rho_n$),
+\item \textit{(for -PN models only)} ratio between the partner removal rate after notification and the removal rate $\frac{\phi}{\psi} \in ]50, 500]$,
+\item \textit{(for -PN models only)} partner notification probability $\upsilon \in ]0.01/\rho, 0.9]$ (the lower bound was picked this way to ensure that the probability to be a notified partner is at least 0.01 $=\rho \upsilon$),
 \item \textit{(for BDEI models only)} becoming-infectious rate (at which an exposed individual becomes infectious) $\mu \in ]0.2, 10]$,
@@
 
-The common parameters were consistent between models, in a sense that the $i$-th tree under the BDSS model had the same $R_e$, $\psi$ and $\rho$ values as the $i$-th trees under BDSS-PN, BD(-PN) and BDEI(-PN) models; the $i$-th tree under the BDSS-PN model had the same $\psi_p$ and $\rho_n$ values as the $i$-th trees under BD-PN and BDEI-PN models.
+The common parameters were consistent between models, in a sense that the $i$-th tree under the BDSS model had the same $R_e$, $\psi$ and $\rho$ values as the $i$-th trees under BDSS-PN, BD(-PN) and BDEI(-PN) models; the $i$-th tree under the BDSS-PN model had the same $\phi$ and $\upsilon$ values as the $i$-th trees under BD-PN and BDEI-PN models.
 
 We then run the PN tests on each tree of these 6 datasets.
-For 100/99/100 out of 100 trees in the BD/BDEI/BDSS dataset the PN test p-value was above $0.05$ (i.e.
-100\%/99\%/100\% true negative and 0\%/1\%/0\% false positive results).
-For 97/99/96 out of 100 trees in the BD-PN/BDEI-PN/BDSS-PN dataset the PN test p-value was below $0.05$ (i.e.
-97\%/99\%/96\% true positive and 3\%/1\%/4\% false negative results).
+For 98/97/95 out of 100 trees in the BD/BDEI/BDSS dataset the PN test p-value was above $0.05$ (i.e.
+98\%/97\%/95\% true negative and 2\%/3\%/5\% false positive results).
+For 95/95/96 out of 100 trees in the BD-PN/BDEI-PN/BDSS-PN dataset the PN test p-value was below $0.05$ (i.e.
+95\%/95\%/96\% true positive and 5\%/5\%/4\% false negative results).
 These results are summarised in Table~\ref{tbl:pntest}.
@@
 \caption{PN test performance on trees generated under BD, BDEI, BDSS, BD-PN, BDEI-PN and BDSS-PN models (100 trees per model).
-We report the number of trees for each the PN-test p-value was $<0.05$, $\geq 0.05$, and its mean p-value for each model dataset.}
+We report the number of trees for which the PN-test p-value was $<0.05$, $\geq 0.05$, and the mean p-value for each model dataset.}
 \begin{tabular}{l|r|r|c}
@@
   \midrule
-\textbf{BD}& 0 & 100 & 0.424 \\
-\textbf{BD-PN}& 97 & 3 & 0.004 \\
+\textbf{BD}& 2 & 98 & 0.479 \\
+\textbf{BD-PN}& 95 & 5 & 0.015 \\
   \midrule
-\textbf{BDEI}& 1 & 99 & 0.493 \\
-\textbf{BDEI-PN}& 99 & 1 & 0.001 \\
+\textbf{BDEI}& 3 & 97 & 0.531 \\
+\textbf{BDEI-PN}& 95 & 5 & 0.008 \\
   \midrule
-\textbf{BDSS}& 0 & 100 & 0.455 \\
-\textbf{BD-PN}& 96 & 4 & 0.004 \\
+\textbf{BDSS}& 5 & 95 & 0.505 \\
+\textbf{BDSS-PN}& 96 & 4 & 0.008 \\
 \end{tabular}
@@
 
-\section{Mathematical formulation of BD-PN model}
+\section{Mathematical formulation of BD and BD-PN models}\label{math}
+
+The BD model can be described with master equations representing the likelihood densities of evolving as on the observed transmission tree, with time $t$ going forward from the root ($t=0$) till the time of the last sampled tip ($t=T$).
+These likelihood densities depend on the probability $U(t)$ of an unobserved transmission tree that started from one individual at time $t$ and evolved till $T$, without this individual nor any of their induced cases being sampled:
+
+\begin{equation}
+\scriptsize
+\begin{cases}
+\dot{U}(t) = &\Big(\lambda + \psi\Big) U(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+    &- \lambda U^2(t) \;  \textit{\color{gray} $\leftarrow$ transmission, followed by unsampled evolutions of both subtrees}\\
+    &- \psi (1 - \rho)\;  \textit{\color{gray} $\leftarrow$ removal without sampling}\\
+U(T) = & 1\;  \textit{\color{gray} $\leftarrow$ the probability to stay unsampled over time 0 is 1} \label{eq:Us}
+\end{cases}
+\end{equation}
+
+With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch that finishes at time $t_i$, starting on this branch at a time $t \leq t_i$.
+The master equations for the BD models were initially developed by ~\citet{Stadler2009}, however what we present here is their branch-specific formulation that we proposed in~\citep{zhukovaFastAccurateMaximumLikelihood2022}:
+In our formulation, ${p}^{(i)}(t)$ describes an evolution along a tree branch $i$, without taking into account the branch's subtree.
+% and provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$.
+These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.
+
+\begin{equation}
+\scriptsize
+\begin{cases}
+\dot{p}^{(i)}(t) = & \Big(\lambda + \psi\Big) p^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+    & - 2 \lambda p^{(i)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where one of the subtrees stayed unsampled}\\
+p^{(i)}(t_i) =  &1\;  \textit{\color{gray} $\leftarrow$ the probability of no change over zero time is 1}
+\end{cases}\label{eq:p}
+\end{equation}
+
+Using this equation, we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ as:
+
+\begin{equation}
+\scriptsize
+\begin{split}
+log L(\mathscr{T}|\Theta) =  &\sum\limits_{i \in tips}  log(\psi \rho)  \textit{\color{gray} $\leftarrow$ sampling of tips} \\
+ &+\sum\limits_{i \in \stackanchor{\tiny\textit{internal}}{\tiny\textit{nodes}}} log \big(2 \lambda p^{(i0)}(t_i)p^{(i1)}(t_i)\big)   \textit{\color{gray} $\leftarrow$ transmission events, followed by child }\\
+ & ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\textit{\color{gray} branch evolutions (each can be a donor)}  \label{eq:mtbd_tree_likelihood}
+\end{split}
+\end{equation}
+
+Importantly, the BD model is asymptotically unidentifiable (see Remark 3.4 in~\citep{Stadler2009}).
+To become identifiable it requires one of its parameters to be fixed.
+In practice, it is often the sampling probability $\rho$, as it may be approximated from epidemiological data (e.g., the proportion of sampled cases among the declared ones) or the infectious time $\frac{1}{\psi}$ (estimated from observations of infected cases).
+
 In the classical BD model with incomplete sampling, a branch of the tree corresponds to a virus evolution between the transmission corresponding to the node at the beginning of this branch and either a sampling event (if the branch is external) or a transmission corresponding to the node at the end of the branch (if it is internal).
@@
 According to our model assumption 2, notified individuals are always observed.
-Hence B must be sampled too (at time $t_B \geq t_A$) and correspond to a tree tip, while the common ancestor of A and B (AB) corresponds to the transmission between A and B.
+Hence B must be sampled too (at time $t_B \geq t_A$) and corresponds to a tree tip, while the common ancestor of A and B (AB) corresponds to the transmission between A and B.
 Therefore the individual on the path between AB and B cannot change (is always B) and only hidden transmissions (from B to someone else) where the recipient subtree stayed unobserved are possible along the branch(es) leading from AB to B (before $t_A$).
@@
 
-We will use an \textbf{oriented probability} $p^{(i,o)}(t)$ for cases when the individual at the end of the branch (or considered time interval) is the same as at time $t$ but might have transmitted to someone else (unobserved) along the way:
+We will use an \textbf{oriented (partner) branch probability} $p_o^{(i)}(t)$ for cases when the individual at the end of the branch (or considered time interval) is the same as at time $t$ but might have transmitted to someone else (unobserved) along the way:
 
@@
 \begin{cases}
-\dot{p}^{(i,o)}(t) = & \Big(\lambda + \psi\Big) p^{(i,o)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-    & - \lambda p^{(i,o)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
-p^{(i,o)}(t_i) = 1\label{eq:p-o}
+\dot{p}_o^{(i)}(t) &=~~  \Big(\lambda + \psi\Big) p_o^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+    &~~~ - \lambda p_o^{(i)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
+p_o^{(i)}(t_i) &=  1\label{eq:p-o}
 \end{cases}
@@
 
-We will use an \textbf{partner probability once notified} $p^{(i,np)}(t)$ for cases when the individual at the end of the branch is the partner who got notified at time $t_n < t$:
+We will use a \textbf{notified partner branch probability} $p_p^{(i)}(t)$ for cases when the individual at the end of the branch is the partner who got notified at time $t_n < t$:
 
@@
 \begin{cases}
-\dot{p}^{(i,np)}(t) = & \psi_p p^{(i,nt)}(t)\; \textit{\color{gray} $\leftarrow$ no sampling in the next infinitesimal time $\Delta t$ }\\
-p^{(i,o)}(t_i) = 1\label{eq:p-np}
+\dot{p}_p^{(i)}(t) &=  \phi p_p^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no sampling in the next infinitesimal time $\Delta t$ }\\
+p_p^{(i)}(t_i) &=  1\label{eq:p-np}
 \end{cases}
@@
 Moreover, according to our model assumption 4, only the most recent partner can get notified.
-Hence, AB must be the parent node of A, and no hidden tranmission is possible along the branch AB-A (as otherwise there would have been a partner of A who is more recent than B).
+Hence, AB must be the parent node of A, and no hidden transmission is possible along the branch AB-A (as otherwise there would have been a partner of A who is more recent than B).
 
-We will use a \textbf{non-hidden probability} $p^{(i,nh)}(t)$ for such cases (when no hidden transmission occurred along the branch):
+We will use a \textbf{notifier branch probability} $p_n^{(i)}(t)$ for such cases (when no hidden transmission occurred along the branch):
 
@@
 \begin{cases}
-\dot{p}^{(i,nh)}(t) = & \Big(\lambda + \psi\Big) p^{(i,nh)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-p^{(i,nh)}(t_i) = 1\label{eq:p-nh}
+\dot{p}_n^{(i)}(t) &=  \Big(\lambda + \psi\Big) p_n^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+p_n^{(i)}(t_i) &= 1\label{eq:p-nh}
 \end{cases}
@@
 
-Note that A could have also ``unsuccessfully'' notified B, if by the time $t_A$, when A decided to notify B, B already got sampled via the standard procedure or was notified by someone else.
+Note that A could have also ``unsuccessfully'' notified B, if by the time $t_A$, when A decided to notify B, B had already got sampled via the standard procedure or got notified by someone else.
 In this case nothing changes for the AB-A branch, and the AB-B branch(es) stay oriented.
@@
 
-%For the BD-PN model we will also need to consider the case where the individual at the end of the branch is the same as at time $t$.
-This corresponds to a notified partner branch, as according to our model assumptions if a partner is notified, he or she is sampled (i.e.
-observed in our tree): Hence the individual at the beginning of the branch (internal node corresponding to transmission) should be the same as at its end (sampled partner if it finishes with a tip, or a transmission where the partner is the donor).
-%
-%
-%We will use an \textbf{oriented probability} $p^{(i,o)}(t)$ for the cases when the individual at the end of the branch is the same as at time $t$:
-%
-%\begin{equation}
-%\scriptsize
-%\begin{cases}
-%\dot{p}^{(i,o)}(t) = & \Big(\lambda + \psi\Big) p^{(i,o)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-%    & - \lambda p^{(i,o)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
-%p^{(i,o)}(t_i) = 1\label{eq:p-o}
-%\end{cases}
-%\end{equation}
-%
-%
-%Additionally we define a \textbf{non-hidden probability} $p^{(i,nh)}(t)$ for the cases when no hidden transmission occurred along the branch:
-%
-%\begin{equation}
-%\scriptsize
-%\begin{cases}
-%\dot{p}^{(i,nh)}(t) = & \Big(\lambda + \psi\Big) p^{(i,nh)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-%p^{(i,nh)}(t_i) = 1\label{eq:p-nh}
-%\end{cases}
-%\end{equation}
-
 The equations~\ref{eq:Us}, \ref{eq:p}, \ref{eq:p-o}-\ref{eq:p-nh} have the following closed form solutions:
@@
 \\
-p^{(i,o)}(t) =  \frac{E_2 + 1}{E_1 + 1}\sqrt{E_3E_4} \\
+p_o^{(i)}(t) =  \frac{E_2 + 1}{E_1 + 1}\sqrt{E_3E_4} \\
 \\
-p^{(i,nh)}(t) =  E_4
+p_n^{(i)}(t) =  E_4
 \\
-p^{(i,np)}(t) =  e^{\psi_p (t -t_i)}
+p_p^{(i)}(t) =  e^{\phi (t -t_i)}
 \end{cases},\\
@@
 
-%Under the BD-PN model, the probability density $q(i)$ of a branch connecting a node $i$ to its parent node $j$ (without taking into account the event at the end of the branch) can be written as:
-%
-%
-%\begin{equation}
-%\scriptsize
-%q(i) = 
-%\begin{cases}
-%q^{(u)}(i) = p^{(i)}(t_j) \textit{\color{gray} ~if $i$ is an \textbf{unnotified non-notifier}}\\
-%q^{(n)}(i) = 
-%p^{(i,nh)}(t_j) \textit{\color{gray} ~~if $i$ is a tip corresponding to a \textbf{notifier}}\\
-%~~~~~~~~~~~~ \textit{\color{gray}(parent is the last transmission, no hidden transmission along the branch)}
-%\\
-%q^{(p)}(i)  = p^{(i, o)}(t_j)  \textit{\color{gray} ~if $i$ is an internal or tip node whose individual will be notified after $t_i$}
-%\\
-%q^{(p)}(i, r)  = p^{(r, o)}(t_j) e^{- \psi_p (t_i - t_r)} \textit{\color{gray} if $i$ is a \textbf{partner} tip notified by $r$ at time $t_r \leq t_i$}
-%\end{cases}
-%\label{eq:branch}
-%\end{equation}
+\subsection{BD-PN tree likelihood density}
 
-Using the formulas~(\ref{eq:ps}), we can calculate the likelihood density of a tree using a pruning algorithm~\citep{10.1093/sysbio/22.3.240}, where for each visited node $i$ we calculate the likelihood density of its subtree if the branch connecting it to its parent node $j$ corresponds to (1) an unnotified individual $l^{(u)}(i)$; or (2) an (eventually) notified partner $l^{(p)}(i, r)$ (notified by the individual corresponding to the tip $r$ at time $t_r$).
-For each visited tip we additionally consider whether the corresponding individual notified their most recent partner or not.
+Using the formulas~(\ref{eq:ps}), we can calculate the likelihood density of a tree using a pruning algorithm~\citep{10.1093/sysbio/22.3.240}.
+For each visited node $i$ we calculate the likelihood density of its subtree for two cases depending on their receiving-a-notification status (shown with the first superscript, $^{\not{n}}$ or $^n$).
+In the first case,  $l^{(\not{n})}(i)$, the branch $ji$ connecting $i$ it to its parent node $j$ corresponds to an unnotified individual.
+In the second case, $l^{(n)}(i, r)$, the branch $ji$ corresponds to an (eventually) notified partner (notified by the individual corresponding to a tip $r$ at time $t_r$).
+If $i$ is a tip, we additionally consider whether the corresponding individual notified their most recent partner or not (shown with the second superscript, $^n$ or $^{\not{n}}$).
+For instance, $l^{(\not{n},n)}(i)$ represents the case where the tip $i$ was not notified but notified their partner.
+Note that if the tip $i$ was notified by their partner $r$, we need to consider two possibilities: (1) when $t_r \leq t_i$, and hence the notification was successful at provoking $i$'s diagnostics and sampling; and (2) when $t_r > t_i$, and hence by the time of notification by $r$ $i$ already got sampled by other means.
 
-%For a tip $i$ in state $si$ these values can be calculated as:
-%\begin{equation}
-%\scriptsize
-%\begin{array}{ll}
-%l^{(u)}(i) &= \begin{cases} l^{(u,n)}(i) = q^{(u)}(i)\psi\rho
-%\rho_{n} \textit{\color{gray} ~~~~~~~~if $i$ notified their most recent patner}\\
-%l^{(u,\not{n})}(i) = q^{(u)}(i)\psi\rho (1-\rho_{n}) \textit{\color{gray} ~if $i$ did not notify}
-%\end{cases}\\
-%l^{(p)}(i, r) &= q^{(p)}(i,r)\psi_p\cdot \begin{cases}
-%\rho_{n} \textit{\color{gray} ~~~~~~~~if $i$ notified their most recent patner}\\
-%(1-\rho_{n}) \textit{\color{gray} ~if $i$ did not notify}
-%\end{cases}\\
-%\end{array}
-%\label{eq:tip}
-%\end{equation}
+The tree $\mathscr{T}$ likelihood density $\mathscr{L}(\mathscr{T}|\Theta)$ under BD-PN model with parameters $\Theta=\{\lambda, \psi, \rho, \phi, \upsilon\}$ can be calculated as $l^{(\not{n})}(root)$, as the root can only correspond to an unnotified individual (since there was no observed transmission before it).
 
-For a tip $i$ with a parent node $j$ these values can be calculated as:
+We will now describe the subtree likelihood density calculation case by case.
+
+\subsubsection*{Case 1: $i$ is a tip}
+
+For a tip $i$ with a parent node $j$ the subtree likelihood density values can be calculated as:
 \begin{equation}
@@
 \begin{array}{ll}
-l^{(u,n)}(i) &= p^{(i,nh)}(t_j)\psi\rho
-\rho_{n} \textit{\color{gray} ~if $i$ was not notified but notified their most recent patner}\\
-l^{(u,\not{n})}(i) &= p^{i}(t_j)\psi\rho (1-\rho_{n}) \textit{\color{gray} ~if $i$ was not notified and did not notify}\\
-l^{(p,n)}(i,r) &= \begin{cases}
-p^{(i, nh)}(t_j)\psi\rho\rho_{n} \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and notified their most recent patner}\\
-p^{(r, nh)}(t_j) e^{- \psi_p (t_i - t_r)} \psi_p \rho_{n} \textit{\color{gray} ~if $i$ got notified at time $t_r \leq t_i$ and notified}
+l^{(\not{n},\not{n})}(i) &= p^{(i)}(t_j)\psi\rho (1-\upsilon) \textit{\color{gray} ~i.e. $i$ was not notified and did not notify}\\
+& \textit{\color{gray} (standard branch evolution followed by sampling and no notification)}\\
+\\
+l^{(\not{n},n)}(i) &= p_n^{(i)}(t_j)\psi\rho
+\upsilon \textit{\color{gray} ~i.e. $i$ was not notified but notified their most recent patner}\\
+& \textit{\color{gray} (notifier branch evolution followed by sampling and notification)}\\
+\\
+l^{(n,\not{n})}(i,r) &= \begin{cases}
+0 \textit{\color{gray} ~if $t_r < t_j$}\\
+\textit{\color{gray} (not possible as once notified partners do not transmit)}\\
+\\
+p_o^{(i)}(t_j)\psi\rho(1 - \upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and did not notify}\\
+\textit{\color{gray} (oriented (partner) branch evolution followed by standard sampling and no notification)}\\
+\\
+p_o^{(r)}(t_j) p_p^{(i)}(t_r) \phi (1-\upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_j \leq t_r \leq t_i$ and did not notify}\\
+\textit{\color{gray} (oriented (partner) branch evolution followed by} \\
+\textit{\color{gray}~notified partner branch evolution,} \\
+\textit{\color{gray}~notification-provoked sampling and no notification)}\\
 \end{cases}\\
-l^{(p,\not{n})}(i,r) &= \begin{cases}
-p^{(i, o)}(t_j)\psi\rho(1 - \rho_{n}) \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and did not notify}\\
-p^{(r, o)}(t_j) e^{- \psi_p (t_i - t_r)} \psi_p (1-\rho_{n}) \textit{\color{gray} ~if $i$ got notified at time $t_r \leq t_i$ and did not notify}
+\\
+l^{(n,n)}(i,r) &= \begin{cases}0 \textit{\color{gray} ~if $t_r < t_j$}\\
+\textit{\color{gray} (not possible as once notified partners do not transmit)}\\
+\\
+p_n^{(i)}(t_j)\psi\rho\upsilon \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and notified their most recent patner}\\
+\textit{\color{gray} (notifier branch evolution followed by standard sampling and  notification)}\\
+\\
+p_n^{(r)}(t_j) p_p^{(i)}(t_r) \phi \upsilon \textit{\color{gray} ~if $i$ got notified at time $t_j \leq t_r \leq t_i$ and notified}\\
+\textit{\color{gray} (notifier branch evolution followed by} \\
+\textit{\color{gray}~notified partner branch evolution,} \\
+\textit{\color{gray}~notification-provoked sampling and notification)}\\
 \end{cases}\\
@@
 
+\subsubsection*{Case 2: $i$ is an internal node} 
 Let us now consider an internal node $i$ with two child nodes, $i_0$ and $i_1$.
+
+\subsubsection*{Case 2.1: $i$ is an unnotified internal node} 
 We will start with the configuration where the $i$'s branch corresponds to an unnotified individual.
-We will consider three possibilities: (i) when both $i_0$ and $i_1$ are internal nodes; (ii) when both $i_0$ and $i_1$ are tips; and (iii) when one of them is an internal node and the other one is a tip.
+Hence the $i$'s branch evolution is represented by $p^{(i)}(t_j)$ with a transmission event at the end of it, where each of the child branches can correspond to the donor (probability density of $2\lambda$).
+We will consider three possibilities: (1) when both $i_0$ and $i_1$ are internal nodes; (2) when both $i_0$ and $i_1$ are tips; and (3) when one of them is an internal node and the other one is a tip.
 
@@
 
-First, let's assume that (i) $i_0$ and $i_1$ are internal. Then none of them can be a notifier (since notifiers correspond to tips), and hence none of them is notified:
+\subsubsection*{Case 2.1.1: $i$ is an unnotified internal node with internal-node children $i0$ and $i1$}
+
+First, let's assume that $i_0$ and $i_1$ are internal. Then none of them can be a notifier (since notifiers correspond to tips), and hence none of them is notified (as $i$ is also unnotified):
 
@@
 \scriptsize
-l^{(u)}(i) = p^{i}(t_j) 2\lambda l^{(u)}(i_0)l^{(u)}(i_1) \label{eq:lu-int-int}
+l^{(\not{n})}(i) = p^{(i)}(t_j) 2\lambda l^{(\not{n})}(i_0)l^{(\not{n})}(i_1) \label{eq:lu-int-int}
 \end{equation}
 
-Secondly, let's assume that (ii) $i_0$ and $i_1$ are tips. Then each of them could have notified the other one:
+\subsubsection*{Case 2.1.2: $i$ is an unnotified internal node with tip children $i0$ and $i1$}
+
+Secondly, let's assume that $i_0$ and $i_1$ are tips. Then each of them could have notified the other one:
 
@@
 \begin{split}
-l^{(u)}(i) = p^{i}(t_j) 2\lambda \cdot 
-\Big(&l^{(u,\not{n})}(i_0)l^{(u,\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ neigher $i_0$ nor $i_1$ notified} \\
-&+ l^{(u,n)}(i_0)l^{(p,\not{n})}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $i_0$ notified $i_1$ and $i_1$ did not notify} \\
-&+ l^{(u,n)}(i_1)l^{(p,\not{n})}(i_0,i_1) \textit{\color{gray}~$\leftarrow$ $i_1$ notified $i_0$ and $i_0$ did not notify} \\
-&+ l^{(p,n)}(i_0,i1)l^{(p,n)}(i_1,i_0)\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ and $i_0$ notified each other} \\
+l^{(\not{n})}(i) = p^{i}(t_j) 2\lambda \cdot 
+\Big(&~~~~l^{(\not{n},\not{n})}(i_0)l^{(\not{n},\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ neigher $i_0$ nor $i_1$ notified} \\
+&+ l^{(\not{n},n)}(i_0)l^{(n,\not{n})}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $i_0$ notified $i_1$ and $i_1$ did not notify} \\
+&+ l^{(n,\not{n})}(i_0,i_1)l^{(\not{n},n)}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ did not notify, but $i_1$ notified $i_0$} \\
+&+ l^{(n,n)}(i_0,i1)l^{(n,n)}(i_1,i_0)~~\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ and $i_0$ notified each other} \\
 \end{split}\label{eq:lu-tip-tip}
@@
 
-Lastly, let's assume that (iii) $i_0$ is a tip and $i_1$ is internal (the scenario where $i_1$ is a tip and $i_0$ is internal can be obtained by swapping the labels).
-Then $i_0$ could have notified $i_1$:
+\subsubsection*{Case 2.1.3: $i$ is an unnotified internal node whose children are a tip and an internal node}
+Lastly, let's assume that $i_0$ is a tip and $i_1$ is internal (the scenario where $i_1$ is a tip and $i_0$ is internal can be obtained by swapping the labels). Then $i_0$ could have notified $i_1$:
 
@@
 \begin{split}
-l^{(u)}(i) = p^{i}(t_j) 2\lambda \cdot 
-\Big(&l^{(u,\not{n})}(i_0)l^{(u)}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ did not notify $i_1$ } \\
-&+ l^{(u,n)}(i_0)l^{(p)}(i_1,i_0)\Big) \textit{\color{gray}~$\leftarrow$ $i_0$ notified $i_1$} \\
+l^{(\not{n})}(i) = p^{i}(t_j) 2\lambda \cdot 
+\Big(&~~~l^{(\not{n},\not{n})}(i_0)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ did not notify $i_1$ } \\
+&+ l^{(\not{n},n)}(i_0)l^{(n)}(i_1,i_0)~~\Big) \textit{\color{gray}~$\leftarrow$ $i_0$ notified $i_1$} \\
 \end{split}\label{eq:lu-int-tip}
@@
 
-%\begin{equation}
-%\scriptsize
-%l^{(u)}(i) = p^{i}(t_j) 2\lambda \\
-%\cdot \begin{cases}
-%l^{(u)}(i_0)l^{(u)}(i_1) \textit{\color{gray} ~if both $i_0$ and $i_1$ are internal} \\
-%l^{(u,\not{n})}(i_0)l^{(u)}(i_1) + l^{(u,n)}(i_0)l^{(p)}(i_1,i_0) \textit{\color{gray}~if $i_0$ is a tip and $i_1$ is internal} \\
-%l^{(u,\not{n})}(i_1)l^{(u)}(i_0) + l^{(u,n)}(i_1)l^{(p)}(i_0,i_1) \textit{\color{gray}~if $i_1$ is a tip and $i_0$ is internal} \\
-%l^{(u,\not{n})}(i_0)l^{(u,\not{n})}(i_1) \\
-%~~~~+ l^{(u,n)}(i_0)l^{(p,\not{n})}(i_1,i_0) + l^{(u,n)}(i_1)l^{(p,\not{n})}(i_0,i_1)\\
-%~~~~+ l^{(p,n)}(i_0,i1)l^{(p,n)}(i_1,i_0) \textit{\color{gray}~if both $i_0$ and $i_1$ are tips} \\
-%\end{cases}
-%%& + is\_tip(i_0)l^{(u,n)}(i_0)l^{(p)}(i_1,i_0)\textit{\color{gray} $\leftarrow$ notification of someone in $i_1$'s subtree by $i_0$} \\
-%%&+ is\_tip(i_1)l^{(u,n)}(i_1)l^{(p)}(i_0,i_1)~\Big) \textit{\color{gray} $\leftarrow$ notification of someone in $i_0$'s subtree by $i_1$},\\
-%%\textit{where } & is\_tip(r) = 1 \textit{ if r is a tip and $= 0$ if r is an internal node.}\label{eq:lu}
-%\end{equation}
+\subsubsection*{Case 2.2: $i$ is an (eventually notified) partner internal node}
 
-In the other case an internal node $i$ corresponds to a partner who will eventually get notified by a node $r$.
-In this case the individual on the path connecting this branch to the notified tip stays the same (the partner).
+In the other case an internal node $i$ corresponds to a partner who will eventually get notified by a tip $r$.
+In this case the individual on the path connecting $i$ to its common ancestor $ir$ with the notifier tip $r$ (corresponding to the transmission between $r$ and $i$) stays the same (the partner).
 Hence, all the transmissions on this path are oriented and correspond to the partner transmitting to another recipient individual.
+In particular, the $i$'s branch evolution is represented by $p_o^{(i)}(t_j)$ with a transmission event at the end of it, where the donor is known (probability density of $\lambda$): The donor corresponds to the subtree containing the tip representing the sampling of the individual $i$.
 %Moreover, this recipient's branch cannot be notified by our partner (as in our model notified partners do not notify further):
-Again, we will consider three possibilities for $i$'s child nodes: (i) when both $i_0$ and $i_1$ are internal nodes; (ii) when both $i_0$ and $i_1$ are tips; and (iii) when one of them is an internal node and the other one is a tip.
+Again, we will consider three possibilities for $i$'s child nodes: (1) when both $i_0$ and $i_1$ are internal nodes; (2) when both $i_0$ and $i_1$ are tips; and (3) when one of them is an internal node and the other one is a tip.
 
@@
 \scriptsize
-l^{(p)}(i, r) = 0 \textit{~if $i$ is internal and $t_i > t_r$ or $i$ is a tip and $t_{parent(i)} > t_r$} \\
+l^{(n)}(i, r) = 0 \textit{\color{gray}~if $i$ is internal and $t_i > t_r$ or $i$ is a tip and $t_{parent(i)} > t_r$} \\
 \label{eq:lp-zero}
@@
 
-First, let's assume that (i) $i_0$ and $i_1$ are internal. Then none of them can be a notifier (since notifiers correspond to tips), and hence exactly one of them is notified (by $r$):
+\subsubsection*{Case 2.2.1: $i$ is a partner internal node with internal-node children $i0$ and $i1$}
+
+First, let's assume that $i_0$ and $i_1$ are internal. Then none of them can be a notifier (since notifiers correspond to tips), and hence exactly one of them is notified (by $r$):
 
@@
 \begin{split}
-l^{(p)}(i, r) = p^{(i,o)}(t_j) \lambda \cdot
-\Big(&l^{(p)}(i_0,r)l^{(u)}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ is notified by $r$} \\
-& + l^{(p)}(i_1,r)l^{(u)}(i_0)\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ is notified by $r$} \\
+l^{(n)}(i, r) = p_o^{(i)}(t_j) \lambda \cdot
+\Big(&~~~l^{(n)}(i_0,r)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ is notified by $r$} \\
+& + l^{(\not{n})}(i_0)l^{(n)}(i_1,r)~~\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ is notified by $r$} \\
  \end{split}
@@
 
-Secondly, let's assume that (ii) $i_0$ and $i_1$ are tips. Then each of them could have notified the other one (in addition to one of them being notified by $r$):
+\subsubsection*{Case 2.2.2: $i$ is a partner internal node with tip children $i0$ and $i1$} 
+Secondly, let's assume that $i_0$ and $i_1$ are tips. Then each of them could have notified the other one (in addition to one of them being notified by $r$):
 
@@
 \begin{split}
-l^{(p)}(i, r) = p^{(i,o)}(t_j) \lambda \cdot
-\Big(&l^{(p,\not{n})}(i_0,first(r,i_1))l^{(u,n)}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by both $r$ and $i_1$, but did not notify $i_1$} \\
-&+ l^{(p,n)}(i_0,first(r,i_1))l^{(p,n)}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by both $r$ and $i_1$, and notified $i_1$} \\
-&+l^{(p,\not{n})}(i_1,first(r,i_0))l^{(u,n)}(i_0) \textit{\color{gray}~$\leftarrow$ $i_1$ got notified by both $r$ and $i_0$, but did not notify $i_0$} \\
-&+ l^{(p,n)}(i_1,first(r,i_0))l^{(p,n)}(i_0, i_1) \textit{\color{gray}~$\leftarrow$ $i_1$ got notified by both $r$ and $i_0$, and notified $i_0$} \\
-&+l^{(p,n)}(i_0,r)l^{(p,\not{n})}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by $r$, and notified $i_1$} \\
-& + l^{(p,n)}(i_1,r)l^{(p,\not{n})}(i_0, i_1) \textit{\color{gray}~$\leftarrow$ $i_1$ got notified by $r$, and notified $i_0$} \\
-&+l^{(p,\not{n})}(i_0,r)l^{(u,\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by $r$, and did not notify $i_1$} \\
-& + l^{(p,\not{n})}(i_1,r)l^{(u,\not{n})}(i_0)\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ got notified by $r$, and did not notify $i_0$} \\
+l^{(n)}(i, r) = p_o^{(i)}(t_j) \lambda \cdot
+\Big(~~~&l^{(n,\not{n})}(i_0,first(r,i_1))l^{(\not{n},n)}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ and $i_1$ notified $i_0$, $i_0$ did not notify} \\
+&+ l^{(n,n)}(i_0,first(r,i_1))l^{(n,n)}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $r$ and $i_1$ notified $i_0$, $i_0$ notified $i_1$} \\
+&+l^{(\not{n},n)}(i_0)l^{(n,\not{n})}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ $r$ and $i_0$ notified $i_1$, $i_1$ did not notify} \\
+&+ l^{(n,n)}(i_0, i_1)l^{(n,n)}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ $r$ and $i_0$ notified $i_1$, $i_1$ notified $i_0$} \\
+&+l^{(n,n)}(i_0,r)l^{(n,\not{n})}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$ notified $i_1$, $i_1$ did not notify} \\
+& + l^{(n,\not{n})}(i_0, i_1)l^{(n,n)}(i_1,r) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_1$ notified $i_0$, $i_0$ did not notify} \\
+&+l^{(n,\not{n})}(i_0,r)l^{(\not{n},\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$ and $i_1$ did not notify} \\
+& + l^{(\not{n},\not{n})}(i_0)l^{(n,\not{n})}(i_1,r)~~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_0$ and $i_1$ did not notify}, \\
 \textit{\color{gray}~where } &\color{gray}first(a,b) = 
@@
 
-Lastly, let's assume that (iii) $i_0$ is a tip and $i_1$ is internal (the scenario where $i_1$ is a tip and $i_0$ is internal can be obtained by swapping the labels).
+\subsubsection*{Case 2.2.3: $i$ is a partner internal node whose children are a tip and an internal node}
+Lastly, let's assume that $i_0$ is a tip and $i_1$ is internal (the scenario where $i_1$ is a tip and $i_0$ is internal can be obtained by swapping the labels).
 Then $i_0$ could have notified $i_1$ (in addition to one of them being notified by $r$):
@@
 \begin{split}
-l^{(p)}(i, r) = p^{(i,o)}(t_j) \lambda \cdot
-\Big(&l^{(p,n)}(i_0,r)l^{(p)}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by $r$, and notified $i_1$} \\
-& + l^{(p,\not{n})}(i_0,r)l^{(u)}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by $r$, and did not notify $i_1$} \\
-& +l^{(p)}(i_1,first(r,i_0))l^{(u, n)}(i_0) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by both $r$ and $i_0$} \\
-& + l^{(p)}(i_1,r)l^{(u, \not{n})}(i_0)\Big) \textit{\color{gray}~$\leftarrow$ $i_0$ got notified by $r$} \\
+l^{(n)}(i, r) = p_o^{(i)}(t_j) \lambda \cdot
+\Big(~~~&l^{(n,n)}(i_0,r)l^{(n)}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$  notified $i_1$} \\
+& + l^{(n,\not{n})}(i_0,r)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$  $r$ notified $i_0$, $i_0$ did not notify} \\
+& +l^{(\not{n}, n)}(i_0)l^{(n)}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ both $r$ and $i_0$ notified $i_1$} \\
+& + l^{(\not{n},\not{n})}(i_0)l^{(n)}(i_1,r)~~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_0$ did not notify}, \\
 \textit{\color{gray}~where } &\color{gray}first(a,b) = 
@@
 
-%\begin{equation}
-%\scriptsize
-%l^{(p)}(i, r) = p^{(i,o)}(t_j) \lambda
-%\begin{cases}
-%l^{(p)}(i_0,r)l^{(u)}(i_1) + l^{(p)}(i_1,r)l^{(u)}(i_0) \textit{\color{gray} ~if both $i_0$ and $i_1$ are internal} \\
-%l^{(p,n)}(i_0,r)l^{(p)}(i_1,i_0) + l^{(p,\not{n})}(i_0,r)l^{(u)}(i_1) \\
-%~~~~~~+ \begin{cases}
-%l^{(p)}(i_1,first(r,i_0))l^{(u, n)}(i_0) \textit{\color{gray} ~if $t_{i0} \geq t_{i1}$}  \\
-%0 \textit{\color{gray} ~if $t_{i0} < t_{i1}$}
-% \end{cases}\\
-%~~~~~~+ l^{(p)}(i_1,r)l^{(u, \not{n})}(i_0) \textit{\color{gray} ~if $i_0$ is a tip and $i_1$ is internal} \\
-%l^{(p,n)}(i_1,r)l^{(p)}(i_0,i_1) + l^{(p,\not{n})}(i_1,r)l^{(u)}(i_0) \\
-%~~~~~~+ \begin{cases}
-%l^{(p)}(i_0,first(r,i_1))l^{(u, n)}(i_1) \textit{\color{gray} ~if $t_{i1} \geq t_{i0}$}  \\
-%0 \textit{\color{gray} ~if $t_{i1} < t_{i0}$}
-% \end{cases}\\
-%~~~~~~+ l^{(p)}(i_0,r)l^{(u), \not{n}}(i_1) \textit{\color{gray} ~if $i_1$ is a tip and $i_0$ is internal} \\
-%l^{(p,\not{n})}(i_0,first(r,i_1))l^{(u,n)}(i_1) + l^{(p,n)}(i_0,first(r,i_1))l^{(p,n)}(i_1, i_0)\\ 
-%~~~~~~+l^{(p,\not{n})}(i_1,first(r,i_0))l^{(u,n)}(i_0) + l^{(p,n)}(i_1,first(r,i_0))l^{(p,n)}(i_0, i_1)\\
-%~~~~~~+l^{(p,n)}(i_0,r)l^{(p,\not{n})}(i_1, i_0) + l^{(p,n)}(i_1,r)l^{(p,\not{n})}(i_0, i_1)\\ 
-%~~~~~~+l^{(p,\not{n})}(i_0,r)l^{(u,\not{n})}(i_1) + l^{(p,\not{n})}(i_1,r)l^{(u,\not{n})}(i_0)
-%\textit{\color{gray} ~if both $i_0$ and $i_1$ are tips} \\
-% \end{cases}
-%\label{eq:lu}
-%\end{equation}
-
-Note that $l^{(p)}(a, b)$ depends on the notifier $b$ and can only be calculated once the notification time is known, i.e.
+Note that $l^{(n)}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time is known, i.e.
 once we have climbed to an internal node corresponding to the transmission between the notifier (corresponding to one of its children, must be a tip) and the partner (the other child).
 The calculation is done recursively by descending into the subtree.
-However, as all the unotified equation parts ($l^{(u)}(i)$) only depend on the corresponding node $i$, they can be pre-calculated and reused to speed up the recursive steps.
-
-%In the partner subtree only the contribution coming from the partner tip branch depends on the notification time.
-Hence while climbing the tree from tips to the root, for each internal node $i$ we can store an array of pre-calculated values for notified internal paths leading to each of its tips.
-The values of this array will store for each tip (e.g.
-$a$) its sampling time $t_a$ and a configuration $C(i,a)$, representing the subtree likelihood provided $a$ corresponds to the same notified partner individual as $i$ (but without the $a$'s tip branch contribution):
-%
-%\begin{equation}
-%\scriptsize
-%\begin{split}
-%C_{sj}(i,a) = &\begin{cases}
-%1 \textit{\color{gray}~~~if $i = a$}
-%\\
-%q^{(p)}_{sj}(i)
-%\sum\limits_{k=1}^{m}\lambda{si,k} \cdot l_{k}^{(u)}\Big(child(i, \not a)\Big)C_{si}\Big(child(i, a),a\Big) \textit{\color{gray}~~otherwise}\\ 
-%\end{cases},
-%\\ &\textit{where child(i,a) is the child of i whose subtree contains a},
-%\\
-% &\textit{~~~~~~~~child(i,$\not a$) is the child of i whose subtree does not contain a.}
-%\end{split}
-%\label{eq:c}
-%\end{equation}
-
-%Note that the path connecting $j$ to the parent of $i$ in this configuration corresponds to the same partner individual, hence this individual is the donor at all transmissions (internal nodes) on this path, hence at the moments of these transmissions the individual's state is known (as we only consider trees with known node states).
-At each internal node in this path the other branch (leading to a non-$j$'s subtree) corresponds to a recipient of the transmission occurring at that node, where the donor is the partner individual corresponding to $j$ (and $i$) and hence will not notify the recipient (as in our model's assumptions notified partners do not notify further).
-
-%Once a potential notifier tip $r$ is identified (i.e., $r$ is a sister tip of the currently visited node $i$), we can calculate $l^{(p)}_{sj}(i, r)$ as:
-%\begin{equation}
-%\scriptsize 
-%l^{(p)}_{sj}(i, r) = \sum\limits_{a \in tips(i)}C_{sj}(i, a) l^{(p)}_{sa}(a, r)\label{eq:lp}
-%\end{equation}
+However, as all the unotified equation parts, $l^{(\not{n})}(i)$, only depend on the corresponding node $i$, they can be pre-calculated and reused to speed up the recursive steps.
 
 \bigskip
-
-The tree $\mathscr{T}$ likelihood density $\mathscr{L}(\mathscr{T}|\Theta)$ under BD-PN model with parameters $\Theta=\{\lambda, \psi, \rho, \psi_p, \rho_n\}$ can be calculated as $l^{(u)}(root)$, as the root can only correspond to an unnotified individual (since there was no observed transmission before it).
 
@@
 The forest $\mathscr{F}$ might also include a certain number $u$ of unobserved trees, for which none of their tips got sampled.
-Forest likelihood formula hence combines the likelihoods of $f$ observed and $u$ hidden trees, and can be represented in logarithmic form~(\ref{eq:forest_likelihood}).
-Tree likelihood formula %[\ref{eq:tree_likelihood}]
+Forest likelihood formula hence combines the likelihoods of $f$ observed and $u$ hidden trees, and can be represented in a logarithmic form~(\ref{eq:forest_likelihood}).
+The tree likelihood formula %[\ref{eq:tree_likelihood}]
 is its special case, where $f=1$ and $u=0$.
@@
 We implemented a parameter estimator for the BD-PN model (which we called bdpn).
-It estimates the BD-PN model parameters  $\Theta = (\lambda,\psi,\psi_p,\rho,\rho_n) \in \mathbb{R}^5$ for a forest $\mathscr{F}$ comprising $f \geq 1$ observed trees in the maximum-likelihood framework, where one of the BD parameters ($\lambda,\psi$ or $\rho$) is fixed (for identifiability reasons).
+It estimates the BD-PN model parameters  $\Theta = (\lambda,\psi,\phi,\rho,\upsilon) \in \mathbb{R}^5$ for a forest $\mathscr{F}$ comprising $f \geq 1$ observed trees in the maximum-likelihood framework, where one of the BD parameters ($\lambda,\psi$ or $\rho$) is fixed (for identifiability reasons).
 
@@
 As with the point estimates, CI estimates were the best when $\psi$ or $\rho$ were fixed, and the worst when $\lambda$ was fixed (see Table~\ref{tbl:ci}).
-The relative width of the CIs ($\frac{value_{97.5\%} - value_{2.5\%}}{value}$) varied between 11\% for $\lambda$ estimates and 32\% for $\psi_p$ estimates (in all the settings).
+The relative width of the CIs ($\frac{value_{97.5\%} - value_{2.5\%}}{value}$) varied between 11\% for $\lambda$ estimates and 32\% for $\phi$ estimates (in all the settings).
 
@@
 Average relative error (and in parentheses average relative bias) are displayed for each parameter and method below their swarmplot.
-For $\rho$ and $\rho_n$ absolute values are shown instead of normalized ones.
+For $\rho$ and $\upsilon$ absolute values are shown instead of normalized ones.
 }
@@
  $\rho$ & 16\% & 55\%  & -- \\
- $\psi_p$ & 70\% & 93\% & 94\% \\
- $\rho_n$ & 21\% & 68\% & 63\% \\
+ $\phi$ & 70\% & 93\% & 94\% \\
+ $\upsilon$ & 21\% & 68\% & 63\% \\
 \bottomrule
@@
 
-Using $\rho=0.58$, we estimated the following parameter values: $\lambda=0.53$, $\psi=0.43$, $\psi_p=23-59$, $\rho_n=0.006-0.007$, which corresponds to $R_e = 1.2$, infectious time $\frac{1}{\psi} = 2.3$ years, and about 6--16 days between partner notification and their sampling.
+Using $\rho=0.58$, we estimated the following parameter values: $\lambda=0.53$, $\psi=0.43$, $\phi=23-59$, $\upsilon=0.006-0.007$, which corresponds to $R_e = 1.2$, infectious time $\frac{1}{\psi} = 2.3$ years, and about 6--16 days between partner notification and their sampling.
 The results are consistent between 10 forests and shown in Table~\ref{tbl:uk}.
@@
 fo-&num.&num.&&time&removal t.&\\
-rest&tips&trees&$R_e$&$\frac{1}{\psi}$ [years]& $\frac{1}{\psi_p}$ [days]&$\rho_n$\\
+rest&tips&trees&$R_e$&$\frac{1}{\psi}$ [years]& $\frac{1}{\phi}$ [days]&$\upsilon$\\
 \toprule
@@
 
-The -PN extension adds two constant parameters to the classical BD model: the probability of partner notification $p_n$ and the partner detection rate $\psi_p$.
+The -PN extension adds two constant parameters to the classical BD model: the probability of partner notification $p_n$ and the partner detection rate $\phi$.
 It can also be applied to a more general, multi-type birth-death (MTBD) model framework, which accounts for heterogeneity in the infected population (see Appendix).
@@
 The obtained estimates of epidemiological parameters are in agreement with what we now know about this epidemic (e.g., $R_e \approx 1.2$ ).
-We estimated a modest partner notification probability $\rho_p \in [0.003-0.1]$, and rapid notified partner detection ($3-18$ days vs $\approx2.3$ years for detection under the standard random procedure).
+We estimated a modest partner notification probability $\upsilon \in [0.003-0.1]$, and rapid notified partner detection ($3-18$ days vs $\approx2.3$ years for detection under the standard random procedure).
 
@@
 For the PN version of this simulator, we add an additional $m+1^{th}$ state for notified partners.
-$m + 1$'s state-change and transmission rates are set to zero: $\mu_{k,m+1} = \mu_{m+1,k} = \lambda_{k,m+1} = \lambda_{m+1,k} = 0\; \forall 1 \leq k \leq m$, while its removal rate is set to the notification removal rate: $\psi_{m+1} = \psi_p >> \psi_k\; \forall 1 \leq k \leq m$.
+$m + 1$'s state-change and transmission rates are set to zero: $\mu_{k,m+1} = \mu_{m+1,k} = \lambda_{k,m+1} = \lambda_{m+1,k} = 0\; \forall 1 \leq k \leq m$, while its removal rate is set to the notification removal rate: $\psi_{m+1} = \phi >> \psi_k\; \forall 1 \leq k \leq m$.
 We keep a mapping of most recent partners $P$ for each infectious individual, and update it for both the donor and the recipient at each transmission event.
@@
 If not, and if $i$ gets sampled upon removal, we draw a value $p$ (uniformly) from an interval $[0, 1[$.
-If $p < \rho_n$, and if $i$'s most recent partner ($j = P[i]$) is not yet sampled (not in $N[state(j)]$), we change $j$'s state to $m + 1$ as in a state-change event.
+If $p < \upsilon$, and if $i$'s most recent partner ($j = P[i]$) is not yet sampled (not in $N[state(j)]$), we change $j$'s state to $m + 1$ as in a state-change event.
+
+\subsubsection*{Code availability}
+The simulator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\citep{Huerta-Cepas2016} and NumPy package for array operations~\citep{harris_array_2020}.
+
+It is available as a command-line program and a Python 3 package via PyPi (\href{https://pypi.org/project/treesimulator}{treesimulator}), and via Docker/Singularity (\href{https://hub.docker.com/r/evolbioinfo/treesimulator/tags}{evolbioinfo/treesimulator}).
+Its source code and the installation and usage documentation are available on GitHub at \href{https://github.com/evolbioinfo/treesimulator}{github.com/evolbioinfo/treesimulator}.
 

### 2024-04-30 bdpn/model_distinguisher.py

@@ -3,7 +3,10 @@ from functools import reduce
 from multiprocessing.pool import ThreadPool
 
 import numpy as np
-from wquantiles import quantile
+import scipy
+from matplotlib import pyplot as plt
+from matplotlib.pyplot import show
+import seaborn as sns
 
 from bdpn.tree_manager import TIME, read_forest, annotate_forest_with_time
 
@@ -24,7 +27,7 @@ class CherryLikeMotif(object):
         :param root: ete.TreeNode, the root of the motif subtree
         :param clustered_tips: list of clustered child tips
         """
-        self.root = root if root else clustered_tips[0].get_common_ancestor(*clustered_tips)
+        self.root = root # if root else clustered_tips[0].get_common_ancestor(*clustered_tips)
         self.clustered_tips = clustered_tips
 
     def __str__(self):
@@ -160,6 +163,189 @@ def pn_test(forest, k=DEFAULT_NEIGHBOURHOOD_SIZE, repetitions=RANDOM_REPETITIONS
     return result
 
 
+def sign_test(forest):
+    """
+    Tests if the input forest was generated under a -PN model.
+
+    The test detects cherries in the forest and sorts them by the times of their roots.
+    For each cherry the test calculates the difference between its tip times,
+    hence obtaining an array of cherry tip differences.
+    It then generates a collection of random cherry tip differences of the same size:
+    It fixed one of the tips for each cherry and then swaps the other tips between neighbouring cherries,
+    such that the other tip of cherry 2i is swapped with the other tip of cherry 2i + 1 (i = 0, 1, ...).
+    (If the total number of cherries is odd, the last three cherries instead of the last two
+    swap their other tips in a cycle). For each hence reshuffled cherry its tip difference is calculated.
+
+    Finally, we calculate the sign test of one by one comparison of real vs reshuffled diffs
+    (-1 if the difference for the i-th cherry is smaller in the real array, 1 if larger, 0 is the same).
+
+    The test therefore reports a probability of partner notification
+    being present in the tree.
+
+    :param forest: list of trees
+    :return: pval
+    """
+    annotate_forest_with_time(forest)
+
+    all_cherries = []
+    for tree in forest:
+        all_cherries.extend(pick_cherries(tree, include_polytomies=False))
+    all_cherries = sorted(all_cherries, key=lambda _: getattr(_.root, TIME))
+
[…]

### 2024-04-30 examples/main_cherry_test.py

@@ -1,8 +1,12 @@
 import logging
 import os
 
-from bdpn.model_distinguisher import pn_test
+from treesimulator.generator import generate
+from treesimulator.mtbd_models import BirthDeathModel, PNModel
+
+from bdpn.model_distinguisher import pn_test, cherry_diff_plot, sign_test, estimate_pn
 from bdpn.tree_manager import read_forest
+from treesimulator import simulate_forest_bd
 
 DATA_DIR = os.path.abspath(os.path.join(os.path.dirname(__file__), '..', 'simulations', 'trees'))
 NWK_BD = os.path.join(DATA_DIR, 'BD', 'tree.{}.nwk')
@@ -12,7 +16,32 @@ min_cherry_num = 100
 if __name__ == '__main__':
     logging.basicConfig(level=logging.INFO, format='%(asctime)s: %(message)s', datefmt="%Y-%m-%d %H:%M:%S")
 
-    for tree_id in range(100):
-        for nwk in (NWK_BD.format(tree_id), NWK_BDPN.format(tree_id)):
-            pval = pn_test(read_forest(nwk), repetitions=1e2)
-            logging.info('Tree {tree}, PN test {res}.'.format(res=pval, tree=nwk.replace(DATA_DIR, '')))
+    forest = read_forest('/home/azhukova/projects/bdpn/trees/D_CRF_19_timetree.resolved.nwk')
+    print(sign_test(forest))
+    # forest = read_forest('/home/azhukova/projects/bdpn/hiv_b_uk/data/forest.2012_2015.0.nwk')
+    # print(sign_test(forest))
+    # print(estimate_pn(forest))
+    # exit()
+
+    bd_model = BirthDeathModel(p=0.7, la=2, psi=1)
+    forest, _, _ = generate(bd_model, min_tips=5000, max_tips=5050)
+    cherry_diff_plot(forest, outfile='bd.jpg')
+    print(sign_test(forest))
+    print(estimate_pn(forest))
+    forest[0].write(outfile='bd.nwk')
+
+    pn_model = PNModel(model=bd_model, pn=0.9, removal_rate=25)
+    forest, _, _ = generate(pn_model, min_tips=5000, max_tips=5050)
+    cherry_diff_plot(forest, outfile='bdpn.jpg')
+    print(sign_test(forest))
+    print(estimate_pn(forest))
+    forest[0].write(outfile='bdpn.nwk')
+
+    # for tree_id in range(1):
[…]

### 2024-04-30 setup.py

@@ -22,8 +22,8 @@ setup(
     install_requires=['six', 'ete3', 'numpy', 'scipy', 'wquantiles'],
     entry_points={
             'console_scripts': [
-                'bdpn_infer = bdpn.bdpn:main',
-                'bd_infer = bdpn.bd:main',
+                'bdpn_infer = bdpn.bdpn_model:main',
+                'bd_infer = bdpn.bd_model:main',
                 'pn_test = bdpn.model_distinguisher:main',
             ]
     },

### 2024-04-30 bdpn/bdpn.py
Rewrote formulas case by case

@@ -1,379 +1,0 @@
-import os
-from multiprocessing.pool import ThreadPool
-
-import numpy as np
-
-from bdpn import bd
-from bdpn.formulas import get_log_p, get_c1, get_c2, get_E2, get_E1, get_log_po, get_log_pnh, get_log_po_from_p_pnh, \
-    get_u
-from bdpn.parameter_estimator import optimize_likelihood_params, rescale_log
-from bdpn.tree_manager import TIME, read_forest, annotate_forest_with_time
-
-PARAMETER_NAMES = np.array(['la', 'psi', 'partner_psi', 'p', 'pn'])
-
-DEFAULT_LOWER_BOUNDS = [bd.DEFAULT_MIN_RATE, bd.DEFAULT_MIN_RATE, bd.DEFAULT_MIN_RATE,
-                        bd.DEFAULT_MIN_PROB, bd.DEFAULT_MIN_PROB]
-DEFAULT_UPPER_BOUNDS = [bd.DEFAULT_MAX_RATE, bd.DEFAULT_MAX_RATE, bd.DEFAULT_MAX_RATE * 1e3,
-                        bd.DEFAULT_MAX_PROB, bd.DEFAULT_MAX_PROB]
-
-
-def loglikelihood(forest, la, psi, psi_n, rho, rho_n, T=None, threads=1):
-    annotate_forest_with_time(forest)
-    if T is None:
-        T = 0
-        for tree in forest:
-            T = max(T, max(getattr(_, TIME) for _ in tree))
-
-    c1 = get_c1(la=la, psi=psi, rho=rho)
-    c2 = get_c2(la=la, psi=psi, c1=c1)
-
-    nodes = []
-    for tree in forest:
-        for n in tree.traverse():
-            nodes.append(n)
-
-    def annotate_node(n):
-        ti = getattr(n, TIME)
-        t = ti - n.dist
-
-        E1 = get_E1(c1=c1, c2=c2, t=t, T=T)
-        E2 = get_E2(c1=c1, c2=c2, ti=ti, T=T)
-
-        log_p = get_log_p(c1=c1, t=t, ti=ti, E1=E1, E2=E2)
-        log_pnh = get_log_pnh(la=la, psi=psi, t=t, ti=ti)
-        log_po = get_log_po_from_p_pnh(log_p=log_p, log_pnh=log_pnh)
-
-        n.add_feature('p', log_p)
-        n.add_feature('po', log_po)
-        if n.is_leaf():
[…]

### 2024-04-30 bdpn/bdpn_model.py

@@ -1,0 +1,354 @@
+import os
+from multiprocessing.pool import ThreadPool
+
+import numpy as np
+
+from bdpn import bd_model
+from bdpn.formulas import get_log_p, get_c1, get_c2, get_E2, get_E1, get_log_po, get_log_pn, get_log_po_from_p_pn, \
+    get_u, get_log_pp
+from bdpn.parameter_estimator import optimize_likelihood_params, rescale_log
+from bdpn.tree_manager import TIME, read_forest, annotate_forest_with_time
+
+PARAMETER_NAMES = np.array(['la', 'psi', 'partner_psi', 'p', 'pn'])
+
+DEFAULT_LOWER_BOUNDS = [bd_model.DEFAULT_MIN_RATE, bd_model.DEFAULT_MIN_RATE, bd_model.DEFAULT_MIN_RATE,
+                        bd_model.DEFAULT_MIN_PROB, bd_model.DEFAULT_MIN_PROB]
+DEFAULT_UPPER_BOUNDS = [bd_model.DEFAULT_MAX_RATE, bd_model.DEFAULT_MAX_RATE, bd_model.DEFAULT_MAX_RATE * 1e3,
+                        bd_model.DEFAULT_MAX_PROB, bd_model.DEFAULT_MAX_PROB]
+
+
+
+
+def loglikelihood(forest, la, psi, phi, rho, upsilon, T=None, threads=1):
+    annotate_forest_with_time(forest)
+    if T is None:
+        T = 0
+        for tree in forest:
+            T = max(T, max(getattr(_, TIME) for _ in tree))
+
+    c1 = get_c1(la=la, psi=psi, rho=rho)
+    c2 = get_c2(la=la, psi=psi, c1=c1)
+
+    log_psi = np.log(psi)
+    log_phi = np.log(phi)
+    log_rho = np.log(rho)
+    log_ups = np.log(upsilon)
+    log_not_ups = np.log(1 - upsilon)
+    log_2 = np.log(2)
+    log_la = np.log(la)
+
+    def process_node(node):
+        ti = getattr(node, TIME)
+        tj = ti - node.dist
+
+        E1 = get_E1(c1=c1, c2=c2, t=tj, T=T)
+        E2 = get_E2(c1=c1, c2=c2, ti=ti, T=T)
+
+        log_p = get_log_p(c1=c1, t=tj, ti=ti, E1=E1, E2=E2)
+        log_po = get_log_po_from_p_pn(log_p=log_p, log_pn=log_pn)
+
[…]

### 2024-04-30 bdpn/formulas.py

@@ -80,24 +80,33 @@ def get_log_po(la, psi, c1, t, ti, E1, E2):
     return np.log(E2 + 1) - np.log(E1 + 1) + (c1 + la + psi) * (t - ti) / 2
 
 
-def get_log_po_from_p_pnh(log_p, log_pnh):
+def get_log_po_from_p_pn(log_p, log_pn):
     """
     Returns log(po(t)), where
     dpo(t)/dt = la + psi * po(t) − la * po(t) * U(t)
     po(ti) = 1
-    if log(p(t)) and log(pnh(t)) are known, where
-    dpnh(t)/dt = la + psi * pnh(t)
-    pnh(ti) = 1,
+    if log(p(t)) and log(pn(t)) are known, where
+    dpn(t)/dt = la + psi * pnh(t)
+    pn(ti) = 1,
     dp(t)/dt = la + psi * p(t) − 2 * la * p(t) * U(t)
     p(ti) = 1
     """
-    return (log_p + log_pnh) / 2
+    return (log_p + log_pn) / 2
 
 
-def get_log_pnh(la, psi, t, ti):
+def get_log_pn(la, psi, t, ti):
     """
-    Returns log(pnh(t)), where
-    dpnh(t)/dt = la + psi * pnh(t)
-    pnh(ti) = 1
+    Returns log(pn(t)), where
+    dpn(t)/dt = la + psi * pn(t)
+    pn(ti) = 1
     """
     return (la + psi) * (t - ti)
+
+
+def get_log_pp(la, phi, t, ti):
+    """
+    Returns log(pp(t)), where
+    dpp(t)/dt = phi * pp(t)
+    pp(ti) = 1
+    """
+    return phi * (t - ti)

### 2024-04-30 simulations/py/likelihood_comparison.py

@@ -1,7 +1,8 @@
 import pandas as pd
 from ete3 import Tree
 
-from bdpn.bdpn import loglikelihood
+from bdpn import bd_model
+from bdpn.bdpn_model import loglikelihood
 import numpy as np
 from scipy.stats import binomtest
 
@@ -32,7 +33,10 @@ if __name__ == "__main__":
         type2lk = {}
         for type in ALL_TYPES:
             la, psi, psi_n, rho, rho_n = df.loc['{}_{}'.format(type, i), ['lambda', 'psi', 'psi_p', 'p', 'pn']]
-            type2lk[type] = loglikelihood([tree], la, psi, psi_n, rho, rho_n)
+            if 'BDPN' in type:
+                type2lk[type] = loglikelihood([tree], la, psi, psi_n, rho, rho_n)
+            else:
+                type2lk[type] = bd_model.loglikelihood([tree], la, psi, rho)
             index.append(i)
         best_lk = max(type2lk.values())
         for _ in ALL_TYPES:

### 2024-04-30 simulations/py/likelihood_plot.py

@@ -1,0 +1,64 @@
+import pandas as pd
+from ete3 import Tree
+
+from bdpn import bd_model
+from bdpn.bdpn_model import loglikelihood
+import numpy as np
+from scipy.stats import binomtest
+
+import matplotlib.pyplot as plt
+
+from mpl_toolkits.mplot3d import axes3d
+
+PYBDEI = 'PyBDEI'
+
+REAL_TYPE = 'real'
+
+
+if __name__ == "__main__":
+    import argparse
+
+    parser = argparse.ArgumentParser(description="Plots likelihoods.")
+    parser.add_argument('--nwk', type=str, help="tree", default="/home/azhukova/projects/bdpn/simulations/trees/BDPN/tree.0.nwk")
+    parser.add_argument('--log', type=str, default="/home/azhukova/projects/bdpn/simulations/trees/BDPN/tree.0.log")
+    params = parser.parse_args()
+
+    tree = Tree(params.nwk)
+
+    df = pd.read_csv(params.log)
+    R0, infectious_time, p, pn, rt = df.iloc[0, :5]
+    la, psi, psip = R0 / infectious_time, 1 / infectious_time, 1 / rt
+
+    pns = np.arange(pn / 10, max(pn * 10, 1), step=(-pn / 10 + max(pn * 10, 1)) / 10)
+    psips = np.arange(psip / 10, psip * 10, step=(-psip / 10 + psip * 10) / 10)
+    Z = -np.ones((pns.shape[0], psips.shape[0]), dtype=float) * np.inf
+    X = np.zeros(Z.shape, dtype=float)
+    Y = np.zeros(Z.shape, dtype=float)
+    z_min, z_max = np.inf, -np.inf
+    for i in range(pns.shape[0]):
+        for j in range(psips.shape[0]):
+            X[i, j] = pns[i]
+            Y[i, j] = psips[j]
+            z = loglikelihood([tree], la, psi, psips[j], p, pns[i])
+            Z[i, j] = z
+            z_min = min(z_min, z)
+            z_max = max(z_max, z)
+
+    ax = plt.figure().add_subplot(projection='3d')
+    # Plot the 3D surface
+    ax.plot_surface(X, Y, Z, edgecolor='royalblue', lw=0.5, alpha=0.3)
+    ax.set(xlim=(pns[0], pns[-1]), ylim=(psips[0], psips[-1]), zlim=(z_min, z_max),
[…]

### 2024-04-30 simulations/py/plot_error_bd.py

@@ -7,11 +7,11 @@ import seaborn as sns
 from matplotlib.offsetbox import TextArea, HPacker, AnchoredOffsetbox, VPacker
 from statsmodels.stats.weightstats import CompareMeans
 
-RATE_PARAMETERS = ['lambda', 'psi', 'psi_p', 'p', 'pn']
-EPIDEMIOLOGIC_PARAMETERS = ['R_naught', 'infectious_time', 'partner_removal_time']
-par2greek = {'lambda': u'\u03bb', 'psi': u'\u03c8', 'psi_p': u'\u03c8p', 'p': '\u03c1', 'pn': '\u03c1n',
+RATE_PARAMETERS = ['lambda', 'psi', 'p']
+EPIDEMIOLOGIC_PARAMETERS = ['R_naught', 'infectious_time']
+par2greek = {'lambda': u'\u03bb', 'psi': u'\u03c8', 'p': '\u03c1',
              'R_naught': u'\u0052\u2080' + '=' + u'\u03bb\u002F\u03c8',
-             'infectious_time': 'infectious time 1' + u'\u002F\u03c8', 'partner_removal_time': 'partner removal time 1' + u'\u002F\u03c8p'}
+             'infectious_time': 'infectious time 1' + u'\u002F\u03c8'}
 PARAMETERS = RATE_PARAMETERS + EPIDEMIOLOGIC_PARAMETERS
 
 if __name__ == "__main__":
@@ -22,14 +22,11 @@ if __name__ == "__main__":
     parser.add_argument('--pdf', type=str, help="plot")
     parser.add_argument('--tab', type=str, help="error table")
     parser.add_argument('--fixed', type=str, default='p', help="fixed parameter", choices=RATE_PARAMETERS)
-    # parser.add_argument('--pn_min', type=float, default=0., help="Only display cases with pn greater or equal to this value")
-    # parser.add_argument('--pn_max', type=float, default=1, help="Only display cases with pn smaller than this value")
     params = parser.parse_args()
 
     df = pd.read_csv(params.estimates, sep='\t', index_col=0)
 
     real_df = df.loc[df['type'] == 'real', :]
-    # real_df = real_df[(real_df['pn'] * real_df['p'] >= params.pn_min) & (real_df['pn'] * real_df['p'] < params.pn_max)]
 
     df = df.loc[df['type'] != 'real', :]
     df = df.loc[real_df.index, :]
@@ -40,12 +37,10 @@ if __name__ == "__main__":
[…]

### 2024-04-30 simulations/py/summary_table_bd.py

@@ -8,9 +8,6 @@ if __name__ == "__main__":
     import argparse
 
     parser = argparse.ArgumentParser(description="Summarize errors.")
-    parser.add_argument('--estimated_p', nargs='*', default=[], type=str, help="estimated parameters")
-    parser.add_argument('--estimated_psi', nargs='*', default=[], type=str, help="estimated parameters")
-    parser.add_argument('--estimated_la', nargs='*', default=[], type=str, help="estimated parameters")
     parser.add_argument('--estimated_p_bd', nargs='*', default=[], type=str, help="estimated parameters")
     parser.add_argument('--estimated_psi_bd', nargs='*', default=[], type=str, help="estimated parameters")
     parser.add_argument('--estimated_la_bd', nargs='*', default=[], type=str, help="estimated parameters")
@@ -27,51 +24,18 @@ if __name__ == "__main__":
                                'psi_p', 'psi_p_min', 'psi_p_max',
                                'R_naught', 'R_naught_min', 'R_naught_max',
                                'infectious_time', 'infectious_time_min', 'infectious_time_max',
-                               'p', 'p_min', 'p_max',
-                               'pn', 'pn_min', 'pn_max',
-                               'partner_removal_time', 'partner_removal_time_min', 'partner_removal_time_max'])
+                               'p', 'p_min', 'p_max'])
 
     for real in params.real:
         i = int(re.findall(r'[0-9]+', real)[0])
         ddf = pd.read_csv(real)
         # R0,infectious time,sampling probability,notification probability,removal time after notification,tips,time,hidden_trees
-        R0, it, p, pn, rt, tips, T, h_trees \
-            = ddf.loc[next(iter(ddf.index)), :]
+        R0, it, p, tips, T, h_trees \
+            = ddf.loc[next(iter(ddf.index)), ['R0', 'infectious time', 'sampling probability', 'tips', 'time', 'hidden_trees']]
         df.loc['{}.real'.format(i),
-               ['R_naught', 'infectious_time', 'partner_removal_time',
-                'lambda', 'psi', 'psi_p', 'p', 'pn', 'sampled_tips', 'type']] \
-            = [R0, it, rt, R0 / it, 1 / it, 1 / rt, p, pn, tips, 'real']
-
-    estimate_list = []
-    if params.estimated_la:
[…]

### 2024-04-30 bdpn/bdpn_model.py
Fixed lk calculations

@@ -2,6 +2,7 @@ import os
 from multiprocessing.pool import ThreadPool
 
 import numpy as np
+import pandas as pd
 
 from bdpn import bd_model
 from bdpn.formulas import get_log_p, get_c1, get_c2, get_E2, get_E1, get_log_po, get_log_pn, get_log_po_from_p_pn, \
@@ -17,27 +18,38 @@ DEFAULT_UPPER_BOUNDS = [bd_model.DEFAULT_MAX_RATE, bd_model.DEFAULT_MAX_RATE, bd
                         bd_model.DEFAULT_MAX_PROB, bd_model.DEFAULT_MAX_PROB]
 
 
+def log_sum(log_summands):
+    """
+    Takes [logX1, ..., logXk] as input and returns log(X1 + ... + Xk) as output, 
+    while taking care of potential under/overflow.
+    
+    :param log_summands: an array of summands in log form
+    :return: log of the sum
+    """
+    result = np.array(log_summands, dtype=np.float64)
+    factors = rescale_log(result)
+    return np.log(np.sum(np.exp(result))) - factors
 
 
 def loglikelihood(forest, la, psi, phi, rho, upsilon, T=None, threads=1):
     annotate_forest_with_time(forest)
-    if T is None:
-        T = 0
-        for tree in forest:
-            T = max(T, max(getattr(_, TIME) for _ in tree))
+    T = get_T(T, forest)
 
     c1 = get_c1(la=la, psi=psi, rho=rho)
     c2 = get_c2(la=la, psi=psi, c1=c1)
 
-    log_psi = np.log(psi)
-    log_phi = np.log(phi)
-    log_rho = np.log(rho)
-    log_ups = np.log(upsilon)
-    log_not_ups = np.log(1 - upsilon)
-    log_2 = np.log(2)
-    log_la = np.log(la)
+    log_la, log_psi, log_phi, log_rho, log_ups, log_not_ups, log_2 = \
+        np.log(la), np.log(psi), np.log(phi), np.log(rho), np.log(upsilon), np.log(1 - upsilon), np.log(2)
 
     def process_node(node):
+        """
+        Calculate the loglikelihood density of this branch being unnotified, including the subtree
+        and put it into node's annotation: 'lx' for internal nodes,
+        'lxx' and 'lxn' (depending on whether the node notified) for tips.
+
+        :param node: tree node whose children (if any) are already processed
+        :return: void, add a node annotation
+        """
         ti = getattr(node, TIME)
         tj = ti - node.dist
 
[…]

### 2024-04-30 bdpn/parameter_estimator.py

@@ -41,6 +41,7 @@ def optimize_likelihood_params(forest, input_parameters, loglikelihood, bounds, 
     :param forest: a list of ete3.Tree trees
     :return: tuple: (the values of optimized parameters, CIs)
     """
+    # print('Bounds are set to {}'.format(bounds))
     annotate_forest_with_time(forest)
     T = 0
     for tree in forest:

### 2024-05-03 article/article.tex
Fixed lk calculations
REmoved unused code

@@
 \scriptsize
-l^{(n)}(i, r) = 0 \textit{\color{gray}~if $i$ is internal and $t_i > t_r$ or $i$ is a tip and $t_{parent(i)} > t_r$} \\
+l^{(n)}(i, r) = 0 \textit{\color{gray}~if $t_i > t_r$} \\
 \label{eq:lp-zero}
@@
 
-We applied bdpn to each of the trees three times: fixing each of the BD parameters ($\lambda,\psi$ or $\rho$) to its real value (for identifiability).
+We applied bdpn to each of the trees fixing the sampling probability $\rho$ to its real value (for identifiability).
+The results with the becoming-uninfectious rate $\psi$ are similar and given in the Supplementary Material.
 

### 2024-05-03 bdpn/bd_model.py

@@ -102,12 +102,13 @@ def infer(forest, la=None, psi=None, p=None,
     bounds[:, 1] = upper_bounds
     start_parameters = get_start_parameters(forest, la, psi, p)
     input_params = np.array([la, psi, p])
-    print('Input parameters {} are fixed: {}'.format(PARAMETER_NAMES[input_params != None],
-                                                     input_params[input_params != None]))
+    print('Fixed input parameter(s): {}'
+          .format(', '.join('{}={:g}'.format(*_)
+                            for _ in zip(PARAMETER_NAMES[input_params != None], input_params[input_params != None]))))
     print('Starting BD parameters: {}'.format(start_parameters))
-    vs, cis = optimize_likelihood_params(forest, input_parameters=input_params,
-                                      loglikelihood=loglikelihood, bounds=bounds[input_params == None],
-                                      start_parameters=start_parameters, cis=ci)
+    vs, cis, lk = optimize_likelihood_params(forest, input_parameters=input_params,
+                                             loglikelihood=loglikelihood, bounds=bounds[input_params == None],
+                                             start_parameters=start_parameters, cis=ci)
     print('Estimated BD parameters: {}'.format(vs))
     return vs, cis
 
@@ -116,7 +117,7 @@ def save_results(vs, cis, log, ci=False):
     os.makedirs(os.path.dirname(os.path.abspath(log)), exist_ok=True)
     with open(log, 'w+') as f:
         f.write(',{}\n'.format(','.join(['R0', 'infectious time', 'sampling probability',
-                                          'transmission rate', 'removal rate'])))
+                                         'transmission rate', 'removal rate'])))
         la, psi, rho = vs
         R0 = la / psi
         rt = 1 / psi

### 2024-05-03 bdpn/bdpn_model.py

@@ -108,7 +108,7 @@ def loglikelihood(forest, la, psi, phi, rho, upsilon, T=None, threads=1):
         tj = ti - tip.dist
 
         if tr < tj:
-            return -np.inf
+            return -np.inf, -np.inf
         if tr > ti:
             return getattr(tip, 'lnx_late'), getattr(tip, 'lnn_late')
         else:
@@ -176,7 +176,7 @@ def loglikelihood(forest, la, psi, phi, rho, upsilon, T=None, threads=1):
     def process_tree(tree):
         for node in tree.traverse('postorder'):
             process_node(node)
-        return getattr(tree, 'lx')
+        return getattr(tree, 'lx' if not tree.is_leaf() else 'lxx')
 
     if threads > 1 and len(forest) > 1:
         with ThreadPool(processes=threads) as pool:
@@ -250,7 +250,7 @@ def main():
     #     la = R * psi
     #     phi = 1 / rt
     #     print('Real parameters: ', np.array([la, psi, phi, p, pn]))
-    #     params.p = p
+    #     params.psi = psi
 
     if params.la is None and params.psi is None and params.p is None:
         raise ValueError('At least one of the BD model parameters (la, psi, p) needs to be specified '
@@ -292,13 +292,16 @@ def infer(forest, la=None, psi=None, phi=None, p=None, pn=None,
     start_parameters = np.array([vs[0], vs[1], vs[1] * 10 if phi is None or phi < 0 else phi,
                                  vs[-1], 0.5 if pn is None or pn <= 0 or pn > 1 else pn])
     input_params = np.array([la, psi, phi, p, pn])
-    print('Input parameters {} are fixed: {}'.format(PARAMETER_NAMES[input_params != None],
-                                                     input_params[input_params != None]))
+    print('Fixed input parameter(s): {}'
+          .format(', '.join('{}={:g}'.format(*_)
+                            for _ in zip(PARAMETER_NAMES[input_params != None], input_params[input_params != None]))))
     print('Starting BDPN parameters: {}'.format(start_parameters))
-    vs, cis = optimize_likelihood_params(forest, input_parameters=input_params,
-                                         loglikelihood=loglikelihood, bounds=bounds[input_params == None],
-                                         start_parameters=start_parameters, cis=ci)
+    vs, cis, lk = optimize_likelihood_params(forest, input_parameters=input_params,
+                                             loglikelihood=loglikelihood, bounds=bounds[input_params == None],
[…]

### 2024-05-03 bdpn/model_distinguisher.py

@@ -1,6 +1,4 @@
 import logging
-from functools import reduce
-from multiprocessing.pool import ThreadPool
 
 import numpy as np
 import scipy
@@ -35,73 +33,6 @@ class CherryLikeMotif(object):
             .format(self.root.name, len(self.clustered_tips), ', '.join(_.name for _ in self.clustered_tips))
 
 
-def nonparametric_cherry_diff(cherries, k=DEFAULT_NEIGHBOURHOOD_SIZE, repetitions=RANDOM_REPETITIONS):
-    """
-    Calculates the branch differences in real (pseudo)cherries and in random ones.
-    Random cherry configuration is generated by replacing each real (pseudo)cherry
-    by a (pseudo)cherry whose tips are randomly drawn from a k-neighbourhood of the real (pseudo)cherry.
-    The k-neighbourhood contains the k cherries whose roots are the closest in time to the (pseudo)cherry in question,
-    including itself.
-
-    We compare the real (pseudo)cherry branch differences
-    to the distribution of those for the random configurations.
-    Each real [d1, d2, ..., dn] vs random [d'1, d'2, ..., d'n] configuration comparison is performed by:
-        * comparing one by one each corresponding cherry and calculating the sign:
-        -1 if the random one if smaller, +1 if the real one is smaller or 0 if they are equal
-        * calculating the sum of signs (so that for similar distributions the sum should be close to 0)
-    Finally we report the proportion of sing sums that are smaller than 1.
-
-    :param repetitions: how many times a random (pseudo)cherry configuration should be generated.
-    :param k: the size of the k-neighbourhood.
-    :return: the proportion, between 0 and 1,
-        of random (pseudo)cherry diffs that have more than half elements that are smaller
-        than the corresponding elements in the real one.
-    """
-    cherries = sorted(cherries, key=lambda _: getattr(_.root, TIME))
-    n = len(cherries)
-
-    def get_diff(b1, b2):
-        return abs(b1.dist - b2.dist)
-
-    def get_diff_distribution(tip_list_collection):
-        data, weights = [], []
-        for tips in tip_list_collection:
-            size = len(tips)
-            # weight is calculated as (size / 2) / C^2_{size} = (size / 2) / (size * (size - 1) / 2) = 1 / (size - 1),
-            # so that in total all possible cherries in this polytomy count as size / 2 cherries
-            weight = 1 / (size - 1)
[…]

### 2024-05-03 bdpn/parameter_estimator.py

@@ -99,29 +99,65 @@ def optimize_likelihood_params(forest, input_parameters, loglikelihood, bounds, 
     optimised_cis[:, 1] = np.array(optimised_parameters)
 
     if cis:
-        lk_threshold = loglikelihood(forest, *optimised_parameters, T, threads=threads) - chi2.ppf(q=0.95, df=1) / 2
+        print('Estimated parameters:', optimised_parameters)
+        print('Estimating CIs...')
+        diff = chi2.ppf(q=0.95, df=1) / 2
+        lk_threshold = loglikelihood(forest, *optimised_parameters, T, threads=threads) - diff
 
-        def binary_search(v_min, v_max, i, lower=True):
-            rps = np.array(optimised_parameters)
+        def binary_search(v_min, v_max, get_lk, lower=True):
             v = v_min + (v_max - v_min) / 2
-            if (v_max - v_min) < 1e-4:
-                return v
-            rps[i] = v
-            lk = loglikelihood(forest, *rps, T, threads=threads) #threads=max(1, threads // len(bounds)))
-            lk_diff = lk - lk_threshold
-            if np.abs(lk_diff) < 1e-4:
+            if (v_max - v_min) < 1e-3:
+                return v_min if lower else v_max
+
+            lk_diff = get_lk(v) - lk_threshold
+            if np.abs(lk_diff) < diff / 100:
                 return v
 
             go_left = (lower and lk_diff > 0) or ((not lower) and lk_diff < 0)
 
             if go_left:
-                return binary_search(v_min, v, i, lower)
-            return binary_search(v, v_max, i, lower)
+                return binary_search(v_min, v, get_lk, lower)
+            return binary_search(v, v_max, get_lk, lower)
 
         def get_ci(args):
             (i, optimised_value), (b_min, b_max) = args
-            optimised_cis[i, 0] = binary_search(b_min, optimised_value, i, True)
-            optimised_cis[i, 1] = binary_search(optimised_value, b_max, i, False)
+            # print('---------')
+            # print(i, optimised_value, (b_min, b_max))
+
+            bs = []
+            bound_iterator = iter(bounds)
+            for bs_i in range(n):
+                if input_parameters[bs_i] is not None:
+                    continue
+                if bs_i == i:
+                    # skip these bounds
+                    next(bound_iterator)
+                    continue
+                bs.append(next(bound_iterator))
+
+            bs = np.array(bs)
+            rps = np.array(optimised_parameters)
[…]

### 2024-05-03 bdpn/tree_manager.py

@@ -80,10 +80,7 @@ def read_forest(tree_path):
 
 def annotate_tree(tree):
     for n in tree.traverse('preorder'):
-        if n.is_root():
-            p_time = 0
-        else:
-            p_time = getattr(n.up, TIME)
+        p_time = 0 if n.is_root() else getattr(n.up, TIME)
         n.add_feature(TIME, p_time + n.dist)
 
 

### 2024-05-03 simulations/py/summary_table.py

@@ -35,12 +35,22 @@ if __name__ == "__main__":
         i = int(re.findall(r'[0-9]+', real)[0])
         ddf = pd.read_csv(real)
         # R0,infectious time,sampling probability,notification probability,removal time after notification,tips,time,hidden_trees
-        R0, it, p, pn, rt, tips, T, h_trees \
-            = ddf.loc[next(iter(ddf.index)), :]
-        df.loc['{}.real'.format(i),
-               ['R_naught', 'infectious_time', 'partner_removal_time',
-                'lambda', 'psi', 'psi_p', 'p', 'pn', 'sampled_tips', 'type']] \
-            = [R0, it, rt, R0 / it, 1 / it, 1 / rt, p, pn, tips, 'real']
+        try:
+            R0, it, p, pn, rt, tips, T, h_trees \
+                = ddf.loc[next(iter(ddf.index)), :]
+            df.loc['{}.real'.format(i),
+                   ['R_naught', 'infectious_time', 'partner_removal_time',
+                    'lambda', 'psi', 'psi_p', 'p', 'pn', 'sampled_tips', 'type']] \
+                = [R0, it, rt, R0 / it, 1 / it, 1 / rt, p, pn, tips, 'real']
+        except:
+            # This is a BD model tree actually
+            R0, it, p, tips, T, h_trees \
+                = ddf.loc[next(iter(ddf.index)), :]
+            df.loc['{}.real'.format(i),
+                   ['R_naught', 'infectious_time',
+                    'lambda', 'psi', 'p', 'sampled_tips', 'type']] \
+                = [R0, it, R0 / it, 1 / it, p, tips, 'real']
+
 
     estimate_list = []
     if params.estimated_la:

### 2024-05-21 article/article.tex
Article updated listing hidden partner cases

@@
 quarantine measures or development of pathogen detection tests), pathogen evolution over time (e.g.
-some SARS-CoV-2 variants are more transmissible than other),  differences between host individuals (e.g.
+some SARS-CoV-2 variants are more transmissible than the others),  differences between host individuals (e.g.
 due to their immune system particularities or to their behaviour).
@@
 
-Another important source of heterogeneity that none of these models accounts for is inflicted by non-random sampling.
-We can propose different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability over time intervals with the BDSKY, however, none of them can account for non-randomness of sampling, in particular due to partner notification (contact tracing).
+Another important source of heterogeneity that none of these models accounts for is inflicted by non-independent sampling.
+We can propose different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability over time intervals with the BDSKY, however, none of them can account for correlation effects in sampling, in particular due to partner notification (contact tracing).
 Partner notification or contact tracing is a process in which contacts of detected infected individuals are identified and offered testing for infection.
@@
 We describe the -PN extension and its assumptions, and present a transmission tree simulator under MTBD-PN models.
-We propose a test for detecting partner notification in pathogen phylogenetic trees.
+We propose a non-parametric test for detecting partner notification in pathogen phylogenetic trees.
 For the simplest representative of this model family, the BD-PN model, we also derive the equations for its likelihood calculation and present a maximum likelihood parameter estimator.
@@
 internal nodes) correspond to pathogen transmissions (Fig.~\ref{fig:tt}).
-The tree branches are measured in units of time, where $T$ is the time that passed between the tree root (the beginning of the (sub-)epidemic) and the last sampled tip.
+The tree branches are measured in units of time.
+Let $T$ denote the time that passed between the tree root (the beginning of the (sub-)epidemic, at time $t=0$) and the last sampled tip (at $t=T$).
 
@@
 \item \textit{effective reproduction number} $R_e = \frac{\lambda}{\psi}$, expected number of individuals directly infected by an infectious case;
-\item \textit{infectious time} $\frac{1}{\psi}$, expected time during which an infectious individual can further spread the epidemic.
+\item \textit{infectious time} $=\frac{1}{\psi}$, expected time during which an infectious individual can further spread the epidemic.
 \end{itemize}
@@
 MTBD models~\citep{Stadler2013a} add population structure to the BD model by allowing different individual states, transmissions between them and state changes.
-A general MTBD model with $m$ individual states has $2m^2 + m$ parameters: An individual in state $k \in \{1, \ldots, m\}$ can be removed at a constant average rate $\psi_k$ (with pathogen sampling probability $\rho_k$), change their state to state $l \in \{1, \ldots, m\}$ at a constant average rate $\mu_{kl}$ (where $\mu_{kk} = 0$), and transmit their pathogen to an individual in state $l$ at a constant average rate $\lambda_{kl}$.
+A general MTBD model with $m$ individual states has $2m^2 + m$ parameters: An individual in state $k \in \{1, \ldots, m\}$ can be removed at a constant average rate $\psi_k$ (with pathogen sampling probability $\rho_k$), change their state to state $l \in \{1, \ldots, m\}$ at a constant average rate $\mu_{kl}$ ($k \neq l$), and transmit their pathogen to an individual in state $r \in \{1, \ldots, m\} $ at a constant average rate $\lambda_{kr}$.
 The time between events % of the same type
-is hence modeled with exponential distribution.
+is hence modeled with exponential distributions.
 
-\subsection{PN extension}
+\subsection{Partner Notification extension}
 
@@
 In the simplest case, the BD-PN model, at the moment of sampling the sampled individual might notify their most recent partner with a constant probability $\upsilon$.
-Upon notification, the partner is removed almost instantaneously (modeled via a constant notified removal rate $\phi >> \psi$).
+The notification implies that four conditions are met: (i) the sampled individual remembers their last partner and accepts to either notify the partner on their own or to give the partner's contacts to the health officials for assisted notification; (ii) a pathogen transmission happened at the moment of contact (either the sampled individual infected the partner or the other way round); (iii) the partner is successfully reached and notified; (iv) unless the partner already became uninfectious by other means by the moment of notification (e.g.
+got diagnosed though the standard procedure), they get tested upon notification.
+The parameter $\upsilon$ hence regroups a complex set of conditions, which would hardly be identifiable separately from only sequence data.
+Upon notification, the partner's pathogen is sampled almost instantaneously (modeled via a constant notified sampling rate $\phi >> \psi$).
+Upon sampling, the partner might notify in their turn.
 BD-PN hence has 3 classical BD and 2 additional parameters:
@@
  \item $\upsilon$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
- \item $\phi >> \psi$ -- constant notified removal rate: a notified individual (if not yet removed via standard procedure) stops being infectious and gets observed with this rate.
+ \item $\phi >> \psi$ -- constant notified sampling rate: a notified individual (if not yet removed via standard procedure) gets observed with this rate.
 This creates an observed tip in the transmission tree.
@@
 
-\item notified individuals are always observed upon removal;\\
+\item notified individuals will eventually get sampled (i.e., their sampling is modeled with a rate $\upsilon$ instead of a combination of a removal rate $\psi$ and a sampling probability $\rho$);\\
 
@@
 %\item notified individuals do not notify further;
-\item after the notification, notified individuals do not transmit further (but might notify in their turn);\\
+\item once notified, the partners do not transmit further;\\
 
-We assume here that the diagnosis and follow-up process works as expected, so that the notified patients are either isolated or provided with treatment or other means of preventing further transmission.
-Note that the most recent partner relationship is not necessarily symmetrical.
-The most recent partner $j$ of individual $i$ might have transmitted their virus further (to individual $k$) after the contact with $i$ and before the notification by $i$.
-Hence, while $j$ is the most recent partner of $i$, $j$'s most recent partner is $k$, and, once notified, $j$ might notify $k$.
+We assume here that once aware of the potential infection, the notified patients take care to avoid further transmission.
+Moreover, we assume that the diagnosis and follow-up process works as expected, so once the notified patients are sampled, they are either isolated or provided with treatment or other means of preventing further transmission.
 
@@
 to up to a certain fixed number of most recent partners, or all the partners within a certain time interval) in the future work.
-It is however useful to make the mathematics behind the model manageable.
+It is however useful to make the mathematics and likelihood calculation under the model manageable.
 In the case of sexually transmitted diseases, notifying only the most recent partner could roughly correspond to notifying the ``main'' partner (e.g.
 spouse).
+
+Note that the ``most recent partner'' relationship is not necessarily symmetrical.
+The most recent partner $j$ of individual $i$ might have transmitted their virus further (to individual $k$) after the contact with $i$ and before the notification by $i$.
+Hence, while $j$ is the most recent partner of $i$, $j$'s most recent partner is $k$, and, once notified, $j$ might notify $k$.
 \end{enumerate}
@@
 It is Gillespie-based, generates state change, transmission and removal times, and only reconstructs the sampled parts of the tree to save memory and increase speed.
-We give its algorithmic details in Appendix.
+We give its algorithmic details in Appendix. The results in the following section show that the trees generated under MTBD and MTBD-PN models are different and easily distinguishable.
 
-\section{PN test}~\label{sec:test}
-To access whether a given tree is generated in a presence of partner notification, we developed a non-parametric test based on cherries (two-tip clades).
+\section{Testing for presence of Partner Notification}~\label{sec:test}
+To access whether a given tree reveals the presence of partner notification, we developed a non-parametric test based on cherries (two-tip clades).
 
@@
 Hence if a transmission tree is complete, a cherry corresponds to a transmission from an infected individual to their most recent partner, followed by sampling of both.
-If there is no partner notification, these two sampling events are independent, while in the presence of partner notification the first sampling might trigger the second one, and they will be close in time.
-For an incomplete transmission tree (where some infected individuals were not sampled), the only difference would be that some cherries might include hidden transmissions.
+If there is no partner notification, these two sampling events are independent, while in the presence of partner notification the first sampling might trigger the second one, and they will be close in time (e.g., the second cherry in Fig.~\ref{fig:tipswap}a).
+For an incomplete transmission tree (where some infected individuals were not sampled), the difference would be that some cherries might include hidden transmissions.
+The notification pattern should however still be observed for some of the cherries.
 
@@
 For each cherry $i\;(1 \leq i \leq N)$ with tips $i0$ and $i1$, we calculate the difference between its tip times $d_i = |t_{i0} - t_{i1}|$.
-We then generate a collection of reshuffled cherries by (1) ordering the cherries by their root dates; (2) for each cherry $i$ randomly selecting one of its tips $i* \in \{i0, i1\}$; (3) swapping the selected tips between the neighboring cherries: the tip $\{2k\}*$ gets swapped with the tip $\{2k-1\}*$ ($1 \leq k \leq N / 2$) as illustrated in Fig.~\ref{fig:tipswap}.
+We then generate a collection of reshuffled cherries by (1) ordering the cherries by their root dates; (2) for each cherry $i$ randomly selecting one of its tips $i* \in \{i0, i1\}$; (3) swapping the selected tips between the neighboring cherries: the tip $2i*$ gets swapped with the tip $2i-1*$ ($1 \leq i \leq N / 2$) as illustrated in Fig.~\ref{fig:tipswap}.
 If the total number of cherries N is odd, instead of swapping the selected tips of the last two cherries, we swap the selected tips of the last three cherries in a cycle (see Fig.~\ref{fig:tipswap}).
-We then calculate the time differences $d_{i'}$ between the tips of these reshuffles cherries.
+We then calculate the time differences $d_{i'}$ between the tips of these shuffles cherries.
 The test then compares $d_i$ to $d_{i'}$: $sign(d_{i'} - d_i) = 1$ if $d_{i'}$ is larger than $d_i$; $= -1$ if it is smaller; or $= 0$ if they are equal (this could happen as in real datasets dates are often truncated, e.g.
@@
 Note that if cherry tip branch lengths are independently drawn from the same distribution, $d_{i'}$ will be smaller than $d_i$ for about half of the cherries.
-However, if some cherries are ``notified'' their real tip differences should be systematically smaller than reshuffled ones.
+However, if some cherries are ``notified'' their real tip differences should be smaller than reshuffled ones.
 We therefore output the pvalue calculated by the sign test.
@@
 
-\subsection*{Code availability}
 We implemented the PN-test in Python 3. It is available as a part of BDPN parameter estimator (see section~\ref{sec:sim}).
@@
 
+\subsection{Master equation and tree likelihood under BD model}
 The BD model can be described with master equations representing the likelihood densities of evolving as on the observed transmission tree, with time $t$ going forward from the root ($t=0$) till the time of the last sampled tip ($t=T$).
@@
 
+\subsection{Master equation and tree likelihood under BD-PN model}
+
+\subsubsection{Tree branches under BD vs BD-PN models}
+
 In the classical BD model with incomplete sampling, a branch of the tree corresponds to a virus evolution between the transmission corresponding to the node at the beginning of this branch and either a sampling event (if the branch is external) or a transmission corresponding to the node at the end of the branch (if it is internal).
@@
 These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.
-$p^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along such a tree branch (with potentially different individuals at times $t$ and (the branch end) $t_i$).
+$p^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along such a tree branch, integrating over all the possibilities, starting from no hidden transmission along the branch and including any number of hidden transmissions with either donor or recipient tree staying unsampled (and hence potentially different individuals at times $t$ and the branch end $t_i$).
 
-In trees generated by the BD-PN model, however, two other types of branches are possible.
-Let us consider a tip corresponding to a person A who got sampled at time $t_A$ and successfully notified their partner B (see Fig.~\ref{fig:pn-branches}).
-According to our model assumption 2, notified individuals are always observed.
-Hence B must be sampled too (at time $t_B \geq t_A$) and corresponds to a tree tip, while the common ancestor of A and B (AB) corresponds to the transmission between A and B.
-Therefore the individual on the path between AB and B cannot change (is always B) and only hidden transmissions (from B to someone else) where the recipient subtree stayed unobserved are possible along the branch(es) leading from AB to B (before $t_A$).
-Note, that non-hidden transmissions are also possible before the notification time $t_A$, and in such a case we know which subtree corresponds to the donor (the one that contains B).
-Between $t_A$ and $t_B$ no transmission is possible on the branch of B as, according to assumption 3, after notification notified individuals do not transmit further.
+In trees generated by the BD-PN model, however, the individuals do not behave in the same way when they are notified or not and, hence, other types of branches are possible.
+As in a real tree we do not know who might have notified whom, we will have to integrate over all the possibilities to calculate the tree likelihood.
+We will do so calculating the likelihood density of a tree using a pruning algorithm~\citep{10.1093/sysbio/22.3.240}, while considering all the combinatorial possibilities for each node.
+For instance, for each tree tip we will consider four possibilities: whether the corresponding individual was notified or not by someone, and on top of that, whether the tip's individual notified or not their most recent partner upon sampling.
+Before describing the combinations behind tree likelihood calculation, we will introduce additional tree branch types.
 
-We will use an \textbf{oriented (partner) branch probability} $p_o^{(i)}(t)$ for cases when the individual at the end of the branch (or considered time interval) is the same as at time $t$ but might have transmitted to someone else (unobserved) along the way:
+\subsubsection{Additional master equations under BD-PN models}
+
+Let us consider a person A who got sampled at time $t_A$ and successfully notified their partner B, who as a consequence got sampled at time $t_B \geq t_A$.
+A and B hence correspond to tree tips, while their most recent common ancestor node AB corresponds to the transmission between A and B (see Fig.~\ref{fig:pn-branches}(1a)).
+
+\begin{figure*}[tbhp]
+\centering 
+%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
+\includegraphics[width=1\textwidth]{Fig_PNbranches}
+\caption{An example of partner notification in a transmission tree: an individual A got sampled at time $t_A$ (violet tip) and notified their most recent partner B (orange), where B is either observed (\textbf{1}) or hidden (\textbf{2-3}).
+\\
+The (observed as in \textbf{1-2} or hidden as in \textbf{3}) internal node AB corresponds to the transmission between A and B.
+The fact that A is the notifier of B implies that there is no hidden event along its branch AB-A (coloured violet), and that in any transmission on the path AB-B (coloured orange) the donor is B.
+Note, that since the most-recent-partner relationship is not always symmetric, B might be the most recent partner of more than one individual and hence might have got notified by someone else too (e.g.
+C in \textbf{b}, who in \textbf{1b,2b} was infected by B, and in \textbf{3b} could be infected by or infect B, who then infected A).
+Moreover, between the moment of B's first notification ($t_A$ for \textbf{a}, $t_C$ for \textbf{b}) and the moment of sampling  of B ($t_B$) there are no hidden events on the branch B (coloured dark red).
+Finally, B might have been removed via a standard procedure before getting notified by their partner(s) (\textbf{c}).\\
+In the panels \textbf{3} B's subtree is fully hidden, and hence the transmission node AB is also unobserved, therefore the A's branch is composed of two parts, where the bottom part (violet) corresponds to A and contains no hidden transmissions, while the top part could correspond or not to A (e.g in \textbf{3b} it corresponds to B) and might contain hidden transmissions.
+\\
+In the panels \textbf{2-3a-b} B is hidden as by the end of the sampling period ($T$) B is not yet sampled, while in \textbf{2-3c} B is hidden as B got removed via the standard procedure without sampling before any notification.}
+\label{fig:pn-branches} 
+\end{figure*}
+
+According to our model assumption 4, only the most recent partner can get notified.
+Hence, AB must be the parent node of A, and no hidden transmission is possible along the branch AB-A (as otherwise there would have been a partner of A who is more recent than B).
+
+We will use a \textbf{notifier branch probability} $p_n^{(A)}(t)$ for such cases (when no hidden transmission occurred along the branch A, coloured violet in Fig~\ref{fig:pn-branches}):
 
@@
 \begin{cases}
-\dot{p}_o^{(i)}(t) &=~~  \Big(\lambda + \psi\Big) p_o^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-    &~~~ - \lambda p_o^{(i)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
-p_o^{(i)}(t_i) &=  1\label{eq:p-o}
+\dot{p}_n^{(A)}(t) &=  \Big(\lambda + \psi\Big) p_n^{(A)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+p_n^{(A)}(t_A) &= 1\label{eq:p-nh}
 \end{cases}
@@
 
-We will use a \textbf{notified partner branch probability} $p_p^{(i)}(t)$ for cases when the individual at the end of the branch is the partner who got notified at time $t_n < t$:
+On the branch of B, no transmission is possible between the moment of notification $t_A$ and the moment of B's sampling $t_B$ as, according to assumption 3, once notified partners do not transmit further.
+We will use a \textbf{notified partner branch probability} $p_p^{(i)}(t)$ for cases when the individual at the end of the branch is the partner who got notified at time prior to $t$ (coloured dark red in Fig~\ref{fig:pn-branches}):
 
@@
 \begin{cases}
-\dot{p}_p^{(i)}(t) &=  \phi p_p^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no sampling in the next infinitesimal time $\Delta t$ }\\
-p_p^{(i)}(t_i) &=  1\label{eq:p-np}
+\dot{p}_p^{(B)}(t) &=  \phi p_p^{(B)}(t)\; \textit{\color{gray} $\leftarrow$ no sampling in the next infinitesimal time $\Delta t$ }\\
+p_p^{(B)}(t_B) &=  1\label{eq:p-np}
 \end{cases}
@@
 
-Moreover, according to our model assumption 4, only the most recent partner can get notified.
-Hence, AB must be the parent node of A, and no hidden transmission is possible along the branch AB-A (as otherwise there would have been a partner of A who is more recent than B).
-
-We will use a \textbf{notifier branch probability} $p_n^{(i)}(t)$ for such cases (when no hidden transmission occurred along the branch):
+As B corresponds to A's partner, the individual on the path between the transmission node AB and the tip B cannot change (is always B) and before the notification time $t_A$ only transmissions from B to someone else are possible along the path leading from AB to B.
+These transmissions can be either hidden (where the recipient tree stayed unobserved) or observed: in such a case we know which subtree corresponds to the donor (the one that contains B).
+We will use an \textbf{oriented (partner) branch probability} $p_o^{(B)}(t)$ for cases when the individual at the end of the branch (or considered time interval, here $t_A$) is the same as at time $t$ but might have transmitted to someone else (unobserved) along the way (coloured orange in Fig~\ref{fig:pn-branches}):
 
@@
 \begin{cases}
-\dot{p}_n^{(i)}(t) &=  \Big(\lambda + \psi\Big) p_n^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-p_n^{(i)}(t_i) &= 1\label{eq:p-nh}
+\dot{p}_o^{(B)}(t, t_A) &=~~  \Big(\lambda + \psi\Big) p_o^{(B)}(t, t_A)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+    &~~~ - \lambda p_o^{(B)}(t, t_A)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission with hidden recipient subtree}\\
+p_o^{(B)}\Big(min(t_A, t_B), t_A\Big) &=  1\label{eq:p-o}
 \end{cases}
@@
 
-Note that A could have also ``unsuccessfully'' notified B, if by the time $t_A$, when A decided to notify B, B had already got sampled via the standard procedure or got notified by someone else.
-In this case nothing changes for the AB-A branch, and the AB-B branch(es) stay oriented.
-
-\begin{figure*}[tbhp]
-\centering 
-%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
-\includegraphics[width=0.6\textwidth]{Fig_PNbranches}
-\caption{An example of partner notification in a transmission tree: an individual A got sampled at time $t_A$ (violet tip) and notified their most recent partner B (orange tip).
-The internal node AB corresponds to the transmission between A and B.
-The fact that A is the notifier of B implies that there is no hidden event along its branch AB-A, and that in any transmission on the path AB-B the donor is B.
-Moreover, between the moment of notification ($t_A$) and the moment of sampling  of B ($t_B$) there are no hidden events on the branch B.}
-\label{fig:pn-branches} 
-\end{figure*}
-
-The equations~\ref{eq:Us}, \ref{eq:p}, \ref{eq:p-o}-\ref{eq:p-nh} have the following closed form solutions:
+The equations~\ref{eq:Us}, \ref{eq:p}, \ref{eq:p-nh}-\ref{eq:p-o} have the following closed form solutions:
 \begin{equation}
@@
 &\begin{cases}
-U(t) = \frac{\lambda + \psi}{2\lambda} +  \frac{c_1}{2\lambda}\Big(\frac{E_1 - 1}{E_1 + 1}\Big)\\
+U(t) = \frac{\lambda + \psi}{2\lambda} +  \frac{c_1}{2\lambda}\Big(\frac{E_1(t) - 1}{E_1(t) + 1}\Big)\\
 \\
-p^{(i)}(t) = \Big(\frac{E_2 + 1}{E_1 + 1}\Big)^2E_3 \\
+p^{(i)}(t) = \Big(\frac{E_2(t_i) + 1}{E_1(t) + 1}\Big)^2E_3(t, t_i) \\
 \\
-p_o^{(i)}(t) =  \frac{E_2 + 1}{E_1 + 1}\sqrt{E_3E_4} \\
+p_o^{(i)}(t, t_r) =  \frac{E_2\big(min(t_i, t_r)\big) + 1}{E_1(t) + 1}\sqrt{E_3\Big(t, min(t_i, t_r)\Big)E_4\Big(t, min(t_i, t_r)\Big)} \\
 \\
-p_n^{(i)}(t) =  E_4
+p_n^{(i)}(t) =  E_4(t, t_i)
 \\
@@
 & \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi},\\
-& ~~~~~~~~E_1 = c_2 e^{c_1 (t -T)},\; E_2 = c_2 e^{c_1 (t_i -T)},\; E_3 = e^{c_1 (t -t_i)}, \; E_4 =  e^{(\lambda + \psi) (t -t_i)}\\
+& ~~~~~~~~E_1(t) = c_2 e^{c_1 (t -T)},\; E_2(t_i) = c_2 e^{c_1 (t_i -T)},\; E_3(t, t_i) = e^{c_1 (t -t_i)}, \; E_4(t, t_i) =  e^{(\lambda + \psi) (t -t_i)}\\
 \end{split}\label{eq:ps}
+\end{equation}
+
+Note that on top of the scenario we just described (shown in Fig.~\ref{fig:pn-branches}(1a)), two other scenarios are possible for a person A who once sampled (at $t_A$) notified their most recent partner B, who got sampled at time $t_B$:
+\begin{enumerate}
+
+\item[1b.] by the time $t_A$, B was already notified by someone else (at time $t_C < t_A$, see Fig.~\ref{fig:pn-branches}(1b)).
+Note, that once notified partners do not transmit further, hence $t_C > t_{AB}$. \\
+
+This would lead to a shorter oriented part on the B branch (finishing at $t_C$), and a longer notified-partner part (between $t_C$ and $t_B$).
+
+\item[1c.] by the time $t_A$, B was already sampled via a standard procedure (at time $t_B < t_A$, hence the branch B would finish before the branch A, as in Fig~\ref{fig:pn-branches}(1c)).
+
+This would lead to an oriented part on the B branch finishing at $t_B$ and no notified-partner part. Note that the branches between AB and B stay oriented as B remains the most recent partner of A.
+\end{enumerate}
+
+Finally, B could also stay unsampled: either because once notified (by A or another partner, who notified them before A did) B has not yet got sampled (Fig~\ref{fig:pn-branches}(2a-b,3a-b)); or because B got removed via the standard procedure without sampling before any notification (Fig.~\ref{fig:pn-branches}(2c,3c)).
+Moreover, between the transmission AB and the moment of B's notification/removal B might have transmitted to someone else (e.g.
+C in Fig.~\ref{fig:pn-branches}(2), BC denoting the transmission).
+If that happened and someone in BC's subtree got sampled, the branch AB-A would still correspond to the notifier A, while the other AB's child branch would partially correspond to B (between AB and BC) and partially to C (starting at BC), as in Fig.~\ref{fig:pn-branches}(2).
+Otherwise (if there was no such transmission or all such subtrees stayed unobserved, as in Fig.~\ref{fig:pn-branches}(3)), the node AB will not be observed either, and A's branch will get extended with its parent's branch.
+Hence only the bottom part (after AB) of the A's branch would correspond to the notifier probability (see Fig.~\ref{fig:pn-branches}(3))).
+
+Note that if A is a notifier, A is always observed, as notification only happens upon sampling.
+
+As in the real trees the sampling is incomplete, and we do not know sampled individual's status (individual who did not notify, notifier, partner notified in time or not), during the tree likelihood calculation, we will have to consider all these scenarios.
+
+In particular, we will use the probability of a \textbf{hidden partner subtree} $U_p(t, t_A)$, which started at time $t$ on a partner branch and evolved till time $T$ without any of its individuals being sampled, provided that the first of the partner's notifiers tried to notify them at time $t_A$ (B's branch in Fig.~\ref{fig:pn-branches}(2-3)).
+This probability consists of two parts: either the partner got notified at time $t_A$ but has not yet got sampled (like in Fig.~\ref{fig:pn-branches}(2a-b,3a-b)), $U_p^{(1)}(t, t_A)$;  or the partner got removed via the standard procedure before the time $t_A$ (like in Fig.~\ref{fig:pn-branches}(2c,3c)), $U_p^{(2)}(t, t_A)$:
+\begin{equation}
+U_p(t, t_A) = U_p^{(1)}(t, t_A) + U_p^{(2)}(t, t_A).
+\end{equation}
+
+The first part can be easily expressed via the previously defined equations: oriented (partner) branch evolution till time $t_A$ followed by notified partner branch evolution till time $T$.
+Note, that neither  $p_o^{(x)}(t, t_A)$ not $p_p^{(x)}(t)$ include an event at the end of the branch, and in this case the partner branch does not end at $t_x = T$:
+\begin{equation}
+U_p^{(1)}(t, t_A) = p_o^{(x)}(t, t_A) p_p^{(x)}(t)\textit{\color{gray}, where $t_x =T$}
+\end{equation}
+
+For the second part we would need to integrate over all possible partner removal times $t_B \in [t, t_A[$, provided that in the interval between $t$ and $t_B$ the partner might have transmitted to someone else (potentially multiple times), whose subtree(s) stayed unobserved till $T$.
+Note that $p_o^{(x)}(t, t_A)$ describes a partner branch evolution between $t$ and $t_A$ with any number of hidden transmissions, no removal along it, and no event at the end of the branch.
+What we need instead is a removal with no sampling at time $\tau$.
+To approximate $U_p^{(2)}(t, t_A)$ we will divide $p_o^{(x)}(t, t_A)$ by the probability of no removal during the time $(t_A - t)$ (i.e.
+$e^{-\psi(t_A - t)}$), multiply it by the probability of removal during that time ($1 - e^{-\psi(t_A - t)}$), and by the probability of not being sampled upon removal ($1 - \rho$):
+
+\begin{equation}
+U_p^{(2)}(t, t_A) \approx p_o^{(x)}(t, t_A) \frac{1 - e^{-\psi(t_A - t)}}{e^{-\psi(t_A - t)}} (1 - \rho)\textit{\color{gray}, where $t_x =T$}
+\end{equation}
+
+It is an approximation as instead of integrating over all possible intervals $[t, t_B[$ for potential hidden transmissions from the partner branch, we use the interval $[t, t_A[$.
+
+We will also need to describe the probability of a\textbf{ mixed branch} $p_{m(s1,s2)}^{(i)}(t,t_h,t_r)$ for branches that include a hidden partner subtree (AB-C branches in Fig.~\ref{fig:pn-branches}(2) and A's tip branch in Fig.~\ref{fig:pn-branches}(3)).
+$s1$ here denotes the type of the top of the branch (e.g.
+between AB and BC in Fig.~\ref{fig:pn-branches}(2), partner) and $s2$ denotes the type of the bottom of the branch (e.g.
+between AB and A in Fig.~\ref{fig:pn-branches}(3), index).
+$t_h$ denotes the time of the hidden partner subtree start, while $t_r$ denotes the first notification time of the hidden partner.
+For instance, in Fig~\ref{fig:pn-branches}(2a) $t_i=t_C$, $t_r=t_A$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(2b) $t_i=t_r=t_C$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(3c) $t_i=t_r=t_A$, $t_h=t_{AB}$.
+
+$p_{m(s1,s2)}^{(i)}(t,t_h,t_r)$ combines three elements: the probability of the top part of the branch finishing with a transmission, the probability of the bottom part of the branch (without taking into account the event at its end) and the probability of the hidden partner subtree:
+
+\begin{equation}
+\begin{split}
+p_{m(s1,s2)}^{(i)}(t,t_h,t_r) &= p_{top=s1}^{(h)}(t)p_{bottom=s2}^{(i)}(t_h)U_p(t_h,t_r)\textit{\color{gray},}\\
+&\textit{\color{gray}where $t_h$ is the time of the hidden partner subtree start,}\\
+&\textit{\color{gray}~~~~~~~~$t_r$ is the first notification time of the hidden partner.}\\
+\end{split}
+\end{equation}\label{eq:mixed}
+
+Note that the bottom part of the branch can either correspond to a notifier (as in Fig~\ref{fig:pn-branches}(2b,3)), or to a standard branch, or to a mixed tip branch starting as a standard branch and finishing as a notifier.
+It cannot correspond to another partner branch as every partner has their notifier sampled, and notifiers only notify one partner (who is already hidden along this branch).
+The bottom branch part evolution can then be expressed with the previously defined equations:
+
+\begin{equation}
+p_{bottom=s2}^{(i)}(t_h) = 
+\begin{cases}
+p_n^{(i)}(t)&\textit{\color{gray} if s2 is a notifier,}\\
+p_{m(-,n)}^{(i)}(t)&\textit{\color{gray} if i is a tip, and s2 type is a standard-notifier mix,}\\
+p^{(i)}(t)&\textit{\color{gray} otherwise (standard state)}
+\end{cases}
+\end{equation}
+
+Finally, the top part of the branch can be in three possible states.
+The first possibility is that it corresponds to the hidden partner's individual.
+This happens if (one of) the hidden partner's notifier(s) does not correspond to the branch's end (as in Fig~\ref{fig:pn-branches}(2,3b)).
+As this notifier is observed, the transmission between them and the partner is also observed and is ancestral to the node at the end of the current branch.
+All the branches on the way connecting this notifier and the partner (including the top part of the current branch) must then correspond to the partner.
+The other possibility is that the only partner's notifier corresponds to the tip at the end of the current branch (as in Fig~\ref{fig:pn-branches}(3a,c)).
+Then the branch can either be in a standard state (unnotifier non-notifier), or a mix between a (another, partially hidden) partner branch and a standard state (corresponding to that other partner's recipient).
+In the first case we will approximate $p_{top=o}^{(h)}(t)$ using $p_o^{(h)}(t)$ in the following way.
+$p_o^{(h)}(t)$ expresses a probability of a partner evolving along the branch between the times $t$ and $t_h$ with no or any number of oriented hidden transmissions, without taking into account the event at the end of it.
+In our case there must be at least one hidden transmission: from the partner at the end of the branch (where the donor tree stayed unsampled).
+If we remove the probability of no event during this time ($e^{-(\lambda + \psi)(t_h - t)}$) from $p_o^{(h)}(t)$, we will obtain a probability of a partner branch with at least one oriented hidden transmission somewhere between $t$ and $t_h$ (the probability of the corresponding hidden tree being included).
+$p_{top=s1}^{(h)}(t)$ expresses almost the same thing, with the difference being that the last hidden transmission must have happened at time $t_h$ and the probability of the hidden partner tree corresponding to it is not included.
+We will hence approximate it as follows:
+\begin{equation}
+p_{top=o}^{(h)}(t) \approx \frac{p_o^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}}{U(t_{h})}
+\end{equation}
+
+We will treat the case when the branch is in the standard state in a similar way (using the standard branch probability $p^{(h)}(t)$ instead of the partner one $p_o^{(h)}(t)$):
+\begin{equation}
+p_{top=-}^{(h)}(t) \approx \frac{p^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}}{U(t_{h})}
+\end{equation}
+
+We will process the mixed case in the same way as the previous two:
+\begin{equation}
+p_{top=m(o,-)}^{(h)}(t) \approx \frac{p_{m(o,-)}^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}}{U(t_{h})}
+\end{equation}
+
+Note that we wrote down equation~(\label{eq:mixed}) as if the start of the hidden partner tree $t_h$ was known.
+In practice, it is not the case as we need to integrate over all the possible times between the branch start $t_{parent(i)}$ and either its end or the first partner's notification (if it was before): $min(t_r,t_i)$.
+To simplify the calculations, we will simply take the time in the middle of this interval instead:
+
+\begin{equation}
+\begin{split}
+p_{m(s1,s2)}^{(i)}(t,t_r) &\approx p_{s1}^{(h)}(t)\Big(p_{s2}^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}\Big)\frac{U_p(t_h,t_r)}{U(t_{h})}\textit{\color{gray},}\\
+&\textit{\color{gray}where $t_h \approx t + \frac{min(t_r, t_i) - t}{2}$,}\\
+&\textit{\color{gray}~~~~~~~~$t_r$ is the first notification time of the hidden partner.}\\
+\end{split}\label{eq:p_mixed}
 \end{equation}
@@
 
-Using the formulas~(\ref{eq:ps}), we can calculate the likelihood density of a tree using a pruning algorithm~\citep{10.1093/sysbio/22.3.240}.
-For each visited node $i$ we calculate the likelihood density of its subtree for two cases depending on their receiving-a-notification status (shown with the first superscript, $^{\not{n}}$ or $^n$).
-In the first case,  $l^{(\not{n})}(i)$, the branch $ji$ connecting $i$ it to its parent node $j$ corresponds to an unnotified individual.
-In the second case, $l^{(n)}(i, r)$, the branch $ji$ corresponds to an (eventually) notified partner (notified by the individual corresponding to a tip $r$ at time $t_r$).
-If $i$ is a tip, we additionally consider whether the corresponding individual notified their most recent partner or not (shown with the second superscript, $^n$ or $^{\not{n}}$).
-For instance, $l^{(\not{n},n)}(i)$ represents the case where the tip $i$ was not notified but notified their partner.
-Note that if the tip $i$ was notified by their partner $r$, we need to consider two possibilities: (1) when $t_r \leq t_i$, and hence the notification was successful at provoking $i$'s diagnostics and sampling; and (2) when $t_r > t_i$, and hence by the time of notification by $r$ $i$ already got sampled by other means.
+Using the formulas~(\ref{eq:ps}) and (\ref{eq:p_mixed}), we can calculate the likelihood density of a tree using a pruning algorithm~\citep{10.1093/sysbio/22.3.240}.
+For each visited node $i$ we calculate the likelihood density of its subtree including the branch $ji$ connecting $i$ to its parent node $j$ for two cases depending on $i$'s receiving-a-notification status (shown with the first superscript, $^{\not{n}}$ or $^n$).
+In the first case,  $l^{(\not{n})}(i)$, the node $i$ corresponds to an unnotified individual.
+In the second case, $l^{(n)}(i, r)$, the branch $ji$ and the node $i$ correspond to an (eventually) notified partner (notified by the individual corresponding to a tip $r$ at time $t_r$).
+If $i$ is a tip, we additionally consider whether there is an observed partner in our tree, whom the individual $i$ notified  (shown with the second superscript, $^n$ or $^{\not{n}}$).
+For instance, $l^{(\not{n},n)}(i)$ represents the case where the tip $i$ was not notified but notified their most recent partner (who is observed).
+%Note that if the tip $i$ was notified by their partner $r$, we need to consider two possibilities: (1) when $t_r \leq t_i$, and hence the notification was successful at provoking $i$'s diagnostics and sampling; and (2) when $t_r > t_i$, and hence by the time of notification by $r$ $i$ already got sampled by other means.
 
@@
 & \textit{\color{gray} (standard branch evolution followed by sampling and no notification)}\\
+&~~+p_{m(-,n)}^{(i)}(t_j,t_i)\psi\rho\upsilon \textit{\color{gray} ~i.e. $i$ was not notified but notified their partner who is hidden}\\
+& \textit{\color{gray} (mixed standard-notifier branch evolution followed by sampling and notification)}\\
 \\
 l^{(\not{n},n)}(i) &= p_n^{(i)}(t_j)\psi\rho
-\upsilon \textit{\color{gray} ~i.e. $i$ was not notified but notified their most recent patner}\\
+\upsilon \textit{\color{gray} ~i.e. $i$ was not notified but notified their most recent patner (observed)}\\
 & \textit{\color{gray} (notifier branch evolution followed by sampling and notification)}\\
+&~~\\
 \\
@@
 \textit{\color{gray} (oriented (partner) branch evolution followed by standard sampling and no notification)}\\
+~~~+p_{m(o,n)}^{(i)}(t_j,t_i)\psi\rho\upsilon \textit{\color{gray} ~i.e. $i$ got notified at $t_r > t_i$, and notified a hidden partner}\\
+ \textit{\color{gray} ~~~(mixed oriented-notifier branch evolution followed by sampling and notification)}\\
 \\
@@
 \textit{\color{gray} (oriented (partner) branch evolution followed by} \\
-\textit{\color{gray}~notified partner branch evolution,} \\
-\textit{\color{gray}~notification-provoked sampling and no notification)}\\
+\textit{\color{gray}~notified partner branch evolution, notification-provoked sampling and no notification)}\\
+~~~+p_{m(o,n)}^{(r)}(t_j,t_i)p_p^{(i)}(t_r)\phi\upsilon \textit{\color{gray} ~i.e. $i$ got notified at $t_j \leq t_r \leq t_i$, and notified a hidden partner}\\
+ \textit{\color{gray} ~~~(mixed oriented-notifier branch evolution followed by} \\
+\textit{\color{gray}~~~~notified partner branch evolution, notification-provoked sampling and notification)}\\
 \end{cases}\\
@@
 Hence the $i$'s branch evolution is represented by $p^{(i)}(t_j)$ with a transmission event at the end of it, where each of the child branches can correspond to the donor (probability density of $2\lambda$).
-We will consider three possibilities: (1) when both $i_0$ and $i_1$ are internal nodes; (2) when both $i_0$ and $i_1$ are tips; and (3) when one of them is an internal node and the other one is a tip.
+We will consider three possibilities: (2.1.1) when both $i_0$ and $i_1$ are internal nodes; (2.1.2) when both $i_0$ and $i_1$ are tips; and (2.1.3) when one of them is an internal node and the other one is a tip.
 
@@
 l^{(\not{n})}(i) = p^{i}(t_j) 2\lambda \cdot 
-\Big(&~~~~l^{(\not{n},\not{n})}(i_0)l^{(\not{n},\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ neigher $i_0$ nor $i_1$ notified} \\
+\Big(&~l^{(\not{n},\not{n})}(i_0)l^{(\not{n},\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ neigher $i_0$ nor $i_1$ notified} \\
 &+ l^{(\not{n},n)}(i_0)l^{(n,\not{n})}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $i_0$ notified $i_1$ and $i_1$ did not notify} \\
 &+ l^{(n,\not{n})}(i_0,i_1)l^{(\not{n},n)}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ did not notify, but $i_1$ notified $i_0$} \\
-&+ l^{(n,n)}(i_0,i1)l^{(n,n)}(i_1,i_0)~~\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ and $i_0$ notified each other} \\
+&+ l^{(n,n)}(i_0,i1)l^{(n,n)}(i_1,i_0)~\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ and $i_0$ notified each other} \\
 \end{split}\label{eq:lu-tip-tip}
@@
 l^{(\not{n})}(i) = p^{i}(t_j) 2\lambda \cdot 
-\Big(&~~~l^{(\not{n},\not{n})}(i_0)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ did not notify $i_1$ } \\
-&+ l^{(\not{n},n)}(i_0)l^{(n)}(i_1,i_0)~~\Big) \textit{\color{gray}~$\leftarrow$ $i_0$ notified $i_1$} \\
+\Big(&~l^{(\not{n},\not{n})}(i_0)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ did not notify $i_1$ } \\
+&+ l^{(\not{n},n)}(i_0)l^{(n)}(i_1,i_0)~\Big) \textit{\color{gray}~$\leftarrow$ $i_0$ notified $i_1$} \\
 \end{split}\label{eq:lu-int-tip}
@@
 
-\subsubsection*{Case 2.2: $i$ is an (eventually notified) partner internal node}
+\subsubsection*{Case 2.2: $i$'s branch is internal and (at least initially) corresponds to an (eventually notified) partner}
 
-In the other case an internal node $i$ corresponds to a partner who will eventually get notified by a tip $r$.
-In this case the individual on the path connecting $i$ to its common ancestor $ir$ with the notifier tip $r$ (corresponding to the transmission between $r$ and $i$) stays the same (the partner).
-Hence, all the transmissions on this path are oriented and correspond to the partner transmitting to another recipient individual.
-In particular, the $i$'s branch evolution is represented by $p_o^{(i)}(t_j)$ with a transmission event at the end of it, where the donor is known (probability density of $\lambda$): The donor corresponds to the subtree containing the tip representing the sampling of the individual $i$.
+In the other case the internal node $i$'s branch starts as a partner who will eventually get notified by a tip $r$.
+There are two possibilities: either the node $i$ corresponds to the partner or the partner is hidden along the $i$'s branch and $i$ corresponds to someone within $i$'s recipient subtree (e.g., if $i=C$ in Fig.~\ref{fig:pn-branches}(2a)).
+In the first case, the $i$'s branch evolution is represented by $p_o^{(i)}(t_j)$ with a transmission event at the end of it, where the donor is known (probability density of $\lambda$): The donor corresponds to the subtree containing the tip representing the sampling of the individual $i$.
+
+In the second case, the $i$'s branch evolution is represented by $p_{m(o,-)}^{(i)}(t_j,t_r)$ with a transmission event at the end of it, where the donor is unknown (probability density of $2\lambda$): any of the subtrees can be the donor.
  %Moreover, this recipient's branch cannot be notified by our partner (as in our model notified partners do not notify further):
-Again, we will consider three possibilities for $i$'s child nodes: (1) when both $i_0$ and $i_1$ are internal nodes; (2) when both $i_0$ and $i_1$ are tips; and (3) when one of them is an internal node and the other one is a tip.
+Again, we will consider three possibilities for $i$'s child nodes: (2.2.1) when both $i_0$ and $i_1$ are internal nodes; (2.2.2) when both $i_0$ and $i_1$ are tips; and (2.2.3) when one of them is an internal node and the other one is a tip.
 
@@
 
-\subsubsection*{Case 2.2.1: $i$ is a partner internal node with internal-node children $i0$ and $i1$}
+\subsubsection*{Case 2.2.1: $i$'s branch is internal and (at least initially) corresponds to a partner with internal-node children $i0$ and $i1$}
 
-First, let's assume that $i_0$ and $i_1$ are internal. Then none of them can be a notifier (since notifiers correspond to tips), and hence exactly one of them is notified (by $r$):
+First, let's assume that $i_0$ and $i_1$ are internal.
+Then none of them can be a notifier (since notifiers correspond to tips), and hence either exactly one of them is notified by $r$, or the partner is hidden:
 
@@
 \begin{split}
-l^{(n)}(i, r) = p_o^{(i)}(t_j) \lambda \cdot
-\Big(&~~~l^{(n)}(i_0,r)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ is notified by $r$} \\
-& + l^{(\not{n})}(i_0)l^{(n)}(i_1,r)~~\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ is notified by $r$} \\
+l^{(n)}(i, r) = &p_o^{(i)}(t_j) \lambda \cdot
+\Big(~l^{(n)}(i_0,r)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ is notified by $r$} \\
+& ~~~~~~~~~~~~~~~+ l^{(\not{n})}(i_0)l^{(n)}(i_1,r)~\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ is notified by $r$} \\
+& + p_{m(o,-)}^{(i)}(t_j,t_r) 2 \lambda l^{(\not{n})}(i_0)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ hidden partner}\\
  \end{split}
@@
 
-\subsubsection*{Case 2.2.2: $i$ is a partner internal node with tip children $i0$ and $i1$} 
-Secondly, let's assume that $i_0$ and $i_1$ are tips. Then each of them could have notified the other one (in addition to one of them being notified by $r$):
+\subsubsection*{Case 2.2.2: $i$'s branch is internal and (at least initially) corresponds to a partner with tip children $i0$ and $i1$} 
+Secondly, let's assume that $i_0$ and $i_1$ are tips. Then each of them could have notified the other one (in addition to one of them potentially being notified by $r$):
 
@@
 \begin{split}
-l^{(n)}(i, r) = p_o^{(i)}(t_j) \lambda \cdot
-\Big(~~~&l^{(n,\not{n})}(i_0,first(r,i_1))l^{(\not{n},n)}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ and $i_1$ notified $i_0$, $i_0$ did not notify} \\
-&+ l^{(n,n)}(i_0,first(r,i_1))l^{(n,n)}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $r$ and $i_1$ notified $i_0$, $i_0$ notified $i_1$} \\
-&+l^{(\not{n},n)}(i_0)l^{(n,\not{n})}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ $r$ and $i_0$ notified $i_1$, $i_1$ did not notify} \\
-&+ l^{(n,n)}(i_0, i_1)l^{(n,n)}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ $r$ and $i_0$ notified $i_1$, $i_1$ notified $i_0$} \\
-&+l^{(n,n)}(i_0,r)l^{(n,\not{n})}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$ notified $i_1$, $i_1$ did not notify} \\
-& + l^{(n,\not{n})}(i_0, i_1)l^{(n,n)}(i_1,r) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_1$ notified $i_0$, $i_0$ did not notify} \\
-&+l^{(n,\not{n})}(i_0,r)l^{(\not{n},\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$ and $i_1$ did not notify} \\
-& + l^{(\not{n},\not{n})}(i_0)l^{(n,\not{n})}(i_1,r)~~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_0$ and $i_1$ did not notify}, \\
-\textit{\color{gray}~where } &\color{gray}first(a,b) = 
+l^{(n)}(i, r) = &p_o^{(i)}(t_j) \lambda \cdot
+\Big(~l^{(n,\not{n})}(i_0,first(r,i_1))l^{(\not{n},n)}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ and $i_1$ notified $i_0$, $i_0$ did not notify} \\
+&~~~~~~~~~~~~~~~+ l^{(n,n)}(i_0,first(r,i_1))l^{(n,n)}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $r$ and $i_1$ notified $i_0$, $i_0$ notified $i_1$} \\
+&~~~~~~~~~~~~~~~+l^{(\not{n},n)}(i_0)l^{(n,\not{n})}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ $r$ and $i_0$ notified $i_1$, $i_1$ did not notify} \\
+&~~~~~~~~~~~~~~~+ l^{(n,n)}(i_0, i_1)l^{(n,n)}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ $r$ and $i_0$ notified $i_1$, $i_1$ notified $i_0$} \\
+&~~~~~~~~~~~~~~~+l^{(n,n)}(i_0,r)l^{(n,\not{n})}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$ notified $i_1$, $i_1$ did not notify} \\
+&~~~~~~~~~~~~~~~+ l^{(n,\not{n})}(i_0, i_1)l^{(n,n)}(i_1,r) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_1$ notified $i_0$, $i_0$ did not notify} \\
+&~~~~~~~~~~~~~~~+l^{(n,\not{n})}(i_0,r)l^{(\not{n},\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$ and $i_1$ did not notify} \\
+&~~~~~~~~~~~~~~~+ l^{(\not{n},\not{n})}(i_0)l^{(n,\not{n})}(i_1,r)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_0$ and $i_1$ did not notify} 
+\\
+&+ p_{m(o,-)}^{(i)}(t_j,t_r) 2 \lambda \cdot
+\Big(~l^{(n,\not{n})}(i_0,i_1)l^{(\not{n},n)}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i_1$ notified $i_0$, $i_0$ did not notify} \\
+&~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~+ l^{(n,n)}(i_0,i_1)l^{(n,n)}(i_1, i_0)\textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i_1$ notified $i_0$, $i_0$ notified $i_1$} \\
+&~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~+l^{(\not{n},n)}(i_0)l^{(n,\not{n})}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i_0$ notified $i_1$, $i_1$ did not notify} \\
+&~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~+l^{(\not{n},\not{n})}(i_0)l^{(\not{n},\not{n})}(i_1)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i_0$ and $i_1$ did not notify} \\
+\\
+&\textit{\color{gray}~where } \color{gray}first(a,b) = 
 \begin{cases}
@@
 
-\subsubsection*{Case 2.2.3: $i$ is a partner internal node whose children are a tip and an internal node}
+\subsubsection*{Case 2.2.3: $i$'s branch is internal and (at least initially) corresponds to a partner, $i$'s children are a tip and an internal node}
 Lastly, let's assume that $i_0$ is a tip and $i_1$ is internal (the scenario where $i_1$ is a tip and $i_0$ is internal can be obtained by swapping the labels).
-Then $i_0$ could have notified $i_1$ (in addition to one of them being notified by $r$):
+Then $i_0$ could have notified $i_1$ (in addition to one of them being potentially notified by $r$):
 
@@
 \begin{split}
-l^{(n)}(i, r) = p_o^{(i)}(t_j) \lambda \cdot
-\Big(~~~&l^{(n,n)}(i_0,r)l^{(n)}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$  notified $i_1$} \\
-& + l^{(n,\not{n})}(i_0,r)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$  $r$ notified $i_0$, $i_0$ did not notify} \\
-& +l^{(\not{n}, n)}(i_0)l^{(n)}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ both $r$ and $i_0$ notified $i_1$} \\
-& + l^{(\not{n},\not{n})}(i_0)l^{(n)}(i_1,r)~~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_0$ did not notify}, \\
+l^{(n)}(i, r) = &p_o^{(i)}(t_j) \lambda \cdot
+\Big(~l^{(n,n)}(i_0,r)l^{(n)}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$  notified $i_1$} \\
+&~~~~~~~~~~~~~~~+ l^{(n,\not{n})}(i_0,r)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$  $r$ notified $i_0$, $i_0$ did not notify} \\
+&~~~~~~~~~~~~~~~+l^{(\not{n}, n)}(i_0)l^{(n)}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ both $r$ and $i_0$ notified $i_1$} \\
+&~~~~~~~~~~~~~~~+ l^{(\not{n},\not{n})}(i_0)l^{(n)}(i_1,r)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_0$ did not notify} 
+\\
+&+ p_{m(o,-)}^{(i)}(t_j,t_r) 2\lambda \cdot
+\Big(~l^{(\not{n},n)}(i_0)l^{(n)}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i_0$ notified $i_1$} \\
+&~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~+ l^{(\not{n},\not{n})}(i_0)l^{(\not{n})}(i_1)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ $i_0$ did not notify}\\
 \textit{\color{gray}~where } &\color{gray}first(a,b) = 
@@
 
-Note that $l^{(n)}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time is known, i.e.
-once we have climbed to an internal node corresponding to the transmission between the notifier (corresponding to one of its children, must be a tip) and the partner (the other child).
+Note that $l^{(n)}(i, r)$ and $p_{m(o,-)}^{(i)}(t_j,t_r)$ depends on the notifier $r$ and can only be calculated once the notification time $t_r$ is known, i.e.
+once we have climbed to an internal node corresponding to the transmission between the notifier $r$ (corresponding to one of its children, must be a tip) and the partner (the other child).
 The calculation is done recursively by descending into the subtree.
@@
 It estimates the BD-PN model parameters  $\Theta = (\lambda,\psi,\phi,\rho,\upsilon) \in \mathbb{R}^5$ for a forest $\mathscr{F}$ comprising $f \geq 1$ observed trees in the maximum-likelihood framework, where one of the BD parameters ($\lambda,\psi$ or $\rho$) is fixed (for identifiability reasons).
+We also implemented the classical BD model estimator, bd.
 
@@
 \subsection*{Code availability}
-The BD-PN parameter estimator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\citep{Huerta-Cepas2016} and NumPy package for array operations~\citep{harris_array_2020}.
+The BD-PN and BD parameter estimators are implemented in Python 3.
+They use ETE 3 framework for tree manipulation~\citep{Huerta-Cepas2016} and NumPy package for array operations~\citep{harris_array_2020}.
 
-It is available as a command-line program and a Python 3 package via PyPi (\href{https://pypi.org/project/bdpn}{bdpn}), and via Docker/Singularity (\href{https://hub.docker.com/r/evolbioinfo/bdpn/tags}{evolbioinfo/bdpn}).
-Its source code and the installation and usage documentation are available on GitHub at \href{https://github.com/evolbioinfo/bdpn}{github.com/evolbioinfo/bdpn}.
+They are available as a command-line program and a Python 3 package via PyPi (\href{https://pypi.org/project/bdpn}{bdpn}), and via Docker/Singularity (\href{https://hub.docker.com/r/evolbioinfo/bdpn/tags}{evolbioinfo/bdpn}).
+The source code and the installation and usage documentation are available on GitHub at \href{https://github.com/evolbioinfo/bdpn}{github.com/evolbioinfo/bdpn}.
 
@@
 We applied bdpn to each of the trees fixing the sampling probability $\rho$ to its real value (for identifiability).
-The results with the becoming-uninfectious rate $\psi$ are similar and given in the Supplementary Material.
+The results with the becoming-uninfectious rate $\psi$ fixed are similar and given in the Supplementary Material.
 
-As expected with a maximum likelihood estimator, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the real parameter values in all the settings.
+As expected with a maximum likelihood estimator, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the real parameter values in both settings.
 
-We calculated the relative error (normalized distance between the estimated and the target values: $\frac{|estimated - target|}{target}$) and the relative bias ($\frac{estimated - target}{target}$) for each parameter on each tree (Fig.~\ref{fig:sim}).
-Average relative errors were the lowest when $\psi$ or $\rho$ were fixed: $\leq 10\%$  for all the parameters.
+We calculated the relative error (normalized distance between the estimated and the target values: $\frac{|estimated - target|}{target}$) and the relative bias ($\frac{estimated - target}{target}$) for each parameter on each tree.
+For $\rho$ and $\upsilon$ absolute values were calculated instead of normalized ones.
+
+To highlight the importance of accounting for partner notification with the model, we also applied the classical BD model estimator bd to these trees.
+Additionally, we applied both BD-PN and BD model parameter estimators to a set of 100 trees with 500--1\,000 tips simulated under the BD model (see subsection~\ref{sec:test:sim} for the data set details).
+The results are shown in  Fig.~\ref{fig:sim}.
+We compared the method accuracy by a paired z-test.
+On the BD data set, bd and bdpn estimators performed equally well in estimation $R_0$ and infectious time $\frac{1}{\psi}$, with an average relative error $\leq 6\%$, and average relative bias $\leq 2\%$.
+The partner notification probability $\upsilon$ estimated on these trees by bdpn was systematically low, with the maximum value of 0.018, and mean value of 0.003.
+Its CI's 2.5\% bound hit the lower bound $0.001$ for all the trees.
+On the BD-PN data set, bdpn accuracy was significantly better (p-value $\leq 0.01$), with errors and biases $\leq 10\%$ for all parameters (vs.
+$36-40\%$ for bd).
+
+%Average relative errors were the lowest when $\psi$ or $\rho$ were fixed: $\leq 10\%$  for all the parameters.
 The worst performance was obtained with $\lambda$ being fixed: in this setting the estimator tends to underestimate $\psi$ and overestimate $\rho$.
+
 %The hardest parameter to estimate in all the settings (relative errors $\approx 15\%$, with a tendency to underestimation).
 
-As with the point estimates, CI estimates were the best when $\psi$ or $\rho$ were fixed, and the worst when $\lambda$ was fixed (see Table~\ref{tbl:ci}).
-The relative width of the CIs ($\frac{value_{97.5\%} - value_{2.5\%}}{value}$) varied between 11\% for $\lambda$ estimates and 32\% for $\phi$ estimates (in all the settings).
+The percentage of the trees for which real parameter values are with the CI estimates and relative CI widths ($\frac{CI_{97.5\%} - CI_{2.5\%}}{\textit{real parameter value}}$) are shown in Table~\ref{tbl:ci}.
+They suggest that the rate parameters are easier to estimate than the probability ones.
 
@@
 \centering 
-\includegraphics[width=1\textwidth]{Fig_errors.eps}
-\caption{Comparison of inference accuracy of bdpn on a data set of 100 trees of 500--1\,000 tips each, with either $\lambda$ (blue), $\psi$ (orange) or $\rho$ (green) parameter being fixed to its real value.
-We show the swarmplots (coloured by fixed parameter) of relative errors for each test tree and parameter, which are measured as the normalized distance between the estimate and the real value.
+\includegraphics[width=1\textwidth]{Fig_errors_p.png}
+\caption{Comparison of inference accuracy of bdpn (yellow) and bd (blue) estimators on a data set of 100 trees of 500--1\,000 tips generated under BD-PN model (top) and a data set of 100 trees of 500--1\,000 tips generated under BD model (bottom), with $\rho$ parameter being fixed to its real value.
+We show the swarmplots (coloured by estimator) of relative errors for each test tree and parameter, which are measured as the normalized distance between the estimate and the real value.
 Average relative error (and in parentheses average relative bias) are displayed for each parameter and method below their swarmplot.
@@
 \small
-\caption{Percentage of simulated trees for which the real parameter values were withing the estimated 95\% CIs. \smallskip}
-\begin{tabular}{c|ccc}
-\textbf{parameter} & \textbf{$\lambda$ fixed} & \textbf{$\psi$ fixed} & \textbf{$\rho$ fixed} \\
+\caption{Percentage of simulated BD-PN trees for which the real parameter values were withing the bdpn-estimated 95\% CIs, and the median CI width: $\frac{CI_{97.5\%} - CI_{2.5\%}}{\textit{real parameter value}}$.
+\smallskip}
+\begin{tabular}{c|cc|cc}
+\textbf{parameter} & \textbf{$\psi$ fixed} & & \textbf{$\rho$ fixed} &  \\
+ & \% within CI & median CI width & \% within CI & median CI width  \\
 \toprule 
- $\lambda$ & -- &  76\% & 96\% \\
- $\psi$ & 22\% & -- & 78\% \\
- $\rho$ & 16\% & 55\%  & -- \\
- $\phi$ & 70\% & 93\% & 94\% \\
- $\upsilon$ & 21\% & 68\% & 63\% \\
+ $\lambda$ &  87\% & 14.2\% & 96\% & 11.6\%\\
+ $\psi$ & -- & -- & 80\% & 28.6\% \\
+ $\rho$ & 66\%  & 39.1\% & -- & -- \\
+ $\phi$ & 93\% & 32.5\% & 94\% & 32.3\% \\
+ $\upsilon$ & 69\% & 21\% & 65\% & 20.7 \% \\
 \bottomrule
@@
 
-The PN-test values were below 0.05 for all the forests (value between $0.01$ and $0.04$), suggesting the presence of partner notification.
+The PN-test values were below 0.05 for all the forests (values between $3 \cdot 10^{-21}$ and $3 \cdot 10^{-13}$), strongly suggesting the presence of partner notification.
 
@@
 
-Using $\rho=0.58$, we estimated the following parameter values: $\lambda=0.53$, $\psi=0.43$, $\phi=23-59$, $\upsilon=0.006-0.007$, which corresponds to $R_e = 1.2$, infectious time $\frac{1}{\psi} = 2.3$ years, and about 6--16 days between partner notification and their sampling.
+Using $\rho=0.58$, we estimated the partner notification probability $\upsilon=0.006-0.007$, %the following parameter values: $\lambda=0.53$, $\psi=0.43$, $\phi=23-59$, $\upsilon=0.006-0.007$, which corresponds to
+$R_e = 1.2$, infectious time $\frac{1}{\psi} = 2.3$ years, and 6--16 days between partner notification and their sampling.
 The results are consistent between 10 forests and shown in Table~\ref{tbl:uk}.
@@
 \toprule
- $1$ & $11\,117$ & $8\,838$& $1.24\;[1.22-1.27]$& $2.34\;[2.29-2.39]$& $8\;[5-12]$& $0.007\;[0.004-0.010]$ \\
- $2$ & $11\,109$ & $8\,833$& $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.39]$& $6\;[4-10]$& $0.006\;[0.003-0.009]$ \\
- $3$ & $11\,122$ & $8\,844$& $1.24\;[1.22-1.27]$& $2.34\;[2.28-2.39]$& $16\;[3-18]$& $0.006\;[0.003-0.010]$ \\
- $4$ & $11\,119$ & $8\,838$& $1.24\;[1.22-1.27]$& $2.34\;[2.28-2.39]$& $7\;[5-11]$& $0.007\;[0.004-0.010]$ \\
- $5$ & $11\,116$ & $8\,829$& $1.24\;[1.22-1.27]$& $2.33\;[2.28-2.39]$& $9\;[6-13]$& $0.006\;[0.003-0.009]$ \\
- $6$ & $11\,111$ & $8\,827$& $1.24\;[1.22-1.27]$& $2.34\;[2.28-2.39]$& $8\;[5-12]$& $0.006\;[0.003-0.009]$ \\
- $7$ & $11\,119$ & $8\,838$& $1.24\;[1.22-1.27]$& $2.33\;[2.28-2.39]$& $7\;[5-11]$& $0.006\;[0.003-0.009]$ \\
- $8$ & $11111$ & $8\,830$& $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.39]$& $6\;[4-11]$& $0.006\;[0.003-0.008]$ \\
- $9$ & $11\,120$ & $8\,833$& $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.39]$& $7\;[4-12]$& $0.006\;[0.003-0.009]$ \\
- $10$ & $11\,112$ & $8\,827$& $1.25\;[1.22-1.27]$& $2.34\;[2.29-2.39]$& $7\;[5-11]$& $0.006\;[0.004-0.009]$ \\
+ $1$ & $11\,117$ & $8\,838$& $1.24\;[1.21-1.27]$& $2.34\;[2.28-2.40]$& $8\;[5-12]$& $0.007\;[0.003-0.010]$ \\
+ $2$ & $11\,109$ & $8\,833$& $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.39]$& $6\;[3-10]$& $0.006\;[0.003-0.009]$ \\
+ $3$ & $11\,122$ & $8\,844$& $1.24\;[1.21-1.30]$& $2.34\;[2.27-2.40]$& $16\;[2-18]$& $0.006\;[0.002-0.015]$ \\
+ $4$ & $11\,119$ & $8\,838$& $1.24\;[1.21-1.27]$& $2.34\;[2.28-2.40]$& $7\;[5-11]$& $0.007\;[0.004-0.011]$ \\
+ $5$ & $11\,116$ & $8\,829$&  $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.39]$& $9\;[5-13]$& $0.006\;[0.003-0.010]$ \\
+ $6$ & $11\,111$ & $8\,827$&  $1.24\;[1.22-1.27]$& $2.34\;[2.28-2.40]$& $8\;[5-12]$& $0.006\;[0.003-0.010]$ \\
+ $7$ & $11\,119$ & $8\,838$& $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.39]$& $7\;[4-11]$& $0.006\;[0.003-0.010]$ \\
+ $8$ & $11111$ & $8\,830$& $1.24\;[1.21-1.27]$& $2.33\;[2.27-2.39]$& $6\;[3-11]$& $0.006\;[0.003-0.008]$ \\
+ $9$ & $11\,120$ & $8\,833$&  $1.24\;[1.21-1.27]$& $2.33\;[2.27-2.39]$& $7\;[4-12]$& $0.006\;[0.003-0.010]$ \\
+ $10$ & $11\,112$ & $8\,827$& $1.25\;[1.22-1.27]$& $2.34\;[2.28-2.40]$& $7\;[4-11]$& $0.006\;[0.003-0.010]$ \\
  \bottomrule
@@
 
-The -PN extension adds two constant parameters to the classical BD model: the probability of partner notification $p_n$ and the partner detection rate $\phi$.
-It can also be applied to a more general, multi-type birth-death (MTBD) model framework, which accounts for heterogeneity in the infected population (see Appendix).
+The -PN extension adds two constant parameters to the classical BD model: the probability of partner notification $\upsilon$ and the partner detection rate $\phi$.
+It can also be applied to a more general, multi-type birth-death (MTBD) model framework, which accounts for heterogeneity in the infected population.
 
-We developed a mathematical framework for calculation of the likelihood density of a transmission tree (approximated by a time-scaled phylogenetic tree) under the BD-PN model, and implemented a maximum-likelihood BD-PN parameter estimator based on this framework.
+We developed a PN-test allowing to distinguish between trees generated under models with and without partner notification. Its application to simulated data showed high sensitivity and specificity.
+
+We developed a mathematical framework for calculation of the likelihood density of a transmission tree (approximated by a time-scaled phylogenetic tree) under the BD-PN model, and implemented a maximum-likelihood BD-PN parameter estimator bdpn based on this framework.
 The advantage of the BD-PN model is that the differential equations describing it have closed-form solutions, which are quick to calculate and do not require numerical approximation (as is the case for more complex MTBD models).
-Tests on simulated data show that our estimator robustly infers model parameter values and their CIs in the setting when the duration of the infectious period $\frac{1}{\psi}$ or the base sampling probability $\rho$ if fixed.
-Fixing one of the model parameters is needed for identifiability.
+Tests on simulated data show that our estimator robustly infers model parameter values and their CIs in the setting when the duration of the infectious period $\frac{1}{\psi}$ or the base sampling probability $\rho$ is fixed.
+Fixing one of the model parameters is needed for model identifiability.
 In practice, $\rho$ and $\frac{1}{\psi}$ may often be approximated from epidemiological data, e.g., as the proportion of sampled cases among the declared ones (for $\rho$), or estimated from observations of infected cases ($\frac{1}{\psi}$).
 
-We also developed a PN-test allowing to distinguish between trees generated under models with and without partner notification.
-Its application to simulated data showed high sensitivity and specificity.
+Importantly, our simulations show that in the presence of partner notification, the classical BD model estimator fails to accurately estimate epidemiological parameters (average relative errors $> 30\%$ for $R_0$ and infectious time $\frac{1}{\psi}$).
+On the contrary, in the absence of partner notification, the BD-PN model estimator correctly identifies $R_0$ and infectious time $\frac{1}{\psi}$ (as well as the BD model estimator), and estimates a very low partner notification probability $\upsilon$, whose confidence intervals hit the lower bound.
 
-We applied the PN-test our estimator to the HIV-1 B epidemic in the UK, between 2012 and 2015.
-The test detected the presence of partner notification.
+We applied the PN-test and our estimator to the HIV-1 B epidemic in the UK, between 2012 and 2015.
+The test detected the strong presence of partner notification.
 The obtained estimates of epidemiological parameters are in agreement with what we now know about this epidemic (e.g., $R_e \approx 1.2$ ).
-We estimated a modest partner notification probability $\upsilon \in [0.003-0.1]$, and rapid notified partner detection ($3-18$ days vs $\approx2.3$ years for detection under the standard random procedure).
+We estimated a modest partner notification probability $\upsilon \approx 0.006$, and rapid notified partner detection ($\approx 8$ days vs $\approx2.3$ years for detection under the standard random procedure).
 
 The BD-PN model is the first step towards accounting for non-random sampling in phylodynamic models.
-In order to reduce its mathematical complexity (having closed-form solutions for its differential equations) and make parameter estimation fast, we made several assumptions: (i) that only detected individuals can notify their partner; (ii) only the most recent partner can get notified; (iii) after the notification, notified partners do not transmit further; and (iv) notified partners are always sampled upon removal (i.e., observed in the transmission tree).
+In order to reduce its mathematical complexity (having closed-form solutions for its differential equations) and make parameter estimation fast, we made several assumptions: (i) that only detected individuals can notify their partners; (ii) only the most recent partner can get notified; (iii) after the notification, notified partners do not transmit further; and (iv) notified partners are always sampled upon removal (i.e., observed in the transmission tree).
 These assumptions may be too simple to fully grasp the real partner notification process and should be relaxed in the future work.

### 2024-05-23 article/article.tex
wrote down all the options

@@
 \begin{equation}
-U_p^{(1)}(t, t_A) = p_o^{(x)}(t, t_A) p_p^{(x)}(t)\textit{\color{gray}, where $t_x =T$}
+U_p^{(1)}(t, t_A) = p_o^{(x)}(t, t_A) p_p^{(x)}(t_A)\textit{\color{gray}, where $t_x =T$}
 \end{equation}
@@
 
-We will also need to describe the probability of a\textbf{ mixed branch} $p_{m(s1,s2)}^{(i)}(t,t_h,t_r)$ for branches that include a hidden partner subtree (AB-C branches in Fig.~\ref{fig:pn-branches}(2) and A's tip branch in Fig.~\ref{fig:pn-branches}(3)).
+We will also need to describe the probability of a\textbf{ mixed branch} $p_{m(s1,s2)}^{(i)}(t,t_r)$ for branches that include a hidden partner subtree (AB-C branches in Fig.~\ref{fig:pn-branches}(2) and A's tip branch in Fig.~\ref{fig:pn-branches}(3)).
 $s1$ here denotes the type of the top of the branch (e.g.
@@
 between AB and A in Fig.~\ref{fig:pn-branches}(3), index).
-$t_h$ denotes the time of the hidden partner subtree start, while $t_r$ denotes the first notification time of the hidden partner.
+$t_r$ denotes the first notification time of the hidden partner.
+Moreover, let's denote the time of the hidden partner subtree start $t_h$.
 For instance, in Fig~\ref{fig:pn-branches}(2a) $t_i=t_C$, $t_r=t_A$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(2b) $t_i=t_r=t_C$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(3c) $t_i=t_r=t_A$, $t_h=t_{AB}$.
 
-$p_{m(s1,s2)}^{(i)}(t,t_h,t_r)$ combines three elements: the probability of the top part of the branch finishing with a transmission, the probability of the bottom part of the branch (without taking into account the event at its end) and the probability of the hidden partner subtree:
+$p_{m(s1,s2)}^{(i)}(t,t_r)$ combines three elements: the probability of the top part of the branch, which finishes with a transmission to or from the hidden partner, the probability of the bottom part of the branch (without taking into account the event at its end) and the probability of the hidden partner subtree:
 
@@
 \begin{split}
-p_{m(s1,s2)}^{(i)}(t,t_h,t_r) &= p_{top=s1}^{(h)}(t)p_{bottom=s2}^{(i)}(t_h)U_p(t_h,t_r)\textit{\color{gray},}\\
+p_{m(s1,s2)}^{(i)}(t,t_r) &= p_{top=s1}^{(h)}(t)p_{bottom=s2}^{(i)}(t_h)U_p(t_h,t_r)\textit{\color{gray},}\\
 &\textit{\color{gray}where $t_h$ is the time of the hidden partner subtree start,}\\
@@
 
-Note that the bottom part of the branch can either correspond to a notifier (as in Fig~\ref{fig:pn-branches}(2b,3)), or to a standard branch, or to a mixed tip branch starting as a standard branch and finishing as a notifier.
+Note that since the branch contains a hidden partner, this partner must have at least one notifier who is observed.
+If the branch is external, the notifier could correspond to its tip (e.g., tip A in Fig.~\ref{fig:pn-branches}(3) and tip C in Fig.~\ref{fig:pn-branches}(2b)) or to a tip in the tip's supertree (the tree without the tip and its branch): For example, such notifiers correspond to tip A in Fig.~\ref{fig:pn-branches}(2b) and tip C in Fig.~\ref{fig:pn-branches}(3b)).
+If the branch is external, the notifier could only correspond to a tip in its node's supertree: For example, tip A who is part of the C's supertree in Fig.~\ref{fig:pn-branches}(2a,c)).
+Having a notifier as the branch's tip implies that the bottom part of the branch corresponds to the notifier.
+Having a notifier in the supertree implies that the top part of the branch corresponds to the partner.
+Hence if the branch is internal, the top part can only correspond to the partner, while the bottom part can only correspond to a standard state (unnotified non-notifier), as the bottom part of such a branch corresponds to the partner's recipient.
+Finally, if the branch is external, and the only partner's notifier corresponds to its tip, then there are two possibilities for its top part: (1a) either it fully corresponds to a standard state (unnotified non-notifier); or (1b) there is another partners' notifier in its supertree, and this other partner is also hidden along the top part of the same branch, hence the top part of the branch is partner-standard (see Fig.~\ref{fig:pn-mixed}).
+If the branch is external, and the partner's notifier(s) is(are) only found in its supertree, then there are two possibilities for its bottom part: (2a) either it fully corresponds to a standard state (unnotified non-notifier); or (2b) there is another partners' notifier at its tip, and this other partner is also hidden along the bottom part of the same branch, hence the bottom part of the branch is standard-notifier (see Fig.~\ref{fig:pn-mixed}).
+Note that (1b) and (2b) represent the same case but from the point of view of two different hidden partners (B and its notifier C for former, and D and its notifier C for the latter in Fig.~\ref{fig:pn-mixed}).
+
+\begin{figure*}[tbhp]
+\centering 
+%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
+\includegraphics[width=0.3\textwidth]{Fig_mixed}
+\caption{An example of mixed tip branch with two hidden partner subtrees: an individual C got sampled at time $t_C$ and notified their most recent partner D.
+D however, stayed unobserved as D did not get sampled by the end of the sampling period (time $T$).
+Before notification (at time $t_C$) D transmitted two times, the first transmission leading to an unobserved recipient subtree (gray dotted triangle) and the second (DB) leading to an observed recipient subtree (which includes a sampled individual A and a hidden individual B).
+The D's recipient (or someone in their subtree if there were hidden transmissions along the path DB-AB) transmitted further (transmission AB).
+Once A got sampled at time $t_A$, they notified their most recent partner B, who however was already removed by then via the standard procedure without sampling (at time $t_B < t_A$).
+Hence the branch CD-A represents a mixed branch with states partner(CD-DB)-standard(DB-AB)-notifier(AB-A).
+}
+\label{fig:pn-mixed} 
+\end{figure*}
+
+%if the branch is external the bottom part of the branch can either correspond to a notifier (as in Fig~\ref{fig:pn-branches}(2b,3)), or to a standard branch, or to a mixed tip branch starting as a standard branch and finishing as a notifier.
 It cannot correspond to another partner branch as every partner has their notifier sampled, and notifiers only notify one partner (who is already hidden along this branch).
@@
 
+We will first describe the one-hidden-partner cases ($p_{m(s1,s2)}^{(i)}(t,t_r)$, where $s1 \in \{o, -\}$, $s2 \in \{n, -\}$ and $-$ stands for the standard (unnotified non-notifier) state), and then proceed with the two-partner case ($p_{m(o,-,i)}^{(i)}(t,t_{r1},t_{r2})$, where $t_{r1}$ is the first notification time for the hidden partner at the top of the branch (e.g.
+$t_C$ in Fig.~\ref{fig:pn-mixed}), $t_{r2}$ is the first notification time for the hidden partner at the top of the branch  (e.g.
+$t_A$ in Fig.~\ref{fig:pn-mixed}).
+
+For the bottom branch part:
+
 \begin{equation}
@@
 \begin{cases}
-p_n^{(i)}(t)&\textit{\color{gray} if s2 is a notifier,}\\
-p_{m(-,n)}^{(i)}(t)&\textit{\color{gray} if i is a tip, and s2 type is a standard-notifier mix,}\\
-p^{(i)}(t)&\textit{\color{gray} otherwise (standard state)}
+p_n^{(i)}(t)&\textit{\color{gray} if $i$ is a notifier tip ($s2=n$),}\\
+p^{(i)}(t)&\textit{\color{gray} otherwise ($s2=-$, the standard state)}\\
 \end{cases}
@@
 
-Finally, the top part of the branch can be in three possible states.
+%Finally, the top part of the branch can be in three possible states.
 The first possibility is that it corresponds to the hidden partner's individual.
@@
 The other possibility is that the only partner's notifier corresponds to the tip at the end of the current branch (as in Fig~\ref{fig:pn-branches}(3a,c)).
-Then the branch can either be in a standard state (unnotifier non-notifier), or a mix between a (another, partially hidden) partner branch and a standard state (corresponding to that other partner's recipient).
-In the first case we will approximate $p_{top=o}^{(h)}(t)$ using $p_o^{(h)}(t)$ in the following way.
+Then the branch can either be in a standard state (unnotifier non-notifier), or a mix between another, partially hidden, partner branch and a standard state (corresponding to that other partner's recipient).
+We will approximate $p_{top=s1}^{(h)}(t)$ using the branch evolution probability of the type corresponding to $s1$  in the following way.
+For example, let us assume that $s1=o$ (observed (partner) branch).
 $p_o^{(h)}(t)$ expresses a probability of a partner evolving along the branch between the times $t$ and $t_h$ with no or any number of oriented hidden transmissions, without taking into account the event at the end of it.
-In our case there must be at least one hidden transmission: from the partner at the end of the branch (where the donor tree stayed unsampled).
+In our case there must be at least one hidden transmission: from the partner at time $t_h$ (where the donor tree stayed unsampled).
 If we remove the probability of no event during this time ($e^{-(\lambda + \psi)(t_h - t)}$) from $p_o^{(h)}(t)$, we will obtain a probability of a partner branch with at least one oriented hidden transmission somewhere between $t$ and $t_h$ (the probability of the corresponding hidden tree being included).
-$p_{top=s1}^{(h)}(t)$ expresses almost the same thing, with the difference being that the last hidden transmission must have happened at time $t_h$ and the probability of the hidden partner tree corresponding to it is not included.
-We will hence approximate it as follows:
+$p_{top=o}^{(h)}(t)$ expresses almost the same thing, with the difference being that the last hidden transmission must have happened at time $t_h$ and the probability of the hidden partner tree corresponding to it is not included.
+We will hence approximate $p_{top=o}^{(h)}(t)$ as follows:
 \begin{equation}
@@
 
-We will process the mixed case in the same way as the previous two:
-\begin{equation}
-p_{top=m(o,-)}^{(h)}(t) \approx \frac{p_{m(o,-)}^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}}{U(t_{h})}
-\end{equation}
-
-Note that we wrote down equation~(\label{eq:mixed}) as if the start of the hidden partner tree $t_h$ was known.
+Note that we wrote down equation~(\ref{eq:mixed}) as if the start of the hidden partner tree $t_h$ was known.
 In practice, it is not the case as we need to integrate over all the possible times between the branch start $t_{parent(i)}$ and either its end or the first partner's notification (if it was before): $min(t_r,t_i)$.
@@
 \begin{split}
-p_{m(s1,s2)}^{(i)}(t,t_r) &\approx p_{s1}^{(h)}(t)\Big(p_{s2}^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}\Big)\frac{U_p(t_h,t_r)}{U(t_{h})}\textit{\color{gray},}\\
+p_{m(s1,s2)}^{(i)}(t,t_r) &\approx \Big(p_{s1}^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}\Big)\frac{U_p(t_h,t_r)}{U(t_{h})}p_{s2}^{(i)}(t_h)\textit{\color{gray},}\\
 &\textit{\color{gray}where $t_h \approx t + \frac{min(t_r, t_i) - t}{2}$,}\\
@@
 
+Finally, we will treat the mixed case with two hidden partners using similar principals:
+\begin{equation}
+\begin{split}
+p_{m(o,-,n)}^{(i)}(t,t_{r1},t_{r2}) &\approx \Big(p_{o}^{(h1)}(t) -e^{-(\lambda + \psi)(t_{h1} - t)}\Big)\frac{U_p(t_{h1},t_{r1})}{U(t_{h1})}\cdot
+\\
+&~~~~\Big(p^{(h2)}(t_{h1}) -e^{-(\lambda + \psi)(t_{h2} - t_{h1})}\Big)\frac{U_p(t_{h2},t_{r2})}{U(t_{h2})}\cdot
+\\
+&~~~~~p_n^{(i)}(t_{h2})\textit{\color{gray},}\\
+&\textit{\color{gray}where $t_{h1} \approx t + \frac{min(t_{r1}, t_i, t_{r2}) - t}{3}$,}\\
+&\textit{\color{gray}~~~~~~~~$t_{h2} \approx t + \frac{2(min(t_{r2}, t_i) - t)}{3}$,}\\
+&\textit{\color{gray}~~~~~~~~$t_{r1}$ is the earliest notification time of the }\\
+&\textit{\color{gray}~~~~~~~~~~~~hidden partner at the top of the branch (o,-),}\\
+&\textit{\color{gray}~~~~~~~~$t_{r2}$ is the earliest notification time of the }\\
+&\textit{\color{gray}~~~~~~~~~~~~hidden partner at the bottom of the branch (-,n).}\\
+\end{split}\label{eq:p_mixed_mixed}
+\end{equation}
+
 \subsection{BD-PN tree likelihood density}
 
-Using the formulas~(\ref{eq:ps}) and (\ref{eq:p_mixed}), we can calculate the likelihood density of a tree using a pruning algorithm~\citep{10.1093/sysbio/22.3.240}.
+Using the formulas~(\ref{eq:ps}), (\ref{eq:p_mixed}) and (\ref{eq:p_mixed_mixed}), we can calculate the likelihood density of a tree using a pruning algorithm~\citep{10.1093/sysbio/22.3.240}.
 For each visited node $i$ we calculate the likelihood density of its subtree including the branch $ji$ connecting $i$ to its parent node $j$ for two cases depending on $i$'s receiving-a-notification status (shown with the first superscript, $^{\not{n}}$ or $^n$).
@@
 0 \textit{\color{gray} ~if $t_r < t_j$}\\
-\textit{\color{gray} (not possible as once notified partners do not transmit)}\\
+\textit{\color{gray} (not possible as once notified partners do not transmit)}
+\\
 \\
 p_o^{(i)}(t_j)\psi\rho(1 - \upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and did not notify}\\
-\textit{\color{gray} (oriented (partner) branch evolution followed by standard sampling and no notification)}\\
-~~~+p_{m(o,n)}^{(i)}(t_j,t_i)\psi\rho\upsilon \textit{\color{gray} ~i.e. $i$ got notified at $t_r > t_i$, and notified a hidden partner}\\
- \textit{\color{gray} ~~~(mixed oriented-notifier branch evolution followed by sampling and notification)}\\
+\textit{\color{gray} (oriented (partner) branch evolution followed by standard sampling and no notification)}
+\\
+~~~+p_{m(o,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r > t_i$, }\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$  did not notify}\\
+ \textit{\color{gray} ~~~(mixed oriented-standard branch evolution followed by sampling and no notification)}\\
+ ~~~+p_{m(o,n)}^{(i)}(t_j,t_i)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
+ \textit{\color{gray} ~~~(mixed oriented-notifier branch evolution followed by sampling and notification)}
+\\
+~~~+p_{m(o,-,n)}^{(i)}(t_j,t_r,t_i)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r > t_i$, }\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
+ \textit{\color{gray} ~~~(mixed oriented-standard-notifier branch evolution followed by sampling and notification)}
+\\
 \\
@@
 \textit{\color{gray} (oriented (partner) branch evolution followed by} \\
-\textit{\color{gray}~notified partner branch evolution, notification-provoked sampling and no notification)}\\
-~~~+p_{m(o,n)}^{(r)}(t_j,t_i)p_p^{(i)}(t_r)\phi\upsilon \textit{\color{gray} ~i.e. $i$ got notified at $t_j \leq t_r \leq t_i$, and notified a hidden partner}\\
- \textit{\color{gray} ~~~(mixed oriented-notifier branch evolution followed by} \\
-\textit{\color{gray}~~~~notified partner branch evolution, notification-provoked sampling and notification)}\\
+\textit{\color{gray}~notified partner branch evolution, notification-provoked sampling and no notification)}
+\\
+~~~+p_{m(o,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_j \leq t_r \leq t_i$, }\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$  did not notify}\\
+ \textit{\color{gray} ~~~(mixed oriented-standard branch evolution followed by sampling and no notification)}\\
+ ~~~+p_{m(o,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
+ \textit{\color{gray} ~~~(mixed oriented-notifier branch evolution followed by sampling and notification)}
+\\
+~~~+p_{m(o,-,n)}^{(i)}(t_j,t_r,t_i)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_j \leq t_r \leq t_i$, }\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
+ \textit{\color{gray} ~~~(mixed oriented-standard-notifier branch evolution followed by sampling and notification)}
 \end{cases}\\

### 2024-05-23 simulations/py/plot_error_both.py

@@ -1,0 +1,176 @@
+from collections import defaultdict
+
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sns
+from matplotlib.offsetbox import TextArea, HPacker, AnchoredOffsetbox, VPacker
+from statsmodels.stats.weightstats import CompareMeans
+
+RATE_PARAMETERS = ['lambda', 'psi', 'psi_p', 'p', 'pn']
+EPIDEMIOLOGIC_PARAMETERS = ['R_naught', 'infectious_time', 'partner_removal_time']
+par2greek = {'lambda': u'\u03bb', 'psi': u'\u03c8', 'psi_p': u'\u03c8p', 'p': '\u03c1', 'pn': '\u03c5',
+             'R_naught': u'\u0052\u2080' + '=' + u'\u03bb\u002F\u03c8',
+             'infectious_time': 'infectious time 1' + u'\u002F\u03c8',
+             'partner_removal_time': 'partner removal time 1' + u'\u002F\u03c6'}
+PARAMETERS = RATE_PARAMETERS + EPIDEMIOLOGIC_PARAMETERS
+
+PS_BDPN = EPIDEMIOLOGIC_PARAMETERS + ['p', 'pn']
+PS_BD = EPIDEMIOLOGIC_PARAMETERS[:-1] + ['p']
+
+if __name__ == "__main__":
+    import argparse
+
+    parser = argparse.ArgumentParser(description="Plots errors.")
+    parser.add_argument('--estimates_bdpn', type=str, help="estimated parameters")
+    parser.add_argument('--estimates_bd', type=str, help="estimated parameters")
+    parser.add_argument('--pdf', type=str, help="plot")
+    parser.add_argument('--fixed', type=str, default='p', help="fixed parameter", choices=RATE_PARAMETERS)
+    # parser.add_argument('--pn_min', type=float, default=0., help="Only display cases with pn greater or equal to this value")
+    # parser.add_argument('--pn_max', type=float, default=1, help="Only display cases with pn smaller than this value")
+    params = parser.parse_args()
+
+    plt.clf()
+
+    fig, (ax1, ax2) = plt.subplots(2, 1, figsize=(15, 12))
+    rc = {'font.size': 12, 'axes.labelsize': 10, 'legend.fontsize': 10, 'axes.titlesize': 10, 'xtick.labelsize': 10,
+          'ytick.labelsize': 10}
[…]

### 2024-05-24 article/article.tex
Implementation updated with hidden partner cases

@@
 
-Note that since the branch contains a hidden partner, this partner must have at least one notifier who is observed.
-If the branch is external, the notifier could correspond to its tip (e.g., tip A in Fig.~\ref{fig:pn-branches}(3) and tip C in Fig.~\ref{fig:pn-branches}(2b)) or to a tip in the tip's supertree (the tree without the tip and its branch): For example, such notifiers correspond to tip A in Fig.~\ref{fig:pn-branches}(2b) and tip C in Fig.~\ref{fig:pn-branches}(3b)).
+Note that since the branch contains a hidden partner, this partner must have at least one notifier (who is observed as all the notifiers).
+If the branch is external, the notifier could correspond to its tip: For example, such notifiers correspond to tip A in Fig.~\ref{fig:pn-branches}(2b) and tip C in Fig.~\ref{fig:pn-branches}(3b)).
 If the branch is external, the notifier could only correspond to a tip in its node's supertree: For example, tip A who is part of the C's supertree in Fig.~\ref{fig:pn-branches}(2a,c)).
@@
 Having a notifier in the supertree implies that the top part of the branch corresponds to the partner.
-Hence if the branch is internal, the top part can only correspond to the partner, while the bottom part can only correspond to a standard state (unnotified non-notifier), as the bottom part of such a branch corresponds to the partner's recipient.
+Hence if the branch is internal, the top part can only correspond to the partner, while the bottom part can only correspond to a standard state (unnotified non-notifier), as the bottom part of such a branch corresponds to the partner's recipient (who could not be notified as the partner is unobserved and could not be a notifier as the branch is not external).
 Finally, if the branch is external, and the only partner's notifier corresponds to its tip, then there are two possibilities for its top part: (1a) either it fully corresponds to a standard state (unnotified non-notifier); or (1b) there is another partners' notifier in its supertree, and this other partner is also hidden along the top part of the same branch, hence the top part of the branch is partner-standard (see Fig.~\ref{fig:pn-mixed}).
-If the branch is external, and the partner's notifier(s) is(are) only found in its supertree, then there are two possibilities for its bottom part: (2a) either it fully corresponds to a standard state (unnotified non-notifier); or (2b) there is another partners' notifier at its tip, and this other partner is also hidden along the bottom part of the same branch, hence the bottom part of the branch is standard-notifier (see Fig.~\ref{fig:pn-mixed}).
-Note that (1b) and (2b) represent the same case but from the point of view of two different hidden partners (B and its notifier C for former, and D and its notifier C for the latter in Fig.~\ref{fig:pn-mixed}).
+If, on the other hand, the branch is external, and the partner's notifier(s) is(are) only found in its supertree, then there are two possibilities for its bottom part: (2a) either it fully corresponds to a standard state (unnotified non-notifier); or (2b) there is another partners' notifier at its tip, and this other partner is also hidden along the bottom part of the same branch, hence the bottom part of the branch is standard-notifier (see Fig.~\ref{fig:pn-mixed}).
+Note that (1b) and (2b) represent the same case but from the point of view of two different hidden partners (B and their notifier C for the former, and D and their notifier C for the latter in Fig.~\ref{fig:pn-mixed}).
 
@@
 
-We will first describe the one-hidden-partner cases ($p_{m(s1,s2)}^{(i)}(t,t_r)$, where $s1 \in \{o, -\}$, $s2 \in \{n, -\}$ and $-$ stands for the standard (unnotified non-notifier) state), and then proceed with the two-partner case ($p_{m(o,-,i)}^{(i)}(t,t_{r1},t_{r2})$, where $t_{r1}$ is the first notification time for the hidden partner at the top of the branch (e.g.
-$t_C$ in Fig.~\ref{fig:pn-mixed}), $t_{r2}$ is the first notification time for the hidden partner at the top of the branch  (e.g.
-$t_A$ in Fig.~\ref{fig:pn-mixed}).
+We will first describe the one-hidden-partner cases ($p_{m(s1,s2)}^{(i)}(t,t_r)$, where $s1 \in \{o, -\}$, $s2 \in \{n, -\}$ and $-$ stands for the standard (unnotified non-notifier) state), and then proceed with the two-partner case ($p_{m(o,-,n)}^{(i)}(t,t_{r})$, where $t_{r}$ is the earliest notification time for the hidden partner at the top of the branch (e.g.
+$t_r=t_C$ in Fig.~\ref{fig:pn-mixed}), the only notification time for the hidden partner at the bottom of the branch being $t_i$  (e.g.
+$t_i=t_A$ in Fig.~\ref{fig:pn-mixed}).
 
@@
 \begin{split}
-p_{m(o,-,n)}^{(i)}(t,t_{r1},t_{r2}) &\approx \Big(p_{o}^{(h1)}(t) -e^{-(\lambda + \psi)(t_{h1} - t)}\Big)\frac{U_p(t_{h1},t_{r1})}{U(t_{h1})}\cdot
-\\
-&~~~~\Big(p^{(h2)}(t_{h1}) -e^{-(\lambda + \psi)(t_{h2} - t_{h1})}\Big)\frac{U_p(t_{h2},t_{r2})}{U(t_{h2})}\cdot
+p_{m(o,-,n)}^{(i)}(t,t_r) &\approx \Big(p_{o}^{(h1)}(t) -e^{-(\lambda + \psi)(t_{h1} - t)}\Big)\frac{U_p(t_{h1},t_{r})}{U(t_{h1})}\cdot
+\\
+&~~~~\Big(p^{(h2)}(t_{h1}) -e^{-(\lambda + \psi)(t_{h2} - t_{h1})}\Big)\frac{U_p(t_{h2},t_{i})}{U(t_{h2})}\cdot
 \\
 &~~~~~p_n^{(i)}(t_{h2})\textit{\color{gray},}\\
-&\textit{\color{gray}where $t_{h1} \approx t + \frac{min(t_{r1}, t_i, t_{r2}) - t}{3}$,}\\
-&\textit{\color{gray}~~~~~~~~$t_{h2} \approx t + \frac{2(min(t_{r2}, t_i) - t)}{3}$,}\\
-&\textit{\color{gray}~~~~~~~~$t_{r1}$ is the earliest notification time of the }\\
-&\textit{\color{gray}~~~~~~~~~~~~hidden partner at the top of the branch (o,-),}\\
-&\textit{\color{gray}~~~~~~~~$t_{r2}$ is the earliest notification time of the }\\
-&\textit{\color{gray}~~~~~~~~~~~~hidden partner at the bottom of the branch (-,n).}\\
+&\textit{\color{gray}where $t_{h1} \approx t + \frac{1}{3}\Big(min(t_{r}, t_i) - t\Big)$,}\\
+&\textit{\color{gray}~~~~~~~~$t_{h2} \approx t + \frac{2}{3}\Big(min(t_{r}, t_i) - t\Big)$,}\\
+&\textit{\color{gray}~~~~~~~~$t_{r}$ is the earliest notification time of the }\\
+&\textit{\color{gray}~~~~~~~~~~~~hidden partner at the top of the branch (o,-).}\\
 \end{split}\label{eq:p_mixed_mixed}
@@
 \\
-~~~+p_{m(o,-,n)}^{(i)}(t_j,t_r,t_i)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r > t_i$, }\\
+~~~+p_{m(o,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r > t_i$, }\\
 \textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
@@
 \\
-~~~+p_{m(o,-,n)}^{(i)}(t_j,t_r,t_i)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_j \leq t_r \leq t_i$, }\\
+~~~+p_{m(o,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_j \leq t_r \leq t_i$, }\\
 \textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
@@
 \begin{split}
-l^{(\not{n})}(i) = p^{i}(t_j) 2\lambda \cdot 
+l^{(\not{n})}(i) = p^{(i)}(t_j) 2\lambda \cdot 
 \Big(&~l^{(\not{n},\not{n})}(i_0)l^{(\not{n},\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ neigher $i_0$ nor $i_1$ notified} \\
@@
 \begin{split}
-l^{(\not{n})}(i) = p^{i}(t_j) 2\lambda \cdot 
+l^{(\not{n})}(i) = p^{(i)}(t_j) 2\lambda \cdot 
 \Big(&~l^{(\not{n},\not{n})}(i_0)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ did not notify $i_1$ } \\
@@
 \toprule 
- $\lambda$ &  87\% & 14.2\% & 96\% & 11.6\%\\
- $\psi$ & -- & -- & 80\% & 28.6\% \\
- $\rho$ & 66\%  & 39.1\% & -- & -- \\
- $\phi$ & 93\% & 32.5\% & 94\% & 32.3\% \\
- $\upsilon$ & 69\% & 21\% & 65\% & 20.7 \% \\
+ $\lambda$ &  91\% & 14.2\% & 96\% & 11.6\%\\
+ $\psi$ & -- & -- & 88\% & 28.9\% \\
+ $\rho$ & 78\%  & 38.9\% & -- & -- \\
+ $\phi$ & 93\% & 32.3\% & 95\% & 32.2\% \\
+ $\upsilon$ & 74\% & 21.8\% & 70\% & 21.7\% \\
 \bottomrule

### 2024-05-24 bdpn/bdpn_model.py

@@ -6,7 +6,7 @@ import pandas as pd
 
 from bdpn import bd_model
 from bdpn.formulas import get_log_p, get_c1, get_c2, get_E2, get_E1, get_log_po, get_log_pn, get_log_po_from_p_pn, \
-    get_u, get_log_pp
+    get_u, get_log_pp, get_log_no_event
 from bdpn.parameter_estimator import optimize_likelihood_params, rescale_log
 from bdpn.tree_manager import TIME, read_forest, annotate_forest_with_time
 
@@ -31,6 +31,174 @@ def log_sum(log_summands):
     return np.log(np.sum(np.exp(result))) - factors
 
 
+def log_subtraction(log_minuend, log_subtrahend):
+    """
+    Takes logX1 and logX2 as input and returns log(X1 - X2) as output,
+    while taking care of potential under/overflow.
+
+    :param log_minuend: logX1 in the formula above
+    :param log_subtrahend: logX2 in the formula above
+    :return: log of the difference
+    """
+    result = np.array([log_minuend, log_subtrahend], dtype=np.float64)
+    factors = rescale_log(result)
+    return np.log(np.sum(np.exp(result) * [1, -1])) - factors
+
+def log_u_p(t, tr, T, la, psi, rho, phi, c1, E1, E2_tr):
+    """
+    Calculates a log probability of a partner subtree evolving between the time t and T unobserved,
+    provided the first partner notification happened at time tr.
+    In practice, two possibilities are summed up:
+    (1) the hidden partner branch has not yet got sampled between the notification and T;
+    (2) the hidden partner branch got removed before the notification (without sampling).
+
+
+    :param t: start time of the hidden partner subtree evolution
+    :param tr: moment of the earliest partner notification
+    :param T: end of sampling time
+    :param E1: E1 = c2 * exp(c1 * (t - T)), where c2 = (c1 + la - psi) / (c1 - la + psi)
+    :param E2_tr: E2_tr = c2 * exp(c1 * (tr - T))
+    :return: log of the hidden partner subtree probability (approximation)
+    """
+    branch_len = tr - t
+
+    # Oriented (partner) branch evolution from t till the moment of notification (tr)
+    log_po = get_log_po(la=la, psi=psi, c1=c1, t=t, ti=tr,
+                        E1=E1, E2=E2_tr)
[…]

### 2024-05-24 bdpn/formulas.py

@@ -97,16 +97,25 @@ def get_log_po_from_p_pn(log_p, log_pn):
 def get_log_pn(la, psi, t, ti):
     """
     Returns log(pn(t)), where
-    dpn(t)/dt = la + psi * pn(t)
+    dpn(t)/dt = (la + psi) * pn(t)
     pn(ti) = 1
     """
-    return (la + psi) * (t - ti)
+    return get_log_no_event(la + psi, t, ti)
 
 
-def get_log_pp(la, phi, t, ti):
+def get_log_no_event(rate, t, ti):
+    """
+    Returns log(pne(t)), where
+    dpne(t)/dt = rate * pne(t)
+    pne(ti) = 1
+    """
+    return -rate * (ti - t)
+
+
+def get_log_pp(phi, t, ti):
     """
     Returns log(pp(t)), where
     dpp(t)/dt = phi * pp(t)
     pp(ti) = 1
     """
-    return phi * (t - ti)
+    return get_log_no_event(phi, t, ti)

### 2024-05-30 article/article.tex
Rewriting article with transmission from notified partners,
simulations updated accordingly

@@
  \item $\upsilon$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
- \item $\phi >> \psi$ -- constant notified sampling rate: a notified individual (if not yet removed via standard procedure) gets observed with this rate.
+ \item $\phi >> \psi$ -- constant notified sampling rate: a notified individual (if not yet removed via standard procedure) gets removed \textit{and observed} with this rate.
 This creates an observed tip in the transmission tree.
@@
 
-This model is a simplification of the notification process, in particular it makes 4 assumptions:
+This model is a simplification of the notification process, in particular it makes 3 assumptions:
 \begin{enumerate}
@@
 
-\item notified individuals will eventually get sampled (i.e., their sampling is modeled with a rate $\upsilon$ instead of a combination of a removal rate $\psi$ and a sampling probability $\rho$);\\
+\item notified individuals will eventually get sampled (i.e., their pathogens are always observed upon removal, unlike the standard case where the observation happens with a sampling probability $\rho$);\\
 
@@
 %\item notified individuals do not notify further;
-\item once notified, the partners do not transmit further;\\
-
-We assume here that once aware of the potential infection, the notified patients take care to avoid further transmission.
+%\item once notified, the partners do not transmit further;\\
+%
+%We assume here that once aware of the potential infection, the notified patients take care to avoid further transmission.
 Moreover, we assume that the diagnosis and follow-up process works as expected, so once the notified patients are sampled, they are either isolated or provided with treatment or other means of preventing further transmission.
@@
 
-The -PN extension can be generalized from BD to MTBD models: at the moment of sampling, an individual of any type can notify their last partner with the probability $\upsilon$, who will then stop transmitting and get removed at a constant removal rate $\phi >> \psi_k \; \forall k \in \{1, \ldots, m\}$.
+The -PN extension can be generalized from BD to MTBD models: at the moment of sampling, an individual of any type can notify their last partner with the probability $\upsilon$, whose removal rate will then get replaced by  a constant sampling rate $\phi >> \psi_k \; \forall k \in \{1, \ldots, m\}$.
 The -PN extension therefore adds two parameters to an MTBD model: $\phi$ and $\upsilon$.
@@
 
-With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch that finishes at time $t_i$, starting on this branch at a time $t \leq t_i$.
+With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch $i$, starting on this branch at a time $t \leq t_i$ till time $t_i$.
 The master equations for the BD models were initially developed by ~\citet{Stadler2009}, however what we present here is their branch-specific formulation that we proposed in~\citep{zhukovaFastAccurateMaximumLikelihood2022}:
 In our formulation, ${p}^{(i)}(t)$ describes an evolution along a tree branch $i$, without taking into account the branch's subtree.
+$t_i$ corresponds to the time of the equation's initial condition (typically at the end of the branch).
 % and provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$.
@@
 \caption{An example of partner notification in a transmission tree: an individual A got sampled at time $t_A$ (violet tip) and notified their most recent partner B (orange), where B is either observed (\textbf{1}) or hidden (\textbf{2-3}).
-\\
 The (observed as in \textbf{1-2} or hidden as in \textbf{3}) internal node AB corresponds to the transmission between A and B.
@@
 Note, that since the most-recent-partner relationship is not always symmetric, B might be the most recent partner of more than one individual and hence might have got notified by someone else too (e.g.
-C in \textbf{b}, who in \textbf{1b,2b} was infected by B, and in \textbf{3b} could be infected by or infect B, who then infected A).
-Moreover, between the moment of B's first notification ($t_A$ for \textbf{a}, $t_C$ for \textbf{b}) and the moment of sampling  of B ($t_B$) there are no hidden events on the branch B (coloured dark red).
-Finally, B might have been removed via a standard procedure before getting notified by their partner(s) (\textbf{c}).\\
-In the panels \textbf{3} B's subtree is fully hidden, and hence the transmission node AB is also unobserved, therefore the A's branch is composed of two parts, where the bottom part (violet) corresponds to A and contains no hidden transmissions, while the top part could correspond or not to A (e.g in \textbf{3b} it corresponds to B) and might contain hidden transmissions.
-\\
+C in \textbf{b}, who in \textbf{1b,2b} was infected by B, and in \textbf{3b} could have been infected by or have infected B, who have then infected A).
+Moreover, between the moment of B's first notification ($t_A$ for \textbf{a}, $t_C$ for \textbf{b}) and the moment of sampling  of B ($t_B$) the removal rate of B becomes $\phi$ instead of $\psi$ (coloured dark red).
+Finally, B might have been removed via a standard procedure before getting notified by their partner(s) (with rate $\psi$ and a probability $\rho$, \textbf{c}).
+In the panels \textbf{3} B's subtree is fully hidden, and hence the transmission node AB is also unobserved, therefore the A's branch is composed of two parts, where the bottom part (violet) corresponds to A and contains no hidden transmissions, while the top part could correspond to A or not  (e.g in \textbf{3b} it corresponds to B) and might contain hidden transmissions.
 In the panels \textbf{2-3a-b} B is hidden as by the end of the sampling period ($T$) B is not yet sampled, while in \textbf{2-3c} B is hidden as B got removed via the standard procedure without sampling before any notification.}
@@
 
-According to our model assumption 4, only the most recent partner can get notified.
+According to our model assumption 3, only the most recent partner can get notified.
 Hence, AB must be the parent node of A, and no hidden transmission is possible along the branch AB-A (as otherwise there would have been a partner of A who is more recent than B).
@@
 \dot{p}_n^{(A)}(t) &=  \Big(\lambda + \psi\Big) p_n^{(A)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-p_n^{(A)}(t_A) &= 1\label{eq:p-nh}
+p_n^{(A)}(t_A) &= 1\label{eq:p-n}
 \end{cases}
@@
 
-On the branch of B, no transmission is possible between the moment of notification $t_A$ and the moment of B's sampling $t_B$ as, according to assumption 3, once notified partners do not transmit further.
-We will use a \textbf{notified partner branch probability} $p_p^{(i)}(t)$ for cases when the individual at the end of the branch is the partner who got notified at time prior to $t$ (coloured dark red in Fig~\ref{fig:pn-branches}):
+As B corresponds to A's partner, the individual on the path between the transmission node AB and the tip B cannot change (is always B) and 
+%before the notification time $t_A$ 
+only transmissions from B to someone else are possible along the path leading from AB to B.
+These transmissions can be either hidden (where the recipient tree stayed unobserved) or observed: in such a case we know which subtree corresponds to the donor (the one that contains B).
+However, depending on whether B already got notified ($t \geq t_A$ in Fig.~\ref{fig:pn-branches}(1a)) or not yet (($t < t_A$ in Fig.~\ref{fig:pn-branches}(1a))) B's removal rate would be different (correspondingly $\phi$ or $\psi$).
+We will use an \textbf{partner branch probability before notification} $p_{pb}^{(x)}(t)$ for cases when the individual at the end of the branch (or considered time interval, here $t_x=t_A$) is the same as at time $t$, has not yet been notified, but might have transmitted to someone else (unobserved) along the way (coloured orange in Fig~\ref{fig:pn-branches}):
 
@@
 \begin{cases}
-\dot{p}_p^{(B)}(t) &=  \phi p_p^{(B)}(t)\; \textit{\color{gray} $\leftarrow$ no sampling in the next infinitesimal time $\Delta t$ }\\
-p_p^{(B)}(t_B) &=  1\label{eq:p-np}
+\dot{p}_{pb}^{(x)}(t) &=~~  \Big(\lambda + \psi\Big) p_{pb}^{(x)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+    &~~~ - \lambda p_{pb}^{(x)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission with hidden recipient subtree}\\
+p_{pb}^{(x)}(t_x) &=  1\label{eq:p-o}
 \end{cases}
@@
 
-As B corresponds to A's partner, the individual on the path between the transmission node AB and the tip B cannot change (is always B) and before the notification time $t_A$ only transmissions from B to someone else are possible along the path leading from AB to B.
-These transmissions can be either hidden (where the recipient tree stayed unobserved) or observed: in such a case we know which subtree corresponds to the donor (the one that contains B).
-We will use an \textbf{oriented (partner) branch probability} $p_o^{(B)}(t)$ for cases when the individual at the end of the branch (or considered time interval, here $t_A$) is the same as at time $t$ but might have transmitted to someone else (unobserved) along the way (coloured orange in Fig~\ref{fig:pn-branches}):
+We will use a \textbf{partner branch probability after notification} $p_{pa}^{(i)}(t)$ for cases when the individual at the end of the branch is the same as at time $t$, by time $t$ has already been notified, and might have transmitted to someone else (unobserved) along the way (coloured dark red in Fig~\ref{fig:pn-branches}):
 
@@
 \begin{cases}
-\dot{p}_o^{(B)}(t, t_A) &=~~  \Big(\lambda + \psi\Big) p_o^{(B)}(t, t_A)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-    &~~~ - \lambda p_o^{(B)}(t, t_A)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission with hidden recipient subtree}\\
-p_o^{(B)}\Big(min(t_A, t_B), t_A\Big) &=  1\label{eq:p-o}
+\dot{p}_{pa}^{(B)}(t) &=~~  \Big(\lambda + \phi\Big) p_{pa}^{(B)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+    &~~~ - \lambda p_{pa}^{(B)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission with hidden recipient subtree}\\
+p_{pa}^{(B)}(t_B) &=  1\label{eq:p-p}
 \end{cases}
@@
 
-The equations~\ref{eq:Us}, \ref{eq:p}, \ref{eq:p-nh}-\ref{eq:p-o} have the following closed form solutions:
+Finally, once sampled, B could have notified their own most recent partner (represented by the most recent hidden subtree in Fig.~\ref{fig:pn-branches}(1a)).
+This would imply that between the moment of transmission between B and their most recent partner ($t_x$) and the moment of B's sampling ($t_B$) there was no event along the B's branch.
+Hence, if B is a notifier, the part of the B's branch after $t_x$ should be modeled with a notifier probability instead (no event along the branch).
+Note that once B is notified, the removal rate becomes $\phi$, and hence a \textbf{both notified and notifier branch probability} $p_b^{(i)}(t)$ applies:
+
+\begin{equation}
+\scriptsize
+\begin{cases}
+\dot{p}_b^{(B)}(t) &=  \Big(\lambda + \phi\Big) p_b^{(B)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+p_b^{(B)}(t_B) &= 1\label{eq:p-b}
+\end{cases}
+\end{equation}
+
+%On the branch of B, no transmission is possible between the moment of notification $t_A$ and the moment of B's sampling $t_B$ as, according to assumption 3, once notified partners do not transmit further.
+We will use a \textbf{notified partner branch probability} $p_{pa}^{(i)}(t)$ for cases when the individual at the end of the branch is the partner who got notified at time prior to $t$ (coloured dark red in Fig~\ref{fig:pn-branches}):
+%
+%\begin{equation}
+%\scriptsize
+%\begin{cases}
+%\dot{p}_p^{(B)}(t) &=  \phi p_{pa}^{(B)}(t)\; \textit{\color{gray} $\leftarrow$ no sampling in the next infinitesimal time $\Delta t$ }\\
+%p_{pa}^{(B)}(t_B) &=  1\label{eq:p-np}
+%\end{cases}
+%\end{equation}
+
+The equations~\ref{eq:Us}, \ref{eq:p}, \ref{eq:p-n}-\ref{eq:p-b} have the following closed form solutions:
 \begin{equation}
@@
 \\
-p_o^{(i)}(t, t_r) =  \frac{E_2\big(min(t_i, t_r)\big) + 1}{E_1(t) + 1}\sqrt{E_3\Big(t, min(t_i, t_r)\Big)E_4\Big(t, min(t_i, t_r)\Big)} \\
-\\
-p_n^{(i)}(t) =  E_4(t, t_i)
-\\
-p_p^{(i)}(t) =  e^{\phi (t -t_i)}
+p_n^{(i)}(t) =  e^{(\lambda + \psi) (t -t_i)}\\
+\\
+p_b^{(i)}(t) =  e^{(\lambda + \phi) (t -t_i)}
+\\
+\\
+p_{pb}^{(i)}(t) =  \sqrt{p^{(i)}(t)p_n^{(i)}(t)} \\
+\\
+p_{pa}^{(i)}(t) =  e^{(\phi-\psi) (t -t_i)} p_{pb}^{(i)}(t) 
 \end{cases},\\
 & \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi},\\
-& ~~~~~~~~E_1(t) = c_2 e^{c_1 (t -T)},\; E_2(t_i) = c_2 e^{c_1 (t_i -T)},\; E_3(t, t_i) = e^{c_1 (t -t_i)}, \; E_4(t, t_i) =  e^{(\lambda + \psi) (t -t_i)}\\
+& ~~~~~~~~E_1(t) = c_2 e^{c_1 (t -T)},\; E_2(t_i) = c_2 e^{c_1 (t_i -T)},\; E_3(t, t_i) = e^{c_1 (t -t_i)}\\
 \end{split}\label{eq:ps}
@@
 \item[1b.] by the time $t_A$, B was already notified by someone else (at time $t_C < t_A$, see Fig.~\ref{fig:pn-branches}(1b)).
-Note, that once notified partners do not transmit further, hence $t_C > t_{AB}$. \\
+%Note, that once notified partners do not transmit further, hence $t_C > t_{AB}$. 
+\\
 
@@
 
-The first part can be easily expressed via the previously defined equations: oriented (partner) branch evolution till time $t_A$ followed by notified partner branch evolution till time $T$.
-Note, that neither  $p_o^{(x)}(t, t_A)$ not $p_p^{(x)}(t)$ include an event at the end of the branch, and in this case the partner branch does not end at $t_x = T$:
-\begin{equation}
-U_p^{(1)}(t, t_A) = p_o^{(x)}(t, t_A) p_p^{(x)}(t_A)\textit{\color{gray}, where $t_x =T$}
+The first part can be easily expressed via the previously defined equations: partner branch evolution before notification till time $t_A$ followed by partner branch evolution after notification till time $T$.
+Note, that neither  $p_{pb}^{(x)}(t, t_A)$ not $p_{pa}^{(x)}(t)$ include an event at the end of the branch, and in this case the partner branch does not end at $t_x = T$:
+\begin{equation}
+U_p^{(1)}(t, t_A) = p_{pb}^{(A)}(t) p_{pa}^{(x)}(t_A)\textit{\color{gray}, where $t_x =T$}
 \end{equation}
@@
 For the second part we would need to integrate over all possible partner removal times $t_B \in [t, t_A[$, provided that in the interval between $t$ and $t_B$ the partner might have transmitted to someone else (potentially multiple times), whose subtree(s) stayed unobserved till $T$.
-Note that $p_o^{(x)}(t, t_A)$ describes a partner branch evolution between $t$ and $t_A$ with any number of hidden transmissions, no removal along it, and no event at the end of the branch.
+Note that $p_{pb}^{(A)}(t)$ describes a partner branch evolution between $t$ and $t_A$ with any number of hidden transmissions, no removal along it, and no event at the end of the branch.
 What we need instead is a removal with no sampling at time $\tau$.
-To approximate $U_p^{(2)}(t, t_A)$ we will divide $p_o^{(x)}(t, t_A)$ by the probability of no removal during the time $(t_A - t)$ (i.e.
+To approximate $U_p^{(2)}(t, t_A)$ we will divide $p_{pb}^{(A)}(t)$ by the probability of no removal during the time $(t_A - t)$ (i.e.
 $e^{-\psi(t_A - t)}$), multiply it by the probability of removal during that time ($1 - e^{-\psi(t_A - t)}$), and by the probability of not being sampled upon removal ($1 - \rho$):
@@
 \begin{equation}
-U_p^{(2)}(t, t_A) \approx p_o^{(x)}(t, t_A) \frac{1 - e^{-\psi(t_A - t)}}{e^{-\psi(t_A - t)}} (1 - \rho)\textit{\color{gray}, where $t_x =T$}
+U_p^{(2)}(t, t_A) \approx p_{pb}^{(A)}(t) \frac{1 - e^{-\psi(t_A - t)}}{e^{-\psi(t_A - t)}} (1 - \rho)
 \end{equation}
@@
 between AB and BC in Fig.~\ref{fig:pn-branches}(2), partner) and $s2$ denotes the type of the bottom of the branch (e.g.
-between AB and A in Fig.~\ref{fig:pn-branches}(3), index).
+between AB and A in Fig.~\ref{fig:pn-branches}(3), notifier).
 $t_r$ denotes the first notification time of the hidden partner.
@@
 \begin{split}
-p_{m(s1,s2)}^{(i)}(t,t_r) &= p_{top=s1}^{(h)}(t)p_{bottom=s2}^{(i)}(t_h)U_p(t_h,t_r)\textit{\color{gray},}\\
+p_{m(s1,s2)}^{(i)}(t,t_r) &= p_{top=s1}^{(h)}(t)U_p(t_h,t_r)p_{bottom=s2}^{(i)}(t_h)\textit{\color{gray},}\\
 &\textit{\color{gray}where $t_h$ is the time of the hidden partner subtree start,}\\
@@
 Having a notifier as the branch's tip implies that the bottom part of the branch corresponds to the notifier.
-Having a notifier in the supertree implies that the top part of the branch corresponds to the partner.
+Having a notifier in the supertree implies that the top part of the branch corresponds to the partner (already notified or not yet).
 Hence if the branch is internal, the top part can only correspond to the partner, while the bottom part can only correspond to a standard state (unnotified non-notifier), as the bottom part of such a branch corresponds to the partner's recipient (who could not be notified as the partner is unobserved and could not be a notifier as the branch is not external).
@@
 If, on the other hand, the branch is external, and the partner's notifier(s) is(are) only found in its supertree, then there are two possibilities for its bottom part: (2a) either it fully corresponds to a standard state (unnotified non-notifier); or (2b) there is another partners' notifier at its tip, and this other partner is also hidden along the bottom part of the same branch, hence the bottom part of the branch is standard-notifier (see Fig.~\ref{fig:pn-mixed}).
-Note that (1b) and (2b) represent the same case but from the point of view of two different hidden partners (B and their notifier C for the former, and D and their notifier C for the latter in Fig.~\ref{fig:pn-mixed}).
+Note that (1b) and (2b) represent the same case but from the point of view of two different hidden partners (B and their notifier A for the former, and D and their notifier C for the latter in Fig.~\ref{fig:pn-mixed}).
 
@@
 
-We will first describe the one-hidden-partner cases ($p_{m(s1,s2)}^{(i)}(t,t_r)$, where $s1 \in \{o, -\}$, $s2 \in \{n, -\}$ and $-$ stands for the standard (unnotified non-notifier) state), and then proceed with the two-partner case ($p_{m(o,-,n)}^{(i)}(t,t_{r})$, where $t_{r}$ is the earliest notification time for the hidden partner at the top of the branch (e.g.
+We will first describe the one-hidden-partner cases ($p_{m(s1,s2)}^{(i)}(t,t_r)$, where $s1 \in \{p, -\}$, $s2 \in \{n, -\}$ and $-$ stands for the standard (unnotified non-notifier) state), and then proceed with the two-partner case ($p_{m(p,-,n)}^{(i)}(t,t_{r})$, where $t_{r}$ is the earliest notification time for the hidden partner at the top of the branch (e.g.
 $t_r=t_C$ in Fig.~\ref{fig:pn-mixed}), the only notification time for the hidden partner at the bottom of the branch being $t_i$  (e.g.
@@
 We will approximate $p_{top=s1}^{(h)}(t)$ using the branch evolution probability of the type corresponding to $s1$  in the following way.
-For example, let us assume that $s1=o$ (observed (partner) branch).
-$p_o^{(h)}(t)$ expresses a probability of a partner evolving along the branch between the times $t$ and $t_h$ with no or any number of oriented hidden transmissions, without taking into account the event at the end of it.
-In our case there must be at least one hidden transmission: from the partner at time $t_h$ (where the donor tree stayed unsampled).
-If we remove the probability of no event during this time ($e^{-(\lambda + \psi)(t_h - t)}$) from $p_o^{(h)}(t)$, we will obtain a probability of a partner branch with at least one oriented hidden transmission somewhere between $t$ and $t_h$ (the probability of the corresponding hidden tree being included).
-$p_{top=o}^{(h)}(t)$ expresses almost the same thing, with the difference being that the last hidden transmission must have happened at time $t_h$ and the probability of the hidden partner tree corresponding to it is not included.
-We will hence approximate $p_{top=o}^{(h)}(t)$ as follows:
-\begin{equation}
-p_{top=o}^{(h)}(t) \approx \frac{p_o^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}}{U(t_{h})}
-\end{equation}
-
-We will treat the case when the branch is in the standard state in a similar way (using the standard branch probability $p^{(h)}(t)$ instead of the partner one $p_o^{(h)}(t)$):
+For example, let us assume that $s1=-$ (standard branch).
+$p^{(h)}(t)$ expresses a probability of evolving along this branch between the times $t$ and $t_h$ with no or any number of hidden transmissions, without taking into account the event at the end of it.
+In our case there must be at least one hidden transmission: to/from the partner at time $t_h$ (where the partner tree stayed unsampled).
+If we remove the probability of no event during this time ($e^{-(\lambda + \psi)(t_h - t)}$) from $p^{(h)}(t)$, we will obtain a probability of a branch with at least one oriented hidden transmission somewhere between $t$ and $t_h$ (the probability of the corresponding hidden tree being included).
+$p_{top=-}^{(h)}(t)$ expresses almost the same thing, with the difference being that the last hidden transmission must have happened at time $t_h$ and the probability of the hidden partner tree corresponding to it is not included.
+We will hence approximate $p_{top=-}^{(h)}(t)$ as follows:
 \begin{equation}
@@
 
+We will treat the case when the branch is in the partner state ($s1=p$) in a similar way:
+\begin{equation}
+\begin{split}
+p_{top=p}^{(h)}(t, t_r) &\approx 
+\begin{cases}
+\frac{p_{pb}^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}}{U(t_{h})}&\textit{\color{gray} if $t_h \leq t_r$}\\
+p_{pb}^{(r)}(t)\frac{p_{pa}^{(h)}(t_r) -e^{-(\lambda + \psi)(t_{h} - t_r)}}{U(t_{h})}&\textit{\color{gray} if $t \leq t_r \leq t_h$}\\
+\frac{p_{pa}^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}}{U(t_{h})}&\textit{\color{gray} if $t > t_r$}\\
+\end{cases}\\
+&\textit{\color{gray}where $t_r$ is the first notification time of the hidden partner.}\\
+\end{split}
+\end{equation}
+
 Note that we wrote down equation~(\ref{eq:mixed}) as if the start of the hidden partner tree $t_h$ was known.
-In practice, it is not the case as we need to integrate over all the possible times between the branch start $t_{parent(i)}$ and either its end or the first partner's notification (if it was before): $min(t_r,t_i)$.
-To simplify the calculations, we will simply take the time in the middle of this interval instead:
-
-\begin{equation}
-\begin{split}
-p_{m(s1,s2)}^{(i)}(t,t_r) &\approx \Big(p_{s1}^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}\Big)\frac{U_p(t_h,t_r)}{U(t_{h})}p_{s2}^{(i)}(t_h)\textit{\color{gray},}\\
-&\textit{\color{gray}where $t_h \approx t + \frac{min(t_r, t_i) - t}{2}$,}\\
-&\textit{\color{gray}~~~~~~~~$t_r$ is the first notification time of the hidden partner.}\\
-\end{split}\label{eq:p_mixed}
+In practice, it is not the case as we need to integrate over all the possible times between the branch start $t_{parent(i)}$ and its end $t_i)$.
+To simplify the calculations, we will take the time in the middle of this interval instead:
+
+\begin{equation}
+t_h \approx t + \frac{t_i - t}{2}
 \end{equation}
@@
 \\
-p_o^{(i)}(t_j)\psi\rho(1 - \upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and did not notify}\\
+p_{pb}^{(i)}(t_j)\psi\rho(1 - \upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and did not notify}\\
 \textit{\color{gray} (oriented (partner) branch evolution followed by standard sampling and no notification)}
@@
 \\
-p_o^{(r)}(t_j) p_p^{(i)}(t_r) \phi (1-\upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_j \leq t_r \leq t_i$ and did not notify}\\
+p_{pb}^{(r)}(t_j) p_{pa}^{(i)}(t_r) \phi (1-\upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_j \leq t_r \leq t_i$ and did not notify}\\
 \textit{\color{gray} (oriented (partner) branch evolution followed by} \\
@@
 \\
-p_n^{(r)}(t_j) p_p^{(i)}(t_r) \phi \upsilon \textit{\color{gray} ~if $i$ got notified at time $t_j \leq t_r \leq t_i$ and notified}\\
+p_n^{(r)}(t_j) p_{pa}^{(i)}(t_r) \phi \upsilon \textit{\color{gray} ~if $i$ got notified at time $t_j \leq t_r \leq t_i$ and notified}\\
 \textit{\color{gray} (notifier branch evolution followed by} \\
@@
 There are two possibilities: either the node $i$ corresponds to the partner or the partner is hidden along the $i$'s branch and $i$ corresponds to someone within $i$'s recipient subtree (e.g., if $i=C$ in Fig.~\ref{fig:pn-branches}(2a)).
-In the first case, the $i$'s branch evolution is represented by $p_o^{(i)}(t_j)$ with a transmission event at the end of it, where the donor is known (probability density of $\lambda$): The donor corresponds to the subtree containing the tip representing the sampling of the individual $i$.
+In the first case, the $i$'s branch evolution is represented by $p_{pb}^{(i)}(t_j)$ with a transmission event at the end of it, where the donor is known (probability density of $\lambda$): The donor corresponds to the subtree containing the tip representing the sampling of the individual $i$.
 
@@
 \begin{split}
-l^{(n)}(i, r) = &p_o^{(i)}(t_j) \lambda \cdot
+l^{(n)}(i, r) = &p_{pb}^{(i)}(t_j) \lambda \cdot
 \Big(~l^{(n)}(i_0,r)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ is notified by $r$} \\
@@
 \begin{split}
-l^{(n)}(i, r) = &p_o^{(i)}(t_j) \lambda \cdot
+l^{(n)}(i, r) = &p_{pb}^{(i)}(t_j) \lambda \cdot
 \Big(~l^{(n,\not{n})}(i_0,first(r,i_1))l^{(\not{n},n)}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ and $i_1$ notified $i_0$, $i_0$ did not notify} \\
@@
 \begin{split}
-l^{(n)}(i, r) = &p_o^{(i)}(t_j) \lambda \cdot
+l^{(n)}(i, r) = &p_{pb}^{(i)}(t_j) \lambda \cdot
 \Big(~l^{(n,n)}(i_0,r)l^{(n)}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$  notified $i_1$} \\
@@
 
-For the PN version of this simulator, we add an additional $m+1^{th}$ state for notified partners.
-$m + 1$'s state-change and transmission rates are set to zero: $\mu_{k,m+1} = \mu_{m+1,k} = \lambda_{k,m+1} = \lambda_{m+1,k} = 0\; \forall 1 \leq k \leq m$, while its removal rate is set to the notification removal rate: $\psi_{m+1} = \phi >> \psi_k\; \forall 1 \leq k \leq m$.
+For the PN version of this simulator, we add additional $m$ states for notified-partner versions of each state.
+$m + k$'s state-change and transmission rates are set to the $k$'s state-change and transmission rates : $\mu_{m+k,m+l} = \mu_{k,l}$; $\lambda_{m+k,m+l} = \lambda_{k,l}$ $\forall 1 \leq  k,l \leq m$.
+The rates between initial and notified states set to zero: ;  $\mu_{m+k,l} = \mu_{k,m+l} = \lambda_{m+k,l} = \lambda_{k,m+l} = 0$.
+The removal rates of notified states are set to $\phi$: $\psi_{m+k} = \phi >> \psi_k$, while their sampling probabilities are set to $1$: $\rho_{m+k} = 1$.
 We keep a mapping of most recent partners $P$ for each infectious individual, and update it for both the donor and the recipient at each transmission event.
 We also keep a mapping between each infectious individual and their state, and update it at each state-change event.
-At each removal event, we check whether the individual $i$ is a notified partner (i.e.
-in state $m+1$).
-If yes, we put them into $N[m+1]$ without performing the sampling probability draw.
-If not, and if $i$ gets sampled upon removal, we draw a value $p$ (uniformly) from an interval $[0, 1[$.
-If $p < \upsilon$, and if $i$'s most recent partner ($j = P[i]$) is not yet sampled (not in $N[state(j)]$), we change $j$'s state to $m + 1$ as in a state-change event.
+At each removal event, if the removed individual $i$ gets sampled upon removal, we draw a value $p$ (uniformly) from an interval $[0, 1[$.
+If $p < \upsilon$, and if $i$'s most recent partner ($j = P[i]$) is not yet sampled (not in $N[state(j)]$) and not yet notified ($J \leq m$), we change $j$'s state to $m + j$ as in a state-change event.
 

### 2024-06-06 article/article.tex
Transmission from notified partners is possible

@@
 &\begin{cases}
-U(t) = \frac{\lambda + \psi}{2\lambda} +  \frac{c_1}{2\lambda}\Big(\frac{E_1(t) - 1}{E_1(t) + 1}\Big)\\
-\\
-p^{(i)}(t) = \Big(\frac{E_2(t_i) + 1}{E_1(t) + 1}\Big)^2E_3(t, t_i) \\
+U(t) = \frac{\lambda + \psi}{2\lambda} +  \frac{c_1}{2\lambda}\Big(\frac{E(t) - 1}{E(t) + 1}\Big)\\
+\\
+p^{(i)}(t) = \Big(\frac{E(t_i) + 1}{E(t) + 1}\Big)^2 e^{c_1 (t -t_i)} \\
 \\
@@
 & \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi},\\
-& ~~~~~~~~E_1(t) = c_2 e^{c_1 (t -T)},\; E_2(t_i) = c_2 e^{c_1 (t_i -T)},\; E_3(t, t_i) = e^{c_1 (t -t_i)}\\
+& ~~~~~~~~E(t) = c_2 e^{c_1 (t -T)} \\
 \end{split}\label{eq:ps}
@@
 &\textit{\color{gray}~~~~~~~~$t_r$ is the first notification time of the hidden partner.}\\
-\end{split}
-\end{equation}\label{eq:mixed}
+\end{split}\label{eq:p_mixed}
+\end{equation}
 
@@
 \begin{cases}
-p_n^{(i)}(t)&\textit{\color{gray} if $i$ is a notifier tip ($s2=n$),}\\
-p^{(i)}(t)&\textit{\color{gray} otherwise ($s2=-$, the standard state)}\\
+p_n^{(i)}(t_h)&\textit{\color{gray} if $i$ is a notifier tip ($s2=n$),}\\
+p^{(i)}(t_h)&\textit{\color{gray} otherwise ($s2=-$, the standard state)}\\
 \end{cases}
@@
 \frac{p_{pb}^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}}{U(t_{h})}&\textit{\color{gray} if $t_h \leq t_r$}\\
-p_{pb}^{(r)}(t)\frac{p_{pa}^{(h)}(t_r) -e^{-(\lambda + \psi)(t_{h} - t_r)}}{U(t_{h})}&\textit{\color{gray} if $t \leq t_r \leq t_h$}\\
-\frac{p_{pa}^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}}{U(t_{h})}&\textit{\color{gray} if $t > t_r$}\\
+p_{pb}^{(r)}(t)\frac{p_{pa}^{(h)}(t_r) -e^{-(\lambda + \phi)(t_{h} - t_r)}}{U(t_{h})}&\textit{\color{gray} if $t \leq t_r \leq t_h$}\\
+\frac{p_{pa}^{(h)}(t) -e^{-(\lambda + \phi)(t_{h} - t)}}{U(t_{h})}&\textit{\color{gray} if $t > t_r$}\\
 \end{cases}\\
@@
 
-Note that we wrote down equation~(\ref{eq:mixed}) as if the start of the hidden partner tree $t_h$ was known.
+Note that we wrote down equation~(\ref{eq:p_mixed}) as if the start of the hidden partner tree $t_h$ was known.
 In practice, it is not the case as we need to integrate over all the possible times between the branch start $t_{parent(i)}$ and its end $t_i)$.
@@
 \begin{split}
-p_{m(o,-,n)}^{(i)}(t,t_r) &\approx \Big(p_{o}^{(h1)}(t) -e^{-(\lambda + \psi)(t_{h1} - t)}\Big)\frac{U_p(t_{h1},t_{r})}{U(t_{h1})}\cdot
-\\
-&~~~~\Big(p^{(h2)}(t_{h1}) -e^{-(\lambda + \psi)(t_{h2} - t_{h1})}\Big)\frac{U_p(t_{h2},t_{i})}{U(t_{h2})}\cdot
-\\
-&~~~~~p_n^{(i)}(t_{h2})\textit{\color{gray},}\\
-&\textit{\color{gray}where $t_{h1} \approx t + \frac{1}{3}\Big(min(t_{r}, t_i) - t\Big)$,}\\
-&\textit{\color{gray}~~~~~~~~$t_{h2} \approx t + \frac{2}{3}\Big(min(t_{r}, t_i) - t\Big)$,}\\
-&\textit{\color{gray}~~~~~~~~$t_{r}$ is the earliest notification time of the }\\
-&\textit{\color{gray}~~~~~~~~~~~~hidden partner at the top of the branch (o,-).}\\
+p_{m(p,-,n)}^{(i)}(t,t_r) &\approx p_{top=p}^{(h1)}(t, t_r)U_p(t_{h1},t_{r}) p_{top=-}^{(h2)}(t_{h1}, t_i)U_p(t_{h2},t_{i}) p_n^{(i)}(t_{h2})\textit{\color{gray},}\\
+&\textit{\color{gray}where $t_{h1} \approx t + \frac{1}{3}(t_i - t)$,}\\
+&\textit{\color{gray}~~~~~~~~$t_{h2} \approx t + \frac{2}{3}(t_i - t)$,}\\
+&\textit{\color{gray}~~~~~~~~$t_{r}$ is the earliest notification time of the partner }\\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~hidden at the top of the branch.}\\
 \end{split}\label{eq:p_mixed_mixed}
@@
 l^{(n,\not{n})}(i,r) &= \begin{cases}
-0 \textit{\color{gray} ~if $t_r < t_j$}\\
-\textit{\color{gray} (not possible as once notified partners do not transmit)}
+%0 \textit{\color{gray} ~if $t_r < t_j$}\\
+%\textit{\color{gray} (not possible as once notified partners do not transmit)}
+%\\
+%\\
+p_{pa}^{(i)}(t_j)\phi(1 - \upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_r < t_j$ and did not notify}\\
+\textit{\color{gray} (notified partner branch evolution followed by notification-provoked sampling and no notification)}
+\\
+~~~+p_{m(pa,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r < t_j$, }\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$  did not notify}\\
+ \textit{\color{gray} ~~~(mixed partner-standard branch evolution followed by sampling and no notification)}\\
+ ~~~+p_{m(pa,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
+ \textit{\color{gray} ~~~(mixed partner-notifier branch evolution followed by sampling and notification)}
+\\
+~~~+p_{m(pa,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r < t_j$, }\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
+ \textit{\color{gray} ~~~(mixed partner-standard-notifier branch evolution followed by sampling and notification)}
 \\
@@
 p_{pb}^{(i)}(t_j)\psi\rho(1 - \upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and did not notify}\\
-\textit{\color{gray} (oriented (partner) branch evolution followed by standard sampling and no notification)}
-\\
-~~~+p_{m(o,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r > t_i$, }\\
+\textit{\color{gray} (partner-before-notification branch evolution followed by standard sampling and no notification)}
+\\
+~~~+p_{m(pb,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r > t_i$, }\\
 \textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$  did not notify}\\
- \textit{\color{gray} ~~~(mixed oriented-standard branch evolution followed by sampling and no notification)}\\
- ~~~+p_{m(o,n)}^{(i)}(t_j,t_i)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
- \textit{\color{gray} ~~~(mixed oriented-notifier branch evolution followed by sampling and notification)}
-\\
-~~~+p_{m(o,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r > t_i$, }\\
+ \textit{\color{gray} ~~~(mixed partner-standard branch evolution followed by sampling and no notification)}\\
+ ~~~+p_{m(pb,n)}^{(i)}(t_j,t_i)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
+ \textit{\color{gray} ~~~(mixed partner-notifier branch evolution followed by sampling and notification)}
+\\
+~~~+p_{m(pb,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r > t_i$, }\\
 \textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
- \textit{\color{gray} ~~~(mixed oriented-standard-notifier branch evolution followed by sampling and notification)}
+ \textit{\color{gray} ~~~(mixed partner-standard-notifier branch evolution followed by sampling and notification)}
 \\
@@
 p_{pb}^{(r)}(t_j) p_{pa}^{(i)}(t_r) \phi (1-\upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_j \leq t_r \leq t_i$ and did not notify}\\
-\textit{\color{gray} (oriented (partner) branch evolution followed by} \\
+\textit{\color{gray} (partner-before-notification branch evolution followed by} \\
 \textit{\color{gray}~notified partner branch evolution, notification-provoked sampling and no notification)}
 \\
-~~~+p_{m(o,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_j \leq t_r \leq t_i$, }\\
+~~~+p_{m(p,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_j \leq t_r \leq t_i$, }\\
 \textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$  did not notify}\\
- \textit{\color{gray} ~~~(mixed oriented-standard branch evolution followed by sampling and no notification)}\\
- ~~~+p_{m(o,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
- \textit{\color{gray} ~~~(mixed oriented-notifier branch evolution followed by sampling and notification)}
-\\
-~~~+p_{m(o,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_j \leq t_r \leq t_i$, }\\
+ \textit{\color{gray} ~~~(mixed partner-standard branch evolution followed by sampling and no notification)}\\
+ ~~~+p_{m(p,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
+ \textit{\color{gray} ~~~(mixed partner-notifier branch evolution followed by sampling and notification)}
+\\
+~~~+p_{m(p,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_j \leq t_r \leq t_i$, }\\
 \textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
- \textit{\color{gray} ~~~(mixed oriented-standard-notifier branch evolution followed by sampling and notification)}
+ \textit{\color{gray} ~~~(mixed partner-standard-notifier branch evolution followed by sampling and notification)}
 \end{cases}\\
 \\
-l^{(n,n)}(i,r) &= \begin{cases}0 \textit{\color{gray} ~if $t_r < t_j$}\\
-\textit{\color{gray} (not possible as once notified partners do not transmit)}\\
+l^{(n,n)}(i,r) &= \begin{cases}
+p_b^{(i)}(t_j)\phi\upsilon \textit{\color{gray} ~if $i$ got notified at time $t_r < t_j$ and notified their most recent patner}\\
+\textit{\color{gray} (notified notifier branch evolution followed by notification-provoked sampling and  notification)}\\
 \\
@@
 \\
-p_n^{(r)}(t_j) p_{pa}^{(i)}(t_r) \phi \upsilon \textit{\color{gray} ~if $i$ got notified at time $t_j \leq t_r \leq t_i$ and notified}\\
-\textit{\color{gray} (notifier branch evolution followed by} \\
-\textit{\color{gray}~notified partner branch evolution,} \\
+p_n^{(r)}(t_j) p_b^{(i)}(t_r) \phi \upsilon \textit{\color{gray} ~if $i$ got notified at time $t_j \leq t_r \leq t_i$ and notified}\\
+\textit{\color{gray} (notifier branch evolution followed by notified notifier branch evolution,} \\
 \textit{\color{gray}~notification-provoked sampling and notification)}\\
@@
 There are two possibilities: either the node $i$ corresponds to the partner or the partner is hidden along the $i$'s branch and $i$ corresponds to someone within $i$'s recipient subtree (e.g., if $i=C$ in Fig.~\ref{fig:pn-branches}(2a)).
-In the first case, the $i$'s branch evolution is represented by $p_{pb}^{(i)}(t_j)$ with a transmission event at the end of it, where the donor is known (probability density of $\lambda$): The donor corresponds to the subtree containing the tip representing the sampling of the individual $i$.
-
-In the second case, the $i$'s branch evolution is represented by $p_{m(o,-)}^{(i)}(t_j,t_r)$ with a transmission event at the end of it, where the donor is unknown (probability density of $2\lambda$): any of the subtrees can be the donor.
+In the first case, the $i$'s branch evolution is represented by the partner branch evolution $p_{p}^{(i)}(t_j, t_r)$:
+
+\begin{equation}
+\scriptsize
+p_{p}^{(i)}(t_j, t_r) =
+\begin{cases}
+p_{pa}^{(i)}(t_j) &\textit{\color{gray} if $t_r < t_j$,}\\
+p_{pb}^{(i)}(t_j) &\textit{\color{gray} if $t_r > t_i$,}\\
+p_{pb}^{(r)}(t_j)p_{pa}^{(i)}(t_r) &\textit{\color{gray} if $t_j \leq t_r \leq t_i$.}\\
+\end{cases}\label{eq:p-p}
+\end{equation}
+This branch has a transmission event at the end of it, where the donor is known (probability density of $\lambda$): The donor corresponds to the subtree containing the tip representing the sampling of the individual $i$.
+
+In the second case, the $i$'s branch evolution is represented by $p_{m(p,-)}^{(i)}(t_j,t_r)$ with a transmission event at the end of it, where the donor is unknown (probability density of $2\lambda$): any of the subtrees can be the donor.
  %Moreover, this recipient's branch cannot be notified by our partner (as in our model notified partners do not notify further):
+
 Again, we will consider three possibilities for $i$'s child nodes: (2.2.1) when both $i_0$ and $i_1$ are internal nodes; (2.2.2) when both $i_0$ and $i_1$ are tips; and (2.2.3) when one of them is an internal node and the other one is a tip.
 
-In any of these scenarios, the notification must have happened after the partner tip branch start (as one of the model assumptions is that partners do not transmit after notification). Hence:
-\begin{equation}
-\scriptsize
-l^{(n)}(i, r) = 0 \textit{\color{gray}~if $t_i > t_r$} \\
-\label{eq:lp-zero}
-\end{equation}
+%In any of these scenarios, the notification must have happened after the partner tip branch start (as one of the model assumptions is that partners do not transmit after notification). Hence:
+%\begin{equation}
+%\scriptsize
+%l^{(n)}(i, r) = 0 \textit{\color{gray}~if $t_i > t_r$} \\
+%\label{eq:lp-zero}
+%\end{equation}
 
@@
 \begin{split}
-l^{(n)}(i, r) = &p_{pb}^{(i)}(t_j) \lambda \cdot
+l^{(n)}(i, r) = &p_{p}^{(i)}(t_j, t_r) \lambda \cdot
 \Big(~l^{(n)}(i_0,r)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ is notified by $r$} \\
-& ~~~~~~~~~~~~~~~+ l^{(\not{n})}(i_0)l^{(n)}(i_1,r)~\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ is notified by $r$} \\
-& + p_{m(o,-)}^{(i)}(t_j,t_r) 2 \lambda l^{(\not{n})}(i_0)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ hidden partner}\\
+& ~~~~~~~~~~~~~~~~~~~+ l^{(\not{n})}(i_0)l^{(n)}(i_1,r)~\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ is notified by $r$} \\
+& + p_{m(p,-)}^{(i)}(t_j,t_r) 2 \lambda l^{(\not{n})}(i_0)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ hidden partner}\\
  \end{split}
@@
 \begin{split}
-l^{(n)}(i, r) = &p_{pb}^{(i)}(t_j) \lambda \cdot
+l^{(n)}(i, r) = &p_{p}^{(i)}(t_j,t_r) \lambda \cdot
 \Big(~l^{(n,\not{n})}(i_0,first(r,i_1))l^{(\not{n},n)}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ and $i_1$ notified $i_0$, $i_0$ did not notify} \\
-&~~~~~~~~~~~~~~~+ l^{(n,n)}(i_0,first(r,i_1))l^{(n,n)}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $r$ and $i_1$ notified $i_0$, $i_0$ notified $i_1$} \\
-&~~~~~~~~~~~~~~~+l^{(\not{n},n)}(i_0)l^{(n,\not{n})}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ $r$ and $i_0$ notified $i_1$, $i_1$ did not notify} \\
-&~~~~~~~~~~~~~~~+ l^{(n,n)}(i_0, i_1)l^{(n,n)}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ $r$ and $i_0$ notified $i_1$, $i_1$ notified $i_0$} \\
-&~~~~~~~~~~~~~~~+l^{(n,n)}(i_0,r)l^{(n,\not{n})}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$ notified $i_1$, $i_1$ did not notify} \\
-&~~~~~~~~~~~~~~~+ l^{(n,\not{n})}(i_0, i_1)l^{(n,n)}(i_1,r) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_1$ notified $i_0$, $i_0$ did not notify} \\
-&~~~~~~~~~~~~~~~+l^{(n,\not{n})}(i_0,r)l^{(\not{n},\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$ and $i_1$ did not notify} \\
-&~~~~~~~~~~~~~~~+ l^{(\not{n},\not{n})}(i_0)l^{(n,\not{n})}(i_1,r)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_0$ and $i_1$ did not notify} 
-\\
-&+ p_{m(o,-)}^{(i)}(t_j,t_r) 2 \lambda \cdot
+&~~~~~~~~~~~~~~~~~~~+ l^{(n,n)}(i_0,first(r,i_1))l^{(n,n)}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $r$ and $i_1$ notified $i_0$, $i_0$ notified $i_1$} \\
+&~~~~~~~~~~~~~~~~~~~+l^{(\not{n},n)}(i_0)l^{(n,\not{n})}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ $r$ and $i_0$ notified $i_1$, $i_1$ did not notify} \\
+&~~~~~~~~~~~~~~~~~~~+ l^{(n,n)}(i_0, i_1)l^{(n,n)}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ $r$ and $i_0$ notified $i_1$, $i_1$ notified $i_0$} \\
+&~~~~~~~~~~~~~~~~~~~+l^{(n,n)}(i_0,r)l^{(n,\not{n})}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$ notified $i_1$, $i_1$ did not notify} \\
+&~~~~~~~~~~~~~~~~~~~+ l^{(n,\not{n})}(i_0, i_1)l^{(n,n)}(i_1,r) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_1$ notified $i_0$, $i_0$ did not notify} \\
+&~~~~~~~~~~~~~~~~~~~+l^{(n,\not{n})}(i_0,r)l^{(\not{n},\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$ and $i_1$ did not notify} \\
+&~~~~~~~~~~~~~~~~~~~+ l^{(\not{n},\not{n})}(i_0)l^{(n,\not{n})}(i_1,r)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_0$ and $i_1$ did not notify} 
+\\
+&+ p_{m(p,-)}^{(i)}(t_j,t_r) 2 \lambda \cdot
 \Big(~l^{(n,\not{n})}(i_0,i_1)l^{(\not{n},n)}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
@@
 \begin{split}
-l^{(n)}(i, r) = &p_{pb}^{(i)}(t_j) \lambda \cdot
+l^{(n)}(i, r) = &p_{p}^{(i)}(t_j,t_r) \lambda \cdot
 \Big(~l^{(n,n)}(i_0,r)l^{(n)}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$  notified $i_1$} \\
-&~~~~~~~~~~~~~~~+ l^{(n,\not{n})}(i_0,r)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$  $r$ notified $i_0$, $i_0$ did not notify} \\
-&~~~~~~~~~~~~~~~+l^{(\not{n}, n)}(i_0)l^{(n)}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ both $r$ and $i_0$ notified $i_1$} \\
-&~~~~~~~~~~~~~~~+ l^{(\not{n},\not{n})}(i_0)l^{(n)}(i_1,r)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_0$ did not notify} 
-\\
-&+ p_{m(o,-)}^{(i)}(t_j,t_r) 2\lambda \cdot
+&~~~~~~~~~~~~~~~~~~~+ l^{(n,\not{n})}(i_0,r)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$  $r$ notified $i_0$, $i_0$ did not notify} \\
+&~~~~~~~~~~~~~~~~~~~+l^{(\not{n}, n)}(i_0)l^{(n)}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ both $r$ and $i_0$ notified $i_1$} \\
+&~~~~~~~~~~~~~~~~~~~+ l^{(\not{n},\not{n})}(i_0)l^{(n)}(i_1,r)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_0$ did not notify} 
+\\
+&+ p_{m(p,-)}^{(i)}(t_j,t_r) 2\lambda \cdot
 \Big(~l^{(\not{n},n)}(i_0)l^{(n)}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
@@
 
-Note that $l^{(n)}(i, r)$ and $p_{m(o,-)}^{(i)}(t_j,t_r)$ depends on the notifier $r$ and can only be calculated once the notification time $t_r$ is known, i.e.
+\bigskip 
+Note that $l^{(n)}(i, r)$, $l^{(n,n)}(i, r)$ and $l^{(n,\not{n})}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time $t_r$ is known, i.e.
 once we have climbed to an internal node corresponding to the transmission between the notifier $r$ (corresponding to one of its children, must be a tip) and the partner (the other child).
 The calculation is done recursively by descending into the subtree.
-However, as all the unotified equation parts, $l^{(\not{n})}(i)$, only depend on the corresponding node $i$, they can be pre-calculated and reused to speed up the recursive steps.
+The tree likelihood calculation hence can take $O(n^2)$ steps in the worst case (ladder tree), where $n$ is the number of tips in the tree.
+However, as all the unotified equation parts ($l^{(\not{n})}(i)$, $l^{(\not{n},n)}(i)$ and $l^{(\not{n},\not{n})}(i)$) only depend on the corresponding node $i$, they can be pre-calculated and reused to speed up the recursive steps.
 

### 2024-06-06 bdpn/bd_model.py

@@ -2,7 +2,7 @@ import os
 
 import numpy as np
 
-from bdpn.formulas import get_c1, get_c2, get_E2, get_E1, get_log_p, get_u
+from bdpn.formulas import get_c1, get_c2, get_E, get_log_p, get_u
 from bdpn.parameter_estimator import optimize_likelihood_params
 from bdpn.tree_manager import TIME, read_forest, annotate_forest_with_time
 
@@ -68,13 +68,13 @@ def loglikelihood(forest, la, psi, rho, T=None, threads=1):
         for n in tree.traverse('preorder'):
             if not n.is_leaf():
                 t = getattr(n, TIME)
-                E1 = get_E1(c1=c1, c2=c2, t=t, T=T)
+                E1 = get_E(c1=c1, c2=c2, t=t, T=T)
                 child1, child2 = n.children
                 ti_1 = getattr(child1, TIME)
                 ti_2 = getattr(child2, TIME)
-                res += get_log_p(c1, t, ti=ti_1, E1=E1, E2=get_E2(c1=c1, c2=c2, ti=ti_1, T=T)) \
-                       + get_log_p(c1, t, ti=ti_2, E1=E1, E2=get_E2(c1=c1, c2=c2, ti=ti_2, T=T))
-    u = get_u(la, psi, c1, E1=get_E1(c1=c1, c2=c2, t=0, T=T))
+                res += get_log_p(c1, t, ti=ti_1, E_t=E1, E_ti=get_E(c1, c2, ti_1, T)) \
+                       + get_log_p(c1, t, ti=ti_2, E_t=E1, E_ti=get_E(c1, c2, ti_2, T))
+    u = get_u(la, psi, c1, E_t=get_E(c1=c1, c2=c2, t=0, T=T))
     return res + len(forest) * u / (1 - u) * np.log(u)
 
 

### 2024-06-06 bdpn/bdpn_model.py

@@ -2,11 +2,10 @@ import os
 from multiprocessing.pool import ThreadPool
 
 import numpy as np
-import pandas as pd
 
 from bdpn import bd_model
-from bdpn.formulas import get_log_p, get_c1, get_c2, get_E2, get_E1, get_log_po, get_log_pn, get_log_po_from_p_pn, \
-    get_u, get_log_pp, get_log_no_event
+from bdpn.formulas import get_log_p, get_c1, get_c2, get_E, get_log_ppb, get_log_pn, get_log_ppb_from_p_pn, \
+    get_u, get_log_no_event, get_log_ppa, get_log_ppa_from_ppb, get_log_pb
 from bdpn.parameter_estimator import optimize_likelihood_params, rescale_log
 from bdpn.tree_manager import TIME, read_forest, annotate_forest_with_time
 
@@ -40,11 +39,14 @@ def log_subtraction(log_minuend, log_subtrahend):
     :param log_subtrahend: logX2 in the formula above
     :return: log of the difference
     """
+    # print(log_subtrahend, log_minuend)
+    # assert (log_subtrahend < log_minuend)
     result = np.array([log_minuend, log_subtrahend], dtype=np.float64)
     factors = rescale_log(result)
     return np.log(np.sum(np.exp(result) * [1, -1])) - factors
 
-def log_u_p(t, tr, T, la, psi, rho, phi, c1, E1, E2_tr):
+
+def log_u_p(t, tr, T, la, psi, rho, phi, c1, c2, E_t, E_tr):
     """
     Calculates a log probability of a partner subtree evolving between the time t and T unobserved,
     provided the first partner notification happened at time tr.
@@ -56,70 +58,53 @@ def log_u_p(t, tr, T, la, psi, rho, phi, c1, E1, E2_tr):
     :param t: start time of the hidden partner subtree evolution
     :param tr: moment of the earliest partner notification
     :param T: end of sampling time
-    :param E1: E1 = c2 * exp(c1 * (t - T)), where c2 = (c1 + la - psi) / (c1 - la + psi)
-    :param E2_tr: E2_tr = c2 * exp(c1 * (tr - T))
+    :param E_t: E_t = c2 * exp(c1 * (t - T)), where c2 = (c1 + la - psi) / (c1 - la + psi)
+    :param E_tr: E_tr = c2 * exp(c1 * (tr - T))
     :return: log of the hidden partner subtree probability (approximation)
     """
[…]

### 2024-06-06 bdpn/formulas.py

@@ -16,79 +16,65 @@ def get_c2(la, psi, c1):
     return (c1 + la_minus_psi) / (c1 - la_minus_psi)
 
 
-def get_E1(c1, c2, t, T):
+def get_E(c1, c2, t, T):
     """
-    Returns E1 = c2 * exp(c1 * (t - T))
+    Returns E = c2 * exp(c1 * (t - T))
     """
     return c2 * np.exp(c1 * (t - T))
 
 
-def get_E2(c1, c2, ti, T):
-    """
-    Returns E2 = c2 * exp(c1 * (ti - T))
-    """
-    return c2 * np.exp(c1 * (ti - T))
-
-
-def get_E3(c1, t, ti):
-    """
-    Returns E3 = exp(c1 * (t - ti))
-    """
-    return np.exp(c1 * (t - ti))
-
-
-def get_u(la, psi, c1, E1):
+def get_u(la, psi, c1, E_t):
     """
     Returns U(t), where
-    dU(t)/dt = la + psi * U(t) − la U^2(t) − psi * (1 − rho)
+    dU(t)/dt = (la + psi) * U(t) − la U^2(t) − psi * (1 − rho)
     U(T) = 1
 
     :param c1: c1 = ((la − psi)^2 + 4 * la * psi * rho)^1/2
-    :param E1: E1 = c2 * exp(c1 * (t - T)), where c2 = (c1 + la - psi) / (c1 - la + psi)
+    :param E_t: E_t = c2 * exp(c1 * (t - T)), where c2 = (c1 + la - psi) / (c1 - la + psi)
     """
     two_la = 2 * la
-    return (la + psi) / two_la + c1 / two_la * ((E1 - 1) / (E1 + 1))
+    return (la + psi) / two_la + c1 / two_la * ((E_t - 1) / (E_t + 1))
 
 
-def get_log_p(c1, t, ti, E1, E2):
+def get_log_p(c1, t, ti, E_t, E_ti):
     """
     Returns log(p(t)), where
-    dp(t)/dt = la + psi * p(t) − 2 * la * p(t) * U(t)
+    dp(t)/dt = (la + psi) * p(t) − 2 * la * p(t) * U(t)
     p(ti) = 1
-    dU(t)/dt = la + psi * U(t) − la U^2(t) − psi * (1 − rho)
+    dU(t)/dt = (la + psi) * U(t) − la U^2(t) − psi * (1 − rho)
     U(T) = 1.
 
[…]

### 2024-06-06 setup.py

@@ -18,7 +18,7 @@ setup(
     author='Anna Zhukova',
     author_email='anna.zhukova@pasteur.fr',
     url='https://github.com/evolbioinfo/bdpn',
-    keywords=['phylogenetics', 'multitype birth-death model', 'partner notification'],
+    keywords=['phylogenetics', 'birth-death model', 'partner notification'],
     install_requires=['six', 'ete3', 'numpy', 'scipy', 'wquantiles'],
     entry_points={
             'console_scripts': [

### 2024-06-06 setup.py
Dependencies fixed

@@ -13,13 +13,13 @@ setup(
         'Topic :: Software Development',
         'Topic :: Software Development :: Libraries :: Python Modules',
     ],
-    version='0.1',
+    version='0.1.1',
     description='Estimation of BDPN parameters from phylogenetic trees.',
     author='Anna Zhukova',
     author_email='anna.zhukova@pasteur.fr',
     url='https://github.com/evolbioinfo/bdpn',
     keywords=['phylogenetics', 'birth-death model', 'partner notification'],
-    install_requires=['six', 'ete3', 'numpy', 'scipy', 'wquantiles'],
+    install_requires=['six', 'ete3', 'numpy', 'scipy', 'biopython'],
     entry_points={
             'console_scripts': [
                 'bdpn_infer = bdpn.bdpn_model:main',

### 2024-06-06 hiv_b_uk/py/summary_table_bd.py
UK and Zurich examples

@@ -1,0 +1,92 @@
+import os
+import re
+
+import pandas as pd
+
+from bdpn.tree_manager import read_forest
+
+
+def latexify_values(ci=True):
+    df['p'] = ' $' + df['p'].apply(str) + '$'
+    if ci:
+        df['infectious_time'] = ' $' + df['infectious_time'].apply(str) + '\;[' + df['infectious_time_min'].apply(
+            str) + '-' + df['infectious_time_max'].apply(str) + ']$'
+        df['R'] = ' $' + df['R'].apply(str) + '\;[' + df['R_min'].apply(str) + '-' + df['R_max'].apply(str) + ']$'
+        df['lambda'] = ' $' + df['lambda'].apply(str) + '\;[' + df['lambda_min'].apply(str) + '-' + df['lambda_max'].apply(
+            str) + ']$'
+        df['psi'] = ' $' + df['psi'].apply(str) + '\;[' + df['psi_min'].apply(str) + '-' + df['psi_max'].apply(
+            str) + ']$ '
+    else:
+        df['infectious_time'] = ' $' + df['infectious_time'].apply(str) + '$'
+        df['R'] = ' $' + df['R'].apply(str) + '$'
+        df['lambda'] = ' $' + df['lambda'].apply(str) + '$'
+        df['psi'] = ' $' + df['psi'].apply(str) + '$ '
+    df['repetition'] = ' $' + df['repetition'].apply(str) + '$ '
+    df['observed_trees'] = ' $' + df['observed_trees'].apply(str) + '$ '
+    df['sampled_tips'] = ' $' + df['sampled_tips'].apply(str) + '$ '
+
+
+if __name__ == "__main__":
+    import argparse
+
+    parser = argparse.ArgumentParser(
+        description="Combines the estimates from different forests/settings into one table.")
+    parser.add_argument('--forests', nargs='+', type=str, help="forests")
+    parser.add_argument('--estimates', nargs='+', type=str, help="estimated parameters")
+    parser.add_argument('--tab', type=str, help="estimate table")
+    params = parser.parse_args()
+
+    df = pd.DataFrame(columns=['repetition', 'sampled_tips', 'observed_trees', 'hidden_trees',
+                               'lambda', 'lambda_min', 'lambda_max',
+                               'psi', 'psi_min', 'psi_max',
+                               'R', 'R_min', 'R_max',
[…]

### 2024-06-06 bdpn/model_distinguisher.py
Fixed dependencies

@@ -2,9 +2,6 @@ import logging
 
 import numpy as np
 import scipy
-from matplotlib import pyplot as plt
-from matplotlib.pyplot import show
-import seaborn as sns
 
 from bdpn.tree_manager import TIME, read_forest, annotate_forest_with_time
 
@@ -120,12 +117,18 @@ def sign_test(forest):
 def cherry_diff_plot(forest, outfile=None):
     """
     Plots cherry tip time differences against cherry root times.
+    Requires matplotlib and seaborn installed.
 
     :param forest: list of trees
     :param outfile: (optional) output file where the plot should be saved.
         If not specified, the plot will be shown instead.
     :return: void
     """
+
+    from matplotlib import pyplot as plt
+    from matplotlib.pyplot import show
+    import seaborn as sns
+
     annotate_forest_with_time(forest)
 
     all_cherries = []

### 2024-06-06 setup.py

@@ -13,7 +13,7 @@ setup(
         'Topic :: Software Development',
         'Topic :: Software Development :: Libraries :: Python Modules',
     ],
-    version='0.1.1',
+    version='0.1.2',
     description='Estimation of BDPN parameters from phylogenetic trees.',
     author='Anna Zhukova',
     author_email='anna.zhukova@pasteur.fr',

### 2024-06-10 article/article_plos.tex
PLOS formatted article

--- 2024-06-06 article/article.tex
@@
-\documentclass[a4paper,10pt]{article}
-\usepackage[utf8]{inputenc}
-\usepackage{amsmath,mathrsfs}\usepackage{amsmath,amsfonts,amssymb,float,graphicx,mathtools,mathrsfs,tikz,amsthm}
+% Template for PLoS
+% Version 3.6 Aug 2022
+%
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended 
+% to minimize problems and delays during our production 
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Once your paper is accepted for publication, 
+% PLEASE REMOVE ALL TRACKED CHANGES in this file 
+% and leave only the final text of your manuscript. 
+% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
+% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
+%
+%
+% There are no restrictions on package use within the LaTeX files except that no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file. 
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig" instead of "Figure".
+% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - spacing/line breaks within cells to alter layout or alignment
+% - do not nest tabular environments (no tabular environments within tabular environments)
+% - no graphics or colored text (cell background color/shading OK)
+% See http://journals.plos.org/plosone/s/tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+%
+% For inline equations, please be sure to include all portions of an equation in the math environment.
+For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+%
+% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column. 
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
+%
+% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
+%
+% % % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
+\usepackage{mathrsfs,amsfonts,mathtools,mathrsfs,amsthm}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage[nopatch=eqnum]{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
 
 \usepackage{standalone}
-\usepackage{caption}
-\usepackage{tikz}
+\usepackage{tikz,tikzsymbols}
 \usetikzlibrary{positioning}
-\usetikzlibrary{calc}
-\usetikzlibrary{backgrounds}
-\usepackage{tikzsymbols}
-\usepackage{hieroglf}
-\usepackage{nameref}
-\usepackage{algpseudocode}
-\usepackage{hyperref}
-\usepackage{natbib}
 \usepackage{booktabs}
@@
 
-%opening
-\title{Epidemiological birth-death models with partner notification}
-\author{Anna Zhukova et al.}
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Remove comment for double spacing
+%\usepackage{setspace} 
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in 
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+%\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+%\setlength{\headheight}{27.023pt}
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\today}
+
+%% Include all macros below
+
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
+
+%% END MACROS SECTION
 
 \begin{document}
-
-\maketitle
-
-%\begin{abstract}
-%
-%\end{abstract}
-
-\section{Introduction}
+\vspace*{0.2in}
+
+% Title must be 250 characters or less.
+\begin{flushleft}
+{\Large
+\textbf\newline{Epidemiological birth-death models with partner notification} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+Anna Zhukova\textsuperscript{1,2,3*},
+Mathieu Moslonka-Lefebvre\textsuperscript{1\textcurrency},
+Olivier Gascuel\textsuperscript{1,4},
+\\
+\bigskip
+\textbf{1} Unit\'e Bioinformatique Evolutive, Institut Pasteur, Université de Paris, Paris, France
+\\
+\textbf{2} Bioinformatics and Biostatistics Hub, Institut Pasteur, Université de Paris, Paris, France
+\\
+\textbf{3} G5 Evolutionary Dynamics of Infectious Diseases, Institut Pasteur, Université de Paris, Paris, France
+\\
+\textbf{4} Institut de Syst\'{e}matique, Evolution, Biodiversit\'{e} (ISYEB) - URM 7205 CNRS, Mus\'{e}um National d'Histoire Naturelle, SU, EPHE \& UA, Paris, France
+\\
+\bigskip
+
+% Current address notes
+\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+% \textcurrency b Insert second current address 
+% \textcurrency c Insert third current address
+
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* anna.zhukova@pasteur.fr
+
+\end{flushleft}
+% Please keep the abstract below 300 words
+\section*{Abstract}
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Curabitur eget porta erat.
+Morbi consectetur est vel gravida pretium.
+Suspendisse ut dui eu ante cursus gravida non sed sem.
+Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
+Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
+Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
+Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
+Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+
+% Please keep the Author Summary between 150 and 200 words
+% Use first person. PLOS ONE authors please skip this step. 
+% Author Summary not valid for PLOS ONE submissions.   
+\section*{Author summary}
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Curabitur eget porta erat.
+Morbi consectetur est vel gravida pretium.
+Suspendisse ut dui eu ante cursus gravida non sed sem.
+Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
+Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
+Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
+Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
+Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+
+\linenumbers
+
+% Use "Eq" instead of "Equation" for equation citations.
+\section*{Introduction}
 The interaction of epidemiological and evolutionary processes leaves a footprint in pathogen genomes.
-Phylodynamics leverages this footprint to estimate epidemiological parameters~\citep{Grenfell2004a,Volz2013}.
+Phylodynamics leverages this footprint to estimate epidemiological parameters~\cite{Grenfell2004a,Volz2013}.
 It relies on models that bridge the gap between traditional epidemiology and sequence data by estimating such parameters as the basic reproduction number, $R_0$, from topology and branch lengths of pathogen phylogenies (i.e.
@@
 Classical epidemiological methods make inferences from incidence curves, which in the presence of import cases might be biased and give appearance of elevated local transmission.
-In contrast, phylodynamic approach can overcome such biases as it uses the transmission structure as one of its information sources~\citep{vaughanEstimatesEarlyOutbreakspecific2024}.
+In contrast, phylodynamic approach can overcome such biases as it uses the transmission structure as one of its information sources~\cite{vaughanEstimatesEarlyOutbreakspecific2024}.
 Therefore, using genomic data with phylodynamic estimations can provide valuable insights for preventing epidemic spreads.
 
-Phylodynamic models can be classified into two main families:  coalescent~\citep{Volz2009a,Drummond2005,Pybus2000a} and birth-death (BD)~\citep{Kendall1948,Maddison2007,Stadler2009,Stadler2010}.
-Coalescent models are often preferred for estimating deterministic population dynamics, however for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\citep{Macpherson2021}.
-In the classic BD model with incomplete sampling~\citep{Stadler2009}, births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g.
+Phylodynamic models can be classified into two main families:  coalescent~\cite{Volz2009a,Drummond2005,Pybus2000a} and birth-death (BD)~\cite{Kendall1948,Maddison2007,Stadler2009,Stadler2010}.
+Coalescent models are often preferred for estimating deterministic population dynamics, however for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\cite{Macpherson2021}.
+In the classic BD model with incomplete sampling~\cite{Stadler2009}, births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g.
 due to healing, self-isolation, starting a treatment, or death, modelled with a constant removal rate).
@@
 
-To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extention of the classical BDS model was developed by~\citet{Stadler2013a}.
+To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extention of the classical BDS model was developed by Stadler \textit{et al.}~\cite{Stadler2013a}.
 MTBD framework allows for different types of individual states (and changes between them, modelled with state change rates).
@@
 Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen (e.g.
-HIV) more than the others~\citep{Stadler2014}.
-
-To allow for heterogeneity over time, \citet{Stadler2013} developed one-state Bayesian birth-death skyline plot (BDSKY) that divides the time into intervals and allows for different piecewise constant rates on them.
-\citet{Kuhnert2016} combined the MTBD model with the BDSKY to allow for both piecewise-constant rate changes over time and multiple individual types.
+HIV) more than the others~\cite{Stadler2014}.
+
+To allow for heterogeneity over time, Stadler \textit{et al.}~\cite{Stadler2013} developed one-state Bayesian birth-death skyline plot (BDSKY) that divides the time into intervals and allows for different piecewise constant rates on them.
+Kühnert \textit{et al.}~{Kuhnert2016} combined the MTBD model with the BDSKY to allow for both piecewise-constant rate changes over time and multiple individual types.
 In particular the skyline approach can be useful to account for changes in the sampling policies (e.g.
@@
 Partner notification or contact tracing is a process in which contacts of detected infected individuals are identified and offered testing for infection.
-It has been applied to various infectious diseases, including syphilis, HIV, tuberculosis, Ebola and COVID-19, and shown to successfully reduce transmission~\citep{el-sadrContactTracingBarriers2022}.
-For instance, for HIV,  partner notification showed increased early referral and initiation of treatment, and several countries include it in their national HIV testing services policies~\citep{worldhealthorganizationCountryPolicyReview2016}.
+It has been applied to various infectious diseases, including syphilis, HIV, tuberculosis, Ebola and COVID-19, and shown to successfully reduce transmission~\cite{el-sadrContactTracingBarriers2022}.
+For instance, for HIV,  partner notification showed increased early referral and initiation of treatment, and several countries include it in their national HIV testing services policies~\cite{worldhealthorganizationCountryPolicyReview2016}.
 
@@
 
-\section{BD-PN Framework}
+\section*{Materials and methods}
 In a pathogen transmission tree $\mathscr{T}$ (approximated by a time-scaled pathogen phylogeny) the tips represent sampled pathogens %, patient state transitions occur along the branches, 
@@
 
-\begin{figure*}[tbhp]
+\begin{figure}[!h]
 \centering 
@@
 \includestandalone[width=0.6\textwidth]{Fig_tree}
-\caption{A transmission tree $\mathscr{T}$ with $n=5$ external nodes (i.e.
+\caption{\textbf{Transmission tree} $\mathscr{T}$ with $n=5$ external nodes (i.e.
 tips, which correspond to sampling events: $0000, 0001, 001, 010, 011$), $n-1=4$ internal nodes (which correspond to transmissions: $0$ (the root) and $00, 01, 000$) and $2n - 2 = 8$ branches (plus the root branch of zero length).
@@
 \label{fig:tt} 
-\end{figure*}
-
-\subsection{BD Model with incomplete sampling}
-Under the basic birth-death model~\citep{Stadler2009} all the individuals are in the same state.
+\end{figure}
+
+\subsection*{Extending Multi-Type Birth-Death models with Partner Notification}
+\subsubsection*{Birth-Death (BD) model with incomplete sampling}
+Under the basic birth-death (BD) model~\cite{Stadler2009} all the individuals are in the same state.
 They can transmit with a constant rate $\lambda$, get removed with a constant rate $\psi$, and their pathogen can be sampled upon removal with a constant probability $\rho$.
@@
 
-\subsection{MTBD models}
-MTBD models~\citep{Stadler2013a} add population structure to the BD model by allowing different individual states, transmissions between them and state changes.
+\subsubsection*{Multi-Type Birth-Death (MTBD) models}
+Multi-Type Birth-Death (MTBD) models~\cite{Stadler2013a} add population structure to the BD model by allowing different individual states, transmissions between them and state changes.
 A general MTBD model with $m$ individual states has $2m^2 + m$ parameters: An individual in state $k \in \{1, \ldots, m\}$ can be removed at a constant average rate $\psi_k$ (with pathogen sampling probability $\rho_k$), change their state to state $l \in \{1, \ldots, m\}$ at a constant average rate $\mu_{kl}$ ($k \neq l$), and transmit their pathogen to an individual in state $r \in \{1, \ldots, m\} $ at a constant average rate $\lambda_{kr}$.
@@
 
-\subsection{Partner Notification extension}
+\subsubsection*{Partner Notification (PN) extension}
 
@@
 
-These assumptions can be easily relaxed in a tree simulator (e.g., to be used with a deep learning estimator~\citep{Voznica2021}, which does not require likelihood calculation).
+These assumptions can be easily relaxed in a tree simulator (e.g., to be used with a deep learning estimator~\cite{Voznica2021}, which does not require likelihood calculation).
 
@@
 
-\subsection{MTBD-PN tree simulator}
+\subsection*{MTBD-PN tree simulator and test datasets}\label{sec:sim}
 
@@
 It is Gillespie-based, generates state change, transmission and removal times, and only reconstructs the sampled parts of the tree to save memory and increase speed.
-We give its algorithmic details in Appendix. The results in the following section show that the trees generated under MTBD and MTBD-PN models are different and easily distinguishable.
-
-\section{Testing for presence of Partner Notification}~\label{sec:test}
+We give its algorithmic details in~\nameref{S1_Appendix}. The results in the following section show that the trees generated under MTBD and MTBD-PN models are different and easily distinguishable.
+
+Using our simulator, we generated trees under three classical MTBD-family models: BD, BDEI (Birth-Death Exposed-Infectious~~\cite{Stadler2014}) and BDSS (Birth-Death with Super Spreading~\cite{Stadler2013a}), and under their -PN versions: BD-PN, BDEI-PN and BDSS-PN models.
+(See Fig.~1 in \cite{Voznica2021} for a graphical representation of BD, BDEI and DBSS models and their parameters).
+
+For each model, we generated 100 trees with 500--1\,000 tips. For each tree the parameter values were drawn randomly within the following bounds:
+\begin{itemize}
+\item $R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
+\item removal rate $\psi \in ]1 / 20, 1 / 5]$,
+\item sampling probability $\rho \in ]0.1, 0.9]$,
+\item \textit{(for -PN models only)} ratio between the partner removal rate after notification and the removal rate $\frac{\phi}{\psi} \in ]50, 500]$,
+\item \textit{(for -PN models only)} partner notification probability $\upsilon \in ]0.01/\rho, 0.9]$ (the lower bound was picked this way to ensure that the probability to be a notified partner is at least 0.01 $=\rho \upsilon$),
+\item \textit{(for BDEI models only)} becoming-infectious rate (at which an exposed individual becomes infectious) $\mu \in ]0.2, 10]$,
+\item \textit{(for BDSS models only)} fraction of superspreaders $f_{SS} = \frac{\lambda_{SS}}{\lambda_{SS} + \lambda_{SN}} \in ]0.05, 0.2]$,
+\item \textit{(for BDSS models only)} superspreading transmission ratio $x_{SS} = \frac{\lambda_{SS}}{\lambda_{NS}} = \frac{\lambda_{SN}}{\lambda_{NN}} \in ]3, 10]$.
+\end{itemize}
+
+The common parameters were consistent between models, in a sense that the $i$-th tree under the BDSS model had the same $R_e$, $\psi$ and $\rho$ values as the $i$-th trees under BDSS-PN, BD(-PN) and BDEI(-PN) models; the $i$-th tree under the BDSS-PN model had the same $\phi$ and $\upsilon$ values as the $i$-th trees under BD-PN and BDEI-PN models.
+
+\subsection*{Testing for presence of Partner Notification}~\label{sec:test}
 To access whether a given tree reveals the presence of partner notification, we developed a non-parametric test based on cherries (two-tip clades).
@@
 Being close in starting time is important as the tree is censored by the end of the sampling period: $t \leq T$.
-Hence the external branches that start close to the end of the sampling period would be on average shorter (as if they are not sampled before $T$ they will stay hidden) than those that start closer to the root~\citep{mooersBranchLengthsBirth2012}.
+Hence the external branches that start close to the end of the sampling period would be on average shorter (as if they are not sampled before $T$ they will stay hidden) than those that start closer to the root~\cite{mooersBranchLengthsBirth2012}.
 %We give control to the user on what should be considered ``some of the cherries'' and ``similar branch lengths'' for their data, via adjustable test parameter values.
@@
 \centering 
-\includegraphics[width=0.8\textwidth]{Fig_cherryswap.png}
+\includegraphics[width=0.5\textwidth]{Fig_cherryswap.png}
 \caption{Reshuffled cherry generation.
@@
 
-We implemented the PN-test in Python 3. It is available as a part of BDPN parameter estimator (see section~\ref{sec:sim}).
-
-\subsection{Performance on simulated data}~\label{sec:test:sim}
-To illustrate the PN test performance, we applied it to the trees simulated under three classical MTBD-family models: BD, BDEI (Birth-Death Exposed-Infectious~~\citep{Stadler2014}) and BDSS (Birth-Death with SuperSpreading~\citep{Stadler2013a}), and under their -PN versions: BD-PN, BDEI-PN and BDSS-PN models.
-(See Fig.~1 in \citet{Voznica2021} for a graphical representation of BD, BDEI and DBSS models and their parameters).
-
-For each model, we generated 100 trees with 500--1\,000 tips with our MTBD/MTBD-PN tree simulator. For each tree the parameter values were drawn randomly within the following bounds:
-\begin{itemize}
-\item $R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
-\item removal rate $\psi \in ]1 / 20, 1 / 5]$,
-\item sampling probability $\rho \in ]0.1, 0.9]$,
-\item \textit{(for -PN models only)} ratio between the partner removal rate after notification and the removal rate $\frac{\phi}{\psi} \in ]50, 500]$,
-\item \textit{(for -PN models only)} partner notification probability $\upsilon \in ]0.01/\rho, 0.9]$ (the lower bound was picked this way to ensure that the probability to be a notified partner is at least 0.01 $=\rho \upsilon$),
-\item \textit{(for BDEI models only)} becoming-infectious rate (at which an exposed individual becomes infectious) $\mu \in ]0.2, 10]$,
-\item \textit{(for BDSS models only)} fraction of superspreaders $f_{SS} = \frac{\lambda_{SS}}{\lambda_{SS} + \lambda_{SN}} \in ]0.05, 0.2]$,
-\item \textit{(for BDSS models only)} superspreading transmission ratio $x_{SS} = \frac{\lambda_{SS}}{\lambda_{NS}} = \frac{\lambda_{SN}}{\lambda_{NN}} \in ]3, 10]$.
-\end{itemize}
-
-The common parameters were consistent between models, in a sense that the $i$-th tree under the BDSS model had the same $R_e$, $\psi$ and $\rho$ values as the $i$-th trees under BDSS-PN, BD(-PN) and BDEI(-PN) models; the $i$-th tree under the BDSS-PN model had the same $\phi$ and $\upsilon$ values as the $i$-th trees under BD-PN and BDEI-PN models.
-
-We then run the PN tests on each tree of these 6 datasets.
+We implemented the PN-test in Python 3. It is available as a part of BDPN parameter estimator (see section~"\nameref{sec:sim}").
+
+\subsection*{Mathematical formulation of BD and BD-PN models}\label{math}
+
+\subsubsection*{Master equation and tree likelihood under BD model}
+The BD model can be described with master equations representing the likelihood densities of evolving as on the observed transmission tree, with time $t$ going forward from the root ($t=0$) till the time of the last sampled tip ($t=T$).
+These likelihood densities depend on the probability $U(t)$ of an unobserved transmission tree that started from one individual at time $t$ and evolved till $T$, without this individual nor any of their induced cases being sampled:
+
+\begin{equation}
+\scriptsize
+\begin{cases}
+\dot{U}(t) = &\Big(\lambda + \psi\Big) U(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+    &- \lambda U^2(t) \;  \textit{\color{gray} $\leftarrow$ transmission, followed by unsampled evolutions of both subtrees}\\
+    &- \psi (1 - \rho)\;  \textit{\color{gray} $\leftarrow$ removal without sampling}\\
+U(T) = & 1\;  \textit{\color{gray} $\leftarrow$ the probability to stay unsampled over time 0 is 1} \label{eq:Us}
+\end{cases}
+\end{equation}
+
+With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch $i$, starting on this branch at a time $t \leq t_i$ till time $t_i$.
+The master equations for the BD models were initially developed by Stadler \textit{et al.}~\cite{Stadler2009}, however what we present here is their branch-specific formulation that we proposed in~\cite{zhukovaFastAccurateMaximumLikelihood2022}:
+In our formulation, ${p}^{(i)}(t)$ describes an evolution along a tree branch $i$, without taking into account the branch's subtree.
+$t_i$ corresponds to the time of the equation's initial condition (typically at the end of the branch).
+% and provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$.
+These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.
+
+\begin{equation}
+\scriptsize
+\begin{cases}
+\dot{p}^{(i)}(t) = & \Big(\lambda + \psi\Big) p^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+    & - 2 \lambda p^{(i)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where one of the subtrees stayed unsampled}\\
+p^{(i)}(t_i) =  &1\;  \textit{\color{gray} $\leftarrow$ the probability of no change over zero time is 1}
+\end{cases}\label{eq:p}
+\end{equation}
+
+Using this equation, we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ as:
+
+\begin{equation}
+\scriptsize
+\begin{split}
+log L(\mathscr{T}|\Theta) =  &\sum\limits_{i \in tips}  log(\psi \rho)  \textit{\color{gray} $\leftarrow$ sampling of tips} \\
+ &+\sum\limits_{i \in \stackanchor{\tiny\textit{internal}}{\tiny\textit{nodes}}} log \big(2 \lambda p^{(i0)}(t_i)p^{(i1)}(t_i)\big)   \textit{\color{gray} $\leftarrow$ transmission events, followed by child }\\
+ & ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\textit{\color{gray} branch evolutions (each can be a donor)}  \label{eq:mtbd_tree_likelihood}
+\end{split}
+\end{equation}
+
+Importantly, the BD model is asymptotically unidentifiable (see Remark 3.4 in~\cite{Stadler2009}).
+To become identifiable it requires one of its parameters to be fixed.
+In practice, it is often the sampling probability $\rho$, as it may be approximated from epidemiological data (e.g., the proportion of sampled cases among the declared ones) or the infectious time $\frac{1}{\psi}$ (estimated from observations of infected cases).
+
+\subsubsection*{Master equation under BD-PN model}
+
+\paragraph{Tree branches under BD vs BD-PN models.}
+
+In the classical BD model with incomplete sampling, a branch of the tree corresponds to a virus evolution between the transmission corresponding to the node at the beginning of this branch and either a sampling event (if the branch is external) or a transmission corresponding to the node at the end of the branch (if it is internal).
+Note, that since sampling is incomplete, hidden transmissions (leading to fully unsampled subtrees) could occur along this branch, where the unsampled subtree could correspond to either donor or recipient.
+Hence, we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at any time $t < t_i$ along the branch.
+These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.
+$p^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along such a tree branch, integrating over all the possibilities, starting from no hidden transmission along the branch and including any number of hidden transmissions with either donor or recipient tree staying unsampled (and hence potentially different individuals at times $t$ and the branch end $t_i$).
+
+In trees generated by the BD-PN model, however, the individuals do not behave in the same way when they are notified or not and, hence, other types of branches are possible.
+As in a real tree we do not know who might have notified whom, we have to integrate over all the possibilities to calculate the tree likelihood.
+We calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}, while considering all the combinatorial possibilities for each node.
+For instance, for each tree tip we consider four possibilities: whether the corresponding individual was notified or not by someone, and on top of that, whether the tip's individual notified or not their most recent partner upon sampling.
+Before describing the combinations behind tree likelihood calculation, we will introduce additional tree branch types.
+
+\paragraph{Additional master equations under BD-PN models.}
+
+Let us consider a person A who got sampled at time $t_A$ and successfully notified their partner B, who as a consequence got sampled at time $t_B \geq t_A$.
+A and B hence correspond to tree tips, while their most recent common ancestor node AB corresponds to the transmission between A and B (see Fig.~\ref{fig:pn-branches}(1a)).
+
+\begin{figure}[h!]
+\centering 
+%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
+\includegraphics[width=1\textwidth]{Fig_PNbranches}
+\caption{\textbf{Examples of partner notification in a transmission tree}.
+Let us assume that an individual A got sampled at time $t_A$ (violet tip) and notified their most recent partner B (orange), where B is either observed (\textbf{1}) or hidden (\textbf{2-3}).
+The (observed as in \textbf{1-2} or hidden as in \textbf{3}) internal node AB corresponds to the transmission between A and B.
+The fact that A is the notifier of B implies that there is no hidden event along its branch AB-A (coloured violet), and that in any transmission on the path AB-B (coloured orange) the donor is B.
+Note, that since the most-recent-partner relationship is not always symmetric, B might be the most recent partner of more than one individual and hence might have got notified by someone else too (e.g.
+C in \textbf{b}, who in \textbf{1b,2b} was infected by B, and in \textbf{3b} could have been infected by or have infected B, who have then infected A).
+Moreover, between the moment of B's first notification ($t_A$ for \textbf{a}, $t_C$ for \textbf{b}) and the moment of sampling  of B ($t_B$) the removal rate of B becomes $\phi$ instead of $\psi$ (coloured dark red).
+Finally, B might have been removed via a standard procedure before getting notified by their partner(s) (with rate $\psi$ and a probability $\rho$, \textbf{c}).
+In the panels \textbf{3} B's subtree is fully hidden, and hence the transmission node AB is also unobserved, therefore the A's branch is composed of two parts, where the bottom part (violet) corresponds to A and contains no hidden transmissions, while the top part could correspond to A or not  (e.g in \textbf{3b} it corresponds to B) and might contain hidden transmissions.
+In the panels \textbf{2-3a-b} B is hidden as by the end of the sampling period ($T$) B is not yet sampled, while in \textbf{2-3c} B is hidden as B got removed via the standard procedure without sampling before any notification.}
+\label{fig:pn-branches} 
+\end{figure}
+
+According to our model assumption 3, only the most recent partner can get notified.
+Hence, AB must be the parent node of A, and no hidden transmission is possible along the branch AB-A (as otherwise there would have been a partner of A who is more recent than B).
+
+We will use a \textbf{notifier branch probability} $p_n^{(A)}(t)$ for such cases (when no hidden transmission occurred along the branch A, coloured violet in Fig~\ref{fig:pn-branches}):
+
+\begin{equation}
+\scriptsize
+\begin{cases}
+\dot{p}_n^{(A)}(t) &=  \Big(\lambda + \psi\Big) p_n^{(A)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+p_n^{(A)}(t_A) &= 1\label{eq:p-n}
+\end{cases}
+\end{equation}
+
+As B corresponds to A's partner, the individual on the path between the transmission node AB and the tip B cannot change (is always B) and 
+%before the notification time $t_A$ 
+only transmissions from B to someone else are possible along the path leading from AB to B.
+These transmissions can be either hidden (where the recipient tree stayed unobserved) or observed: in both cases we know which subtree corresponds to the donor (the one that contains B).
+However, depending on whether B already got notified ($t \geq t_A$ in Fig.~\ref{fig:pn-branches}(1a)) or not yet (($t < t_A$ in Fig.~\ref{fig:pn-branches}(1a))) B's removal rate is different (is correspondingly $\phi$ or $\psi$).
+We will use a \textbf{partner branch probability before notification} $p_{pb}^{(x)}(t)$ for cases when the individual at the end of the branch (or considered time interval, here $t_x=t_A$) is the same as at time $t$, has not yet been notified, but might have transmitted to someone else (unobserved) along the way (coloured orange in Fig~\ref{fig:pn-branches}):
+
+\begin{equation}
+\scriptsize
+\begin{cases}
+\dot{p}_{pb}^{(x)}(t) &=~~  \Big(\lambda + \psi\Big) p_{pb}^{(x)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+    &~~~ - \lambda p_{pb}^{(x)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission with hidden recipient subtree}\\
+p_{pb}^{(x)}(t_x) &=  1\label{eq:p-o}
+\end{cases}
+\end{equation}
+
+We will use a \textbf{partner branch probability after notification} $p_{pa}^{(i)}(t)$ for cases when the individual at the end of the branch is the same as at time $t$, by time $t$ has already been notified, and might have transmitted to someone else (unobserved) along the way (coloured dark red in Fig~\ref{fig:pn-branches}):
+
+\begin{equation}
+\scriptsize
+\begin{cases}
+\dot{p}_{pa}^{(B)}(t) &=~~  \Big(\lambda + \phi\Big) p_{pa}^{(B)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+    &~~~ - \lambda p_{pa}^{(B)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission with hidden recipient subtree}\\
+p_{pa}^{(B)}(t_B) &=  1\label{eq:p-p}
+\end{cases}
+\end{equation}
+
+Finally, once sampled, B could have notified their own most recent partner (represented by the most recent hidden subtree in Fig.~\ref{fig:pn-branches}(1a)).
+This would imply that between the moment of transmission between B and their most recent partner ($t_x$) and the moment of B's sampling ($t_B$) there was no event along the B's branch.
+Hence, if B is a notifier, the part of the B's branch after $t_x$ should be modeled with a notifier probability instead (no event along the branch).
+Note that once B is notified, the removal rate becomes $\phi$, and hence a \textbf{both notified and notifier branch probability} $p_b^{(i)}(t)$ applies:
+
+\begin{equation}
+\scriptsize
+\begin{cases}
+\dot{p}_b^{(B)}(t) &=  \Big(\lambda + \phi\Big) p_b^{(B)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
+p_b^{(B)}(t_B) &= 1\label{eq:p-b}
+\end{cases}
+\end{equation}
+
+%On the branch of B, no transmission is possible between the moment of notification $t_A$ and the moment of B's sampling $t_B$ as, according to assumption 3, once notified partners do not transmit further.
+We will use a \textbf{notified partner branch probability} $p_{pa}^{(i)}(t)$ for cases when the individual at the end of the branch is the partner who got notified at time prior to $t$ (coloured dark red in Fig~\ref{fig:pn-branches}):
+%
+%\begin{equation}
+%\scriptsize
+%\begin{cases}
+%\dot{p}_p^{(B)}(t) &=  \phi p_{pa}^{(B)}(t)\; \textit{\color{gray} $\leftarrow$ no sampling in the next infinitesimal time $\Delta t$ }\\
+%p_{pa}^{(B)}(t_B) &=  1\label{eq:p-np}
+%\end{cases}
+%\end{equation}
+
+Note that on top of the scenario we just described (shown in Fig.~\ref{fig:pn-branches}(1a)), two other scenarios are possible for a person A who once sampled (at $t_A$) notified their most recent partner B, who got sampled at time $t_B$:
+\begin{enumerate}
+
+\item[1b.] by the time $t_A$, B was already notified by someone else (at time $t_C < t_A$, see Fig.~\ref{fig:pn-branches}(1b)).
+%Note, that once notified partners do not transmit further, hence $t_C > t_{AB}$. 
+\\
+
+This would lead to a shorter partner-before-notification part on the B branch (finishing at $t_C$), and a longer notified-partner part (between $t_C$ and $t_B$).
+
+\item[1c.] by the time $t_A$, B was already sampled via a standard procedure (at time $t_B < t_A$, hence the branch B would finish before the branch A, as in Fig~\ref{fig:pn-branches}(1c)).
+
+This would lead to the partner-before-notification part on the B branch finishing at $t_B$ and no notified-partner part.
+Note that the branches between AB and B stay oriented as B remains the most recent partner of A.
+\end{enumerate}
+
+The equations~\ref{eq:Us}, \ref{eq:p}, \ref{eq:p-n}-\ref{eq:p-b} have the following closed form solutions:
+\begin{equation}
+\scriptsize
+\begin{split}
+&\begin{cases}
+U(t) = \frac{\lambda + \psi}{2\lambda} +  \frac{c_1}{2\lambda}\Big(\frac{E(t) - 1}{E(t) + 1}\Big)\\
+\\
+p^{(i)}(t) = \Big(\frac{E(t_i) + 1}{E(t) + 1}\Big)^2 e^{c_1 (t -t_i)} \\
+\\
+p_n^{(i)}(t) =  e^{(\lambda + \psi) (t -t_i)}\\
+\\
+p_b^{(i)}(t) =  e^{(\lambda + \phi) (t -t_i)}
+\\
+\\
+p_{pb}^{(i)}(t) =  \sqrt{p^{(i)}(t)p_n^{(i)}(t)} \\
+\\
+p_{pa}^{(i)}(t) =  e^{(\phi-\psi) (t -t_i)} p_{pb}^{(i)}(t) 
+\end{cases},\\
+& \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi},\\
+& ~~~~~~~~E(t) = c_2 e^{c_1 (t -T)} \\
+\end{split}\label{eq:ps}
+\end{equation}
+
+Note that if A is a notifier, A is always observed, as notification only happens upon sampling.
+However, B could have stay unsampled: either because once notified (by A or another partner, who notified them before A did) B has not yet got sampled (Fig~\ref{fig:pn-branches}(2a-b,3a-b)); or because B got removed via the standard procedure without sampling before any notification (Fig.~\ref{fig:pn-branches}(2c,3c)).
+Moreover, between the transmission AB and the moment of B's notification/removal B might have transmitted to someone else (e.g.
+C in Fig.~\ref{fig:pn-branches}(2), BC denoting the transmission).
+If that happened and someone in BC's subtree got sampled, the branch AB-A would still correspond to the notifier A, while the other AB's child branch would partially correspond to B (between AB and BC) and partially to C (starting at BC), as in Fig.~\ref{fig:pn-branches}(2).
+Otherwise (if there was no such transmission or all such subtrees stayed unobserved, as in Fig.~\ref{fig:pn-branches}(3)), the node AB will not be observed either, and A's branch will get extended with its parent's branch.
+Hence only the bottom part (after AB) of the A's branch would correspond to the notifier probability (see Fig.~\ref{fig:pn-branches}(3))).
+
+To model the cases when B is not sampled, we will need to describe the probability of a\textbf{ mixed branch} $p_{m(s1,s2)}^{(i)}(t,t_r)$ for branches that include a hidden partner subtree (AB-C branches in Fig.~\ref{fig:pn-branches}(2) and A's tip branch in Fig.~\ref{fig:pn-branches}(3)).
+$s1$ here denotes the type of the top of the branch (e.g.
+between AB and BC in Fig.~\ref{fig:pn-branches}(2), partner) and $s2$ denotes the type of the bottom of the branch (e.g.
+between AB and A in Fig.~\ref{fig:pn-branches}(3), notifier).
+Let's denote the first notification time of the hidden partner as $t_r$, and the time of the hidden partner subtree start as $t_h$, while as usual $t_i$ stands for the time at the branch end.
+For instance, in Fig~\ref{fig:pn-branches}(2a) $t_i=t_C$, $t_r=t_A$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(2b) $t_i=t_r=t_C$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(3c) $t_i=t_r=t_A$, $t_h=t_{AB}$.
+
+$p_{m(s1,s2)}^{(i)}(t,t_r)$ combines three elements: the probability of the top part of the branch $p_{top=s1}^{(h)}(t)$, which finishes with a transmission to or from the hidden partner, the probability of the bottom part of the branch $p_{bottom=s2}^{(i)}(t_h)$ (without taking into account the event at its end) and the probability of the hidden partner subtree $U_p(t_h,t_r)$.
+We describe the details behind approximation of $p_{m(s1,s2)}^{(i)}(t,t_r)$ in~\nameref{S2_Appendix}, and give the result below:
+
+\begin{equation}
+\begin{split}
+p_{m(s1,s2)}^{(i)}(t,t_r) &= p_{top=s1}^{(h)}(t, t_r)\frac{U_p(t_h,t_r)}{U(t_h)}p_{bottom=s2}^{(i)}(t_h)\textit{\color{gray},}\\
+&\textit{\color{gray}where $t_{h} \approx t + \frac{(t_i - t)}{2}$ is an approximation of the time}\\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~of the hidden partner subtree start,}\\
+&\textit{\color{gray}~~~~~~~~$t_{r}$ is the earliest notification time of the (first) partner}\\
+&U_p(t_h, t_r) \approx 
+\begin{cases}
+p_{pb}^{(r)}(t_h) \frac{1 - e^{-\psi(t_r - t_h)}}{e^{-\psi(t_r - t_h)}} (1 - \rho) + p_{pb}^{(r)}(t_h) p_{pa}^{(x)}(t_r) & \textit{\color{gray}if $t_h < t_r$}\\
+p_{pa}^{(x)}(t_h) & \textit{\color{gray}if $t_h \geq t_r$}\\
+\end{cases}\\
+&\textit{\color{gray}~~~~~~~where $t_x =T$}\\
+&\\
+& p_{top=s1}^{(h)}(t, t_r) \approx 
+\begin{cases}
+p^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}&\textit{\color{gray} if $s1=-$ (standard state)}\\
+& \\
+p_{pb}^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}&\textit{\color{gray} if $s1=p$ (partner) and}\\
+&\textit{\color{gray} ~~ $t_h \leq t_r$}\\
+p_{pb}^{(r)}(t)\Big(p_{pa}^{(h)}(t_r) -e^{-(\lambda + \phi)(t_{h} - t_r)}\Big)&\textit{\color{gray} if $s1=p$, $t \leq t_r \leq t_h$}\\
+&\\
+p_{pa}^{(h)}(t) -e^{-(\lambda + \phi)(t_{h} - t)}&\textit{\color{gray} if $s1=p$ and $t > t_r$}\\
+\end{cases}\\
+&\\
+& p_{bottom=s2}^{(i)}(t_h) = 
+\begin{cases}
+p_n^{(i)}(t_h)&\textit{\color{gray} if $i$ is a notifier tip ($s2=n$)}\\
+p^{(i)}(t_h)&\textit{\color{gray} otherwise ($s2=-$, the standard state)}\\
+\end{cases}\\
+&\\
+p_{m(p,-,n)}^{(i)}(t,t_r) &\approx p_{top=p}^{(h1)}(t, t_r)\frac{U_p(t_{h1},t_{r})}{U(t_{h1})} p_{top=-}^{(h2)}(t_{h1}, t_i)\frac{U_p(t_{h2},t_{i})}{U(t_{h2})} p_n^{(i)}(t_{h2})\textit{\color{gray},}\\
+&\textit{\color{gray}where $t_{h1} \approx t + \frac{1}{3}(t_i - t)$,}\\
+&\textit{\color{gray}~~~~~~~~$t_{h2} \approx t + \frac{2}{3}(t_i - t)$,}\\
+&\textit{\color{gray}~~~~~~~~$t_{r}$ is the earliest notification time of the partner }\\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~hidden at the top of the branch.}\\
+\end{split}\label{eq:p_mixed}
+\end{equation}
+
+In the real trees we do not know sampled individual's status (individual who did not notify, notifier, partner notified in time or not), and hence during the tree likelihood calculation, we will have to consider all these scenarios.
+
+\subsubsection*{Tree likelihood under BD-PN model}
+
+Using the formulas~(\ref{eq:ps}), (\ref{eq:p_mixed}) and (\ref{eq:p_mixed_mixed}), we can calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}.
+For each visited node $i$ we calculate the likelihood density of its subtree including the branch $ji$ connecting $i$ to its parent node $j$ for two cases depending on $i$'s receiving-a-notification status (shown with the first superscript, $^{\not{n}}$ or $^n$).
+In the first case,  $l^{(\not{n})}(i)$, the node $i$ corresponds to an unnotified individual.
+In the second case, $l^{(n)}(i, r)$, the branch $ji$ and the node $i$ correspond to an (eventually) notified partner (notified by the individual corresponding to a tip $r$ at time $t_r$).
+If $i$ is a tip, we additionally consider whether there is an observed partner in our tree, whom the individual $i$ notified  (shown with the second superscript, $^n$ or $^{\not{n}}$).
+For instance, $l^{(\not{n},n)}(i)$ represents the case where the tip $i$ was not notified but notified their most recent partner (who is observed).
+%Note that if the tip $i$ was notified by their partner $r$, we need to consider two possibilities: (1) when $t_r \leq t_i$, and hence the notification was successful at provoking $i$'s diagnostics and sampling; and (2) when $t_r > t_i$, and hence by the time of notification by $r$ $i$ already got sampled by other means.
+
+The tree $\mathscr{T}$ likelihood density $L(\mathscr{T}|\Theta)$ under BD-PN model with parameters $\Theta=\{\lambda, \psi, \rho, \phi, \upsilon\}$ can be calculated as $l^{(\not{n})}(root)$, as the root can only correspond to an unnotified individual (since there was no observed transmission before it).
+
+We describe the subtree likelihood density calculation case by case in~~\nameref{S2_Appendix}.
+
+\bigskip
+
+%The computational of $\mathscr{L}(\mathscr{T}|\Theta)$ includes $O(N^2)$ resolutions of system~(\ref{eq:p}) (for different times and initial conditions) in the worst case (caterpillar tree) and $O(NlogN)$ in the best case (for a balanced tree), where $N$ is the number of tips in $\mathscr{T}$.
+
+\subsubsection*{Extension to forests}
+
+As we previously described in~\cite{zhukovaFastAccurateMaximumLikelihood2022}, the likelihood calculation with MTBD models can be easily extended to forests.
+This extension applies also to the BD-PN model.
+Forests are useful for cases when a (sub-)epidemic started with several infected individuals (e.g.
+due to multiple pathogen introductions to a country of interest or due to a change of health policies leading to a change in parameter values).
+
+In this case the (sub-)epidemic leads to a forest $\mathscr{F}$ of $f$ observed trees: $\mathscr{T}_1, \ldots, \mathscr{T}_f$.
+The forest $\mathscr{F}$ might also include a certain number $u$ of unobserved trees, for which none of their tips got sampled.
+Forest likelihood formula hence combines the likelihoods of $f$ observed and $u$ hidden trees, and can be represented in a logarithmic form~(\ref{eq:forest_likelihood}).
+The tree likelihood formula %[\ref{eq:tree_likelihood}]
+is its special case, where $f=1$ and $u=0$.
+
+%As our application to Ebola data (see~Application) shows, $u$ is an important parameter: while the results are not very sensible to slight $u$ variations, ignoring it completely (e.g.
+using $u=0$ instead of $u \approx 500$ in our example) can change the inferences.
+
+\begin{equation}
+\begin{split}
+logL(\mathscr{F}|\Theta,u)&=u\,logU_{hidden}(\Theta) + \sum\limits_{j=1}^f logL(\mathscr{T}_j|\Theta), \\
+\textit{ where }& U_{hidden}(\Theta)=U(t_{start}),\\
+\textit{ and }& t_{start} \textit{ is the (potentially averaged) start time of the hidden trees.}
+\end{split} \label{eq:forest_likelihood} 
+\end{equation}
+
+For given model parameter values $\Theta$ we can estimate the number of hidden trees $u$ from the number of observed trees $f$ as:
+\begin{equation}
+u = f \frac{U_{hidden}(\Theta)}{1 - U_{hidden}(\Theta)}\label{eq:u} 
+\end{equation}
+
+\subsection*{BD-PN parameter and CI estimator}\label{sec:sim}
+We implemented a parameter estimator for the BD-PN model (which we called bdpn).
+It estimates the BD-PN model parameters  $\Theta = (\lambda,\psi,\phi,\rho,\upsilon) \in \mathbb{R}^5$ for a forest $\mathscr{F}$ comprising $f \geq 1$ observed trees in the maximum-likelihood framework, where one of the BD parameters ($\lambda,\psi$ or $\rho$) is fixed (for identifiability reasons).
+We also implemented the classical BD model estimator, bd.
+
+Once the optimal parameter values are found, we calculate their confidence intervals (CIs) using Wilks' method~\cite{Wilks1938}.
+For each non-fixed parameter $p \in \Theta$, we calculate its $95\%$-CI as including the values $\tilde{p}$ such that $log L(\mathscr{F}|\Theta_{opt|p=\tilde{p}}) > log L(\mathscr{F}| \Theta_{opt}) - \chi^2_1(0.95) / 2$, where $\chi^2_1(0.95)$ is the value of chi-squared distribution with 1 degree of freedom corresponding to the significance level of $0.95$ (i.e., $\sim3.84$).
+$\Theta_{opt|p=\tilde{p}}$ corresponds to the maximum-likelihood value for the other non-fixed parameters when $p = \tilde{p}$.
+
+\subsubsection*{Code availability}
+The BD-PN and BD parameter estimators are implemented in Python 3.
+They use ETE 3 framework for tree manipulation~\cite{Huerta-Cepas2016} and NumPy package for array operations~\cite{harris_array_2020}.
+
+They are available as a command-line program and a Python 3 package via PyPi (\href{https://pypi.org/project/bdpn}{bdpn}), and via Docker/Singularity (\href{https://hub.docker.com/r/evolbioinfo/bdpn/tags}{evolbioinfo/bdpn}).
+The source code and the installation and usage documentation are available on GitHub at \href{https://github.com/evolbioinfo/bdpn}{github.com/evolbioinfo/bdpn}.
+
+% Results and Discussion can be combined.
+\section*{Results}
+
+\subsubsection*{PN test is highly specific and sensitive at detecting partner notification}~\label{sec:test:sim}
+To illustrate the PN test performance, we applied it to the six 100-tree simulated datasets, corresponding to the three classical MTBD-family models: BD, BDEI (Birth-Death Exposed-Infectious~~\cite{Stadler2014}) and BDSS (Birth-Death with Super Spreading~\cite{Stadler2013a}), and to their -PN versions: BD-PN, BDEI-PN and BDSS-PN (see~"\nameref{sec:sim}" for dataset details).
+
 For 98/97/95 out of 100 trees in the BD/BDEI/BDSS dataset the PN test p-value was above $0.05$ (i.e.
@@
 
-\section{Mathematical formulation of BD and BD-PN models}\label{math}
-
-\subsection{Master equation and tree likelihood under BD model}
-The BD model can be described with master equations representing the likelihood densities of evolving as on the observed transmission tree, with time $t$ going forward from the root ($t=0$) till the time of the last sampled tip ($t=T$).
-These likelihood densities depend on the probability $U(t)$ of an unobserved transmission tree that started from one individual at time $t$ and evolved till $T$, without this individual nor any of their induced cases being sampled:
-
-\begin{equation}
-\scriptsize
-\begin{cases}
-\dot{U}(t) = &\Big(\lambda + \psi\Big) U(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-    &- \lambda U^2(t) \;  \textit{\color{gray} $\leftarrow$ transmission, followed by unsampled evolutions of both subtrees}\\
-    &- \psi (1 - \rho)\;  \textit{\color{gray} $\leftarrow$ removal without sampling}\\
-U(T) = & 1\;  \textit{\color{gray} $\leftarrow$ the probability to stay unsampled over time 0 is 1} \label{eq:Us}
-\end{cases}
-\end{equation}
-
-With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch $i$, starting on this branch at a time $t \leq t_i$ till time $t_i$.
-The master equations for the BD models were initially developed by ~\citet{Stadler2009}, however what we present here is their branch-specific formulation that we proposed in~\citep{zhukovaFastAccurateMaximumLikelihood2022}:
-In our formulation, ${p}^{(i)}(t)$ describes an evolution along a tree branch $i$, without taking into account the branch's subtree.
-$t_i$ corresponds to the time of the equation's initial condition (typically at the end of the branch).
-% and provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$.
-These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.
-
-\begin{equation}
-\scriptsize
-\begin{cases}
-\dot{p}^{(i)}(t) = & \Big(\lambda + \psi\Big) p^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-    & - 2 \lambda p^{(i)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where one of the subtrees stayed unsampled}\\
-p^{(i)}(t_i) =  &1\;  \textit{\color{gray} $\leftarrow$ the probability of no change over zero time is 1}
-\end{cases}\label{eq:p}
-\end{equation}
-
-Using this equation, we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ as:
-
-\begin{equation}
-\scriptsize
-\begin{split}
-log L(\mathscr{T}|\Theta) =  &\sum\limits_{i \in tips}  log(\psi \rho)  \textit{\color{gray} $\leftarrow$ sampling of tips} \\
- &+\sum\limits_{i \in \stackanchor{\tiny\textit{internal}}{\tiny\textit{nodes}}} log \big(2 \lambda p^{(i0)}(t_i)p^{(i1)}(t_i)\big)   \textit{\color{gray} $\leftarrow$ transmission events, followed by child }\\
- & ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\textit{\color{gray} branch evolutions (each can be a donor)}  \label{eq:mtbd_tree_likelihood}
-\end{split}
-\end{equation}
-
-Importantly, the BD model is asymptotically unidentifiable (see Remark 3.4 in~\citep{Stadler2009}).
-To become identifiable it requires one of its parameters to be fixed.
-In practice, it is often the sampling probability $\rho$, as it may be approximated from epidemiological data (e.g., the proportion of sampled cases among the declared ones) or the infectious time $\frac{1}{\psi}$ (estimated from observations of infected cases).
-
-\subsection{Master equation and tree likelihood under BD-PN model}
-
-\subsubsection{Tree branches under BD vs BD-PN models}
-
-In the classical BD model with incomplete sampling, a branch of the tree corresponds to a virus evolution between the transmission corresponding to the node at the beginning of this branch and either a sampling event (if the branch is external) or a transmission corresponding to the node at the end of the branch (if it is internal).
-Note, that since sampling is incomplete, hidden transmissions (leading to fully unsampled subtrees) could occur along this branch, where the unsampled subtree could correspond to either donor or recipient.
-Hence, we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at any time $t < t_i$ along the branch.
-These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.
-$p^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along such a tree branch, integrating over all the possibilities, starting from no hidden transmission along the branch and including any number of hidden transmissions with either donor or recipient tree staying unsampled (and hence potentially different individuals at times $t$ and the branch end $t_i$).
-
-In trees generated by the BD-PN model, however, the individuals do not behave in the same way when they are notified or not and, hence, other types of branches are possible.
-As in a real tree we do not know who might have notified whom, we will have to integrate over all the possibilities to calculate the tree likelihood.
-We will do so calculating the likelihood density of a tree using a pruning algorithm~\citep{10.1093/sysbio/22.3.240}, while considering all the combinatorial possibilities for each node.
-For instance, for each tree tip we will consider four possibilities: whether the corresponding individual was notified or not by someone, and on top of that, whether the tip's individual notified or not their most recent partner upon sampling.
-Before describing the combinations behind tree likelihood calculation, we will introduce additional tree branch types.
-
-\subsubsection{Additional master equations under BD-PN models}
-
-Let us consider a person A who got sampled at time $t_A$ and successfully notified their partner B, who as a consequence got sampled at time $t_B \geq t_A$.
-A and B hence correspond to tree tips, while their most recent common ancestor node AB corresponds to the transmission between A and B (see Fig.~\ref{fig:pn-branches}(1a)).
-
-\begin{figure*}[tbhp]
-\centering 
-%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
-\includegraphics[width=1\textwidth]{Fig_PNbranches}
-\caption{An example of partner notification in a transmission tree: an individual A got sampled at time $t_A$ (violet tip) and notified their most recent partner B (orange), where B is either observed (\textbf{1}) or hidden (\textbf{2-3}).
-The (observed as in \textbf{1-2} or hidden as in \textbf{3}) internal node AB corresponds to the transmission between A and B.
-The fact that A is the notifier of B implies that there is no hidden event along its branch AB-A (coloured violet), and that in any transmission on the path AB-B (coloured orange) the donor is B.
-Note, that since the most-recent-partner relationship is not always symmetric, B might be the most recent partner of more than one individual and hence might have got notified by someone else too (e.g.
-C in \textbf{b}, who in \textbf{1b,2b} was infected by B, and in \textbf{3b} could have been infected by or have infected B, who have then infected A).
-Moreover, between the moment of B's first notification ($t_A$ for \textbf{a}, $t_C$ for \textbf{b}) and the moment of sampling  of B ($t_B$) the removal rate of B becomes $\phi$ instead of $\psi$ (coloured dark red).
-Finally, B might have been removed via a standard procedure before getting notified by their partner(s) (with rate $\psi$ and a probability $\rho$, \textbf{c}).
-In the panels \textbf{3} B's subtree is fully hidden, and hence the transmission node AB is also unobserved, therefore the A's branch is composed of two parts, where the bottom part (violet) corresponds to A and contains no hidden transmissions, while the top part could correspond to A or not  (e.g in \textbf{3b} it corresponds to B) and might contain hidden transmissions.
-In the panels \textbf{2-3a-b} B is hidden as by the end of the sampling period ($T$) B is not yet sampled, while in \textbf{2-3c} B is hidden as B got removed via the standard procedure without sampling before any notification.}
-\label{fig:pn-branches} 
-\end{figure*}
-
-According to our model assumption 3, only the most recent partner can get notified.
-Hence, AB must be the parent node of A, and no hidden transmission is possible along the branch AB-A (as otherwise there would have been a partner of A who is more recent than B).
-
-We will use a \textbf{notifier branch probability} $p_n^{(A)}(t)$ for such cases (when no hidden transmission occurred along the branch A, coloured violet in Fig~\ref{fig:pn-branches}):
-
-\begin{equation}
-\scriptsize
-\begin{cases}
-\dot{p}_n^{(A)}(t) &=  \Big(\lambda + \psi\Big) p_n^{(A)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-p_n^{(A)}(t_A) &= 1\label{eq:p-n}
-\end{cases}
-\end{equation}
-
-As B corresponds to A's partner, the individual on the path between the transmission node AB and the tip B cannot change (is always B) and 
-%before the notification time $t_A$ 
-only transmissions from B to someone else are possible along the path leading from AB to B.
-These transmissions can be either hidden (where the recipient tree stayed unobserved) or observed: in such a case we know which subtree corresponds to the donor (the one that contains B).
-However, depending on whether B already got notified ($t \geq t_A$ in Fig.~\ref{fig:pn-branches}(1a)) or not yet (($t < t_A$ in Fig.~\ref{fig:pn-branches}(1a))) B's removal rate would be different (correspondingly $\phi$ or $\psi$).
-We will use an \textbf{partner branch probability before notification} $p_{pb}^{(x)}(t)$ for cases when the individual at the end of the branch (or considered time interval, here $t_x=t_A$) is the same as at time $t$, has not yet been notified, but might have transmitted to someone else (unobserved) along the way (coloured orange in Fig~\ref{fig:pn-branches}):
-
-\begin{equation}
-\scriptsize
-\begin{cases}
-\dot{p}_{pb}^{(x)}(t) &=~~  \Big(\lambda + \psi\Big) p_{pb}^{(x)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-    &~~~ - \lambda p_{pb}^{(x)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission with hidden recipient subtree}\\
-p_{pb}^{(x)}(t_x) &=  1\label{eq:p-o}
-\end{cases}
-\end{equation}
-
-We will use a \textbf{partner branch probability after notification} $p_{pa}^{(i)}(t)$ for cases when the individual at the end of the branch is the same as at time $t$, by time $t$ has already been notified, and might have transmitted to someone else (unobserved) along the way (coloured dark red in Fig~\ref{fig:pn-branches}):
-
-\begin{equation}
-\scriptsize
-\begin{cases}
-\dot{p}_{pa}^{(B)}(t) &=~~  \Big(\lambda + \phi\Big) p_{pa}^{(B)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-    &~~~ - \lambda p_{pa}^{(B)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission with hidden recipient subtree}\\
-p_{pa}^{(B)}(t_B) &=  1\label{eq:p-p}
-\end{cases}
-\end{equation}
-
-Finally, once sampled, B could have notified their own most recent partner (represented by the most recent hidden subtree in Fig.~\ref{fig:pn-branches}(1a)).
-This would imply that between the moment of transmission between B and their most recent partner ($t_x$) and the moment of B's sampling ($t_B$) there was no event along the B's branch.
-Hence, if B is a notifier, the part of the B's branch after $t_x$ should be modeled with a notifier probability instead (no event along the branch).
-Note that once B is notified, the removal rate becomes $\phi$, and hence a \textbf{both notified and notifier branch probability} $p_b^{(i)}(t)$ applies:
-
-\begin{equation}
-\scriptsize
-\begin{cases}
-\dot{p}_b^{(B)}(t) &=  \Big(\lambda + \phi\Big) p_b^{(B)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
-p_b^{(B)}(t_B) &= 1\label{eq:p-b}
-\end{cases}
-\end{equation}
-
-%On the branch of B, no transmission is possible between the moment of notification $t_A$ and the moment of B's sampling $t_B$ as, according to assumption 3, once notified partners do not transmit further.
-We will use a \textbf{notified partner branch probability} $p_{pa}^{(i)}(t)$ for cases when the individual at the end of the branch is the partner who got notified at time prior to $t$ (coloured dark red in Fig~\ref{fig:pn-branches}):
-%
-%\begin{equation}
-%\scriptsize
-%\begin{cases}
-%\dot{p}_p^{(B)}(t) &=  \phi p_{pa}^{(B)}(t)\; \textit{\color{gray} $\leftarrow$ no sampling in the next infinitesimal time $\Delta t$ }\\
-%p_{pa}^{(B)}(t_B) &=  1\label{eq:p-np}
-%\end{cases}
-%\end{equation}
-
-The equations~\ref{eq:Us}, \ref{eq:p}, \ref{eq:p-n}-\ref{eq:p-b} have the following closed form solutions:
-\begin{equation}
-\scriptsize
-\begin{split}
-&\begin{cases}
-U(t) = \frac{\lambda + \psi}{2\lambda} +  \frac{c_1}{2\lambda}\Big(\frac{E(t) - 1}{E(t) + 1}\Big)\\
-\\
-p^{(i)}(t) = \Big(\frac{E(t_i) + 1}{E(t) + 1}\Big)^2 e^{c_1 (t -t_i)} \\
-\\
-p_n^{(i)}(t) =  e^{(\lambda + \psi) (t -t_i)}\\
-\\
-p_b^{(i)}(t) =  e^{(\lambda + \phi) (t -t_i)}
-\\
-\\
-p_{pb}^{(i)}(t) =  \sqrt{p^{(i)}(t)p_n^{(i)}(t)} \\
-\\
-p_{pa}^{(i)}(t) =  e^{(\phi-\psi) (t -t_i)} p_{pb}^{(i)}(t) 
-\end{cases},\\
-& \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi},\\
-& ~~~~~~~~E(t) = c_2 e^{c_1 (t -T)} \\
-\end{split}\label{eq:ps}
-\end{equation}
-
-Note that on top of the scenario we just described (shown in Fig.~\ref{fig:pn-branches}(1a)), two other scenarios are possible for a person A who once sampled (at $t_A$) notified their most recent partner B, who got sampled at time $t_B$:
-\begin{enumerate}
-
-\item[1b.] by the time $t_A$, B was already notified by someone else (at time $t_C < t_A$, see Fig.~\ref{fig:pn-branches}(1b)).
-%Note, that once notified partners do not transmit further, hence $t_C > t_{AB}$. 
-\\
-
-This would lead to a shorter oriented part on the B branch (finishing at $t_C$), and a longer notified-partner part (between $t_C$ and $t_B$).
-
-\item[1c.] by the time $t_A$, B was already sampled via a standard procedure (at time $t_B < t_A$, hence the branch B would finish before the branch A, as in Fig~\ref{fig:pn-branches}(1c)).
-
-This would lead to an oriented part on the B branch finishing at $t_B$ and no notified-partner part. Note that the branches between AB and B stay oriented as B remains the most recent partner of A.
-\end{enumerate}
-
-Finally, B could also stay unsampled: either because once notified (by A or another partner, who notified them before A did) B has not yet got sampled (Fig~\ref{fig:pn-branches}(2a-b,3a-b)); or because B got removed via the standard procedure without sampling before any notification (Fig.~\ref{fig:pn-branches}(2c,3c)).
-Moreover, between the transmission AB and the moment of B's notification/removal B might have transmitted to someone else (e.g.
-C in Fig.~\ref{fig:pn-branches}(2), BC denoting the transmission).
-If that happened and someone in BC's subtree got sampled, the branch AB-A would still correspond to the notifier A, while the other AB's child branch would partially correspond to B (between AB and BC) and partially to C (starting at BC), as in Fig.~\ref{fig:pn-branches}(2).
-Otherwise (if there was no such transmission or all such subtrees stayed unobserved, as in Fig.~\ref{fig:pn-branches}(3)), the node AB will not be observed either, and A's branch will get extended with its parent's branch.
-Hence only the bottom part (after AB) of the A's branch would correspond to the notifier probability (see Fig.~\ref{fig:pn-branches}(3))).
-
-Note that if A is a notifier, A is always observed, as notification only happens upon sampling.
-
-As in the real trees the sampling is incomplete, and we do not know sampled individual's status (individual who did not notify, notifier, partner notified in time or not), during the tree likelihood calculation, we will have to consider all these scenarios.
-
-In particular, we will use the probability of a \textbf{hidden partner subtree} $U_p(t, t_A)$, which started at time $t$ on a partner branch and evolved till time $T$ without any of its individuals being sampled, provided that the first of the partner's notifiers tried to notify them at time $t_A$ (B's branch in Fig.~\ref{fig:pn-branches}(2-3)).
-This probability consists of two parts: either the partner got notified at time $t_A$ but has not yet got sampled (like in Fig.~\ref{fig:pn-branches}(2a-b,3a-b)), $U_p^{(1)}(t, t_A)$;  or the partner got removed via the standard procedure before the time $t_A$ (like in Fig.~\ref{fig:pn-branches}(2c,3c)), $U_p^{(2)}(t, t_A)$:
-\begin{equation}
-U_p(t, t_A) = U_p^{(1)}(t, t_A) + U_p^{(2)}(t, t_A).
-\end{equation}
-
-The first part can be easily expressed via the previously defined equations: partner branch evolution before notification till time $t_A$ followed by partner branch evolution after notification till time $T$.
-Note, that neither  $p_{pb}^{(x)}(t, t_A)$ not $p_{pa}^{(x)}(t)$ include an event at the end of the branch, and in this case the partner branch does not end at $t_x = T$:
-\begin{equation}
-U_p^{(1)}(t, t_A) = p_{pb}^{(A)}(t) p_{pa}^{(x)}(t_A)\textit{\color{gray}, where $t_x =T$}
-\end{equation}
-
-For the second part we would need to integrate over all possible partner removal times $t_B \in [t, t_A[$, provided that in the interval between $t$ and $t_B$ the partner might have transmitted to someone else (potentially multiple times), whose subtree(s) stayed unobserved till $T$.
-Note that $p_{pb}^{(A)}(t)$ describes a partner branch evolution between $t$ and $t_A$ with any number of hidden transmissions, no removal along it, and no event at the end of the branch.
-What we need instead is a removal with no sampling at time $\tau$.
-To approximate $U_p^{(2)}(t, t_A)$ we will divide $p_{pb}^{(A)}(t)$ by the probability of no removal during the time $(t_A - t)$ (i.e.
-$e^{-\psi(t_A - t)}$), multiply it by the probability of removal during that time ($1 - e^{-\psi(t_A - t)}$), and by the probability of not being sampled upon removal ($1 - \rho$):
-
-\begin{equation}
-U_p^{(2)}(t, t_A) \approx p_{pb}^{(A)}(t) \frac{1 - e^{-\psi(t_A - t)}}{e^{-\psi(t_A - t)}} (1 - \rho)
-\end{equation}
-
-It is an approximation as instead of integrating over all possible intervals $[t, t_B[$ for potential hidden transmissions from the partner branch, we use the interval $[t, t_A[$.
-
-We will also need to describe the probability of a\textbf{ mixed branch} $p_{m(s1,s2)}^{(i)}(t,t_r)$ for branches that include a hidden partner subtree (AB-C branches in Fig.~\ref{fig:pn-branches}(2) and A's tip branch in Fig.~\ref{fig:pn-branches}(3)).
-$s1$ here denotes the type of the top of the branch (e.g.
-between AB and BC in Fig.~\ref{fig:pn-branches}(2), partner) and $s2$ denotes the type of the bottom of the branch (e.g.
-between AB and A in Fig.~\ref{fig:pn-branches}(3), notifier).
-$t_r$ denotes the first notification time of the hidden partner.
-Moreover, let's denote the time of the hidden partner subtree start $t_h$.
-For instance, in Fig~\ref{fig:pn-branches}(2a) $t_i=t_C$, $t_r=t_A$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(2b) $t_i=t_r=t_C$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(3c) $t_i=t_r=t_A$, $t_h=t_{AB}$.
-
-$p_{m(s1,s2)}^{(i)}(t,t_r)$ combines three elements: the probability of the top part of the branch, which finishes with a transmission to or from the hidden partner, the probability of the bottom part of the branch (without taking into account the event at its end) and the probability of the hidden partner subtree:
-
-\begin{equation}
-\begin{split}
-p_{m(s1,s2)}^{(i)}(t,t_r) &= p_{top=s1}^{(h)}(t)U_p(t_h,t_r)p_{bottom=s2}^{(i)}(t_h)\textit{\color{gray},}\\
-&\textit{\color{gray}where $t_h$ is the time of the hidden partner subtree start,}\\
-&\textit{\color{gray}~~~~~~~~$t_r$ is the first notification time of the hidden partner.}\\
-\end{split}\label{eq:p_mixed}
-\end{equation}
-
-Note that since the branch contains a hidden partner, this partner must have at least one notifier (who is observed as all the notifiers).
-If the branch is external, the notifier could correspond to its tip: For example, such notifiers correspond to tip A in Fig.~\ref{fig:pn-branches}(2b) and tip C in Fig.~\ref{fig:pn-branches}(3b)).
-If the branch is external, the notifier could only correspond to a tip in its node's supertree: For example, tip A who is part of the C's supertree in Fig.~\ref{fig:pn-branches}(2a,c)).
-Having a notifier as the branch's tip implies that the bottom part of the branch corresponds to the notifier.
-Having a notifier in the supertree implies that the top part of the branch corresponds to the partner (already notified or not yet).
-Hence if the branch is internal, the top part can only correspond to the partner, while the bottom part can only correspond to a standard state (unnotified non-notifier), as the bottom part of such a branch corresponds to the partner's recipient (who could not be notified as the partner is unobserved and could not be a notifier as the branch is not external).
-Finally, if the branch is external, and the only partner's notifier corresponds to its tip, then there are two possibilities for its top part: (1a) either it fully corresponds to a standard state (unnotified non-notifier); or (1b) there is another partners' notifier in its supertree, and this other partner is also hidden along the top part of the same branch, hence the top part of the branch is partner-standard (see Fig.~\ref{fig:pn-mixed}).
-If, on the other hand, the branch is external, and the partner's notifier(s) is(are) only found in its supertree, then there are two possibilities for its bottom part: (2a) either it fully corresponds to a standard state (unnotified non-notifier); or (2b) there is another partners' notifier at its tip, and this other partner is also hidden along the bottom part of the same branch, hence the bottom part of the branch is standard-notifier (see Fig.~\ref{fig:pn-mixed}).
-Note that (1b) and (2b) represent the same case but from the point of view of two different hidden partners (B and their notifier A for the former, and D and their notifier C for the latter in Fig.~\ref{fig:pn-mixed}).
-
-\begin{figure*}[tbhp]
-\centering 
-%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
-\includegraphics[width=0.3\textwidth]{Fig_mixed}
-\caption{An example of mixed tip branch with two hidden partner subtrees: an individual C got sampled at time $t_C$ and notified their most recent partner D.
-D however, stayed unobserved as D did not get sampled by the end of the sampling period (time $T$).
-Before notification (at time $t_C$) D transmitted two times, the first transmission leading to an unobserved recipient subtree (gray dotted triangle) and the second (DB) leading to an observed recipient subtree (which includes a sampled individual A and a hidden individual B).
-The D's recipient (or someone in their subtree if there were hidden transmissions along the path DB-AB) transmitted further (transmission AB).
-Once A got sampled at time $t_A$, they notified their most recent partner B, who however was already removed by then via the standard procedure without sampling (at time $t_B < t_A$).
-Hence the branch CD-A represents a mixed branch with states partner(CD-DB)-standard(DB-AB)-notifier(AB-A).
-}
-\label{fig:pn-mixed} 
-\end{figure*}
-
-%if the branch is external the bottom part of the branch can either correspond to a notifier (as in Fig~\ref{fig:pn-branches}(2b,3)), or to a standard branch, or to a mixed tip branch starting as a standard branch and finishing as a notifier.
-It cannot correspond to another partner branch as every partner has their notifier sampled, and notifiers only notify one partner (who is already hidden along this branch).
-The bottom branch part evolution can then be expressed with the previously defined equations:
-
-We will first describe the one-hidden-partner cases ($p_{m(s1,s2)}^{(i)}(t,t_r)$, where $s1 \in \{p, -\}$, $s2 \in \{n, -\}$ and $-$ stands for the standard (unnotified non-notifier) state), and then proceed with the two-partner case ($p_{m(p,-,n)}^{(i)}(t,t_{r})$, where $t_{r}$ is the earliest notification time for the hidden partner at the top of the branch (e.g.
-$t_r=t_C$ in Fig.~\ref{fig:pn-mixed}), the only notification time for the hidden partner at the bottom of the branch being $t_i$  (e.g.
-$t_i=t_A$ in Fig.~\ref{fig:pn-mixed}).
-
-For the bottom branch part:
-
-\begin{equation}
-p_{bottom=s2}^{(i)}(t_h) = 
-\begin{cases}
-p_n^{(i)}(t_h)&\textit{\color{gray} if $i$ is a notifier tip ($s2=n$),}\\
-p^{(i)}(t_h)&\textit{\color{gray} otherwise ($s2=-$, the standard state)}\\
-\end{cases}
-\end{equation}
-
-%Finally, the top part of the branch can be in three possible states.
-The first possibility is that it corresponds to the hidden partner's individual.
-This happens if (one of) the hidden partner's notifier(s) does not correspond to the branch's end (as in Fig~\ref{fig:pn-branches}(2,3b)).
-As this notifier is observed, the transmission between them and the partner is also observed and is ancestral to the node at the end of the current branch.
-All the branches on the way connecting this notifier and the partner (including the top part of the current branch) must then correspond to the partner.
-The other possibility is that the only partner's notifier corresponds to the tip at the end of the current branch (as in Fig~\ref{fig:pn-branches}(3a,c)).
-Then the branch can either be in a standard state (unnotifier non-notifier), or a mix between another, partially hidden, partner branch and a standard state (corresponding to that other partner's recipient).
-We will approximate $p_{top=s1}^{(h)}(t)$ using the branch evolution probability of the type corresponding to $s1$  in the following way.
-For example, let us assume that $s1=-$ (standard branch).
-$p^{(h)}(t)$ expresses a probability of evolving along this branch between the times $t$ and $t_h$ with no or any number of hidden transmissions, without taking into account the event at the end of it.
-In our case there must be at least one hidden transmission: to/from the partner at time $t_h$ (where the partner tree stayed unsampled).
-If we remove the probability of no event during this time ($e^{-(\lambda + \psi)(t_h - t)}$) from $p^{(h)}(t)$, we will obtain a probability of a branch with at least one oriented hidden transmission somewhere between $t$ and $t_h$ (the probability of the corresponding hidden tree being included).
-$p_{top=-}^{(h)}(t)$ expresses almost the same thing, with the difference being that the last hidden transmission must have happened at time $t_h$ and the probability of the hidden partner tree corresponding to it is not included.
-We will hence approximate $p_{top=-}^{(h)}(t)$ as follows:
-\begin{equation}
-p_{top=-}^{(h)}(t) \approx \frac{p^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}}{U(t_{h})}
-\end{equation}
-
-We will treat the case when the branch is in the partner state ($s1=p$) in a similar way:
-\begin{equation}
-\begin{split}
-p_{top=p}^{(h)}(t, t_r) &\approx 
-\begin{cases}
-\frac{p_{pb}^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}}{U(t_{h})}&\textit{\color{gray} if $t_h \leq t_r$}\\
-p_{pb}^{(r)}(t)\frac{p_{pa}^{(h)}(t_r) -e^{-(\lambda + \phi)(t_{h} - t_r)}}{U(t_{h})}&\textit{\color{gray} if $t \leq t_r \leq t_h$}\\
-\frac{p_{pa}^{(h)}(t) -e^{-(\lambda + \phi)(t_{h} - t)}}{U(t_{h})}&\textit{\color{gray} if $t > t_r$}\\
-\end{cases}\\
-&\textit{\color{gray}where $t_r$ is the first notification time of the hidden partner.}\\
-\end{split}
-\end{equation}
-
-Note that we wrote down equation~(\ref{eq:p_mixed}) as if the start of the hidden partner tree $t_h$ was known.
-In practice, it is not the case as we need to integrate over all the possible times between the branch start $t_{parent(i)}$ and its end $t_i)$.
-To simplify the calculations, we will take the time in the middle of this interval instead:
-
-\begin{equation}
-t_h \approx t + \frac{t_i - t}{2}
-\end{equation}
-
-Finally, we will treat the mixed case with two hidden partners using similar principals:
-\begin{equation}
-\begin{split}
-p_{m(p,-,n)}^{(i)}(t,t_r) &\approx p_{top=p}^{(h1)}(t, t_r)U_p(t_{h1},t_{r}) p_{top=-}^{(h2)}(t_{h1}, t_i)U_p(t_{h2},t_{i}) p_n^{(i)}(t_{h2})\textit{\color{gray},}\\
-&\textit{\color{gray}where $t_{h1} \approx t + \frac{1}{3}(t_i - t)$,}\\
-&\textit{\color{gray}~~~~~~~~$t_{h2} \approx t + \frac{2}{3}(t_i - t)$,}\\
-&\textit{\color{gray}~~~~~~~~$t_{r}$ is the earliest notification time of the partner }\\
-&\textit{\color{gray}~~~~~~~~~~~~~~~~~~hidden at the top of the branch.}\\
-\end{split}\label{eq:p_mixed_mixed}
-\end{equation}
-
-\subsection{BD-PN tree likelihood density}
-
-Using the formulas~(\ref{eq:ps}), (\ref{eq:p_mixed}) and (\ref{eq:p_mixed_mixed}), we can calculate the likelihood density of a tree using a pruning algorithm~\citep{10.1093/sysbio/22.3.240}.
-For each visited node $i$ we calculate the likelihood density of its subtree including the branch $ji$ connecting $i$ to its parent node $j$ for two cases depending on $i$'s receiving-a-notification status (shown with the first superscript, $^{\not{n}}$ or $^n$).
-In the first case,  $l^{(\not{n})}(i)$, the node $i$ corresponds to an unnotified individual.
-In the second case, $l^{(n)}(i, r)$, the branch $ji$ and the node $i$ correspond to an (eventually) notified partner (notified by the individual corresponding to a tip $r$ at time $t_r$).
-If $i$ is a tip, we additionally consider whether there is an observed partner in our tree, whom the individual $i$ notified  (shown with the second superscript, $^n$ or $^{\not{n}}$).
-For instance, $l^{(\not{n},n)}(i)$ represents the case where the tip $i$ was not notified but notified their most recent partner (who is observed).
-%Note that if the tip $i$ was notified by their partner $r$, we need to consider two possibilities: (1) when $t_r \leq t_i$, and hence the notification was successful at provoking $i$'s diagnostics and sampling; and (2) when $t_r > t_i$, and hence by the time of notification by $r$ $i$ already got sampled by other means.
-
-The tree $\mathscr{T}$ likelihood density $\mathscr{L}(\mathscr{T}|\Theta)$ under BD-PN model with parameters $\Theta=\{\lambda, \psi, \rho, \phi, \upsilon\}$ can be calculated as $l^{(\not{n})}(root)$, as the root can only correspond to an unnotified individual (since there was no observed transmission before it).
-
-We will now describe the subtree likelihood density calculation case by case.
-
-\subsubsection*{Case 1: $i$ is a tip}
-
-For a tip $i$ with a parent node $j$ the subtree likelihood density values can be calculated as:
-\begin{equation}
-\scriptsize
-\begin{array}{ll}
-l^{(\not{n},\not{n})}(i) &= p^{(i)}(t_j)\psi\rho (1-\upsilon) \textit{\color{gray} ~i.e. $i$ was not notified and did not notify}\\
-& \textit{\color{gray} (standard branch evolution followed by sampling and no notification)}\\
-&~~+p_{m(-,n)}^{(i)}(t_j,t_i)\psi\rho\upsilon \textit{\color{gray} ~i.e. $i$ was not notified but notified their partner who is hidden}\\
-& \textit{\color{gray} (mixed standard-notifier branch evolution followed by sampling and notification)}\\
-\\
-l^{(\not{n},n)}(i) &= p_n^{(i)}(t_j)\psi\rho
-\upsilon \textit{\color{gray} ~i.e. $i$ was not notified but notified their most recent patner (observed)}\\
-& \textit{\color{gray} (notifier branch evolution followed by sampling and notification)}\\
-&~~\\
-\\
-l^{(n,\not{n})}(i,r) &= \begin{cases}
-%0 \textit{\color{gray} ~if $t_r < t_j$}\\
-%\textit{\color{gray} (not possible as once notified partners do not transmit)}
-%\\
-%\\
-p_{pa}^{(i)}(t_j)\phi(1 - \upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_r < t_j$ and did not notify}\\
-\textit{\color{gray} (notified partner branch evolution followed by notification-provoked sampling and no notification)}
-\\
-~~~+p_{m(pa,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r < t_j$, }\\
-\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$  did not notify}\\
- \textit{\color{gray} ~~~(mixed partner-standard branch evolution followed by sampling and no notification)}\\
- ~~~+p_{m(pa,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
- \textit{\color{gray} ~~~(mixed partner-notifier branch evolution followed by sampling and notification)}
-\\
-~~~+p_{m(pa,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r < t_j$, }\\
-\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
- \textit{\color{gray} ~~~(mixed partner-standard-notifier branch evolution followed by sampling and notification)}
-\\
-\\
-p_{pb}^{(i)}(t_j)\psi\rho(1 - \upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and did not notify}\\
-\textit{\color{gray} (partner-before-notification branch evolution followed by standard sampling and no notification)}
-\\
-~~~+p_{m(pb,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r > t_i$, }\\
-\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$  did not notify}\\
- \textit{\color{gray} ~~~(mixed partner-standard branch evolution followed by sampling and no notification)}\\
- ~~~+p_{m(pb,n)}^{(i)}(t_j,t_i)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
- \textit{\color{gray} ~~~(mixed partner-notifier branch evolution followed by sampling and notification)}
-\\
-~~~+p_{m(pb,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r > t_i$, }\\
-\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
- \textit{\color{gray} ~~~(mixed partner-standard-notifier branch evolution followed by sampling and notification)}
-\\
-\\
-p_{pb}^{(r)}(t_j) p_{pa}^{(i)}(t_r) \phi (1-\upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_j \leq t_r \leq t_i$ and did not notify}\\
-\textit{\color{gray} (partner-before-notification branch evolution followed by} \\
-\textit{\color{gray}~notified partner branch evolution, notification-provoked sampling and no notification)}
-\\
-~~~+p_{m(p,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_j \leq t_r \leq t_i$, }\\
-\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$  did not notify}\\
- \textit{\color{gray} ~~~(mixed partner-standard branch evolution followed by sampling and no notification)}\\
- ~~~+p_{m(p,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
- \textit{\color{gray} ~~~(mixed partner-notifier branch evolution followed by sampling and notification)}
-\\
-~~~+p_{m(p,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_j \leq t_r \leq t_i$, }\\
-\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
- \textit{\color{gray} ~~~(mixed partner-standard-notifier branch evolution followed by sampling and notification)}
-\end{cases}\\
-\\
-l^{(n,n)}(i,r) &= \begin{cases}
-p_b^{(i)}(t_j)\phi\upsilon \textit{\color{gray} ~if $i$ got notified at time $t_r < t_j$ and notified their most recent patner}\\
-\textit{\color{gray} (notified notifier branch evolution followed by notification-provoked sampling and  notification)}\\
-\\
-p_n^{(i)}(t_j)\psi\rho\upsilon \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and notified their most recent patner}\\
-\textit{\color{gray} (notifier branch evolution followed by standard sampling and  notification)}\\
-\\
-p_n^{(r)}(t_j) p_b^{(i)}(t_r) \phi \upsilon \textit{\color{gray} ~if $i$ got notified at time $t_j \leq t_r \leq t_i$ and notified}\\
-\textit{\color{gray} (notifier branch evolution followed by notified notifier branch evolution,} \\
-\textit{\color{gray}~notification-provoked sampling and notification)}\\
-\end{cases}\\
-\end{array}
-\label{eq:tip}
-\end{equation}
-
-\subsubsection*{Case 2: $i$ is an internal node} 
-Let us now consider an internal node $i$ with two child nodes, $i_0$ and $i_1$.
-
-\subsubsection*{Case 2.1: $i$ is an unnotified internal node} 
-We will start with the configuration where the $i$'s branch corresponds to an unnotified individual.
-Hence the $i$'s branch evolution is represented by $p^{(i)}(t_j)$ with a transmission event at the end of it, where each of the child branches can correspond to the donor (probability density of $2\lambda$).
-We will consider three possibilities: (2.1.1) when both $i_0$ and $i_1$ are internal nodes; (2.1.2) when both $i_0$ and $i_1$ are tips; and (2.1.3) when one of them is an internal node and the other one is a tip.
-
-%Note that a notification will only be possible if the notifying branch is external as the notifier only notifies the most recent partner, i.e.
-the transmission corresponds to the notifier's parent node.
-
-\subsubsection*{Case 2.1.1: $i$ is an unnotified internal node with internal-node children $i0$ and $i1$}
-
-First, let's assume that $i_0$ and $i_1$ are internal. Then none of them can be a notifier (since notifiers correspond to tips), and hence none of them is notified (as $i$ is also unnotified):
-
-\begin{equation}
-\scriptsize
-l^{(\not{n})}(i) = p^{(i)}(t_j) 2\lambda l^{(\not{n})}(i_0)l^{(\not{n})}(i_1) \label{eq:lu-int-int}
-\end{equation}
-
-\subsubsection*{Case 2.1.2: $i$ is an unnotified internal node with tip children $i0$ and $i1$}
-
-Secondly, let's assume that $i_0$ and $i_1$ are tips. Then each of them could have notified the other one:
-
-\begin{equation}
-\scriptsize
-\begin{split}
-l^{(\not{n})}(i) = p^{(i)}(t_j) 2\lambda \cdot 
-\Big(&~l^{(\not{n},\not{n})}(i_0)l^{(\not{n},\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ neigher $i_0$ nor $i_1$ notified} \\
-&+ l^{(\not{n},n)}(i_0)l^{(n,\not{n})}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $i_0$ notified $i_1$ and $i_1$ did not notify} \\
-&+ l^{(n,\not{n})}(i_0,i_1)l^{(\not{n},n)}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ did not notify, but $i_1$ notified $i_0$} \\
-&+ l^{(n,n)}(i_0,i1)l^{(n,n)}(i_1,i_0)~\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ and $i_0$ notified each other} \\
-\end{split}\label{eq:lu-tip-tip}
-\end{equation}
-
-\subsubsection*{Case 2.1.3: $i$ is an unnotified internal node whose children are a tip and an internal node}
-Lastly, let's assume that $i_0$ is a tip and $i_1$ is internal (the scenario where $i_1$ is a tip and $i_0$ is internal can be obtained by swapping the labels). Then $i_0$ could have notified $i_1$:
-
-\begin{equation}
-\scriptsize
-\begin{split}
-l^{(\not{n})}(i) = p^{(i)}(t_j) 2\lambda \cdot 
-\Big(&~l^{(\not{n},\not{n})}(i_0)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ did not notify $i_1$ } \\
-&+ l^{(\not{n},n)}(i_0)l^{(n)}(i_1,i_0)~\Big) \textit{\color{gray}~$\leftarrow$ $i_0$ notified $i_1$} \\
-\end{split}\label{eq:lu-int-tip}
-\end{equation}
-
-\subsubsection*{Case 2.2: $i$'s branch is internal and (at least initially) corresponds to an (eventually notified) partner}
-
-In the other case the internal node $i$'s branch starts as a partner who will eventually get notified by a tip $r$.
-There are two possibilities: either the node $i$ corresponds to the partner or the partner is hidden along the $i$'s branch and $i$ corresponds to someone within $i$'s recipient subtree (e.g., if $i=C$ in Fig.~\ref{fig:pn-branches}(2a)).
-In the first case, the $i$'s branch evolution is represented by the partner branch evolution $p_{p}^{(i)}(t_j, t_r)$:
-
-\begin{equation}
-\scriptsize
-p_{p}^{(i)}(t_j, t_r) =
-\begin{cases}
-p_{pa}^{(i)}(t_j) &\textit{\color{gray} if $t_r < t_j$,}\\
-p_{pb}^{(i)}(t_j) &\textit{\color{gray} if $t_r > t_i$,}\\
-p_{pb}^{(r)}(t_j)p_{pa}^{(i)}(t_r) &\textit{\color{gray} if $t_j \leq t_r \leq t_i$.}\\
-\end{cases}\label{eq:p-p}
-\end{equation}
-This branch has a transmission event at the end of it, where the donor is known (probability density of $\lambda$): The donor corresponds to the subtree containing the tip representing the sampling of the individual $i$.
-
-In the second case, the $i$'s branch evolution is represented by $p_{m(p,-)}^{(i)}(t_j,t_r)$ with a transmission event at the end of it, where the donor is unknown (probability density of $2\lambda$): any of the subtrees can be the donor.
- %Moreover, this recipient's branch cannot be notified by our partner (as in our model notified partners do not notify further):
-
-Again, we will consider three possibilities for $i$'s child nodes: (2.2.1) when both $i_0$ and $i_1$ are internal nodes; (2.2.2) when both $i_0$ and $i_1$ are tips; and (2.2.3) when one of them is an internal node and the other one is a tip.
-
-%In any of these scenarios, the notification must have happened after the partner tip branch start (as one of the model assumptions is that partners do not transmit after notification). Hence:
-%\begin{equation}
-%\scriptsize
-%l^{(n)}(i, r) = 0 \textit{\color{gray}~if $t_i > t_r$} \\
-%\label{eq:lp-zero}
-%\end{equation}
-
-\subsubsection*{Case 2.2.1: $i$'s branch is internal and (at least initially) corresponds to a partner with internal-node children $i0$ and $i1$}
-
-First, let's assume that $i_0$ and $i_1$ are internal.
-Then none of them can be a notifier (since notifiers correspond to tips), and hence either exactly one of them is notified by $r$, or the partner is hidden:
-
-\begin{equation}
-\scriptsize
-\begin{split}
-l^{(n)}(i, r) = &p_{p}^{(i)}(t_j, t_r) \lambda \cdot
-\Big(~l^{(n)}(i_0,r)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ is notified by $r$} \\
-& ~~~~~~~~~~~~~~~~~~~+ l^{(\not{n})}(i_0)l^{(n)}(i_1,r)~\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ is notified by $r$} \\
-& + p_{m(p,-)}^{(i)}(t_j,t_r) 2 \lambda l^{(\not{n})}(i_0)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ hidden partner}\\
- \end{split}
-\label{eq:lp-int-int}
-\end{equation}
-
-\subsubsection*{Case 2.2.2: $i$'s branch is internal and (at least initially) corresponds to a partner with tip children $i0$ and $i1$} 
-Secondly, let's assume that $i_0$ and $i_1$ are tips. Then each of them could have notified the other one (in addition to one of them potentially being notified by $r$):
-
-\begin{equation}
-\scriptsize
-\begin{split}
-l^{(n)}(i, r) = &p_{p}^{(i)}(t_j,t_r) \lambda \cdot
-\Big(~l^{(n,\not{n})}(i_0,first(r,i_1))l^{(\not{n},n)}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ and $i_1$ notified $i_0$, $i_0$ did not notify} \\
-&~~~~~~~~~~~~~~~~~~~+ l^{(n,n)}(i_0,first(r,i_1))l^{(n,n)}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $r$ and $i_1$ notified $i_0$, $i_0$ notified $i_1$} \\
-&~~~~~~~~~~~~~~~~~~~+l^{(\not{n},n)}(i_0)l^{(n,\not{n})}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ $r$ and $i_0$ notified $i_1$, $i_1$ did not notify} \\
-&~~~~~~~~~~~~~~~~~~~+ l^{(n,n)}(i_0, i_1)l^{(n,n)}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ $r$ and $i_0$ notified $i_1$, $i_1$ notified $i_0$} \\
-&~~~~~~~~~~~~~~~~~~~+l^{(n,n)}(i_0,r)l^{(n,\not{n})}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$ notified $i_1$, $i_1$ did not notify} \\
-&~~~~~~~~~~~~~~~~~~~+ l^{(n,\not{n})}(i_0, i_1)l^{(n,n)}(i_1,r) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_1$ notified $i_0$, $i_0$ did not notify} \\
-&~~~~~~~~~~~~~~~~~~~+l^{(n,\not{n})}(i_0,r)l^{(\not{n},\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$ and $i_1$ did not notify} \\
-&~~~~~~~~~~~~~~~~~~~+ l^{(\not{n},\not{n})}(i_0)l^{(n,\not{n})}(i_1,r)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_0$ and $i_1$ did not notify} 
-\\
-&+ p_{m(p,-)}^{(i)}(t_j,t_r) 2 \lambda \cdot
-\Big(~l^{(n,\not{n})}(i_0,i_1)l^{(\not{n},n)}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
-&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i_1$ notified $i_0$, $i_0$ did not notify} \\
-&~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~+ l^{(n,n)}(i_0,i_1)l^{(n,n)}(i_1, i_0)\textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
-&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i_1$ notified $i_0$, $i_0$ notified $i_1$} \\
-&~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~+l^{(\not{n},n)}(i_0)l^{(n,\not{n})}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
-&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i_0$ notified $i_1$, $i_1$ did not notify} \\
-&~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~+l^{(\not{n},\not{n})}(i_0)l^{(\not{n},\not{n})}(i_1)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
-&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i_0$ and $i_1$ did not notify} \\
-\\
-&\textit{\color{gray}~where } \color{gray}first(a,b) = 
-\begin{cases}
-a \textit{~if $t_a \leq t_b$}\\
-b \textit{~if $t_a > t_b$.}
-\end{cases} 
- \end{split}
-\label{eq:lu-tip-tip}
-\end{equation}
-
-\subsubsection*{Case 2.2.3: $i$'s branch is internal and (at least initially) corresponds to a partner, $i$'s children are a tip and an internal node}
-Lastly, let's assume that $i_0$ is a tip and $i_1$ is internal (the scenario where $i_1$ is a tip and $i_0$ is internal can be obtained by swapping the labels).
-Then $i_0$ could have notified $i_1$ (in addition to one of them being potentially notified by $r$):
-
-\begin{equation}
-\scriptsize
-\begin{split}
-l^{(n)}(i, r) = &p_{p}^{(i)}(t_j,t_r) \lambda \cdot
-\Big(~l^{(n,n)}(i_0,r)l^{(n)}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$  notified $i_1$} \\
-&~~~~~~~~~~~~~~~~~~~+ l^{(n,\not{n})}(i_0,r)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$  $r$ notified $i_0$, $i_0$ did not notify} \\
-&~~~~~~~~~~~~~~~~~~~+l^{(\not{n}, n)}(i_0)l^{(n)}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ both $r$ and $i_0$ notified $i_1$} \\
-&~~~~~~~~~~~~~~~~~~~+ l^{(\not{n},\not{n})}(i_0)l^{(n)}(i_1,r)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_0$ did not notify} 
-\\
-&+ p_{m(p,-)}^{(i)}(t_j,t_r) 2\lambda \cdot
-\Big(~l^{(\not{n},n)}(i_0)l^{(n)}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
-&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i_0$ notified $i_1$} \\
-&~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~+ l^{(\not{n},\not{n})}(i_0)l^{(\not{n})}(i_1)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
-&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ $i_0$ did not notify}\\
-\textit{\color{gray}~where } &\color{gray}first(a,b) = 
-\begin{cases}
-a \textit{~if $t_a \leq t_b$}\\
-b \textit{~if $t_a > t_b$.}
-\end{cases} 
- \end{split}
-\label{eq:lp-tip-int}
-\end{equation}
-
-\bigskip 
-Note that $l^{(n)}(i, r)$, $l^{(n,n)}(i, r)$ and $l^{(n,\not{n})}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time $t_r$ is known, i.e.
-once we have climbed to an internal node corresponding to the transmission between the notifier $r$ (corresponding to one of its children, must be a tip) and the partner (the other child).
-The calculation is done recursively by descending into the subtree.
-The tree likelihood calculation hence can take $O(n^2)$ steps in the worst case (ladder tree), where $n$ is the number of tips in the tree.
-However, as all the unotified equation parts ($l^{(\not{n})}(i)$, $l^{(\not{n},n)}(i)$ and $l^{(\not{n},\not{n})}(i)$) only depend on the corresponding node $i$, they can be pre-calculated and reused to speed up the recursive steps.
-
-\bigskip
-
-%The computational of $\mathscr{L}(\mathscr{T}|\Theta)$ includes $O(N^2)$ resolutions of system~(\ref{eq:p}) (for different times and initial conditions) in the worst case (caterpillar tree) and $O(NlogN)$ in the best case (for a balanced tree), where $N$ is the number of tips in $\mathscr{T}$.
-
-\subsection{Extension to forests}
-
-As we previously described in~\citep{zhukovaFastAccurateMaximumLikelihood2022}, the likelihood calculation with MTBD models can be easily extended to forests.
-This extension applies also to the BD-PN model.
-Forests are useful for cases when a (sub-)epidemic started with several infected individuals (e.g.
-due to multiple pathogen introductions to a country of interest or due to a change of health policies leading to a change in parameter values).
-
-In this case the (sub-)epidemic leads to a forest $\mathscr{F}$ of $f$ observed trees: $\mathscr{T}_1, \ldots, \mathscr{T}_f$.
-The forest $\mathscr{F}$ might also include a certain number $u$ of unobserved trees, for which none of their tips got sampled.
-Forest likelihood formula hence combines the likelihoods of $f$ observed and $u$ hidden trees, and can be represented in a logarithmic form~(\ref{eq:forest_likelihood}).
-The tree likelihood formula %[\ref{eq:tree_likelihood}]
-is its special case, where $f=1$ and $u=0$.
-
-%As our application to Ebola data (see~Application) shows, $u$ is an important parameter: while the results are not very sensible to slight $u$ variations, ignoring it completely (e.g.
-using $u=0$ instead of $u \approx 500$ in our example) can change the inferences.
-
-\begin{equation}
-\begin{split}
-logL(\mathscr{F}|\Theta,u)&=u\,logU_{hidden}(\Theta) + \sum\limits_{j=1}^f logL(\mathscr{T}_j|\Theta), \\
-\textit{ where }& U_{hidden}(\Theta)=\sum\limits_{s=1}^{d}\pi_s U_s(t_{start}),\\
-\textit{ and }& t_{start} \textit{ is the (potentially averaged) start time of the hidden trees.}
-\end{split} \label{eq:forest_likelihood} 
-\end{equation}
-
-For given model parameter values $\Theta$ we can estimate the number of hidden trees $u$ from the number of observed trees $f$ as:
-\begin{equation}
-u = f \frac{U_{hidden}(\Theta)}{1 - U_{hidden}(\Theta)}\label{eq:u} 
-\end{equation}
-
-\section{BD-PN parameter and CI estimator}\label{sec:sim}
-We implemented a parameter estimator for the BD-PN model (which we called bdpn).
-It estimates the BD-PN model parameters  $\Theta = (\lambda,\psi,\phi,\rho,\upsilon) \in \mathbb{R}^5$ for a forest $\mathscr{F}$ comprising $f \geq 1$ observed trees in the maximum-likelihood framework, where one of the BD parameters ($\lambda,\psi$ or $\rho$) is fixed (for identifiability reasons).
-We also implemented the classical BD model estimator, bd.
-
-Once the optimal parameter values are found, we calculate their confidence intervals (CIs) using Wilks' method~\citep{Wilks1938}.
-For each non-fixed parameter $p \in \Theta$, we calculate its $95\%$-CI as including the values $\tilde{p}$ such that $log L(\mathscr{F}|\Theta_{opt|p=\tilde{p}}) > log L(\mathscr{F}| \Theta_{opt}) - \chi^2_1(0.95) / 2$, where $\chi^2_1(0.95)$ is the value of chi-squared distribution with 1 degree of freedom corresponding to the significance level of $0.95$ (i.e., $\sim3.84$).
-$\Theta_{opt|p=\tilde{p}}$ corresponds to the maximum-likelihood value for the other non-fixed parameters when $p = \tilde{p}$.
-
-\subsection*{Code availability}
-The BD-PN and BD parameter estimators are implemented in Python 3.
-They use ETE 3 framework for tree manipulation~\citep{Huerta-Cepas2016} and NumPy package for array operations~\citep{harris_array_2020}.
-
-They are available as a command-line program and a Python 3 package via PyPi (\href{https://pypi.org/project/bdpn}{bdpn}), and via Docker/Singularity (\href{https://hub.docker.com/r/evolbioinfo/bdpn/tags}{evolbioinfo/bdpn}).
-The source code and the installation and usage documentation are available on GitHub at \href{https://github.com/evolbioinfo/bdpn}{github.com/evolbioinfo/bdpn}.
-
-\subsection{Performance on simulated data}
-
-To assess the performance of our maximum-likelihood estimator for the BD-PN model, we used the 100 trees with 500--1\,000 tips simulated under the BD-PN model as described in subsection~\ref{sec:test:sim}.
+\subsubsection*{Accounting for partner notification is crucial for accurate parameter estimation}
+
+We assessed the performance of our maximum-likelihood estimator for the BD-PN model on the 100 trees from the the BD-PN simulated data set (see~"\nameref{sec:sim}" for dataset details).
 
@@
 The results with the becoming-uninfectious rate $\psi$ fixed are similar and given in the Supplementary Material.
-
 As expected with a maximum likelihood estimator, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the real parameter values in both settings.
 
-We calculated the relative error (normalized distance between the estimated and the target values: $\frac{|estimated - target|}{target}$) and the relative bias ($\frac{estimated - target}{target}$) for each parameter on each tree.
+To highlight the importance of accounting for partner notification with the model, we also applied the classical BD model estimator bd to these trees.
+Additionally, we applied both BD-PN and BD model parameter estimators to the dataset simulated under BD model.
+The results are shown in  Fig.~\ref{fig:sim}.
+%We calculated the relative error (normalized distance between the estimated and the target values: $\frac{|estimated - target|}{target}$) and the relative bias ($\frac{estimated - target}{target}$) for each parameter on each tree.
 For $\rho$ and $\upsilon$ absolute values were calculated instead of normalized ones.
-
-To highlight the importance of accounting for partner notification with the model, we also applied the classical BD model estimator bd to these trees.
-Additionally, we applied both BD-PN and BD model parameter estimators to a set of 100 trees with 500--1\,000 tips simulated under the BD model (see subsection~\ref{sec:test:sim} for the data set details).
-The results are shown in  Fig.~\ref{fig:sim}.
 We compared the method accuracy by a paired z-test.
@@
 
-\begin{figure}[!pht]
+\begin{figure}[h!]
 \centering 
 \includegraphics[width=1\textwidth]{Fig_errors_p.png}
-\caption{Comparison of inference accuracy of bdpn (yellow) and bd (blue) estimators on a data set of 100 trees of 500--1\,000 tips generated under BD-PN model (top) and a data set of 100 trees of 500--1\,000 tips generated under BD model (bottom), with $\rho$ parameter being fixed to its real value.
+\caption{\textbf{Comparison of inference accuracy of bdpn (yellow) and bd (blue) estimators} on a data set of 100 trees of 500--1\,000 tips generated under BD-PN model (top) and a data set of 100 trees of 500--1\,000 tips generated under BD model (bottom), with $\rho$ parameter being fixed to its real value.
 We show the swarmplots (coloured by estimator) of relative errors for each test tree and parameter, which are measured as the normalized distance between the estimate and the real value.
@@
 
-\subsection{Application: HIV-B epidemic in the UK}
+\subsubsection*{Application 1: HIV-1 B epidemic in the UK}
 We applied our estimator to asses the partner notification in HIV-infected patients in the UK.
-We used the phylogenetic tree from our recent study of HIV drug resistance in the UK~\citep{zhukovaModelingDrugResistance2023}.
+We used the phylogenetic tree from our recent study of HIV drug resistance in the UK~\cite{zhukovaModelingDrugResistance2023}.
 The tree tips represent viruses sampled from 40\,055 individuals between 1996 and 2016.
-The samples used for the tree reconstruction were obtained from the UK HIV Drug resistance database~\citep{Dunn2007}.
+The samples used for the tree reconstruction were obtained from the UK HIV Drug resistance database~\cite{Dunn2007}.
 The sampling date metadata for these samples were month-specific (e.g.
 Aug 2015), hence potentially creating simultaneous sampling dates for cases, which were sampled on different days of the month.
-To avoid this issue we randomly selected a date within the specified month for each sample, and time-scaled the tree with LSD2~\citep{To2016} (v2.4.1, strict molecular clock with outlier removal).
+To avoid this issue we randomly selected a date within the specified month for each sample, and time-scaled the tree with LSD2~\cite{To2016} (v2.4.1, strict molecular clock with outlier removal).
 We repeated this time-scaling procedure 10 times, obtaining 10 trees with slightly different branch lengths and root dates between Oct 5 and Nov 11, 1962.
 
-Between 2012 and 2015, in the UK antiretroviral treatment (ART) of HIV-infected individuals  was initiated when their CD4 count dropped below 350 cells/mL~\citep{williamsBritishHIVAssociation2012}.
-However ``if a patient with a CD4 cell count > 350 cells/mL wishes to start ART to reduce the risk of transmission to partners, this decision is respected and ART is started''~\citep{williamsBritishHIVAssociation2012}.
+Between 2012 and 2015, in the UK antiretroviral treatment (ART) of HIV-infected individuals  was initiated when their CD4 count dropped below 350 cells/mL~\cite{williamsBritishHIVAssociation2012}.
+However ``if a patient with a CD4 cell count $>$ 350 cells/mL wishes to start ART to reduce the risk of transmission to partners, this decision is respected and ART is started''~\cite{williamsBritishHIVAssociation2012}.
 Before 2012, treatment started with an even lower CD4 (i.e.
 later).
-The British HIV Association 2015 guidelines recommend all individuals with suspected or diagnosed primary HIV infection ``are offered immediate ART''~\citep{churchillBritishHIVAssociation2016}.
+The British HIV Association 2015 guidelines recommend all individuals with suspected or diagnosed primary HIV infection ``are offered immediate ART''~\cite{churchillBritishHIVAssociation2016}.
 To have a more homogeneous setting in terms of access to ART, awareness of the HIV infection, etc., we cut the dated trees between 2012 and 2015.
@@
 
-The PN-test values were below 0.05 for all the forests (values between $3 \cdot 10^{-21}$ and $3 \cdot 10^{-13}$), strongly suggesting the presence of partner notification.
+The PN-test values were below 0.05 for all the forests (values between $7 \cdot 10^{-18}$ and $5 \cdot 10^{-12}$), strongly suggesting the presence of partner notification.
 
@@
 Moreover these document report about 400 HIV-related deaths per year.
-According to our estimate~\citep{zhukovaModelingDrugResistance2023}, 66.5\% of the HIV-infected individuals in the UK were infected with HIV-B.
+According to our estimate~\cite{zhukovaModelingDrugResistance2023}, 66.5\% of the HIV-infected individuals in the UK were infected with HIV-B.
 We hence estimated that in 2015 67\,298 (64\,837, 70\,291) people were living with HIV-B.
@@
 
-Using $\rho=0.58$, we estimated the partner notification probability $\upsilon=0.006-0.007$, %the following parameter values: $\lambda=0.53$, $\psi=0.43$, $\phi=23-59$, $\upsilon=0.006-0.007$, which corresponds to
-$R_e = 1.2$, infectious time $\frac{1}{\psi} = 2.3$ years, and 6--16 days between partner notification and their sampling.
+Using $\rho=0.58$, we estimated the partner notification probability $\upsilon=0.006-0.008$, %the following parameter values: $\lambda=0.53$, $\psi=0.43$, $\phi=23-59$, $\upsilon=0.006-0.007$, which corresponds to
+$R_e = 1.2$, infectious time $\frac{1}{\psi} = 2.3$ years, and 15--16 days between partner notification and their sampling.
 The results are consistent between 10 forests and shown in Table~\ref{tbl:uk}.
@@
 
-\begin{table}
-\caption{BD-PN model parameters estimated for UK HIV-1 B epidemic between 2012 and 2015, assuming that our data represents 58\% of infected cases ($\rho=0.58$).}
-\label{tbl:uk}
+For comparison, parameter estimation on the same forests with the BD model (with $\rho$ fixed to $0.58$) gave unrealistic results: $R_e = 0.5$ (suggesting contained epidemic) and very short infectious time $\frac{1}{\psi} = 0.4$ years.
+These results were robust across the 10 forests, see \nameref{S1_Table}.
+As with the results on simulated data, this  highlights the importance of accounting for partner notification during parameter estimation.
+
+\begin{table}[!ht]
+\begin{adjustwidth}{-0.8in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\centering
+\caption{
+{\bf BD-PN model parameters estimated for UK HIV-1 B epidemic between 2012 and 2015}, assuming that our data represents 58\% of infected cases ($\rho=0.58$).}
+
 \begin{tabular}{c|cc|cccc}
@@
 \toprule
- $1$ & $11\,117$ & $8\,838$& $1.24\;[1.21-1.27]$& $2.34\;[2.28-2.40]$& $8\;[5-12]$& $0.007\;[0.003-0.010]$ \\
- $2$ & $11\,109$ & $8\,833$& $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.39]$& $6\;[3-10]$& $0.006\;[0.003-0.009]$ \\
- $3$ & $11\,122$ & $8\,844$& $1.24\;[1.21-1.30]$& $2.34\;[2.27-2.40]$& $16\;[2-18]$& $0.006\;[0.002-0.015]$ \\
- $4$ & $11\,119$ & $8\,838$& $1.24\;[1.21-1.27]$& $2.34\;[2.28-2.40]$& $7\;[5-11]$& $0.007\;[0.004-0.011]$ \\
- $5$ & $11\,116$ & $8\,829$&  $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.39]$& $9\;[5-13]$& $0.006\;[0.003-0.010]$ \\
- $6$ & $11\,111$ & $8\,827$&  $1.24\;[1.22-1.27]$& $2.34\;[2.28-2.40]$& $8\;[5-12]$& $0.006\;[0.003-0.010]$ \\
- $7$ & $11\,119$ & $8\,838$& $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.39]$& $7\;[4-11]$& $0.006\;[0.003-0.010]$ \\
- $8$ & $11111$ & $8\,830$& $1.24\;[1.21-1.27]$& $2.33\;[2.27-2.39]$& $6\;[3-11]$& $0.006\;[0.003-0.008]$ \\
- $9$ & $11\,120$ & $8\,833$&  $1.24\;[1.21-1.27]$& $2.33\;[2.27-2.39]$& $7\;[4-12]$& $0.006\;[0.003-0.010]$ \\
- $10$ & $11\,112$ & $8\,827$& $1.25\;[1.22-1.27]$& $2.34\;[2.28-2.40]$& $7\;[4-11]$& $0.006\;[0.003-0.010]$ \\
+  $1$ & $11\,117$ & $8\,838$ & $1.24\;[1.22-1.29]$& $2.34\;[2.25-2.40]$& $15\;[3-17]$& $0.008\;[0.004-0.016]$ \\
+ $2$ & $11\,109$ & $8\,833$ & $1.24\;[1.21-1.27]$& $2.33\;[2.26-2.39]$& $16\;[1-18]$& $0.006\;[0.002-0.014]$ \\
+ $3$ & $11\,122$ & $8\,844$ & $1.24\;[1.21-1.29]$& $2.34\;[2.24-2.40]$& $15\;[3-17]$& $0.008\;[0.004-0.016]$ \\
+ $4$ & $11\,119$ & $8\,838$ & $1.24\;[1.21-1.29]$& $2.34\;[2.24-2.40]$& $15\;[3-17]$& $0.008\;[0.004-0.016]$ \\
+ $5$ & $11\,116$ & $8\,829$ & $1.24\;[1.22-1.27]$& $2.33\;[2.26-2.40]$& $16\;[3-18]$& $0.006\;[0.002-0.013]$ \\
+ $6$ & $11\,111$ & $8\,827$ & $1.24\;[1.21-1.27]$& $2.34\;[2.25-2.40]$& $16\;[3-18]$& $0.007\;[0.002-0.014]$ \\
+ $7$ & $11\,119$ & $8\,838$ & $1.24\;[1.21-1.27]$& $2.33\;[2.26-2.39]$& $15\;[2-18]$& $0.006\;[0.002-0.014]$ \\
+ $8$ & $11\,111$ & $8\,830$ & $1.24\;[1.19-1.27]$& $2.33\;[2.25-2.39]$& $16\;[2-18]$& $0.006\;[0.002-0.014]$ \\
+ $9$ & $11\,120$ & $8\,833$ & $1.24\;[1.21-1.27]$& $2.33\;[2.26-2.39]$& $16\;[2-18]$& $0.006\;[0.002-0.014]$ \\
+ $10$ & $11\,112$ & $8\,827$ & $1.25\;[1.22-1.29]$& $2.34\;[2.25-2.40]$& $16\;[2-18]$& $0.007\;[0.002-0.015]$ \\
  \bottomrule
  \end{tabular}
+%\begin{flushleft} We assumed that our data represents 58\% of infected cases: $\rho=0.58$.
+%\end{flushleft}
+\label{tbl:uk}
+\end{adjustwidth}
 \end{table}
 
-\section{Discussion}
-\label{disc}
-
+\subsubsection*{Application 2: Men-having-sex-with-men (MSM) HIV-1 B epidemic in Zurich}
+For comparison with the UK case, we applied our estimator to asses the partner notification for HIV-1 subtype B epidemic in a different setting: in men-having-sex-with-men (MSM) community from Zurich.
+We used the phylogenetic tree including 200 samples from the largest MSM HIV cluster in Zurich reconstructed by Rasmussen \textit{et al.}~\cite{Rasmussen2017} from the sequences collected as a part of the Swiss Cohort Study~\cite{swisshivcohortstudyCohortProfileSwiss2010a} between 1988 and 2014.
+As this data set was already small, we did not cut it to obtain a more homogeneous treatment policy, as we did for the UK.
+
+The PN test detected partner notification in this tree (p-value=0.046).
+Following Voznica~\textit{et al.}~\cite{Voznica2021}, we fixed the sampling probability for this tree at 25\%: $\rho=0.25$.
+Our bdpn estimator inferred the partner notification probability $\upsilon=0.07$ (CI $[0.006-0.190]$),
+$R_e = 1.4$ (CI $[1.2-1.6]$), infectious time $\frac{1}{\psi} = 8$ (CI $[6-10]$) years, and 3.4 months (CI [5 days -- 15 months]) between partner notification and their sampling.
+
+Overall the estimates on this much smaller dataset show larger CIs than those on the larger UK data set.
+The CIs estimated on the two data sets intersect for the effective reproductive number $R_e$, while the infectious time is larger for the Zurich data set (CI of $6-10$ years vs $2.2-2.4$ years for the UK data set).
+This is however expected as the Zurich data set covers a longer, earlier and non-homogeneous time interval (between 1974, the root of the tree, and 2012), including times with no or limited access to treatment, when the infectious time corresponded to progression to AIDS ($\approx 10$ years).
+The CIs for notification probability $\upsilon$ and partner removal time $\frac{1}{\phi}$ also intersect on the two data sets, though they do not have to, as the notification policy might differ between the two countries, and especially during the long and non-homogeneous time interval represented by Zurich data.
+The point estimates suggest stronger partner notification probability and longer notified partner removal time in Zurich than in the UK.
+
+\section*{Discussion}
 We proposed an extension of the phylodynamic birth-death model that accounts for non-random sampling due to partner notification (BD-PN).
@@
 
-We applied the PN-test and our estimator to the HIV-1 B epidemic in the UK, between 2012 and 2015.
-The test detected the strong presence of partner notification.
-The obtained estimates of epidemiological parameters are in agreement with what we now know about this epidemic (e.g., $R_e \approx 1.2$ ).
-We estimated a modest partner notification probability $\upsilon \approx 0.006$, and rapid notified partner detection ($\approx 8$ days vs $\approx2.3$ years for detection under the standard random procedure).
+We applied the PN-test and our estimator to the HIV-1 B epidemic in the UK, between 2012 and 2015 and to the HIV-1 B MSM cluster in Zurich.
+The test detected presence of partner notification in both cases.
+The obtained estimates of epidemiological parameters are in agreement with what we now know about the HIV-1 B epidemic (e.g., $R_e$ between $1.2$ and $1.6$ ).
+We estimated a modest partner notification probability ($1\%$ in the UK, $7\%$ in Zurich), and rapid notified partner detection ($\approx 16$ days vs $\approx 2.3$ years for detection under the standard random procedure in the UK; and $\approx 3.4$ months vs $\approx  8$ years for detection under the standard random procedure in Zurich).
 
 The BD-PN model is the first step towards accounting for non-random sampling in phylodynamic models.
-In order to reduce its mathematical complexity (having closed-form solutions for its differential equations) and make parameter estimation fast, we made several assumptions: (i) that only detected individuals can notify their partners; (ii) only the most recent partner can get notified; (iii) after the notification, notified partners do not transmit further; and (iv) notified partners are always sampled upon removal (i.e., observed in the transmission tree).
+In order to reduce its mathematical complexity (having closed-form solutions for its differential equations) and make parameter estimation fast, we made several assumptions: (i) that only detected individuals can notify their partners; (ii) only the most recent partner can get notified; (iii) notified partners are always sampled upon removal (i.e., observed in the transmission tree).
 These assumptions may be too simple to fully grasp the real partner notification process and should be relaxed in the future work.
 Another needed extension should account for population heterogeneity (i.e., MTBD-PN).
-As it is much easier to implement a tree simulator with a complex model than to derive its likelihood, a promising path to take for future -PN model developments would be using deep learning (e.g., adding new models to PhyloDeep~\citep{Voznica2021}).
-
-\bibliographystyle{plainnat}
-\bibliography{refs}
-
-\section{Appendix}
-
-\subsection{MTBD-PN tree simulator}
+As it is much easier to implement a tree simulator with a complex model (e.g., as we did for MTBD-PN models) than to derive its likelihood, a promising path to take for future -PN model developments would be using deep learning (e.g., adding new models to PhyloDeep~\cite{Voznica2021}).
+
+Finally, what we presented here is a specific case of an epidemiological model.
+However, it might be useful in other contexts as well.
+For instance in macroevolution, a detection of a new species might lead to faster detection of closely related ones.
+This process could be modeled in the same way as partner notification.
+
+\section*{Supporting information}
+
+% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+%\paragraph*{S1 Fig.}
+%\label{S1_Fig}
+%{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
+%
+%
+%
+%
+%
+%\paragraph*{S2 Fig.}
+%\label{S2_Fig}
+%{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+%
+%\paragraph*{S1 File.}
+%\label{S1_File}
+%{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+%
+%\paragraph*{S1 Video.}
+%\label{S1_Video}
+%{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+
+\paragraph*{S1 Appendix}
+\label{S1_Appendix}
+
+{\bf MTBD-PN tree simulator.} The appendix describes the tree simulator that we developed for generating sampled transmission trees under MTBD and MTBD-PN models.
+
+\subsection*{MTBD-PN tree simulator}
 
@@
 \subsubsection*{Code availability}
-The simulator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\citep{Huerta-Cepas2016} and NumPy package for array operations~\citep{harris_array_2020}.
+The simulator is implemented in Python 3. It uses ETE 3 framework for tree manipulation~\cite{Huerta-Cepas2016} and NumPy package for array operations~\cite{harris_array_2020}.
 
@@
 
+\paragraph*{S2 Appendix.}
+\label{S2_Appendix}
+{\bf Calculation of a probability of a branch containing a hidden partner.} The appendix describes the approximation of the probability of having a branch in the tree that contains one or two hidden partners.
+
+\subsection*{Mixed branch probability}
+
+To model the cases when a partner B is not sampled, we will need the probability of a \textbf{hidden partner subtree} $U_p(t, t_A)$, which started at time $t$ on a partner branch and evolved till time $T$ without any of its individuals being sampled, provided that the first of the partner's notifiers tried to notify them at time $t_A$ (B's branch in Fig.~\ref{fig:pn-branches}(2a,c,3a,c)).
+This probability consists of two parts: either the partner got notified at time $t_A$ but has not yet got sampled (like in Fig.~\ref{fig:pn-branches}(2a,3a)), $U_p^{(1)}(t, t_A)$;  or the partner got removed via the standard procedure before the time $t_A$ (like in Fig.~\ref{fig:pn-branches}(2c,3c)), $U_p^{(2)}(t, t_A)$:
+\begin{equation}
+U_p(t, t_A) = U_p^{(1)}(t, t_A) + U_p^{(2)}(t, t_A).
+\end{equation}
+
+The first part can be easily expressed via the previously defined equations: partner branch evolution before notification till time $t_A$ (if $t_A > t$) followed by partner branch evolution after notification till time $T$.
+Note, that neither  $p_{pb}^{(x)}(t, t_A)$ not $p_{pa}^{(x)}(t_A)$ include an event at the end of the branch, and in this case the partner branch does not end at $t_x = T$:
+\begin{equation}
+\begin{split}
+U_p^{(1)}(t, t_A) &= 
+\begin{cases}
+p_{pb}^{(A)}(t) p_{pa}^{(x)}(t_A) & \textit{\color{gray} if $t_A > t$}\\
+p_{pa}^{(x)}(t) & \textit{\color{gray} if $t_A \leq t$}
+\end{cases}{\color{gray},}\\
+&\textit{\color{gray}where $t_x =T$}\\
+\end{split}
+\end{equation}
+
+The second scenario is only possible if $t_A > t$.
+To calculate it, we would need to integrate over all possible partner removal times $t_B \in [t, t_A[$, provided that in the interval between $t$ and $t_B$ the partner might have transmitted to someone else (potentially multiple times), whose subtree(s) stayed unobserved till $T$.
+Note that $p_{pb}^{(A)}(t)$ describes a partner branch evolution between $t$ and $t_A$ with any number of hidden transmissions, no removal along it, and no event at the end of the branch.
+What we need instead is a removal with no sampling at time $\tau$.
+To approximate $U_p^{(2)}(t, t_A)$ we will divide $p_{pb}^{(A)}(t)$ by the probability of no removal during the time $(t_A - t)$ (i.e.
+$e^{-\psi(t_A - t)}$), multiply it by the probability of removal during that time ($1 - e^{-\psi(t_A - t)}$), and by the probability of not being sampled upon removal ($1 - \rho$):
+
+\begin{equation}
+U_p^{(2)}(t, t_A) \approx 
+\begin{cases}
+p_{pb}^{(A)}(t) \frac{1 - e^{-\psi(t_A - t)}}{e^{-\psi(t_A - t)}} (1 - \rho)& \textit{\color{gray} if $t_A > t$}\\
+0 & \textit{\color{gray} if $t_A \leq t$}
+\end{cases}
+\end{equation}
+
+%It is an approximation as instead of integrating over all possible intervals $[t, t_B[$ for the times of hidden transmissions from the partner branch, we use the interval $[t, t_A[$.
+
+We will also need to describe the probability of a\textbf{ mixed branch} $p_{m(s1,s2)}^{(i)}(t,t_r)$ for branches that include a hidden partner subtree (AB-C branches in Fig.~\ref{fig:pn-branches}(2) and A's tip branch in Fig.~\ref{fig:pn-branches}(3)).
+$s1$ here denotes the type of the top of the branch (e.g.
+between AB and BC in Fig.~\ref{fig:pn-branches}(2), partner) and $s2$ denotes the type of the bottom of the branch (e.g.
+between AB and A in Fig.~\ref{fig:pn-branches}(3), notifier).
+Let's denote the first notification time of the hidden partner as $t_r$, and the time of the hidden partner subtree start as $t_h$, while as usual $t_i$ stands for the time at the branch end.
+For instance, in Fig~\ref{fig:pn-branches}(2a) $t_i=t_C$, $t_r=t_A$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(2b) $t_i=t_r=t_C$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(3c) $t_i=t_r=t_A$, $t_h=t_{AB}$.
+
+$p_{m(s1,s2)}^{(i)}(t,t_r)$ combines two elements: the probability of the top part of the branch $p_{top=s1}^{(h)}(t, t_r)$, which finishes with a transmission to or from the hidden partner at time $t_h$ and the probability of the bottom part of the branch (without taking into account the event at its end) $p_{bottom=s2}^{(i)}(t_h)$.
+
+Note that since the branch contains a hidden partner, this partner must have at least one notifier (who is observed as all the notifiers).
+If the branch is external, the notifier could correspond to its tip: For example, such notifiers correspond to tip A in Fig.~\ref{fig:pn-branches}(2b) and tip C in Fig.~\ref{fig:pn-branches}(3b).
+If the branch is external, the notifier could only correspond to a tip in its node's supertree: For example, tip A who is part of the C's supertree in Fig.~\ref{fig:pn-branches}(2a,c).
+Having a notifier as the branch's tip implies that the bottom part of the branch corresponds to the notifier.
+Having a notifier in the supertree implies that the top part of the branch corresponds to the partner (already notified or not yet).
+Hence if the branch is internal, the top part can only correspond to the partner, while the bottom part can only correspond to a standard state (unnotified non-notifier), as the bottom part of such a branch corresponds to the partner's recipient (who could not be notified as the partner is unobserved and could not be a notifier as the branch is not external).
+Finally, if the branch is external, and the only partner's notifier corresponds to its tip, then there are two possibilities for its top part: (1a) either it fully corresponds to a standard state (unnotified non-notifier); or (1b) there is another partners' notifier in its supertree, and this other partner is also hidden along the top part of the same branch, hence the top part of the branch is partner-standard (see Fig.~\ref{fig:pn-mixed}).
+If, on the other hand, the branch is external, and the partner's notifier(s) is(are) only found in its supertree, then there are two possibilities for its bottom part: (2a) either it fully corresponds to a standard state (unnotified non-notifier); or (2b) there is another partners' notifier at its tip, and this other partner is also hidden along the bottom part of the same branch, hence the bottom part of the branch is standard-notifier (see Fig.~\ref{fig:pn-mixed}).
+Note that (1b) and (2b) represent the same case but from the point of view of two different hidden partners (B and their notifier A for the former, and D and their notifier C for the latter in Fig.~\ref{fig:pn-mixed}).
+
+\begin{figure*}[tbhp]
+\centering 
+%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
+\includegraphics[width=0.3\textwidth]{Fig_mixed}
+\caption{An example of mixed tip branch with two hidden partner subtrees: an individual C got sampled at time $t_C$ and notified their most recent partner D.
+D however, stayed unobserved as D did not get sampled by the end of the sampling period (time $T$).
+Before notification (at time $t_C$) D transmitted two times, the first transmission leading to an unobserved recipient subtree (gray dotted triangle) and the second (DB) leading to an observed recipient subtree (which includes a sampled individual A and a hidden individual B).
+The D's recipient (or someone in their subtree if there were hidden transmissions along the path DB-AB) transmitted further (transmission AB).
+Once A got sampled at time $t_A$, they notified their most recent partner B, who however was already removed by then via the standard procedure without sampling (at time $t_B < t_A$).
+Hence the branch CD-A represents a mixed branch with states partner(CD-DB)-standard(DB-AB)-notifier(AB-A).
+}
+\label{fig:pn-mixed} 
+\end{figure*}
+
+%if the branch is external the bottom part of the branch can either correspond to a notifier (as in Fig~\ref{fig:pn-branches}(2b,3)), or to a standard branch, or to a mixed tip branch starting as a standard branch and finishing as a notifier.
+It cannot correspond to another partner branch as every partner has their notifier sampled, and notifiers only notify one partner (who is already hidden along this branch).
+The bottom branch part evolution can then be expressed with the previously defined equations:
+
+We will first describe the one-hidden-partner case $p_{m(s1,s2)}^{(i)}(t,t_r)$ (where $s1 \in \{p, -\}$, $s2 \in \{n, -\}$ and $-$ stands for the standard (unnotified non-notifier) state), and then proceed with the two-partner case $p_{m(p,-,n)}^{(i)}(t,t_{r})$, where $t_{r}$ is the earliest notification time for the hidden partner at the top of the branch (e.g.
+$t_r=t_C$ in Fig.~\ref{fig:pn-mixed}), the only notification time for the hidden partner at the bottom of the branch being $t_i$  (e.g.
+$t_i=t_A$ in Fig.~\ref{fig:pn-mixed}).
+
+For the bottom branch part:
+
+\begin{equation}
+p_{bottom=s2}^{(i)}(t_h) = 
+\begin{cases}
+p_n^{(i)}(t_h)&\textit{\color{gray} if $i$ is a notifier tip ($s2=n$),}\\
+p^{(i)}(t_h)&\textit{\color{gray} otherwise ($s2=-$, the standard state)}\\
+\end{cases}
+\end{equation}
+
+We will approximate $p_{top=s1}^{(h)}(t)$ using the branch evolution probability of the type corresponding to $s1$  in the following way.
+For example, let us assume that $s1=-$ (standard branch).
+$p^{(h)}(t)$ expresses a probability of evolving along this branch between the times $t$ and $t_h$ with no or any number of hidden transmissions, without taking into account the event at the end of it.
+In our case there must be at least one hidden transmission: to/from the partner at time $t_h$ (where the partner tree stayed unsampled).
+If we remove the probability of no event during this time ($e^{-(\lambda + \psi)(t_h - t)}$) from $p^{(h)}(t)$, we will obtain a probability of a branch with at least one oriented hidden transmission somewhere between $t$ and $t_h$ (the probability of the corresponding hidden tree being included).
+$p_{top=-}^{(h)}(t)$ expresses almost the same thing, with the difference being that the last hidden transmission must have happened at time $t_h$ and correspond to the hidden partner.
+% the probability of the hidden partner tree corresponding to it is not included.
+We will hence approximate $p_{top=-}^{(h)}(t)$ as follows: 
+\begin{equation}
+p_{top=-}^{(h)}(t, t_r) \approx p^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}
+\end{equation}
+
+We will treat the case when the branch is in the partner state ($s1=p$) in a similar way:
+\begin{equation}
+\begin{split}
+p_{top=p}^{(h)}(t, t_r) &\approx 
+\begin{cases}
+p_{pb}^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}&\textit{\color{gray} if $t_h \leq t_r$}\\
+p_{pb}^{(r)}(t)\Big(p_{pa}^{(h)}(t_r) -e^{-(\lambda + \phi)(t_{h} - t_r)}\Big)&\textit{\color{gray} if $t \leq t_r \leq t_h$}\\
+p_{pa}^{(h)}(t) -e^{-(\lambda + \phi)(t_{h} - t)}&\textit{\color{gray} if $t > t_r$}\\
+\end{cases}\\
+&\textit{\color{gray}where $t_r$ is the first notification time of the hidden partner.}\\
+\end{split}
+\end{equation}
+
+Finally, we will account for the fact that the hidden tree starting at time $t_h$ corresponds to the hidden partner by dividing the expression by $U(t_h)$ (a standard hidden tree) and multiplying by $U_p(t_h, t_r)$ (partner hidden tree).
+
+Putting everything together, and approximating the time of the start of the hidden partner tree $t_h$ with the middle of the branch, we obtain:
+
+\begin{equation}
+\begin{split}
+p_{m(s1,s2)}^{(i)}(t,t_r) &= p_{top=s1}^{(h)}(t, t_r)\frac{U_p(t_h,t_r)}{U(t_h)}p_{bottom=s2}^{(i)}(t_h)\textit{\color{gray},}\\
+&\textit{\color{gray}where $t_{h} \approx t + \frac{(t_i - t)}{2}$ is an approximation of the time}\\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~of the hidden partner subtree start,}\\
+&\textit{\color{gray}~~~~~~~~$t_{r}$ is the earliest notification time of the hidden partner.}\\
+\end{split}\label{eq:p_mixed}
+\end{equation}
+
+Finally, we approximate the mixed case with two hidden partners using similar principals:
+\begin{equation}
+\begin{split}
+p_{m(p,-,n)}^{(i)}(t,t_r) &\approx p_{top=p}^{(h1)}(t, t_r)\frac{U_p(t_{h1},t_{r})}{U(t_{h1})} p_{top=-}^{(h2)}(t_{h1}, t_i)\frac{U_p(t_{h2},t_{i})}{U(t_{h2})} p_n^{(i)}(t_{h2})\textit{\color{gray},}\\
+&\textit{\color{gray}where $t_{h1} \approx t + \frac{1}{3}(t_i - t)$,}\\
+&\textit{\color{gray}~~~~~~~~$t_{h2} \approx t + \frac{2}{3}(t_i - t)$,}\\
+&\textit{\color{gray}~~~~~~~~$t_{r}$ is the earliest notification time of the partner }\\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~hidden at the top of the branch.}\\
+\end{split}\label{eq:p_mixed_mixed}
+\end{equation}
+
+\paragraph*{S3 Appendix.}
+\label{S3_Appendix}
+{\bf Tree likelihood calculation.} The appendix describes tree likelihood calculation case by case, accounting for all possible notifier-partner configurations.
+
+\subsection*{Tree likelihood calculation under BD-PN model}
+
+\subsubsection*{Case 1: $i$ is a tip}
+
+For a tip $i$ with a parent node $j$ the subtree likelihood density values can be calculated as:
+\begin{equation}
+\scriptsize
+\begin{array}{ll}
+l^{(\not{n},\not{n})}(i) &= p^{(i)}(t_j)\psi\rho (1-\upsilon) \textit{\color{gray} ~i.e. $i$ was not notified and did not notify}\\
+& \textit{\color{gray} (standard branch evolution followed by sampling and no notification)}\\
+&~~+p_{m(-,n)}^{(i)}(t_j,t_i)\psi\rho\upsilon \textit{\color{gray} ~i.e. $i$ was not notified but notified their partner who is hidden}\\
+& \textit{\color{gray} (mixed standard-notifier branch evolution followed by sampling and notification)}\\
+\\
+l^{(\not{n},n)}(i) &= p_n^{(i)}(t_j)\psi\rho
+\upsilon \textit{\color{gray} ~i.e. $i$ was not notified but notified their most recent patner (observed)}\\
+& \textit{\color{gray} (notifier branch evolution followed by sampling and notification)}\\
+&~~\\
+\\
+l^{(n,\not{n})}(i,r) &= \begin{cases}
+%0 \textit{\color{gray} ~if $t_r < t_j$}\\
+%\textit{\color{gray} (not possible as once notified partners do not transmit)}
+%\\
+%\\
+p_{pa}^{(i)}(t_j)\phi(1 - \upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_r < t_j$ and did not notify}\\
+\textit{\color{gray} (notified partner branch evolution followed by notification-provoked sampling and no notification)}
+\\
+~~~+p_{m(pa,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r < t_j$, }\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$  did not notify}\\
+ \textit{\color{gray} ~~~(mixed partner-standard branch evolution followed by sampling and no notification)}\\
+ ~~~+p_{m(pa,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
+ \textit{\color{gray} ~~~(mixed partner-notifier branch evolution followed by sampling and notification)}
+\\
+~~~+p_{m(pa,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r < t_j$, }\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
+ \textit{\color{gray} ~~~(mixed partner-standard-notifier branch evolution followed by sampling and notification)}
+\\
+\\
+p_{pb}^{(i)}(t_j)\psi\rho(1 - \upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and did not notify}\\
+\textit{\color{gray} (partner-before-notification branch evolution followed by standard sampling and no notification)}
+\\
+~~~+p_{m(pb,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r > t_i$, }\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$  did not notify}\\
+ \textit{\color{gray} ~~~(mixed partner-standard branch evolution followed by sampling and no notification)}\\
+ ~~~+p_{m(pb,n)}^{(i)}(t_j,t_i)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
+ \textit{\color{gray} ~~~(mixed partner-notifier branch evolution followed by sampling and notification)}
+\\
+~~~+p_{m(pb,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r > t_i$, }\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
+ \textit{\color{gray} ~~~(mixed partner-standard-notifier branch evolution followed by sampling and notification)}
+\\
+\\
+p_{pb}^{(r)}(t_j) p_{pa}^{(i)}(t_r) \phi (1-\upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_j \leq t_r \leq t_i$ and did not notify}\\
+\textit{\color{gray} (partner-before-notification branch evolution followed by} \\
+\textit{\color{gray}~notified partner branch evolution, notification-provoked sampling and no notification)}
+\\
+~~~+p_{m(p,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_j \leq t_r \leq t_i$, }\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$  did not notify}\\
+ \textit{\color{gray} ~~~(mixed partner-standard branch evolution followed by sampling and no notification)}\\
+ ~~~+p_{m(p,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
+ \textit{\color{gray} ~~~(mixed partner-notifier branch evolution followed by sampling and notification)}
+\\
+~~~+p_{m(p,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_j \leq t_r \leq t_i$, }\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
+ \textit{\color{gray} ~~~(mixed partner-standard-notifier branch evolution followed by sampling and notification)}
+\end{cases}\\
+\\
+l^{(n,n)}(i,r) &= \begin{cases}
+p_b^{(i)}(t_j)\phi\upsilon \textit{\color{gray} ~if $i$ got notified at time $t_r < t_j$ and notified their most recent patner}\\
+\textit{\color{gray} (notified notifier branch evolution followed by notification-provoked sampling and  notification)}\\
+\\
+p_n^{(i)}(t_j)\psi\rho\upsilon \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and notified their most recent patner}\\
+\textit{\color{gray} (notifier branch evolution followed by standard sampling and  notification)}\\
+\\
+p_n^{(r)}(t_j) p_b^{(i)}(t_r) \phi \upsilon \textit{\color{gray} ~if $i$ got notified at time $t_j \leq t_r \leq t_i$ and notified}\\
+\textit{\color{gray} (notifier branch evolution followed by notified notifier branch evolution,} \\
+\textit{\color{gray}~notification-provoked sampling and notification)}\\
+\end{cases}\\
+\end{array}
+\label{eq:tip}
+\end{equation}
+
+\subsubsection*{Case 2: $i$ is an internal node} 
+Let us now consider an internal node $i$ with two child nodes, $i_0$ and $i_1$.
+
+\subsubsection*{Case 2.1: $i$ is an unnotified internal node} 
+We will start with the configuration where the $i$'s branch corresponds to an unnotified individual.
+Hence the $i$'s branch evolution is represented by $p^{(i)}(t_j)$ with a transmission event at the end of it, where each of the child branches can correspond to the donor (probability density of $2\lambda$).
+We will consider three possibilities: (2.1.1) when both $i_0$ and $i_1$ are internal nodes; (2.1.2) when both $i_0$ and $i_1$ are tips; and (2.1.3) when one of them is an internal node and the other one is a tip.
+
+%Note that a notification will only be possible if the notifying branch is external as the notifier only notifies the most recent partner, i.e.
+the transmission corresponds to the notifier's parent node.
+
+\subsubsection*{Case 2.1.1: $i$ is an unnotified internal node with internal-node children $i0$ and $i1$}
+
+First, let's assume that $i_0$ and $i_1$ are internal. Then none of them can be a notifier (since notifiers correspond to tips), and hence none of them is notified (as $i$ is also unnotified):
+
+\begin{equation}
+\scriptsize
+l^{(\not{n})}(i) = p^{(i)}(t_j) 2\lambda l^{(\not{n})}(i_0)l^{(\not{n})}(i_1) \label{eq:lu-int-int}
+\end{equation}
+
+\subsubsection*{Case 2.1.2: $i$ is an unnotified internal node with tip children $i0$ and $i1$}
+
+Secondly, let's assume that $i_0$ and $i_1$ are tips. Then each of them could have notified the other one:
+
+\begin{equation}
+\scriptsize
+\begin{split}
+l^{(\not{n})}(i) = p^{(i)}(t_j) 2\lambda \cdot 
+\Big(&~l^{(\not{n},\not{n})}(i_0)l^{(\not{n},\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ neigher $i_0$ nor $i_1$ notified} \\
+&+ l^{(\not{n},n)}(i_0)l^{(n,\not{n})}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $i_0$ notified $i_1$ and $i_1$ did not notify} \\
+&+ l^{(n,\not{n})}(i_0,i_1)l^{(\not{n},n)}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ did not notify, but $i_1$ notified $i_0$} \\
+&+ l^{(n,n)}(i_0,i1)l^{(n,n)}(i_1,i_0)~\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ and $i_0$ notified each other} \\
+\end{split}\label{eq:lu-tip-tip}
+\end{equation}
+
+\subsubsection*{Case 2.1.3: $i$ is an unnotified internal node whose children are a tip and an internal node}
+Lastly, let's assume that $i_0$ is a tip and $i_1$ is internal (the scenario where $i_1$ is a tip and $i_0$ is internal can be obtained by swapping the labels). Then $i_0$ could have notified $i_1$:
+
+\begin{equation}
+\scriptsize
+\begin{split}
+l^{(\not{n})}(i) = p^{(i)}(t_j) 2\lambda \cdot 
+\Big(&~l^{(\not{n},\not{n})}(i_0)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ did not notify $i_1$ } \\
+&+ l^{(\not{n},n)}(i_0)l^{(n)}(i_1,i_0)~\Big) \textit{\color{gray}~$\leftarrow$ $i_0$ notified $i_1$} \\
+\end{split}\label{eq:lu-int-tip}
+\end{equation}
+
+\subsubsection*{Case 2.2: $i$'s branch is internal and (at least initially) corresponds to an (eventually notified) partner}
+
+In the other case the internal node $i$'s branch starts as a partner who will eventually get notified by a tip $r$.
+There are two possibilities: either the node $i$ corresponds to the partner or the partner is hidden along the $i$'s branch and $i$ corresponds to someone within $i$'s recipient subtree (e.g., if $i=C$ in Fig.~\ref{fig:pn-branches}(2a)).
+In the first case, the $i$'s branch evolution is represented by the partner branch evolution $p_{p}^{(i)}(t_j, t_r)$:
+
+\begin{equation}
+\scriptsize
+p_{p}^{(i)}(t_j, t_r) =
+\begin{cases}
+p_{pa}^{(i)}(t_j) &\textit{\color{gray} if $t_r < t_j$,}\\
+p_{pb}^{(i)}(t_j) &\textit{\color{gray} if $t_r > t_i$,}\\
+p_{pb}^{(r)}(t_j)p_{pa}^{(i)}(t_r) &\textit{\color{gray} if $t_j \leq t_r \leq t_i$.}\\
+\end{cases}\label{eq:p-p}
+\end{equation}
+This branch has a transmission event at the end of it, where the donor is known (probability density of $\lambda$): The donor corresponds to the subtree containing the tip representing the sampling of the individual $i$.
+
+In the second case, the $i$'s branch evolution is represented by $p_{m(p,-)}^{(i)}(t_j,t_r)$ with a transmission event at the end of it, where the donor is unknown (probability density of $2\lambda$): any of the subtrees can be the donor.
+ %Moreover, this recipient's branch cannot be notified by our partner (as in our model notified partners do not notify further):
+
+Again, we will consider three possibilities for $i$'s child nodes: (2.2.1) when both $i_0$ and $i_1$ are internal nodes; (2.2.2) when both $i_0$ and $i_1$ are tips; and (2.2.3) when one of them is an internal node and the other one is a tip.
+
+%In any of these scenarios, the notification must have happened after the partner tip branch start (as one of the model assumptions is that partners do not transmit after notification). Hence:
+%\begin{equation}
+%\scriptsize
+%l^{(n)}(i, r) = 0 \textit{\color{gray}~if $t_i > t_r$} \\
+%\label{eq:lp-zero}
+%\end{equation}
+
+\subsubsection*{Case 2.2.1: $i$'s branch is internal and (at least initially) corresponds to a partner with internal-node children $i0$ and $i1$}
+
+First, let's assume that $i_0$ and $i_1$ are internal.
+Then none of them can be a notifier (since notifiers correspond to tips), and hence either exactly one of them is notified by $r$, or the partner is hidden:
+
+\begin{equation}
+\scriptsize
+\begin{split}
+l^{(n)}(i, r) = &p_{p}^{(i)}(t_j, t_r) \lambda \cdot
+\Big(~l^{(n)}(i_0,r)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $i_0$ is notified by $r$} \\
+& ~~~~~~~~~~~~~~~~~~~+ l^{(\not{n})}(i_0)l^{(n)}(i_1,r)~\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ is notified by $r$} \\
+& + p_{m(p,-)}^{(i)}(t_j,t_r) 2 \lambda l^{(\not{n})}(i_0)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ hidden partner}\\
+ \end{split}
+\label{eq:lp-int-int}
+\end{equation}
+
+\subsubsection*{Case 2.2.2: $i$'s branch is internal and (at least initially) corresponds to a partner with tip children $i0$ and $i1$} 
+Secondly, let's assume that $i_0$ and $i_1$ are tips. Then each of them could have notified the other one (in addition to one of them potentially being notified by $r$):
+
+\begin{equation}
+\scriptsize
+\begin{split}
+l^{(n)}(i, r) = &p_{p}^{(i)}(t_j,t_r) \lambda \cdot
+\Big(~l^{(n,\not{n})}(i_0,first(r,i_1))l^{(\not{n},n)}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ and $i_1$ notified $i_0$, $i_0$ did not notify} \\
+&~~~~~~~~~~~~~~~~~~~+ l^{(n,n)}(i_0,first(r,i_1))l^{(n,n)}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $r$ and $i_1$ notified $i_0$, $i_0$ notified $i_1$} \\
+&~~~~~~~~~~~~~~~~~~~+l^{(\not{n},n)}(i_0)l^{(n,\not{n})}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ $r$ and $i_0$ notified $i_1$, $i_1$ did not notify} \\
+&~~~~~~~~~~~~~~~~~~~+ l^{(n,n)}(i_0, i_1)l^{(n,n)}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ $r$ and $i_0$ notified $i_1$, $i_1$ notified $i_0$} \\
+&~~~~~~~~~~~~~~~~~~~+l^{(n,n)}(i_0,r)l^{(n,\not{n})}(i_1, i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$ notified $i_1$, $i_1$ did not notify} \\
+&~~~~~~~~~~~~~~~~~~~+ l^{(n,\not{n})}(i_0, i_1)l^{(n,n)}(i_1,r) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_1$ notified $i_0$, $i_0$ did not notify} \\
+&~~~~~~~~~~~~~~~~~~~+l^{(n,\not{n})}(i_0,r)l^{(\not{n},\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$ and $i_1$ did not notify} \\
+&~~~~~~~~~~~~~~~~~~~+ l^{(\not{n},\not{n})}(i_0)l^{(n,\not{n})}(i_1,r)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_0$ and $i_1$ did not notify} 
+\\
+&+ p_{m(p,-)}^{(i)}(t_j,t_r) 2 \lambda \cdot
+\Big(~l^{(n,\not{n})}(i_0,i_1)l^{(\not{n},n)}(i_1) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i_1$ notified $i_0$, $i_0$ did not notify} \\
+&~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~+ l^{(n,n)}(i_0,i_1)l^{(n,n)}(i_1, i_0)\textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i_1$ notified $i_0$, $i_0$ notified $i_1$} \\
+&~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~+l^{(\not{n},n)}(i_0)l^{(n,\not{n})}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i_0$ notified $i_1$, $i_1$ did not notify} \\
+&~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~+l^{(\not{n},\not{n})}(i_0)l^{(\not{n},\not{n})}(i_1)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i_0$ and $i_1$ did not notify} \\
+\\
+&\textit{\color{gray}~where } \color{gray}first(a,b) = 
+\begin{cases}
+a \textit{~if $t_a \leq t_b$}\\
+b \textit{~if $t_a > t_b$.}
+\end{cases} 
+ \end{split}
+\label{eq:lu-tip-tip}
+\end{equation}
+
+\subsubsection*{Case 2.2.3: $i$'s branch is internal and (at least initially) corresponds to a partner, $i$'s children are a tip and an internal node}
+Lastly, let's assume that $i_0$ is a tip and $i_1$ is internal (the scenario where $i_1$ is a tip and $i_0$ is internal can be obtained by swapping the labels).
+Then $i_0$ could have notified $i_1$ (in addition to one of them being potentially notified by $r$):
+
+\begin{equation}
+\scriptsize
+\begin{split}
+l^{(n)}(i, r) = &p_{p}^{(i)}(t_j,t_r) \lambda \cdot
+\Big(~l^{(n,n)}(i_0,r)l^{(n)}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_0$, $i_0$  notified $i_1$} \\
+&~~~~~~~~~~~~~~~~~~~+ l^{(n,\not{n})}(i_0,r)l^{(\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$  $r$ notified $i_0$, $i_0$ did not notify} \\
+&~~~~~~~~~~~~~~~~~~~+l^{(\not{n}, n)}(i_0)l^{(n)}(i_1,first(r,i_0)) \textit{\color{gray}~$\leftarrow$ both $r$ and $i_0$ notified $i_1$} \\
+&~~~~~~~~~~~~~~~~~~~+ l^{(\not{n},\not{n})}(i_0)l^{(n)}(i_1,r)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified $i_1$, $i_0$ did not notify} 
+\\
+&+ p_{m(p,-)}^{(i)}(t_j,t_r) 2\lambda \cdot
+\Big(~l^{(\not{n},n)}(i_0)l^{(n)}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i_0$ notified $i_1$} \\
+&~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~+ l^{(\not{n},\not{n})}(i_0)l^{(\not{n})}(i_1)~\Big) \textit{\color{gray}~$\leftarrow$ $r$ notified a hidden partner,} \\
+&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ $i_0$ did not notify}\\
+\textit{\color{gray}~where } &\color{gray}first(a,b) = 
+\begin{cases}
+a \textit{~if $t_a \leq t_b$}\\
+b \textit{~if $t_a > t_b$.}
+\end{cases} 
+ \end{split}
+\label{eq:lp-tip-int}
+\end{equation}
+
+\bigskip 
+Note that $l^{(n)}(i, r)$, $l^{(n,n)}(i, r)$ and $l^{(n,\not{n})}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time $t_r$ is known, i.e.
+once we have climbed to an internal node corresponding to the transmission between the notifier $r$ (corresponding to one of its children, must be a tip) and the partner (the other child).
+The calculation is done recursively by descending into the subtree.
+The tree likelihood calculation hence can take $O(n^2)$ steps in the worst case (ladder tree), where $n$ is the number of tips in the tree.
+However, as all the unotified equation parts ($l^{(\not{n})}(i)$, $l^{(\not{n},n)}(i)$ and $l^{(\not{n},\not{n})}(i)$) only depend on the corresponding node $i$, they can be pre-calculated and reused to speed up the recursive steps.
+
+\paragraph*{S1 Table.}
+\label{S1_Table}
+{\bf BD model parameters estimated for UK HIV-1 B epidemic between 2012 and 2015}, assuming that our data represents 58\% of infected cases ($\rho=0.58$).
+
+\begin{table}[!ht]
+%\begin{adjustwidth}{-0.8in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\centering
+\caption{
+{\bf BD model parameters estimated for UK HIV-1 B epidemic between 2012 and 2015}, assuming that our data represents 58\% of infected cases ($\rho=0.58$).}
+
+\begin{tabular}{c|cc|cc}
+&&&&infectious\\
+fo-&num.&num.&&time\\
+rest&tips&trees&$R_e$&$\frac{1}{\psi}$ [years]\\
+\toprule
+ $1$ & $11\,117$ & $8\,838$ & $0.53\;[0.52-0.55]$& $0.37\;[0.36-0.37]$ \\
+ $2$ & $11\,109$ & $8\,833$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
+ $3$ & $11\,122$ & $8\,844$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
+ $4$ & $11\,119$ & $8\,838$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
+ $5$ & $11\,116$ & $8\,829$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
+ $6$ & $11\,111$ & $8\,827$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
+ $7$ & $11\,119$ & $8\,838$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
+ $8$ & $11\,111$ & $8\,830$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
+ $9$ & $11\,120$ & $8\,833$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
+ $10$ & $11\,112$ & $8\,827$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
+ \bottomrule
+ \end{tabular}
+%\begin{flushleft} We assumed that our data represents 58\% of infected cases: $\rho=0.58$.
+%\end{flushleft}
+\label{tbl:uk}
+%\end{adjustwidth}
+\end{table}
+
+\section*{Acknowledgments}
+The authors would like to thank Dr Jakub Voznica for fruitful discussions on partner notification mechanisms.
+
+\nolinenumbers
+
+\bibliography{refs}
+
+% Either type in your references using
+% \begin{thebibliography}{}
+% \bibitem{}
+% Text
+% \end{thebibliography}
+%
+% or
+%
+% Compile your BiBTeX database using our plos2015.bst
+% style file and paste the contents of your .bbl file
+% here. See http://journals.plos.org/plosone/s/latex for 
+% step-by-step instructions.
+% 
+%\begin{thebibliography}{10}
+%
+%\bibitem{bib1}
+%Conant GC, Wolfe KH.
+%\newblock {{T}urning a hobby into a job: how duplicated genes find new
+%  functions}.
+%\newblock Nat Rev Genet. 2008 Dec;9(12):938--950.
+%
+%\bibitem{bib2}
+%Ohno S.
+%\newblock Evolution by gene duplication.
+%\newblock London: George Alien \& Unwin Ltd. Berlin, Heidelberg and New York:
+%  Springer-Verlag.; 1970.
+%
+%\bibitem{bib3}
+%Magwire MM, Bayer F, Webster CL, Cao C, Jiggins FM.
+%\newblock {{S}uccessive increases in the resistance of {D}rosophila to viral
+%  infection through a transposon insertion followed by a {D}uplication}.
+%\newblock PLoS Genet. 2011 Oct;7(10):e1002337.
+%
+%\end{thebibliography}
+
 \end{document}

### 2024-06-10 bdpn/bd_model.py
Added likelihood calculation programs

@@ -163,5 +163,27 @@ def main():
     save_results(vs, cis, params.log, ci=params.ci)
 
 
+def loglikelihood_main():
+    """
+    Entry point for tree likelihood estimation with the BD model with command-line arguments.
+    :return: void
+    """
+    import argparse
+
+    parser = \
+        argparse.ArgumentParser(description="Calculate BD likelihood on a given forest for given parameter values.")
+    parser.add_argument('--la', required=True, type=float, help="transmission rate")
+    parser.add_argument('--psi', required=True, type=float, help="removal rate")
+    parser.add_argument('--rho', required=True, type=float, help='sampling probability')
+    parser.add_argument('--nwk', required=True, type=str, help="input tree file")
+    params = parser.parse_args()
+
+    forest = read_forest(params.nwk)
+    lk = loglikelihood(forest, la=params.la, psi=params.psi, rho=params.rho)
+    print(lk)
+
+
 if '__main__' == __name__:
     main()
+
+

### 2024-06-10 bdpn/bdpn_model.py

@@ -492,6 +492,28 @@ def main():
     save_results(vs, cis, params.log, ci=params.ci)
 
 
+def loglikelihood_main():
+    """
+    Entry point for tree likelihood estimation with the BDPN model with command-line arguments.
+    :return: void
+    """
+    import argparse
+
+    parser = \
+        argparse.ArgumentParser(description="Calculate BDPN likelihood on a given forest for given parameter values.")
+    parser.add_argument('--la', required=True, type=float, help="transmission rate")
+    parser.add_argument('--psi', required=True, type=float, help="removal rate")
+    parser.add_argument('--rho', required=True, type=float, help='sampling probability')
+    parser.add_argument('--upsilon', required=True, type=float, help='notification probability')
+    parser.add_argument('--phi', required=True, type=float, help='partner removal rate')
+    parser.add_argument('--nwk', required=True, type=str, help="input tree file")
+    params = parser.parse_args()
+
+    forest = read_forest(params.nwk)
+    lk = loglikelihood(forest, la=params.la, psi=params.psi, rho=params.rho, phi=params.phi, upsilon=params.upsilon)
+    print(lk)
+
+
 def infer(forest, la=None, psi=None, phi=None, p=None, pn=None,
           lower_bounds=DEFAULT_LOWER_BOUNDS, upper_bounds=DEFAULT_UPPER_BOUNDS, ci=False, **kwargs):
     """

### 2024-06-10 setup.py

@@ -24,6 +24,8 @@ setup(
             'console_scripts': [
                 'bdpn_infer = bdpn.bdpn_model:main',
                 'bd_infer = bdpn.bd_model:main',
+                'bdpn_loglikelihood = bdpn.bdpn_model:loglikelihood_main',
+                'bd_loglikelihood = bdpn.bd_model:loglikelihood_main',
                 'pn_test = bdpn.model_distinguisher:main',
             ]
     },

### 2024-06-14 bdpn/bd_model.py
Version 0.1.3

@@ -110,6 +110,8 @@ def infer(forest, la=None, psi=None, p=None,
                                              loglikelihood=loglikelihood, bounds=bounds[input_params == None],
                                              start_parameters=start_parameters, cis=ci)
     print('Estimated BD parameters: {}'.format(vs))
+    if ci:
+        print('Estimated CIs:\n{}'.format(cis))
     return vs, cis
 
 

### 2024-06-14 bdpn/bdpn_model.py

@@ -46,143 +46,6 @@ def log_subtraction(log_minuend, log_subtrahend):
     return np.log(np.sum(np.exp(result) * [1, -1])) - factors
 
 
-def log_u_p(t, tr, T, la, psi, rho, phi, c1, c2, E_t, E_tr):
-    """
-    Calculates a log probability of a partner subtree evolving between the time t and T unobserved,
-    provided the first partner notification happened at time tr.
-    In practice, two possibilities are summed up:
-    (1) the hidden partner branch has not yet got sampled between the notification and T;
-    (2) the hidden partner branch got removed before the notification (without sampling).
-
-
-    :param t: start time of the hidden partner subtree evolution
-    :param tr: moment of the earliest partner notification
-    :param T: end of sampling time
-    :param E_t: E_t = c2 * exp(c1 * (t - T)), where c2 = (c1 + la - psi) / (c1 - la + psi)
-    :param E_tr: E_tr = c2 * exp(c1 * (tr - T))
-    :return: log of the hidden partner subtree probability (approximation)
-    """
-    E_T = get_E(c1, c2, T, T)
-    if tr > t:
-        branch_len = tr - t
-
-        # Partner branch evolution from t till the moment of notification (tr)
-        log_ppb = get_log_ppb(la=la, psi=psi, c1=c1, t=t, ti=tr, E_t=E_t, E_ti=E_tr)
-        # Notified-partner branch evolution after tr
-        log_ppa = get_log_ppa(la, psi, phi, c1, tr, T, E_t=E_tr, E_ti=E_T)
-        log_hidden_partner_not_sampled = log_ppb + log_ppa
-
-        # Hidden partner branch removed before the notification:
-        # divide the oriented branch evolution
-        # by no-removal-event-along-the-branch probability (e^{-psi * branch_len})
-        # and multiply by the removal-event-along-the-branch probability (1 - e^{-psi * branch_len})
-        # with no sampling, in a log form
-        log_hidden_partner_sampled_before = log_ppb \
-                                            + psi * branch_len + np.log(1 - np.exp(-psi * branch_len)) \
-                                            + np.log(1 - rho)
-
[…]

### 2024-06-14 bdpn/parameter_estimator.py

@@ -95,8 +95,12 @@ def optimize_likelihood_params(forest, input_parameters, loglikelihood, bounds, 
 
     n = len(optimised_parameters)
     optimised_cis = np.zeros((n, 2))
-    optimised_cis[:, 0] = np.array(optimised_parameters)
-    optimised_cis[:, 1] = np.array(optimised_parameters)
+    bound_iterator = iter(bounds)
+    for bs_i in range(n):
+        if input_parameters[bs_i] is not None:
+            optimised_cis[bs_i, :] = input_parameters[bs_i], input_parameters[bs_i]
+        else:
+            optimised_cis[bs_i, :] = next(bound_iterator)
 
     if cis:
         print('Estimated parameters:', optimised_parameters)
@@ -120,12 +124,12 @@ def optimize_likelihood_params(forest, input_parameters, loglikelihood, bounds, 
             return binary_search(v, v_max, get_lk, lower)
 
         def get_ci(args):
-            (i, optimised_value), (b_min, b_max) = args
+            (i, optimised_value) = args
             # print('---------')
             # print(i, optimised_value, (b_min, b_max))
 
             bs = []
-            bound_iterator = iter(bounds)
+            bound_iterator = iter(optimised_cis)
             for bs_i in range(n):
                 if input_parameters[bs_i] is not None:
                     continue
@@ -145,17 +149,8 @@ def optimize_likelihood_params(forest, input_parameters, loglikelihood, bounds, 
                 return optimize_likelihood_params(forest, ip, loglikelihood, bounds=bs, start_parameters=rps,
                                                   cis=False, threads=threads)[-1]
 
-            # Try to narrow down the bounds a bit
-            v = max(b_min, optimised_value / 2)
-            if get_lk(v) < lk_threshold:
-                b_min = v
-
-            v = min(b_max, 1.5 * optimised_value)
-            if get_lk(v) < lk_threshold:
-                b_max = v
-
-            optimised_cis[i, 0] = binary_search(b_min, optimised_value, get_lk, True)
-            optimised_cis[i, 1] = binary_search(optimised_value, b_max, get_lk, False)
+            optimised_cis[i, 0] = binary_search(optimised_cis[i, 0], optimised_value, get_lk, True)
+            optimised_cis[i, 1] = binary_search(optimised_value, optimised_cis[i, 1], get_lk, False)
             # print(i, optimised_cis[i, :])
             # print('---------')
 
[…]

### 2024-06-14 setup.py

@@ -13,7 +13,7 @@ setup(
         'Topic :: Software Development',
         'Topic :: Software Development :: Libraries :: Python Modules',
     ],
-    version='0.1.2',
+    version='0.1.3',
     description='Estimation of BDPN parameters from phylogenetic trees.',
     author='Anna Zhukova',
     author_email='anna.zhukova@pasteur.fr',

### 2024-07-05 bdpn/parameter_estimator.py
v0.1.13: fixed within bound constraint

@@ -50,7 +50,7 @@ def optimize_likelihood_params(forest, T, input_parameters, loglikelihood_functi
                                                   np.finfo(np.float64).eps))
 
     def get_real_params_from_optimised(ps):
-        # ps = np.maximum(np.minimum(ps, optimised_bounds[:, 1]), optimised_bounds[:, 0])
+        ps = np.maximum(np.minimum(ps, optimised_bounds[:, 1]), optimised_bounds[:, 0])
         ps[optimise_as_logs] = np.power(10, ps[optimise_as_logs])
         result = np.array(input_parameters)
         result[optimised_parameter_mask] = ps

### 2024-07-05 setup.py

@@ -13,7 +13,7 @@ setup(
         'Topic :: Software Development',
         'Topic :: Software Development :: Libraries :: Python Modules',
     ],
-    version='0.1.12',
+    version='0.1.13',
     description='Estimation of BDPN parameters from phylogenetic trees.',
     author='Anna Zhukova',
     author_email='anna.zhukova@pasteur.fr',

### 2024-07-05 upload_new_version.sh

@@ -1,4 +1,4 @@
 rm -rf bdpn.egg-info build dist
 python setup.py sdist bdist_wheel
 twine upload dist/* && \
-sudo docker build -t evolbioinfo/bdpn:v0.1.12 -f Dockerfile . && sudo docker login && sudo docker push evolbioinfo/bdpn:v0.1.12
+sudo docker build -t evolbioinfo/bdpn:v0.1.13 -f Dockerfile . && sudo docker login && sudo docker push evolbioinfo/bdpn:v0.1.13

### 2024-07-05 article/article_plos.tex
Version 0.1.3
Article abstract

@@
 Anna Zhukova\textsuperscript{1,2,3*},
-Mathieu Moslonka-Lefebvre\textsuperscript{1\textcurrency},
+%Mathieu Moslonka-Lefebvre\textsuperscript{1\textcurrency},
 Olivier Gascuel\textsuperscript{1,4},
@@
 % Current address notes
-\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
 % \textcurrency b Insert second current address 
@@
 \section*{Abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+Phylodynamics bridges the gap between classical epidemiology and pathogen genome sequence data by estimating epidemiological parameters from time-scaled pathogen phylogenetic trees.
+The models used in phylodynamics typically assume that the sampling procedure is independent between infected individuals.
+However this assumption does not hold for many epidemics, in particular for such sexually transmitted infections as HIV-1, for which partner notification schemes are included in health policies of many countries.
+
+We developed an extension of phylodynamic multi-type birth-death (MTBD) models with partner notification (PN), and a simulator to generate trees under MTBD and MTBD-PN models.
+We proposed a non-parametric test for detecting partner notification in pathogen phylogenetic trees.
+Its application to simulated data showed that it is both highly specific and sensitive.
+For the simplest representative of the MTBD-PN family, the BD-PN model, we developed a maximum likelihood parameter and CI estimator.
+It performed accurate estimations on simulated data, and detected partner notification in HIV-1 B epidemics in Zurich and the UK.
+Importantly, we showed that not accounting for partner notification when it is present leads to bias in parameter estimation with the BD model, while BD-PN parameter estimator performs well both in presence and in absence of partner notification.
+
+Our PN test, MTBD-PN tree simulator and BD-PN parameter estimator are freely available at \href{github.com/evolbioinfo/treesimulator}{https://github.com/evolbioinfo/treesimulator} and \href{github.com/evolbioinfo/bdpn}{https://github.com/evolbioinfo/bdpn}.
 
@@
 \section*{Author summary}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+Phylodynamic models can estimate epidemiological parameters from pathogen phylogenetic trees (i.e., genealogies).
+These models typically assume that sampling of pathogen sequences from detected cases is independent between infected individuals.
+However this assumption does not hold for many epidemics, in particular for such sexually transmitted infections as HIV-1, for which partner notification (contact tracing) is included in health policies of many countries.
+
+We developed a phylodynamic model accounting for partner notification and a test for detection of partner notification in pathogen phylogenetic trees.
+They showed good performance both on simulated and real data (HIV-1 B epidemics in Zurich and the UK).
+Importantly, we showed that not accounting for partner notification when it is present leads to bias in parameter estimation, while our new estimator performs well both in presence and in absence of partner notification.
 
@@
 Phylodynamics leverages this footprint to estimate epidemiological parameters~\cite{Grenfell2004a,Volz2013}.
-It relies on models that bridge the gap between traditional epidemiology and sequence data by estimating such parameters as the basic reproduction number, $R_0$, from topology and branch lengths of pathogen phylogenies (i.e.
-genealogies of the pathogen population, approximating the transmission trees) combined with metadata on the samples.
+It relies on models that bridge the gap between traditional epidemiology and sequence data by estimating such parameters as the basic reproduction number, $R_0$, from topology and branch lengths of pathogen phylogenies (i.e., genealogies of the pathogen population, approximating the transmission trees) combined with metadata on the samples.
 Classical epidemiological methods make inferences from incidence curves, which in the presence of import cases might be biased and give appearance of elevated local transmission.
@@
 Coalescent models are often preferred for estimating deterministic population dynamics, however for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\cite{Macpherson2021}.
-In the classic BD model with incomplete sampling~\cite{Stadler2009}, births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g.
-due to healing, self-isolation, starting a treatment, or death, modelled with a constant removal rate).
-
-The epidemic spread and detection can however be non-homogenous and depend on different factors, including governmental health policies (e.g.
-quarantine measures or development of pathogen detection tests), pathogen evolution over time (e.g.
-some SARS-CoV-2 variants are more transmissible than the others),  differences between host individuals (e.g.
-due to their immune system particularities or to their behaviour).
-
-To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extention of the classical BDS model was developed by Stadler \textit{et al.}~\cite{Stadler2013a}.
+In the classic BD model with incomplete sampling~\cite{Stadler2009}, births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g., due to healing, self-isolation, starting a treatment, or death, modelled with a constant removal rate).
+
+The epidemic spread and detection can however be non-homogenous and depend on different factors, including governmental health policies (e.g., quarantine measures or development of pathogen detection tests), pathogen evolution over time (e.g., some SARS-CoV-2 variants are more transmissible than the others),  differences between host individuals (e.g., due to their immune system particularities or to their behaviour).
+
+To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extention of the classical BD model was developed by Stadler \textit{et al.}~\cite{Stadler2013a}.
 MTBD framework allows for different types of individual states (and changes between them, modelled with state change rates).
-These models are phylodynamic analogies of compartmental models in classical epidemiology (e.g.
-SIR, Susceptible-Infectious-Removed).
-Examples of the MTBD family models include the birth-death exposed-infectious (BDEI) model, which was designed for pathogens featuring an incubation period between the moments of infection and of becoming infectious (e.g.
-Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen (e.g.
-HIV) more than the others~\cite{Stadler2014}.
+These models are phylodynamic analogies of compartmental models in classical epidemiology (e.g., SIR, Susceptible-Infectious-Removed).
+Examples of the MTBD family models include the birth-death exposed-infectious (BDEI) model, which was designed for pathogens featuring an incubation period between the moments of infection and of becoming infectious (e.g., Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen (e.g., HIV) more than the others~\cite{Stadler2014}.
 
 To allow for heterogeneity over time, Stadler \textit{et al.}~\cite{Stadler2013} developed one-state Bayesian birth-death skyline plot (BDSKY) that divides the time into intervals and allows for different piecewise constant rates on them.
-Kühnert \textit{et al.}~{Kuhnert2016} combined the MTBD model with the BDSKY to allow for both piecewise-constant rate changes over time and multiple individual types.
-In particular the skyline approach can be useful to account for changes in the sampling policies (e.g.
-before/after HIV discovery or before/after PCR test spread for SARS-CoV-2).
+Kühnert \textit{et al.}~\cite{Kuhnert2016} combined the MTBD model with the BDSKY to allow for both piecewise-constant rate changes over time and multiple individual types.
+In particular the skyline approach can be useful to account for changes in the sampling policies (e.g., before/after HIV discovery or before/after PCR test spread for SARS-CoV-2).
 
 Another important source of heterogeneity that none of these models accounts for is inflicted by non-independent sampling.
-We can propose different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability over time intervals with the BDSKY, however, none of them can account for correlation effects in sampling, in particular due to partner notification (contact tracing).
+One can set different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability between time intervals with the BDSKY, however, none of these models can account for correlation effects in sampling, in particular due to partner notification (contact tracing).
 Partner notification or contact tracing is a process in which contacts of detected infected individuals are identified and offered testing for infection.
@@
 For instance, for HIV,  partner notification showed increased early referral and initiation of treatment, and several countries include it in their national HIV testing services policies~\cite{worldhealthorganizationCountryPolicyReview2016}.
+
+\bigskip
 
@@
 For the simplest representative of this model family, the BD-PN model, we also derive the equations for its likelihood calculation and present a maximum likelihood parameter estimator.
-We test our BD-PN parameter estimator on simulated data, and apply it to the study of HIV-1 B epidemic in the UK.
+We test our BD-PN parameter estimator on simulated data, and apply it to the study of HIV-1 B epidemics in Zurich and in the UK.
+Importantly, we show that not accounting for partner notification when it is present leads to bias in parameter estimation with the BD model, while BD-PN parameter estimator performs well both in presence and in absence of partner notification.
 
@@
 In a pathogen transmission tree $\mathscr{T}$ (approximated by a time-scaled pathogen phylogeny) the tips represent sampled pathogens %, patient state transitions occur along the branches, 
-while bifurcations (i.e.
-internal nodes) correspond to pathogen transmissions (Fig.~\ref{fig:tt}).
-The tree branches are measured in units of time.
-Let $T$ denote the time that passed between the tree root (the beginning of the (sub-)epidemic, at time $t=0$) and the last sampled tip (at $t=T$).
-
-\begin{figure}[!h]
-\centering 
-%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
-\includestandalone[width=0.6\textwidth]{Fig_tree}
-\caption{\textbf{Transmission tree} $\mathscr{T}$ with $n=5$ external nodes (i.e.
-tips, which correspond to sampling events: $0000, 0001, 001, 010, 011$), $n-1=4$ internal nodes (which correspond to transmissions: $0$ (the root) and $00, 01, 000$) and $2n - 2 = 8$ branches (plus the root branch of zero length).
+while bifurcations (i.e., internal nodes) correspond to pathogen transmissions. % (Fig.~\ref{fig:tt}). 
+The tree branches are measured in units of time, and the time goes forward, from the tree root (representing the beginning of the (sub-)epidemic, at time $t=0$) to the time of the last sampled tip (at $t=T$).
+
+%\begin{figure}[!h]
+%\centering 
+%%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
+%\includestandalone[width=0.6\textwidth]{Fig_tree}
+%\caption{\textbf{Transmission tree} $\mathscr{T}$ with $n=5$ external nodes (i.e., tips, which correspond to sampling events: $0000, 0001, 001, 010, 011$), $n-1=4$ internal nodes (which correspond to transmissions: $0$ (the root) and $00, 01, 000$) and $2n - 2 = 8$ branches (plus the root branch of zero length).
 %The lengths of the branches are shown on the right of each branch: $t_i$ is the lengths of the branch connecting the node $i$ to its parent.
-Time $t$ starts at the root of the tree ($t=0$) and goes till the last sampled tip.
-The times of the nodes are shown on the left, e.g.
-$t_{0001}$ is the time of tip $0001$ (when $0001$'s pathogen was sampled).
+%Time $t$ starts at the root of the tree ($t=0$) and goes till the last sampled tip.
+The times of the nodes are shown on the left, e.g., $t_{0001}$ is the time of tip $0001$ (when $0001$'s pathogen was sampled).
 $T$ corresponds to the end of the sampling period (when the most recent tip, $0000$, was sampled).}
-\label{fig:tt} 
-\end{figure}
+%\label{fig:tt} 
+%\end{figure}
 
@@
 Under this model we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ (see \nameref{math}).
-The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g.
-by finding parameter values that maximize it, in maximum likelihood framework).
+The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g., by finding parameter values that maximize it, in maximum likelihood framework).
 In particular, it permits inference of the following epidemiological parameters:
@@
 
-In this section we propose an %non-Markovian 
+We propose an %non-Markovian 
 extension of the MTBD models that adds partner notification (PN).
 In the simplest case, the BD-PN model, at the moment of sampling the sampled individual might notify their most recent partner with a constant probability $\upsilon$.
-The notification implies that four conditions are met: (i) the sampled individual remembers their last partner and accepts to either notify the partner on their own or to give the partner's contacts to the health officials for assisted notification; (ii) a pathogen transmission happened at the moment of contact (either the sampled individual infected the partner or the other way round); (iii) the partner is successfully reached and notified; (iv) unless the partner already became uninfectious by other means by the moment of notification (e.g.
-got diagnosed though the standard procedure), they get tested upon notification.
+The notification implies that four conditions are met: (i) the sampled individual remembers their last partner and accepts to either notify the partner on their own or to give the partner's contacts to health-care providers for assisted notification; (ii) a pathogen transmission happened at the moment of contact (either the sampled individual infected the partner or the other way round); (iii) the partner is successfully reached and notified; (iv) unless the partner already became uninfectious by other means by the moment of notification (e.g., got diagnosed though the standard procedure), they get tested upon notification.
 The parameter $\upsilon$ hence regroups a complex set of conditions, which would hardly be identifiable separately from only sequence data.
@@
  \item $\upsilon$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
- \item $\phi >> \psi$ -- constant notified sampling rate: a notified individual (if not yet removed via standard procedure) gets removed \textit{and observed} with this rate.
+ \item $\phi >> \psi$ -- constant notified sampling rate: a notified individual (if not yet removed via the standard procedure) gets removed \textit{and observed} with this rate.
 This creates an observed tip in the transmission tree.
@@
 
-\item notified individuals will eventually get sampled (i.e., their pathogens are always observed upon removal, unlike the standard case where the observation happens with a sampling probability $\rho$);\\
-
-Realistically again, we assume here that the diagnosis and follow-up process works as expected, sampling notified patients.
+\item successfully notified individuals get sampled (i.e., their pathogens are always observed upon removal, unlike the standard case where the observation happens with a sampling probability $\rho$);\\
+
+Realistically again, we assume here that the diagnostic and follow-up process works as expected, sampling notified patients.
 
@@
 
-This assumption might need to be removed or relaxed (e.g.
-to up to a certain fixed number of most recent partners, or all the partners within a certain time interval) in the future work.
+This assumption might need to be removed or relaxed (e.g., to up to a certain fixed number of most recent partners, or all the partners within a certain time interval) in the future work.
 It is however useful to make the mathematics and likelihood calculation under the model manageable.
-In the case of sexually transmitted diseases, notifying only the most recent partner could roughly correspond to notifying the ``main'' partner (e.g.
-spouse).
+In the case of sexually transmitted diseases, notifying only the most recent partner could roughly correspond to notifying the ``main'' partner (e.g., spouse).
 
 Note that the ``most recent partner'' relationship is not necessarily symmetrical.
-The most recent partner $j$ of individual $i$ might have transmitted their virus further (to individual $k$) after the contact with $i$ and before the notification by $i$.
+The most recent partner $j$ of individual $i$ might have transmitted their virus further (to individual $k$) after the contact with $i$.
 Hence, while $j$ is the most recent partner of $i$, $j$'s most recent partner is $k$, and, once notified, $j$ might notify $k$.
@@
 It is Gillespie-based, generates state change, transmission and removal times, and only reconstructs the sampled parts of the tree to save memory and increase speed.
-We give its algorithmic details in~\nameref{S1_Appendix}. The results in the following section show that the trees generated under MTBD and MTBD-PN models are different and easily distinguishable.
+We give its algorithmic details in~\nameref{S1_Appendix}.
+The results in section~"\nameref{sec:test:sim}" show that the trees generated under MTBD and MTBD-PN models are different and easily distinguishable.
 
@@
 \subsection*{Testing for presence of Partner Notification}~\label{sec:test}
-To access whether a given tree reveals the presence of partner notification, we developed a non-parametric test based on cherries (two-tip clades).
+To access whether a given tree reveals partner notification, we developed a non-parametric test based on cherries (two-tip clades).
 
@@
 Being close in starting time is important as the tree is censored by the end of the sampling period: $t \leq T$.
-Hence the external branches that start close to the end of the sampling period would be on average shorter (as if they are not sampled before $T$ they will stay hidden) than those that start closer to the root~\cite{mooersBranchLengthsBirth2012}.
+Hence the external branches that start close to the end of the sampling period would be on average shorter (as unless sampled before $T$ they will stay hidden) than those that start closer to the root~\cite{mooersBranchLengthsBirth2012}.
 %We give control to the user on what should be considered ``some of the cherries'' and ``similar branch lengths'' for their data, via adjustable test parameter values.
@@
 We then generate a collection of reshuffled cherries by (1) ordering the cherries by their root dates; (2) for each cherry $i$ randomly selecting one of its tips $i* \in \{i0, i1\}$; (3) swapping the selected tips between the neighboring cherries: the tip $2i*$ gets swapped with the tip $2i-1*$ ($1 \leq i \leq N / 2$) as illustrated in Fig.~\ref{fig:tipswap}.
-If the total number of cherries N is odd, instead of swapping the selected tips of the last two cherries, we swap the selected tips of the last three cherries in a cycle (see Fig.~\ref{fig:tipswap}).
+(If the total number of cherries N is odd, instead of swapping the selected tips of the last two cherries, we swap the selected tips of the last three cherries in a cycle as in Fig.~\ref{fig:tipswap}).
 We then calculate the time differences $d_{i'}$ between the tips of these shuffles cherries.
-The test then compares $d_i$ to $d_{i'}$: $sign(d_{i'} - d_i) = 1$ if $d_{i'}$ is larger than $d_i$; $= -1$ if it is smaller; or $= 0$ if they are equal (this could happen as in real datasets dates are often truncated, e.g.
-to month or even a year).
+The test then compares $d_i$ to $d_{i'}$: $sign(d_{i'} - d_i) = 1$ if $d_{i'}$ is larger than $d_i$; $= -1$ if it is smaller; or $= 0$ if they are equal (this could happen as in real datasets dates are often truncated, e.g., to month or even a year).
 Note that if cherry tip branch lengths are independently drawn from the same distribution, $d_{i'}$ will be smaller than $d_i$ for about half of the cherries.
@@
 \textbf{a.} Cherries of the original tree sorted by their root times (times of the roots of the two oldest cherries, $t_{r1}$ and $t_{r2}$, are shown on the left).
-The randomly selected tip is shown with a bold branch for each cherry (e.g.
-the tip on the right for the leftmost cherry).
+The randomly selected tip is shown with a bold branch for each cherry (e.g., the tip on the right for the leftmost cherry).
 The tip sampling time differences are shown for the two leftmost cherries ($d_1$ and $d_2$).
@@
 
-We implemented the PN-test in Python 3. It is available as a part of BDPN parameter estimator (see section~"\nameref{sec:sim}").
+We implemented the PN-test in Python 3. It is available as a part of BDPN parameter estimator ($pn\_test$, see section~"\nameref{sec:sim}").
 
@@
 
-With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch $i$, starting on this branch at a time $t \leq t_i$ till time $t_i$.
+With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch $i$ till time $t_i$, starting on this branch at time $t \leq t_i$.
 The master equations for the BD models were initially developed by Stadler \textit{et al.}~\cite{Stadler2009}, however what we present here is their branch-specific formulation that we proposed in~\cite{zhukovaFastAccurateMaximumLikelihood2022}:
 In our formulation, ${p}^{(i)}(t)$ describes an evolution along a tree branch $i$, without taking into account the branch's subtree.
-$t_i$ corresponds to the time of the equation's initial condition (typically at the end of the branch).
+$t_i$ corresponds to the time of the equation's initial condition (typically at the end of the branch $i$).
 % and provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$.
@@
 Hence, we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at any time $t < t_i$ along the branch.
-These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.
+These two individuals could be different if a hidden transmission where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.
 $p^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along such a tree branch, integrating over all the possibilities, starting from no hidden transmission along the branch and including any number of hidden transmissions with either donor or recipient tree staying unsampled (and hence potentially different individuals at times $t$ and the branch end $t_i$).
@@
 %\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
-\includegraphics[width=1\textwidth]{Fig_PNbranches}
+\includegraphics[width=0.9\textwidth]{Fig_PNbranches}
 \caption{\textbf{Examples of partner notification in a transmission tree}.
-Let us assume that an individual A got sampled at time $t_A$ (violet tip) and notified their most recent partner B (orange), where B is either observed (\textbf{1}) or hidden (\textbf{2-3}).
+Let us assume that an individual A got sampled at time $t_A$ (violet tip) and notified their most recent partner B (orange), who is either observed (\textbf{1}) or hidden (\textbf{2-3}).
 The (observed as in \textbf{1-2} or hidden as in \textbf{3}) internal node AB corresponds to the transmission between A and B. 
-The fact that A is the notifier of B implies that there is no hidden event along its branch AB-A (coloured violet), and that in any transmission on the path AB-B (coloured orange) the donor is B.
-Note, that since the most-recent-partner relationship is not always symmetric, B might be the most recent partner of more than one individual and hence might have got notified by someone else too (e.g.
-C in \textbf{b}, who in \textbf{1b,2b} was infected by B, and in \textbf{3b} could have been infected by or have infected B, who have then infected A).
-Moreover, between the moment of B's first notification ($t_A$ for \textbf{a}, $t_C$ for \textbf{b}) and the moment of sampling  of B ($t_B$) the removal rate of B becomes $\phi$ instead of $\psi$ (coloured dark red).
+The fact that A is the notifier of B implies that there is no hidden event along its branch AB-A (violet), and that in any transmission on the path AB-B (orange) the donor is B. 
+Note, that since the most-recent-partner relationship is not always symmetric, B might be the most recent partner of more than one individual and hence might have got notified by someone else too (e.g., by C in \textbf{b}, who in \textbf{1b,2b} was infected by B, and in \textbf{3b} could have been infected by or have infected B, who have then infected A).
+Moreover, between the moment of B's first notification ($t_A$ for \textbf{a}, $t_C$ for \textbf{b}) and the moment of sampling  of B ($t_B$) the removal rate of B becomes $\phi$ instead of $\psi$ (dark red).
 Finally, B might have been removed via a standard procedure before getting notified by their partner(s) (with rate $\psi$ and a probability $\rho$, \textbf{c}).
-In the panels \textbf{3} B's subtree is fully hidden, and hence the transmission node AB is also unobserved, therefore the A's branch is composed of two parts, where the bottom part (violet) corresponds to A and contains no hidden transmissions, while the top part could correspond to A or not  (e.g in \textbf{3b} it corresponds to B) and might contain hidden transmissions.
-In the panels \textbf{2-3a-b} B is hidden as by the end of the sampling period ($T$) B is not yet sampled, while in \textbf{2-3c} B is hidden as B got removed via the standard procedure without sampling before any notification.}
+In the panels \textbf{3} B's subtree is fully hidden, and hence the transmission node AB is also unobserved.
+Therefore the A's branch is composed of two parts, where the bottom part (violet) corresponds to A and contains no hidden transmissions, while the top part could correspond to A or not  (e.g in \textbf{3b} it corresponds to B) and might contain hidden transmissions.
+In the panels \textbf{2a-b,3a-b} B is hidden as by the end of the sampling period ($t=T$) B is not yet sampled, while in \textbf{2-3c} B is hidden as B got removed via the standard procedure without sampling before any notification.
+\\
+A's branch is modeled with a notifier branch probability~(\ref{eq:p-n}) in panels \textbf{1-2}, and with a mixed branch probability~(\ref{eq:p_mixed}) where the bottom part corresponds to the notifier (A) in panels \textbf{3}.
+If B did not notify their most recent partner upon sampling in panel \textbf{1}, branches on the path AB-B are modeled with a combination of partner probability before notification (orange, (\ref{eq:p-pb})) and partner probability after notification (dark red, (\ref{eq:p-pa})).
+Otherwise, the part after the transmission from B to their last partner is modeled with either a both notified and notifier branch probability~(\ref{eq:p-b}) (as in \textbf{1a-b}) or a notifier branch probability~(\ref{eq:p-n}) (as in \textbf{1c}), depending on the B's notified status at that time.
+In the panels \textbf{2}, the AB-C branch is modeled with a mixed branch probability~(\ref{eq:p_mixed}), where the top part corresponds to the partner state (B).
+}
 \label{fig:pn-branches} 
@@
 These transmissions can be either hidden (where the recipient tree stayed unobserved) or observed: in both cases we know which subtree corresponds to the donor (the one that contains B).
-However, depending on whether B already got notified ($t \geq t_A$ in Fig.~\ref{fig:pn-branches}(1a)) or not yet (($t < t_A$ in Fig.~\ref{fig:pn-branches}(1a))) B's removal rate is different (is correspondingly $\phi$ or $\psi$).
+However, depending on whether B already got notified ($t \geq t_A$ in Fig.~\ref{fig:pn-branches}(1a)) or not yet ($t < t_A$ in Fig.~\ref{fig:pn-branches}(1a)), B's removal rate is different (is correspondingly $\phi$ or $\psi$).
 We will use a \textbf{partner branch probability before notification} $p_{pb}^{(x)}(t)$ for cases when the individual at the end of the branch (or considered time interval, here $t_x=t_A$) is the same as at time $t$, has not yet been notified, but might have transmitted to someone else (unobserved) along the way (coloured orange in Fig~\ref{fig:pn-branches}):
@@
     &~~~ - \lambda p_{pb}^{(x)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission with hidden recipient subtree}\\
-p_{pb}^{(x)}(t_x) &=  1\label{eq:p-o}
+p_{pb}^{(x)}(t_x) &=  1\label{eq:p-pb}
 \end{cases}
@@
     &~~~ - \lambda p_{pa}^{(B)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission with hidden recipient subtree}\\
-p_{pa}^{(B)}(t_B) &=  1\label{eq:p-p}
+p_{pa}^{(B)}(t_B) &=  1\label{eq:p-pa}
 \end{cases}
@@
 
-%On the branch of B, no transmission is possible between the moment of notification $t_A$ and the moment of B's sampling $t_B$ as, according to assumption 3, once notified partners do not transmit further.
-We will use a \textbf{notified partner branch probability} $p_{pa}^{(i)}(t)$ for cases when the individual at the end of the branch is the partner who got notified at time prior to $t$ (coloured dark red in Fig~\ref{fig:pn-branches}):
-%
-%\begin{equation}
-%\scriptsize
-%\begin{cases}
-%\dot{p}_p^{(B)}(t) &=  \phi p_{pa}^{(B)}(t)\; \textit{\color{gray} $\leftarrow$ no sampling in the next infinitesimal time $\Delta t$ }\\
-%p_{pa}^{(B)}(t_B) &=  1\label{eq:p-np}
-%\end{cases}
-%\end{equation}
+The equations~\ref{eq:Us}, \ref{eq:p}, \ref{eq:p-n}-\ref{eq:p-b} have the following closed form solutions:
+\begin{equation}
+\scriptsize
+\begin{split}
+&\begin{cases}
+U(t) = \frac{\lambda + \psi}{2\lambda} +  \frac{c_1}{2\lambda}\Big(\frac{E(t) - 1}{E(t) + 1}\Big)\\
+\\
+p^{(i)}(t) = \Big(\frac{E(t_i) + 1}{E(t) + 1}\Big)^2 e^{c_1 (t -t_i)} \\
+\\
+p_n^{(i)}(t) =  e^{(\lambda + \psi) (t -t_i)}\\
+\\
+p_b^{(i)}(t) =  e^{(\lambda + \phi) (t -t_i)}
+\\
+\\
+p_{pb}^{(i)}(t) =  \sqrt{p^{(i)}(t)p_n^{(i)}(t)} \\
+\\
+p_{pa}^{(i)}(t) =  e^{(\phi-\psi) (t -t_i)} p_{pb}^{(i)}(t) 
+\end{cases},\\
+& \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi}, E(t) = c_2 e^{c_1 (t -T)} \\
+\end{split}\label{eq:ps}
+\end{equation}
 
@@
 
-\item[1b.] by the time $t_A$, B was already notified by someone else (at time $t_C < t_A$, see Fig.~\ref{fig:pn-branches}(1b)).
-%Note, that once notified partners do not transmit further, hence $t_C > t_{AB}$. 
-\\
-
-This would lead to a shorter partner-before-notification part on the B branch (finishing at $t_C$), and a longer notified-partner part (between $t_C$ and $t_B$).
+\item[1b.] by the time $t_A$, B was already notified by someone else (e.g., at time $t_C < t_A$ in  Fig.~\ref{fig:pn-branches}(1b)).
+
+This would lead to a shorter partner-before-notification part on the B's branch (finishing at $t_C$), and a longer notified-partner part (between $t_C$ and $t_B$).
 
@@
 
-The equations~\ref{eq:Us}, \ref{eq:p}, \ref{eq:p-n}-\ref{eq:p-b} have the following closed form solutions:
-\begin{equation}
-\scriptsize
-\begin{split}
-&\begin{cases}
-U(t) = \frac{\lambda + \psi}{2\lambda} +  \frac{c_1}{2\lambda}\Big(\frac{E(t) - 1}{E(t) + 1}\Big)\\
-\\
-p^{(i)}(t) = \Big(\frac{E(t_i) + 1}{E(t) + 1}\Big)^2 e^{c_1 (t -t_i)} \\
-\\
-p_n^{(i)}(t) =  e^{(\lambda + \psi) (t -t_i)}\\
-\\
-p_b^{(i)}(t) =  e^{(\lambda + \phi) (t -t_i)}
-\\
-\\
-p_{pb}^{(i)}(t) =  \sqrt{p^{(i)}(t)p_n^{(i)}(t)} \\
-\\
-p_{pa}^{(i)}(t) =  e^{(\phi-\psi) (t -t_i)} p_{pb}^{(i)}(t) 
-\end{cases},\\
-& \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi},\\
-& ~~~~~~~~E(t) = c_2 e^{c_1 (t -T)} \\
-\end{split}\label{eq:ps}
-\end{equation}
-
 Note that if A is a notifier, A is always observed, as notification only happens upon sampling.
 However, B could have stay unsampled: either because once notified (by A or another partner, who notified them before A did) B has not yet got sampled (Fig~\ref{fig:pn-branches}(2a-b,3a-b)); or because B got removed via the standard procedure without sampling before any notification (Fig.~\ref{fig:pn-branches}(2c,3c)).
-Moreover, between the transmission AB and the moment of B's notification/removal B might have transmitted to someone else (e.g.
-C in Fig.~\ref{fig:pn-branches}(2), BC denoting the transmission).
+Moreover, between the transmission AB and the moment of B's notification/removal, B might have transmitted to someone else (e.g., C in Fig.~\ref{fig:pn-branches}(2), BC denoting the transmission).
 If that happened and someone in BC's subtree got sampled, the branch AB-A would still correspond to the notifier A, while the other AB's child branch would partially correspond to B (between AB and BC) and partially to C (starting at BC), as in Fig.~\ref{fig:pn-branches}(2).
 Otherwise (if there was no such transmission or all such subtrees stayed unobserved, as in Fig.~\ref{fig:pn-branches}(3)), the node AB will not be observed either, and A's branch will get extended with its parent's branch.
-Hence only the bottom part (after AB) of the A's branch would correspond to the notifier probability (see Fig.~\ref{fig:pn-branches}(3))).
+Hence only the bottom part (after AB) of the A's branch would correspond to the notifier probability (see Fig.~\ref{fig:pn-branches}(3)).
 
 To model the cases when B is not sampled, we will need to describe the probability of a\textbf{ mixed branch} $p_{m(s1,s2)}^{(i)}(t,t_r)$ for branches that include a hidden partner subtree (AB-C branches in Fig.~\ref{fig:pn-branches}(2) and A's tip branch in Fig.~\ref{fig:pn-branches}(3)).
-$s1$ here denotes the type of the top of the branch (e.g.
-between AB and BC in Fig.~\ref{fig:pn-branches}(2), partner) and $s2$ denotes the type of the bottom of the branch (e.g.
-between AB and A in Fig.~\ref{fig:pn-branches}(3), notifier).
-Let's denote the first notification time of the hidden partner as $t_r$, and the time of the hidden partner subtree start as $t_h$, while as usual $t_i$ stands for the time at the branch end.
+$s1$ here denotes the type of the top of the branch (e.g., between AB and BC in Fig.~\ref{fig:pn-branches}(2), partner) and $s2$ denotes the type of the bottom of the branch (e.g., between AB and A in Fig.~\ref{fig:pn-branches}(3), notifier).
+Let us denote the first notification time of the hidden partner as $t_r$, and the time of the hidden partner subtree start as $t_h$, while as usual $t_i$ stands for the time at the branch end.
 For instance, in Fig~\ref{fig:pn-branches}(2a) $t_i=t_C$, $t_r=t_A$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(2b) $t_i=t_r=t_C$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(3c) $t_i=t_r=t_A$, $t_h=t_{AB}$.
@@
 
-In the real trees we do not know sampled individual's status (individual who did not notify, notifier, partner notified in time or not), and hence during the tree likelihood calculation, we will have to consider all these scenarios.
+In real trees we do not know sampled individual's status (individual who did not notify, notifier, partner notified in time or not), and hence during the tree likelihood calculation, we will have to consider all these scenarios.
 
@@
 
-Using the formulas~(\ref{eq:ps}), (\ref{eq:p_mixed}) and (\ref{eq:p_mixed_mixed}), we can calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}.
+Using the formulas~(\ref{eq:ps}) and (\ref{eq:p_mixed}), we can calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}.
 For each visited node $i$ we calculate the likelihood density of its subtree including the branch $ji$ connecting $i$ to its parent node $j$ for two cases depending on $i$'s receiving-a-notification status (shown with the first superscript, $^{\not{n}}$ or $^n$).
@@
 This extension applies also to the BD-PN model.
-Forests are useful for cases when a (sub-)epidemic started with several infected individuals (e.g.
-due to multiple pathogen introductions to a country of interest or due to a change of health policies leading to a change in parameter values).
-
+Forests are useful for cases when a (sub-)epidemic started with several infected individuals (e.g., due to multiple pathogen introductions to a country of interest or due to a change of health policies leading to a change in parameter values).
 In this case the (sub-)epidemic leads to a forest $\mathscr{F}$ of $f$ observed trees: $\mathscr{T}_1, \ldots, \mathscr{T}_f$.
@@
 
-%As our application to Ebola data (see~Application) shows, $u$ is an important parameter: while the results are not very sensible to slight $u$ variations, ignoring it completely (e.g.
-using $u=0$ instead of $u \approx 500$ in our example) can change the inferences.
+%As our application to Ebola data (see~Application) shows, $u$ is an important parameter: while the results are not very sensible to slight $u$ variations, ignoring it completely (e.g., using $u=0$ instead of $u \approx 500$ in our example) can change the inferences.
 
@@
 Once the optimal parameter values are found, we calculate their confidence intervals (CIs) using Wilks' method~\cite{Wilks1938}.
-For each non-fixed parameter $p \in \Theta$, we calculate its $95\%$-CI as including the values $\tilde{p}$ such that $log L(\mathscr{F}|\Theta_{opt|p=\tilde{p}}) > log L(\mathscr{F}| \Theta_{opt}) - \chi^2_1(0.95) / 2$, where $\chi^2_1(0.95)$ is the value of chi-squared distribution with 1 degree of freedom corresponding to the significance level of $0.95$ (i.e., $\sim3.84$).
-$\Theta_{opt|p=\tilde{p}}$ corresponds to the maximum-likelihood value for the other non-fixed parameters when $p = \tilde{p}$.
+For each non-fixed parameter $p \in \Theta$, we calculate its $95\%$-CI as $[\tilde{p} \in \mathbb{R}\; |\; log L(\mathscr{F}|\Theta_{opt|p=\tilde{p}}) > log L(\mathscr{F}| \Theta_{opt}) - \frac{\chi^2_1(0.95)}{2}]$, where $\chi^2_1(0.95) \approx 3.84$ is the value of chi-squared distribution with 1 degree of freedom corresponding to the significance level of $0.95$, while $\Theta_{opt|p=\tilde{p}}$ corresponds to the maximum-likelihood value for the other non-fixed parameters when $p = \tilde{p}$.
 
@@
 
+\subsection*{UK HIV-1 B tree reconstruction and sampling probability estimation}\label{uk_tree}
+To asses the partner notification in HIV-infected patients in the UK, we used the phylogenetic tree from our recent study of HIV drug resistance in the UK~\cite{zhukovaModelingDrugResistance2023}.
+The tree tips represent viruses sampled from 40\,055 individuals between 1996 and 2016.
+The samples used for the tree reconstruction were obtained from the UK HIV Drug resistance database~\cite{Dunn2007}.
+The sampling date metadata for these samples were month-specific (e.g., Aug 2015), hence potentially creating simultaneous sampling dates for cases, which were sampled on different days of the month.
+To avoid this issue we randomly selected a date within the specified month for each sample, and time-scaled the tree with LSD2~\cite{To2016} (v2.4.1, strict molecular clock with outlier removal).
+We repeated this time-scaling procedure 10 times, obtaining 10 trees with slightly different branch lengths and root dates between Oct 5 and Nov 11, 1962.
+
+Between 2012 and 2015, in the UK ART of HIV-infected individuals  was initiated when their CD4 count dropped below 350 cells/mL~\cite{williamsBritishHIVAssociation2012}.
+However ``if a patient with a CD4 cell count $>$ 350 cells/mL wishes to start ART to reduce the risk of transmission to partners, this decision is respected and ART is started''~\cite{williamsBritishHIVAssociation2012}.
+Before 2012, treatment started with an even lower CD4 (i.e., later).
+The British HIV Association 2015 guidelines recommend all individuals with suspected or diagnosed primary HIV infection ``are offered immediate ART''~\cite{churchillBritishHIVAssociation2016}.
+To have a more homogeneous setting in terms of access to ART, awareness of the HIV infection, etc., we cut the dated trees between 2012 and 2015.
+
+To estimate the BD-PN parameter values, we needed to fix one of the parameters for model identifiability.
+We therefore estimated the base sampling probability $\rho$ as follows.
+According to the \href{https://webarchive.nationalarchives.gov.uk/ukgwa/20181112132123mp_/https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/602942/HIV_in_the_UK_report.pdf}{``Towards elimination of HIV transmission, AIDS and HIV-related deaths in the UK''}
+ report, the estimated total number of people living with HIV in the UK in 2015 was 101\,200.
+% (97\,500 to 105\,700).
+% in 2012 this number was 98\,400 (93\,500-104\,300) [\href{HIV in the United Kingdom: 2013 Report}{https://webarchive.nationalarchives.gov.uk/ukgwa/20181112133715mp_/https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/326601/HIV_annual_report_2013.pdf}].
+Moreover these document report about 400 HIV-related deaths per year.
+According to our estimate~\cite{zhukovaModelingDrugResistance2023}, 66.5\% of the HIV-infected individuals in the UK were infected with HIV-B.
+We hence estimated that in 2015 67\,298 % (64\,837, 70\,291)
+people were living with HIV-B. Our trees contained 39\,047 tips sampled by the end of 2015, and therefore account for about 58\% %(56\%, 60\%) 
+of the total number of individuals living with HIV-B in the UK.
+We hence fixed the sampling probability $\rho$ to 0.58.
+%However this estimate does not take into account deaths of the HIV-infected individuals over the years, which would increase the total number of HIV-infected individuals by 2015, and decrease the proportion of individuals represented by our tree.
+We therefore tested several sampling probabilities $\rho=$0.6; 0.5; 0.4.
+
 % Results and Discussion can be combined.
@@
 
-For 98/97/95 out of 100 trees in the BD/BDEI/BDSS dataset the PN test p-value was above $0.05$ (i.e.
-98\%/97\%/95\% true negative and 2\%/3\%/5\% false positive results).
-For 95/95/96 out of 100 trees in the BD-PN/BDEI-PN/BDSS-PN dataset the PN test p-value was below $0.05$ (i.e.
-95\%/95\%/96\% true positive and 5\%/5\%/4\% false negative results).
+For 98/97/95 out of 100 trees in the BD/BDEI/BDSS dataset the PN test p-value was above $0.05$ (i.e., 98\%/97\%/95\% true negative and 2\%/3\%/5\% false positive results).
+For 95/95/96 out of 100 trees in the BD-PN/BDEI-PN/BDSS-PN dataset the PN test p-value was below $0.05$ (i.e., 95\%/95\%/96\% true positive and 5\%/5\%/4\% false negative results).
 These results are summarised in Table~\ref{tbl:pntest}.
@@
 
-We assessed the performance of our maximum-likelihood estimator for the BD-PN model on the 100 trees from the the BD-PN simulated data set (see~"\nameref{sec:sim}" for dataset details).
+We assessed the performance of our maximum-likelihood estimator for the BD-PN model bdpn on the 100 trees from the the BD-PN simulated data set (see~"\nameref{sec:sim}" for dataset details).
 
 We applied bdpn to each of the trees fixing the sampling probability $\rho$ to its real value (for identifiability).
-The results with the becoming-uninfectious rate $\psi$ fixed are similar and given in the Supplementary Material.
-As expected with a maximum likelihood estimator, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the real parameter values in both settings.
+%The results with the becoming-uninfectious rate $\psi$ fixed are similar and given in the Supplementary Material.
+As expected with a maximum likelihood estimator, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the real parameter values for all the trees in both settings.
 
 To highlight the importance of accounting for partner notification with the model, we also applied the classical BD model estimator bd to these trees.
-Additionally, we applied both BD-PN and BD model parameter estimators to the dataset simulated under BD model.
+Additionally, we applied both bdpn and bd estimators to the dataset simulated under the BD model.
 The results are shown in  Fig.~\ref{fig:sim}.
@@
 
-The percentage of the trees for which real parameter values are with the CI estimates and relative CI widths ($\frac{CI_{97.5\%} - CI_{2.5\%}}{\textit{real parameter value}}$) are shown in Table~\ref{tbl:ci}.
+The percentage of the trees for which real parameter values were with the CI estimates and relative CI widths ($\frac{CI_{97.5\%} - CI_{2.5\%}}{\textit{real parameter value}}$) are shown in Table~\ref{tbl:ci}.
 They suggest that the rate parameters are easier to estimate than the probability ones.
@@
 \small
-\caption{Percentage of simulated BD-PN trees for which the real parameter values were withing the bdpn-estimated 95\% CIs, and the median CI width: $\frac{CI_{97.5\%} - CI_{2.5\%}}{\textit{real parameter value}}$.
-\smallskip}
-\begin{tabular}{c|cc|cc}
-\textbf{parameter} & \textbf{$\psi$ fixed} & & \textbf{$\rho$ fixed} &  \\
- & \% within CI & median CI width & \% within CI & median CI width  \\
+\caption{Percentage of simulated BD-PN trees for which the real parameter values were withing the bdpn-estimated 95\% CIs, and the median CI width: $\frac{CI_{97.5\%} - CI_{2.5\%}}{\textit{real parameter value}}$  \smallskip}
+\begin{tabular}{c|cc}
+\textbf{parameter} & \% within CI & median CI width  \\
 \toprule 
- $\lambda$ &  91\% & 14.2\% & 96\% & 11.6\%\\
- $\psi$ & -- & -- & 88\% & 28.9\% \\
- $\rho$ & 78\%  & 38.9\% & -- & -- \\
- $\phi$ & 93\% & 32.3\% & 95\% & 32.2\% \\
- $\upsilon$ & 74\% & 21.8\% & 70\% & 21.7\% \\
+ $\lambda$ & 96\% & 11.6\%\\
+ $\psi$ & 88\% & 28.9\% \\
+ $\rho$ & fixed & -- \\
+ $\phi$ & 95\% & 32.2\% \\
+ $\upsilon$ & 70\% & 21.7\% \\
 \bottomrule
@@
 
-\subsubsection*{Application 1: HIV-1 B epidemic in the UK}
-We applied our estimator to asses the partner notification in HIV-infected patients in the UK.
-We used the phylogenetic tree from our recent study of HIV drug resistance in the UK~\cite{zhukovaModelingDrugResistance2023}.
-The tree tips represent viruses sampled from 40\,055 individuals between 1996 and 2016.
-The samples used for the tree reconstruction were obtained from the UK HIV Drug resistance database~\cite{Dunn2007}.
-The sampling date metadata for these samples were month-specific (e.g.
-Aug 2015), hence potentially creating simultaneous sampling dates for cases, which were sampled on different days of the month.
-To avoid this issue we randomly selected a date within the specified month for each sample, and time-scaled the tree with LSD2~\cite{To2016} (v2.4.1, strict molecular clock with outlier removal).
-We repeated this time-scaling procedure 10 times, obtaining 10 trees with slightly different branch lengths and root dates between Oct 5 and Nov 11, 1962.
-
-Between 2012 and 2015, in the UK antiretroviral treatment (ART) of HIV-infected individuals  was initiated when their CD4 count dropped below 350 cells/mL~\cite{williamsBritishHIVAssociation2012}.
-However ``if a patient with a CD4 cell count $>$ 350 cells/mL wishes to start ART to reduce the risk of transmission to partners, this decision is respected and ART is started''~\cite{williamsBritishHIVAssociation2012}.
-Before 2012, treatment started with an even lower CD4 (i.e.
-later).
-The British HIV Association 2015 guidelines recommend all individuals with suspected or diagnosed primary HIV infection ``are offered immediate ART''~\cite{churchillBritishHIVAssociation2016}.
-To have a more homogeneous setting in terms of access to ART, awareness of the HIV infection, etc., we cut the dated trees between 2012 and 2015.
+\subsubsection*{Application 1: Men-having-sex-with-men (MSM) HIV-1 B epidemic in Zurich}
+We applied our estimator to asses the partner notification for HIV-1 subtype B epidemic among men-having-sex-with-men (MSM) in Zurich.
+We used the phylogenetic tree of 200 samples reconstructed by Rasmussen \textit{et al.}~\cite{Rasmussen2017} from the sequences collected as a part of the Swiss Cohort Study~\cite{swisshivcohortstudyCohortProfileSwiss2010a} between 1988 and 2014.
+This tree was also previously analyzed by Voznica~\textit{et al.}~\cite{Voznica2021} using a Birth-Death model with SuperSpreading (BDSS).
+
+The PN test detected partner notification in this tree (p-value=0.046).
+Following Voznica~\textit{et al.}~\cite{Voznica2021}, we fixed the sampling probability for this tree at 25\%: $\rho=0.25$.
+Our bdpn estimator inferred the partner notification probability $\upsilon=0.07$ CI $[0.006-0.190]$,
+$R_e = 1.4$ $[1.2-1.6]$, infectious time $\frac{1}{\psi} = 8$ $[6-10]$ years, and 3.4 months [5 days -- 15 months] between partner notification and their sampling.
+These results globally agree (CIs intersect) with the $R_e$ estimate using a coalescent model from Rasmussen \textit{et al.}~\cite{Rasmussen2017} (between 1.0 and 2.5), and the estimates using the BDSS model from Voznica~\textit{et al.}~\cite{Voznica2021}: $R_e$ of 1.6 and 1.7 and an infectious period of 10.2 and 9.8 years (depending on the tree representation used for the deep learner).
+
+Note that this tree covers a long non-homogeneous time interval (between 1974, the root of the tree, and 2012), including times with no or limited access to treatment, when the infectious time corresponded to progression to AIDS ($\approx 10$ years), as well as later times, when antiretroviral treatment (ART) became available, and hence the infectious period would correspond to the time till ART start, as, when taken properly, it prevents transmission.
+Our estimate (8 years) represent a mix of these scenarios.
+This also holds for the other estimated parameters, whose values represent a somewhat averaged estimate over the time period.
+As this data set was already small, we did not cut it to obtain a more homogeneous treatment policy (as we did in Application 2).
+
+\subsubsection*{Application 2: HIV-1 B epidemic in the UK}
+We also applied our estimator to asses the partner notification in HIV-infected individuals in the UK.
+We used the phylogenetic tree from our recent study of HIV drug resistance in the UK~\cite{zhukovaModelingDrugResistance2023}, reconstructed from HIV-1 B samples of 40\,055 individuals sampled between 1996 and 2016 and obtained from the UK HIV Drug resistance database~\cite{Dunn2007}.
+
+The sampling dates in the UK HIV Drug resistance database are month-specific, hence to avoid potentially creating simultaneous sampling dates for cases, which were sampled on different days of the month, we randomly selected a date within the specified month for each sample before time-scaling the tree.
+We repeated the time-scaling procedure 10 times, obtaining 10 trees with slightly different branch lengths and root dates between Oct 5 and Nov 11, 1962 (see~"\nameref{uk_tree}").
+We cut this tree between 2012 and 2015, as this period featured a homogeneous ART start policy in the UK (CD4 cell count below 350 cells/mL~\cite{williamsBritishHIVAssociation2012}).
 We then applied the PN-test and estimated the BD-PN parameters on these 2012--2015 forests.
@@
 The PN-test values were below 0.05 for all the forests (values between $7 \cdot 10^{-18}$ and $5 \cdot 10^{-12}$), strongly suggesting the presence of partner notification.
-
-To estimate the BD-PN parameter values, we needed to fix one of the parameters for model identifiability.
-We therefore estimated the base sampling probability $\rho$ as follows.
-According to the \href{https://webarchive.nationalarchives.gov.uk/ukgwa/20181112132123mp_/https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/602942/HIV_in_the_UK_report.pdf}{``Towards elimination of HIV transmission, AIDS and HIV-related deaths in the UK''}
- report, the estimated total number of people living with HIV in the UK in 2015 was 101\,200 (97\,500 to 105\,700).
-% in 2012 this number was 98\,400 (93\,500-104\,300) [\href{HIV in the United Kingdom: 2013 Report}{https://webarchive.nationalarchives.gov.uk/ukgwa/20181112133715mp_/https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/326601/HIV_annual_report_2013.pdf}].
-Moreover these document report about 400 HIV-related deaths per year.
-According to our estimate~\cite{zhukovaModelingDrugResistance2023}, 66.5\% of the HIV-infected individuals in the UK were infected with HIV-B.
-We hence estimated that in 2015 67\,298 (64\,837, 70\,291) people were living with HIV-B.
-Our trees contained 39\,047 tips sampled by the end of 2015, and therefore account for about 58\% (56\%, 60\%) of the total number of individuals living with HIV-B in the UK.
-We hence fixed the sampling probability $\rho$ to 0.58.
-%However this estimate does not take into account deaths of the HIV-infected individuals over the years, which would increase the total number of HIV-infected individuals by 2015, and decrease the proportion of individuals represented by our tree.
-We therefore tested several sampling probabilities $\rho=$0.6; 0.5; 0.4.
-
-Using $\rho=0.58$, we estimated the partner notification probability $\upsilon=0.006-0.008$, %the following parameter values: $\lambda=0.53$, $\psi=0.43$, $\phi=23-59$, $\upsilon=0.006-0.007$, which corresponds to
+Using sampling probability $\rho=0.58$ (see~"\nameref{uk_tree}" for its estimation), we estimated the partner notification probability $\upsilon=0.006-0.008$, %the following parameter values: $\lambda=0.53$, $\psi=0.43$, $\phi=23-59$, $\upsilon=0.006-0.007$, which corresponds to
 $R_e = 1.2$, infectious time $\frac{1}{\psi} = 2.3$ years, and 15--16 days between partner notification and their sampling.
-The results are consistent between 10 forests and shown in Table~\ref{tbl:uk}.
+The results were consistent between 10 forests and shown in Table~\ref{tbl:uk}.
 %The results for $\rho=0.4$ and for $\rho=0.6$ were similar: $R_e$ of respectively $1.5$ and
@@
 
-The average time of HIV progression to AIDs in the absence of treatment is 10 years, which could serve as an estimate of the infectious time.
-However, in the UK, 95\% of infected population is on antiretroviral treatment, which, when taken properly, prevents transmission.
+In the UK 95\% of infected population is on antiretroviral treatment, which, when taken properly, prevents transmission.
 Hence, the average infectious time represents the average time before starting treatment.
 Our estimate is approximately 2.3 years.
-
-For comparison, parameter estimation on the same forests with the BD model (with $\rho$ fixed to $0.58$) gave unrealistic results: $R_e = 0.5$ (suggesting contained epidemic) and very short infectious time $\frac{1}{\psi} = 0.4$ years.
-These results were robust across the 10 forests, see \nameref{S1_Table}.
-As with the results on simulated data, this  highlights the importance of accounting for partner notification during parameter estimation.
 
@@
 
-\subsubsection*{Application 2: Men-having-sex-with-men (MSM) HIV-1 B epidemic in Zurich}
-For comparison with the UK case, we applied our estimator to asses the partner notification for HIV-1 subtype B epidemic in a different setting: in men-having-sex-with-men (MSM) community from Zurich.
-We used the phylogenetic tree including 200 samples from the largest MSM HIV cluster in Zurich reconstructed by Rasmussen \textit{et al.}~\cite{Rasmussen2017} from the sequences collected as a part of the Swiss Cohort Study~\cite{swisshivcohortstudyCohortProfileSwiss2010a} between 1988 and 2014.
-As this data set was already small, we did not cut it to obtain a more homogeneous treatment policy, as we did for the UK.
-
-The PN test detected partner notification in this tree (p-value=0.046).
-Following Voznica~\textit{et al.}~\cite{Voznica2021}, we fixed the sampling probability for this tree at 25\%: $\rho=0.25$.
-Our bdpn estimator inferred the partner notification probability $\upsilon=0.07$ (CI $[0.006-0.190]$),
-$R_e = 1.4$ (CI $[1.2-1.6]$), infectious time $\frac{1}{\psi} = 8$ (CI $[6-10]$) years, and 3.4 months (CI [5 days -- 15 months]) between partner notification and their sampling.
-
-Overall the estimates on this much smaller dataset show larger CIs than those on the larger UK data set.
-The CIs estimated on the two data sets intersect for the effective reproductive number $R_e$, while the infectious time is larger for the Zurich data set (CI of $6-10$ years vs $2.2-2.4$ years for the UK data set).
-This is however expected as the Zurich data set covers a longer, earlier and non-homogeneous time interval (between 1974, the root of the tree, and 2012), including times with no or limited access to treatment, when the infectious time corresponded to progression to AIDS ($\approx 10$ years).
-The CIs for notification probability $\upsilon$ and partner removal time $\frac{1}{\phi}$ also intersect on the two data sets, though they do not have to, as the notification policy might differ between the two countries, and especially during the long and non-homogeneous time interval represented by Zurich data.
+For comparison, parameter estimation on the same forests with the BD model (with $\rho$ fixed to $0.58$) gave unrealistic results: $R_e = 0.5$ (suggesting contained epidemic) and very short infectious time $\frac{1}{\psi} = 0.4$ years.
+These results were robust across the 10 forests, see \nameref{S1_Table}.
+As with the results on simulated data, this  highlights the importance of accounting for partner notification during parameter estimation.
+
+\bigskip
+
+Overall the estimates on the much smaller Zurich dataset show larger CIs than those on the larger UK one.
+The CIs estimated on the two data sets intersect for the effective reproductive number $R_e$, while the infectious time is larger for Zurich (CI of $6-10$ years vs $2.2-2.4$ years for the UK).
+This is however expected as the Zurich tree covers a longer, earlier and non-homogeneous time interval (between 1974 and 2012), including times with no or limited access to treatment, when the infectious time corresponded to progression to AIDS ($\approx 10$ years).
+The CIs for notification probability $\upsilon$ and partner removal time $\frac{1}{\phi}$ also intersect on the two data sets, though they do not have to, as the notification policy might differ between the two countries, and especially during the long and non-homogeneous time interval represented by the Zurich tree.
 The point estimates suggest stronger partner notification probability and longer notified partner removal time in Zurich than in the UK.
@@
 We proposed an extension of the phylodynamic birth-death model that accounts for non-random sampling due to partner notification (BD-PN).
-In real-life epidemics, detected infected individuals might notify the people whom they think they might have infected or got infected by, and who in turn might get tested and detected in a much faster and non-random way.
+In real-life epidemics, detected infected individuals might notify the people whom they think they might have infected or got infected by, and who in turn might get tested and detected much faster than via the standard procedure.
 Our model accounts for this scenario.
@@
 We developed a mathematical framework for calculation of the likelihood density of a transmission tree (approximated by a time-scaled phylogenetic tree) under the BD-PN model, and implemented a maximum-likelihood BD-PN parameter estimator bdpn based on this framework.
-The advantage of the BD-PN model is that the differential equations describing it have closed-form solutions, which are quick to calculate and do not require numerical approximation (as is the case for more complex MTBD models).
+%The advantage of the BD-PN model is that the differential equations describing it have closed-form solutions, which are quick to calculate and do not require numerical approximation (as is the case for more complex MTBD models).
 Tests on simulated data show that our estimator robustly infers model parameter values and their CIs in the setting when the duration of the infectious period $\frac{1}{\psi}$ or the base sampling probability $\rho$ is fixed.
@@
 
-We applied the PN-test and our estimator to the HIV-1 B epidemic in the UK, between 2012 and 2015 and to the HIV-1 B MSM cluster in Zurich.
+We applied the PN-test and our estimator to the HIV-1 B MSM cluster in Zurich and to the HIV-1 B epidemic in the UK between 2012 and 2015.
 The test detected presence of partner notification in both cases.
-The obtained estimates of epidemiological parameters are in agreement with what we now know about the HIV-1 B epidemic (e.g., $R_e$ between $1.2$ and $1.6$ ).
-We estimated a modest partner notification probability ($1\%$ in the UK, $7\%$ in Zurich), and rapid notified partner detection ($\approx 16$ days vs $\approx 2.3$ years for detection under the standard random procedure in the UK; and $\approx 3.4$ months vs $\approx  8$ years for detection under the standard random procedure in Zurich).
+The obtained estimates of epidemiological parameters are in agreement with what we now know about the HIV-1 B epidemic (e.g., $R_e$ between $1.2$ and $1.6$).
+We estimated a modest partner notification probability ($7\%$ in Zurich, $1\%$ in the UK), and rapid notified partner detection ($\approx 3.4$ months vs $\approx  8$ years for detection under the standard procedure in Zurich; and $\approx 16$ days vs $\approx 2.3$ years for detection under the standard procedure in the UK).
+
+\bigskip
 
@@
 The simulator iterates through occurring events till either time or sampled tip number limit is reached. 
-It keeps updating an array $C = [c_1, \ldots, c_m]$ of counts of currently infected individuals (i.e.
-non-removed) in different states; an array $S = [s_1, \ldots, s_m]$ of counts of sampled individuals in different states; as well as a mapping $M$ between states and ids of currently infected individuals in those states, and a mapping $N$ between states and ids of sampled individuals in those states.
+It keeps updating an array $C = [c_1, \ldots, c_m]$ of counts of currently infected individuals (i.e., non-removed) in different states; an array $S = [s_1, \ldots, s_m]$ of counts of sampled individuals in different states; as well as a mapping $M$ between states and ids of currently infected individuals in those states, and a mapping $N$ between states and ids of sampled individuals in those states.
 In the beginning ($t=0$) the only infected individual corresponds to the root: $c_k = 0 \;\forall k \neq state(root), \;c_{state(root)} = 1; \;M_k = \emptyset \; \forall k \neq state(root)$, $ \; M_{state(root)} = \{root\}$, and no individual is sampled: $s_k=0,\;N_k = \emptyset \; \forall 1 \leq k \leq m$.
@@
 At step 2, to chose the type of the event, we draw a value $v$ (uniformly) from an interval $[0, r[$.
-This interval can be seen as composed of subintervals corresponding to each possible event, where the width of each subinterval is defined by the corresponding rate and infected individual count, e.g.
-the subinterval corresponding to a transmission from $k$ to $r$ has a width of $\lambda_{kr}c_k$.
+This interval can be seen as composed of subintervals corresponding to each possible event, where the width of each subinterval is defined by the corresponding rate and infected individual count, e.g., the subinterval corresponding to a transmission from $k$ to $l$ has a width of $\lambda_{kl}c_k$.
 The subinterval in which $v$ is located defines the event type.
@@
 At step 3, we randomly draw an individual $i$ of type $k$ (selected at step 2) from $M[k]$, and proceed depending on the event type.
-If it is a state-change from $k$ to $r$, then we decrease $c_k$ by one, increase $c_r$ by one, remove $i$ from $M[k]$ and put $i$ into $M[r]$.
-If it is a transmission from $k$ to $r$, then we increase $c_r$ by one, create a new id $j$ for the recipient and put $j$ into $M[r]$.
+If it is a state-change from $k$ to $l$, then we decrease $c_k$ by one, increase $c_l$ by one, remove $i$ from $M[k]$ and put $i$ into $M[l]$.
+If it is a transmission from $k$ to $l$, then we increase $c_l$ by one, create a new id $j$ for the recipient and put $j$ into $M[l]$.
 If it is a removal, then we decrease $c_k$ by one, remove $i$ from $M[k]$, and draw a value $p$ (uniformly) from an interval $[0, 1[$: if $p < \rho$, the individual gets sampled and we put $i$ into $N[k]$.
@@
 For the PN version of this simulator, we add additional $m$ states for notified-partner versions of each state.
-$m + k$'s state-change and transmission rates are set to the $k$'s state-change and transmission rates : $\mu_{m+k,m+l} = \mu_{k,l}$; $\lambda_{m+k,m+l} = \lambda_{k,l}$ $\forall 1 \leq  k,l \leq m$.
-The rates between initial and notified states set to zero: ;  $\mu_{m+k,l} = \mu_{k,m+l} = \lambda_{m+k,l} = \lambda_{k,m+l} = 0$.
-The removal rates of notified states are set to $\phi$: $\psi_{m+k} = \phi >> \psi_k$, while their sampling probabilities are set to $1$: $\rho_{m+k} = 1$.
+$m + k$'s state-change rates are set to the $k$'s state-change rates : $\mu_{m+k,m+l} = \mu_{k,l}\; \forall 1 \leq  k,l \leq m$, while the state-change rates between initial and notified states are set to zero: $\mu_{m+k,l} = \mu_{k,m+l} = 0$.
+Only the transmissions to non-notified states are allowed (and the ones from notified states are set to the corresponding non-notified rates): $\lambda_{m+k,l} = \lambda_{k,l}; \;\lambda_{m+k,l+k} = \lambda_{k,m+l} =0 \; \forall 1 \leq  k,l \leq m$.
+The removal rates of notified states are set to $\phi$: $\psi_{m+k} = \phi >> \psi_k$, while their sampling probabilities are set to $1$: $\rho_{m+k} = 1 \; \forall 1 \leq  k \leq m$.
 We keep a mapping of most recent partners $P$ for each infectious individual, and update it for both the donor and the recipient at each transmission event.
@@
 At each removal event, if the removed individual $i$ gets sampled upon removal, we draw a value $p$ (uniformly) from an interval $[0, 1[$.
-If $p < \upsilon$, and if $i$'s most recent partner ($j = P[i]$) is not yet sampled (not in $N[state(j)]$) and not yet notified ($J \leq m$), we change $j$'s state to $m + j$ as in a state-change event.
+If $p < \upsilon$, and if $i$'s most recent partner ($j = P[i]$) is not yet sampled (not in $N[state(j)]$) and not yet notified ($j \leq m$), we change $j$'s state to $m + j$ as in a state-change event.
 
@@
 \begin{equation}
-\begin{split}
-U_p^{(1)}(t, t_A) &= 
+U_p^{(1)}(t, t_A) = 
 \begin{cases}
@@
 p_{pa}^{(x)}(t) & \textit{\color{gray} if $t_A \leq t$}
-\end{cases}{\color{gray},}\\
-&\textit{\color{gray}where $t_x =T$}\\
-\end{split}
+\end{cases}\textit{\color{gray}, where $t_x =T$}
 \end{equation}
@@
 What we need instead is a removal with no sampling at time $\tau$.
-To approximate $U_p^{(2)}(t, t_A)$ we will divide $p_{pb}^{(A)}(t)$ by the probability of no removal during the time $(t_A - t)$ (i.e.
-$e^{-\psi(t_A - t)}$), multiply it by the probability of removal during that time ($1 - e^{-\psi(t_A - t)}$), and by the probability of not being sampled upon removal ($1 - \rho$):
+To approximate $U_p^{(2)}(t, t_A)$ we will divide $p_{pb}^{(A)}(t)$ by the probability of no removal during the time $(t_A - t)$ (i.e., $e^{-\psi(t_A - t)}$), multiply it by the probability of removal during that time ($1 - e^{-\psi(t_A - t)}$), and by the probability of not being sampled upon removal ($1 - \rho$):
 
@@
 
-We will also need to describe the probability of a\textbf{ mixed branch} $p_{m(s1,s2)}^{(i)}(t,t_r)$ for branches that include a hidden partner subtree (AB-C branches in Fig.~\ref{fig:pn-branches}(2) and A's tip branch in Fig.~\ref{fig:pn-branches}(3)).
-$s1$ here denotes the type of the top of the branch (e.g.
-between AB and BC in Fig.~\ref{fig:pn-branches}(2), partner) and $s2$ denotes the type of the bottom of the branch (e.g.
-between AB and A in Fig.~\ref{fig:pn-branches}(3), notifier).
-Let's denote the first notification time of the hidden partner as $t_r$, and the time of the hidden partner subtree start as $t_h$, while as usual $t_i$ stands for the time at the branch end.
+We now describe the probability of a\textbf{ mixed branch} $p_{m(s1,s2)}^{(i)}(t,t_r)$ for branches that include a hidden partner subtree (AB-C branches in Fig.~\ref{fig:pn-branches}(2) and A's tip branches in Fig.~\ref{fig:pn-branches}(3)).
+$s1$ here denotes the type of the top of the branch (e.g., between AB and BC in Fig.~\ref{fig:pn-branches}(2), partner) and $s2$ denotes the type of the bottom of the branch (e.g., between AB and A in Fig.~\ref{fig:pn-branches}(3), notifier).
+Let us denote the first notification time of the hidden partner as $t_r$, and the time of the hidden partner subtree start as $t_h$, while as usual $t_i$ stands for the time at the branch end.
 For instance, in Fig~\ref{fig:pn-branches}(2a) $t_i=t_C$, $t_r=t_A$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(2b) $t_i=t_r=t_C$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(3c) $t_i=t_r=t_A$, $t_h=t_{AB}$.
@@
 Note that since the branch contains a hidden partner, this partner must have at least one notifier (who is observed as all the notifiers).
-If the branch is external, the notifier could correspond to its tip: For example, such notifiers correspond to tip A in Fig.~\ref{fig:pn-branches}(2b) and tip C in Fig.~\ref{fig:pn-branches}(3b).
-If the branch is external, the notifier could only correspond to a tip in its node's supertree: For example, tip A who is part of the C's supertree in Fig.~\ref{fig:pn-branches}(2a,c).
+The notifier could correspond to a tip in the branch's supertree: For example, tip A who is part of the C's supertree in Fig.~\ref{fig:pn-branches}(2a,c).
+If the branch is external, the notifier could also correspond to its own tip: For example, such notifiers correspond to tip A in Fig.~\ref{fig:pn-branches}(2b) and tip C in Fig.~\ref{fig:pn-branches}(3b).
 Having a notifier as the branch's tip implies that the bottom part of the branch corresponds to the notifier.
@@
 
-\begin{figure*}[tbhp]
+\begin{figure*}[h]
 \centering 
@@
 
-We will first describe the one-hidden-partner case $p_{m(s1,s2)}^{(i)}(t,t_r)$ (where $s1 \in \{p, -\}$, $s2 \in \{n, -\}$ and $-$ stands for the standard (unnotified non-notifier) state), and then proceed with the two-partner case $p_{m(p,-,n)}^{(i)}(t,t_{r})$, where $t_{r}$ is the earliest notification time for the hidden partner at the top of the branch (e.g.
-$t_r=t_C$ in Fig.~\ref{fig:pn-mixed}), the only notification time for the hidden partner at the bottom of the branch being $t_i$  (e.g.
-$t_i=t_A$ in Fig.~\ref{fig:pn-mixed}).
+We will first describe the one-hidden-partner case $p_{m(s1,s2)}^{(i)}(t,t_r)$ (where $s1 \in \{p, -\}$, $s2 \in \{n, -\}$ and $-$ stands for the standard (unnotified non-notifier) state), and then proceed with the two-partner case $p_{m(p,-,n)}^{(i)}(t,t_{r})$, where $t_{r}$ is the earliest notification time for the partner hidden at the top of the branch (e.g., $t_r=t_C$ in Fig.~\ref{fig:pn-mixed}), the only notification time for the partner hidden at the bottom of the branch being $t_i$  (e.g., $t_i=t_A$ in Fig.~\ref{fig:pn-mixed}).
 
@@
 In our case there must be at least one hidden transmission: to/from the partner at time $t_h$ (where the partner tree stayed unsampled).
-If we remove the probability of no event during this time ($e^{-(\lambda + \psi)(t_h - t)}$) from $p^{(h)}(t)$, we will obtain a probability of a branch with at least one oriented hidden transmission somewhere between $t$ and $t_h$ (the probability of the corresponding hidden tree being included).
+If we remove the probability of no event during this time ($e^{-(\lambda + \psi)(t_h - t)}$) from $p^{(h)}(t)$, we will obtain a probability of a branch with at least one hidden transmission somewhere between $t$ and $t_h$ (the probability of the corresponding hidden tree being included).
 $p_{top=-}^{(h)}(t)$ expresses almost the same thing, with the difference being that the last hidden transmission must have happened at time $t_h$ and correspond to the hidden partner.
@@
 &\textit{\color{gray}~~~~~~~~$t_{r}$ is the earliest notification time of the hidden partner.}\\
-\end{split}\label{eq:p_mixed}
+\end{split}\label{eq:p_mixed_eq}
 \end{equation}
@@
 \begin{array}{ll}
-l^{(\not{n},\not{n})}(i) &= p^{(i)}(t_j)\psi\rho (1-\upsilon) \textit{\color{gray} ~i.e. $i$ was not notified and did not notify}\\
+l^{(\not{n},\not{n})}(i) &= p^{(i)}(t_j)\psi\rho (1-\upsilon) \textit{\color{gray} ~i.e., $i$ was not notified and did not notify}\\
 & \textit{\color{gray} (standard branch evolution followed by sampling and no notification)}\\
-&~~+p_{m(-,n)}^{(i)}(t_j,t_i)\psi\rho\upsilon \textit{\color{gray} ~i.e. $i$ was not notified but notified their partner who is hidden}\\
-& \textit{\color{gray} (mixed standard-notifier branch evolution followed by sampling and notification)}\\
-\\
+&~~+p_{m(-,n)}^{(i)}(t_j,t_i)\psi\rho\upsilon \textit{\color{gray} ~i.e., $i$ was not notified but notified their partner who is hidden}\\
+& \textit{\color{gray} (mixed standard-notifier branch evolution followed by sampling and notification)}
+\end{array}
+\label{eq:tipxx}
+\end{equation}
+
+\begin{equation}
+\scriptsize
+\begin{array}{ll}
 l^{(\not{n},n)}(i) &= p_n^{(i)}(t_j)\psi\rho
-\upsilon \textit{\color{gray} ~i.e. $i$ was not notified but notified their most recent patner (observed)}\\
-& \textit{\color{gray} (notifier branch evolution followed by sampling and notification)}\\
-&~~\\
-\\
+\upsilon \textit{\color{gray} ~i.e., $i$ was not notified but notified their most recent patner (observed)}\\
+& \textit{\color{gray} (notifier branch evolution followed by sampling and notification)}
+\end{array}
+\label{eq:tipxx}
+\end{equation}
+
+\begin{equation}
+\scriptsize
+\begin{array}{ll}
 l^{(n,\not{n})}(i,r) &= \begin{cases}
-%0 \textit{\color{gray} ~if $t_r < t_j$}\\
-%\textit{\color{gray} (not possible as once notified partners do not transmit)}
-%\\
-%\\
 p_{pa}^{(i)}(t_j)\phi(1 - \upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_r < t_j$ and did not notify}\\
-\textit{\color{gray} (notified partner branch evolution followed by notification-provoked sampling and no notification)}
-\\
-~~~+p_{m(pa,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r < t_j$, }\\
-\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$  did not notify}\\
+\textit{\color{gray} (notified partner branch evolution then notification-provoked sampling and no notification)}
+\\
+~~~+p_{m(pa,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e., branch $i$ contains a hidden partner notified by $r$}\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~at $t_r < t_j$, $i$  did not notify}\\
  \textit{\color{gray} ~~~(mixed partner-standard branch evolution followed by sampling and no notification)}\\
- ~~~+p_{m(pa,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
+ ~~~+p_{m(pa,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon \textit{\color{gray} ~i.e., branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
  \textit{\color{gray} ~~~(mixed partner-notifier branch evolution followed by sampling and notification)}
 \\
-~~~+p_{m(pa,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r < t_j$, }\\
-\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
- \textit{\color{gray} ~~~(mixed partner-standard-notifier branch evolution followed by sampling and notification)}
+~~~+p_{m(pa,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e., branch $i$ contains a hidden partner notified by $r$}\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~at $t_r < t_j$, $i$ notified another hidden partner}\\
+ \textit{\color{gray} ~~~(mixed partner-standard-notifier branch evolution then sampling and notification)}
 \\
@@
 p_{pb}^{(i)}(t_j)\psi\rho(1 - \upsilon) \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and did not notify}\\
-\textit{\color{gray} (partner-before-notification branch evolution followed by standard sampling and no notification)}
-\\
-~~~+p_{m(pb,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r > t_i$, }\\
-\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$  did not notify}\\
+\textit{\color{gray} (partner-before-notification branch evolution then standard sampling and no notification)}
+\\
+~~~+p_{m(pb,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e., branch $i$ contains a hidden partner notified by $r$}\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~at $t_r > t_i$,  $i$  did not notify}\\
  \textit{\color{gray} ~~~(mixed partner-standard branch evolution followed by sampling and no notification)}\\
- ~~~+p_{m(pb,n)}^{(i)}(t_j,t_i)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
+ ~~~+p_{m(pb,n)}^{(i)}(t_j,t_i)\psi\rho\upsilon \textit{\color{gray} ~i.e., branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
  \textit{\color{gray} ~~~(mixed partner-notifier branch evolution followed by sampling and notification)}
 \\
-~~~+p_{m(pb,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_r > t_i$, }\\
+~~~+p_{m(pb,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e., branch $i$ contains a hidden partner notified by $r$ at $t_r > t_i$, }\\
 \textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
- \textit{\color{gray} ~~~(mixed partner-standard-notifier branch evolution followed by sampling and notification)}
+ \textit{\color{gray} ~~~(mixed partner-standard-notifier branch evolution then sampling and notification)}
 \\
@@
 \\
-~~~+p_{m(p,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_j \leq t_r \leq t_i$, }\\
-\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$  did not notify}\\
+~~~+p_{m(p,-)}^{(i)}(t_j,t_r)\psi\rho(1-\upsilon) \textit{\color{gray} ~i.e., branch $i$ contains a hidden partner notified by $r$}\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~at $t_j \leq t_r \leq t_i$, $i$  did not notify}\\
  \textit{\color{gray} ~~~(mixed partner-standard branch evolution followed by sampling and no notification)}\\
- ~~~+p_{m(p,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
+ ~~~+p_{m(p,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon \textit{\color{gray} ~i.e., branch $i$ contains a hidden partner notified both by $r$ and $i$}\\
  \textit{\color{gray} ~~~(mixed partner-notifier branch evolution followed by sampling and notification)}
 \\
-~~~+p_{m(p,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e. branch $i$ contains a hidden partner notified by $r$ at $t_j \leq t_r \leq t_i$, }\\
-\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$i$ notified another hidden partner}\\
- \textit{\color{gray} ~~~(mixed partner-standard-notifier branch evolution followed by sampling and notification)}
-\end{cases}\\
-\\
+~~~+p_{m(p,-,n)}^{(i)}(t_j,t_r)\psi\rho\upsilon  \textit{\color{gray} ~i.e., branch $i$ contains a hidden partner notified by $r$}\\
+\textit{\color{gray} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~at $t_j \leq t_r \leq t_i$, $i$ notified another hidden partner}\\
+ \textit{\color{gray} ~~~(mixed partner-standard-notifier branch evolution then sampling and notification)}
+\end{cases}
+\end{array}
+\label{eq:tipxx}
+\end{equation}
+
+\begin{equation}
+\scriptsize
+\begin{array}{ll}
 l^{(n,n)}(i,r) &= \begin{cases}
 p_b^{(i)}(t_j)\phi\upsilon \textit{\color{gray} ~if $i$ got notified at time $t_r < t_j$ and notified their most recent patner}\\
-\textit{\color{gray} (notified notifier branch evolution followed by notification-provoked sampling and  notification)}\\
+\textit{\color{gray} (notified notifier branch evolution then notification-provoked sampling and  notification)}\\
 \\
@@
 
-%Note that a notification will only be possible if the notifying branch is external as the notifier only notifies the most recent partner, i.e.
-the transmission corresponds to the notifier's parent node.
+%Note that a notification will only be possible if the notifying branch is external as the notifier only notifies the most recent partner, i.e., the transmission corresponds to the notifier's parent node.
 
@@
 
-\bigskip 
-Note that $l^{(n)}(i, r)$, $l^{(n,n)}(i, r)$ and $l^{(n,\not{n})}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time $t_r$ is known, i.e.
-once we have climbed to an internal node corresponding to the transmission between the notifier $r$ (corresponding to one of its children, must be a tip) and the partner (the other child).
-The calculation is done recursively by descending into the subtree.
-The tree likelihood calculation hence can take $O(n^2)$ steps in the worst case (ladder tree), where $n$ is the number of tips in the tree.
-However, as all the unotified equation parts ($l^{(\not{n})}(i)$, $l^{(\not{n},n)}(i)$ and $l^{(\not{n},\not{n})}(i)$) only depend on the corresponding node $i$, they can be pre-calculated and reused to speed up the recursive steps.
+\subsubsection*{Tree likelihood calculation with a pruning algorithm} 
+Note that $l^{(n)}(i, r)$, $l^{(n,n)}(i, r)$ and $l^{(n,\not{n})}(i, r)$ depend on the notifier $r$ and the notification time $t_r$.
+Note that for a node $i$ any tip $r$ that is sister to any internal node on the path between $i$ and the tree root could be its potential notifier.
+During the tree likelihood calculation we first annotate each node $i$ with a set of potential notifiers $N_i$.
+For that we first add tips' to their parent nodes'  potential notifier sets.
+We then perform a preorder tree traversal, where for every visited internal node $i$ we add its potential notifiers to its child nodes $i0$ and $i1$: $N_{i0} := N_{i0} \cup N_i - \{i0\}$, $N_{i1} := N_{i1} \cup N_i - \{i1\}$ (removing the child itself, as if it is a tip it will be found among its parent's potential notifiers).
+This preprocessing is only performed once, takes time O(N) where N is the number of tips in the tree and is negligible with respect to the rest of the parameter inference procedure.
+
+ Then we perform a postorder tree traversal, where for each node $i$ we calculate $l^{(\not{n}, \not{n})}(i)$, $l^{(\not{n}, n)}(i)$, $l^{(n, \not{n})}(i, r)$ and $l^{(n, n)}(i, r)$ $\forall r \in N_i$ if it is a tip, or $l^{(\not{n})}(i)$ and $l^{(n)}(i, r)$ $\forall r \in N_i$ if it is internal.
+This part takes O(R + N) time, where O(N) accounts for calculation of $l^{(\not{n})}(i)$ for all the $N - 1$ internal tree nodes and of $l^{(\not{n}, \not{n})}(i)$ and $l^{(\not{n}, n)}(i)$ for $N$ tips, while R is the sum of numbers of potential notifiers over all tree nodes and O(R) accounts for calculation of $l^{(n)}(i, r)$, $l^{(n, \not{n})}(i, r)$ and $l^{(n, n)}(i, r)$.
+
+For the best case (a perfectly balanced tree, where each tip is in a cherry), the number of potential notifiers is one for the tips (i.e., the sister tip) and zero for the internal nodes.
+Hence $R_b = N$ and the likelihood calculation time is linear: O(N).
+The worst case  is represented by a caterpillar tree.
+Each internal node has $k$ potential notifiers (tips in its supertree), where $k$ is the node's depth in the tree, the root's depth being 0 and the deepest internal node's depth being $N-2$.
+Moreover each tip has as many notifiers as its parent node, apart from the two deepest tips who have $N-1$ notifier each instead of $N-2$ as they are in a cherry and hence could have notified each other).
+This hence sums up to $R_c = 2\big(1 + 2 + \ldots + (N-3)\big) + 2(N-1) = N^2 - 2N + 2$ and the time is hence quadratic: $O(N^2)$.
+On our BDPN simulated dataset of 100 trees with N=500-1\,000 tips each, $R \in [2N, 38N]$ with a median at $7N$, suggesting a likelihood calculation time close to the linear one.
 
@@
 \section*{Acknowledgments}
-The authors would like to thank Dr Jakub Voznica for fruitful discussions on partner notification mechanisms.
+The authors would like to thank Dr Mathieu Moslonka-Lefebvre who initiated this project by proposing to model partner notification.
+He worked on a different version of the PN model (which is not part of this manuscript) and on the initial version of the PN test.
+We would like to thank Dr Jakub Voznica for fruitful discussions on partner notification mechanisms.
+We would also like to thank the HPC cluster team in Institut Pasteur for their support with computer simulations.
 

### 2024-07-08 hiv_b_uk/py/likelihood_stats.py
using v0.1.14 in UK

@@ -1,0 +1,21 @@
+import os
+import re
+
+
+if __name__ == "__main__":
+    import argparse
+
+    parser = argparse.ArgumentParser(description="Compares likelihoods.")
+    parser.add_argument('--likelihoods_bd', type=str, nargs='+', help="likelihood values")
+    parser.add_argument('--likelihoods_bdpn', type=str, nargs='+', help="likelihood values")
+    parser.add_argument('--log', type=str, help="likelihood stats")
+    params = parser.parse_args()
+
+    with open(params.log, 'w+') as f:
+        f.write('\tloglikelihood BDPN\tloglikelihood BD\tloglikelihood ratio\n')
+        for lk_bd_file, lk_bdpn_file in zip(params.likelihoods_bd, params.likelihoods_bdpn):
+            basename = os.path.basename(lk_bd_file)
+            rep = int(re.findall(r'\.(\d+)\.[a-z]+[=]', basename)[0]) + 1
+            lk_bd = float(open(lk_bd_file).read())
+            lk_bdpn = float(open(lk_bdpn_file).read())
+            f.write('{}\t{}\t{}\t{}\n'.format(rep, lk_bdpn, lk_bd, 2 * (lk_bdpn - lk_bd)))

### 2024-07-08 article/article_plos.tex
Aricle update

@@
 \usepackage[table]{xcolor}
+\usepackage{multirow}
 
@@
 
-We implemented the PN-test in Python 3. It is available as a part of BDPN parameter estimator ($pn\_test$, see section~"\nameref{sec:sim}").
+We implemented the PN-test in Python 3. It is available as a part of BDPN parameter estimator ($pn\_test$, see section~"\nameref{sec:tool}").
 
@@
 
-\subsection*{BD-PN parameter and CI estimator}\label{sec:sim}
+\subsection*{BD-PN parameter and CI estimator}\label{sec:tool}
 We implemented a parameter estimator for the BD-PN model (which we called bdpn).
@@
 
-\subsubsection*{PN test is highly specific and sensitive at detecting partner notification}~\label{sec:test:sim}
+\subsubsection*{PN test is highly specific and sensitive at detecting partner notification}~\label{sec:test}
 To illustrate the PN test performance, we applied it to the six 100-tree simulated datasets, corresponding to the three classical MTBD-family models: BD, BDEI (Birth-Death Exposed-Infectious~~\cite{Stadler2014}) and BDSS (Birth-Death with Super Spreading~\cite{Stadler2013a}), and to their -PN versions: BD-PN, BDEI-PN and BDSS-PN (see~"\nameref{sec:sim}" for dataset details).
 
-For 98/97/95 out of 100 trees in the BD/BDEI/BDSS dataset the PN test p-value was above $0.05$ (i.e., 98\%/97\%/95\% true negative and 2\%/3\%/5\% false positive results).
-For 95/95/96 out of 100 trees in the BD-PN/BDEI-PN/BDSS-PN dataset the PN test p-value was below $0.05$ (i.e., 95\%/95\%/96\% true positive and 5\%/5\%/4\% false negative results).
+For 100/98/98 out of 100 trees in the BD/BDEI/BDSS dataset the PN test p-value was above $0.05$ (i.e., 100\%/98\%/98\% true negative and 0\%/2\%/2\% false positive results).
+For 95/93/94 out of 100 trees in the BD-PN/BDEI-PN/BDSS-PN dataset the PN test p-value was below $0.05$ (i.e., 95\%/93\%/94\% true positive and 5\%/7\%/6\% false negative results).
 These results are summarised in Table~\ref{tbl:pntest}.
@@
   \midrule
-\textbf{BD}& 2 & 98 & 0.479 \\
-\textbf{BD-PN}& 95 & 5 & 0.015 \\
+\textbf{BD}& 0 & 100 & 0.538 \\
+\textbf{BD-PN}& 95 & 5 & 0.007 \\
   \midrule
-\textbf{BDEI}& 3 & 97 & 0.531 \\
-\textbf{BDEI-PN}& 95 & 5 & 0.008 \\
+\textbf{BDEI}& 2 & 98 & 0.539 \\
+\textbf{BDEI-PN}& 93 & 7 & 0.027 \\
   \midrule
-\textbf{BDSS}& 5 & 95 & 0.505 \\
-\textbf{BDSS-PN}& 96 & 4 & 0.008 \\
+\textbf{BDSS}& 2 & 98 & 0.515 \\
+\textbf{BDSS-PN}& 94 & 6 & 0.022 \\
 \end{tabular}
@@
 
-We applied bdpn to each of the trees fixing the sampling probability $\rho$ to its real value (for identifiability).
+We applied bdpn to each of the trees fixing the sampling probability $\rho$ to its true value (for identifiability).
 %The results with the becoming-uninfectious rate $\psi$ fixed are similar and given in the Supplementary Material.
-As expected with a maximum likelihood estimator, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the real parameter values for all the trees in both settings.
+As expected with a maximum likelihood estimator, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the true parameter values for all the trees in both settings.
 
@@
 We compared the method accuracy by a paired z-test.
-On the BD data set, bd and bdpn estimators performed equally well in estimation $R_0$ and infectious time $\frac{1}{\psi}$, with an average relative error $\leq 6\%$, and average relative bias $\leq 2\%$.
-The partner notification probability $\upsilon$ estimated on these trees by bdpn was systematically low, with the maximum value of 0.018, and mean value of 0.003.
-Its CI's 2.5\% bound hit the lower bound $0.001$ for all the trees.
-On the BD-PN data set, bdpn accuracy was significantly better (p-value $\leq 0.01$), with errors and biases $\leq 10\%$ for all parameters (vs.
-$36-40\%$ for bd).
+\textbf{On the BD data set}, bd and bdpn estimators performed equally well in estimation $R_0$ and infectious time $\frac{1}{\psi}$, with an average relative error $\leq 5\%$, and average relative bias $\leq 1\%$.
+The partner notification probability $\upsilon$ estimated on these trees by bdpn was systematically low, with the maximum value of 0.015, and mean value of 0.001.
+Moreover, its true value (zero) was within the CIs for all the trees.
+\textbf{On the BD-PN data set}, bdpn accuracy was significantly better (p-value $\leq 0.01$), with errors and biases $\leq 11\%$ for all the parameters (vs.
+$12-70\%$ for bd).
+bd estimator tends to significantly overestimate the removal rate (average relative bias of 70\%), which could be explained by the fact that the BD model has only one removal rate to accommodate a non-homogeneous population containing both standard cases and notified partners with a higher removal rate $\phi$.
+However, bd also systematically overestimates the transmission rate (average relative bias of 11\%), while this parameter is homogeneous in the population.
 
@@
 
-The percentage of the trees for which real parameter values were with the CI estimates and relative CI widths ($\frac{CI_{97.5\%} - CI_{2.5\%}}{\textit{real parameter value}}$) are shown in Table~\ref{tbl:ci}.
-They suggest that the rate parameters are easier to estimate than the probability ones.
+The percentage of the trees for which true parameter values were within the CI estimates and relative CI widths ($\frac{CI_{97.5\%} - CI_{2.5\%}}{\textit{true parameter value}}$) are shown in Table~\ref{tbl:ci}.
+For the CIs as for the point estimates, bdpn and bd estimators performed equally well on the BD trees, while on the BD-PN trees the bd estimator performed significantly worse (only 44\% and 16\% of CIs included respectively the true transmission rate $\lambda$ and removal rate $\psi$ values versus 94\% and 81\% for bdpn).
 
 \begin{figure}[h!]
+\begin{adjustwidth}{-1in}{0in}
 \centering 
-\includegraphics[width=1\textwidth]{Fig_errors_p.png}
-\caption{\textbf{Comparison of inference accuracy of bdpn (yellow) and bd (blue) estimators} on a data set of 100 trees of 500--1\,000 tips generated under BD-PN model (top) and a data set of 100 trees of 500--1\,000 tips generated under BD model (bottom), with $\rho$ parameter being fixed to its real value.
-We show the swarmplots (coloured by estimator) of relative errors for each test tree and parameter, which are measured as the normalized distance between the estimate and the real value.
-Average relative error (and in parentheses average relative bias) are displayed for each parameter and method below their swarmplot.
-For $\rho$ and $\upsilon$ absolute values are shown instead of normalized ones.
-}
+\includegraphics[width=1.2\textwidth]{Fig_errors_p.png}
+\end{adjustwidth}
+\caption{\textbf{Comparison of inference accuracy of bdpn (yellow) and bd (blue) estimators} on a data set of 100 trees of 500--1\,000 tips generated under BD-PN model (top) and a data set of 100 trees of 500--1\,000 tips generated under BD model (bottom), with $\rho$ parameter being fixed to its true value.
+We show the swarmplots (coloured by estimator) of errors for each test tree and parameter (epidemiological on the left, rates in the middle and probability on the right).
+For the epidemiological and rate parameters relative errors are shown (normalized distance between the estimate and the true value), for the partner notification probability $\upsilon$ the absolute errors are shown (distance between the estimate and the true value, where for the BD model the true value is zero).
+Average error and in parentheses average bias (absolute for probability, relative otherwise) are displayed for each parameter and method below their swarmplot.}
 \label{fig:sim} 
@@
 \small
-\caption{Percentage of simulated BD-PN trees for which the real parameter values were withing the bdpn-estimated 95\% CIs, and the median CI width: $\frac{CI_{97.5\%} - CI_{2.5\%}}{\textit{real parameter value}}$  \smallskip}
-\begin{tabular}{c|cc}
-\textbf{parameter} & \% within CI & median CI width  \\
+\caption{Percentage of simulated BD-PN and BD trees for which the true parameter values were withing the bdpn- and bd-estimated 95\% CIs, and the median CI width: $\frac{CI_{97.5\%} - CI_{2.5\%}}{\textit{true parameter value}}$.
+(The sampling probability $\rho$ was fixed to its true value in all the settings.)  \smallskip}
+\begin{tabular}{c|c|cc|cc}
+& & \multicolumn{2}{c|}{\textbf{bdpn estimator}} &  \multicolumn{2}{c}{\textbf{bd estimator}} \\
+& \textbf{parameter} & \multicolumn{1}{c|}{\% within CI} & median CI width & \multicolumn{1}{c|}{\% within CI} & median CI width \\
 \toprule 
- $\lambda$ & 96\% & 11.6\%\\
- $\psi$ & 88\% & 28.9\% \\
- $\rho$ & fixed & -- \\
- $\phi$ & 95\% & 32.2\% \\
- $\upsilon$ & 70\% & 21.7\% \\
+& $\lambda$ & 94\% & 11.4\%& 44\% & 12.6\%\\
+BD-PN & $\psi$ & 81\% & 25.6\% & 16\% & 29.4\%\\
+%& $\rho$ & \textit{fixed} & -- & \textit{fixed} & 11.6\%\\
+model &  $\phi$ & 93\% & 35.0\% & -- & --\\
+& $\upsilon$ & 71\% & 24.5\% & -- & --\\
+\midrule
+BD & $\lambda$ & 97\% & 10.3\%& 97\% & 10.3\%\\
+model & $\psi$ & 95\% & 23.2\% & 95\% & 23.1\%\\
+%& $\rho$ & \textit{fixed} & -- & \textit{fixed} & --\\
+& $\upsilon(=0)$ & 100\% & 0.004* & -- & --\\
 \bottomrule
 \end{tabular}
+\begin{flushleft} *absolute width ($CI_{97.5\%} - CI_{2.5\%}$) is given instead of the relative one, as the true value is zero.
+\end{flushleft}
 \label{tbl:ci}
@@
 
-The PN test detected partner notification in this tree (p-value=0.046).
+The PN test detected partner notification in this tree (p-value=0.014).
 Following Voznica~\textit{et al.}~\cite{Voznica2021}, we fixed the sampling probability for this tree at 25\%: $\rho=0.25$.
-Our bdpn estimator inferred the partner notification probability $\upsilon=0.07$ CI $[0.006-0.190]$,
-$R_e = 1.4$ $[1.2-1.6]$, infectious time $\frac{1}{\psi} = 8$ $[6-10]$ years, and 3.4 months [5 days -- 15 months] between partner notification and their sampling.
+Our bdpn estimator inferred the partner notification probability $\upsilon=0.09$ CI $[0.008-0.226]$,
+$R_e = 1.43$ $[1.26-1.66]$, infectious time $\frac{1}{\psi} = 8$ $[6-11]$ years, and 4.9 months [6 days -- 1.4 years] between partner notification and their sampling.
 These results globally agree (CIs intersect) with the $R_e$ estimate using a coalescent model from Rasmussen \textit{et al.}~\cite{Rasmussen2017} (between 1.0 and 2.5), and the estimates using the BDSS model from Voznica~\textit{et al.}~\cite{Voznica2021}: $R_e$ of 1.6 and 1.7 and an infectious period of 10.2 and 9.8 years (depending on the tree representation used for the deep learner).
@@
 
+For comparison, parameter estimation on the same tree with the BD model gave slightly lower estimate of $R_e$ and a slightly shorter infectious time, though with intersecting CIs ($R_e = 1.32$ $[1.19-1.47]$, $\frac{1}{\psi} = 7$ $[6-9]$ years), which is in line with underestimation of $R_e$ and $\frac{1}{\psi}$ with bd on the BD-PN simulated data.
+The likelihood ratio test suggests the BDPN model: loglikelihood under the BD-PN-estimated parameters was $l_{BDPN} = -1\,140$ vs.
+$l_{BD}=-1\,146$ under the BD parameters, which gives the ratio $-2 (l_{BD} - l_{BDPN}) =	12 > 5.99 \approx \chi^2_2(0.95)$ (the value of chi-squared distribution with 2 degrees of freedom corresponding to the significance level of $0.95$).
+% unrealistic results: $R_e = 0.5$ (suggesting contained epidemic) and very short infectious time $\frac{1}{\psi} = 0.4$ years.
+These results were robust across the 10 forests, see \nameref{S1_Table}.
+As with the results on simulated data, this  highlights the importance of accounting for partner notification during parameter estimation.
+
 \subsubsection*{Application 2: HIV-1 B epidemic in the UK}
@@
 
-The PN-test values were below 0.05 for all the forests (values between $7 \cdot 10^{-18}$ and $5 \cdot 10^{-12}$), strongly suggesting the presence of partner notification.
+The PN-test values were below 0.05 for all the forests (values between $1 \cdot 10^{-18}$ and $2 \cdot 10^{-14}$), strongly suggesting the presence of partner notification.
 Using sampling probability $\rho=0.58$ (see~"\nameref{uk_tree}" for its estimation), we estimated the partner notification probability $\upsilon=0.006-0.008$, %the following parameter values: $\lambda=0.53$, $\psi=0.43$, $\phi=23-59$, $\upsilon=0.006-0.007$, which corresponds to
-$R_e = 1.2$, infectious time $\frac{1}{\psi} = 2.3$ years, and 15--16 days between partner notification and their sampling.
+$R_e = 1.2-1.3$, infectious time $\frac{1}{\psi} = 2.3-2.4$ years, and 7--11 days between partner notification and their sampling.
 The results were consistent between 10 forests and shown in Table~\ref{tbl:uk}.
@@
 \toprule
-  $1$ & $11\,117$ & $8\,838$ & $1.24\;[1.22-1.29]$& $2.34\;[2.25-2.40]$& $15\;[3-17]$& $0.008\;[0.004-0.016]$ \\
- $2$ & $11\,109$ & $8\,833$ & $1.24\;[1.21-1.27]$& $2.33\;[2.26-2.39]$& $16\;[1-18]$& $0.006\;[0.002-0.014]$ \\
- $3$ & $11\,122$ & $8\,844$ & $1.24\;[1.21-1.29]$& $2.34\;[2.24-2.40]$& $15\;[3-17]$& $0.008\;[0.004-0.016]$ \\
- $4$ & $11\,119$ & $8\,838$ & $1.24\;[1.21-1.29]$& $2.34\;[2.24-2.40]$& $15\;[3-17]$& $0.008\;[0.004-0.016]$ \\
- $5$ & $11\,116$ & $8\,829$ & $1.24\;[1.22-1.27]$& $2.33\;[2.26-2.40]$& $16\;[3-18]$& $0.006\;[0.002-0.013]$ \\
- $6$ & $11\,111$ & $8\,827$ & $1.24\;[1.21-1.27]$& $2.34\;[2.25-2.40]$& $16\;[3-18]$& $0.007\;[0.002-0.014]$ \\
- $7$ & $11\,119$ & $8\,838$ & $1.24\;[1.21-1.27]$& $2.33\;[2.26-2.39]$& $15\;[2-18]$& $0.006\;[0.002-0.014]$ \\
- $8$ & $11\,111$ & $8\,830$ & $1.24\;[1.19-1.27]$& $2.33\;[2.25-2.39]$& $16\;[2-18]$& $0.006\;[0.002-0.014]$ \\
- $9$ & $11\,120$ & $8\,833$ & $1.24\;[1.21-1.27]$& $2.33\;[2.26-2.39]$& $16\;[2-18]$& $0.006\;[0.002-0.014]$ \\
- $10$ & $11\,112$ & $8\,827$ & $1.25\;[1.22-1.29]$& $2.34\;[2.25-2.40]$& $16\;[2-18]$& $0.007\;[0.002-0.015]$ \\
+  $1$ & $11\,117$ & $8\,838$ & $1.24\;[1.22-1.27]$& $2.34\;[2.28-2.40]$& $8\;[5-12]$& $0.008\;[0.005-0.012]$ \\
+ $2$ & $11\,109$ & $8\,833$ & $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.40]$& $7\;[4-11]$& $0.007\;[0.003-0.011]$ \\
+ $3$ & $11\,122$ & $8\,844$ & $1.24\;[1.22-1.27]$& $2.34\;[2.28-2.39]$& $8\;[5-11]$& $0.008\;[0.005-0.012]$ \\
+ $4$ & $11\,119$ & $8\,838$ & $1.30\;[1.26-1.32]$& $2.40\;[2.27-2.43]$& $11\;[3-13]$& $0.006\;[0.002-0.016]$ \\
+ $5$ & $11\,116$ & $8\,829$ & $1.24\;[1.22-1.27]$& $2.34\;[2.28-2.39]$& $9\;[6-13]$& $0.007\;[0.004-0.011]$ \\
+ $6$ & $11\,111$ & $8\,827$ & $1.25\;[1.22-1.27]$& $2.34\;[2.28-2.40]$& $8\;[5-12]$& $0.007\;[0.004-0.011]$ \\
+ $7$ & $11\,119$ & $8\,838$ & $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.39]$& $8\;[5-12]$& $0.007\;[0.004-0.011]$ \\
+ $8$ & $11\,111$ & $8\,830$ & $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.40]$& $7\;[4-11]$& $0.007\;[0.003-0.011]$ \\
+ $9$ & $11\,120$ & $8\,833$ & $1.24\;[1.21-1.27]$& $2.33\;[2.28-2.39]$& $8\;[4-12]$& $0.007\;[0.003-0.011]$ \\
+ $10$ & $11\,112$ & $8\,827$ & $1.25\;[1.22-1.28]$& $2.34\;[2.28-2.40]$& $8\;[5-11]$& $0.008\;[0.004-0.012]$ \\
  \bottomrule
@@
 
-For comparison, parameter estimation on the same forests with the BD model (with $\rho$ fixed to $0.58$) gave unrealistic results: $R_e = 0.5$ (suggesting contained epidemic) and very short infectious time $\frac{1}{\psi} = 0.4$ years.
+For comparison, parameter estimation on the same forests with the BD model gave similar results ($R_e = 1.2$, $\frac{1}{\psi} = 2.3$ years), however the likelihood ratio test suggests the BDPN model.
+For instance, on forest 1, loglikelihood under the BD-PN-estimated parameters was $l_{BDPN} = -28\,984$ vs.
+$l_{BD}=-29\,002$ under the BD parameters, which gives the ratio $-2 (l_{BD} - l_{BDPN}) =	36 > 5.99 \approx \chi^2_2(0.95)$.
+% unrealistic results: $R_e = 0.5$ (suggesting contained epidemic) and very short infectious time $\frac{1}{\psi} = 0.4$ years.
 These results were robust across the 10 forests, see \nameref{S1_Table}.
@@
 
-Importantly, our simulations show that in the presence of partner notification, the classical BD model estimator fails to accurately estimate epidemiological parameters (average relative errors $> 30\%$ for $R_0$ and infectious time $\frac{1}{\psi}$).
-On the contrary, in the absence of partner notification, the BD-PN model estimator correctly identifies $R_0$ and infectious time $\frac{1}{\psi}$ (as well as the BD model estimator), and estimates a very low partner notification probability $\upsilon$, whose confidence intervals hit the lower bound.
+Importantly, our simulations show that in the presence of partner notification, the classical BD model estimator fails to accurately estimate epidemiological parameters (average relative errors $> 25\%$ for $R_0$ and infectious time $\frac{1}{\psi}$).
+On the contrary, in the absence of partner notification, the BD-PN model estimator correctly identifies $R_0$ and infectious time $\frac{1}{\psi}$ (as well as the BD model estimator), and estimates a very low partner notification probability $\upsilon$, whose confidence intervals include zero.
 
@@
 The test detected presence of partner notification in both cases.
-The obtained estimates of epidemiological parameters are in agreement with what we now know about the HIV-1 B epidemic (e.g., $R_e$ between $1.2$ and $1.6$).
-We estimated a modest partner notification probability ($7\%$ in Zurich, $1\%$ in the UK), and rapid notified partner detection ($\approx 3.4$ months vs $\approx  8$ years for detection under the standard procedure in Zurich; and $\approx 16$ days vs $\approx 2.3$ years for detection under the standard procedure in the UK).
+The obtained estimates of epidemiological parameters are in agreement with what we now know about the HIV-1 B epidemic (e.g., $R_e$ between $1.2$ and $1.7$).
+We estimated a modest partner notification probability ($9\%$ in Zurich, $1\%$ in the UK), and rapid notified partner detection ($\approx 5$ months vs $\approx  8$ years for detection under the standard procedure in Zurich; and $\approx 8$ days vs $\approx 2.3$ years for detection under the standard procedure in the UK).
 
@@
 
-\paragraph*{S1 Table.}
-\label{S1_Table}
-{\bf BD model parameters estimated for UK HIV-1 B epidemic between 2012 and 2015}, assuming that our data represents 58\% of infected cases ($\rho=0.58$).
-
-\begin{table}[!ht]
-%\begin{adjustwidth}{-0.8in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\centering
-\caption{
-{\bf BD model parameters estimated for UK HIV-1 B epidemic between 2012 and 2015}, assuming that our data represents 58\% of infected cases ($\rho=0.58$).}
-
-\begin{tabular}{c|cc|cc}
-&&&&infectious\\
-fo-&num.&num.&&time\\
-rest&tips&trees&$R_e$&$\frac{1}{\psi}$ [years]\\
-\toprule
- $1$ & $11\,117$ & $8\,838$ & $0.53\;[0.52-0.55]$& $0.37\;[0.36-0.37]$ \\
- $2$ & $11\,109$ & $8\,833$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
- $3$ & $11\,122$ & $8\,844$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
- $4$ & $11\,119$ & $8\,838$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
- $5$ & $11\,116$ & $8\,829$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
- $6$ & $11\,111$ & $8\,827$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
- $7$ & $11\,119$ & $8\,838$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
- $8$ & $11\,111$ & $8\,830$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
- $9$ & $11\,120$ & $8\,833$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
- $10$ & $11\,112$ & $8\,827$ & $0.54\;[0.52-0.55]$& $0.37\;[0.36-0.38]$\\
- \bottomrule
- \end{tabular}
-%\begin{flushleft} We assumed that our data represents 58\% of infected cases: $\rho=0.58$.
-%\end{flushleft}
-\label{tbl:uk}
-%\end{adjustwidth}
-\end{table}
+%\paragraph*{S1 Table.}
+%\label{S1_Table}
+%{\bf BD model parameters estimated for UK HIV-1 B epidemic between 2012 and 2015}, assuming that our data represents 58\% of infected cases ($\rho=0.58$).
+%
+%
+%\begin{table}[!ht]
+%%\begin{adjustwidth}{-0.8in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+%\centering
+%\caption{
+%{\bf BD model parameters estimated for UK HIV-1 B epidemic between 2012 and 2015}, assuming that our data represents 58\% of infected cases ($\rho=0.58$).}
+%
+%\begin{tabular}{c|cc|cc}
+%&&&&infectious\\
+%fo-&num.&num.&&time\\
+%rest&tips&trees&$R_e$&$\frac{1}{\psi}$ [years]\\
+%\toprule
+% $1$ & $11\,117$ & $8\,838$ & $1.23\;[1.20-1.26]$& $2.32\;[2.26-2.38]$\\
+% $2$ & $11\,109$ & $8\,833$ & $1.23\;[1.20-1.26]$& $2.32\;[2.26-2.38]$\\
+% $3$ & $11\,122$ & $8\,844$ & $1.23\;[1.21-1.26]$& $2.32\;[2.26-2.38]$\\
+% $4$ & $11\,119$ & $8\,838$ & $1.23\;[1.20-1.26]$& $2.32\;[2.26-2.38]$\\
+% $5$ & $11\,116$ & $8\,829$ & $1.23\;[1.21-1.26]$& $2.32\;[2.26-2.38]$\\
+% $6$ & $11\,111$ & $8\,827$ & $1.24\;[1.21-1.26]$& $2.32\;[2.26-2.38]$\\
+% $7$ & $11\,119$ & $8\,838$ & $1.23\;[1.20-1.26]$& $2.32\;[2.26-2.38]$\\
+% $8$ & $11\,111$ & $8\,830$ & $1.23\;[1.20-1.26]$& $2.32\;[2.26-2.38]$\\
+% $9$ & $11\,120$ & $8\,833$ & $1.23\;[1.20-1.26]$& $2.32\;[2.26-2.38]$\\
+% $10$ & $11\,112$ & $8\,827$ & $1.24\;[1.21-1.26]$& $2.32\;[2.26-2.38]$\\
+% \bottomrule
+% \end{tabular}
+%%\begin{flushleft} We assumed that our data represents 58\% of infected cases: $\rho=0.58$.
+%%\end{flushleft}
+%\label{tbl:uk}
+%%\end{adjustwidth}
+%\end{table}
 

### 2024-07-08 simulations/py/plot_error_both.py
Better error fig

@@ -15,8 +15,12 @@ par2greek = {'lambda': u'\u03bb', 'psi': u'\u03c8', 'phi': u'\u03c6', 'p': '\u03
              'partner_removal_time': 'partner removal time 1' + u'\u002F\u03c6'}
 PARAMETERS = RATE_PARAMETERS + EPIDEMIOLOGIC_PARAMETERS
 
-PS_BDPN = EPIDEMIOLOGIC_PARAMETERS + ['p', 'upsilon']
-PS_BD = EPIDEMIOLOGIC_PARAMETERS[:-1] + ['p', 'upsilon']
+PS_BDPN_EPID = EPIDEMIOLOGIC_PARAMETERS
+PS_BD_EPID = EPIDEMIOLOGIC_PARAMETERS[:-1]
+PS_BDPN_RATES = ['lambda', 'psi', 'phi']
+PS_BD_RATES = ['lambda', 'psi']
+
+PROBS = ['p', 'upsilon']
 
 if __name__ == "__main__":
     import argparse
@@ -32,13 +36,14 @@ if __name__ == "__main__":
 
     plt.clf()
 
-    fig, (ax1, ax2) = plt.subplots(2, 1, figsize=(15, 12))
+    fig, ((ax1, ax2, ax3), (ax4, ax5, ax6)) = plt.subplots(2, 3, figsize=(25, 9), gridspec_kw={'width_ratios': [4, 4, 1]})
     rc = {'font.size': 12, 'axes.labelsize': 10, 'legend.fontsize': 10, 'axes.titlesize': 10, 'xtick.labelsize': 10,
           'ytick.labelsize': 10}
     sns.set(style="whitegrid")
     sns.set(rc=rc)
 
-    for est, ax, pars in ((params.estimates_bdpn, ax1, PS_BDPN), (params.estimates_bd, ax2, PS_BD)):
+    for est, ax, pars in ((params.estimates_bdpn, ax1, PS_BDPN_EPID), (params.estimates_bdpn, ax2, PS_BDPN_RATES), (params.estimates_bdpn, ax3, PROBS),
+                          (params.estimates_bd, ax4, PS_BD_EPID), (params.estimates_bd, ax5, PS_BD_RATES), (params.estimates_bd, ax6, PROBS)):
         df = pd.read_csv(est, sep='\t', index_col=0)
 
         real_df = df.loc[df['type'] == 'real', :]
@@ -102,7 +107,7 @@ if __name__ == "__main__":
[…]

### 2024-09-02 article/article_plos.tex
Rewrote the PN extension justifications

@@
 Its application to simulated data showed that it is both highly specific and sensitive.
-For the simplest representative of the MTBD-PN family, the BD-PN model, we developed a maximum likelihood parameter and CI estimator.
+For the simplest representative of the MTBD-PN family, the BD-PN model, we solved the differential equations and proposed a closed form solution for the likelihood function, which we implemented in a software program to estimate the model parameters and their confidence intervals from phylogenetic trees.
 It performed accurate estimations on simulated data, and detected partner notification in HIV-1 B epidemics in Zurich and the UK.
@@
 \section*{Author summary}
-Phylodynamic models can estimate epidemiological parameters from pathogen phylogenetic trees (i.e., genealogies).
+Phylodynamic models can estimate epidemiological parameters (e.g., the number of secondary infections $R_e$) from pathogen phylogenetic trees (i.e., genealogies, estimated from sampled pathogen genomic sequences).
 These models typically assume that sampling of pathogen sequences from detected cases is independent between infected individuals.
@@
 
-We developed a phylodynamic model accounting for partner notification and a test for detection of partner notification in pathogen phylogenetic trees.
-They showed good performance both on simulated and real data (HIV-1 B epidemics in Zurich and the UK).
-Importantly, we showed that not accounting for partner notification when it is present leads to bias in parameter estimation, while our new estimator performs well both in presence and in absence of partner notification.
+We developed a phylodynamic model accounting for partner notification and a test for the detection of partner notification in pathogen phylogenetic trees.
+Our test and epidemiological parameter estimator showed good performance both on simulated and real data.
+We detected the presence of partner notification in the HIV-1 B epidemics in Zurich and the UK, and corrected the previous parameter estimates made without accounting for partner notification.
+%  Importantly, we showed that not accounting for partner notification when it is present leads to bias in parameter estimation, while our new estimator performs well both in presence and in absence of partner notification.
 
@@
 In contrast, phylodynamic approach can overcome such biases as it uses the transmission structure as one of its information sources~\cite{vaughanEstimatesEarlyOutbreakspecific2024}.
-Therefore, using genomic data with phylodynamic estimations can provide valuable insights for preventing epidemic spreads.
+Using genomic data with phylodynamic estimations can provide valuable insights for preventing epidemic spreads.
 
 Phylodynamic models can be classified into two main families:  coalescent~\cite{Volz2009a,Drummond2005,Pybus2000a} and birth-death (BD)~\cite{Kendall1948,Maddison2007,Stadler2009,Stadler2010}.
-Coalescent models are often preferred for estimating deterministic population dynamics, however for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\cite{Macpherson2021}.
+Coalescent models are often preferred for estimating deterministic population dynamics, however, for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\cite{Macpherson2021}.
 In the classic BD model with incomplete sampling~\cite{Stadler2009}, births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g., due to healing, self-isolation, starting a treatment, or death, modelled with a constant removal rate).
 
-The epidemic spread and detection can however be non-homogenous and depend on different factors, including governmental health policies (e.g., quarantine measures or development of pathogen detection tests), pathogen evolution over time (e.g., some SARS-CoV-2 variants are more transmissible than the others),  differences between host individuals (e.g., due to their immune system particularities or to their behaviour).
-
-To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extention of the classical BD model was developed by Stadler \textit{et al.}~\cite{Stadler2013a}.
-MTBD framework allows for different types of individual states (and changes between them, modelled with state change rates).
-These models are phylodynamic analogies of compartmental models in classical epidemiology (e.g., SIR, Susceptible-Infectious-Removed).
-Examples of the MTBD family models include the birth-death exposed-infectious (BDEI) model, which was designed for pathogens featuring an incubation period between the moments of infection and of becoming infectious (e.g., Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen (e.g., HIV) more than the others~\cite{Stadler2014}.
+However, the spread and detection or an epidemic can be non-homogenous and depend on different factors, including governmental health policies (e.g., quarantine measures or development of pathogen detection tests), pathogen evolution over time (e.g., some SARS-CoV-2 variants are more transmissible than the others),  differences between host individuals (e.g., due to their immune system particularities or to their behaviour).
+
+To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extension of the classical BD model was developed by Stadler \textit{et al.}~\cite{Stadler2013a}.
+The MTBD framework allows for different types of individual states (and changes between them, modelled with state change rates).
+MTBD models are phylodynamic analogies of compartmental models in classical epidemiology (e.g., SIR, Susceptible-Infectious-Removed).
+Examples of MTBD models include the birth-death exposed-infectious (BDEI) model, which was designed for pathogens featuring an incubation period between the moments of infection and of becoming infectious (e.g., Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen (e.g., HIV) more than the others~\cite{Stadler2014}.
 
@@
 
-Another important source of heterogeneity that none of these models accounts for is inflicted by non-independent sampling.
-One can set different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability between time intervals with the BDSKY, however, none of these models can account for correlation effects in sampling, in particular due to partner notification (contact tracing).
-Partner notification or contact tracing is a process in which contacts of detected infected individuals are identified and offered testing for infection.
+Another important source of heterogeneity that none of these models accounts for is caused by non-independent sampling.
+One can set different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability between time intervals with the BDSKY, however, none of these models can account for correlation effects in sampling, in particular due to partner notification (PN).
+PN (or contact tracing) is a process in which contacts of detected infected individuals are identified and offered testing for infection.
 It has been applied to various infectious diseases, including syphilis, HIV, tuberculosis, Ebola and COVID-19, and shown to successfully reduce transmission~\cite{el-sadrContactTracingBarriers2022}.
@@
 
-In this study we propose an extension of MTBD models that allows for modeling non-random sampling due to partner notification (PN), MTBD-PN.
+In this study we propose an extension of MTBD models that allows for modeling non-random sampling due to partner notification, MTBD-PN.
 We describe the -PN extension and its assumptions, and present a transmission tree simulator under MTBD-PN models.
 We propose a non-parametric test for detecting partner notification in pathogen phylogenetic trees.
-For the simplest representative of this model family, the BD-PN model, we also derive the equations for its likelihood calculation and present a maximum likelihood parameter estimator.
+For the simplest representative of this model family, the BD-PN model, we also derive the equations (and their closed-form solution) for its likelihood calculation and present a maximum-likelihood parameter estimator.
 We test our BD-PN parameter estimator on simulated data, and apply it to the study of HIV-1 B epidemics in Zurich and in the UK.
@@
 Under this model we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ (see \nameref{math}).
-The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g., by finding parameter values that maximize it, in maximum likelihood framework).
-In particular, it permits inference of the following epidemiological parameters:
+The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g., by finding parameter values that maximize it, in maximum-likelihood framework).
+In particular, it permits the estimation of the following epidemiological parameters:
 
@@
 extension of the MTBD models that adds partner notification (PN).  
-In the simplest case, the BD-PN model, at the moment of sampling the sampled individual might notify their most recent partner with a constant probability $\upsilon$.
-The notification implies that four conditions are met: (i) the sampled individual remembers their last partner and accepts to either notify the partner on their own or to give the partner's contacts to health-care providers for assisted notification; (ii) a pathogen transmission happened at the moment of contact (either the sampled individual infected the partner or the other way round); (iii) the partner is successfully reached and notified; (iv) unless the partner already became uninfectious by other means by the moment of notification (e.g., got diagnosed though the standard procedure), they get tested upon notification.
-The parameter $\upsilon$ hence regroups a complex set of conditions, which would hardly be identifiable separately from only sequence data.
-Upon notification, the partner's pathogen is sampled almost instantaneously (modeled via a constant notified sampling rate $\phi >> \psi$).
+In the partner notification process, an identified infected index case notifies their potentially infected contacts (e.g., sexual or drug injection partners for HIV), either on their own (patient referral) or with a hep of a health service provider (provider referral, when a health adviser notifies the partners without identifying the index patient; or contract referral, when if the index case fails to notify partners within a specified time period, the health adviser proceeds to provider referral)~\cite{wardPartnerNotification2014}.
+The notified partners are then provided with appropriate tests and treatment.
+Therefore, a successful detection of a new case B via partner notification by an index case A implies several conditions to be met: (i) A remembers the contact with B and  either notifies B on their own or gives the B's contacts to a health advisor for assisted notification; (ii) B is successfully reached and notified; (iii) B agrees to take a test; (iv) B is indeed infected.
+For instance, according to the HIV report from the Public Health England~\cite{aHIVUnitedKingdom2014}, in England in 2013, reaching the stage (ii) had an estimated probability of about 50\%; reaching the stage (iii) -- of about 41\% (as 83\% of all the patients who attended a clinic through PN had a reported HIV test); and the stage (iv) -- of less than 3\%.
+
+All these conditions are hardly identifiable separately from only sequence data, hence in our model we combine them into one, ``an infected person B being successfully notified by A'', and model it with a constant notification probability $\upsilon$ which is applied upon sampling of A.
+
+For simplicity, a positive draw of $\upsilon$ models a slightly modified scenario: (i) A remembers the contact with B and  either notifies B on their own or gives the B's contacts to a health advisor for assisted notification; (iv') B is indeed infected, moreover, \textit{the infection of either A or B is caused by the pathogen transmission during their contact};  and either (ii) B is successfully reached and notified and (iii) agrees to take a test; or (ii') \textit{by the time of notification, B is already removed by other means} (already got sampled or stopped being infectious without sampling, e.g., died).
+
+The condition (iv') in contrast to (iv) excludes the possibility of notifying partners if there was no pathogen transmission during the contact with them.
+We additionally simplify the model by allowing to notify only the last such partner.
+From now on, we will use the term ``partner'' to denote a partner, during a contact with whom a pathogen transmission happened.
+
+Upon notification, if not yet removed,  the partner's pathogen is sampled almost instantaneously (modeled via a constant notified sampling rate $\phi >> \psi$).
 Upon sampling, the partner might notify in their turn.
+
 BD-PN hence has 3 classical BD and 2 additional parameters:
@@
 We give its algorithmic details in~\nameref{S1_Appendix}.
-The results in section~"\nameref{sec:test:sim}" show that the trees generated under MTBD and MTBD-PN models are different and easily distinguishable.
+The results in section~"\nameref{sec:test}" show that the trees generated under MTBD and MTBD-PN models are different and easily distinguishable.
 
@@
 \item $R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
-\item removal rate $\psi \in ]1 / 20, 1 / 5]$,
+\item removal rate $\psi \in ]0.05, 0.2]$,
 \item sampling probability $\rho \in ]0.1, 0.9]$,
@@
 
-The common parameters were consistent between models, in a sense that the $i$-th tree under the BDSS model had the same $R_e$, $\psi$ and $\rho$ values as the $i$-th trees under BDSS-PN, BD(-PN) and BDEI(-PN) models; the $i$-th tree under the BDSS-PN model had the same $\phi$ and $\upsilon$ values as the $i$-th trees under BD-PN and BDEI-PN models.
+The common parameters were consistent between models, in the sense that the $i$-th tree under the BDSS model had the same $R_e$, $\psi$ and $\rho$ values as the $i$-th trees under BDSS-PN, BD(-PN) and BDEI(-PN) models; the $i$-th tree under the BDSS-PN model had the same $\phi$ and $\upsilon$ values as the $i$-th trees under BD-PN and BDEI-PN models, etc.
 
@@
 (If the total number of cherries N is odd, instead of swapping the selected tips of the last two cherries, we swap the selected tips of the last three cherries in a cycle as in Fig.~\ref{fig:tipswap}).
-We then calculate the time differences $d_{i'}$ between the tips of these shuffles cherries.
+We then calculate the time differences $d_{i'}$ between the tips of these reshuffled cherries.
 The test then compares $d_i$ to $d_{i'}$: $sign(d_{i'} - d_i) = 1$ if $d_{i'}$ is larger than $d_i$; $= -1$ if it is smaller; or $= 0$ if they are equal (this could happen as in real datasets dates are often truncated, e.g., to month or even a year).
@@
 %The results with the becoming-uninfectious rate $\psi$ fixed are similar and given in the Supplementary Material.
-As expected with a maximum likelihood estimator, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the true parameter values for all the trees in both settings.
+As expected with a maximum-likelihood estimator, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the true parameter values for all the trees in both settings.
 

### 2024-09-05 simulations/py/summary_table_lk_ratio.py
Likelihood ratio tests on simulated data

@@ -1,0 +1,41 @@
+import re
+
+import numpy as np
+import pandas as pd
+
+if __name__ == "__main__":
+    import argparse
+
+    parser = argparse.ArgumentParser(description="Summarize likelihood ratios.")
+    parser.add_argument('--pn', nargs='+', type=str, help="BD-PN model likelihoods")
+    parser.add_argument('--no_pn', nargs='+', type=str, help="BD model likelihoods")
+    parser.add_argument('--tab', type=str, help="summary table")
+    params = parser.parse_args()
+
+    df = pd.DataFrame(columns=['lk', 'lk_PN'])
+
+    for log in params.no_pn:
+        i = int(re.findall(r'[0-9]+', log)[-1])
+        with open(log, 'r') as f:
+            val = float(f.readline().strip('\n'))
+        df.loc[i, 'lk'] = val
+
+    for log in params.pn:
+        i = int(re.findall(r'[0-9]+', log)[-1])
+        with open(log, 'r') as f:
+            val = float(f.readline().strip('\n'))
+        df.loc[i, 'lk_PN'] = val
+
+    df['lk_ratio'] = -2 * (df['lk'] - df['lk_PN'])
+    df['above_threshold'] = df['lk_ratio'] > 5.99
+
+    df.sort_index(inplace=True)
+    mean_lk_r = df['lk_ratio'].mean()
+    median_lk_r = np.median(df['lk_ratio'])
+
+    df.loc['total', 'above_threshold'] = sum(df['above_threshold'])
+    df.loc['mean', 'lk_ratio'] = mean_lk_r
+    df.loc['median', 'lk_ratio'] = median_lk_r
+
+    df.to_csv(params.tab, sep='\t', index=True, index_label='tree')
+

### 2024-09-05 article/article_plos.tex
Article update (final version?)

@@
 {\Large
-\textbf\newline{Epidemiological birth-death models with partner notification} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{Accounting for partner notification in epidemiological birth-death models} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 \\
-Anna Zhukova\textsuperscript{1,2,3*},
+Anna Zhukova\textsuperscript{1,2*},
 %Mathieu Moslonka-Lefebvre\textsuperscript{1\textcurrency},
-Olivier Gascuel\textsuperscript{1,4},
+Olivier Gascuel\textsuperscript{3},
 \\
 \bigskip
-\textbf{1} Unit\'e Bioinformatique Evolutive, Institut Pasteur, Université de Paris, Paris, France
+%\textbf{1} Unit\'e Bioinformatique Evolutive, Institut Pasteur, Université de Paris, Paris, France
+%\\
+\textbf{1} G5 Evolutionary Dynamics of Infectious Diseases, Institut Pasteur, Université de Paris, Paris, France
 \\
@@
 \\
-\textbf{3} G5 Evolutionary Dynamics of Infectious Diseases, Institut Pasteur, Université de Paris, Paris, France
-\\
-\textbf{4} Institut de Syst\'{e}matique, Evolution, Biodiversit\'{e} (ISYEB) - URM 7205 CNRS, Mus\'{e}um National d'Histoire Naturelle, SU, EPHE \& UA, Paris, France
+\textbf{3} Institut de Syst\'{e}matique, Evolution, Biodiversit\'{e} (ISYEB) - URM 7205 CNRS, Mus\'{e}um National d'Histoire Naturelle, SU, EPHE \& UA, Paris, France
 \\
@@
 The models used in phylodynamics typically assume that the sampling procedure is independent between infected individuals.
-However this assumption does not hold for many epidemics, in particular for such sexually transmitted infections as HIV-1, for which partner notification schemes are included in health policies of many countries.
+However, this assumption does not hold for many epidemics, in particular for such sexually transmitted infections as HIV-1, for which partner notification schemes are included in health policies of many countries.
 
@@
 Its application to simulated data showed that it is both highly specific and sensitive.
-For the simplest representative of the MTBD-PN family, the BD-PN model, we solved the differential equations and proposed a closed form solution for the likelihood function, which we implemented in a software program to estimate the model parameters and their confidence intervals from phylogenetic trees.
+For the simplest representative of the MTBD-PN family, the BD-PN model, we solved the differential equations and proposed a closed form solution for the likelihood function.
+We implemented it in a program, which estimates the model parameters and their confidence intervals from phylogenetic trees.
 It performed accurate estimations on simulated data, and detected partner notification in HIV-1 B epidemics in Zurich and the UK.
@@
 
-Our PN test, MTBD-PN tree simulator and BD-PN parameter estimator are freely available at \href{github.com/evolbioinfo/treesimulator}{https://github.com/evolbioinfo/treesimulator} and \href{github.com/evolbioinfo/bdpn}{https://github.com/evolbioinfo/bdpn}.
+Our PN test, MTBD-PN tree simulator and BD-PN parameter estimator are freely available at \href{https://github.com/evolbioinfo/treesimulator}{github.com/evolbioinfo/treesimulator} and \href{https://github.com/evolbioinfo/bdpn}{github.com/evolbioinfo/bdpn}.
 
@@
 \section*{Author summary}
-Phylodynamic models can estimate epidemiological parameters (e.g., the number of secondary infections $R_e$) from pathogen phylogenetic trees (i.e., genealogies, estimated from sampled pathogen genomic sequences).
-These models typically assume that sampling of pathogen sequences from detected cases is independent between infected individuals.
-However this assumption does not hold for many epidemics, in particular for such sexually transmitted infections as HIV-1, for which partner notification (contact tracing) is included in health policies of many countries.
+Phylodynamic models can estimate epidemiological parameters such as the number of secondary infections $R_e$ from pathogen phylogenetic trees (i.e., genealogies, inferred from pathogen genomic sequences).
+These models do not account for partner notification (contact tracing), and instead assume that detection of new cases is independent between infected individuals.
+However, especially for sexually transmitted infections, partner notification plays an important role and is included in health policies of many countries.
 
@@
 It relies on models that bridge the gap between traditional epidemiology and sequence data by estimating such parameters as the basic reproduction number, $R_0$, from topology and branch lengths of pathogen phylogenies (i.e., genealogies of the pathogen population, approximating the transmission trees) combined with metadata on the samples.
-Classical epidemiological methods make inferences from incidence curves, which in the presence of import cases might be biased and give appearance of elevated local transmission.
+Classical epidemiological methods make inferences from incidence curves, which in the presence of imported cases might be biased and give appearance of elevated local transmission.
 In contrast, phylodynamic approach can overcome such biases as it uses the transmission structure as one of its information sources~\cite{vaughanEstimatesEarlyOutbreakspecific2024}.
-Using genomic data with phylodynamic estimations can provide valuable insights for preventing epidemic spreads.
+Using genomic data with phylodynamic estimations can provide valuable insights for preventing epidemic spread.
 
@@
 
-However, the spread and detection or an epidemic can be non-homogenous and depend on different factors, including governmental health policies (e.g., quarantine measures or development of pathogen detection tests), pathogen evolution over time (e.g., some SARS-CoV-2 variants are more transmissible than the others),  differences between host individuals (e.g., due to their immune system particularities or to their behaviour).
+However, the spread and detection of an epidemic can be non-homogenous and depend on different factors, including governmental health policies (e.g., quarantine measures or development of pathogen detection tests), pathogen evolution over time (e.g., some SARS-CoV-2 variants are more transmissible than the others),  differences between host individuals (e.g., due to their immune system particularities or to their behaviour).
 
@@
 Another important source of heterogeneity that none of these models accounts for is caused by non-independent sampling.
-One can set different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability between time intervals with the BDSKY, however, none of these models can account for correlation effects in sampling, in particular due to partner notification (PN).
+One can set different sampling probabilities for different types of individuals in the MTBD framework, or change of the sampling probability between time intervals with the BDSKY, however, none of these models can account for correlation effects in sampling, in particular due to partner notification (PN).
 PN (or contact tracing) is a process in which contacts of detected infected individuals are identified and offered testing for infection.
@@
 Under this model we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ (see \nameref{math}).
-The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g., by finding parameter values that maximize it, in maximum-likelihood framework).
+The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g., by finding parameter values that maximize it, in a maximum-likelihood framework).
 In particular, it permits the estimation of the following epidemiological parameters:
@@
 Upon $i$'s removal, their pathogen can be sampled (and hence observed as a tip in the transmission tree) with a probability $\rho$.
+
+Note that the BD model is asymptotically unidentifiable (see Remark 3.4 in~\cite{Stadler2009}).
+To become identifiable it requires one of its parameters to be fixed.
+In practice, it is often the sampling probability $\rho$, as it may be approximated from epidemiological data (e.g., the proportion of sampled cases among the declared ones) or the infectious time $\frac{1}{\psi}$ (estimated from observations of infected cases).
 
@@
 extension of the MTBD models that adds partner notification (PN).  
-In the partner notification process, an identified infected index case notifies their potentially infected contacts (e.g., sexual or drug injection partners for HIV), either on their own (patient referral) or with a hep of a health service provider (provider referral, when a health adviser notifies the partners without identifying the index patient; or contract referral, when if the index case fails to notify partners within a specified time period, the health adviser proceeds to provider referral)~\cite{wardPartnerNotification2014}.
+In the partner notification process, an identified infected index case notifies their potentially infected contacts (e.g., sexual or drug injection partners for HIV), either on their own (patient referral) or with a help of a health service provider (provider referral, when a health adviser notifies the partners without identifying the index patient; or contract referral, when if the index case fails to notify partners within a specified time period, the health adviser proceeds to provider referral)~\cite{wardPartnerNotification2014}.
 The notified partners are then provided with appropriate tests and treatment.
 Therefore, a successful detection of a new case B via partner notification by an index case A implies several conditions to be met: (i) A remembers the contact with B and  either notifies B on their own or gives the B's contacts to a health advisor for assisted notification; (ii) B is successfully reached and notified; (iii) B agrees to take a test; (iv) B is indeed infected.
-For instance, according to the HIV report from the Public Health England~\cite{aHIVUnitedKingdom2014}, in England in 2013, reaching the stage (ii) had an estimated probability of about 50\%; reaching the stage (iii) -- of about 41\% (as 83\% of all the patients who attended a clinic through PN had a reported HIV test); and the stage (iv) -- of less than 3\%.
+For instance, according to the report from the Public Health England~\cite{aHIVUnitedKingdom2014}, in England in 2013 for HIV, reaching the stage (ii) had an estimated probability of about 50\%; reaching the stage (iii) -- of about 41\% (as 83\% of all the patients who attended a clinic through PN had a reported HIV test); and the stage (iv) -- of less than 3\%.
 
@@
 
-For simplicity, a positive draw of $\upsilon$ models a slightly modified scenario: (i) A remembers the contact with B and  either notifies B on their own or gives the B's contacts to a health advisor for assisted notification; (iv') B is indeed infected, moreover, \textit{the infection of either A or B is caused by the pathogen transmission during their contact};  and either (ii) B is successfully reached and notified and (iii) agrees to take a test; or (ii') \textit{by the time of notification, B is already removed by other means} (already got sampled or stopped being infectious without sampling, e.g., died).
-
-The condition (iv') in contrast to (iv) excludes the possibility of notifying partners if there was no pathogen transmission during the contact with them.
+For simplicity, a positive draw with probability $\upsilon$ models a slightly modified scenario: (i) A remembers the contact with B and  either notifies B on their own or gives the B's contacts to a health advisor for assisted notification; (ii', similar to iv in the previous scenario) B is indeed infected, moreover, \textit{the infection of either A or B is caused by the pathogen transmission during their contact};  and either (iii', same as ii) B is successfully reached and notified and (iv', same as iii) agrees to take a test (which at the end of the considered time period might not yet be available in the database); or (iii'') \textit{by the time of notification, B is already removed by other means} (already got sampled or stopped being infectious without sampling, e.g., died).
+
+The condition (ii') in contrast to (iv) excludes the possibility of notifying partners if there was no pathogen transmission during the contact with them (and thus mostly focuses on regular partners).
 We additionally simplify the model by allowing to notify only the last such partner.
@@
 \begin{itemize}
- \item $\lambda$ -- constant transmission rate: an infected individual transmits their pathogen to a newly infected individual (which corresponds to an internal node in the transmission tree);
- \item $\psi$ -- constant removal rate: an infected individual stops being infectious at this rate. This corresponds to a tip in the transmission tree (observed or hidden);
- \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
- \item $\upsilon$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
+ \item $\lambda$ -- constant transmission rate: an infected individual transmits their pathogen to a newly infected individual with this rate. This leads to an internal node in the transmission tree;
+ \item $\psi$ -- constant removal rate: an infected individual stops being infectious at this rate. This leads to a tip in the transmission tree (observed or hidden);
+ \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability. This leads to a tip in the transmission tree being observed;
+ \item $\upsilon$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability.
+This leads to the removal rate change for the partner branch in the transmission tree;
  \item $\phi >> \psi$ -- constant notified sampling rate: a notified individual (if not yet removed via the standard procedure) gets removed \textit{and observed} with this rate.
-This creates an observed tip in the transmission tree.
+This leads to an observed tip in the transmission tree.
 \end{itemize}
@@
 
+In the case of sexually transmitted diseases, notifying only the most recent partner could roughly correspond to notifying the ``main'' partner (e.g., spouse).
 This assumption might need to be removed or relaxed (e.g., to up to a certain fixed number of most recent partners, or all the partners within a certain time interval) in the future work.
 It is however useful to make the mathematics and likelihood calculation under the model manageable.
-In the case of sexually transmitted diseases, notifying only the most recent partner could roughly correspond to notifying the ``main'' partner (e.g., spouse).
 
@@
 
+\smallskip
+
 The -PN extension can be generalized from BD to MTBD models: at the moment of sampling, an individual of any type can notify their last partner with the probability $\upsilon$, whose removal rate will then get replaced by  a constant sampling rate $\phi >> \psi_k \; \forall k \in \{1, \ldots, m\}$.
@@
 We give its algorithmic details in~\nameref{S1_Appendix}.
-The results in section~"\nameref{sec:test}" show that the trees generated under MTBD and MTBD-PN models are different and easily distinguishable.
+The results in section ``\nameref{sec:test}'' show that the trees generated under MTBD and MTBD-PN models are different and easily distinguishable.
 
 Using our simulator, we generated trees under three classical MTBD-family models: BD, BDEI (Birth-Death Exposed-Infectious~~\cite{Stadler2014}) and BDSS (Birth-Death with Super Spreading~\cite{Stadler2013a}), and under their -PN versions: BD-PN, BDEI-PN and BDSS-PN models.
-(See Fig.~1 in \cite{Voznica2021} for a graphical representation of BD, BDEI and DBSS models and their parameters).
+(One can see Fig.~1 in \cite{Voznica2021} for a graphical representation of BD, BDEI and DBSS models and their parameters).
 
@@
 We then calculate the time differences $d_{i'}$ between the tips of these reshuffled cherries.
-The test then compares $d_i$ to $d_{i'}$: $sign(d_{i'} - d_i) = 1$ if $d_{i'}$ is larger than $d_i$; $= -1$ if it is smaller; or $= 0$ if they are equal (this could happen as in real datasets dates are often truncated, e.g., to month or even a year).
+The test then compares $d_i$ to $d_{i'}$: $sign(d_{i'} - d_i) = 1$ if $d_{i'}$ is larger than $d_i$; $= -1$ if it is smaller; or $= 0$ if they are equal (this could happen as in real datasets dates are often truncated, e.g., to a month or even a year).
 Note that if cherry tip branch lengths are independently drawn from the same distribution, $d_{i'}$ will be smaller than $d_i$ for about half of the cherries.
@@
 
-We implemented the PN-test in Python 3. It is available as a part of BDPN parameter estimator ($pn\_test$, see section~"\nameref{sec:tool}").
+We implemented the PN-test in Python 3. It is available as a part of BD-PN parameter estimator ($pn\_test$, see section~``\nameref{sec:tool}'').
 
@@
 With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch $i$ till time $t_i$, starting on this branch at time $t \leq t_i$.
-The master equations for the BD models were initially developed by Stadler \textit{et al.}~\cite{Stadler2009}, however what we present here is their branch-specific formulation that we proposed in~\cite{zhukovaFastAccurateMaximumLikelihood2022}:
+The master equations for the BD model were initially developed by Stadler \textit{et al.}~\cite{Stadler2009}, however what we present here is their branch-specific formulation that we proposed in~\cite{zhukovaFastAccurateMaximumLikelihood2022}:
 In our formulation, ${p}^{(i)}(t)$ describes an evolution along a tree branch $i$, without taking into account the branch's subtree.
@@
 
-Importantly, the BD model is asymptotically unidentifiable (see Remark 3.4 in~\cite{Stadler2009}).
-To become identifiable it requires one of its parameters to be fixed.
-In practice, it is often the sampling probability $\rho$, as it may be approximated from epidemiological data (e.g., the proportion of sampled cases among the declared ones) or the infectious time $\frac{1}{\psi}$ (estimated from observations of infected cases).
-
 \subsubsection*{Master equation under BD-PN model}
@@
 These two individuals could be different if a hidden transmission where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.
-$p^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along such a tree branch, integrating over all the possibilities, starting from no hidden transmission along the branch and including any number of hidden transmissions with either donor or recipient tree staying unsampled (and hence potentially different individuals at times $t$ and the branch end $t_i$).
+$p^{(i)}(t)$ in equation~(\ref{eq:p}) describes an evolution along such a tree branch, integrating over all the possibilities, starting from no hidden transmission along the branch and including any number of hidden transmissions with either donor or recipient tree staying unsampled (and hence potentially different individuals at times $t$ and the branch end $t_i$).
 
@@
 
-\paragraph{Additional master equations under BD-PN models.}
+\paragraph{Additional master equations under BD-PN model.}
 
@@
 The (observed as in \textbf{1-2} or hidden as in \textbf{3}) internal node AB corresponds to the transmission between A and B. 
-The fact that A is the notifier of B implies that there is no hidden event along its branch AB-A (violet), and that in any transmission on the path AB-B (orange) the donor is B. 
+The fact that A is the notifier of B implies that there is no hidden event along its branch AB-A (violet), and that in any transmission on the path AB-B (orange or dark-red) the donor is B. 
 Note, that since the most-recent-partner relationship is not always symmetric, B might be the most recent partner of more than one individual and hence might have got notified by someone else too (e.g., by C in \textbf{b}, who in \textbf{1b,2b} was infected by B, and in \textbf{3b} could have been infected by or have infected B, who have then infected A).
@@
 Therefore the A's branch is composed of two parts, where the bottom part (violet) corresponds to A and contains no hidden transmissions, while the top part could correspond to A or not  (e.g in \textbf{3b} it corresponds to B) and might contain hidden transmissions.
-In the panels \textbf{2a-b,3a-b} B is hidden as by the end of the sampling period ($t=T$) B is not yet sampled, while in \textbf{2-3c} B is hidden as B got removed via the standard procedure without sampling before any notification.
+In the panels \textbf{2a-b,3a-b} B is hidden as by the end of the sampling period ($t=T$) B is not yet sampled, while in \textbf{2c,3c} B is hidden as B got removed via the standard procedure without sampling before any notification.
 \\
@@
 
-We will use a \textbf{notifier branch probability} $p_n^{(A)}(t)$ for such cases (when no hidden transmission occurred along the branch A, coloured violet in Fig~\ref{fig:pn-branches}):
+We will use a \textbf{notifier branch probability} $p_n^{(A)}(t)$ for such cases (when no hidden transmission occurred along the branch A, coloured violet in Fig.~\ref{fig:pn-branches}):
 
@@
 However, depending on whether B already got notified ($t \geq t_A$ in Fig.~\ref{fig:pn-branches}(1a)) or not yet ($t < t_A$ in Fig.~\ref{fig:pn-branches}(1a)), B's removal rate is different (is correspondingly $\phi$ or $\psi$).
-We will use a \textbf{partner branch probability before notification} $p_{pb}^{(x)}(t)$ for cases when the individual at the end of the branch (or considered time interval, here $t_x=t_A$) is the same as at time $t$, has not yet been notified, but might have transmitted to someone else (unobserved) along the way (coloured orange in Fig~\ref{fig:pn-branches}):
+We will use a \textbf{partner branch probability before notification} $p_{pb}^{(x)}(t)$ for cases when the individual at the end of the branch (or considered time interval, here $t_x=t_A$) is the same as at time $t$, has not yet been notified, but might have transmitted to someone else (unobserved) along the way (coloured orange in Fig.~\ref{fig:pn-branches}):
 
@@
 
-We will use a \textbf{partner branch probability after notification} $p_{pa}^{(i)}(t)$ for cases when the individual at the end of the branch is the same as at time $t$, by time $t$ has already been notified, and might have transmitted to someone else (unobserved) along the way (coloured dark red in Fig~\ref{fig:pn-branches}):
+We will use a \textbf{partner branch probability after notification} $p_{pa}^{(i)}(t)$ for cases when the individual at the end of the branch is the same as at time $t$, by time $t$ has already been notified, and might have transmitted to someone else (unobserved) along the way (coloured dark red in Fig.~\ref{fig:pn-branches}):
 
@@
 
-\item[1c.] by the time $t_A$, B was already sampled via a standard procedure (at time $t_B < t_A$, hence the branch B would finish before the branch A, as in Fig~\ref{fig:pn-branches}(1c)).
+\item[1c.] by the time $t_A$, B was already sampled via a standard procedure (at time $t_B < t_A$, hence the branch B would finish before the branch A, as in Fig.~\ref{fig:pn-branches}(1c)).
 
@@
 Note that if A is a notifier, A is always observed, as notification only happens upon sampling.
-However, B could have stay unsampled: either because once notified (by A or another partner, who notified them before A did) B has not yet got sampled (Fig~\ref{fig:pn-branches}(2a-b,3a-b)); or because B got removed via the standard procedure without sampling before any notification (Fig.~\ref{fig:pn-branches}(2c,3c)).
+However, B could have stay unsampled: either because once notified (by A or another partner, who notified them before A did) B has not yet got sampled (Fig.~\ref{fig:pn-branches}(2a-b,3a-b)); or because B got removed via the standard procedure without sampling before any notification (Fig.~\ref{fig:pn-branches}(2c,3c)).
 Moreover, between the transmission AB and the moment of B's notification/removal, B might have transmitted to someone else (e.g., C in Fig.~\ref{fig:pn-branches}(2), BC denoting the transmission).
@@
 
-To model the cases when B is not sampled, we will need to describe the probability of a\textbf{ mixed branch} $p_{m(s1,s2)}^{(i)}(t,t_r)$ for branches that include a hidden partner subtree (AB-C branches in Fig.~\ref{fig:pn-branches}(2) and A's tip branch in Fig.~\ref{fig:pn-branches}(3)).
+To model the cases when B is not sampled, we will need to describe the probability of a \textbf{mixed branch} $p_{m(s1,s2)}^{(i)}(t,t_r)$ for branches that include a hidden partner subtree (AB-C branches in Fig.~\ref{fig:pn-branches}(2) and A's tip branch in Fig.~\ref{fig:pn-branches}(3)).
 $s1$ here denotes the type of the top of the branch (e.g., between AB and BC in Fig.~\ref{fig:pn-branches}(2), partner) and $s2$ denotes the type of the bottom of the branch (e.g., between AB and A in Fig.~\ref{fig:pn-branches}(3), notifier).
 Let us denote the first notification time of the hidden partner as $t_r$, and the time of the hidden partner subtree start as $t_h$, while as usual $t_i$ stands for the time at the branch end.
-For instance, in Fig~\ref{fig:pn-branches}(2a) $t_i=t_C$, $t_r=t_A$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(2b) $t_i=t_r=t_C$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(3c) $t_i=t_r=t_A$, $t_h=t_{AB}$.
+For instance, in Fig.~\ref{fig:pn-branches}(2a) $t_i=t_C$, $t_r=t_A$, $t_h=t_{BC}$; in Fig.~\ref{fig:pn-branches}(2b) $t_i=t_r=t_C$, $t_h=t_{BC}$; in Fig.~\ref{fig:pn-branches}(3c) $t_i=t_r=t_A$, $t_h=t_{AB}$.
 
 $p_{m(s1,s2)}^{(i)}(t,t_r)$ combines three elements: the probability of the top part of the branch $p_{top=s1}^{(h)}(t)$, which finishes with a transmission to or from the hidden partner, the probability of the bottom part of the branch $p_{bottom=s2}^{(i)}(t_h)$ (without taking into account the event at its end) and the probability of the hidden partner subtree $U_p(t_h,t_r)$.
-We describe the details behind approximation of $p_{m(s1,s2)}^{(i)}(t,t_r)$ in~\nameref{S2_Appendix}, and give the result below:
-
-\begin{equation}
-\begin{split}
-p_{m(s1,s2)}^{(i)}(t,t_r) &= p_{top=s1}^{(h)}(t, t_r)\frac{U_p(t_h,t_r)}{U(t_h)}p_{bottom=s2}^{(i)}(t_h)\textit{\color{gray},}\\
-&\textit{\color{gray}where $t_{h} \approx t + \frac{(t_i - t)}{2}$ is an approximation of the time}\\
-&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~of the hidden partner subtree start,}\\
-&\textit{\color{gray}~~~~~~~~$t_{r}$ is the earliest notification time of the (first) partner}\\
-&U_p(t_h, t_r) \approx 
-\begin{cases}
-p_{pb}^{(r)}(t_h) \frac{1 - e^{-\psi(t_r - t_h)}}{e^{-\psi(t_r - t_h)}} (1 - \rho) + p_{pb}^{(r)}(t_h) p_{pa}^{(x)}(t_r) & \textit{\color{gray}if $t_h < t_r$}\\
-p_{pa}^{(x)}(t_h) & \textit{\color{gray}if $t_h \geq t_r$}\\
-\end{cases}\\
-&\textit{\color{gray}~~~~~~~where $t_x =T$}\\
-&\\
-& p_{top=s1}^{(h)}(t, t_r) \approx 
-\begin{cases}
-p^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}&\textit{\color{gray} if $s1=-$ (standard state)}\\
-& \\
-p_{pb}^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}&\textit{\color{gray} if $s1=p$ (partner) and}\\
-&\textit{\color{gray} ~~ $t_h \leq t_r$}\\
-p_{pb}^{(r)}(t)\Big(p_{pa}^{(h)}(t_r) -e^{-(\lambda + \phi)(t_{h} - t_r)}\Big)&\textit{\color{gray} if $s1=p$, $t \leq t_r \leq t_h$}\\
-&\\
-p_{pa}^{(h)}(t) -e^{-(\lambda + \phi)(t_{h} - t)}&\textit{\color{gray} if $s1=p$ and $t > t_r$}\\
-\end{cases}\\
-&\\
-& p_{bottom=s2}^{(i)}(t_h) = 
-\begin{cases}
-p_n^{(i)}(t_h)&\textit{\color{gray} if $i$ is a notifier tip ($s2=n$)}\\
-p^{(i)}(t_h)&\textit{\color{gray} otherwise ($s2=-$, the standard state)}\\
-\end{cases}\\
-&\\
-p_{m(p,-,n)}^{(i)}(t,t_r) &\approx p_{top=p}^{(h1)}(t, t_r)\frac{U_p(t_{h1},t_{r})}{U(t_{h1})} p_{top=-}^{(h2)}(t_{h1}, t_i)\frac{U_p(t_{h2},t_{i})}{U(t_{h2})} p_n^{(i)}(t_{h2})\textit{\color{gray},}\\
-&\textit{\color{gray}where $t_{h1} \approx t + \frac{1}{3}(t_i - t)$,}\\
-&\textit{\color{gray}~~~~~~~~$t_{h2} \approx t + \frac{2}{3}(t_i - t)$,}\\
-&\textit{\color{gray}~~~~~~~~$t_{r}$ is the earliest notification time of the partner }\\
-&\textit{\color{gray}~~~~~~~~~~~~~~~~~~hidden at the top of the branch.}\\
-\end{split}\label{eq:p_mixed}
-\end{equation}
+We describe the details behind calculation of $p_{m(s1,s2)}^{(i)}(t,t_r)$ in~\nameref{S2_Appendix}.
+%, and give the result below:
+
+%\begin{equation}
+%\begin{split}
+%p_{m(s1,s2)}^{(i)}(t,t_r) &= p_{top=s1}^{(h)}(t, t_r)\frac{U_p(t_h,t_r)}{U(t_h)}p_{bottom=s2}^{(i)}(t_h)\textit{\color{gray},}\\
+%&\textit{\color{gray}where $t_{h} \approx t + \frac{(t_i - t)}{2}$ is an approximation of the time}\\
+%&\textit{\color{gray}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~of the hidden partner subtree start,}\\
+%&\textit{\color{gray}~~~~~~~~$t_{r}$ is the earliest notification time of the (first) partner}\\
+%&U_p(t_h, t_r) \approx 
+%\begin{cases}
+%p_{pb}^{(r)}(t_h) \frac{1 - e^{-\psi(t_r - t_h)}}{e^{-\psi(t_r - t_h)}} (1 - \rho) + p_{pb}^{(r)}(t_h) p_{pa}^{(x)}(t_r) & \textit{\color{gray}if $t_h < t_r$}\\
+%p_{pa}^{(x)}(t_h) & \textit{\color{gray}if $t_h \geq t_r$}\\
+%\end{cases}\\
+%&\textit{\color{gray}~~~~~~~where $t_x =T$}\\
+%&\\
+%& p_{top=s1}^{(h)}(t, t_r) \approx 
+%\begin{cases}
+%p^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}&\textit{\color{gray} if $s1=-$ (standard state)}\\
+%& \\
+%p_{pb}^{(h)}(t) -e^{-(\lambda + \psi)(t_{h} - t)}&\textit{\color{gray} if $s1=p$ (partner) and}\\
+%&\textit{\color{gray} ~~ $t_h \leq t_r$}\\
+%p_{pb}^{(r)}(t)\Big(p_{pa}^{(h)}(t_r) -e^{-(\lambda + \phi)(t_{h} - t_r)}\Big)&\textit{\color{gray} if $s1=p$, $t \leq t_r \leq t_h$}\\
+%&\\
+%p_{pa}^{(h)}(t) -e^{-(\lambda + \phi)(t_{h} - t)}&\textit{\color{gray} if $s1=p$ and $t > t_r$}\\
+%\end{cases}\\
+%&\\
+%& p_{bottom=s2}^{(i)}(t_h) = 
+%\begin{cases}
+%p_n^{(i)}(t_h)&\textit{\color{gray} if $i$ is a notifier tip ($s2=n$)}\\
+%p^{(i)}(t_h)&\textit{\color{gray} otherwise ($s2=-$, standard state)}\\
+%\end{cases}\\
+%&\\
+%p_{m(p,-,n)}^{(i)}(t,t_r) &\approx p_{top=p}^{(h1)}(t, t_r)\frac{U_p(t_{h1},t_{r})}{U(t_{h1})} p_{top=-}^{(h2)}(t_{h1}, t_i)\frac{U_p(t_{h2},t_{i})}{U(t_{h2})} p_n^{(i)}(t_{h2})\textit{\color{gray},}\\
+%&\textit{\color{gray}where $t_{h1} \approx t + \frac{1}{3}(t_i - t)$,}\\
+%&\textit{\color{gray}~~~~~~~~$t_{h2} \approx t + \frac{2}{3}(t_i - t)$,}\\
+%&\textit{\color{gray}~~~~~~~~$t_{r}$ is the earliest notification time of the partner }\\
+%&\textit{\color{gray}~~~~~~~~~~~~~~~~~~hidden at the top of the branch.}\\
+%\end{split}\label{eq:p_mixed}
+%\end{equation}
+
+\smallskip
 
@@
 
-Using the formulas~(\ref{eq:ps}) and (\ref{eq:p_mixed}), we can calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}.
+Using the branch probability formulas described above% ~(\ref{eq:ps}) and (\ref{eq:p_mixed})
+, we can calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}.
 For each visited node $i$ we calculate the likelihood density of its subtree including the branch $ji$ connecting $i$ to its parent node $j$ for two cases depending on $i$'s receiving-a-notification status (shown with the first superscript, $^{\not{n}}$ or $^n$).
@@
 
-\subsection*{UK HIV-1 B tree reconstruction and sampling probability estimation}\label{uk_tree}
+\subsection*{Zurich men-having-sex-with-men HIV-1 B tree and sampling probability}\label{zurich_tree}
+We applied our estimator to asses the partner notification for HIV-1 subtype B epidemic among men-having-sex-with-men (MSM) in Zurich.
+We used the phylogenetic tree of 200 samples reconstructed by Rasmussen \textit{et al.}~\cite{Rasmussen2017} from the sequences collected as a part of the Swiss Cohort Study~\cite{swisshivcohortstudyCohortProfileSwiss2010a} between 1988 and 2014.
+This tree was also previously analyzed by Voznica~\textit{et al.}~\cite{Voznica2021} using a Birth-Death model with SuperSpreading (BDSS).
+Following Voznica~\textit{et al.}~\cite{Voznica2021}, we fixed the sampling probability for this tree at 25\%: $\rho=0.25$.
+
+\subsection*{UK HIV-1 B forest reconstruction and sampling probability estimation}\label{uk_tree}
 To asses the partner notification in HIV-infected patients in the UK, we used the phylogenetic tree from our recent study of HIV drug resistance in the UK~\cite{zhukovaModelingDrugResistance2023}.
@@
 The samples used for the tree reconstruction were obtained from the UK HIV Drug resistance database~\cite{Dunn2007}.
-The sampling date metadata for these samples were month-specific (e.g., Aug 2015), hence potentially creating simultaneous sampling dates for cases, which were sampled on different days of the month.
+The sampling date metadata for these samples were expressed in months (e.g., Aug 2015), hence potentially creating simultaneous sampling dates for cases that were sampled on different days of the month.
 To avoid this issue we randomly selected a date within the specified month for each sample, and time-scaled the tree with LSD2~\cite{To2016} (v2.4.1, strict molecular clock with outlier removal).
-We repeated this time-scaling procedure 10 times, obtaining 10 trees with slightly different branch lengths and root dates between Oct 5 and Nov 11, 1962.
-
-Between 2012 and 2015, in the UK ART of HIV-infected individuals  was initiated when their CD4 count dropped below 350 cells/mL~\cite{williamsBritishHIVAssociation2012}.
+We repeated random date resolution and time-scaling 10 times, obtaining 10 trees with slightly different branch lengths and root dates between Oct 5 and Nov 11, 1962.
+
+Between 2012 and 2015, in the UK antiretroviral treatment (ART) of HIV-infected individuals  was initiated when their CD4 count dropped below 350 cells/mL~\cite{williamsBritishHIVAssociation2012}.
 However ``if a patient with a CD4 cell count $>$ 350 cells/mL wishes to start ART to reduce the risk of transmission to partners, this decision is respected and ART is started''~\cite{williamsBritishHIVAssociation2012}.
@@
 We hence estimated that in 2015 67\,298 % (64\,837, 70\,291)
-people were living with HIV-B. Our trees contained 39\,047 tips sampled by the end of 2015, and therefore account for about 58\% %(56\%, 60\%) 
+people were living with HIV-B. Our trees contained 39\,047 tips sampled by the end of 2015, and therefore account for about 58\%  ($=39\,047/67\,298 \cdot 100\%$) %(56\%, 60\%) 
 of the total number of individuals living with HIV-B in the UK.
@@
 \subsubsection*{PN test is highly specific and sensitive at detecting partner notification}~\label{sec:test}
-To illustrate the PN test performance, we applied it to the six 100-tree simulated datasets, corresponding to the three classical MTBD-family models: BD, BDEI (Birth-Death Exposed-Infectious~~\cite{Stadler2014}) and BDSS (Birth-Death with Super Spreading~\cite{Stadler2013a}), and to their -PN versions: BD-PN, BDEI-PN and BDSS-PN (see~"\nameref{sec:sim}" for dataset details).
+To illustrate the PN test performance, we applied it to the six 100-tree simulated datasets, corresponding to the three classical MTBD-family models: BD, BDEI (Birth-Death Exposed-Infectious~~\cite{Stadler2014}) and BDSS (Birth-Death with Super Spreading~\cite{Stadler2013a}), and to their -PN versions: BD-PN, BDEI-PN and BDSS-PN (see~``\nameref{sec:sim}'' for dataset details).
 
@@
 These results are summarised in Table~\ref{tbl:pntest}.
-Our test therefore showed both high specificity and sensitivity.
+Our test therefore showed both high specificity and sensitivity (99\% and 94\% respectively on the combined, 600-tree, data set).
 
@@
 
-We assessed the performance of our maximum-likelihood estimator for the BD-PN model bdpn on the 100 trees from the the BD-PN simulated data set (see~"\nameref{sec:sim}" for dataset details).
+We assessed the performance of our maximum-likelihood estimator for the BD-PN model bdpn on the 100 trees from the the BD-PN simulated data set (see ``\nameref{sec:sim}'' for dataset details).
 
@@
 %The results with the becoming-uninfectious rate $\psi$ fixed are similar and given in the Supplementary Material.
-As expected with a maximum-likelihood estimator, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the true parameter values for all the trees in both settings.
+%As expected with a maximum-likelihood estimator, 
+The tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the true parameter values for all the trees in both settings.
 
@@
 We compared the method accuracy by a paired z-test.
-\textbf{On the BD data set}, bd and bdpn estimators performed equally well in estimation $R_0$ and infectious time $\frac{1}{\psi}$, with an average relative error $\leq 5\%$, and average relative bias $\leq 1\%$.
+\textbf{On the BD data set}, bd and bdpn estimators performed equally well in estimating $R_0$ and infectious time $\frac{1}{\psi}$, with an average relative error $\leq 5\%$, and average relative bias $\leq 1\%$.
 The partner notification probability $\upsilon$ estimated on these trees by bdpn was systematically low, with the maximum value of 0.015, and mean value of 0.001.
@@
 
+Finally, we performed a likelihood ratio test on the BD and BD-PN trees with the parameters estimated by the bd and bdpn estimators: $-2 (l_{BD} - l_{BDPN}) \lessgtr 5.99$ (where 5.99 is the value of chi-squared distribution with 2 degrees of freedom corresponding to the significance level of $0.95$).
+For all the 200 trees the test selected the correct model.
+
 \subsubsection*{Application 1: Men-having-sex-with-men (MSM) HIV-1 B epidemic in Zurich}
-We applied our estimator to asses the partner notification for HIV-1 subtype B epidemic among men-having-sex-with-men (MSM) in Zurich.
-We used the phylogenetic tree of 200 samples reconstructed by Rasmussen \textit{et al.}~\cite{Rasmussen2017} from the sequences collected as a part of the Swiss Cohort Study~\cite{swisshivcohortstudyCohortProfileSwiss2010a} between 1988 and 2014.
-This tree was also previously analyzed by Voznica~\textit{et al.}~\cite{Voznica2021} using a Birth-Death model with SuperSpreading (BDSS).
-
-The PN test detected partner notification in this tree (p-value=0.014).
-Following Voznica~\textit{et al.}~\cite{Voznica2021}, we fixed the sampling probability for this tree at 25\%: $\rho=0.25$.
-Our bdpn estimator inferred the partner notification probability $\upsilon=0.09$ CI $[0.008-0.226]$,
+We applied our estimator to asses the partner notification for HIV-1 subtype B epidemic among men-having-sex-with-men (MSM) in Zurich, using a 200-sample time-scaled phylogenetic tree from the study by Rasmussen \textit{et al.}~\cite{Rasmussen2017}, and fixing the sampling probability to 25\%: $\rho=0.25$ (see ``\nameref{zurich_tree}'' for more details).
+
+The PN test detected partner notification in this tree (p-value $=0.014$). Our bdpn estimator inferred the partner notification probability $\upsilon=0.09$ CI $[0.008-0.226]$,  
 $R_e = 1.43$ $[1.26-1.66]$, infectious time $\frac{1}{\psi} = 8$ $[6-11]$ years, and 4.9 months [6 days -- 1.4 years] between partner notification and their sampling.
-These results globally agree (CIs intersect) with the $R_e$ estimate using a coalescent model from Rasmussen \textit{et al.}~\cite{Rasmussen2017} (between 1.0 and 2.5), and the estimates using the BDSS model from Voznica~\textit{et al.}~\cite{Voznica2021}: $R_e$ of 1.6 and 1.7 and an infectious period of 10.2 and 9.8 years (depending on the tree representation used for the deep learner).
+These results globally agree (CIs intersect) with the $R_e$ estimate for this data using a coalescent model from Rasmussen \textit{et al.}~\cite{Rasmussen2017} (between 1.0 and 2.5), and the estimates using the BDSS model from Voznica~\textit{et al.}~\cite{Voznica2021}: $R_e$ of 1.6 and 1.7 and an infectious period of 10.2 and 9.8 years (depending on the tree representation used for the deep learner).
 
@@
 
-For comparison, parameter estimation on the same tree with the BD model gave slightly lower estimate of $R_e$ and a slightly shorter infectious time, though with intersecting CIs ($R_e = 1.32$ $[1.19-1.47]$, $\frac{1}{\psi} = 7$ $[6-9]$ years), which is in line with underestimation of $R_e$ and $\frac{1}{\psi}$ with bd on the BD-PN simulated data.
-The likelihood ratio test suggests the BDPN model: loglikelihood under the BD-PN-estimated parameters was $l_{BDPN} = -1\,140$ vs.
-$l_{BD}=-1\,146$ under the BD parameters, which gives the ratio $-2 (l_{BD} - l_{BDPN}) =	12 > 5.99 \approx \chi^2_2(0.95)$ (the value of chi-squared distribution with 2 degrees of freedom corresponding to the significance level of $0.95$).
+For comparison, parameter estimation on the same tree with the BD model gave a slightly lower estimate of $R_e$ and a slightly shorter infectious time, though with intersecting CIs ($R_e = 1.32$ $[1.19-1.47]$, $\frac{1}{\psi} = 7$ $[6-9]$ years), which is in line with underestimation of $R_e$ and $\frac{1}{\psi}$ with bd on the BD-PN simulated data.
+The likelihood ratio test suggested the BD-PN model: loglikelihood under the BD-PN-estimated parameters was $l_{BDPN} = -1\,140$ vs.
+$l_{BD}=-1\,146$ under the BD parameters, which gives the ratio of $-2 (l_{BD} - l_{BDPN}) =	12 > 5.99 \approx \chi^2_2(0.95)$.
+% (the value of chi-squared distribution with 2 degrees of freedom corresponding to the significance level of $0.95$).
 % unrealistic results: $R_e = 0.5$ (suggesting contained epidemic) and very short infectious time $\frac{1}{\psi} = 0.4$ years.
@@
 \subsubsection*{Application 2: HIV-1 B epidemic in the UK}
-We also applied our estimator to asses the partner notification in HIV-infected individuals in the UK.
-We used the phylogenetic tree from our recent study of HIV drug resistance in the UK~\cite{zhukovaModelingDrugResistance2023}, reconstructed from HIV-1 B samples of 40\,055 individuals sampled between 1996 and 2016 and obtained from the UK HIV Drug resistance database~\cite{Dunn2007}.
-
-The sampling dates in the UK HIV Drug resistance database are month-specific, hence to avoid potentially creating simultaneous sampling dates for cases, which were sampled on different days of the month, we randomly selected a date within the specified month for each sample before time-scaling the tree.
-We repeated the time-scaling procedure 10 times, obtaining 10 trees with slightly different branch lengths and root dates between Oct 5 and Nov 11, 1962 (see~"\nameref{uk_tree}").
-We cut this tree between 2012 and 2015, as this period featured a homogeneous ART start policy in the UK (CD4 cell count below 350 cells/mL~\cite{williamsBritishHIVAssociation2012}).
-We then applied the PN-test and estimated the BD-PN parameters on these 2012--2015 forests.
-
-The PN-test values were below 0.05 for all the forests (values between $1 \cdot 10^{-18}$ and $2 \cdot 10^{-14}$), strongly suggesting the presence of partner notification.
-Using sampling probability $\rho=0.58$ (see~"\nameref{uk_tree}" for its estimation), we estimated the partner notification probability $\upsilon=0.006-0.008$, %the following parameter values: $\lambda=0.53$, $\psi=0.43$, $\phi=23-59$, $\upsilon=0.006-0.007$, which corresponds to
+We also applied our estimator to asses the partner notification in HIV-1 subtype B infected individuals in the UK between 2012 and 2015.
+This time period was chosen to represent a homogeneous treatment policy  (treatment start at CD4 count below 350 cells/mL~\cite{williamsBritishHIVAssociation2012}).
+We used 10 forests reconstructed from HIV-1 B sequences from the UK HIV Drug resistance database~\cite{Dunn2007} (see ``\nameref{uk_tree}'' for details).
+
+The PN-test values were below 0.05 for all the forests (p-values between $1 \cdot 10^{-18}$ and $2 \cdot 10^{-14}$), strongly suggesting the presence of partner notification.
+Using sampling probability $\rho=0.58$ (see ``\nameref{uk_tree}'' for its estimation), we estimated the partner notification probability $\upsilon=0.006-0.008$, %the following parameter values: $\lambda=0.53$, $\psi=0.43$, $\phi=23-59$, $\upsilon=0.006-0.007$, which corresponds to
 $R_e = 1.2-1.3$, infectious time $\frac{1}{\psi} = 2.3-2.4$ years, and 7--11 days between partner notification and their sampling.
-The results were consistent between 10 forests and shown in Table~\ref{tbl:uk}.
+The results were consistent between 10 forests and are shown in Table~\ref{tbl:uk}.
 %The results for $\rho=0.4$ and for $\rho=0.6$ were similar: $R_e$ of respectively $1.5$ and
@@
 
-In the UK 95\% of infected population is on antiretroviral treatment, which, when taken properly, prevents transmission.
-Hence, the average infectious time represents the average time before starting treatment.
+In the UK in 2015 the majority of infected population (72\%~\cite{aHIVUnitedKingdom2016}) was on antiretroviral treatment, which, when taken properly, prevents transmission.
+Hence, the average infectious time roughly represents the average time before starting treatment.
 Our estimate is approximately 2.3 years.
+According to the HIV report from the Public Health England~\cite{aHIVUnitedKingdom2014}, in England in 2013, less than 3\% of sampled infected individuals were detected though partner notification.
+Our estimate is even lower, potentially because as our model does not allow for notifying partners during the contact with whom there was no transmission, nor notifying multiple partners.
 
@@
 
-For comparison, parameter estimation on the same forests with the BD model gave similar results ($R_e = 1.2$, $\frac{1}{\psi} = 2.3$ years), however the likelihood ratio test suggests the BDPN model.
+For comparison, parameter estimation on the same forests with the BD model gave similar results ($R_e = 1.2$, $\frac{1}{\psi} = 2.3$ years), however the likelihood ratio test suggests the BD-PN model.
 For instance, on forest 1, loglikelihood under the BD-PN-estimated parameters was $l_{BDPN} = -28\,984$ vs.
@@
 %The advantage of the BD-PN model is that the differential equations describing it have closed-form solutions, which are quick to calculate and do not require numerical approximation (as is the case for more complex MTBD models).
-Tests on simulated data show that our estimator robustly infers model parameter values and their CIs in the setting when the duration of the infectious period $\frac{1}{\psi}$ or the base sampling probability $\rho$ is fixed.
-Fixing one of the model parameters is needed for model identifiability.
-In practice, $\rho$ and $\frac{1}{\psi}$ may often be approximated from epidemiological data, e.g., as the proportion of sampled cases among the declared ones (for $\rho$), or estimated from observations of infected cases ($\frac{1}{\psi}$).
+Tests on simulated data show that our estimator robustly infers model parameter values and their CIs in the setting when %the duration of the infectious period $\frac{1}{\psi}$ or 
+the base sampling probability $\rho$ is fixed.
+Fixing one of the model parameters is needed for model identifiability and in practice, $\rho$ may often be approximated from epidemiological data, e.g., as the proportion of sampled cases among the declared ones.
 
@@
 We estimated a modest partner notification probability ($9\%$ in Zurich, $1\%$ in the UK), and rapid notified partner detection ($\approx 5$ months vs $\approx  8$ years for detection under the standard procedure in Zurich; and $\approx 8$ days vs $\approx 2.3$ years for detection under the standard procedure in the UK).
+Note that the notification probability represents ``successful notification'', which combines several conditions: the index case remembering and agreeing to notify the partner, the partner being reached, agreeing to take a test, being indeed infected, and moreover the transmission happening during the contact between the index and the partner.
+This explains the seemingly low values.
 
@@
 The BD-PN model is the first step towards accounting for non-random sampling in phylodynamic models.
-In order to reduce its mathematical complexity (having closed-form solutions for its differential equations) and make parameter estimation fast, we made several assumptions: (i) that only detected individuals can notify their partners; (ii) only the most recent partner can get notified; (iii) notified partners are always sampled upon removal (i.e., observed in the transmission tree).
+In order to reduce its mathematical complexity (having closed-form solutions for its differential equations) and make parameter estimation fast, we combined several conditions of the complex partner notification process into one ``successful notification'' probability $\rho$ (see above), and made several additional assumptions: (i) only sampled individuals can notify their partners; (ii) only the most recent partner can get notified; (iii) once notified, partners are always sampled upon removal (however, potentially after the end of the considered sampling period).
 These assumptions may be too simple to fully grasp the real partner notification process and should be relaxed in the future work.
@@
 
-Finally, what we presented here is a specific case of an epidemiological model.
-However, it might be useful in other contexts as well.
-For instance in macroevolution, a detection of a new species might lead to faster detection of closely related ones.
+Finally, what we presented here is a specific case of an epidemiological model. However, it might be useful in other contexts as well.
+For instance, the State Speciation and Extinction (SSE) family of models~\cite{Maddison2007,fitzjohnEstimatingTraitDependentSpeciation2009,goldbergTempoModePlant2012} in the macroevolutionary domain are mathematically very similar to the epidemiological MTBD models~\cite{macphersonUnifyingPhylogeneticBirth2022}.
+In macroevolution, a detection of a new species might lead to faster detection of closely related ones.
 This process could be modeled in the same way as partner notification.
@@
 
-At step 2, to chose the type of the event, we draw a value $v$ (uniformly) from an interval $[0, r[$.
+At step 2, to chose the type of the event, we draw a value $w$ (uniformly) from an interval $[0, r[$.
 This interval can be seen as composed of subintervals corresponding to each possible event, where the width of each subinterval is defined by the corresponding rate and infected individual count, e.g., the subinterval corresponding to a transmission from $k$ to $l$ has a width of $\lambda_{kl}c_k$.
-The subinterval in which $v$ is located defines the event type.
+The subinterval in which $w$ is located defines the event type.
 
@@
 For transmission and sampling events we also record their time and ids of the involved individuals. These values are used to reconstruct the sampled parts of the tree once the simulation is finished.
+
+\smallskip
 
@@
 
-To model the cases when a partner B is not sampled, we will need the probability of a \textbf{hidden partner subtree} $U_p(t, t_A)$, which started at time $t$ on a partner branch and evolved till time $T$ without any of its individuals being sampled, provided that the first of the partner's notifiers tried to notify them at time $t_A$ (B's branch in Fig.~\ref{fig:pn-branches}(2a,c,3a,c)).
-This probability consists of two parts: either the partner got notified at time $t_A$ but has not yet got sampled (like in Fig.~\ref{fig:pn-branches}(2a,3a)), $U_p^{(1)}(t, t_A)$;  or the partner got removed via the standard procedure before the time $t_A$ (like in Fig.~\ref{fig:pn-branches}(2c,3c)), $U_p^{(2)}(t, t_A)$:
+To model the cases when a partner B is not sampled, we need to calculate the probability of a \textbf{hidden partner subtree} $U_p(t, t_A)$, which started at time $t$ on a partner branch and evolved till time $T$ without any of its individuals being sampled, provided that the first of the partner's notifiers tried to notify them at time $t_A$ (B's branch in Fig.~\ref{fig:pn-branches}(2a,c,3a,c)).
+This probability consists of two parts: either the partner got notified at time $t_A$ but has not got sampled by the end of the sampling period $T$ (like in Fig.~\ref{fig:pn-branches}(2a,3a)), $U_p^{(1)}(t, t_A)$;  or the partner got removed via the standard procedure before the time $t_A$ (like in Fig.~\ref{fig:pn-branches}(2c,3c)), $U_p^{(2)}(t, t_A)$:
 \begin{equation}
@@
 Note that $p_{pb}^{(A)}(t)$ describes a partner branch evolution between $t$ and $t_A$ with any number of hidden transmissions, no removal along it, and no event at the end of the branch.
-What we need instead is a removal with no sampling at time $\tau$.
+What we need instead is a removal with no sampling at time $t_B$.
 To approximate $U_p^{(2)}(t, t_A)$ we will divide $p_{pb}^{(A)}(t)$ by the probability of no removal during the time $(t_A - t)$ (i.e., $e^{-\psi(t_A - t)}$), multiply it by the probability of removal during that time ($1 - e^{-\psi(t_A - t)}$), and by the probability of not being sampled upon removal ($1 - \rho$):
@@
 Let us denote the first notification time of the hidden partner as $t_r$, and the time of the hidden partner subtree start as $t_h$, while as usual $t_i$ stands for the time at the branch end.
-For instance, in Fig~\ref{fig:pn-branches}(2a) $t_i=t_C$, $t_r=t_A$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(2b) $t_i=t_r=t_C$, $t_h=t_{BC}$; in Fig~\ref{fig:pn-branches}(3c) $t_i=t_r=t_A$, $t_h=t_{AB}$.
+For instance, in Fig.~\ref{fig:pn-branches}(2a) $t_i=t_C$, $t_r=t_A$, $t_h=t_{BC}$; in Fig.~\ref{fig:pn-branches}(2b) $t_i=t_r=t_C$, $t_h=t_{BC}$; in Fig.~\ref{fig:pn-branches}(3c) $t_i=t_r=t_A$, $t_h=t_{AB}$.
 
@@
 Hence if the branch is internal, the top part can only correspond to the partner, while the bottom part can only correspond to a standard state (unnotified non-notifier), as the bottom part of such a branch corresponds to the partner's recipient (who could not be notified as the partner is unobserved and could not be a notifier as the branch is not external).
-Finally, if the branch is external, and the only partner's notifier corresponds to its tip, then there are two possibilities for its top part: (1a) either it fully corresponds to a standard state (unnotified non-notifier); or (1b) there is another partners' notifier in its supertree, and this other partner is also hidden along the top part of the same branch, hence the top part of the branch is partner-standard (see Fig.~\ref{fig:pn-mixed}).
-If, on the other hand, the branch is external, and the partner's notifier(s) is(are) only found in its supertree, then there are two possibilities for its bottom part: (2a) either it fully corresponds to a standard state (unnotified non-notifier); or (2b) there is another partners' notifier at its tip, and this other partner is also hidden along the bottom part of the same branch, hence the bottom part of the branch is standard-notifier (see Fig.~\ref{fig:pn-mixed}).
+Finally, if the branch is external, and the only partner's notifier corresponds to its tip, then there are two possibilities for its top part: (1a) either it fully corresponds to a standard state (unnotified non-notifier); or (1b) there is another partner's notifier in its supertree, and this other partner is also hidden along the top part of the same branch, hence the top part of the branch is partner-standard (see Fig.~\ref{fig:pn-mixed}).
+If, on the other hand, the branch is external, and the partner's notifier(s) is(are) only found in its supertree, then there are two possibilities for its bottom part: (2a) either it fully corresponds to a standard state (unnotified non-notifier); or (2b) there is another partner's notifier at its tip, and this other partner is also hidden along the bottom part of the same branch, hence the bottom part of the branch is standard-notifier (see Fig.~\ref{fig:pn-mixed}).
 Note that (1b) and (2b) represent the same case but from the point of view of two different hidden partners (B and their notifier A for the former, and D and their notifier C for the latter in Fig.~\ref{fig:pn-mixed}).
@@
 \includegraphics[width=0.3\textwidth]{Fig_mixed}
-\caption{An example of mixed tip branch with two hidden partner subtrees: an individual C got sampled at time $t_C$ and notified their most recent partner D.
+\caption{\textbf{An example of mixed tip branch with two hidden partner subtrees}: an individual C got sampled at time $t_C$ and notified their most recent partner D.
 D however, stayed unobserved as D did not get sampled by the end of the sampling period (time $T$).
-Before notification (at time $t_C$) D transmitted two times, the first transmission leading to an unobserved recipient subtree (gray dotted triangle) and the second (DB) leading to an observed recipient subtree (which includes a sampled individual A and a hidden individual B).
-The D's recipient (or someone in their subtree if there were hidden transmissions along the path DB-AB) transmitted further (transmission AB).
+Before notification (at time $t_C$) D transmitted two times, the latter transmission leading to an unobserved recipient subtree (gray dotted triangle) and the former transmission (DB) leading to an observed recipient subtree (which includes a sampled individual A and a hidden individual B).
+The D's recipient A transmitted further (transmission AB).
 Once A got sampled at time $t_A$, they notified their most recent partner B, who however was already removed by then via the standard procedure without sampling (at time $t_B < t_A$).
@@
 
-%if the branch is external the bottom part of the branch can either correspond to a notifier (as in Fig~\ref{fig:pn-branches}(2b,3)), or to a standard branch, or to a mixed tip branch starting as a standard branch and finishing as a notifier.
+%if the branch is external the bottom part of the branch can either correspond to a notifier (as in Fig.~\ref{fig:pn-branches}(2b,3)), or to a standard branch, or to a mixed tip branch starting as a standard branch and finishing as a notifier.
 It cannot correspond to another partner branch as every partner has their notifier sampled, and notifiers only notify one partner (who is already hidden along this branch).
@@
 
-We will first describe the one-hidden-partner case $p_{m(s1,s2)}^{(i)}(t,t_r)$ (where $s1 \in \{p, -\}$, $s2 \in \{n, -\}$ and $-$ stands for the standard (unnotified non-notifier) state), and then proceed with the two-partner case $p_{m(p,-,n)}^{(i)}(t,t_{r})$, where $t_{r}$ is the earliest notification time for the partner hidden at the top of the branch (e.g., $t_r=t_C$ in Fig.~\ref{fig:pn-mixed}), the only notification time for the partner hidden at the bottom of the branch being $t_i$  (e.g., $t_i=t_A$ in Fig.~\ref{fig:pn-mixed}).
+We will first describe the one-hidden-partner case $p_{m(s1,s2)}^{(i)}(t,t_r)$ (where $s1 \in \{p, -\}$, $s2 \in \{n, -\}$ and $-$ stands for the standard (unnotified non-notifier) state), and then proceed with the two-hidden-partners case $p_{m(p,-,n)}^{(i)}(t,t_{r})$, where $t_{r}$ is the earliest notification time for the partner hidden at the top of the branch (e.g., $t_r=t_C$ in Fig.~\ref{fig:pn-mixed}), the only notification time for the partner hidden at the bottom of the branch being $t_i$  (e.g., $t_i=t_A$ in Fig.~\ref{fig:pn-mixed}).
 
@@
 In our case there must be at least one hidden transmission: to/from the partner at time $t_h$ (where the partner tree stayed unsampled).
-If we remove the probability of no event during this time ($e^{-(\lambda + \psi)(t_h - t)}$) from $p^{(h)}(t)$, we will obtain a probability of a branch with at least one hidden transmission somewhere between $t$ and $t_h$ (the probability of the corresponding hidden tree being included).
+If we remove the probability of no event during this time ($e^{-(\lambda + \psi)(t_h - t)}$) from $p^{(h)}(t)$, we  obtain a probability of a branch with at least one hidden transmission somewhere between $t$ and $t_h$ (the probability of the corresponding hidden tree being included).
 $p_{top=-}^{(h)}(t)$ expresses almost the same thing, with the difference being that the last hidden transmission must have happened at time $t_h$ and correspond to the hidden partner.
@@
 
-\subsubsection*{Tree likelihood calculation with a pruning algorithm} 
+\subsubsection*{Tree likelihood calculation with a pruning algorithm and its time complexity} 
 Note that $l^{(n)}(i, r)$, $l^{(n,n)}(i, r)$ and $l^{(n,\not{n})}(i, r)$ depend on the notifier $r$ and the notification time $t_r$.
@@
  Then we perform a postorder tree traversal, where for each node $i$ we calculate $l^{(\not{n}, \not{n})}(i)$, $l^{(\not{n}, n)}(i)$, $l^{(n, \not{n})}(i, r)$ and $l^{(n, n)}(i, r)$ $\forall r \in N_i$ if it is a tip, or $l^{(\not{n})}(i)$ and $l^{(n)}(i, r)$ $\forall r \in N_i$ if it is internal.
-This part takes O(R + N) time, where O(N) accounts for calculation of $l^{(\not{n})}(i)$ for all the $N - 1$ internal tree nodes and of $l^{(\not{n}, \not{n})}(i)$ and $l^{(\not{n}, n)}(i)$ for $N$ tips, while R is the sum of numbers of potential notifiers over all tree nodes and O(R) accounts for calculation of $l^{(n)}(i, r)$, $l^{(n, \not{n})}(i, r)$ and $l^{(n, n)}(i, r)$.
+This part takes O(R + N) time, where O(N) accounts for calculation of $l^{(\not{n})}(i)$ for all the $N - 1$ internal tree nodes and of $l^{(\not{n}, \not{n})}(i)$ and $l^{(\not{n}, n)}(i)$ for $N$ tips, while R is the sum of numbers of potential notifiers over all tree nodes and O(R) accounts for calculation of $l^{(n)}(i, r)$, $l^{(n, \not{n})}(i, r)$ and $l^{(n, n)}(i, r)$ $\forall i\; \forall r$.
 
 For the best case (a perfectly balanced tree, where each tip is in a cherry), the number of potential notifiers is one for the tips (i.e., the sister tip) and zero for the internal nodes.
-Hence $R_b = N$ and the likelihood calculation time is linear: O(N).
-The worst case  is represented by a caterpillar tree.
+Hence $R_{balanced} = N$ and the likelihood calculation time is linear: O(N).
+The worst case  is represented by a caterpillar tree (where each internal node has a tip child).
 Each internal node has $k$ potential notifiers (tips in its supertree), where $k$ is the node's depth in the tree, the root's depth being 0 and the deepest internal node's depth being $N-2$.
-Moreover each tip has as many notifiers as its parent node, apart from the two deepest tips who have $N-1$ notifier each instead of $N-2$ as they are in a cherry and hence could have notified each other).
-This hence sums up to $R_c = 2\big(1 + 2 + \ldots + (N-3)\big) + 2(N-1) = N^2 - 2N + 2$ and the time is hence quadratic: $O(N^2)$.
-On our BDPN simulated dataset of 100 trees with N=500-1\,000 tips each, $R \in [2N, 38N]$ with a median at $7N$, suggesting a likelihood calculation time close to the linear one.
+Moreover each tip has as many notifiers as its parent node, apart from the two deepest tips who have $N-1$ notifiers each instead of $N-2$ as they are in a cherry and hence could have notified each other.
+This hence sums up to $R_{caterpillar} = 2\big(1 + 2 + \ldots + (N-3)\big) + 2(N-1) = N^2 - 2N + 2$ and the time is hence quadratic: $O(N^2)$.
+On our BD-PN simulated dataset of 100 trees with N=500-1\,000 tips each, $R \in [2N, 38N]$ with a median at $7N$, suggesting a likelihood calculation time close to the linear one.
 

### 2024-09-05 article/article_plos.tex
minor

@@
 Under this model we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ (see \nameref{math}).
-The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g., by finding parameter values that maximize it, in a maximum-likelihood framework).
+The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g., by finding parameter values that maximize it, in the maximum-likelihood framework).
 In particular, it permits the estimation of the following epidemiological parameters:
@@
 extension of the MTBD models that adds partner notification (PN).  
-In the partner notification process, an identified infected index case notifies their potentially infected contacts (e.g., sexual or drug injection partners for HIV), either on their own (patient referral) or with a help of a health service provider (provider referral, when a health adviser notifies the partners without identifying the index patient; or contract referral, when if the index case fails to notify partners within a specified time period, the health adviser proceeds to provider referral)~\cite{wardPartnerNotification2014}.
+In the partner notification process, an identified infected index patient notifies their potentially infected contacts (e.g., sexual or drug injection partners for HIV), either on their own (patient referral) or with the help of a health service provider (provider referral, when a health adviser notifies the partners without identifying the index patient; or contract referral, when if the index patient fails to notify partners within a specified time period, the health adviser proceeds to provider referral)~\cite{wardPartnerNotification2014}.
 The notified partners are then provided with appropriate tests and treatment.
-Therefore, a successful detection of a new case B via partner notification by an index case A implies several conditions to be met: (i) A remembers the contact with B and  either notifies B on their own or gives the B's contacts to a health advisor for assisted notification; (ii) B is successfully reached and notified; (iii) B agrees to take a test; (iv) B is indeed infected.
+Therefore, a successful detection of a new case B via partner notification by an index patient A implies several conditions to be met: (i) A remembers the contact with B and  either notifies B on their own or gives the B's contacts to a health advisor for assisted notification; (ii) B is successfully reached and notified; (iii) B agrees to take a test; (iv) B is indeed infected.
 For instance, according to the report from the Public Health England~\cite{aHIVUnitedKingdom2014}, in England in 2013 for HIV, reaching the stage (ii) had an estimated probability of about 50\%; reaching the stage (iii) -- of about 41\% (as 83\% of all the patients who attended a clinic through PN had a reported HIV test); and the stage (iv) -- of less than 3\%.
@@
  \item $\upsilon$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability.
-This leads to the removal rate change for the partner branch in the transmission tree;
+This leads to a removal rate change for the partner branch in the transmission tree;
  \item $\phi >> \psi$ -- constant notified sampling rate: a notified individual (if not yet removed via the standard procedure) gets removed \textit{and observed} with this rate.
@@
 %The results with the becoming-uninfectious rate $\psi$ fixed are similar and given in the Supplementary Material.
-%As expected with a maximum-likelihood estimator, 
-The tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the true parameter values for all the trees in both settings.
+As expected, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the true parameter values for all the trees in both settings.
+This shows the consistent behavior of our likelihood calculation and optimization.
 
@@
 
-The PN test detected partner notification in this tree (p-value $=0.014$). Our bdpn estimator inferred the partner notification probability $\upsilon=0.09$ CI $[0.008-0.226]$,  
+The PN test detected partner notification in this tree (p-value $=0.014$). Our bdpn estimator inferred the partner notification probability $\upsilon=0.09$ 95\%-CI $[0.008-0.226]$,  
 $R_e = 1.43$ $[1.26-1.66]$, infectious time $\frac{1}{\psi} = 8$ $[6-11]$ years, and 4.9 months [6 days -- 1.4 years] between partner notification and their sampling.
@@
 
-For comparison, parameter estimation on the same tree with the BD model gave a slightly lower estimate of $R_e$ and a slightly shorter infectious time, though with intersecting CIs ($R_e = 1.32$ $[1.19-1.47]$, $\frac{1}{\psi} = 7$ $[6-9]$ years), which is in line with underestimation of $R_e$ and $\frac{1}{\psi}$ with bd on the BD-PN simulated data.
+For comparison, parameter estimation on the same tree with the BD model gave a slightly lower estimate of $R_e$ and a slightly shorter infectious time, though with intersecting CIs ($R_e = 1.32$ $[1.19-1.47]$, $\frac{1}{\psi} = 7$ $[6-9]$ years), which is in line with underestimation of $R_e$ and $\frac{1}{\psi}$ with the bd estimator on the BD-PN simulated data.
 The likelihood ratio test suggested the BD-PN model: loglikelihood under the BD-PN-estimated parameters was $l_{BDPN} = -1\,140$ vs.
@@
 According to the HIV report from the Public Health England~\cite{aHIVUnitedKingdom2014}, in England in 2013, less than 3\% of sampled infected individuals were detected though partner notification.
-Our estimate is even lower, potentially because as our model does not allow for notifying partners during the contact with whom there was no transmission, nor notifying multiple partners.
+Our estimate is even lower, potentially because our model does not allow for notifying partners with whom there was no transmission, nor notifying multiple partners.
 
@@
 We estimated a modest partner notification probability ($9\%$ in Zurich, $1\%$ in the UK), and rapid notified partner detection ($\approx 5$ months vs $\approx  8$ years for detection under the standard procedure in Zurich; and $\approx 8$ days vs $\approx 2.3$ years for detection under the standard procedure in the UK).
-Note that the notification probability represents ``successful notification'', which combines several conditions: the index case remembering and agreeing to notify the partner, the partner being reached, agreeing to take a test, being indeed infected, and moreover the transmission happening during the contact between the index and the partner.
+Note that the notification probability represents ``successful notification'', which combines several conditions: the index patient remembering and agreeing to notify the partner, the partner being reached, agreeing to take a test, being indeed infected, and moreover the transmission happening during the contact between the index and the partner.
 This explains the seemingly low values.
@@
 The BD-PN model is the first step towards accounting for non-random sampling in phylodynamic models.
-In order to reduce its mathematical complexity (having closed-form solutions for its differential equations) and make parameter estimation fast, we combined several conditions of the complex partner notification process into one ``successful notification'' probability $\rho$ (see above), and made several additional assumptions: (i) only sampled individuals can notify their partners; (ii) only the most recent partner can get notified; (iii) once notified, partners are always sampled upon removal (however, potentially after the end of the considered sampling period).
+In order to reduce its mathematical complexity (having closed-form solutions for its differential equations) and make parameter estimation fast, we combined several conditions of the complex partner notification process into one ``successful notification'' probability $\upsilon$ (see above), and made several additional assumptions: (i) only sampled individuals can notify their partners; (ii) only the most recent partner can get notified; (iii) once notified, partners are always sampled upon removal (however, potentially after the end of the considered sampling period).
 These assumptions may be too simple to fully grasp the real partner notification process and should be relaxed in the future work.
@@
 Finally, what we presented here is a specific case of an epidemiological model. However, it might be useful in other contexts as well.
-For instance, the State Speciation and Extinction (SSE) family of models~\cite{Maddison2007,fitzjohnEstimatingTraitDependentSpeciation2009,goldbergTempoModePlant2012} in the macroevolutionary domain are mathematically very similar to the epidemiological MTBD models~\cite{macphersonUnifyingPhylogeneticBirth2022}.
+For instance, the State Speciation and Extinction (SSE) family of models~\cite{Maddison2007,fitzjohnEstimatingTraitDependentSpeciation2009,goldbergTempoModePlant2012} in the macroevolution domain are mathematically very similar to the epidemiological MTBD models~\cite{macphersonUnifyingPhylogeneticBirth2022}.
 In macroevolution, a detection of a new species might lead to faster detection of closely related ones.
@@
 \section*{Acknowledgments}
-The authors would like to thank Dr Mathieu Moslonka-Lefebvre who initiated this project by proposing to model partner notification.
+The authors would like to thank Dr Mathieu Moslonka-Lefebvre who initiated this project by proposing to study partner notification.
 He worked on a different version of the PN model (which is not part of this manuscript) and on the initial version of the PN test.
@@
 We would also like to thank the HPC cluster team in Institut Pasteur for their support with computer simulations.
+O.G.
+was supported by the Paris Artificial Intelligence Research Institute (PRAIRIE, ANR-19-P3IA-0001).
 
